### 2012-10-06 paper/paper.tex
Initial content.

\documentclass{llncs}
%
\begin{document}
\title{Optimizing Abstract$^2$ Machines}
%
\frontmatter          % for the preliminaries
%
\author{J.~Ian Johnson\inst{1} \and Matthew Might\inst{2} \and David Van Horn\inst{1}}
\institute{Northeastern University \and University of Utah}

\maketitle              % typeset the title of the contribution

\begin{abstract}
The abstracting abstract machines approach to program analysis makes
it easy to design and implement sound and computable abstract
interpretations of higher-order programs.
%
However, the straightforward implementation is generally
not very fast.
%
This article contributes a step by step process for going from a naive
abstract$^2$ machine to an efficient program analyzer that is two
orders of magnitude faster than the starting point, yet just as
precise.
\end{abstract}
%
\section{Introduction}

\section{Starting point}
\subsection{An abstract$^2$ machine for PCF}
\subsection{An example: Analyzing Church number computations}
\subsection{Store widening}
\subsection{Baseline evaluation}

\section{Precision preserving recipe}
\subsection{Fixed-point specialization}
\subsection{Lazy non-determinism}
\subsection{Abstract compilation}
\subsection{Computing with store changes}

\section{Evaluation}

\section{Related work}

\section{Conclusion}

Abstract machines are not only a good model for rapid analysis
development, they also can be systematically developed into efficient
algorithms.

\end{document}

### 2012-10-06 paper/paper.tex
Initial numbers in outline.

@@
 This article contributes a step by step process for going from a naive
-abstract$^2$ machine to an efficient program analyzer that is two
-orders of magnitude faster than the starting point, yet just as
-precise.
+abstract$^2$ machine to an efficient program analyzer that is (nearly)
+three orders of magnitude faster than the starting point, yet just as
+precise (and purely functional).
 \end{abstract}
@@
 \subsection{Store widening}
+
 \subsection{Baseline evaluation}
 
+Wide Store: cpu time: 551571 real time: 571319 gc time: 4003
+
 \section{Precision preserving recipe}
+
+\subsection{Lazy non-determinism}
+
+Lazy:
+   cpu time: 32481 real time: 32881 gc time: 547
+
+\subsection{Abstract compilation}
+
+Compile:
+   cpu time: 255397 real time: 261532 gc time: 2947
+
+\noindent
+Compile + Lazy:
+   cpu time: 31173 real time: 31642 gc time: 739
+
 \subsection{Fixed-point specialization}
-\subsection{Lazy non-determinism}
-\subsection{Abstract compilation}
+
+%% Should be commutted to beginning
+
+Special + Compile + Lazy:
+   cpu time: 14212 real time: 14681 gc time: 823
+
 \subsection{Computing with store changes}
 
+Delta Store + Special + Compile + Lazy:
+   cpu time: 668 real time: 686 gc time: 41
+
 \section{Evaluation}
+
+Other benchmarks
 

### 2012-10-07 paper/paper.tex
Progress on initial formalism in paper.

@@
 \documentclass{llncs}
+\usepackage{alltt}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{xltxtra}
+\setmonofont[Scale=MatchLowercase]{DejaVu Sans Mono}
 %
+\input{preamble}
 \begin{document}
@@
 
+\subsection{Notation and prerequisites}
+
+Prereqs: Semantics Engineering, AAM.
+
+Notation: concrete examples of programs under analysis is given in
+(monochrome) Scheme notation.  Code is given in (syntax colored)
+Racket.
+
+\newpage
 \section{Starting point}
-\subsection{An abstract$^2$ machine for PCF}
+\subsection{An abstract$^2$ machine for ISWIM}
+
+\begin{figure}
+\begin{align*}
+\ev{\svar{x},\rho,\kappa,\sigma} &\machstep
+\co{\kappa,v,\sigma}
+\text{ where }v \in \msto(\menv(x))
+\\
+\ev{\snum{n},\menv,\mkont,\msto} &\machstep
+\co{\mkont,n,\msto}
+\\
+\ev{\slam\mvar\mexp,\menv,\mkont,\msto} &\machstep
+\co{\mkont,\clos{\mvar,\mexp,\menv}}
+\\
+%% \ev{\srec\mvarf\mvar\mexp,\menv,\mkont,\msto} &\machstep
+%% \co{\mkont,\rlos{\mvarf,\mvar,\mexp,\menv}}
+%% \\
+\ev{\sapp{\mexpi0}{\mexpi1},\menv,\mkont,\msto} &\machstep
+\ev{\mexpi{0},\menv,\kar{\mexpi{1},\menv,a},\msto'}
+\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
+\\
+\ev{\sif{\mexpi0}{\mexpi1}{\mexpi2},\menv,\mkont,\msto} &\machstep
+\ev{\mexpi0,\menv,\kif{\mexpi1,\mexpi2,\menv,\maddr},\msto'}
+\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
+\\
+\ev{\sopu\mop\mexp,\menv,\mkont,\msto} &\machstep
+\ev{\mexp,\menv,\kuop{\mop,\maddr},\msto'}
+\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
+%%\\
+%% \ev{\sopb\mop{\mexpi0}{\mexpi1},\menv,\mkont,\msto} &\machstep
+%% \ev{\mexpi0,\menv,\kbopa{\mexpi1,\menv,\maddr},\msto'}
+%% \text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
+\\[2mm]
+\co{\kmt,\mval,\msto} &\machstep
+\ans{\msto,\mval}
+\\
+\co{\kar{\mexp,\menv,\maddr},\mval,\msto} & \machstep
+\ev{\mexp,\menv,\kfn{\mval,\maddr},\msto}
+\\
+\co{\kfn{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
+\ap{\mval,\mvalx{u},\mkont,\msto}
+\text{ where }\mkont \in \msto(\maddr)
+\\
+\co{\kif{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
+\ev{\mexpi0,\menv,\mkont,\msto}
+\text{ where }\mkont\in\msto(\maddr)
+\\
+\co{\kif{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
+\ev{\mexpi1,\menv,\mkont,\msto}
+\text{ where }\mkont\in\msto(\maddr)
+\\
+\co{\kuop{\mop,\maddr},\mval,\msto} & \machstep
+\co{\mkont,\mval',\msto}
+\text{ where }\mkont\in\msto(\maddr)
+\text{ and } \mval'\in\delta(\mop,\mval)
+%% \\
+%% \co{\kbopa{\mop,\mexp,\menv,\maddr},\mval,\msto} & \machstep
+%% \ev{\mexp,\menv,\kbopb{\mop,\mval,\maddr},\msto}
+%% \\
+%% \co{\kbopb{\mop,\mvalx{u},\maddr},\mval,\msto} & \machstep
+%% \co{\mkont,\mval',\msto}
+%% \text{ where }\mkont=\msto(\maddr)
+%% \text{ and } \mval'=\delta(\mop,\mval,\mvalx{u})
+\\[2mm]
+\ap{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \machstep
+\ev{\mexp,\menv',\mkont,\msto'}
+\text{ where }\menv',\msto' = \mathit{bind}(\msto,\mvar,\mval)
+\\
+%% \ap{\rlos{\mvarf,\mvar,\mexp,\menv},\mval,\mkont} & \machstep
+%% \ev{\mexp,\menv',\mkont,\msto'}
+%% \text{ where }\menv',\msto' = \mathit{bind}(\msto,\langle\mvar,\mval\rangle,\langle\mvarf,
+%% \rlos{\mvarf,\mvar,\mexp,\menv}\rangle)
+\end{align*}
+\caption{Abstract$^2$ machine for ISWIM}
+\label{fig:aam}
+\end{figure}
+
+\paragraph{Concrete interpretation:}
+
+To obtain a concrete interpreter from the relation given in
+Figure~\ref{fig:aam}, set the implicit parameters as follows:
+\begin{gather}
+\begin{align*}
+\mathit{push}(l,\msto,\mkont) &= \maddr,\sigma[\maddr\mapsto\{\mkont\}]
+\mbox{ where }\maddr \notin\msto
+\\
+\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\sigma[\maddr\mapsto\{\mval\}]
+\end{align*}
+\\[2mm]
+\begin{align*}
+\mnum+1 &\in \delta(\saddone,\mnum) &
+\mnum-1 &\in \delta(\ssubone,\mnum)\\
+\strue &\in \delta(\szerohuh,\szero) &
+\sfalse &\in \delta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+\end{align*}
+\end{gather}
+
+\paragraph{Abstract interpretation:}
+
+To obtain an abstract interpreter from the relation given in
+Figure~\ref{fig:aam}, set the implicit parameters as follows:
+\begin{gather}
+\begin{align*}
+\widehat{\mathit{push}}(l,\msto,\mkont) &= l,\sigma\sqcup[l\mapsto\{\mkont\}]
+\mbox{ where }\maddr \notin\msto
+\\
+\widehat{\mathit{bind}}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
+\end{align*}
+\\[2mm]
+\begin{align*}
+\sNum &\in \hat\delta(\saddone,\mnum) &
+\sNum &\in \hat\delta(\ssubone,\mnum)\\
+\strue &\in \hat\delta(\szerohuh,\sNum) &
+\sfalse &\in \hat\delta(\szerohuh,\sNum)\\
+\strue &\in \hat\delta(\szerohuh,\szero) &
+\sfalse &\in \hat\delta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+\end{align*}
+\end{gather}
+
+\begin{figure}
+\begin{alltt}
+  ;; State \(\rightarrow\) Setof State
+  (define (step state)
+    (match state
+      [(ev σ e ρ k)
+       (match e
+         [(var l x)           (for/set ((v (lookup ρ σ x))) (co σ k v))]
+         [(num l n)           (set (co σ k n))]
+         [(bln l b)           (set (co σ k b))]
+         [(lam l x e)         (set (co σ k (clos l x e ρ)))]
+         [(rec f (lam l x e)) (set (co σ k (rlos l f x e ρ)))]
+         [(app l f e)
+          (define-values (σ* a) (push state))
+          (set (ev σ* f ρ (ar e ρ a)))]
+         [(ife l e0 e1 e2)
+          (define-values (σ* a) (push state))
+          (set (ev σ* e0 ρ (ifk e1 e2 ρ a)))]
+         [(1op l o e)
+          (define-values (σ* a) (push state))
+          (set (ev σ* e ρ (1opk o a)))])]
+      [(co σ k v)
+       (match k
+         ['mt (set (ans σ v))]
+         [(ar e ρ l) (set (ev σ e ρ (fn v l)))]
+         [(fn f l)   (for/set ((k (get-cont σ l))) (ap σ f v k))]
+         [(fi c a ρ l)
+          (for/set ((k (get-cont σ l)))
+            (ev σ (if v c a) ρ k))]
+         [(oa o l)
+          (for*/set ((k (get-cont σ l))
+                     (v (delta o (list v))))
+            (co σ k v))]
+      [(ap σ fun a k)
+       (match fun
+         [(clos l x e ρ)
+          (define-values (ρ* σ*) (bind state))
+          (set (ev σ* e ρ* k))]
+         [_ (set)])]
+\end{alltt}
+\caption{Implementation of machine transition relation.}
+\end{figure}
+
 \subsection{An example: Analyzing Church number computations}
+
+\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
+
+\begin{alltt}
+;; multiplication distributes over addition
+((church=? ((mult \church2) ((plus \church1) \church3)))
+ ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))))
+\end{alltt}
+
+Where \syntax{church=?} is an equality function for Church numerals
+written in terms of recursion and an iterative \syntax{pred} function.
+The \syntax{plus} and \syntax{mult} functions are written as usual.
+
+This program reaches $N$ states under concrete interpretation.
+
 \subsection{Store widening}

### 2012-10-12 paper/paper.tex
Filling in bits of paper.

@@
 
-\paragraph{Concrete interpretation:}
-
-To obtain a concrete interpreter from the relation given in
-Figure~\ref{fig:aam}, set the implicit parameters as follows:
-\begin{gather}
+\paragraph{Concrete interpretation} can be characterized by setting the implicit
+parameters of the relation given in Figure~\ref{fig:aam} as follows:
 \begin{align*}
-\mathit{push}(l,\msto,\mkont) &= \maddr,\sigma[\maddr\mapsto\{\mkont\}]
+\mathit{push}(l,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
 \mbox{ where }\maddr \notin\msto
 \\
-\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\sigma[\maddr\mapsto\{\mval\}]
+\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
 \end{align*}
-\\[2mm]
+The resulting relation is non-deterministic in its choice of
+addresses, however it must always choose a fresh address when
+allocating a continuation or variable binding.  If we consider machine
+states equivalent up to consistent renaming, this relation defines
+a deterministic machine.  (The relation is really a function.)
+
+\paragraph{Abstract interpretation} can be characterized by setting the implicit
+parameters just as above, but dropping the $\maddr \not\in \msto$
+condition.  This family of interpereters is also non-deterministic in
+choices of addresses, but it is free to choose addresses that are
+already in use.  Consequently, the machines may be non-deterministic
+when mutiple values reside in a store location.
+
+It is important to recognize from this definition that \emph{any}
+allocation strategy is an abstract interpretation.  In particular,
+concrete intepretation is a kind of abstract interpretation.  So is an
+interpretation that allocates a single cell into which all bindings
+and continuations are stored.  On the one hand is an abstract
+intepretation that is non-computable and gives only the ground truth
+of a programs behavior; on the other is an abstract interpretation
+that is easy to compute but gives little information.  Useful program
+analyses lay somewhere in between and can be characterized by their
+choice of address representation and allocation strategy.
+Uniform \(k\)-CFA is one such analysis.
+
+\paragraph{Uniform \(k\)-CFA} can be characterized by the following allocation
+strategy:
+
+\begin{align*}
+\mathit{push} &= \dots\\
+\mathit{bind} &= \dots
+\end{align*}
+
+\paragraph{Primitives}
+
+%% \begin{align*}
+%% \widehat{\mathit{push}}(\maddr,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
+%% % \mbox{ where }\maddr \notin\msto
+%% \\
+%% \widehat{\mathit{bind}}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
+%% \end{align*}
+
 \begin{align*}
@@
 \end{align*}
-\end{gather}
 
-\paragraph{Abstract interpretation:}
-
-To obtain an abstract interpreter from the relation given in
-Figure~\ref{fig:aam}, set the implicit parameters as follows:
-\begin{gather}
-\begin{align*}
-\widehat{\mathit{push}}(l,\msto,\mkont) &= l,\sigma\sqcup[l\mapsto\{\mkont\}]
-\mbox{ where }\maddr \notin\msto
-\\
-\widehat{\mathit{bind}}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
-\end{align*}
-\\[2mm]
 \begin{align*}
@@
 \end{align*}
-\end{gather}
 
@@
 
+We can eliminate the {\tt ev} states from the run-time interpretation
+of a program by specializing the machine transition relation to the
+program being analyzed.  This eliminates interpretative overhead by
+first compiling the program into ``byte code'' instructions.
+
+The essence of the compilation effect can be seen by consider an example
+such as
+\[
+\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
+\]
+which makes the following transitions:
+\begin{align}
+& \ev{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\mkont,\msto_0}\\
+\machstep\; &
+\ev{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\kar{\mexp_3,\menv,\maddr_1},\msto_1}
+\\
+\machstep\; &
+\ev{\sapp\mvar{\mexp_1},\menv,\kar{\mexp_2,\menv,\maddr_2},\msto_2}
+\\
+\machstep\; &
+\ev{\mvar, \menv,\kar{\mexp_1,\menv,\maddr_3},\msto_3} % {\mexp_2}
+\\
+\machstep\; &
+\co{\kar{\mexp_1,\menv},\mval,\msto_4} % {\mexp_1}{\mexp_2}
+\mbox{ where } \mval \in \msto(\menv(\maddr))
+\end{align}
+
+where $\msto_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
+[\maddr_2 \mapsto \kar{\mexp_3,\menv,\maddr_1}]
+[\maddr_3 \mapsto \kar{\mexp_2,\menv,\maddr_2}]$.
+
+Notice that the only difference between concrete and abstract interpretation
+is which addresses are pushed.
+
 Compile:
@@
 
+\appendix
+\section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
+
+This machine's allocation strategy mimics the Uniform k-CFA analysis
+in Principles of Program Analysis, which is defined in terms of
+``$\delta$ contours.''  However, because the machine represetation makes
+context explicit via continuations, we can calculate these contours
+rather than thread them throught the evaluator.  In other words, we
+can use the CESK* machine without modification to obtain Uniform k-CFA
+by way of a simple allocation strategy.  (In this way, it's a
+simplification of the presentation in JFP.)
+
+NNH uses a coinductive acceptability relation to specify Uniform
+k-CFA:
+
+\[
+   C,R \models^{ce}_\delta E
+\]
+
+The cache and global environment form a finite store-like structure
+holding bindings and return values.  The contour environment ce maps
+variables to locations in R which contains their bindings, just as the
+environment of the CESK* machine does.  The current contour delta is a
+string of application labels describing the enclosing context under
+which this term is being analyzed (or evaluated).  If you view the
+acceptability relation as a big-step evalator, the
+$(\widehat C,\widehat\rho)$ component should be seen as a global
+store ce is the environment mapping variables to their locations.
+
+Starting form the initial configuration for a program and iterating
+the machine transition relation until reaching a fixpoint of reachable
+states will \emph{underestimate} the acceptability relation of Uniform
+k-CFA.  You can recover acceptability by feeding this store back into
+the initial configuration and iterating again.  Repeating this process
+until a complete run of the program reaches no new states will be the
+least solution that is acceptable.
+
+HOWEVER.  Why should we care about acceptability?  What this
+machine computes is safe.  In other words, it computes a more
+precise characterization of the run-time behavior of a program.  In
+doing is so, it actually saves work (as can be seen above).
+
+An Example:
+
+\begin{verbatim}
+ (let ((id (lambda (x) x)))
+   (begin (id 1) (id 2)))
+\end{verbatim}
+
+Under Uniform 0-CFA, we would have:
+\[
+   [{\tt x} \mapsto \{{\tt 1}, {\tt 2}\}] \in \widehat\rho
+\]
+
+in the least solution to $\models$.  This says that, when run, 'x' is
+bound to 1 or 2.
+
+Under the machine semantics using a 0CFA allocation policy, the trace
+semantics of the machine show that x is bound to 1, and that at some
+later point, x becomes bound to 1 or 2.  Moreover, the machine would
+show that (id 1) evaluates to 1 and only 1, while Uniform 0CFA must
+give that (id 1) is either 1 or 2 to be acceptable.  We don't see any
+value in these kinds of false flows that are due to the global and
+timeless aspects of C,R which acceptability requires the heap to be
+both finite and unchanging over the course of abstract
+interpretation. (Another view of the difference: the machine abstracts
+a program's execution as a \emph{finite state machine} that mimics the
+machine interpretation of the program; the aceptability relation of
+Uniform \(k\)-CFA abstracts a program's execution as a \emph{finite
+  map} that mimics the big-step evaluator: from terms to (sets of)
+values.)
+
+\section{Pushdown Analysis}
+
+It is straightforward to instantiate a \emph{pushdown} abstraction by
+bounding only the variable binding portion of the heap, but using a
+unique allocation strategy for continuations.  Such a strategy
+abstracts a program's execution as a \emph{pushdown automata}
+that mimics the machine interpretation.  This strategy therefore
+models the abstract stack in a true stack like fashion and always
+properly matches function calls with their return.
+
+Although such analyses can be formulated straightforwardly in the
+abstract machine approach, it is not clear all of the techniques of
+this paper can be applied to similar effect in the pushdown context.
+The main problem is calculation of an analysis can no longer be
+computed as the fixed point of the machine transition relation.
+Although there are several implementations (CFA2,ICFP'12), they
+operate at speeds roughly on par with our starting point: unoptimized
+store widened machines.
+
 \end{document}

### 2012-10-14 paper/paper.tex
Setting up definitions.

@@
 
+ISWIM is a family of programming languages parameterized by a set of
+base values and operations.  To make things concrete, we consider a
+member of the ISWIM family with integers, booleans, and a few
+operations.
+
+\begin{figure}
+\[
+\begin{array}{l@{\qquad}rcl}
+\text{Expressions} & \mathit{e} &=& \svar[^\mlab]\mvar\\
+&&|& \slit[^\mlab]\mlit\\
+&&|& \slam[^\mlab]\mvar\mexp\\
+&&|& \sapp[^\mlab]\mexp\mexp \\
+&&|& \sif[^\mlab]\mexp\mexp\mexp \\
+\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
+\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
+\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
+\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
+\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
+\end{array}
+\]
+\caption{Syntax of ISWIM}
+\label{fig:syntax}
+\end{figure}
+
+Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.
+Figure~\ref{fig:aam} defines the semantics of ISWIM as a machine
+model.  Evaluation is defined as the set of states reachable by the
+reflexive, transitive closure of the machine transition relation.  The
+machine is a very slight variation on a standard abstract machine for
+ISWIM in ``eval, continue, apply'' form.
+
+Compared with the standard machine semantics, this definition is
+different in the following ways:
+\begin{itemize}
+\item the store maps addresses to \emph{sets} of values,
+\item continuations are heap-allocated,
+\item there are ``contour values'' (written $\mcntr$) and syntax
+  labels ($\mlab$) threaded through the computation, and
+\item the machine is implicity parameterized by the functions
+  $\mathit{push}$, $\mathit{bind}$, and $\interpdelta$.
+\end{itemize}
+
 \begin{figure}
 \begin{align*}
-\ev{\svar{x},\rho,\kappa,\sigma} &\machstep
-\co{\kappa,v,\sigma}
-\text{ where }v \in \msto(\menv(x))
+\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\epsilon}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
+\\[2mm]
+%% EVAL
+\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
+\co{\mkont,\mval,\msto}
+\text{ where }\mval \in \msto(\menv(\mvar))
 \\
-\ev{\snum{n},\menv,\mkont,\msto} &\machstep
-\co{\mkont,n,\msto}
+\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
+\co{\mkont,\mlit,\msto}
 \\
-\ev{\slam\mvar\mexp,\menv,\mkont,\msto} &\machstep
-\co{\mkont,\clos{\mvar,\mexp,\menv}}
+\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
+\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
 \\
-%% \ev{\srec\mvarf\mvar\mexp,\menv,\mkont,\msto} &\machstep
-%% \co{\mkont,\rlos{\mvarf,\mvar,\mexp,\menv}}
-%% \\
-\ev{\sapp{\mexpi0}{\mexpi1},\menv,\mkont,\msto} &\machstep
-\ev{\mexpi{0},\menv,\kar{\mexpi{1},\menv,a},\msto'}
-\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
+\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
+\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
+\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
 \\
-\ev{\sif{\mexpi0}{\mexpi1}{\mexpi2},\menv,\mkont,\msto} &\machstep
-\ev{\mexpi0,\menv,\kif{\mexpi1,\mexpi2,\menv,\maddr},\msto'}
-\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
-\\
-\ev{\sopu\mop\mexp,\menv,\mkont,\msto} &\machstep
-\ev{\mexp,\menv,\kuop{\mop,\maddr},\msto'}
-\text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
-%%\\
-%% \ev{\sopb\mop{\mexpi0}{\mexpi1},\menv,\mkont,\msto} &\machstep
-%% \ev{\mexpi0,\menv,\kbopa{\mexpi1,\menv,\maddr},\msto'}
-%% \text{ where }a,\msto' = \mathit{push}(l,\msto,\mkont)
+\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
+\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
+\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
 \\[2mm]
+%% CONTINUE
 \co{\kmt,\mval,\msto} &\machstep
@@
 \\
-\co{\kar{\mexp,\menv,\maddr},\mval,\msto} & \machstep
-\ev{\mexp,\menv,\kfn{\mval,\maddr},\msto}
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
+\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
 \\
-\co{\kfn{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
-\ap{\mval,\mvalx{u},\mkont,\msto}
+\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
+\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
 \text{ where }\mkont \in \msto(\maddr)
 \\
-\co{\kif{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
-\ev{\mexpi0,\menv,\mkont,\msto}
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
+\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
 \text{ where }\mkont\in\msto(\maddr)
 \\
-\co{\kif{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
-\ev{\mexpi1,\menv,\mkont,\msto}
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
+\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
 \text{ where }\mkont\in\msto(\maddr)
+\\[2mm]
+%% APPLY
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
+\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
+\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
 \\
-\co{\kuop{\mop,\maddr},\mval,\msto} & \machstep
+\ap{\mop,\mval,\msto,\mkont} & \machstep
 \co{\mkont,\mval',\msto}
 \text{ where }\mkont\in\msto(\maddr)
-\text{ and } \mval'\in\delta(\mop,\mval)
-%% \\
-%% \co{\kbopa{\mop,\mexp,\menv,\maddr},\mval,\msto} & \machstep
-%% \ev{\mexp,\menv,\kbopb{\mop,\mval,\maddr},\msto}
-%% \\
-%% \co{\kbopb{\mop,\mvalx{u},\maddr},\mval,\msto} & \machstep
-%% \co{\mkont,\mval',\msto}
-%% \text{ where }\mkont=\msto(\maddr)
-%% \text{ and } \mval'=\delta(\mop,\mval,\mvalx{u})
-\\[2mm]
-\ap{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \machstep
-\ev{\mexp,\menv',\mkont,\msto'}
-\text{ where }\menv',\msto' = \mathit{bind}(\msto,\mvar,\mval)
-\\
-%% \ap{\rlos{\mvarf,\mvar,\mexp,\menv},\mval,\mkont} & \machstep
-%% \ev{\mexp,\menv',\mkont,\msto'}
-%% \text{ where }\menv',\msto' = \mathit{bind}(\msto,\langle\mvar,\mval\rangle,\langle\mvarf,
-%% \rlos{\mvarf,\mvar,\mexp,\menv}\rangle)
+\text{ and } \mval'\in\interpdelta(\mop,\mval)
 \end{align*}
@@
 choice of address representation and allocation strategy.
+
+%% We now have a framework for describing program analysis for the ISWIM
+%% family of languages, whereby approximation of both control and
+%% environment structure is regulated by the heap and allocation
+%% policies.
+
 Uniform \(k\)-CFA is one such analysis.

### 2012-10-14 paper/paper.tex
Definition for compilation.
Code for generic fixpoint and widening.

@@
 \usepackage{amssymb}
+\usepackage{stmaryrd}
 \usepackage{xltxtra}
@@
 
+\subsection{Generic fixpoint calculator}
+
+We compute the semantics of a program by iterating the state
+transition relation until a fixed point in the reachable states is
+reached.
+
+\begin{verbatim}
+;; appl : (∀ (X) ((X -> (Setof X)) -> ((Setof X) -> (Setof X))))
+(define ((appl f) s)
+  (for/fold ([i (set)])
+    ([x (in-set s)])
+    (set-union i (f x))))
+
+;; Calculate fixpoint of (appl f).
+;; fix : (∀ (X) ((X -> (Setof X)) (Setof X) -> (Setof X)))
+(define (fix f s)
+  (let loop ((accum (set)) (front s))
+    (if (set-empty? front)
+        accum
+        (let ((new-front ((appl f) front)))
+          (loop (set-union accum front)
+                (set-subtract new-front accum))))))
+\end{verbatim}
+
 \subsection{Store widening}
+
+\begin{verbatim}
+;; State^ = (cons (Set Conf) Store)
+
+;; (State -> Setof State) -> State^ -> { State^ }
+(define ((wide-step step) state)
+  (match state
+    [(cons cs σ)
+     (define ss ((appl step)
+                 (for/set ([c cs]) (c->s c σ))))
+     (set (cons (for/set ([s ss]) (s->c s))
+                (join-stores ss)))]))
+\end{verbatim}
 
@@
 
+\begin{figure}
+\begin{align*}
+\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\mval,\msto} \text{ where }\mval\in\msto(\menv(\mvar))
+\\
+\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
+\co{\mkont,\mlit,\msto}
+\\
+\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
+\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
+\\
+\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
+\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
+\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
+\\
+\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
+\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
+\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
+\end{align*}
+\caption{Compilation}
+\end{figure}
+
+\begin{figure}
+\begin{align*}
+\mathit{eval}(\mexp) &= \{ \mstate\ |\ \compile{\mexp}(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
+\\[2mm]
+%% CONTINUE
+\co{\kmt,\mval,\msto} &\machstep
+\ans{\msto,\mval}
+\\
+\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \machstep
+\mcomp^\mcntr(\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr})
+\\
+\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
+\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
+\text{ where }\mkont \in \msto(\maddr)
+\\
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \machstep
+\mcompi0^\mcntr(\menv,\msto,\mkont)
+\text{ where }\mkont\in\msto(\maddr)
+\\
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \machstep
+\mcompi1^\mcntr(\menv,\msto,\mkont)
+\text{ where }\mkont\in\msto(\maddr)
+\\[2mm]
+%% APPLY
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \machstep
+\mcomp^{\mcntr'}(\menv',\msto',\mkont)
+\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
+\\
+\ap{\mop,\mval,\msto,\mkont} & \machstep
+\co{\mkont,\mval',\msto}
+\text{ where }\mkont\in\msto(\maddr)
+\text{ and } \mval'\in\interpdelta(\mop,\mval)
+\end{align*}
+\caption{Abstract$^2$ machine for compiled ISWIM}
+\label{fig:caam}
+\end{figure}
+
 Compile:
@@
 \section{Related work}
+
+Boucher and Feeley, CC'96 introduce the idea of abstract compilation
+by closure generation.  (Feeley and Lapalme.)
 

### 2012-10-14 paper/paper.tex
Some citations.

@@
 
-Prereqs: Semantics Engineering, AAM.
+Prereqs: Semantics Engineering \cite{dvanhorn:Felleisen2009Semantics},
+AAM \cite{dvanhorn:VanHorn2011Abstracting}
+\cite{dvanhorn:VanHorn2012Systematic}.
 
@@
 
+\cite{dvanhorn:Earl2012Introspective}
+
+\cite{dvanhorn:wright-jagannathan-toplas98}
+
 Other benchmarks
@@
 
-Boucher and Feeley, CC'96 introduce the idea of abstract compilation
-by closure generation.  (Feeley and Lapalme.)
+Boucher and Feeley \cite{dvanhorn:Boucher1996Abstract} introduced the
+idea of \emph{abstract compilation}, which used closure generation
+\cite{dvanhorn:Feeley1987Using} to improve the performance of control
+flow analysis.  We have adapted the closure generation technique from
+composition evaluators to abstract machines and applied it to similar
+effect.
 
@@
 
+\bibliographystyle{plain}
+\bibliography{dvh-bibliography}
+
 \appendix
 \section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
+
+\cite{dvanhorn:nielson-nielson-popl97} \cite{dvanhorn:Neilson:1999}
 
@@
 operate at speeds roughly on par with our starting point: unoptimized
-store widened machines.
+store widened
+machines. \cite{dvanhorn:Earl2012Introspective,dvanhorn:Vardoulakis2011CFA2}
 

### 2012-11-01 paper/paper.tex

@@
-\documentclass{llncs}
+\documentclass[preprint]{sigplanconf}
 \usepackage{alltt}
@@
 \usepackage{stmaryrd}
-\usepackage{xltxtra}
-\setmonofont[Scale=MatchLowercase]{DejaVu Sans Mono}
+% \usepackage{xltxtra}
+% \setmonofont[Scale=MatchLowercase]{DejaVu Sans Mono}
 %
 \input{preamble}
+
 \begin{document}
-\title{Optimizing Abstract$^2$ Machines}
-%
-\frontmatter          % for the preliminaries
-%
-\author{J.~Ian Johnson\inst{1} \and Matthew Might\inst{2} \and David Van Horn\inst{1}}
-\institute{Northeastern University \and University of Utah}
 
-\maketitle              % typeset the title of the contribution
+\conferenceinfo{WXYZ '05}{date, City.} 
+\copyrightyear{2005} 
+\copyrightdata{[to be supplied]}
+
+% \titlebanner{banner above paper title}        % These are ignored unless
+% \preprintfooter{short description of paper}   % 'preprint' option specified.
+
+\title{Optimizing Abstract$^{\text 2}$ Machines}
+
+%% \authorinfo{J. Ian Johnson}
+%%            {Northeastern University}
+%%            {ianj@ccs.neu.edu}
+%% \authorinfo{Matthew Might}
+%%            {University of Utah}
+%%            {might@cs.utah.edu}
+%% \authorinfo{David Van Horn}
+%%            {Northeastern University}
+%%            {dvanhorn@ccs.neu.edu}
+\authorinfo{}{}{}
+
+\maketitle
 
 \begin{abstract}
+Recently \emph{abstracting abstract machines} has been proposed as a
+systematic approach to designing sound and computable program analyses
+for higher-order programming languages.  The approach has been applied
+to a wide variety of languages with features widely considered
+difficult to analyze
+
 The abstracting abstract machines approach to program analysis makes
@@
 \end{abstract}
-%
+
+%% \category{CR-number}{subcategory}{third-level}
+
+%% \terms
+%% term1, term2
+
+%% \keywords
+%% keyword1, keyword2
+
 \section{Introduction}
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{align*}
-\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\epsilon}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
+\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
 \\[2mm]
@@
 \label{fig:aam}
-\end{figure}
+\end{figure*}
 
@@
 \begin{align*}
-\mathit{push}(l,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
+\mathit{push}(\ell,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
 \mbox{ where }\maddr \notin\msto
@@
 \mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
+\mbox{ where }\maddr \notin\msto
 \end{align*}
@@
 
+\paragraph 0CFA
+
+\begin{align*}
+\mathit{push}(\ell,\msto,\mkont) &= \ell,\msto\sqcup[\ell\mapsto\{\mkont\}]
+\\
+\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
+\end{align*}
+
 \paragraph{Primitives}
@@
        (match e
-         [(var l x)           (for/set ((v (lookup ρ σ x))) (co σ k v))]
-         [(num l n)           (set (co σ k n))]
-         [(bln l b)           (set (co σ k b))]
-         [(lam l x e)         (set (co σ k (clos l x e ρ)))]
-         [(rec f (lam l x e)) (set (co σ k (rlos l f x e ρ)))]
-         [(app l f e)
+         [(var\(\superscript\ell\) x)           (for/set ((v (lookup ρ σ x))) (co σ k v))]
+         [(lit\(\superscript\ell\) l)           (set (co σ k n))]
+         [(lam\(\superscript\ell\) x e)         (set (co σ k (clos x e ρ)))]
+         [(app\(\superscript\ell\) f e)
           (define-values (σ* a) (push state))
           (set (ev σ* f ρ (ar e ρ a)))]
-         [(ife l e0 e1 e2)
+         [(ife\(\superscript\ell\) e0 e1 e2)
           (define-values (σ* a) (push state))
-          (set (ev σ* e0 ρ (ifk e1 e2 ρ a)))]
-         [(1op l o e)
-          (define-values (σ* a) (push state))
-          (set (ev σ* e ρ (1opk o a)))])]
+          (set (ev σ* e0 ρ (ifk e1 e2 ρ a)))])]
       [(co σ k v)
@@
          ['mt (set (ans σ v))]
-         [(ar e ρ l) (set (ev σ e ρ (fn v l)))]
-         [(fn f l)   (for/set ((k (get-cont σ l))) (ap σ f v k))]
-         [(fi c a ρ l)
+         [(ar\(\superscript\ell\) e ρ) (set (ev σ e ρ (fn v l)))]
+         [(fn\(\superscript\ell\) f)   (for/set ((k (get-cont σ l))) (ap σ f v k))]
+         [(fi\(\superscript\ell\) c a ρ)
           (for/set ((k (get-cont σ l)))
-            (ev σ (if v c a) ρ k))]
-         [(oa o l)
-          (for*/set ((k (get-cont σ l))
-                     (v (delta o (list v))))
-            (co σ k v))]
+            (ev σ (if v c a) ρ k))])]
       [(ap σ fun a k)
@@
           (set (ev σ* e ρ* k))]
-         [_ (set)])]
+         [(? op? o)
+          (for*/set ((k (get-cont σ l))
+                     (v (\(\Delta\) o (list v))))
+            (co σ k v))
+         [_ (set)]))]))
 \end{alltt}
@@
 \newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
+
+\begin{alltt}
+(define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))
+
+(define pred
+  (\(\lambda\) (n)
+    (\(\lambda\) (rf)
+      (\(\lambda\) (rx)
+        (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
+          (\(\lambda\) (i) rx))
+         (\(\lambda\) (id) id))))))
+\end{alltt}
 
@@
 
+\newpage
 \subsection{Generic fixpoint calculator}
+
+\begin{alltt}
+;; Expr \(\rightarrow\) Setof State
+(define (eval e)
+  (fix step (set (inj e))))
+\end{alltt}
 
@@
 
+\newpage
 \subsection{Fixed-point specialization}
 
+\begin{alltt}
+;; State^ -> State^
+;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
+(define (wide-step-specialized state)
+  (match state
+    [(cons σ cs)
+     (define-values (cs* σ*)
+       (for/fold ([cs* (set)] [σ* σ])
+         ([c cs])
+         (match (step-compiled^ (cons σ c))
+           [(cons σ** cs**)
+            (values (set-union cs* cs**) (join-store σ* σ**))])))
+     (cons σ* (set-union cs cs*))]))
+\end{alltt}
 %% Should be commutted to beginning
@@
 \subsection{Computing with store changes}
+
+\begin{alltt}
+;; State^ -> State^
+;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
+(define (wide-step-specialized state)
+  (match state
+    [(cons σ cs)
+     (define-values (cs* ∆)
+       (for/fold ([cs* (set)] [∆* '()])
+         ([c cs])
+         (match (step-compiled^ (cons σ c))
+           [(cons ∆** cs**)
+            (values (set-union cs* cs**) (append ∆** ∆*))])))
+     (cons (update ∆ σ) (set-union cs cs*))]))
+\end{alltt}
+
+\[
+\{ \widehat{v_3}, \widehat{v_5}, \widehat{v_6} \} = 0001011
+\]
+
+\[
+\sigma = \mbox{\tt \#(0$_0$,\dots,0$_{|P|}$)}
+\]
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{splncs}
 \bibliography{dvh-bibliography}

### 2012-11-01 paper/paper.tex

@@
 
+\subsection{Another problem with acceptability: Temporal ignorance}
+
+The small-step approach to static analysis brings subtle yet important temporal
+richness not found in classical analyses for higher-order programs.
+%
+Classical analyses (ultimately) compute judgments on program terms and
+contexts, e.g., at expression $e$, value $x$ may have value $v$.
+%
+The judgments do not relate the order in which expressions and context may be
+evaluated in a program, e.g., a classical analysis has nothing to say with
+regard to question like, ``Do we always evaluate $e_1$ before $e_2$?'' or ``Is
+it always the case that a file handle is opened, read and then closed in that
+order?''
+
+Small-step analyses, by their nature, encode the temporal relationships between
+abstract states.
+%
+It is sensible to make temporal queries of a small-step analysis.
+%
+Of course, this does not come for free: respecting temporal order imposes an
+order in which states and terms may be evaluated \emph{during} the analysis.
+%
+Classical analyses can (and do) evaluate expressions in any order, or in some
+cases, even in parallel~\cite{might:Prabhu:2010:EigenCFA}.
+%
+Relaxing that restriction on order affords additional optimizations that we
+have \emph{not} performed.
+
+We avoid sacrificing order not simply because we are interested in the
+questions it allows us to ask, but because considering temporal order actually
+improves the precision of the analysis itself.
+
 \section{Pushdown Analysis}

### 2012-11-01 paper/paper.tex
Filled in some of the implementation section.

--- 2012-11-01 paper/paper.tex
@@
 
+The first four techniques reside in the purely functional realm. The
+following three techniques use imperative constructs to avoid allocation
+overhead.
+
 \subsection{Lazy non-determinism}
@@
 
+% For David to fill in.
+\subsection{Imperative global store and worklist}
+...
+\subsection{Pre-allocated global store}
+...
+
+% Ian's sections: generators, implementation and compound data
+\section{Implementation}
+
+All the above techniques can be combined into a single implementation
+of the small step reduction relation and small variations of a general
+fixpoint combinator. We implementated all of these ideas in an
+analysis of R4RS scheme in just 3000 lines of Racket. Most lines are
+spent listing all the primitives and library functions (and their
+abstract and concrete implementations) of R4RS. Here we describe the
+unifying vocabulary for writing such a framework, and some design
+decisions related to scaling to a language with compound data.
+
+\subsection{Vocabulary of Parameterized Optimizations}
+
+We combine the interface of generators ~\cite{ianjohnson:cluhistory}
+with a special lambda form ({\tt $\lambda$\%}) that wraps the meanings
+of {\tt ev} states. We also have a special {\tt do} form that performs
+behind-the-scenes state passing and iteration much like {\tt do} in
+Haskell. In the presence of the different optimization
+techniques, {\tt generator}, {\tt yield}, {\tt $\lambda$\%} and {\tt do}
+change meaning slightly, but keep the same interface as their names
+suggest.
+
+\begin{figure}
+\begin{alltt}
+\(e-dispatch =\)
+ (match e
+    [(var\(\superscript\ell\) x)
+     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
+      (do (\(\msto\)) ((v (delay (lookup-env \(\menv\) x))))
+        (yield (co \(\msto\) k v))))]
+    [(lit\(\superscript\ell\) l)
+     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
+      (do (\(\msto\)) () (yield (co \(\msto\) k n))))]
+    [(lam\(\superscript\ell\) x e)
+     (define c (compile e))
+     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
+      (do (\(\msto\)) () (yield (co \(\msto\) k (clos x c \(\menv\))))))]
+    [(app\(\superscript\ell\) f e)
+     (define cf (compile f))
+     (define ce (compile e))
+     (\(\lambda\)\% (\(\msto\) \(\menv\) k)
+      (define-values (\(\msto\)* a) (push state))
+      (yield (ev \(\msto\)* cf \(\menv\) (ar c \(\menv\) a))))]
+    [(ife\(\superscript\ell\) e0 e1 e2)
+      (define c0 (compile e0))
+      (define c1 (compile e1))
+      (define c2 (compile e2))
+      (\(\lambda\)\% (\(\msto\) \(\menv\) k)
+        (do (\(\msto\)) [(\(\msto\)* a) (push state)]
+         (yield (ev \(\msto\)* c0 \(\menv\) (ifk c1 c2 \(\menv\) a)))))])
+
+  (define (step state)
+    (match state
+      [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
+      [(co \(\msto\) k v)
+       (match k
+         ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
+           (yield (ans \(\msto\) v))))]
+         [(ar\(\superscript\ell\) e \(\menv\))
+          (generator (do (\(\msto\)) ()
+            (yield (ev \(\msto\) e \(\menv\) (fn v l)))))
+         [(fn\(\superscript\ell\) f)
+          (generator (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)])
+           (yield (ap \(\msto\) f v k))))]
+         [(fi\(\superscript\ell\) c a \(\menv\))
+          (generator
+           (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
+                    [v <- (force \(\msto\) v)])
+             (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
+      [(ap \(\msto\) fun a k)
+       (match fun
+         [(clos l x e \(\menv\))
+          (generator
+           (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
+             (yield (ev \(\msto\)* e \(\menv\)* k))))]
+         [(? op? o)
+          (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
+                   [v <- (force \(\msto\) v)]
+                   [v (\(\Delta\) o (list v))])
+            (yield (co \(\msto\) k v)))
+         [_ (generator (do (\(\msto\)) () (continue)))]))]))
+\end{alltt}
+\caption{Parameterized Abstract$^2$ machine for ISWIM}
+\label{fig:paam}
+\end{figure}
+
+\begin{figure}
+\begin{alltt}
+(define (compile e) \(e-dispatch\))
+(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\machstep\) (\(\lambda\) (\(\msto\) \(\menv\) k) body \ldots)
+(yield (ev \(\msto\) c . rest)) \(\longmapsto\) (c \(\msto\) . rest)
+(yield e) \(\longmapsto\) (yield-meaning e)
+\end{alltt}
+\caption{Meaning of forms if compiled}
+\label{fig:cfm}
+\end{figure}
+
+\begin{figure}
+\begin{alltt}
+(define (compile e) e)
+(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\longmapsto\) (let () body ...)
+(yield e) \(\longmapsto\) (yield-meaning e)
+\end{alltt}
+\caption{Meaning of forms if not compiled}
+\label{fig:ncfm}
+\end{figure}
+
+Why generators? Relations in general cannot be written as computable
+functions, but in our case we have a decidable, finitely
+non-deterministic relation. There are several implementation
+strategies for relations, some more effective than others. For
+instance, a silly implementation strategy would be defining a binary
+function that decides membership in the relation. We want the
+definition to drive the interpretor, so we might naturally decide to
+write a function that takes a state and returns a set of all the next
+states. As we described in the imperative workset section, this
+intermediate allocation is unnecessary and wasteful. What we really
+want is a way of pulling on the reduction relation for more next
+states as the fixpoint is ready for them.
+
+For intermediate sets, {\tt yield} is {\tt return} in the set
+monad. For the imperative worklist, {\tt yield} performs mutation
+behind the scenes. Indeed, we could use real generators! This is the
+only case where {\tt generator} constructs a nullary generator; in the
+other cases it evaluates to its body. Racket's generators are not
+tuned for performance, so we actually lose efficiency in this case,
+but the unifying vocabulary is instructive nonetheless.
+
+The {\tt $\lambda$\%} form constructs a real closure in the
+compilation case and evaluates to its body otherwise.
+
+The {\tt do} form has the special job of maintaining the current store
+and accumulated states (if indeed those are being passed around). It
+uses Racket's syntax parameters ~\cite{ianjohnson:eli/stxparam} to
+track the most recent bindings of these accumulators so that {\tt
+  yield-meaning} can refer to them (say {\tt $\sigma$p} and {\tt
+  statep}). In the case of state passing and set-accumulated states, its
+definition is simply
+
+\begin{center}
+\begin{alltt}
+(yield-meaning e) \(\longmapsto\) (values \(\sigma\)p (set-add statep e))
+\end{alltt}
+\end{center}
+
+The linguistic elements described above were all reusable to implement
+the various languages' abstract interpreters. One such language was
+R4RS Scheme, which has compound data (e.g. vectors and conses).
+
+\subsection{Abstracting Compound Data}
+
+We discovered that the choice of abstract domain for literal data
+(such as the large quoted axiom database in the Boyer benchmark)
+drastically affected the performance of our analysis on the benchmark
+suite. It is worth discussing the ``obvious'' and na\"ive abstraction
+that is natural to a reader of
+~\citep{dvanhorn:VanHorn2011Abstracting}, and other easy-to-implement
+yet effective approaches.
+
+TODO: {\tt (cons A-addr D-addr)}, {\tt qpair}, {\tt opaque-atom}.
+
 \section{Evaluation}

### 2012-11-02 paper/paper.tex
Elaborated on compound data abstraction.

--- 2012-11-01 paper/paper.tex
@@
 
-TODO: {\tt (cons A-addr D-addr)}, {\tt qpair}, {\tt opaque-atom}.
+\subsubsection{Na\"ively}
+
+The uniform way they approach a simple abstraction strategy is to cut
+recursion out of the data definition by tying the recursive knot
+through the abstract store. For Scheme, the grammar for values looks like the following:
+
+\begin{alltt}
+Value ::= #t | #f | (cons Value Value) | '() | ...
+\end{alltt}
+
+Upon evaluating a {\tt cons} application, we instead allocate two
+addresses $a$ and $d$, join them to the respective values in the
+store, and return the flattened {\tt (cons $a$ $d$)} value. Since
+these addresses are all distinguished at different syntactic
+callsites, and quoted lists really are sugar for a sequence of calls
+to {\tt cons}, this abstraction explodes the value space. Analyzing a
+function that counts the number of atoms in a literal s-expression
+would actually interpret that function at least that number of times
+(more because of intermediate conses). Indeed, even in our fastest
+parameterization, this abstraction causes the analysis of Boyer to be
+430 times slower than the approach we will now describe.
+
+\subsubsection{Less na\"ively and less precisely}
+
+The number of syntactic uses of {\tt cons} versus implicit uses via
+literal lists is dramatically smaller in typical Scheme programs. We
+use the above abstraction for these syntactic uses, but choose a
+different strategy for quoted lists. Quotation is special because it
+cannot introduce function values or mutable values, which is important
+to enhancing our technique's soundness to conceptual complexity ratio.
+
+There are a few steps to consider:
+\begin{itemize}
+ \item{Define special value lattice elements for compound data domains that can be quoted
+       (e.g. {\tt Pair} for $\left\lbrace ({\tt cons} a_i d_i)\right\rbrace_i$, {\tt IVector} for immutable vectors, etc.)}
+ \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
+ \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
+       and similar definitions for immutable vectors.}
+ \item{Extend the $\Delta$ axioms to include conservative meaning for these new values (e.g.
+{\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
+\end{itemize}
+
+\begin{figure}
+\begin{align*}
+\Delta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
+\Delta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
+\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
+                            &\qquad \sqcup[d \mapsto \lbrace \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\rbrace] \\
+\bigsqcup(v) &= v \\
+\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
+merge(v, v) &= v \\
+merge(n, m) &= {\tt Number} \\
+merge(n, v) &= {\tt QData} \\
+merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt Pair} \\
+merge({\tt Pair}, ({\tt cons}\ a\ d)) &= {\tt Pair} \\
+merge({\tt Pair}, v) &= {\tt QData} \\
+\vdots
+\end{align*}
+\caption{Quoted list primitive}
+\label{fig:qlist}
+\end{figure}
+
+The alternative to {\tt QData} is to go higher in the lattice to
+$\top$, which has more complicated meaning (Shivers' escape semantics) and
+is overly approximate.
 

### 2012-11-05 paper/paper.tex

@@
 \\
-\ap{\mop,\mval,\msto,\mkont} & \machstep
+\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
 \co{\mkont,\mval',\msto}
@@
 \begin{align*}
-\mathit{push}(\ell,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
+\mathit{push}(\mlab,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
 \mbox{ where }\maddr \notin\msto
@@
 \begin{align*}
-\mathit{push}(\ell,\msto,\mkont) &= \ell,\msto\sqcup[\ell\mapsto\{\mkont\}]
+\mathit{push}(\mlab,\msto,\mkont) &= \mlab,\msto\sqcup[\mlab\mapsto\{\mkont\}]
 \\
@@
 \begin{align*}
-\mnum+1 &\in \delta(\saddone,\mnum) &
-\mnum-1 &\in \delta(\ssubone,\mnum)\\
-\strue &\in \delta(\szerohuh,\szero) &
-\sfalse &\in \delta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+\mnum+1 &\in \interpdelta(\saddone,\mnum) &
+\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
+\strue &\in \interpdelta(\szerohuh,\szero) &
+\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
 \end{align*}
@@
 \begin{align*}
-\sNum &\in \hat\delta(\saddone,\mnum) &
-\sNum &\in \hat\delta(\ssubone,\mnum)\\
-\strue &\in \hat\delta(\szerohuh,\sNum) &
-\sfalse &\in \hat\delta(\szerohuh,\sNum)\\
-\strue &\in \hat\delta(\szerohuh,\szero) &
-\sfalse &\in \hat\delta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+\sNum &\in \hat\interpdelta(\saddone,\mnum) &
+\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
+\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
+\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
+\strue &\in \hat\interpdelta(\szerohuh,\szero) &
+\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
 \end{align*}
+
+HERE: ρ
 
@@
   ;; State \(\rightarrow\) Setof State
-  (define (step state)
-    (match state
-      [(ev σ e ρ k)
-       (match e
-         [(var\(\superscript\ell\) x)           (for/set ((v (lookup ρ σ x))) (co σ k v))]
-         [(lit\(\superscript\ell\) l)           (set (co σ k n))]
-         [(lam\(\superscript\ell\) x e)         (set (co σ k (clos x e ρ)))]
-         [(app\(\superscript\ell\) f e)
-          (define-values (σ* a) (push state))
-          (set (ev σ* f ρ (ar e ρ a)))]
-         [(ife\(\superscript\ell\) e0 e1 e2)
-          (define-values (σ* a) (push state))
-          (set (ev σ* e0 ρ (ifk e1 e2 ρ a)))])]
-      [(co σ k v)
-       (match k
-         ['mt (set (ans σ v))]
-         [(ar\(\superscript\ell\) e ρ) (set (ev σ e ρ (fn v l)))]
-         [(fn\(\superscript\ell\) f)   (for/set ((k (get-cont σ l))) (ap σ f v k))]
-         [(fi\(\superscript\ell\) c a ρ)
-          (for/set ((k (get-cont σ l)))
-            (ev σ (if v c a) ρ k))])]
-      [(ap σ fun a k)
+  (define (step \(\mstate\))
+    (match \(\mstate\)
+      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
+       (match \(\mexp\)
+         [(var\(\superscript\mlab\) \(\mvar\))
+          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
+            (co \(\msto\) \(\mkont\) \(\mval\)))]
+         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
+         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
+         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
+          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
+          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
+         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
+          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
+          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
+      [(co \(\msto\) \(\mkont\) \(\mval\))
+       (match \(\mkont\)
+         ['mt (set (ans \(\msto\) \(\mval\)))]
+         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
+         [(fn\(\superscript\mlab\) \(\mval'\))
+          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
+            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
+         [(fi\(\superscript\mlab\) c a \(\rho\))
+          (for/set ((k (get-cont \(\msto\) l)))
+            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
+      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
        (match fun
-         [(clos l x e ρ)
-          (define-values (ρ* σ*) (bind state))
-          (set (ev σ* e ρ* k))]
-         [(? op? o)
-          (for*/set ((k (get-cont σ l))
-                     (v (\(\Delta\) o (list v))))
-            (co σ k v))
+         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
+          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
+          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
+         [(op \(\mop\))
+          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
+                     (\(\mval\) (\(\Delta\) \(\mop\) \(\mval\))))
+            (co \(\msto\) \(\mkont\) \(\mval\)))]
          [_ (set)]))]))
@@
 
+%% \begin{alltt}
+%% (define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))
+
+%% (define pred
+%%   (\(\lambda\) (n)
+%%     (\(\lambda\) (rf)
+%%       (\(\lambda\) (rx)
+%%         (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
+%%           (\(\lambda\) (i) rx))
+%%          (\(\lambda\) (id) id))))))
+%% \end{alltt}
+
+%% (define \church0 (\(\lambda\) (f0) (\(\lambda\) (x0) x0)))
+%% (define \church1 (\(\lambda\) (f1) (\(\lambda\) (x1) (f1 x1))))
+%% (define \church2 (\(\lambda\) (f2) (\(\lambda\) (x2) (f2 (f2 x2)))))
+%% (define \church3 (\(\lambda\) (f3) (\(\lambda\) (x3) (f3 (f3 (f3 x3))))))
+
+\begin{figure}
 \begin{alltt}
-(define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))
+(define ((((plus p1) p2) pf) x)
+  ((p1 pf) ((p2 pf) x)))
 
-(define pred
-  (\(\lambda\) (n)
-    (\(\lambda\) (rf)
-      (\(\lambda\) (rx)
+(define (((mult m1) m2) mf)
+  (m2 (m1 mf)))
+
+(define (((pred n) rf) rx)
   (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
-          (\(\lambda\) (i) rx))
-         (\(\lambda\) (id) id))))))
-\end{alltt}
+    (\(\lambda\) (ignored) rx))
+   (\(\lambda\) (id) id)))
 
-\begin{alltt}
+(define (zero? n)
+  ((n (\(\lambda\) (zx) #f)) #t))
+
+(define ((church=? e1) e2)
+  (if (zero? e1) (zero? e2)
+      (if (zero? e2) #f
+          ((church=? (pred e1)) (pred e2)))))
+
 ;; multiplication distributes over addition
 ((church=? ((mult \church2) ((plus \church1) \church3)))
- ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))))
+ ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
 \end{alltt}
+\caption{Church-numeral calculation}
+\end{figure}
 
@@
   (match state
-    [(cons cs σ)
+    [(cons cs \(\msto\))
      (define ss ((appl step)
-                 (for/set ([c cs]) (c->s c σ))))
+                 (for/set ([c cs]) (c->s c \(\msto\)))))
      (set (cons (for/set ([s ss]) (s->c s))
@@
 \compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
+\\
+&
 \text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
@@
 \compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
-\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
+\\
+&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
 \end{align*}
@@
   (match state
-    [(cons σ cs)
-     (define-values (cs* σ*)
-       (for/fold ([cs* (set)] [σ* σ])
+    [(cons \(\msto\) cs)
+     (define-values (cs* \(\msto\)*)
+       (for/fold ([cs* (set)] [\(\msto\)* \(\msto\)])
          ([c cs])
-         (match (step-compiled^ (cons σ c))
-           [(cons σ** cs**)
-            (values (set-union cs* cs**) (join-store σ* σ**))])))
-     (cons σ* (set-union cs cs*))]))
+         (match (step-compiled^ (cons \(\msto\) c))
+           [(cons \(\msto\)** cs**)
+            (values (set-union cs* cs**) (join-store \(\msto\)* \(\msto\)**))])))
+     (cons \(\msto\)* (set-union cs cs*))]))
 \end{alltt}
@@
   (match state
-    [(cons σ cs)
+    [(cons \(\msto\) cs)
      (define-values (cs* ∆)
@@
          ([c cs])
-         (match (step-compiled^ (cons σ c))
+         (match (step-compiled^ (cons \(\msto\) c))
            [(cons ∆** cs**)
             (values (set-union cs* cs**) (append ∆** ∆*))])))
-     (cons (update ∆ σ) (set-union cs cs*))]))
+     (cons (update ∆ \(\msto\)) (set-union cs cs*))]))
 \end{alltt}
@@
 
-We combine the interface of generators ~\cite{ianjohnson:cluhistory}
+We combine the interface of generators~\cite{ianjohnson:cluhistory}
 with a special lambda form ({\tt $\lambda$\%}) that wraps the meanings
@@
  (match e
-    [(var\(\superscript\ell\) x)
+    [(var\(\superscript\mlab\) x)
      (\(\lambda\)\% (\(\msto\) \(\menv\) k)
@@
         (yield (co \(\msto\) k v))))]
-    [(lit\(\superscript\ell\) l)
+    [(lit\(\superscript\mlab\) l)
      (\(\lambda\)\% (\(\msto\) \(\menv\) k)
       (do (\(\msto\)) () (yield (co \(\msto\) k n))))]
-    [(lam\(\superscript\ell\) x e)
+    [(lam\(\superscript\mlab\) x e)
      (define c (compile e))
@@
       (do (\(\msto\)) () (yield (co \(\msto\) k (clos x c \(\menv\))))))]
-    [(app\(\superscript\ell\) f e)
+    [(app\(\superscript\mlab\) f e)
      (define cf (compile f))
@@
       (yield (ev \(\msto\)* cf \(\menv\) (ar c \(\menv\) a))))]
-    [(ife\(\superscript\ell\) e0 e1 e2)
+    [(ife\(\superscript\mlab\) e0 e1 e2)
       (define c0 (compile e0))
@@
            (yield (ans \(\msto\) v))))]
-         [(ar\(\superscript\ell\) e \(\menv\))
+         [(ar\(\superscript\mlab\) e \(\menv\))
           (generator (do (\(\msto\)) ()
             (yield (ev \(\msto\) e \(\menv\) (fn v l)))))
-         [(fn\(\superscript\ell\) f)
+         [(fn\(\superscript\mlab\) f)
           (generator (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)])
            (yield (ap \(\msto\) f v k))))]
-         [(fi\(\superscript\ell\) c a \(\menv\))
+         [(fi\(\superscript\mlab\) c a \(\menv\))
           (generator
@@
 \bibliographystyle{splncs}
-\bibliography{dvh-bibliography}
+\bibliography{bibliography}
 
@@
 \begin{verbatim}
- (let ((id (lambda (x) x)))
+ (let ((id (\(\lambda\) (x) x)))
    (begin (id 1) (id 2)))

### 2012-11-07 paper/paper.tex

@@
           (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
-                     (\(\mval\) (\(\Delta\) \(\mop\) \(\mval\))))
+                     (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
             (co \(\msto\) \(\mkont\) \(\mval\)))]
@@
          (match (step-compiled^ (cons \(\msto\) c))
-           [(cons ∆** cs**)
-            (values (set-union cs* cs**) (append ∆** ∆*))])))
-     (cons (update ∆ \(\msto\)) (set-union cs cs*))]))
+           [(cons \(\msdiff\)** cs**)
+            (values (set-union cs* cs**) (append \(\msdiff\)** \(\msdiff\)*))])))
+     (cons (update \(\msdiff\) \(\msto\)) (set-union cs cs*))]))
 \end{alltt}
@@
                  [v <- (force \(\msto\) v)]
-                   [v (\(\Delta\) o (list v))])
+                 [v (\(\interpdelta\) o (list v))])
           (yield (co \(\msto\) k v)))
@@
 Why generators? Relations in general cannot be written as computable
-functions, but in our case we have a decidable, finitely
-non-deterministic relation. There are several implementation
-strategies for relations, some more effective than others. For
-instance, a silly implementation strategy would be defining a binary
-function that decides membership in the relation. We want the
-definition to drive the interpretor, so we might naturally decide to
-write a function that takes a state and returns a set of all the next
-states. As we described in the imperative workset section, this
-intermediate allocation is unnecessary and wasteful. What we really
-want is a way of pulling on the reduction relation for more next
-states as the fixpoint is ready for them.
+functions, but in our case we have a decidable relation with finitely
+many pairs sharing the same left component. There are several
+implementation strategies for such relations, some more effective than
+others. For instance, a silly implementation strategy would be
+defining a binary function that decides membership in the relation. We
+want the definition to drive the interpreter, so we might naturally
+decide to write a function that takes a state and returns a set of all
+the next states. As we described in the imperative workset section,
+this intermediate allocation is unnecessary and wasteful. What we
+really want is a way of pulling on the reduction relation for more
+next states as the fixpoint is ready for them.
 
@@
  \item{Define special value lattice elements for compound data domains that can be quoted
-       (e.g. {\tt Pair} for $\left\lbrace ({\tt cons} a_i d_i)\right\rbrace_i$, {\tt IVector} for immutable vectors, etc.)}
+       (e.g. {\tt QPair} for $\left\lbrace ({\tt cons} a_i d_i)\right\rbrace_i$, {\tt QVector} for immutable vectors, etc.)}
  \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
@@
        and similar definitions for immutable vectors.}
- \item{Extend the $\Delta$ axioms to include conservative meaning for these new values (e.g.
+ \item{Extend the $\interpdelta$ axioms to include conservative meaning for these new values (e.g.
 {\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
@@
 \begin{align*}
-\Delta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
-\Delta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
+\interpdelta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
+\interpdelta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
 \qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
@@
 merge(n, v) &= {\tt QData} \\
-merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt Pair} \\
-merge({\tt Pair}, ({\tt cons}\ a\ d)) &= {\tt Pair} \\
-merge({\tt Pair}, v) &= {\tt QData} \\
+merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
+merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
+merge({\tt QPair}, v) &= {\tt QData} \\
 \vdots

### 2012-11-07 paper/paper.tex
Filled in the details of uniform k-CFA

@@
 \ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
-\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
+\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
 \\
@@
 \begin{align*}
-\mathit{push}(\mlab,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
+\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
 \mbox{ where }\maddr \notin\msto
 \\
-\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
+\mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
 \mbox{ where }\maddr \notin\msto
@@
 \begin{align*}
-\mathit{push} &= \dots\\
-\mathit{bind} &= \dots
+\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
+  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
+\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
+                                           \msto\sqcup[\maddr \mapsto
+                                             \{ \mval\}],
+                                           \mcntr' \\
+\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
+              \maddr &= x\mcntr' \\
+%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
+              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
+              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
 \end{align*}
+%% Redundant
+%% \paragraph 0CFA
 
-\paragraph 0CFA
-
-\begin{align*}
-\mathit{push}(\mlab,\msto,\mkont) &= \mlab,\msto\sqcup[\mlab\mapsto\{\mkont\}]
-\\
-\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
-\end{align*}
+%% \begin{align*}
+%% \mathit{push}(\mlab,\msto,\mkont) &= \mlab,\msto\sqcup[\mlab\mapsto\{\mkont\}]
+%% \\
+%% \mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
+%% \end{align*}
 
@@
 
-The first four techniques reside in the purely functional realm. The
-following three techniques use imperative constructs to avoid allocation
-overhead.
+We discuss a variety of techniques for increasing performance by
+decreasing the interpretive overhead of abstract interpretation, all
+in the purely functional realm. These techniques are the following:
+\begin{itemize}
+ \item{shrink state space by quotients (lazy non-determinism)}
+ \item{shrink state space by removing {\tt ev} states (abstract
+   compilation)}
+ \item{decrease intermediate data structures by specializing our
+fixpoint operator to work with widened machine configurations
+(fixpoint specialization)}
+ \item{decrease intermediate data structures by operating on store
+   changes rather than whole stores (store deltas)}
+\end{itemize}
+Each technique requires minimal changes to the reduction
+relation/fixpoint, and each improves performance by integer factors.
 
@@
 
+The first problem that we realized with the AAM approach was that much
+of the non-determinism is spurious. For example, in a function
+application,
+\begin{alltt}
+(f x y z w)
+\end{alltt}
+where all variables have been joined with multiple values, each
+variable reference transitions to several new states that comprise the
+cartesian product of all the values of {\tt x}, {\tt y}, {\tt z} and
+{\tt w}, only to be joined back together in {\tt f}'s bindings. For
+this reason, we {\it delay} the non-determinism until it is needed in
+{\it strict contexts} (such as the guard for an {\tt if} a called
+procedure, or a numerical primitive application).
+
+For this, we introduce a new value for a delayed lookup that acts to
+quotient the state space by the values at the delayed address.
+
+\begin{align*}
+force(\clos{x,e,\menv},\msto) &= \{\clos{x,e,\menv}\}\\
+force(\saddr\maddralt,\msto) &= \msto(\maddralt)\\
+\ev{\svar{x},\menv,\mkont,\msto} &\machstep\; 
+\co{\mkont,\saddr{\menv(x)},\msto} \\
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
+&\machstep\;
+\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{v,\maddr}} \\
+\mbox{where } &v' \in force(v,\msto)
+bind^\mcntr_\mlab(\msto,x,\menv,v) =
+\end{align*}
 Lazy:
@@
 
+\section{Sacrificing precision for performance}
+
+Each step in the fixpointer for the above techniques requires joining
+several (possibly large) stores together. Even with immutable hashes
+that utilize sharing, this is memory-intensive. Worse, this
+contributes an obscene fraction of the runtime. We reuse a technique
+created in Shivers' dissertation~\cite{ianjohnson:Shivers:1991:CFA},
+timestamp approximation, that removes the need for joining whole
+stores by manipulating a global store imperatively. This loses
+precision because the starting store for each state in a machine
+configuration can change between stepping all states and introduce
+more flows than originally possible. The example in figure
+\ref{fig:timestamp} illustrates this loss in precision. {\bf TODO:}
+EXPLANATION OF LOSS. In many applications, this is acceptable for the
+performance improvement.
+
+\begin{figure}
+TODO
+\caption{Precision lossy program in timestamp approximation}
+\label{fig:timestamp}
+\end{figure}
+
 % For David to fill in.
@@
 All the above techniques can be combined into a single implementation
-of the small step reduction relation and small variations of a general
-fixpoint combinator. We implementated all of these ideas in an
-analysis of R4RS scheme in just 3000 lines of Racket. Most lines are
-spent listing all the primitives and library functions (and their
-abstract and concrete implementations) of R4RS. Here we describe the
-unifying vocabulary for writing such a framework, and some design
-decisions related to scaling to a language with compound data.
+of the small step reduction relation and simple variations of a
+general fixpoint combinator. We implementated all of these ideas in an
+analysis of core Scheme (call-by-value lambda calculus extended with
+variable-arity functions, call/cc, vectors, conses, recursive binding,
+various primitives) in just 3000 lines of Racket. Most lines are spent
+defining the primitives and expected library functions used in our
+benchmarks. Here we describe the unifying vocabulary for writing such
+a framework, and some design decisions related to scaling to a
+language with compound data.
 
@@
 The linguistic elements described above were all reusable to implement
-the various languages' abstract interpreters. One such language was
-R4RS Scheme, which has compound data (e.g. vectors and conses).
+the various languages' abstract interpreters. The language of our
+benchmarks, Scheme, has compound data (e.g. vectors and conses), which
+introduces choices for abstraction.
 
@@
  \item{Define special value lattice elements for compound data domains that can be quoted
-       (e.g. {\tt QPair} for $\left\lbrace ({\tt cons} a_i d_i)\right\rbrace_i$, {\tt QVector} for immutable vectors, etc.)}
+       (e.g. {\tt QPair} for $\left\{ ({\tt cons} a_i d_i)\right\}_i$, {\tt QVector} for immutable vectors, etc.)}
  \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
@@
 \qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
-                            &\qquad \sqcup[d \mapsto \lbrace \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\rbrace] \\
+                            &\qquad \sqcup[d \mapsto \{ \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\}] \\
 \bigsqcup(v) &= v \\

### 2012-11-07 paper/paper.tex
Added same explanation for laziness and used unicode to make code fit.

@@
 \begin{alltt}
-  ;; State \(\rightarrow\) Setof State
+  ;; State \(\to\) Set[State]
   (define (step \(\mstate\))
@@
 \begin{alltt}
-;; Expr \(\rightarrow\) Setof State
+;; Expr \(\to\) Set[State]
 (define (eval e)
@@
 
-\begin{verbatim}
-;; appl : (∀ (X) ((X -> (Setof X)) -> ((Setof X) -> (Setof X))))
+\begin{alltt}
+;; \(\forall\) (X) ((X \(\to\) Set[X]) \(\to\) (Set[X] \(\to\) Set[X]))
 (define ((appl f) s)
@@
 ;; Calculate fixpoint of (appl f).
-;; fix : (∀ (X) ((X -> (Setof X)) (Setof X) -> (Setof X)))
+;; \(\forall\) (X) ((X \(\to\) Set[X]) Set[X] \(\to\) Set[X])
 (define (fix f s)
@@
                 (set-subtract new-front accum))))))
-\end{verbatim}
+\end{alltt}
 
@@
 
-\begin{verbatim}
-;; State^ = (cons (Set Conf) Store)
-
-;; (State -> Setof State) -> State^ -> { State^ }
+\begin{alltt}
+;; \(\widehat{State}\) = (cons (Set Conf) Store)
+
+;; (State \(\to\) Set[State]) \(\to\) \(\widehat{State}\) \(\to\) \(\widehat{State}\)
 (define ((wide-step step) state)
@@
                 (join-stores ss)))]))
-\end{verbatim}
+\end{alltt}
 
@@
 \begin{align*}
-force(\clos{x,e,\menv},\msto) &= \{\clos{x,e,\menv}\}\\
+force(\clos{\mvar,\mexp,\menv},\msto) &= \{\clos{\mvar,\mexp,\menv}\}\\
 force(\saddr\maddralt,\msto) &= \msto(\maddralt)\\
-\ev{\svar{x},\menv,\mkont,\msto} &\machstep\; 
-\co{\mkont,\saddr{\menv(x)},\msto} \\
+\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\; 
+\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
 \co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
 &\machstep\;
-\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{v,\maddr}} \\
-\mbox{where } &v' \in force(v,\msto)
-bind^\mcntr_\mlab(\msto,x,\menv,v) =
+\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}} \\
+\mbox{where } &\mval' \in force(\mval,\msto) \\
+bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
+  \maddr], \msto\sqcup[\maddr \mapsto force(\mval, \msto)], \mcntr' \\
+\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
+              \maddr &= \mvar\mcntr'
 \end{align*}
+
+The force for the second rule is necessary to keep the same type for
+the \(\kfn{}\) continuation, but we would rather use an address. This
+introduces (acceptably) more imprecision, but has the same precision as an ANF
+analysis\footnote{Exercise for the reader.}.
+
 Lazy:
@@
 \ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \machstep
-\mcomp^{\mcntr'}(\menv',\msto',\mkont)
-\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
+\mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
+\text{ where }\menv',\msto',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
 \\
 \ap{\mop,\mval,\msto,\mkont} & \machstep
-\co{\mkont,\mval',\msto}
-\text{ where }\mkont\in\msto(\maddr)
+\co{\mkont,\mval',\msto} \\
+\text{ where }\mkont &\in\msto(\maddr)
 \text{ and } \mval'\in\interpdelta(\mop,\mval)
@@
 \begin{alltt}
-;; State^ -> State^
-;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
+;; \(\widehat{State} \to \widehat{State}\)
 (define (wide-step-specialized state)
@@
      (define-values (cs* \(\msto\)*)
-       (for/fold ([cs* (set)] [\(\msto\)* \(\msto\)])
+       (for/fold ([cs* \(\varnothing\)] [\(\msto\)* \(\msto\)])
          ([c cs])
@@
            [(cons \(\msto\)** cs**)
-            (values (set-union cs* cs**) (join-store \(\msto\)* \(\msto\)**))])))
-     (cons \(\msto\)* (set-union cs cs*))]))
+            (values (\(\cup\) cs* cs**) (\(\sqcup\)-store \(\msto\)* \(\msto\)**))])))
+     (cons \(\msto\)* (\(\cup\) cs cs*))]))
 \end{alltt}
@@
     [(cons \(\msto\) cs)
-     (define-values (cs* ∆)
-       (for/fold ([cs* (set)] [∆* '()])
+     (define-values (cs* \(\msdiff\))
+       (for/fold ([cs* (set)] [\(\msdiff\)* '()])
          ([c cs])
@@
            [(cons \(\msdiff\)** cs**)
-            (values (set-union cs* cs**) (append \(\msdiff\)** \(\msdiff\)*))])))
-     (cons (update \(\msdiff\) \(\msto\)) (set-union cs cs*))]))
+            (values (\(\cup\) cs* cs**) (++ \(\msdiff\)** \(\msdiff\)*))])))
+     (cons (update \(\msdiff\) \(\msto\)) (\(\cup\) cs cs*))]))
 \end{alltt}
@@
 
-\begin{verbatim}
+\begin{alltt}
  (let ((id (\(\lambda\) (x) x)))
    (begin (id 1) (id 2)))
-\end{verbatim}
+\end{alltt}
 

### 2012-11-08 paper/paper.tex

@@
 \usepackage{alltt}
-\usepackage{amsmath}
+\usepackage{amsmath,amsthm}
 \usepackage{amssymb}
 \usepackage{stmaryrd}
+
+\newtheorem{theorem}{Theorem}
 % \usepackage{xltxtra}
@@
 
+\appendix
+\section{Proof: laziness is precision-preserving}
+Given widened machine configurations, we can show that lazy
+non-determinism is precision-preserving in the cases that application
+positions are store-allocated and not. We first show the latter since
+it uses less cluttered transition rules. The high level is that the
+``fan-out'' of non-determinism is collapsed back in the store (in the
+store-allocated applications case) or delayed a single step by
+laziness that would have happened in a step of strictness.
+
+Details of lazy machine given as a diff from the strict machine. \\
+State space of $lazy-\widehat{CESK}^*_t$ varies just in
+$\mathit{Value}$ (storable values do not change):
+\begin{align*}
+\mval \in \mathit{Value} &::= z \mid b \mid o \mid \\saddr{\maddr} \\
+\end{align*}
+
+We define a relation between strict and lazy machines that illustrates
+the quotient that the lazy states
+
+\newcommand{\lapprx}[3]{#1 \approx_{#2} #3}
+\newcommand{\apprx}[3]{#1 \sim_{#2} #3}
+\newcommand{\capprx}[2]{#1 \sim #2}
+
+\begin{align*}
+\ev{\mexp, \menv, \mkont, \mcntr} \in cs &\Rightarrow \lapprx{\ev{\mexp, \menv,\mkont, \mcntr}}{\msto}{cs} \\
+\co{\mkont, \mval} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\co{\mkont, \mval}}{\msto}{cs} \\
+\ap{\mval_0, \mval_1, \mkont} \in cs \mbox{ and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \lapprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{cs} \\
+\ev{\mexp, \menv, \mkont} \in lcs &\Rightarrow \apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs} \\
+\co{\mkont, \mval} \in lcs \mbox { and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\co{\mkont, \mval'}}{\msto}{lcs} \\
+\ap{\mval_0, \mval_1, \mkont} \in lcs \mbox { and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \apprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{lcs} \\
+\forall lc \in lcs, \lapprx{lc}{\msto}{cs}, \mbox{ and }\forall c \in cs, \apprx{c}{\msto}{lcs}. &\Rightarrow \capprx{(lcs, \msto)}{(cs, \msto)}
+\end{align*}
+
+We also need metafunctions for adding and removing a store component
+from states. These are straightforwardly defined, so just name them
+$wn$ and $nw$ respectively.
+%% c2vc(<e^l r k d>,s) = <e^l r s k d>
+%% c2vc(<v k d>,s) = <v s k d>
+%% vc2c(<e^l r s k d>) = <e^l r k d>
+%% vc2c(<v s k d>) = <v k d>
+
+The property we show is that all states stay related through
+reduction. The main difficulty is in showing the resulting stores are
+in fact equal. We do this by showing each individual reduction takes
+equal stores to equal stores, and that the union of all these
+reductions is the full reduction.
+
+\begin{theorem}
+$\forall lcs,cs,lcs',\msto,\msto'$ if $\capprx{(lcs, \msto)}{(cs,\msto)}$
+  and $(lcs,\msto) \machstep (lcs',\msto')$ then there exists $cs'$
+  such that $(cs,\msto) \machstep (cs',\msto')$
+\end{theorem}
+\begin{proof}
+...
+\end{proof}
+
+%% --Eval
+%% <x r s k d> --> <v s k d> where v in s(r(x))  [VAR]
+%% <(e0^l0 e1^l1)^{l,lf,la} r s k d> --> <e0^l0 r s\sqcup[ld |-> {k}] ar^{l,lf,la}(e1^l1 r ld) d> [APP]
+%% <(lambda x e) r s k d> --> <clos(x e r) s k d> [CLOS]
+%% --Continue
+%% <v s ar^{l,lf,la}(e^le r a) d> --> <e^le r s\sqcup[lfd |-> {v}] fn^{l,la}(lfd, a)> [CO]
+%% --Apply
+%% <v s fn^{l,la}(fa, ka) d> --> <e^fl r[x |-> xd'] s'[xd' |-> s'(lad)] k d'> [AP]
+%%   where clos(x e^fl r) in s(fa)
+%%         k in s(ka)
+%%         s' = s\sqcup[lad |-> {v}]
+%%         d' = truncate(ld, K)
+
+%% This extra binding corresponds directly to the intermediate
+%% bindings introduced by ANF.
+%% Matt's ANF analyses likely have all the flavours of our "laziness" technique, but this is how we get our chocolate with our peanut butter as Olin would say.
+
+%% Compare this to lazyCESK^*t:
+%% v ::= clos(x e r) | addr(a)
+%% Clos ::= clos(x e r)
+%% s : a -> P(Clos)
+%% others the same
+
+%% force(s, addr(a)) = s(a)
+%% force(s, clos(x e r)) = {clos(x e r)}
+
+%% --Eval
+%% <x r s k d> --> <addr(r(x)) s k d>
+%% others the same
+%% --Continue
+%% <v s ar^{l,lf,la}(e^le r a) d> --> <e^le r s\sqcup[lfd |-> force(s,v)] fn^{l,la}(lfd, a)>
+%% --Apply
+%% <v s fn^{l,la}(fa, ka) d> --> <e^fl r[x |-> xd'] s'[xd' |-> s'(lad)] k d'>
+%%   where clos(x e^fl r) in s(fa)
+%%         k in s(ka)
+%%         s' = s\sqcup[lad |-> force(s, v)]
+%%         d' = truncate(ld, K)
+%% The same globalizing stuff applies.
+
+%% Let us relate the state spaces of the two abstract machines for the lambda calculus in the following way
+
+%% Then we can show that if lazyconf ~ conf and lazyconf ==> lazyconf' then there exists a conf' such that conf ==> conf' and lazyconf' ~ conf'
+%% Let lc in lazyconf.cs be arbitrary (call lazyconf.s, s)
+%% By cases on lc:
+%% <x r k d>:
+%% Premises:
+%% <x r k d> in conf.cs
+%% Thus:
+%% Since
+%% <x r s k d> -->lazy   <addr(r(x)) s k d>
+%% <x r s k d> -->strict <v s k d> where v in s(r(x))
+%% By definition of force, v in force(addr(r(x)), s)
+%% Thus the first rule of ~_s applies and the second rule of l~_s applies.
+
+%% <(lambda x e) r k d>:
+%% Trivial
+
+%% <(e0^l0 e1^l1)^{l,lf,la} r k d>:
+%% Premises:
+%% <(e0^l0 e1^l1)^{l,lf,la} r k d> in conf.cs
+%% Thus:
+%% <(e0^l0 e1^l1)^{l,lf,la} r s k d> -->lazy   <e0^l0 r s\sqcup[ld |-> {k}] ar^{l,lf,la}(e1^l1 r ld) d>
+%% <(e0^l0 e1^l1)^{l,lf,la} r s k d> -->strict <e0^l0 r s\sqcup[ld |-> {k}] ar^{l,lf,la}(e1^l1 r ld) d>
+
+%% Stores are same in this case.
+
+%% <v ar^{l,lf,la}(e^le r a) d>:
+%% Premises:
+%% For all v' in force(v, s),
+%%   <v' ar^{l,lf,la}(e^le r a) d> in conf.cs
+%% Thus:
+%% <v ar^{l,lf,la}(e^le r ka) d> -->lazy   <e^le r s\sqcup[lfd |-> force(v, s)] fn^{l,la}(lfd ka) d>
+%% for all v' in force(v, s),
+%% <v' ar^{l,lf,la}(e^le r ka) d> -->strict <e^le r s\sqcup[lfd |-> {v'}] fn^{l,la}(lfd ka) d>
+%% Thus the union of all the strict stores will equal the lazy store.
+
+%% <v fn^{l,la}(fa ka) d>:
+%% Premises:
+%% For all v' in force(v, s),
+%%   <v' fn^{l,la}(fa ka) d> in conf.cs
+%% Thus
+%% <v fn^{l,la}(fa ka) d> -->lazy <e^fl r[x |-> xd'] s'\sqcup[xd' |-> s'(lad)] k d'>
+%% where k in s(ka)
+%%       clos(x e^fl r) in s(fa)
+%%       d' = truncate(ld, K)
+%%       s' = s\sqcup[lad |-> force(v, s)]
+%% For all v' in force(v,s),
+%% <v' fn^{l,la}(fa ka) d> -->strict <e^fl r[x |-> xd'] s'\sqcup[xd' |-> s'(lad)] k d'>
+%% where k in s(ka)
+%%       clos(x e^fl r) in s(fa)
+%%       d' = truncate(ld, K)
+%%       s' = s\sqcup[lad |-> {v'}]
+
+%% Thus the union of all the strict stores will equal the lazy store.
+
 \end{document}

### 2012-11-08 paper/paper.tex
Resolve conflicts in paper.

@@
-\documentclass[preprint]{sigplanconf}
+\documentclass[preprint,onecolumn,9pt]{onecol} %{sigplanconf}
 \usepackage{alltt}
@@
 \usepackage{stmaryrd}
+\usepackage{url}
 
@@
 \begin{abstract}
-Recently \emph{abstracting abstract machines} has been proposed as a
-systematic approach to designing sound and computable program analyses
-for higher-order programming languages.  The approach has been applied
-to a wide variety of languages with features widely considered
-difficult to analyze
-
-The abstracting abstract machines approach to program analysis makes
-it easy to design and implement sound and computable abstract
-interpretations of higher-order programs.
-%
-However, the straightforward implementation is generally
-not very fast.
-%
+Abstracting abstract machines has recently been proposed as a
+lightweight approach to designing sound and computable program
+analyses.  The approach involves deriving abstract interpreters from
+existing machine semantics and has been applied to a variety of
+languages with features widely considered difficult to analyze.
+Although analyzers are straightforward to build under this approach,
+the resulting analyzer is prohibitively inefficient.
+
 This article contributes a step by step process for going from a naive
-abstract$^2$ machine to an efficient program analyzer that is (nearly)
-three orders of magnitude faster than the starting point, yet just as
-precise (and purely functional).
+analyzer derived under the abstracting abstract machine approach, to
+an efficient program analyzer.
 \end{abstract}
@@
 \section{Introduction}
+
+The \emph{abstracting abstract machines}
+approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
+to deriving program analyses provides a straightforward way of
+transforming a programming language semantics in the form of an
+abstract machine, into a family of abstract interpreters parameterized
+over policies for regulating the analytic precision of the control,
+environment, store, and base value domains.  By taking a
+machine-oriented view of computation, it becomes possible to design,
+verify, and implement program analyzers for realistic language
+features typically considered difficult to model.  The approach was
+originally applied to features such as higher-order functions,
+stack-inspection, exceptions, laziness, first-class continuations, and
+garbage collection.  It has since been used to verify actor-based
+parallelism~\cite{local:DOsualdo:12A} and behavioral
+contracts~\cite{FIXME:tobin-hochstadt}; it has been used to model
+Dalvik, Erlang~\cite{local:DOsualdo:12B},
+JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, Racket, and
+Coq~\cite{local:harvard}.
+
+The primary strength of the approach is that abstract interpreters can
+be easily derived through a small number of steps from existing
+machine models.% such as the CESK machine or the Lazy Krivine machine.
+Moreover, since the relationships between abstract machines and
+higher-level semantic models---such as definitional
+interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational
+semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction
+semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well
+understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
+these semantic models to sound program analyzers.
+
+%% Using this approach has also lead to new insights in pushdown
+%% abstractions of programs capable of precisely modeling the program
+%% stack, even in the presence of control effects.
 
@@
 
-\bibliographystyle{splncs}
-\bibliography{bibliography}
+\bibliographystyle{plain}
+\bibliography{local,bibliography}
 

### 2012-11-09 paper/paper.tex
Proof checkpoint.

@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
 % \usepackage{xltxtra}
@@
 \begin{align*}
-force(\clos{\mvar,\mexp,\menv},\msto) &= \{\clos{\mvar,\mexp,\menv}\}\\
-force(\saddr\maddralt,\msto) &= \msto(\maddralt)\\
+force(\msto,\clos{\mvar,\mexp,\menv}) &= \{\clos{\mvar,\mexp,\menv}\}\\
+force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
 \ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\; 
@@
 \ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}} \\
-\mbox{where } &\mval' \in force(\mval,\msto) \\
+\mbox{where } &\mval' \in force(\msto,\mval) \\
 bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
-  \maddr], \msto\sqcup[\maddr \mapsto force(\mval, \msto)], \mcntr' \\
+  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
 \mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
@@
 \begin{align*}
-\mval \in \mathit{Value} &::= z \mid b \mid o \mid \\saddr{\maddr} \\
+\mval \in \mathit{Value} &::= z \mid b \mid o \mid \saddr{\maddr} \\
 \end{align*}
@@
 reduction. The main difficulty is in showing the resulting stores are
-in fact equal. We do this by showing each individual reduction takes
-equal stores to equal stores, and that the union of all these
-reductions is the full reduction.
-
-\begin{theorem}
+in fact equal. We do this by showing each lazy reduction corresponds
+to a set of strict reductions, for which the union of their output
+stores equal the lazy store. Also for strict equality, every strict
+reduction corresponds to a lazy reduction (immediate from the relation).
+
+\begin{lemma}[Partition step]
+$\forall lc, lc', cs, \msto, \msto'.$ if $\lapprx{lc}{\msto}{cs}$ and $wn(lc,\msto) \machstep wn(lc', \msto')$ then
+$\exists cs'. \forall c \in cs. \exists c' \in cs', \msto^c. wn(c,\msto) \machstep wn(c', \msto^c)$ and $\msto' = \bigcup\limits_{c \in cs}{\msto^c}$ and $\lapprx{lc'}{\msto'}{cs'}$.
+\end{lemma}
+\begin{proof}
+By cases on $wn(lc,\msto) \machstep wn(lc', \msto')$:
+\begin{itemize}
+\item{Case $\ev{\svar{\mvar}, \menv, \mkont} \machstep
+            \co{\mkont, \maddr{\menv(\mvar)}, \msto}$: \\
+      If $\mvar \in dom(\menv)$ and $\menv(x) \in dom(\msto)$, then $wn(lc, \msto)$ steps
+      \begin{itemize} 
+       \item{lazily to $\co{\saddr{\menv(\mvar)}, \msto, \mkont}$}
+       \item{strictly to each of $\{\co{\mval, \msto, \mkont} \mid \mval \in \msto(\menv(\mvar))\}$}
+      \end{itemize}
+      By definition of ...}
+\item{Case $\ev{\slit{\mlit}, \menv, \msto, \mkont} \machstep
+            \co{\mkont, \mlit, \msto}$: \\
+      Immeditate.}
+\item{Case $\ev{\slam{\mvar}{\mexp}, \menv, \mkont} \machstep
+            \co{\mkont, \clos{\mvar, \mexp, \menv}, \msto}$: \\
+      Immediate}
+\item{Case $\ev[^\mcntr]{\sapp[^\mlab]{\mexp_0}{\mexp_1}, \menv, \mkont} \machstep
+            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kar[^\mcntr_\mlab]{\mexp_1, \menv, \maddr}}$: \\
+      Where $a, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$
+      ...}
+\item{Case $\ev[^\mcntr]{\sif[^\mlab]{\mexp_0}{\mexp_1}{\mexp_2}, \menv, \mkont} \machstep
+            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kif[^\mcntr]{\mexp_1, \mexp_2, \menv, \maddr}}$: \\
+      Where $a, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$            
+      ...}
+\item{Case $\co{\kmt, \mval, \msto} \machstep \ans{\msto, \mval'}$: \\
+      Where $\mval' \in force(\msto, \mval)$
+      ...}
+\item{Case $\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval}$: \\
+      ...}
+\item{Case $\co{\kfn[^\mcntr_\mlab]{\mvalx{u}, \maddr}, \mval}$: \\
+      ...}
+\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \strue}$: \\
+      ...}
+\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \sfalse}$: \\
+      ...}
+\item{Case $\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval, \mkont}$: \\
+      ...}
+\item{Case $\ap[^\mcntr_\mlab]{\mop, \mval, \mkont}$: \\
+      ...}
+\end{itemize}
+\end{proof}
+
+\begin{theorem}[Laziness preserves precision]
 $\forall lcs,cs,lcs',\msto,\msto'$ if $\capprx{(lcs, \msto)}{(cs,\msto)}$
@@
 \begin{proof}
+Let $lc \in lcs$ be arbitrary and let $\hat{c} \subseteq cs$ be the
+smallest set such that $\lapprx{lc}{\msto}{\hat{c}}$. \\
 ...
@@
 %% <x r s k d> -->strict <v s k d> where v in s(r(x))
-%% By definition of force, v in force(addr(r(x)), s)
+%% By definition of force, v in force(s, addr(r(x)))
 %% Thus the first rule of ~_s applies and the second rule of l~_s applies.
@@
 %% Premises:
-%% For all v' in force(v, s),
+%% For all v' in force(s, v),
 %%   <v' ar^{l,lf,la}(e^le r a) d> in conf.cs
 %% Thus:
-%% <v ar^{l,lf,la}(e^le r ka) d> -->lazy   <e^le r s\sqcup[lfd |-> force(v, s)] fn^{l,la}(lfd ka) d>
-%% for all v' in force(v, s),
+%% <v ar^{l,lf,la}(e^le r ka) d> -->lazy   <e^le r s\sqcup[lfd |-> force(s, v)] fn^{l,la}(lfd ka) d>
+%% for all v' in force(s,v),
 %% <v' ar^{l,lf,la}(e^le r ka) d> -->strict <e^le r s\sqcup[lfd |-> {v'}] fn^{l,la}(lfd ka) d>
@@
 %% Premises:
-%% For all v' in force(v, s),
+%% For all v' in force(s,v),
 %%   <v' fn^{l,la}(fa ka) d> in conf.cs
@@
 %%       d' = truncate(ld, K)
-%%       s' = s\sqcup[lad |-> force(v, s)]
-%% For all v' in force(v,s),
+%%       s' = s\sqcup[lad |-> force(s,v)]
+%% For all v' in force(s,v),
 %% <v' fn^{l,la}(fa ka) d> -->strict <e^fl r[x |-> xd'] s'\sqcup[xd' |-> s'(lad)] k d'>

### 2012-11-09 paper/paper.tex
Initial plot of numbers.
Some more citations.

--- 2012-11-08 paper/paper.tex
@@
-\documentclass[preprint,onecolumn,9pt]{onecol} %{sigplanconf}
+\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol} 
 \usepackage{alltt}
@@
 \usepackage{url}
+\usepackage{graphicx}
 
@@
 stack-inspection, exceptions, laziness, first-class continuations, and
-garbage collection.  It has since been used to verify actor-based
-parallelism~\cite{local:DOsualdo:12A} and behavioral
-contracts~\cite{FIXME:tobin-hochstadt}; it has been used to model
-Dalvik, Erlang~\cite{local:DOsualdo:12B},
-JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, Racket, and
+garbage collection.  It has since been used to verify actor-
+\cite{local:DOsualdo:12A} and
+thread-based~\cite{dvanhorn:Might2011Family} parallelism and
+behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
+has been used to model Dalvik, Erlang~\cite{local:DOsualdo:12B},
+JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467},
+Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}, and
 Coq~\cite{local:harvard}.
@@
 
+\begin{figure*}
+\begin{center}
+\includegraphics[width=6in]{rel-time.ps}
+\end{center}
+\end{figure*}
+
 \cite{dvanhorn:Earl2012Introspective}

### 2012-11-09 paper/paper.tex

--- 2012-11-09 paper/paper.tex
@@
 \section{Evaluation}
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=6in]{rel-time.ps}
+\end{center}
+\end{figure*}
 

### 2012-11-09 paper/paper.tex
Currently stuck in the proof since partitioning isn't helping do case analysis on ifs and applications.

@@
 be easily derived through a small number of steps from existing
-machine models.% such as the CESK machine or the Lazy Krivine machine.
+machine models. % such as the CESK machine or the Lazy Krivine machine.
 Moreover, since the relationships between abstract machines and
@@
 \co{\mkont,\mval',\msto}
-\text{ where }\mkont\in\msto(\maddr)
-\text{ and } \mval'\in\interpdelta(\mop,\mval)
+\text{ where } \mval'\in\interpdelta(\mop,\mval)
 \end{align*}
@@
     ([x (in-set s)])
-    (set-union i (f x))))
+    (\(\cup\) i (f x))))
 
@@
         (let ((new-front ((appl f) front)))
-          (loop (set-union accum front)
-                (set-subtract new-front accum))))))
+          (loop (\(\cup\) accum front)
+                (new-front . \(\setminus\) . accum))))))
 \end{alltt}
@@
      (set (cons (for/set ([s ss]) (s->c s))
-                (join-stores ss)))]))
+                (\(\sqcup\)-stores ss)))]))
 \end{alltt}
@@
 \ap{\mval_0, \mval_1, \mkont} \in cs \mbox{ and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \lapprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{cs} \\
+\ans{\mval'} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\ans{\mval}}{\msto}{cs} \\
 \ev{\mexp, \menv, \mkont} \in lcs &\Rightarrow \apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs} \\
@@
 \ap{\mval_0, \mval_1, \mkont} \in lcs \mbox { and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \apprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{lcs} \\
+\ans{\mval} \in lcs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\ans{\mval'}}{\msto}{lcs} \\
 \forall lc \in lcs, \lapprx{lc}{\msto}{cs}, \mbox{ and }\forall c \in cs, \apprx{c}{\msto}{lcs}. &\Rightarrow \capprx{(lcs, \msto)}{(cs, \msto)}
@@
 \begin{lemma}[Partition step]
-$\forall lc, lc', cs, \msto, \msto'.$ if $\lapprx{lc}{\msto}{cs}$ and $wn(lc,\msto) \machstep wn(lc', \msto')$ then
-$\exists cs'. \forall c \in cs. \exists c' \in cs', \msto^c. wn(c,\msto) \machstep wn(c', \msto^c)$ and $\msto' = \bigcup\limits_{c \in cs}{\msto^c}$ and $\lapprx{lc'}{\msto'}{cs'}$.
+$\forall lc, lcs', cs, \msto, \msto'.$
+ if  $\capprx{(\{lc\},\msto)}{(cs,\msto)}$
+ and $\forall lc' \in lcs'. wn(lc,\msto) \machstep wn(lc', \msto')$ then
+ $\exists cs'. \forall c \in cs. \exists c' \in cs', \msto^c$
+ such that $wn(c,\msto) \machstep wn(c', \msto^c)$
+ and       $\msto' = \bigcup\limits_{c \in cs}{\msto^c}$
+ and       $\capprx{(lcs',\msto')}{(cs',\msto')}$.
 \end{lemma}
 \begin{proof}
+Let $lc' \in lcs'$ be arbitrary.
 By cases on $wn(lc,\msto) \machstep wn(lc', \msto')$:
 \begin{itemize}
-\item{Case $\ev{\svar{\mvar}, \menv, \mkont} \machstep
-            \co{\mkont, \maddr{\menv(\mvar)}, \msto}$: \\
-      If $\mvar \in dom(\menv)$ and $\menv(x) \in dom(\msto)$, then $wn(lc, \msto)$ steps
-      \begin{itemize} 
-       \item{lazily to $\co{\saddr{\menv(\mvar)}, \msto, \mkont}$}
-       \item{strictly to each of $\{\co{\mval, \msto, \mkont} \mid \mval \in \msto(\menv(\mvar))\}$}
-      \end{itemize}
-      By definition of ...}
-\item{Case $\ev{\slit{\mlit}, \menv, \msto, \mkont} \machstep
+\item{Case $\ev{\svar{\mvar}, \menv, \mkont, \msto} \machstep \co{\mkont,
+    \saddr{\menv(\mvar)}, \msto}$: \\ By definition of $\machstep$,
+  $wn(lc,\msto)$ steps strictly to each of $cs' = \{\co{\mval, \mkont} \mid \mval \in \msto(\menv(\mvar))\}$.  By definition of
+  $\capprx{}{}$, $cs = \{lc\}$. The stores are the same, and by
+  definition of $\lapprx{}{\msto}{}$, $\lapprx{lc'}{\msto}{cs'}$.}
+\item{Case $\ev{\slit{\mlit}, \menv, \msto, \mkont, \msto} \machstep
             \co{\mkont, \mlit, \msto}$: \\
       Immeditate.}
-\item{Case $\ev{\slam{\mvar}{\mexp}, \menv, \mkont} \machstep
+\item{Case $\ev{\slam{\mvar}{\mexp}, \menv, \mkont, \msto} \machstep
             \co{\mkont, \clos{\mvar, \mexp, \menv}, \msto}$: \\
       Immediate}
-\item{Case $\ev[^\mcntr]{\sapp[^\mlab]{\mexp_0}{\mexp_1}, \menv, \mkont} \machstep
-            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kar[^\mcntr_\mlab]{\mexp_1, \menv, \maddr}}$: \\
-      Where $a, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$
+\item{Case $\ev[^\mcntr]{\sapp[^\mlab]{\mexp_0}{\mexp_1}, \menv, \mkont, \msto} \machstep
+            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kar[^\mcntr_\mlab]{\mexp_1, \menv, \maddr}, \msto'}$: \\
+      Where $\maddr, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$ \\
+      Immediate.}
+\item{Case $\ev[^\mcntr]{\sif[^\mlab]{\mexp_0}{\mexp_1}{\mexp_2}, \menv, \mkont, \msto} \machstep
+            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kif[^\mcntr]{\mexp_1, \mexp_2, \menv, \maddr}, \msto'}$: \\
+      Where $\maddr, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$ \\
+      Immediate.}
+\item{Case $\co{\kmt, \mval, \msto} \machstep \ans{\msto, \mval'}$: \\
+      Where $\mval' \in force(\msto, \mval)$ \\
+      By definition of $\capprx{}{}$, $cs = \{\co{\kmt, \mval'} \mid \mval \in force(\msto, \mval)\}$.
+      By definition of $\machstep$, $wn(lc,\msto)$ steps strictly to
+      each of $cs' = \{\ans{\mval}\}$. The stores are the same.
+      By definition of $\lapprx{}{\msto}{}$, $\forall lc' \in lcs'. \lapprx{lc'}{\msto}{cs'}$.
+      By definition of $\apprx{}{\msto}{}$, $\forall c' \in cs'. \apprx{c'}{\msto}{lcs'}$.
+      Thus $\capprx{(lcs',\msto)}{(cs',\msto)}$.}
+\item{Case $\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval, \msto} \machstep
+            \ev[^\mcntr]{\mexp, \menv, \kfn[^\mcntr_\mlab]{\mval',\maddr}, \msto}$: \\
+      Where $v' \in force(\msto, \mval)$. \\
+      By definition of $\apprx{}{\msto}{}$, 
+        $cs = \{\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval'} \mid \mval \in force(\msto, \mval)\}$.
+      By definition of $\machstep, \lapprx{}{\msto}{}$,
+        $cs' = \{\ev[^\mcntr]{\mexp, \menv, \kfn[^\mcntr_\mlab]{\mval',\maddr}}
+            \mid \co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval'} \in cs\}$.
+      The stores are the same, and the previous imply $\capprx{(lcs',\msto)}{(cs',\msto)}$.}
+\item{Case $\co{\kfn[^\mcntr_\mlab]{\mvalx{u}, \maddr}, \mval, \msto} \machstep
+            \ap{\mvalx{u}, \mval, \mkont, \msto}$: \\
+      Where $\mkont \in \msto(\maddr)$. \\
+      Immediate.}
+\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
+            \ev{\mexp_i, \menv, \mkont, \msto}$: \\
+      Where $\mkont \in \msto(\maddr)$. \\
+      By cases on $\{\strue, \sfalse\} \cap force(\msto, \mval)$: \\
+ \begin{itemize}
+  \item{Case $\{\strue, \sfalse\}$:}
+  \item{Case $\{\strue\}$:}
+  \item{Case $\{\sfalse\}$:}
+ \end{itemize}
  ...}
-\item{Case $\ev[^\mcntr]{\sif[^\mlab]{\mexp_0}{\mexp_1}{\mexp_2}, \menv, \mkont} \machstep
-            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kif[^\mcntr]{\mexp_1, \mexp_2, \menv, \maddr}}$: \\
-      Where $a, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$            
+%% \item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
+%%             \ev{\mexp_1, \menv, \mkont, \msto}$: \\
+%%       Where $\sfalse \in force(\msto, \mval)$ and $\mkont \in \msto(\maddr)$. \\
+%%       ...}
+\item{Case $\ap[^\mcntr_\mlab]{\mvalx{u}, \mval, \mkont,\msto} \machstep
+            \ev[^{\mcntr'}]{\mexp, \menv', \mkont, \msto'}$: \\
+      {\bf BREAKDOWN}
+      Where $\clos{\mvar, \mexp, \menv} \in force(\msto, \mvalx{u})$
+        and $\menv', \msto', \mcntr' = \mathit{bind}^\mcntr_\mlab(\menv, \msto, \mvar, \mval)$
       ...}
-\item{Case $\co{\kmt, \mval, \msto} \machstep \ans{\msto, \mval'}$: \\
-      Where $\mval' \in force(\msto, \mval)$
-      ...}
-\item{Case $\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval}$: \\
-      ...}
-\item{Case $\co{\kfn[^\mcntr_\mlab]{\mvalx{u}, \maddr}, \mval}$: \\
-      ...}
-\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \strue}$: \\
-      ...}
-\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \sfalse}$: \\
-      ...}
-\item{Case $\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval, \mkont}$: \\
-      ...}
-\item{Case $\ap[^\mcntr_\mlab]{\mop, \mval, \mkont}$: \\
+\item{Case $\ap[^\mcntr_\mlab]{\mvalx{u}, \mval, \mkont,\msto} \machstep
+            \co{\mval'', \mkont, \msto}$: \\
+      {\bf BREAKDOWN}
+      Where $\mop \in force(\msto, \mvalx{u})$, $\mval' \in force(\msto, \mval)$
+        and $\mval'' \in\interpdelta(\mop,\mval')$. \\
       ...}

### 2012-11-10 paper/paper.tex
Changed some wording for precision preserving intro.

@@
 \documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
-\usepackage{alltt}
+\usepackage{alltt,mathpartir}
 \usepackage{amsmath,amsthm}
@@
 We discuss a variety of techniques for increasing performance by
-decreasing the interpretive overhead of abstract interpretation, all
-in the purely functional realm. These techniques are the following:
+decreasing the interpretive overhead of abstract interpretation. They
+come in two flavors: decreasing the size of the represented state
+space [SS], and decreasing intermediate data structures [DS].
+
 \begin{itemize}
- \item{shrink state space by quotients (lazy non-determinism)}
- \item{shrink state space by removing {\tt ev} states (abstract
-   compilation)}
- \item{decrease intermediate data structures by specializing our
-fixpoint operator to work with widened machine configurations
-(fixpoint specialization)}
- \item{decrease intermediate data structures by operating on store
-   changes rather than whole stores (store deltas)}
+ \item{[SS] quotient by sets of values (lazy non-determinism)}
+ \item{[SS] remove {\tt ev} states (abstract compilation)}
+ \item{[DS] specialize the fixpoint operator to work with widened
+   machine configurations (fixpoint specialization)}
+ \item{[DS] operate on store changes rather than whole stores (store deltas)}
+ \item{[DS] represent the store as a global, mutable hash and
+   determine a state's store by the number of changes made to the
+   global store at the time of visiting (timestamp)}
+ \item{[DS] use a global mutable workset}
+ \item{[DS] represent the global store as a global, mutable vector in
+   the 0-CFA case. (preallocation)}
 \end{itemize}
 Each technique requires minimal changes to the reduction
-relation/fixpoint, and each improves performance by integer factors.
+relation/fixpoint, and each improves performance by integer
+factors. The first four remain in the purely functional realm. The
+final three use mutation in small, localized places, so they do not
+increase conceptual complexity drastically.
 
@@
 
-\subsection{Computing with store changes}
+\subsection{Store deltas}
 
@@
 % For David to fill in.
-\subsection{Imperative global store and worklist}
+\subsection{Timestamp}
 ...
@@
 We define a relation between strict and lazy machines that illustrates
-the quotient that the lazy states
-
+the quotient that the lazy states embody.
+
+\newcommand{\vapprx}[3]{#1 \cong_{#2} #3}
+\newcommand{\kapprx}[3]{#1 \cong_{#2} #3}
 \newcommand{\lapprx}[3]{#1 \approx_{#2} #3}
@@
 
-\begin{align*}
-\ev{\mexp, \menv, \mkont, \mcntr} \in cs &\Rightarrow \lapprx{\ev{\mexp, \menv,\mkont, \mcntr}}{\msto}{cs} \\
-\co{\mkont, \mval} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\co{\mkont, \mval}}{\msto}{cs} \\
-\ap{\mval_0, \mval_1, \mkont} \in cs \mbox{ and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \lapprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{cs} \\
-\ans{\mval'} \in cs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \lapprx{\ans{\mval}}{\msto}{cs} \\
-\ev{\mexp, \menv, \mkont} \in lcs &\Rightarrow \apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs} \\
-\co{\mkont, \mval} \in lcs \mbox { and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\co{\mkont, \mval'}}{\msto}{lcs} \\
-\ap{\mval_0, \mval_1, \mkont} \in lcs \mbox { and } \mval_i' \in force(\msto, \mval_i) &\Rightarrow \apprx{\ap{\mval_0', \mval_1', \mkont}}{\msto}{lcs} \\
-\ans{\mval} \in lcs \mbox{ and } \mval' \in force(\msto, \mval) &\Rightarrow \apprx{\ans{\mval'}}{\msto}{lcs} \\
-\forall lc \in lcs, \lapprx{lc}{\msto}{cs}, \mbox{ and }\forall c \in cs, \apprx{c}{\msto}{lcs}. &\Rightarrow \capprx{(lcs, \msto)}{(cs, \msto)}
-\end{align*}
+\begin{mathpar}
+\inferrule{\mval' \in force(\msto, \mval)}
+          {\vapprx{\mval}{\msto}{\mval'}} \quad
+\inferrule{\ans{\mval} \in cs}{\lapprx{\ans{\mval}}{\msto}{cs}} \quad
+\inferrule{\ev{\mexp, \menv, \mkont, \mcntr} \in cs}
+          {\lapprx{\ev{\mexp, \menv,\mkont, \mcntr}}{\msto}{cs}} \\
+\inferrule{\co{\mkont, \mval'} \in cs \\
+           \vapprx{\mval}{\msto}{\mval'}}
+          {\lapprx{\co{\mkont, \mval}}{\msto}{cs}} \quad
+\inferrule{\ap{\mval_0, \mval_1, \mkont} \in cs \\
+           \vapprx{\mval_1}{\msto}{\mval_1'}}
+          {\lapprx{\ap{\mval_0, \mval_1', \mkont}}{\msto}{cs}} \\
+\inferrule{\ev{\mexp, \menv, \mkont} \in lcs}
+          {\apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs}} \quad
+\inferrule{\ans{\mval} \in lcs \\
+           \vapprx{\mval}{\msto}{\mval'}}
+          {\apprx{\ans{\mval'}}{\msto}{lcs}} \\
+\inferrule{\co{\mkont, \mval} \in lcs \\
+           \vapprx{\mval}{\msto}{\mval'}}
+          {\apprx{\co{\mkont, \mval'}}{\msto}{lcs}} \quad
+\inferrule{\ap{\mval_0, \mval_1, \mkont} \in lcs \\
+           \vapprx{\mval_1}{\msto}{\mval_1'}}
+          {\apprx{\ap{\mval_0, \mval_1', \mkont}}{\msto}{lcs}} \\
+\inferrule{\forall lc \in lcs, \lapprx{lc}{\msto}{cs} \\
+           \forall c \in cs, \apprx{c}{\msto}{lcs}}
+          {\capprx{(lcs, \msto)}{(cs, \msto)}}
+\end{mathpar}
 
@@
 \item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
-            \ev{\mexp_i, \menv, \mkont, \msto}$: \\
+            \ev{\mexp_0, \menv, \mkont, \msto}$: \\
       Where $\mkont \in \msto(\maddr)$. \\
-      By cases on $\{\strue, \sfalse\} \cap force(\msto, \mval)$: \\
- \begin{itemize}
-  \item{Case $\{\strue, \sfalse\}$:}
-  \item{Case $\{\strue\}$:}
-  \item{Case $\{\sfalse\}$:}
- \end{itemize}
- ...}
-%% \item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
-%%             \ev{\mexp_1, \menv, \mkont, \msto}$: \\
-%%       Where $\sfalse \in force(\msto, \mval)$ and $\mkont \in \msto(\maddr)$. \\
-%%       ...}
-\item{Case $\ap[^\mcntr_\mlab]{\mvalx{u}, \mval, \mkont,\msto} \machstep
+      By definition of $\machstep$, $\strue \in force(\msto, \mval)$, thus by definition of
+      $\capprx{}{}$, the strict machine takes the same step.}
+\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
+            \ev{\mexp_1, \menv, \mkont, \msto}$: \\
+      Where $\mkont \in \msto(\maddr)$. \\
+      By definition of $\machstep$, $\sfalse \in force(\msto, \mval)$, thus by definition of
+      $\capprx{}{}$, the strict machine takes the same step.}
+\item{Case $\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval, \mkont,\msto} \machstep
             \ev[^{\mcntr'}]{\mexp, \menv', \mkont, \msto'}$: \\
-      {\bf BREAKDOWN}
-      Where $\clos{\mvar, \mexp, \menv} \in force(\msto, \mvalx{u})$
-        and $\menv', \msto', \mcntr' = \mathit{bind}^\mcntr_\mlab(\menv, \msto, \mvar, \mval)$
-      ...}
-\item{Case $\ap[^\mcntr_\mlab]{\mvalx{u}, \mval, \mkont,\msto} \machstep
+      Where $\menv', \msto', \mcntr' = \mathit{bind}^\mcntr_\mlab(\menv, \msto, \mvar, \mval)$
+      By definition of $\apprx{}{\msto}{}$,
+        $cs = \{\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval', \mkont} \mid
+                \mval' \in force(\msto, \mval)\}$.
+      Thus $cs' = \{\ev[^{\mcntr'}]{\mexp, \menv', \mkont}\}$ (clearly $\lapprx{lc'}{\msto'}{cs'}$)  and
+      for all $c \in cs$, $wn(c,\msto) \machstep wn(c',\msto\sqcup[\mvar\mcntr' \mapsto \{c.\mval\}])$.
+      Thus the union of these stores is equal to $\msto'$.}
+\item{Case $\ap[^\mcntr_\mlab]{\mop, \mval, \mkont,\msto} \machstep
             \co{\mval'', \mkont, \msto}$: \\
-      {\bf BREAKDOWN}
-      Where $\mop \in force(\msto, \mvalx{u})$, $\mval' \in force(\msto, \mval)$
+      Where $\mval' \in force(\msto, \mval)$
         and $\mval'' \in\interpdelta(\mop,\mval')$. \\
-      ...}
+      By definition of $\capprx{}{}$ and $\interpdelta$,
+        $cs = \{\ap[^\mcntr_\mlab]{\mop, \mvalx{u}, \mkont} \mid
+                \mvalx{u} \in force(\msto, \mval), 
+                \mval'' \in\interpdelta(\mop,\mvalx{u})\}$.
+      Thus the strict machine takes the same step.}
 \end{itemize}
@@
 \begin{proof}
-Let $lc \in lcs$ be arbitrary and let $\hat{c} \subseteq cs$ be the
-smallest set such that $\lapprx{lc}{\msto}{\hat{c}}$. \\
-...
+By definition of $\machstep$,
+$lcs' = \{lc' \mid \exists \msto^{lc}. wn(lc,\msto) \machstep wn(lc', \msto^{lc}), lc \in lcs\}$
+For each $lc \in lcs$ let $\hat{c}^{lc} \subseteq cs$ be the
+smallest set such that $\lapprx{lc}{\msto}{\hat{c}^{lc}}$. \\ By the
+previous lemma, there exists a $\hat{c}^{lc}{}'$ such that $\capprx{(\{lc' \mid wn(lc,\msto) \machstep
+  wn(lc', \msto^{lc})\}, \msto^{lc})}{(\hat{c}',\msto^{lc})}$. Let
+$cs' = \bigcup\limits_{lc \in lcs}{\hat{c}^{lc}{}'}$ and $\msto' = \bigcup\limits_{lc \in lcs}{\msto^{lc}}$.
+Since each step individually is related, the union is related: $\capprx{lcs'}{\msto'}{cs'}$.
+By definition of $\machstep$, $(cs, \msto) \machstep (cs', \msto')$ and $(lcs, \msto) \machstep (lcs', \msto')$.
 \end{proof}

### 2012-11-11 paper/paper.tex

@@
 We discuss a variety of techniques for increasing performance by
-decreasing the interpretive overhead of abstract interpretation. They
-come in two flavors: decreasing the size of the represented state
-space [SS], and decreasing intermediate data structures [DS].
-
+decreasing the interpretive overhead of abstract interpretation, all
+in the purely functional realm. These techniques are the following:
 \begin{itemize}
- \item{[SS] quotient by sets of values (lazy non-determinism)}
- \item{[SS] remove {\tt ev} states (abstract compilation)}
- \item{[DS] specialize the fixpoint operator to work with widened
-   machine configurations (fixpoint specialization)}
- \item{[DS] operate on store changes rather than whole stores (store deltas)}
- \item{[DS] represent the store as a global, mutable hash and
-   determine a state's store by the number of changes made to the
-   global store at the time of visiting (timestamp)}
- \item{[DS] use a global mutable workset}
- \item{[DS] represent the global store as a global, mutable vector in
-   the 0-CFA case. (preallocation)}
+ \item{shrink state space by quotients (lazy non-determinism)}
+ \item{shrink state space by removing {\tt ev} states (abstract
+   compilation)}
+ \item{decrease intermediate data structures by specializing our
+fixpoint operator to work with widened machine configurations
+(fixpoint specialization)}
+ \item{decrease intermediate data structures by operating on store
+   changes rather than whole stores (store deltas)}
 \end{itemize}
 Each technique requires minimal changes to the reduction
-relation/fixpoint, and each improves performance by integer
-factors. The first four remain in the purely functional realm. The
-final three use mutation in small, localized places, so they do not
-increase conceptual complexity drastically.
+relation/fixpoint, and each improves performance by integer factors.
 
@@
 
-\subsection{Store deltas}
+\subsection{Computing with store changes}
 
@@
 % For David to fill in.
-\subsection{Timestamp}
+\subsection{Imperative global store and worklist}
 ...

### 2012-11-11 paper/paper.tex
Added prose for store widening, generic fixpoint and the baseline.

--- 2012-11-10 paper/paper.tex
@@
 \co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
-\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
+\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
 \text{ where }\mkont \in \msto(\maddr)
@@
 
-\begin{alltt}
-;; Expr \(\to\) Set[State]
-(define (eval e)
-  (fix step (set (inj e))))
-\end{alltt}
-
 We compute the semantics of a program by iterating the state
 transition relation until a fixed point in the reachable states is
-reached.
-
-\begin{alltt}
-;; \(\forall\) (X) ((X \(\to\) Set[X]) \(\to\) (Set[X] \(\to\) Set[X]))
-(define ((appl f) s)
-  (for/fold ([i (set)])
-    ([x (in-set s)])
-    (\(\cup\) i (f x))))
-
-;; Calculate fixpoint of (appl f).
-;; \(\forall\) (X) ((X \(\to\) Set[X]) Set[X] \(\to\) Set[X])
-(define (fix f s)
-  (let loop ((accum (set)) (front s))
-    (if (set-empty? front)
-        accum
-        (let ((new-front ((appl f) front)))
-          (loop (\(\cup\) accum front)
-                (new-front . \(\setminus\) . accum))))))
-\end{alltt}
+reached. We begin with a generic fixpoint combinator, ${\mathcal F}$:
+
+\begin{align*}
+{\mathcal F}(f)(X) &= {\mathcal G}(f, X, \varnothing) \\
+{\mathcal G}(f, \varnothing, S) &= S \\
+{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup f(X)) \\
+\end{align*}
+
+The $\machstep$ relation is most naturally written as a function from
+a state to a set of all next states. We want to find a fixpoint of
+$\machstep$ on the state space, however, so we must lift it to work on
+sets of states. The lifting is a natural one:
+
+\begin{align*}
+lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
+\end{align*}
+
+Then $eval(e) = {\mathcal F}(lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.
+
+If we run this on our Church numeral example, we must be
+patient. Indeed, the state space is finite, but it is exponential in
+the size of the program. With $k = 0$, the store can be in one of ${\mathcal
+  O}(2^n)$ many configurations! We take a step toward a polynomial
+complexity baseline by lifting the store out so that there is one
+shared amongst all states.
 
@@
 
-\begin{alltt}
-;; \(\widehat{State}\) = (cons (Set Conf) Store)
-
-;; (State \(\to\) Set[State]) \(\to\) \(\widehat{State}\) \(\to\) \(\widehat{State}\)
-(define ((wide-step step) state)
-  (match state
-    [(cons cs \(\msto\))
-     (define ss ((appl step)
-                 (for/set ([c cs]) (c->s c \(\msto\)))))
-     (set (cons (for/set ([s ss]) (s->c s))
-                (\(\sqcup\)-stores ss)))]))
-\end{alltt}
+A common technique to faster convergence in flow analyses is to share
+a common store that is monotonically updated as interpretation
+continues. In the ISWIM language with only unary operators, this makes
+our 0-CFA quartic in the size of the program. Not great, but
+polynomial.
+
+We can reuse the reduction relation we have defined, but with some
+scaffolding around it. A state in the store-widened machine is a pair
+of a set of ISWIM states without a store, and a store. A step of this
+machine will pair the store with each of the states in this set,
+reduce, and combine the output stores to create the next. Let $wn :
+\widehat{State} \times Store \to State$ be the metafunction that adds
+a store to a storeless state. Formally,
+
+\begin{align*}
+(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
+\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c', \msto^c), c \in cs\} \\
+             \msto' &= \bigcup\{\msto^c \mid wn(c, \msto) \machstep wn(c', \msto^c), c \in cs\}
+\end{align*}
 
 \subsection{Baseline evaluation}
+
+The final approximation we make to get to our baseline analysis is
+store-allocating results of application sub-expressions. Notice that
+the $\kfn{}$ continuation stores a value - this makes the space of
+continuations quadratic rather than linear in 0-CFA. We instead
+allocate an address for the value position when we create the
+continuation. This address and the tail address are both determined by
+the label of the application point, so the space becomes linear and
+the overall complexity drops to cubic. Note that this is a critical
+abstraction in languages with n-ary functions, since otherwise the
+continuation space becomes ${\mathcal O}(n^n)$!
+
+\begin{align*}
+\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\msto,\mkont} &\machstep
+\ev{\mexpi{0},\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
+\\
+\co{\kar{\mexp,\mlab},\mval,\msto} & \machstep
+\ev{\mexp,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
+\\
+\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
+\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
+\\
+\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
+\\
+\ap[_\mlab]{\clos{\mvar,\mexp},\mlab^a, \msto,\mkont} & \machstep
+\ev{\mexp,\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
+\\
+\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
+\co{\mkont,\mval',\msto}
+\\
+\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
+\end{align*}
+
+This extra store-allocation is effectively naming all intermediate
+results, and thus the precision aligns with an analysis specialized to
+either ANF or CPS.
 

### 2012-11-11 paper/paper.tex
Added visualizations of laziness and strictness.

@@
 \widehat{State} \times Store \to State$ be the metafunction that adds
-a store to a storeless state. Formally,
+a store to a storeless state, $nw$ removes the store, and $c.\msto$
+projects the store out of a state $c$. Formally,
 
@@
 (cs, \msto) &\widehat{\machstep} (cs', \msto') \\
-\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c', \msto^c), c \in cs\} \\
-             \msto' &= \bigcup\{\msto^c \mid wn(c, \msto) \machstep wn(c', \msto^c), c \in cs\}
+\mbox{where } CS' &= lift(\machstep)(\{ wn(c, \msto) \mid c \in cs\}) \\
+              cs' &= \{nw(C') \mid C' \in CS'\} \\
+              \msto' &= \bigcup\{C'.\msto \mid C' \in CS'\}
 \end{align*}
@@
 \begin{align*}
-\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\msto,\mkont} &\machstep
-\ev{\mexpi{0},\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
-\\
-\co{\kar{\mexp,\mlab},\mval,\msto} & \machstep
-\ev{\mexp,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
+\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
+\ev{\mexpi{0},\menv,\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
+\\
+\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
+\ev{\mexp,\menv,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
 \\
@@
 \\
-\ap[_\mlab]{\clos{\mvar,\mexp},\mlab^a, \msto,\mkont} & \machstep
-\ev{\mexp,\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
+\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
+\ev{\mexp,\menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
 \\
@@
 \begin{alltt}
-(f x y z w)
+(f x y)
 \end{alltt}
+
 where all variables have been joined with multiple values, each
 variable reference transitions to several new states that comprise the
-cartesian product of all the values of {\tt x}, {\tt y}, {\tt z} and
-{\tt w}, only to be joined back together in {\tt f}'s bindings. For
-this reason, we {\it delay} the non-determinism until it is needed in
-{\it strict contexts} (such as the guard for an {\tt if} a called
-procedure, or a numerical primitive application).
+cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
+only to be joined back together in the application position. We would see transitions of the following shape:
+\begin{center}
+\includegraphics[scale=0.3]{fanout.pdf}
+\end{center}
+For this reason, we {\it delay} the non-determinism
+until it is needed in {\it strict contexts} (such as the guard for an
+{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
+\begin{center}
+\includegraphics[scale=0.3]{lazy.pdf}
+\end{center}
 
@@
 \begin{align*}
-force(\msto,\clos{\mvar,\mexp,\menv}) &= \{\clos{\mvar,\mexp,\menv}\}\\
 force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
+force(\msto,\mval) &= \{\mval\}\\
 \ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
 \co{\mkont,\saddr{\menv(\mvar)},\msto} \\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
 &\machstep\;
-\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}} \\
-\mbox{where } &\mval' \in force(\msto,\mval) \\
+\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
 bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
@@
 \end{align*}
-
-The force for the second rule is necessary to keep the same type for
-the \(\kfn{}\) continuation, but we would rather use an address. This
-introduces (acceptably) more imprecision, but has the same precision as an ANF
-analysis\footnote{Exercise for the reader.}.
 

### 2012-11-11 paper/paper.tex
More prose for store deltas and timestamping.

@@
 
+Each step in the fixpointer for the above techniques requires joining
+several (possibly large) stores together. Even with immutable hashes
+that utilize sharing, this is memory-intensive. Worse, this
+contributes an obscene fraction of the runtime.
+
 \begin{alltt}
@@
 
-\section{Sacrificing precision for performance}
-
-Each step in the fixpointer for the above techniques requires joining
-several (possibly large) stores together. Even with immutable hashes
-that utilize sharing, this is memory-intensive. Worse, this
-contributes an obscene fraction of the runtime. We reuse a technique
-created in Shivers' dissertation~\cite{ianjohnson:Shivers:1991:CFA},
-timestamp approximation, that removes the need for joining whole
-stores by manipulating a global store imperatively. This loses
+\subsection{Timestamping stores}
+
+The store delta technique removes overhead of joining large stores,
+but for determining if a state has been visited before, we still have
+to compare entire stores - an expensive computation.  We modify a
+technique created in Shivers'
+dissertation~\cite{ianjohnson:Shivers:1991:CFA}, timestamp
+approximation, that removes the need for joining whole stores by
+manipulating a global store imperatively. The original technique loses
 precision because the starting store for each state in a machine
 configuration can change between stepping all states and introduce
-more flows than originally possible. The example in figure
-\ref{fig:timestamp} illustrates this loss in precision. {\bf TODO:}
-EXPLANATION OF LOSS. In many applications, this is acceptable for the
-performance improvement.
-
-\begin{figure}
+more flows than originally possible. However, by delaying the store
+changes until after all steps have completed, we can maintain our
+precision. In a variety of experiments, we observed that the overhead
+for storing store-deltas was outweighed by the precision keeping flows
+in check. That is to say, the analysis was faster {\it and} more precise.
+
+TODO: technical detail
+
+\subsection{Pre-allocating the store}
 TODO
-\caption{Precision lossy program in timestamp approximation}
-\label{fig:timestamp}
-\end{figure}
-
-% For David to fill in.
-\subsection{Timestamp}
-...
-\subsection{Pre-allocated global store}
-...
 
@@
         (yield (ans \(\msto\) v))))]
-       [(ar\(\superscript\mlab\) e \(\menv\))
-        (generator (do (\(\msto\)) ()
-          (yield (ev \(\msto\) e \(\menv\) (fn v l)))))
-       [(fn\(\superscript\mlab\) f)
-        (generator (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)])
-         (yield (ap \(\msto\) f v k))))]
-       [(fi\(\superscript\mlab\) c a \(\menv\))
+       [(ar\(\superscript\mcntr\subscript\mlab\) e \(\menv\) a)
         (generator
-         (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
+         (do (\(\msto\)) ([\(\msto*\) (join \(\msto\) a\(\superscript{f}\) (force \(\msto\) v))])
+           (yield (ev \(\msto*\) e \(\menv\) (fn\(\superscript\mcntr\subscript\mlab\) a\(\superscript{f}\) a)))))
+       [(fn\(\superscript\mcntr\subscript\mlab\) fa a)
+        (generator
+         (do (\(\msto\)) ([\(\msto*\) (join \(\msto\) a\(\superscript{a}\) (force \(\msto\) v))]
+                  [k <- (get \(\msto\) a)]
+                  [f <- (get \(\msto\) fa)])
+           (yield (ap\(\superscript\mcntr\subscript\mlab\) \(\msto*\) f a\(\superscript{a}\) k))))]
+       [(fi\(\superscript\mcntr\subscript\mlab\) c a \(\menv\))
+        (generator
+         (do (\(\msto\)) ([k <- (get \(\msto\) l)]
                   [v <- (force \(\msto\) v)])
            (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
-    [(ap \(\msto\) fun a k)
+    [(ap\(\superscript\mcntr\subscript\mlab\) \(\msto\) fun a k)
      (match fun
-       [(clos l x e \(\menv\))
+       [(clos x e \(\menv\))
         (generator
@@
        [(? op? o)
-        (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
+        (do (\(\msto\)) ([k <- (get \(\msto\) l)]
                  [v <- (force \(\msto\) v)]

### 2012-11-11 paper/paper.tex
Removed fixpoint specialization step.

@@
 (cs, \msto) &\widehat{\machstep} (cs', \msto') \\
-\mbox{where } CS' &= lift(\machstep)(\{ wn(c, \msto) \mid c \in cs\}) \\
-              cs' &= \{nw(C') \mid C' \in CS'\} \\
-              \msto' &= \bigcup\{C'.\msto \mid C' \in CS'\}
+\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
+              \msto' &= \bigcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
 \end{align*}
@@
  \item{[SS] remove {\tt ev} states (abstract compilation)}
- \item{[DS] specialize the fixpoint operator to work with widened
-   machine configurations (fixpoint specialization)}
  \item{[DS] operate on store changes rather than whole stores (store deltas)}
@@
 relation/fixpoint, and each improves performance by integer
-factors. The first four remain in the purely functional realm. The
+factors. The first three remain in the purely functional realm. The
 final three use mutation in small, localized places, so they do not
@@
 \newpage
-\subsection{Fixed-point specialization}
-
-\begin{alltt}
-;; \(\widehat{State} \to \widehat{State}\)
-(define (wide-step-specialized state)
-  (match state
-    [(cons \(\msto\) cs)
-     (define-values (cs* \(\msto\)*)
-       (for/fold ([cs* \(\varnothing\)] [\(\msto\)* \(\msto\)])
-         ([c cs])
-         (match (step-compiled^ (cons \(\msto\) c))
-           [(cons \(\msto\)** cs**)
-            (values (\(\cup\) cs* cs**) (\(\sqcup\)-store \(\msto\)* \(\msto\)**))])))
-     (cons \(\msto\)* (\(\cup\) cs cs*))]))
-\end{alltt}
-%% Should be commutted to beginning
-
-Special + Compile + Lazy:
-   cpu time: 14212 real time: 14681 gc time: 823
 

### 2012-11-11 paper/paper.tex
Some related work and intro.

--- 2012-11-11 paper/paper.tex
@@
 behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
-has been used to model Dalvik, Erlang~\cite{local:DOsualdo:12B},
-JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467},
-Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}, and
-Coq~\cite{local:harvard}.
+has been used to model Coq~\cite{local:harvard},
+Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
+JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
+Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.
 
@@
 be easily derived through a small number of steps from existing
-machine models. % such as the CESK machine or the Lazy Krivine machine.
-Moreover, since the relationships between abstract machines and
+machine models.  Since the relationships between abstract machines and
 higher-level semantic models---such as definitional
@@
 understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
-these semantic models to sound program analyzers.
+these high-level semantic models to sound program analyzers in a
+systematic way.  Moreover, since these analyses so closely resemble a
+language's interpreter (a) implementing an analysis requires little
+more than implementing an interpreter, (b) a single implementation can
+serve as both an interpreter and analyzer, and (c) verifying the
+correctness of the implementation is straightforward.
+
+However, there is a considerable weakness with the approach: an
+analyzer designed and implemented by following the AAM recipe is
+prohibitively inefficient without both further approximation and
+further implementation effort.
+
+In this article, we develop a systematic approach to deriving the
+feasible implementation of an abstract machine based analyzer.
 
@@
 
+\includegraphics[width=2.5in]{rel-time-church.ps}
+
 \subsection{Notation and prerequisites}
@@
 
-Notation: concrete examples of programs under analysis is given in
-(monochrome) Scheme notation.  Code is given in (syntax colored)
-Racket.
+%% Notation: concrete examples of programs under analysis is given in
+%% (monochrome) Scheme notation.  Code is given in (syntax colored)
+%% Racket.
 
@@
 \co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
-\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
+\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
 \text{ where }\mkont \in \msto(\maddr)
@@
 \begin{alltt}
-  ;; State \(\to\) Set[State]
+  ;; State \(\to\) \(\Set(\)State\()\)
   (define (step \(\mstate\))
@@
 
-%% \begin{alltt}
-%% (define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))
-
-%% (define pred
-%%   (\(\lambda\) (n)
-%%     (\(\lambda\) (rf)
-%%       (\(\lambda\) (rx)
-%%         (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
-%%           (\(\lambda\) (i) rx))
-%%          (\(\lambda\) (id) id))))))
-%% \end{alltt}
-
-%% (define \church0 (\(\lambda\) (f0) (\(\lambda\) (x0) x0)))
-%% (define \church1 (\(\lambda\) (f1) (\(\lambda\) (x1) (f1 x1))))
-%% (define \church2 (\(\lambda\) (f2) (\(\lambda\) (x2) (f2 (f2 x2)))))
-%% (define \church3 (\(\lambda\) (f3) (\(\lambda\) (x3) (f3 (f3 (f3 x3))))))
-
 \begin{figure}
 \begin{alltt}
+(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
+(define (((mult m1) m2) mf) (m2 (m1 mf)))
 (define ((((plus p1) p2) pf) x)
   ((p1 pf) ((p2 pf) x)))
-
-(define (((mult m1) m2) mf)
-  (m2 (m1 mf)))
 
@@
 
-(define (zero? n)
-  ((n (\(\lambda\) (zx) #f)) #t))
-
 (define ((church=? e1) e2)
@@
 \caption{Church-numeral calculation}
+\label{fig:church}
 \end{figure}
 
-Where \syntax{church=?} is an equality function for Church numerals
-written in terms of recursion and an iterative \syntax{pred} function.
-The \syntax{plus} and \syntax{mult} functions are written as usual.
-
-This program reaches $N$ states under concrete interpretation.
+Figure~\ref{fig:church} gives a small example of computing with Church
+numerals.  The example, while small, is informative: (1) it can be
+written in almost any modern programming language\footnote{It is
+  written here using Scheme notation and makes use of the shorthands:
+%
+\begin{alltt}
+   (define (((f x) y) z) ...) \(\equiv\)
+
+   (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
+\end{alltt}
+
+\begin{alltt}
+   \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
+\end{alltt}
+},
+%
+and (2) it stresses an analyzer's ability to deal with complicated
+environment and control structure arising from the use of higher-order
+functions to model arithmetic.
+
+Our baseline analyzer takes NN:NN minutes to analyze this program
+(without store widening it times out after 30 minutes).  This is on
+par with the performance of Sergey,~et al.'s analyzer, which takes
+NN:NN minutes.
 
@@
 
+\begin{alltt}
+;; Expr \(\to\) \(\Set(\)State\()\)
+(define (eval e)
+  (fix step (set (inj e))))
+\end{alltt}
+
 We compute the semantics of a program by iterating the state
 transition relation until a fixed point in the reachable states is
-reached. We begin with a generic fixpoint combinator, ${\mathcal F}$:
-
-\begin{align*}
-{\mathcal F}(f)(X) &= {\mathcal G}(f, X, \varnothing) \\
-{\mathcal G}(f, \varnothing, S) &= S \\
-{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup f(X)) \\
-\end{align*}
-
-The $\machstep$ relation is most naturally written as a function from
-a state to a set of all next states. We want to find a fixpoint of
-$\machstep$ on the state space, however, so we must lift it to work on
-sets of states. The lifting is a natural one:
-
-\begin{align*}
-lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
-\end{align*}
-
-Then $eval(e) = {\mathcal F}(lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.
-
-If we run this on our Church numeral example, we must be
-patient. Indeed, the state space is finite, but it is exponential in
-the size of the program. With $k = 0$, the store can be in one of ${\mathcal
-  O}(2^n)$ many configurations! We take a step toward a polynomial
-complexity baseline by lifting the store out so that there is one
-shared amongst all states.
+reached.
+
+\begin{alltt}
+;; \(\forall\) (X) ((X \(\to\) \(\Set(\)X\()\)) \(\to\) (\(\Set(\)X\()\) \(\to\) \(\Set(\)X\()\)))
+(define ((appl f) s)
+  (for/fold ([i (set)])
+    ([x (in-set s)])
+    (\(\cup\) i (f x))))
+
+;; Calculate fixpoint of (appl f).
+;; \(\forall\) (X) ((X \(\to\) Set[X]) Set[X] \(\to\) Set[X])
+(define (fix f s)
+  (let loop ((accum (set)) (front s))
+    (if (set-empty? front)
+        accum
+        (let ((new-front ((appl f) front)))
+          (loop (\(\cup\) accum front)
+                (new-front . \(\setminus\) . accum))))))
+\end{alltt}
 
@@
 
-A common technique to faster convergence in flow analyses is to share
-a common store that is monotonically updated as interpretation
-continues. In the ISWIM language with only unary operators, this makes
-our 0-CFA quartic in the size of the program. Not great, but
-polynomial.
-
-We can reuse the reduction relation we have defined, but with some
-scaffolding around it. A state in the store-widened machine is a pair
-of a set of ISWIM states without a store, and a store. A step of this
-machine will pair the store with each of the states in this set,
-reduce, and combine the output stores to create the next. Let $wn :
-\widehat{State} \times Store \to State$ be the metafunction that adds
-a store to a storeless state, $nw$ removes the store, and $c.\msto$
-projects the store out of a state $c$. Formally,
-
-\begin{align*}
-(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
-\mbox{where } CS' &= lift(\machstep)(\{ wn(c, \msto) \mid c \in cs\}) \\
-              cs' &= \{nw(C') \mid C' \in CS'\} \\
-              \msto' &= \bigcup\{C'.\msto \mid C' \in CS'\}
-\end{align*}
+\begin{alltt}
+;; \(\widehat{State}\) = (cons (Set Conf) Store)
+
+;; (State \(\to\) \(\Set(\)State\()\)) \(\to\) \(\widehat{State}\) \(\to\) \(\widehat{State}\)
+(define ((wide-step step) state)
+  (match state
+    [(cons cs \(\msto\))
+     (define ss ((appl step)
+                 (for/set ([c cs]) (c->s c \(\msto\)))))
+     (set (cons (for/set ([s ss]) (s->c s))
+                (\(\sqcup\)-stores ss)))]))
+\end{alltt}
 
@@
 
-The final approximation we make to get to our baseline analysis is
-store-allocating results of application sub-expressions. Notice that
-the $\kfn{}$ continuation stores a value - this makes the space of
-continuations quadratic rather than linear in 0-CFA. We instead
-allocate an address for the value position when we create the
-continuation. This address and the tail address are both determined by
-the label of the application point, so the space becomes linear and
-the overall complexity drops to cubic. Note that this is a critical
-abstraction in languages with n-ary functions, since otherwise the
-continuation space becomes ${\mathcal O}(n^n)$!
-
-\begin{align*}
-\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
-\ev{\mexpi{0},\menv,\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
-\\
-\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
-\ev{\mexp,\menv,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
-\\
-\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
-\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
-\\
-\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
-\\
-\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
-\ev{\mexp,\menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
-\\
-\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
-\co{\mkont,\mval',\msto}
-\\
-\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
-\end{align*}
-
-This extra store-allocation is effectively naming all intermediate
-results, and thus the precision aligns with an analysis specialized to
-either ANF or CPS.
-
 Wide Store: cpu time: 551571 real time: 571319 gc time: 4003
@@
 \begin{alltt}
-(f x y)
-\end{alltt}
-
+(f x y z w)
+\end{alltt}
 where all variables have been joined with multiple values, each
 variable reference transitions to several new states that comprise the
-cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
-only to be joined back together in the application position. We would see transitions of the following shape:
-\begin{center}
-\includegraphics[scale=0.3]{fanout.pdf}
-\end{center}
-For this reason, we {\it delay} the non-determinism
-until it is needed in {\it strict contexts} (such as the guard for an
-{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
-\begin{center}
-\includegraphics[scale=0.3]{lazy.pdf}
-\end{center}
+cartesian product of all the values of {\tt x}, {\tt y}, {\tt z} and
+{\tt w}, only to be joined back together in {\tt f}'s bindings. For
+this reason, we {\it delay} the non-determinism until it is needed in
+{\it strict contexts} (such as the guard for an {\tt if} a called
+procedure, or a numerical primitive application).
 
@@
 \begin{align*}
+force(\msto,\clos{\mvar,\mexp,\menv}) &= \{\clos{\mvar,\mexp,\menv}\}\\
 force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
-force(\msto,\mval) &= \{\mval\}\\
 \ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
 \co{\mkont,\saddr{\menv(\mvar)},\msto} \\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
 &\machstep\;
-\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
+\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}} \\
+\mbox{where } &\mval' \in force(\msto,\mval) \\
 bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
@@
 \end{align*}
+
+The force for the second rule is necessary to keep the same type for
+the \(\kfn{}\) continuation, but we would rather use an address. This
+introduces (acceptably) more imprecision, but has the same precision as an ANF
+analysis\footnote{Exercise for the reader.}.
 
@@
 and accumulated states (if indeed those are being passed around). It
-uses Racket's syntax parameters ~\cite{ianjohnson:eli/stxparam} to
+uses Racket's syntax parameters ~\cite{dvanhorn:eli/stxparam} to
 track the most recent bindings of these accumulators so that {\tt
@@
 
+\paragraph{Constraint-based program analysis for higher-order languages}
+
+Constraint-based program analyses
+(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
+typically compute sets of abstract values for each program point.
+These values approximate the values arising at run-time for each
+program point.  Value sets are computed as the least solution to a set
+of constraints (either inclusion or equality constraints).  The
+constraints must be designed and proved as a sound approximation of
+the semantics.  Efficient implementations of these kinds of analyses
+often take the form of worklist-based graph algorithms for contraint
+solving, and are thus quite different from the interpreter
+implementation.  The approach thus requires effort in constraint
+system design and implementation, and the resulting system require
+verification effort to prove the constraint system is sound and that
+the implementation is correct.
+
+This effort increases substantially as the comlexity of the analyzed
+language increases.  Both the work of maintaining the concrete
+semantics and constraint system (and the relations between them) must
+be scaled simultaneously.  However, constraint systems, which have
+been extensively studied in their own right, enjoy efficient
+implementation techniques and can be expressed in declarative logic
+languages that are heavily
+optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
+Consequently, constraint-based analyses can be computed quickly.  For
+example, Jagannathan and Wright's polymorphic splitting
+implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
+the church numeral example of figure~\ref{fig:church} in 7
+milliseconds, about 25 times faster than the fastest implementation
+considered here.  These analyses compute very different things (as
+detailed in section~\ref{sec:accept}), so the performance comparison
+is not apples-to-apples.  We view the primary contribution of this
+work as a systematic path that eases the design, verification, and
+implementation of analyses using the abstracting abstract machine
+approach to within a factor of performant constraint-based analyses.
+
 \section{Conclusion}
@@
 
+%% \acks Sam Tobin-Hochstadt for encouragement and feedback -- he was
+%% the first to prompt me to look into how make effective
+%% implementations of the AAM approach.
+
+%% NSF, DARPA
+
 \bibliographystyle{plain}
@@
 \section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
+\label{sec:accept}
 

### 2012-11-11 paper/paper.tex

--- 2012-11-11 paper/paper.tex
@@
 
-HERE: ρ
-
 \begin{figure}
 \begin{alltt}
-  ;; State \(\to\) Set[State]
+  ;; State \(\to\) \(\Set(\)State\()\)
   (define (step \(\mstate\))
@@
    (\(\lambda\) (id) id)))
-
-(define (zero? n)
-  ((n (\(\lambda\) (zx) #f)) #t))
 

### 2012-11-11 paper/paper.tex
Checkpoint intro and at a glance draft.

@@
 
-\includegraphics[width=2.5in]{rel-time-church.ps}
-
-\subsection{Notation and prerequisites}
-
-Prereqs: Semantics Engineering \cite{dvanhorn:Felleisen2009Semantics},
-AAM \cite{dvanhorn:VanHorn2011Abstracting}
-\cite{dvanhorn:VanHorn2012Systematic}.
+%% \subsection{Notation and prerequisites}
+
+%% Prereqs: Semantics Engineering \cite{dvanhorn:Felleisen2009Semantics},
+%% AAM \cite{dvanhorn:VanHorn2011Abstracting}
+%% \cite{dvanhorn:VanHorn2012Systematic}.
 
@@
 
-\newpage
-\section{Starting point}
-\subsection{An abstract$^2$ machine for ISWIM}
+\section{At a Glance}
+
+In the following sections, we develop an analyzer for a core
+functional language and follow a series of steps of abstraction and
+optimization.  At each step we evaluate the implementation on an
+example that computes with Church numerals, shown in
+figure~\ref{fig:church}.
+%% \footnote{It is
+%%   written here using Scheme notation and makes use of the shorthands:
+%% %
+%% \begin{alltt}
+%%    (define (((f x) y) z) ...) \(\equiv\)
+%%
+%%    (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
+%% \end{alltt}
+%%
+%% \begin{alltt}
+%%    \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
+%% \end{alltt}}
+The example, while small, is informative:
+\begin{enumerate}
+\item it can be written in most modern programming languages, and
+%
+\item it stresses an analyzer's ability to deal with complicated
+environment and control structure arising from the use of higher-order
+functions to model arithmetic.
+\end{enumerate}
+
+We start, in section~\ref{sec:aam}, by developing an abstract
+interpreter according to the abstracting abstract machines approach
+and observe the resulting implementation cannot analyze the example in
+under 30 minutes.  In section~\ref{sec:baseline}, we perform a further
+abstraction by widening the store.  The resulting analyzer sacrifices
+precision for speed and is able to analyze the example in about 1
+minute.  Without this abstraction step the analysis is exponential due
+to per-state store variance.  This step is described by Van Horn and
+Might and is necessary to make even small examples feasible.  We
+therefore take the widened interpreter as the baseline for our
+evaluation.
+
+In section~\ref{sec:opt} we give a series of simple abstractions and
+implementation techniques that, in total, drop the analysis time to
+only 138 milliseconds, a speedup of 475 times.
+Figure~\ref{fig:churchtime} shows the step-wise improvement of the
+analysis time for this example.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=2.5in]{rel-time-church.ps}
+\end{center}
+\caption{Relative run-time of
+optimized analyers to baseline for Church numeral example.}
+\label{fig:churchtime}
+\end{figure}
+
+In section~\ref{sec:impl}, we describe a more complete implementation
+scaled up and capable of analyzing Scheme benchmark programs.
+
+In section~\ref{sec:eval} we evaluate this implementation against a
+set of benchmark programs drawn from the literature.  Many benchmarks
+cause the baseline analyzer to take longer than 30 minutes or to
+consume more 1 gigabit of memory, at which point the analysis is
+stopped.  This is the case for the largest benchmark program, which is
+3,500 lines of code and takes under a minute in the most optimized
+analyer.  For those benchmarks that did complete on the baseline, the
+optimized analyzer outperformed the baseline by a factor of
+% 475
+two to
+% 4,382
+three orders of magnitude.
+
+\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
+
+\begin{figure}
+\begin{alltt}
+(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
+(define (((mult m1) m2) mf) (m2 (m1 mf)))
+(define ((((plus p1) p2) pf) x)
+  ((p1 pf) ((p2 pf) x)))
+
+(define (((pred n) rf) rx)
+  (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
+    (\(\lambda\) (ignored) rx))
+   (\(\lambda\) (id) id)))
+
+(define ((church=? e1) e2)
+  (if (zero? e1) (zero? e2)
+      (if (zero? e2) #f
+          ((church=? (pred e1)) (pred e2)))))
+
+;; multiplication distributes over addition
+((church=? ((mult \church2) ((plus \church1) \church3)))
+ ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
+\end{alltt}
+\caption{Church-numeral calculation}
+\label{fig:church}
+\end{figure}
+
+Section~\ref{sec:related} relates this work to the literature and
+section~\ref{sec:conclusion} concludes.
+
+\section{An abstract$^{\text 2}$ machine for ISWIM}
+\label{sec:aam}
 
@@
 \end{align*}
-%% Redundant
-%% \paragraph 0CFA
-
-%% \begin{align*}
-%% \mathit{push}(\mlab,\msto,\mkont) &= \mlab,\msto\sqcup[\mlab\mapsto\{\mkont\}]
-%% \\
-%% \mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
-%% \end{align*}
 
 \paragraph{Primitives}
-
-%% \begin{align*}
-%% \widehat{\mathit{push}}(\maddr,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
-%% % \mbox{ where }\maddr \notin\msto
-%% \\
-%% \widehat{\mathit{bind}}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
-%% \end{align*}
 
@@
 
-\subsection{An example: Analyzing Church number computations}
-
-\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
-
-\begin{figure}
-\begin{alltt}
-(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
-(define (((mult m1) m2) mf) (m2 (m1 mf)))
-(define ((((plus p1) p2) pf) x)
-  ((p1 pf) ((p2 pf) x)))
-
-(define (((pred n) rf) rx)
-  (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
-    (\(\lambda\) (ignored) rx))
-   (\(\lambda\) (id) id)))
-
-(define ((church=? e1) e2)
-  (if (zero? e1) (zero? e2)
-      (if (zero? e2) #f
-          ((church=? (pred e1)) (pred e2)))))
-
-;; multiplication distributes over addition
-((church=? ((mult \church2) ((plus \church1) \church3)))
- ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
-\end{alltt}
-\caption{Church-numeral calculation}
-\label{fig:church}
-\end{figure}
-
-Figure~\ref{fig:church} gives a small example of computing with Church
-numerals.  The example, while small, is informative: (1) it can be
-written in almost any modern programming language\footnote{It is
-  written here using Scheme notation and makes use of the shorthands:
-%
-\begin{alltt}
-   (define (((f x) y) z) ...) \(\equiv\)
-
-   (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
-\end{alltt}
-
-\begin{alltt}
-   \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
-\end{alltt}
-},
-%
-and (2) it stresses an analyzer's ability to deal with complicated
-environment and control structure arising from the use of higher-order
-functions to model arithmetic.
-
-Our baseline analyzer takes NN:NN minutes to analyze this program
-(without store widening it times out after 30 minutes).  This is on
-par with the performance of Sergey,~et al.'s analyzer, which takes
-NN:NN minutes.
-
 \newpage
+\section{Baseline analyzer}
+\label{sec:baseline}
+
 \subsection{Generic fixpoint calculator}
+\label{sec:fixpoint}
 
@@
 \begin{align*}
-lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
+\lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
 \end{align*}
 
-Then $eval(e) = {\mathcal F}(lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.
+Then $\eval(e) = {\mathcal F}(\lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.
 
@@
 \subsection{Store widening}
+\label{sec:storewiden}
 
@@
 \subsection{Baseline evaluation}
+\label{sec:baselineeval}
 
@@
 \section{Precision preserving recipe}
+\label{sec:opt}
 
@@
 \section{Implementation}
+\label{sec:impl}
 
@@
 \section{Evaluation}
+\label{sec:eval}
 
@@
 \section{Related work}
+\label{sec:related}
 
@@
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2012-11-11 paper/paper.tex
Toward more numbers.
More on store deltas.

--- 2012-11-11 paper/paper.tex
@@
 behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
-has been used to model Coq~\cite{local:harvard},
-Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
-JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
-Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.
+has been used to model Dalvik, Erlang~\cite{local:DOsualdo:12B},
+JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467},
+Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}, and
+Coq~\cite{local:harvard}.
 
@@
 be easily derived through a small number of steps from existing
-machine models.  Since the relationships between abstract machines and
+machine models. % such as the CESK machine or the Lazy Krivine machine.
+Moreover, since the relationships between abstract machines and
 higher-level semantic models---such as definitional
@@
 understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
-these high-level semantic models to sound program analyzers in a
-systematic way.  Moreover, since these analyses so closely resemble a
-language's interpreter (a) implementing an analysis requires little
-more than implementing an interpreter, (b) a single implementation can
-serve as both an interpreter and analyzer, and (c) verifying the
-correctness of the implementation is straightforward.
-
-However, there is a considerable weakness with the approach: an
-analyzer designed and implemented by following the AAM recipe is
-prohibitively inefficient without both further approximation and
-further implementation effort.
-
-In this article, we develop a systematic approach to deriving the
-feasible implementation of an abstract machine based analyzer.
+these semantic models to sound program analyzers.
 
@@
 
-\includegraphics[width=2.5in]{rel-time-church.ps}
-
 \subsection{Notation and prerequisites}
@@
 
-%% Notation: concrete examples of programs under analysis is given in
-%% (monochrome) Scheme notation.  Code is given in (syntax colored)
-%% Racket.
+Notation: concrete examples of programs under analysis is given in
+(monochrome) Scheme notation.  Code is given in (syntax colored)
+Racket.
 
@@
 
+HERE: ρ
+
 \begin{figure}
 \begin{alltt}
-  ;; State \(\to\) \(\Set(\)State\()\)
+  ;; State \(\to\) Set[State]
   (define (step \(\mstate\))
@@
 
+%% \begin{alltt}
+%% (define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))
+
+%% (define pred
+%%   (\(\lambda\) (n)
+%%     (\(\lambda\) (rf)
+%%       (\(\lambda\) (rx)
+%%         (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
+%%           (\(\lambda\) (i) rx))
+%%          (\(\lambda\) (id) id))))))
+%% \end{alltt}
+
+%% (define \church0 (\(\lambda\) (f0) (\(\lambda\) (x0) x0)))
+%% (define \church1 (\(\lambda\) (f1) (\(\lambda\) (x1) (f1 x1))))
+%% (define \church2 (\(\lambda\) (f2) (\(\lambda\) (x2) (f2 (f2 x2)))))
+%% (define \church3 (\(\lambda\) (f3) (\(\lambda\) (x3) (f3 (f3 (f3 x3))))))
+
 \begin{figure}
 \begin{alltt}
-(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
-(define (((mult m1) m2) mf) (m2 (m1 mf)))
 (define ((((plus p1) p2) pf) x)
   ((p1 pf) ((p2 pf) x)))
+
+(define (((mult m1) m2) mf)
+  (m2 (m1 mf)))
 
@@
    (\(\lambda\) (id) id)))
+
+(define (zero? n)
+  ((n (\(\lambda\) (zx) #f)) #t))
 
@@
 \caption{Church-numeral calculation}
-\label{fig:church}
 \end{figure}
 
-Figure~\ref{fig:church} gives a small example of computing with Church
-numerals.  The example, while small, is informative: (1) it can be
-written in almost any modern programming language\footnote{It is
-  written here using Scheme notation and makes use of the shorthands:
-%
-\begin{alltt}
-   (define (((f x) y) z) ...) \(\equiv\)
-
-   (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
-\end{alltt}
-
-\begin{alltt}
-   \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
-\end{alltt}
-},
-%
-and (2) it stresses an analyzer's ability to deal with complicated
-environment and control structure arising from the use of higher-order
-functions to model arithmetic.
-
-Our baseline analyzer takes NN:NN minutes to analyze this program
-(without store widening it times out after 30 minutes).  This is on
-par with the performance of Sergey,~et al.'s analyzer, which takes
-NN:NN minutes.
+Where \syntax{church=?} is an equality function for Church numerals
+written in terms of recursion and an iterative \syntax{pred} function.
+The \syntax{plus} and \syntax{mult} functions are written as usual.
+
+This program reaches $N$ states under concrete interpretation.
 
@@
 
-The essence of the compilation effect can be seen by consider an example
+The essence of the compilation effect can be seen by considering an example
 such as
@@
 \mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
-\text{ where }\menv',\msto',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
+\text{ where }\menv',\msto',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
 \\
@@
 Each step in the fixpointer for the above techniques requires joining
-several (possibly large) stores together. Even with immutable hashes
+several, commonly large, stores together. Even with immutable hashes
 that utilize sharing, this is memory-intensive. Worse, this
-contributes an obscene fraction of the runtime.
-
-\begin{alltt}
-;; \(\widehat{State} \to \widehat{State}\)
-(define (wide-step-specialized state)
-  (match state
-    [(cons \(\msto\) cs)
-     (define-values (cs* \(\msdiff\))
-       (for/fold ([cs* (set)] [\(\msdiff\)* '()])
-         ([c cs])
-         (match (step-compiled^ (cons \(\msto\) c))
-           [(cons \(\msdiff\)** cs**)
-            (values (\(\cup\) cs* cs**) (++ \(\msdiff\)** \(\msdiff\)*))])))
-     (cons (update \(\msdiff\) \(\msto\)) (\(\cup\) cs cs*))]))
-\end{alltt}
-
-\[
-\{ \widehat{v_3}, \widehat{v_5}, \widehat{v_6} \} = 0001011
-\]
-
-\[
-\sigma = \mbox{\tt \#(0$_0$,\dots,0$_{|P|}$)}
-\]
+contributes an obscene fraction of the runtime. If we instead track
+the changes to the store made over all steps and afterwards update the
+store, we end up with much less allocation and computational overhead.
+
+We represent changes as a list of pairs of address to set of storable
+values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
+$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins as empty for
+each step. Applying the changes is straightforward:
+
+\begin{align*}
+\mtlst(\msto) &= \msto \\
+\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}(\msto) &= \msdiff(\msto)\sqcup[\maddr \mapsto \mval{s}]
+\end{align*}
+
+The reduction relation does have to change to accomodate this
+technique, but only in this regard. The whole store is still given so
+that lookups are possible. Example rule:
+
+\begin{align*}
+bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
+\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
+               \maddr &= \mvar\mcntr' \\
+(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
+(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
+\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
+\end{align*}
+
+We also lift $\machstep$ differently to accomodate for this asymmetry
+in the input and output. For each state that is stepped, we feed the
+output changes to the next so that all changes get accumulated.
+
+\begin{align*}
+(cs, \msto) &\widehat{\machstep} (cs\cup cs', \msdiff(\msto)) \\
+\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
+step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
+step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
+cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
+\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
+\end{align*}
 
@@
 and accumulated states (if indeed those are being passed around). It
-uses Racket's syntax parameters ~\cite{dvanhorn:eli/stxparam} to
+uses Racket's syntax parameters ~\cite{ianjohnson:eli/stxparam} to
 track the most recent bindings of these accumulators so that {\tt
@@
 
-\paragraph{Constraint-based program analysis for higher-order languages}
-
-Constraint-based program analyses
-(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
-typically compute sets of abstract values for each program point.
-These values approximate the values arising at run-time for each
-program point.  Value sets are computed as the least solution to a set
-of constraints (either inclusion or equality constraints).  The
-constraints must be designed and proved as a sound approximation of
-the semantics.  Efficient implementations of these kinds of analyses
-often take the form of worklist-based graph algorithms for contraint
-solving, and are thus quite different from the interpreter
-implementation.  The approach thus requires effort in constraint
-system design and implementation, and the resulting system require
-verification effort to prove the constraint system is sound and that
-the implementation is correct.
-
-This effort increases substantially as the comlexity of the analyzed
-language increases.  Both the work of maintaining the concrete
-semantics and constraint system (and the relations between them) must
-be scaled simultaneously.  However, constraint systems, which have
-been extensively studied in their own right, enjoy efficient
-implementation techniques and can be expressed in declarative logic
-languages that are heavily
-optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
-Consequently, constraint-based analyses can be computed quickly.  For
-example, Jagannathan and Wright's polymorphic splitting
-implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
-the church numeral example of figure~\ref{fig:church} in 7
-milliseconds, about 25 times faster than the fastest implementation
-considered here.  These analyses compute very different things (as
-detailed in section~\ref{sec:accept}), so the performance comparison
-is not apples-to-apples.  We view the primary contribution of this
-work as a systematic path that eases the design, verification, and
-implementation of analyses using the abstracting abstract machine
-approach to within a factor of performant constraint-based analyses.
-
 \section{Conclusion}
@@
 
-%% \acks Sam Tobin-Hochstadt for encouragement and feedback -- he was
-%% the first to prompt me to look into how make effective
-%% implementations of the AAM approach.
-
-%% NSF, DARPA
-
 \bibliographystyle{plain}
@@
 \section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
-\label{sec:accept}
 

### 2012-11-11 paper/paper.tex

--- 2012-11-11 paper/paper.tex
@@
 \end{figure*}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3.2in]{abs-time.ps}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3.2in]{state-per-sec.ps}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=3.2in]{peak-mem.ps}
+\end{center}
+\end{figure}
 

### 2012-11-11 paper/paper.tex
Bug in generic fixpoint definition.
Spelled out the meaning of join.
The semantics is also parameterized by join.

@@
 \item the machine is implicity parameterized by the functions
-  $\mathit{push}$, $\mathit{bind}$, and $\interpdelta$.
+  $\mathit{push}$, $\mathit{bind}$, $\interpdelta$ and $\sqcup$.
 \end{itemize}
@@
               \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
+\end{align*}
+For the duration of this paper, $\sqcup$ is a pointwise lifting of $\cup$:
+\begin{align*}
+\msto\sqcup \varnothing &= \msto \\
+\msto\sqcup \msto'[\maddr \mapsto S] &= \lambda \maddralt. \left\{ 
+  \begin{array}{ll}
+    \msto(\maddr)\cup S & \mbox{if } \maddralt = \maddr, \maddr\in dom(\msto) \\
+    S & \mbox{if } \maddralt = \maddr, \maddr\notin dom(\msto) \\
+    (\msto\sqcup\msto')(\maddralt) & \mbox{otherwise}
+  \end{array}\right.
 \end{align*}
@@
 {\mathcal G}(f, \varnothing, S) &= S \\
-{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup f(X)) \\
+{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup X) \\
 \end{align*}
@@
 \mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
-              \msto' &= \bigcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
+              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
 \end{align*}

### 2012-11-11 paper/paper.tex
Typo: comlexity

@@
 application,
-\begin{alltt}
-(f x y)
-\end{alltt}
+\begin{center}{\tt (f x y)}\end{center}
 
@@
 
-Delta Store + Special + Compile + Lazy:
+Delta Store + Compile + Lazy:
    cpu time: 668 real time: 686 gc time: 41
@@
 
-This effort increases substantially as the comlexity of the analyzed
+This effort increases substantially as the complexity of the analyzed
 language increases.  Both the work of maintaining the concrete

### 2012-11-11 paper/paper.tex
ispell

@@
 \caption{Relative run-time of
-optimized analyers to baseline for Church numeral example.}
+optimized analyzers to baseline for Church numeral example.}
 \label{fig:churchtime}
@@
 3,500 lines of code and takes under a minute in the most optimized
-analyer.  For those benchmarks that did complete on the baseline, the
+analyzer.  For those benchmarks that did complete on the baseline, the
 optimized analyzer outperformed the baseline by a factor of
@@
   labels ($\mlab$) threaded through the computation, and
-\item the machine is implicity parameterized by the functions
+\item the machine is implicitly parameterized by the functions
   $\mathit{push}$, $\mathit{bind}$, $\interpdelta$ and $\sqcup$.
@@
 parameters just as above, but dropping the $\maddr \not\in \msto$
-condition.  This family of interpereters is also non-deterministic in
+condition.  This family of interpreters is also non-deterministic in
 choices of addresses, but it is free to choose addresses that are
 already in use.  Consequently, the machines may be non-deterministic
-when mutiple values reside in a store location.
+when multiple values reside in a store location.
 
@@
 allocation strategy is an abstract interpretation.  In particular,
-concrete intepretation is a kind of abstract interpretation.  So is an
+concrete interpretation is a kind of abstract interpretation.  So is an
 interpretation that allocates a single cell into which all bindings
 and continuations are stored.  On the one hand is an abstract
-intepretation that is non-computable and gives only the ground truth
+interpretation that is non-computable and gives only the ground truth
 of a programs behavior; on the other is an abstract interpretation
@@
 \end{align*}
-For the duration of this paper, $\sqcup$ is a pointwise lifting of $\cup$:
+For the duration of this paper, $\sqcup$ is a point-wise lifting of $\cup$:
 \begin{align*}
@@
 contributes an obscene fraction of the runtime. If we instead track
-the changes to the store made over all steps and afterwards update the
+the changes to the store made over all steps and afterward update the
 store, we end up with much less allocation and computational overhead.
 
-We represent changes as a list of pairs of address to set of storable
+We represent changes as a list of pairs of address to set of storeable
 values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
@@
 
-The reduction relation does have to change to accomodate this
+The reduction relation does have to change to accommodate this
 technique, but only in this regard. The whole store is still given so
@@
 
-We also lift $\machstep$ differently to accomodate for this asymmetry
+We also lift $\machstep$ differently to accommodate for this asymmetry
 in the input and output. For each state that is stepped, we feed the
@@
 of the small step reduction relation and simple variations of a
-general fixpoint combinator. We implementated all of these ideas in an
+general fixpoint combinator. We implemented all of these ideas in an
 analysis of core Scheme (call-by-value lambda calculus extended with
@@
 the semantics.  Efficient implementations of these kinds of analyses
-often take the form of worklist-based graph algorithms for contraint
+often take the form of worklist-based graph algorithms for constraint
 solving, and are thus quite different from the interpreter

### 2012-11-11 paper/paper.tex
Comment out some old numbers.

@@
 
-Wide Store: cpu time: 551571 real time: 571319 gc time: 4003
+% Wide Store: cpu time: 551571 real time: 571319 gc time: 4003
 
@@
 
-Lazy:
-   cpu time: 32481 real time: 32881 gc time: 547
+% Lazy:  cpu time: 32481 real time: 32881 gc time: 547
 
@@
 
-Compile:
-   cpu time: 255397 real time: 261532 gc time: 2947
-
-\noindent
-Compile + Lazy:
-   cpu time: 31173 real time: 31642 gc time: 739
+% Compile: cpu time: 255397 real time: 261532 gc time: 2947
+
+% \noindent
+% Compile + Lazy: cpu time: 31173 real time: 31642 gc time: 739
 
@@
 
-Delta Store + Compile + Lazy:
-   cpu time: 668 real time: 686 gc time: 41
+%% Delta Store + Compile + Lazy:
+%%    cpu time: 668 real time: 686 gc time: 41
 

### 2012-11-11 paper/paper.tex
Timestamp and preallocation sections filled out.

--- 2012-11-11 paper/paper.tex
@@
 
-\subsection{Timestamping stores}
+\subsection{Timestamping an imperative store}
 
 The store delta technique removes overhead of joining large stores,
-but for determining if a state has been visited before, we still have
-to compare entire stores - an expensive computation.  We modify a
-technique created in Shivers'
+but for determining if a state has been visited before (vis. $f(X)
+\setminus S$ in ${\mathcal G}$), we still have to compare entire stores - an
+expensive computation.  We modify a technique created in Shivers'
 dissertation~\cite{ianjohnson:Shivers:1991:CFA}, timestamp
@@
 for storing store-deltas was outweighed by the precision keeping flows
-in check. That is to say, the analysis was faster {\it and} more precise.
-
-TODO: technical detail
+in check. That is to say, the analysis was faster {\it and} more
+precise.
+
+We pair this technique with a mutable workset. That is, to avoid the
+intermediate sets introduced by the $State \to \Set(State)$ signature
+of $\machstep$, we change the fixpoint algorithm to use a global
+workset, {\tt todo}, that $\machstep$ imperatively updates with the
+states that it reaches.
+
+In total, we have six globals:
+\begin{itemize}
+ \item{{\tt $\msto$}, the store}
+ \item{{\tt todo}, the workset}
+ \item{{\tt seen}, a map from state to the timestamp at which it was last seen}
+ \item{{\tt $\msdiff$}, the store changes for the current step}
+ \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
+ \item{{\tt T}, the timestamp of the store}
+\end{itemize}
+
+To ensure termination, we guard adding $c$ to {\tt todo} by the following check:
+\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
+If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.
+
+After all steps complete, we apply $\msdiff$ to $\msto$ and increase
+{\tt T} as long as there was at least one change in $\msdiff$. This
+logic is only sound if we know that each $\ttuple{\maddr}{\mval{s}}$
+in $\msdiff$ would change the value of $\maddr$ in the current
+store. Thus, we also guard additions to $\msdiff$ so that only updates
+that would change the store are permitted. Each time $\msdiff$ is
+successfully extended, we set {\tt $\msdiff$?} to true. Before each
+individual step, we set it to false.
+
+\begin{figure}
+\begin{alltt}
+eval(\(e\)) :=
+  \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
+  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \kmt\))
+  while(true):
+    if todo = \(\varnothing\): return (keys(seen), \(\msto\))
+    else:
+      let old := todo
+      todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
+      foreach c \(\in\) old: \(\msdiff\)? := false; c()
+      unless \(\msdiff\) = \(\mtlst\): T += 1; \(\msdiff(\msto)\)
+\end{alltt}
+\caption{Imperative fixpoint algorithm}
+\label{fig:imperative}
+\end{figure}
 
 \subsection{Pre-allocating the store}
-TODO
+Hashes are not the most efficent for raw speed. If we commit to using
+sequential natural numbers for variables and expression labels, by the
+time we finish parsing the input program, we will know exactly how
+many addresses we will need. Thus the store can be represented by a
+vector the size of the largest label assigned, plus 1. This technique
+only works in 0-CFA, since there are many more addresses that can be
+allocated in 1-CFA and higher. To support general $k$-CFA in a vector
+representation, we need a growable vector and a map from addresses to
+assigned indices. When assignable indices run out, we grow the
+vector. The cost of this map is less than a hash representation for
+the whole store since typically there are more references to an address
+than allocations of it.
 

### 2012-11-11 paper/paper.tex

@@
 Although analyzers are straightforward to build under this approach,
-the resulting analyzer is prohibitively inefficient.
+the result is prohibitively inefficient.
 
@@
 
-The \emph{abstracting abstract machines}
+The \emph{abstracting abstract machines} (AAM)
 approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
@@
 to per-state store variance.  This step is described by Van Horn and
-Might and is necessary to make even small examples feasible.  We
-therefore take the widened interpreter as the baseline for our
-evaluation.
+Might~\cite[\S 3.5--6]{dvanhorn:VanHorn2012Systematic} and is
+necessary to make even small examples feasible.  We therefore take the
+widened interpreter as the baseline for our evaluation.
 
@@
 %% NSF, DARPA
+
+\paragraph{Acknowledgments}
+
+We thank Suresh Jagannathan for providing source code to the
+polymorphic splitting
+analyzer~\cite{dvanhorn:wright-jagannathan-toplas98} and Ilya Sergey
+for the introspective pushdown
+analyzer~\cite{dvanhorn:Earl2012Introspective}.
 

### 2012-11-12 paper/paper.tex
Unfold title.  Cut transition function code.

@@
 
-\title{Optimizing Abstract$^{\text 2}$ Machines}
+\title{Optimizing Abstract Abstract Machines}
 
@@
 
-\begin{figure}
-\begin{alltt}
-  ;; State \(\to\) \(\Set(\)State\()\)
-  (define (step \(\mstate\))
-    (match \(\mstate\)
-      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
-       (match \(\mexp\)
-         [(var\(\superscript\mlab\) \(\mvar\))
-          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
-            (co \(\msto\) \(\mkont\) \(\mval\)))]
-         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
-         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
-         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
-          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
-          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
-         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
-          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
-          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
-      [(co \(\msto\) \(\mkont\) \(\mval\))
-       (match \(\mkont\)
-         ['mt (set (ans \(\msto\) \(\mval\)))]
-         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
-         [(fn\(\superscript\mlab\) \(\mval'\))
-          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
-            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
-         [(fi\(\superscript\mlab\) c a \(\rho\))
-          (for/set ((k (get-cont \(\msto\) l)))
-            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
-      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
-       (match fun
-         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
-          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
-          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
-         [(op \(\mop\))
-          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
-                     (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
-            (co \(\msto\) \(\mkont\) \(\mval\)))]
-         [_ (set)]))]))
-\end{alltt}
-\caption{Implementation of machine transition relation.}
-\end{figure}
+%% \begin{figure}
+%% \begin{alltt}
+%%   ;; State \(\to\) \(\Set(\)State\()\)
+%%   (define (step \(\mstate\))
+%%     (match \(\mstate\)
+%%       [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
+%%        (match \(\mexp\)
+%%          [(var\(\superscript\mlab\) \(\mvar\))
+%%           (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
+%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
+%%          [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
+%%          [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
+%%          [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
+%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
+%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
+%%          [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
+%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
+%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
+%%       [(co \(\msto\) \(\mkont\) \(\mval\))
+%%        (match \(\mkont\)
+%%          ['mt (set (ans \(\msto\) \(\mval\)))]
+%%          [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
+%%          [(fn\(\superscript\mlab\) \(\mval'\))
+%%           (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
+%%             (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
+%%          [(fi\(\superscript\mlab\) c a \(\rho\))
+%%           (for/set ((k (get-cont \(\msto\) l)))
+%%             (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
+%%       [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
+%%        (match fun
+%%          [(clos l \(\mvar\) \(\mexp\) \(\rho\))
+%%           (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
+%%           (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
+%%          [(op \(\mop\))
+%%           (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
+%%                      (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
+%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
+%%          [_ (set)]))]))
+%% \end{alltt}
+%% \caption{Implementation of machine transition relation.}
+%% \end{figure}
 

### 2012-11-12 paper/appendix.tex
Appendix in a file, but not in paper.

--- 2012-11-11 paper/paper.tex
@@
-\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
-\usepackage{alltt,mathpartir}
-\usepackage{amsmath,amsthm}
-\usepackage{amssymb}
-\usepackage{stmaryrd}
-\usepackage{url}
-\usepackage{graphicx}
-
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-% \usepackage{xltxtra}
-% \setmonofont[Scale=MatchLowercase]{DejaVu Sans Mono}
-%
-\input{preamble}
-
-\begin{document}
-
-\conferenceinfo{WXYZ '05}{date, City.}
-\copyrightyear{2005}
-\copyrightdata{[to be supplied]}
-
-% \titlebanner{banner above paper title}        % These are ignored unless
-% \preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Optimizing Abstract$^{\text 2}$ Machines}
-
-%% \authorinfo{J. Ian Johnson}
-%%            {Northeastern University}
-%%            {ianj@ccs.neu.edu}
-%% \authorinfo{Matthew Might}
-%%            {University of Utah}
-%%            {might@cs.utah.edu}
-%% \authorinfo{David Van Horn}
-%%            {Northeastern University}
-%%            {dvanhorn@ccs.neu.edu}
-\authorinfo{}{}{}
-
-\maketitle
-
-\begin{abstract}
-Abstracting abstract machines has recently been proposed as a
-lightweight approach to designing sound and computable program
-analyses.  The approach involves deriving abstract interpreters from
-existing machine semantics and has been applied to a variety of
-languages with features widely considered difficult to analyze.
-Although analyzers are straightforward to build under this approach,
-the resulting analyzer is prohibitively inefficient.
-
-This article contributes a step by step process for going from a naive
-analyzer derived under the abstracting abstract machine approach, to
-an efficient program analyzer.
-\end{abstract}
-
-%% \category{CR-number}{subcategory}{third-level}
-
-%% \terms
-%% term1, term2
-
-%% \keywords
-%% keyword1, keyword2
-
-\section{Introduction}
-
-The \emph{abstracting abstract machines}
-approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
-to deriving program analyses provides a straightforward way of
-transforming a programming language semantics in the form of an
-abstract machine, into a family of abstract interpreters parameterized
-over policies for regulating the analytic precision of the control,
-environment, store, and base value domains.  By taking a
-machine-oriented view of computation, it becomes possible to design,
-verify, and implement program analyzers for realistic language
-features typically considered difficult to model.  The approach was
-originally applied to features such as higher-order functions,
-stack-inspection, exceptions, laziness, first-class continuations, and
-garbage collection.  It has since been used to verify actor-
-\cite{local:DOsualdo:12A} and
-thread-based~\cite{dvanhorn:Might2011Family} parallelism and
-behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
-has been used to model Coq~\cite{local:harvard},
-Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
-JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
-Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.
-
-The primary strength of the approach is that abstract interpreters can
-be easily derived through a small number of steps from existing
-machine models.  Since the relationships between abstract machines and
-higher-level semantic models---such as definitional
-interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational
-semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction
-semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well
-understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
-these high-level semantic models to sound program analyzers in a
-systematic way.  Moreover, since these analyses so closely resemble a
-language's interpreter (a) implementing an analysis requires little
-more than implementing an interpreter, (b) a single implementation can
-serve as both an interpreter and analyzer, and (c) verifying the
-correctness of the implementation is straightforward.
-
-However, there is a considerable weakness with the approach: an
-analyzer designed and implemented by following the AAM recipe is
-prohibitively inefficient without both further approximation and
-further implementation effort.
-
-In this article, we develop a systematic approach to deriving the
-feasible implementation of an abstract machine based analyzer.
-
-%% Using this approach has also lead to new insights in pushdown
-%% abstractions of programs capable of precisely modeling the program
-%% stack, even in the presence of control effects.
-
-%% \subsection{Notation and prerequisites}
-
-%% Prereqs: Semantics Engineering \cite{dvanhorn:Felleisen2009Semantics},
-%% AAM \cite{dvanhorn:VanHorn2011Abstracting}
-%% \cite{dvanhorn:VanHorn2012Systematic}.
-
-%% Notation: concrete examples of programs under analysis is given in
-%% (monochrome) Scheme notation.  Code is given in (syntax colored)
-%% Racket.
-
-\section{At a Glance}
-
-In the following sections, we develop an analyzer for a core
-functional language and follow a series of steps of abstraction and
-optimization.  At each step we evaluate the implementation on an
-example that computes with Church numerals, shown in
-figure~\ref{fig:church}.
-%% \footnote{It is
-%%   written here using Scheme notation and makes use of the shorthands:
-%% %
-%% \begin{alltt}
-%%    (define (((f x) y) z) ...) \(\equiv\)
-%%
-%%    (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
-%% \end{alltt}
-%%
-%% \begin{alltt}
-%%    \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
-%% \end{alltt}}
-The example, while small, is informative:
-\begin{enumerate}
-\item it can be written in most modern programming languages, and
-%
-\item it stresses an analyzer's ability to deal with complicated
-environment and control structure arising from the use of higher-order
-functions to model arithmetic.
-\end{enumerate}
-
-We start, in section~\ref{sec:aam}, by developing an abstract
-interpreter according to the abstracting abstract machines approach
-and observe the resulting implementation cannot analyze the example in
-under 30 minutes.  In section~\ref{sec:baseline}, we perform a further
-abstraction by widening the store.  The resulting analyzer sacrifices
-precision for speed and is able to analyze the example in about 1
-minute.  Without this abstraction step the analysis is exponential due
-to per-state store variance.  This step is described by Van Horn and
-Might and is necessary to make even small examples feasible.  We
-therefore take the widened interpreter as the baseline for our
-evaluation.
-
-In section~\ref{sec:opt} we give a series of simple abstractions and
-implementation techniques that, in total, drop the analysis time to
-only 138 milliseconds, a speedup of 475 times.
-Figure~\ref{fig:churchtime} shows the step-wise improvement of the
-analysis time for this example.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2.5in]{rel-time-church.ps}
-\end{center}
-\caption{Relative run-time of
-optimized analyzers to baseline for Church numeral example.}
-\label{fig:churchtime}
-\end{figure}
-
-In section~\ref{sec:impl}, we describe a more complete implementation
-scaled up and capable of analyzing Scheme benchmark programs.
-
-In section~\ref{sec:eval} we evaluate this implementation against a
-set of benchmark programs drawn from the literature.  Many benchmarks
-cause the baseline analyzer to take longer than 30 minutes or to
-consume more 1 gigabit of memory, at which point the analysis is
-stopped.  This is the case for the largest benchmark program, which is
-3,500 lines of code and takes under a minute in the most optimized
-analyzer.  For those benchmarks that did complete on the baseline, the
-optimized analyzer outperformed the baseline by a factor of
-% 475
-two to
-% 4,382
-three orders of magnitude.
-
-\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
-
-\begin{figure}
-\begin{alltt}
-(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
-(define (((mult m1) m2) mf) (m2 (m1 mf)))
-(define ((((plus p1) p2) pf) x)
-  ((p1 pf) ((p2 pf) x)))
-
-(define (((pred n) rf) rx)
-  (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
-    (\(\lambda\) (ignored) rx))
-   (\(\lambda\) (id) id)))
-
-(define ((church=? e1) e2)
-  (if (zero? e1) (zero? e2)
-      (if (zero? e2) #f
-          ((church=? (pred e1)) (pred e2)))))
-
-;; multiplication distributes over addition
-((church=? ((mult \church2) ((plus \church1) \church3)))
- ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
-\end{alltt}
-\caption{Church-numeral calculation}
-\label{fig:church}
-\end{figure}
-
-Section~\ref{sec:related} relates this work to the literature and
-section~\ref{sec:conclusion} concludes.
-
-\section{An abstract$^{\text 2}$ machine for ISWIM}
-\label{sec:aam}
-
-ISWIM is a family of programming languages parameterized by a set of
-base values and operations.  To make things concrete, we consider a
-member of the ISWIM family with integers, booleans, and a few
-operations.
-
-\begin{figure}
-\[
-\begin{array}{l@{\qquad}rcl}
-\text{Expressions} & \mathit{e} &=& \svar[^\mlab]\mvar\\
-&&|& \slit[^\mlab]\mlit\\
-&&|& \slam[^\mlab]\mvar\mexp\\
-&&|& \sapp[^\mlab]\mexp\mexp \\
-&&|& \sif[^\mlab]\mexp\mexp\mexp \\
-\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
-\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
-\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
-\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
-\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
-\end{array}
-\]
-\caption{Syntax of ISWIM}
-\label{fig:syntax}
-\end{figure}
-
-Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.
-Figure~\ref{fig:aam} defines the semantics of ISWIM as a machine
-model.  Evaluation is defined as the set of states reachable by the
-reflexive, transitive closure of the machine transition relation.  The
-machine is a very slight variation on a standard abstract machine for
-ISWIM in ``eval, continue, apply'' form.
-
-Compared with the standard machine semantics, this definition is
-different in the following ways:
-\begin{itemize}
-\item the store maps addresses to \emph{sets} of values,
-\item continuations are heap-allocated,
-\item there are ``contour values'' (written $\mcntr$) and syntax
-  labels ($\mlab$) threaded through the computation, and
-\item the machine is implicitly parameterized by the functions
-  $\mathit{push}$, $\mathit{bind}$, $\interpdelta$ and $\sqcup$.
-\end{itemize}
-
-\begin{figure*}
-\begin{align*}
-\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
-\\[2mm]
-%% EVAL
-\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
-\co{\mkont,\mval,\msto}
-\text{ where }\mval \in \msto(\menv(\mvar))
-\\
-\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
-\co{\mkont,\mlit,\msto}
-\\
-\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
-\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
-\\
-\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
-\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
-\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
-\\
-\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
-\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
-\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
-\\[2mm]
-%% CONTINUE
-\co{\kmt,\mval,\msto} &\machstep
-\ans{\msto,\mval}
-\\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
-\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
-\\
-\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
-\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
-\text{ where }\mkont \in \msto(\maddr)
-\\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
-\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
-\text{ where }\mkont\in\msto(\maddr)
-\\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
-\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
-\text{ where }\mkont\in\msto(\maddr)
-\\[2mm]
-%% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
-\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
-\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
-\\
-\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
-\co{\mkont,\mval',\msto}
-\text{ where } \mval'\in\interpdelta(\mop,\mval)
-\end{align*}
-\caption{Abstract$^2$ machine for ISWIM}
-\label{fig:aam}
-\end{figure*}
-
-\paragraph{Concrete interpretation} can be characterized by setting the implicit
-parameters of the relation given in Figure~\ref{fig:aam} as follows:
-\begin{align*}
-\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
-\mbox{ where }\maddr \notin\msto
-\\
-\mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
-\mbox{ where }\maddr \notin\msto
-\end{align*}
-The resulting relation is non-deterministic in its choice of
-addresses, however it must always choose a fresh address when
-allocating a continuation or variable binding.  If we consider machine
-states equivalent up to consistent renaming, this relation defines
-a deterministic machine.  (The relation is really a function.)
-
-\paragraph{Abstract interpretation} can be characterized by setting the implicit
-parameters just as above, but dropping the $\maddr \not\in \msto$
-condition.  This family of interpreters is also non-deterministic in
-choices of addresses, but it is free to choose addresses that are
-already in use.  Consequently, the machines may be non-deterministic
-when multiple values reside in a store location.
-
-It is important to recognize from this definition that \emph{any}
-allocation strategy is an abstract interpretation.  In particular,
-concrete interpretation is a kind of abstract interpretation.  So is an
-interpretation that allocates a single cell into which all bindings
-and continuations are stored.  On the one hand is an abstract
-interpretation that is non-computable and gives only the ground truth
-of a programs behavior; on the other is an abstract interpretation
-that is easy to compute but gives little information.  Useful program
-analyses lay somewhere in between and can be characterized by their
-choice of address representation and allocation strategy.
-
-%% We now have a framework for describing program analysis for the ISWIM
-%% family of languages, whereby approximation of both control and
-%% environment structure is regulated by the heap and allocation
-%% policies.
-
-Uniform \(k\)-CFA is one such analysis.
-
-\paragraph{Uniform \(k\)-CFA} can be characterized by the following allocation
-strategy:
-
-\begin{align*}
-\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
-  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
-\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
-                                           \msto\sqcup[\maddr \mapsto
-                                             \{ \mval\}],
-                                           \mcntr' \\
-\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
-              \maddr &= x\mcntr' \\
-%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
-              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
-              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
-\end{align*}
-For the duration of this paper, $\sqcup$ is a point-wise lifting of $\cup$:
-\begin{align*}
-\msto\sqcup \varnothing &= \msto \\
-\msto\sqcup \msto'[\maddr \mapsto S] &= \lambda \maddralt. \left\{
-  \begin{array}{ll}
-    \msto(\maddr)\cup S & \mbox{if } \maddralt = \maddr, \maddr\in dom(\msto) \\
-    S & \mbox{if } \maddralt = \maddr, \maddr\notin dom(\msto) \\
-    (\msto\sqcup\msto')(\maddralt) & \mbox{otherwise}
-  \end{array}\right.
-\end{align*}
-
-\paragraph{Primitives}
-
-\begin{align*}
-\mnum+1 &\in \interpdelta(\saddone,\mnum) &
-\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
-\strue &\in \interpdelta(\szerohuh,\szero) &
-\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
-\end{align*}
-
-\begin{align*}
-\sNum &\in \hat\interpdelta(\saddone,\mnum) &
-\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
-\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
-\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
-\strue &\in \hat\interpdelta(\szerohuh,\szero) &
-\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
-\end{align*}
-
-\begin{figure}
-\begin{alltt}
-  ;; State \(\to\) \(\Set(\)State\()\)
-  (define (step \(\mstate\))
-    (match \(\mstate\)
-      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
-       (match \(\mexp\)
-         [(var\(\superscript\mlab\) \(\mvar\))
-          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
-            (co \(\msto\) \(\mkont\) \(\mval\)))]
-         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
-         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
-         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
-          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
-          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
-         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
-          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
-          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
-      [(co \(\msto\) \(\mkont\) \(\mval\))
-       (match \(\mkont\)
-         ['mt (set (ans \(\msto\) \(\mval\)))]
-         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
-         [(fn\(\superscript\mlab\) \(\mval'\))
-          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
-            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
-         [(fi\(\superscript\mlab\) c a \(\rho\))
-          (for/set ((k (get-cont \(\msto\) l)))
-            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
-      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
-       (match fun
-         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
-          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
-          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
-         [(op \(\mop\))
-          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
-                     (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
-            (co \(\msto\) \(\mkont\) \(\mval\)))]
-         [_ (set)]))]))
-\end{alltt}
-\caption{Implementation of machine transition relation.}
-\end{figure}
-
-\newpage
-\section{Baseline analyzer}
-\label{sec:baseline}
-
-\subsection{Generic fixpoint calculator}
-\label{sec:fixpoint}
-
-We compute the semantics of a program by iterating the state
-transition relation until a fixed point in the reachable states is
-reached. We begin with a generic fixpoint combinator, ${\mathcal F}$:
-
-\begin{align*}
-{\mathcal F}(f)(X) &= {\mathcal G}(f, X, \varnothing) \\
-{\mathcal G}(f, \varnothing, S) &= S \\
-{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup X) \\
-\end{align*}
-
-The $\machstep$ relation is most naturally written as a function from
-a state to a set of all next states. We want to find a fixpoint of
-$\machstep$ on the state space, however, so we must lift it to work on
-sets of states. The lifting is a natural one:
-
-\begin{align*}
-\lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
-\end{align*}
-
-Then $\eval(e) = {\mathcal F}(\lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.
-
-If we run this on our Church numeral example, we must be
-patient. Indeed, the state space is finite, but it is exponential in
-the size of the program. With $k = 0$, the store can be in one of ${\mathcal
-  O}(2^n)$ many configurations! We take a step toward a polynomial
-complexity baseline by lifting the store out so that there is one
-shared amongst all states.
-
-\subsection{Store widening}
-\label{sec:storewiden}
-
-A common technique to faster convergence in flow analyses is to share
-a common store that is monotonically updated as interpretation
-continues. In the ISWIM language with only unary operators, this makes
-our 0-CFA quartic in the size of the program. Not great, but
-polynomial.
-
-We can reuse the reduction relation we have defined, but with some
-scaffolding around it. A state in the store-widened machine is a pair
-of a set of ISWIM states without a store, and a store. A step of this
-machine will pair the store with each of the states in this set,
-reduce, and combine the output stores to create the next. Let $wn :
-\widehat{State} \times Store \to State$ be the metafunction that adds
-a store to a storeless state, $nw$ removes the store, and $c.\msto$
-projects the store out of a state $c$. Formally,
-
-\begin{align*}
-(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
-\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
-              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
-\end{align*}
-
-\subsection{Baseline evaluation}
-\label{sec:baselineeval}
-
-The final approximation we make to get to our baseline analysis is
-store-allocating results of application sub-expressions. Notice that
-the $\kfn{}$ continuation stores a value - this makes the space of
-continuations quadratic rather than linear in 0-CFA. We instead
-allocate an address for the value position when we create the
-continuation. This address and the tail address are both determined by
-the label of the application point, so the space becomes linear and
-the overall complexity drops to cubic. Note that this is a critical
-abstraction in languages with n-ary functions, since otherwise the
-continuation space becomes ${\mathcal O}(n^n)$!
-
-\begin{align*}
-\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
-\ev{\mexpi{0},\menv,\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
-\\
-\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
-\ev{\mexp,\menv,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
-\\
-\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
-\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
-\\
-\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
-\\
-\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
-\ev{\mexp,\menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
-\\
-\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
-\co{\mkont,\mval',\msto}
-\\
-\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
-\end{align*}
-
-This extra store-allocation is effectively naming all intermediate
-results, and thus the precision aligns with an analysis specialized to
-either ANF or CPS.
-
-% Wide Store: cpu time: 551571 real time: 571319 gc time: 4003
-
-\section{Precision preserving recipe}
-\label{sec:opt}
-
-We discuss a variety of techniques for increasing performance by
-decreasing the interpretive overhead of abstract interpretation. They
-come in two flavors: decreasing the size of the represented state
-space [SS], and decreasing intermediate data structures [DS].
-
-\begin{itemize}
- \item{[SS] quotient by sets of values (lazy non-determinism)}
- \item{[SS] remove {\tt ev} states (abstract compilation)}
- \item{[DS] operate on store changes rather than whole stores (store deltas)}
- \item{[DS] represent the store as a global, mutable hash and
-   determine a state's store by the number of changes made to the
-   global store at the time of visiting (timestamp)}
- \item{[DS] use a global mutable workset}
- \item{[DS] represent the global store as a global, mutable vector in
-   the 0-CFA case. (preallocation)}
-\end{itemize}
-Each technique requires minimal changes to the reduction
-relation/fixpoint, and each improves performance by integer
-factors. The first three remain in the purely functional realm. The
-final three use mutation in small, localized places, so they do not
-increase conceptual complexity drastically.
-
-\subsection{Lazy non-determinism}
-
-The first problem that we realized with the AAM approach was that much
-of the non-determinism is spurious. For example, in a function
-application,
-\begin{center}{\tt (f x y)}\end{center}
-
-where all variables have been joined with multiple values, each
-variable reference transitions to several new states that comprise the
-cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
-only to be joined back together in the application position. We would see transitions of the following shape:
-\begin{center}
-\includegraphics[scale=0.3]{fanout.pdf}
-\end{center}
-For this reason, we {\it delay} the non-determinism
-until it is needed in {\it strict contexts} (such as the guard for an
-{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
-\begin{center}
-\includegraphics[scale=0.3]{lazy.pdf}
-\end{center}
-
-For this, we introduce a new value for a delayed lookup that acts to
-quotient the state space by the values at the delayed address.
-
-\begin{align*}
-force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
-force(\msto,\mval) &= \{\mval\}\\
-\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
-\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
-&\machstep\;
-\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
-bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
-  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
-\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
-              \maddr &= \mvar\mcntr'
-\end{align*}
-
-% Lazy:  cpu time: 32481 real time: 32881 gc time: 547
-
-\subsection{Abstract compilation}
-
-We can eliminate the {\tt ev} states from the run-time interpretation
-of a program by specializing the machine transition relation to the
-program being analyzed. This eliminates interpretative overhead by
-first compiling the program into ``byte code'' instructions.
-
-The essence of the compilation effect can be seen by considering an example
-such as
-\[
-\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
-\]
-which makes the following transitions:
-\begin{align}
-& \ev{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\mkont,\msto_0}\\
-\machstep\; &
-\ev{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\kar{\mexp_3,\menv,\maddr_1},\msto_1}
-\\
-\machstep\; &
-\ev{\sapp\mvar{\mexp_1},\menv,\kar{\mexp_2,\menv,\maddr_2},\msto_2}
-\\
-\machstep\; &
-\ev{\mvar, \menv,\kar{\mexp_1,\menv,\maddr_3},\msto_3} % {\mexp_2}
-\\
-\machstep\; &
-\co{\kar{\mexp_1,\menv},\mval,\msto_4} % {\mexp_1}{\mexp_2}
-\mbox{ where } \mval \in \msto(\menv(\maddr))
-\end{align}
-
-where $\msto_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
-[\maddr_2 \mapsto \kar{\mexp_3,\menv,\maddr_1}]
-[\maddr_3 \mapsto \kar{\mexp_2,\menv,\maddr_2}]$.
-
-Notice that the only difference between concrete and abstract interpretation
-is which addresses are pushed.
-
-\begin{figure}
-\begin{align*}
-\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\mval,\msto} \text{ where }\mval\in\msto(\menv(\mvar))
-\\
-\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
-\co{\mkont,\mlit,\msto}
-\\
-\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
-\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
-\\
-\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
-\\
-&
-\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
-\\
-\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
-\\
-&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
-\end{align*}
-\caption{Compilation}
-\end{figure}
-
-\begin{figure}
-\begin{align*}
-\mathit{eval}(\mexp) &= \{ \mstate\ |\ \compile{\mexp}(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
-\\[2mm]
-%% CONTINUE
-\co{\kmt,\mval,\msto} &\machstep
-\ans{\msto,\mval}
-\\
-\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \machstep
-\mcomp^\mcntr(\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr})
-\\
-\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
-\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
-\text{ where }\mkont \in \msto(\maddr)
-\\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \machstep
-\mcompi0^\mcntr(\menv,\msto,\mkont)
-\text{ where }\mkont\in\msto(\maddr)
-\\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \machstep
-\mcompi1^\mcntr(\menv,\msto,\mkont)
-\text{ where }\mkont\in\msto(\maddr)
-\\[2mm]
-%% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \machstep
-\mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
-\text{ where }\menv',\msto',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
-\\
-\ap{\mop,\mval,\msto,\mkont} & \machstep
-\co{\mkont,\mval',\msto} \\
-\text{ where }\mkont &\in\msto(\maddr)
-\text{ and } \mval'\in\interpdelta(\mop,\mval)
-\end{align*}
-\caption{Abstract$^2$ machine for compiled ISWIM}
-\label{fig:caam}
-\end{figure}
-
-% Compile: cpu time: 255397 real time: 261532 gc time: 2947
-
-% \noindent
-% Compile + Lazy: cpu time: 31173 real time: 31642 gc time: 739
-
-\newpage
-
-\subsection{Store deltas}
-
-Each step in the fixpointer for the above techniques requires joining
-several, commonly large, stores together. Even with immutable hashes
-that utilize sharing, this is memory-intensive. Worse, this
-contributes an obscene fraction of the runtime. If we instead track
-the changes to the store made over all steps and afterward update the
-store, we end up with much less allocation and computational overhead.
-
-We represent changes as a list of pairs of address to set of storeable
-values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
-$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins as empty for
-each step. Applying the changes is straightforward:
-
-\begin{align*}
-\mtlst(\msto) &= \msto \\
-\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}(\msto) &= \msdiff(\msto)\sqcup[\maddr \mapsto \mval{s}]
-\end{align*}
-
-The reduction relation does have to change to accommodate this
-technique, but only in this regard. The whole store is still given so
-that lookups are possible. Example rule:
-
-\begin{align*}
-bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
-\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
-               \maddr &= \mvar\mcntr' \\
-(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
-(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
-\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
-\end{align*}
-
-We also lift $\machstep$ differently to accommodate for this asymmetry
-in the input and output. For each state that is stepped, we feed the
-output changes to the next so that all changes get accumulated.
-
-\begin{align*}
-(cs, \msto) &\widehat{\machstep} (cs\cup cs', \msdiff(\msto)) \\
-\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
-step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
-step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
-cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
-\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
-\end{align*}
-
-%% Delta Store + Compile + Lazy:
-%%    cpu time: 668 real time: 686 gc time: 41
-
-\subsection{Timestamping an imperative store}
-
-The store delta technique removes overhead of joining large stores,
-but for determining if a state has been visited before (vis. $f(X)
-\setminus S$ in ${\mathcal G}$), we still have to compare entire stores - an
-expensive computation.  We modify a technique created in Shivers'
-dissertation~\cite{ianjohnson:Shivers:1991:CFA}, timestamp
-approximation, that removes the need for joining whole stores by
-manipulating a global store imperatively. The original technique loses
-precision because the starting store for each state in a machine
-configuration can change between stepping all states and introduce
-more flows than originally possible. However, by delaying the store
-changes until after all steps have completed, we can maintain our
-precision. In a variety of experiments, we observed that the overhead
-for storing store-deltas was outweighed by the precision keeping flows
-in check. That is to say, the analysis was faster {\it and} more
-precise.
-
-We pair this technique with a mutable workset. That is, to avoid the
-intermediate sets introduced by the $State \to \Set(State)$ signature
-of $\machstep$, we change the fixpoint algorithm to use a global
-workset, {\tt todo}, that $\machstep$ imperatively updates with the
-states that it reaches.
-
-In total, we have six globals:
-\begin{itemize}
- \item{{\tt $\msto$}, the store}
- \item{{\tt todo}, the workset}
- \item{{\tt seen}, a map from state to the timestamp at which it was last seen}
- \item{{\tt $\msdiff$}, the store changes for the current step}
- \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
- \item{{\tt T}, the timestamp of the store}
-\end{itemize}
-
-To ensure termination, we guard adding $c$ to {\tt todo} by the following check:
-\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
-If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.
-
-After all steps complete, we apply $\msdiff$ to $\msto$ and increase
-{\tt T} as long as there was at least one change in $\msdiff$. This
-logic is only sound if we know that each $\ttuple{\maddr}{\mval{s}}$
-in $\msdiff$ would change the value of $\maddr$ in the current
-store. Thus, we also guard additions to $\msdiff$ so that only updates
-that would change the store are permitted. Each time $\msdiff$ is
-successfully extended, we set {\tt $\msdiff$?} to true. Before each
-individual step, we set it to false.
-
-\begin{figure}
-\begin{alltt}
-eval(\(e\)) :=
-  \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
-  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \kmt\))
-  while(true):
-    if todo = \(\varnothing\): return (keys(seen), \(\msto\))
-    else:
-      let old := todo
-      todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
-      foreach c \(\in\) old: \(\msdiff\)? := false; c()
-      unless \(\msdiff\) = \(\mtlst\): T += 1; \(\msdiff(\msto)\)
-\end{alltt}
-\caption{Imperative fixpoint algorithm}
-\label{fig:imperative}
-\end{figure}
-
-\subsection{Pre-allocating the store}
-Hashes are not the most efficent for raw speed. If we commit to using
-sequential natural numbers for variables and expression labels, by the
-time we finish parsing the input program, we will know exactly how
-many addresses we will need. Thus the store can be represented by a
-vector the size of the largest label assigned, plus 1. This technique
-only works in 0-CFA, since there are many more addresses that can be
-allocated in 1-CFA and higher. To support general $k$-CFA in a vector
-representation, we need a growable vector and a map from addresses to
-assigned indices. When assignable indices run out, we grow the
-vector. The cost of this map is less than a hash representation for
-the whole store since typically there are more references to an address
-than allocations of it.
-
-% Ian's sections: generators, implementation and compound data
-\section{Implementation}
-\label{sec:impl}
-
-All the above techniques can be combined into a single implementation
-of the small step reduction relation and simple variations of a
-general fixpoint combinator. We implemented all of these ideas in an
-analysis of core Scheme (call-by-value lambda calculus extended with
-variable-arity functions, call/cc, vectors, conses, recursive binding,
-various primitives) in just 3000 lines of Racket. Most lines are spent
-defining the primitives and expected library functions used in our
-benchmarks. Here we describe the unifying vocabulary for writing such
-a framework, and some design decisions related to scaling to a
-language with compound data.
-
-\subsection{Vocabulary of Parameterized Optimizations}
-
-We combine the interface of generators~\cite{ianjohnson:cluhistory}
-with a special lambda form ({\tt $\lambda$\%}) that wraps the meanings
-of {\tt ev} states. We also have a special {\tt do} form that performs
-behind-the-scenes state passing and iteration much like {\tt do} in
-Haskell. In the presence of the different optimization
-techniques, {\tt generator}, {\tt yield}, {\tt $\lambda$\%} and {\tt do}
-change meaning slightly, but keep the same interface as their names
-suggest.
-
-\begin{figure}
-\begin{alltt}
-\(e-dispatch =\)
-(match e
-  [(var\(\superscript\mlab\) x)
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-    (do (\(\msto\)) ((v (delay (lookup-env \(\menv\) x))))
-      (yield (co \(\msto\) k v))))]
-  [(lit\(\superscript\mlab\) l)
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-    (do (\(\msto\)) () (yield (co \(\msto\) k n))))]
-  [(lam\(\superscript\mlab\) x e)
-   (define c (compile e))
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-    (do (\(\msto\)) () (yield (co \(\msto\) k (clos x c \(\menv\))))))]
-  [(app\(\superscript\mlab\) f e)
-   (define cf (compile f))
-   (define ce (compile e))
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-    (define-values (\(\msto\)* a) (push state))
-    (yield (ev \(\msto\)* cf \(\menv\) (ar c \(\menv\) a))))]
-  [(ife\(\superscript\mlab\) e0 e1 e2)
-   (define c0 (compile e0))
-   (define c1 (compile e1))
-   (define c2 (compile e2))
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-     (do (\(\msto\)) [(\(\msto\)* a) (push state)]
-      (yield (ev \(\msto\)* c0 \(\menv\) (ifk c1 c2 \(\menv\) a)))))])
-
-(define (step state)
-  (match state
-    [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
-    [(co \(\msto\) k v)
-     (match k
-       ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
-        (yield (ans \(\msto\) v))))]
-       [(ar\(\superscript\mcntr\subscript\mlab\) e \(\menv\) a)
-        (generator
-         (do (\(\msto\)) ([\(\msto*\) (join \(\msto\) a\(\superscript{f}\) (force \(\msto\) v))])
-           (yield (ev \(\msto*\) e \(\menv\) (fn\(\superscript\mcntr\subscript\mlab\) a\(\superscript{f}\) a)))))
-       [(fn\(\superscript\mcntr\subscript\mlab\) fa a)
-        (generator
-         (do (\(\msto\)) ([\(\msto*\) (join \(\msto\) a\(\superscript{a}\) (force \(\msto\) v))]
-                  [k <- (get \(\msto\) a)]
-                  [f <- (get \(\msto\) fa)])
-           (yield (ap\(\superscript\mcntr\subscript\mlab\) \(\msto*\) f a\(\superscript{a}\) k))))]
-       [(fi\(\superscript\mcntr\subscript\mlab\) c a \(\menv\))
-        (generator
-         (do (\(\msto\)) ([k <- (get \(\msto\) l)]
-                  [v <- (force \(\msto\) v)])
-           (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
-    [(ap\(\superscript\mcntr\subscript\mlab\) \(\msto\) fun a k)
-     (match fun
-       [(clos x e \(\menv\))
-        (generator
-         (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
-           (yield (ev \(\msto\)* e \(\menv\)* k))))]
-       [(? op? o)
-        (do (\(\msto\)) ([k <- (get \(\msto\) l)]
-                 [v <- (force \(\msto\) v)]
-                 [v (\(\interpdelta\) o (list v))])
-          (yield (co \(\msto\) k v)))
-       [_ (generator (do (\(\msto\)) () (continue)))]))]))
-\end{alltt}
-\caption{Parameterized Abstract$^2$ machine for ISWIM}
-\label{fig:paam}
-\end{figure}
-
-\begin{figure}
-\begin{alltt}
-(define (compile e) \(e-dispatch\))
-(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\machstep\) (\(\lambda\) (\(\msto\) \(\menv\) k) body \ldots)
-(yield (ev \(\msto\) c . rest)) \(\longmapsto\) (c \(\msto\) . rest)
-(yield e) \(\longmapsto\) (yield-meaning e)
-\end{alltt}
-\caption{Meaning of forms if compiled}
-\label{fig:cfm}
-\end{figure}
-
-\begin{figure}
-\begin{alltt}
-(define (compile e) e)
-(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\longmapsto\) (let () body ...)
-(yield e) \(\longmapsto\) (yield-meaning e)
-\end{alltt}
-\caption{Meaning of forms if not compiled}
-\label{fig:ncfm}
-\end{figure}
-
-Why generators? Relations in general cannot be written as computable
-functions, but in our case we have a decidable relation with finitely
-many pairs sharing the same left component. There are several
-implementation strategies for such relations, some more effective than
-others. For instance, a silly implementation strategy would be
-defining a binary function that decides membership in the relation. We
-want the definition to drive the interpreter, so we might naturally
-decide to write a function that takes a state and returns a set of all
-the next states. As we described in the imperative workset section,
-this intermediate allocation is unnecessary and wasteful. What we
-really want is a way of pulling on the reduction relation for more
-next states as the fixpoint is ready for them.
-
-For intermediate sets, {\tt yield} is {\tt return} in the set
-monad. For the imperative worklist, {\tt yield} performs mutation
-behind the scenes. Indeed, we could use real generators! This is the
-only case where {\tt generator} constructs a nullary generator; in the
-other cases it evaluates to its body. Racket's generators are not
-tuned for performance, so we actually lose efficiency in this case,
-but the unifying vocabulary is instructive nonetheless.
-
-The {\tt $\lambda$\%} form constructs a real closure in the
-compilation case and evaluates to its body otherwise.
-
-The {\tt do} form has the special job of maintaining the current store
-and accumulated states (if indeed those are being passed around). It
-uses Racket's syntax parameters ~\cite{dvanhorn:eli/stxparam} to
-track the most recent bindings of these accumulators so that {\tt
-  yield-meaning} can refer to them (say {\tt $\sigma$p} and {\tt
-  statep}). In the case of state passing and set-accumulated states, its
-definition is simply
-
-\begin{center}
-\begin{alltt}
-(yield-meaning e) \(\longmapsto\) (values \(\sigma\)p (set-add statep e))
-\end{alltt}
-\end{center}
-
-The linguistic elements described above were all reusable to implement
-the various languages' abstract interpreters. The language of our
-benchmarks, Scheme, has compound data (e.g. vectors and conses), which
-introduces choices for abstraction.
-
-\subsection{Abstracting Compound Data}
-
-We discovered that the choice of abstract domain for literal data
-(such as the large quoted axiom database in the Boyer benchmark)
-drastically affected the performance of our analysis on the benchmark
-suite. It is worth discussing the ``obvious'' and na\"ive abstraction
-that is natural to a reader of
-~\citep{dvanhorn:VanHorn2011Abstracting}, and other easy-to-implement
-yet effective approaches.
-
-\subsubsection{Na\"ively}
-
-The uniform way they approach a simple abstraction strategy is to cut
-recursion out of the data definition by tying the recursive knot
-through the abstract store. For Scheme, the grammar for values looks like the following:
-
-\begin{alltt}
-Value ::= #t | #f | (cons Value Value) | '() | ...
-\end{alltt}
-
-Upon evaluating a {\tt cons} application, we instead allocate two
-addresses $a$ and $d$, join them to the respective values in the
-store, and return the flattened {\tt (cons $a$ $d$)} value. Since
-these addresses are all distinguished at different syntactic
-callsites, and quoted lists really are sugar for a sequence of calls
-to {\tt cons}, this abstraction explodes the value space. Analyzing a
-function that counts the number of atoms in a literal s-expression
-would actually interpret that function at least that number of times
-(more because of intermediate conses). Indeed, even in our fastest
-parameterization, this abstraction causes the analysis of Boyer to be
-430 times slower than the approach we will now describe.
-
-\subsubsection{Less na\"ively and less precisely}
-
-The number of syntactic uses of {\tt cons} versus implicit uses via
-literal lists is dramatically smaller in typical Scheme programs. We
-use the above abstraction for these syntactic uses, but choose a
-different strategy for quoted lists. Quotation is special because it
-cannot introduce function values or mutable values, which is important
-to enhancing our technique's soundness to conceptual complexity ratio.
-
-There are a few steps to consider:
-\begin{itemize}
- \item{Define special value lattice elements for compound data domains that can be quoted
-       (e.g. {\tt QPair} for $\left\{ ({\tt cons} a_i d_i)\right\}_i$, {\tt QVector} for immutable vectors, etc.)}
- \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
- \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
-       and similar definitions for immutable vectors.}
- \item{Extend the $\interpdelta$ axioms to include conservative meaning for these new values (e.g.
-{\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
-\end{itemize}
-
-\begin{figure}
-\begin{align*}
-\interpdelta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
-\interpdelta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
-\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
-                            &\qquad \sqcup[d \mapsto \{ \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\}] \\
-\bigsqcup(v) &= v \\
-\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
-merge(v, v) &= v \\
-merge(n, m) &= {\tt Number} \\
-merge(n, v) &= {\tt QData} \\
-merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
-merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
-merge({\tt QPair}, v) &= {\tt QData} \\
-\vdots
-\end{align*}
-\caption{Quoted list primitive}
-\label{fig:qlist}
-\end{figure}
-
-The alternative to {\tt QData} is to go higher in the lattice to
-$\top$, which has more complicated meaning (Shivers' escape semantics) and
-is overly approximate.
-
-\section{Evaluation}
-\label{sec:eval}
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=6in]{rel-time.ps}
-\end{center}
-\end{figure*}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3.2in]{abs-time.ps}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3.2in]{state-per-sec.ps}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3.2in]{peak-mem.ps}
-\end{center}
-\end{figure}
-
-\cite{dvanhorn:Earl2012Introspective}
-
-\cite{dvanhorn:wright-jagannathan-toplas98}
-
-Other benchmarks
-
-\section{Related work}
-\label{sec:related}
-
-Boucher and Feeley \cite{dvanhorn:Boucher1996Abstract} introduced the
-idea of \emph{abstract compilation}, which used closure generation
-\cite{dvanhorn:Feeley1987Using} to improve the performance of control
-flow analysis.  We have adapted the closure generation technique from
-composition evaluators to abstract machines and applied it to similar
-effect.
-
-\paragraph{Constraint-based program analysis for higher-order languages}
-
-Constraint-based program analyses
-(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
-typically compute sets of abstract values for each program point.
-These values approximate the values arising at run-time for each
-program point.  Value sets are computed as the least solution to a set
-of constraints (either inclusion or equality constraints).  The
-constraints must be designed and proved as a sound approximation of
-the semantics.  Efficient implementations of these kinds of analyses
-often take the form of worklist-based graph algorithms for constraint
-solving, and are thus quite different from the interpreter
-implementation.  The approach thus requires effort in constraint
-system design and implementation, and the resulting system require
-verification effort to prove the constraint system is sound and that
-the implementation is correct.
-
-This effort increases substantially as the complexity of the analyzed
-language increases.  Both the work of maintaining the concrete
-semantics and constraint system (and the relations between them) must
-be scaled simultaneously.  However, constraint systems, which have
-been extensively studied in their own right, enjoy efficient
-implementation techniques and can be expressed in declarative logic
-languages that are heavily
-optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
-Consequently, constraint-based analyses can be computed quickly.  For
-example, Jagannathan and Wright's polymorphic splitting
-implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
-the church numeral example of figure~\ref{fig:church} in 7
-milliseconds, about 25 times faster than the fastest implementation
-considered here.  These analyses compute very different things (as
-detailed in section~\ref{sec:accept}), so the performance comparison
-is not apples-to-apples.  We view the primary contribution of this
-work as a systematic path that eases the design, verification, and
-implementation of analyses using the abstracting abstract machine
-approach to within a factor of performant constraint-based analyses.
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-Abstract machines are not only a good model for rapid analysis
-development, they also can be systematically developed into efficient
-algorithms.
-
-%% \acks Sam Tobin-Hochstadt for encouragement and feedback -- he was
-%% the first to prompt me to look into how make effective
-%% implementations of the AAM approach.
-
-%% NSF, DARPA
-
-\bibliographystyle{plain}
-\bibliography{local,bibliography}
-
 \appendix
@@
 
-\appendix
 \section{Proof: laziness is precision-preserving}
@@
 %% Thus the union of all the strict stores will equal the lazy store.
-
-\end{document}

### 2012-11-12 paper/paper.tex
Appendix in a file, but not in paper.

@@
 
-\appendix
-\section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
-\label{sec:accept}
-
-\cite{dvanhorn:nielson-nielson-popl97} \cite{dvanhorn:Neilson:1999}
-
-This machine's allocation strategy mimics the Uniform k-CFA analysis
-in Principles of Program Analysis, which is defined in terms of
-``$\delta$ contours.''  However, because the machine represetation makes
-context explicit via continuations, we can calculate these contours
-rather than thread them throught the evaluator.  In other words, we
-can use the CESK* machine without modification to obtain Uniform k-CFA
-by way of a simple allocation strategy.  (In this way, it's a
-simplification of the presentation in JFP.)
-
-NNH uses a coinductive acceptability relation to specify Uniform
-k-CFA:
-
-\[
-   C,R \models^{ce}_\delta E
-\]
-
-The cache and global environment form a finite store-like structure
-holding bindings and return values.  The contour environment ce maps
-variables to locations in R which contains their bindings, just as the
-environment of the CESK* machine does.  The current contour delta is a
-string of application labels describing the enclosing context under
-which this term is being analyzed (or evaluated).  If you view the
-acceptability relation as a big-step evalator, the
-$(\widehat C,\widehat\rho)$ component should be seen as a global
-store ce is the environment mapping variables to their locations.
-
-Starting form the initial configuration for a program and iterating
-the machine transition relation until reaching a fixpoint of reachable
-states will \emph{underestimate} the acceptability relation of Uniform
-k-CFA.  You can recover acceptability by feeding this store back into
-the initial configuration and iterating again.  Repeating this process
-until a complete run of the program reaches no new states will be the
-least solution that is acceptable.
-
-HOWEVER.  Why should we care about acceptability?  What this
-machine computes is safe.  In other words, it computes a more
-precise characterization of the run-time behavior of a program.  In
-doing is so, it actually saves work (as can be seen above).
-
-An Example:
-
-\begin{alltt}
- (let ((id (\(\lambda\) (x) x)))
-   (begin (id 1) (id 2)))
-\end{alltt}
-
-Under Uniform 0-CFA, we would have:
-\[
-   [{\tt x} \mapsto \{{\tt 1}, {\tt 2}\}] \in \widehat\rho
-\]
-
-in the least solution to $\models$.  This says that, when run, 'x' is
-bound to 1 or 2.
-
-Under the machine semantics using a 0CFA allocation policy, the trace
-semantics of the machine show that x is bound to 1, and that at some
-later point, x becomes bound to 1 or 2.  Moreover, the machine would
-show that (id 1) evaluates to 1 and only 1, while Uniform 0CFA must
-give that (id 1) is either 1 or 2 to be acceptable.  We don't see any
-value in these kinds of false flows that are due to the global and
-timeless aspects of C,R which acceptability requires the heap to be
-both finite and unchanging over the course of abstract
-interpretation. (Another view of the difference: the machine abstracts
-a program's execution as a \emph{finite state machine} that mimics the
-machine interpretation of the program; the aceptability relation of
-Uniform \(k\)-CFA abstracts a program's execution as a \emph{finite
-  map} that mimics the big-step evaluator: from terms to (sets of)
-values.)
-
-\subsection{Another problem with acceptability: Temporal ignorance}
-
-The small-step approach to static analysis brings subtle yet important temporal
-richness not found in classical analyses for higher-order programs.
-%
-Classical analyses (ultimately) compute judgments on program terms and
-contexts, e.g., at expression $e$, value $x$ may have value $v$.
-%
-The judgments do not relate the order in which expressions and context may be
-evaluated in a program, e.g., a classical analysis has nothing to say with
-regard to question like, ``Do we always evaluate $e_1$ before $e_2$?'' or ``Is
-it always the case that a file handle is opened, read and then closed in that
-order?''
-
-Small-step analyses, by their nature, encode the temporal relationships between
-abstract states.
-%
-It is sensible to make temporal queries of a small-step analysis.
-%
-Of course, this does not come for free: respecting temporal order imposes an
-order in which states and terms may be evaluated \emph{during} the analysis.
-%
-Classical analyses can (and do) evaluate expressions in any order, or in some
-cases, even in parallel~\cite{might:Prabhu:2010:EigenCFA}.
-%
-Relaxing that restriction on order affords additional optimizations that we
-have \emph{not} performed.
-
-We avoid sacrificing order not simply because we are interested in the
-questions it allows us to ask, but because considering temporal order actually
-improves the precision of the analysis itself.
-
-\section{Pushdown Analysis}
-
-It is straightforward to instantiate a \emph{pushdown} abstraction by
-bounding only the variable binding portion of the heap, but using a
-unique allocation strategy for continuations.  Such a strategy
-abstracts a program's execution as a \emph{pushdown automata}
-that mimics the machine interpretation.  This strategy therefore
-models the abstract stack in a true stack like fashion and always
-properly matches function calls with their return.
-
-Although such analyses can be formulated straightforwardly in the
-abstract machine approach, it is not clear all of the techniques of
-this paper can be applied to similar effect in the pushdown context.
-The main problem is calculation of an analysis can no longer be
-computed as the fixed point of the machine transition relation.
-Although there are several implementations (CFA2,ICFP'12), they
-operate at speeds roughly on par with our starting point: unoptimized
-store widened
-machines. \cite{dvanhorn:Earl2012Introspective,dvanhorn:Vardoulakis2011CFA2}
-
-\appendix
-\section{Proof: laziness is precision-preserving}
-Given widened machine configurations, we can show that lazy
-non-determinism is precision-preserving in the cases that application
-positions are store-allocated and not. We first show the latter since
-it uses less cluttered transition rules. The high level is that the
-``fan-out'' of non-determinism is collapsed back in the store (in the
-store-allocated applications case) or delayed a single step by
-laziness that would have happened in a step of strictness.
-
-Details of lazy machine given as a diff from the strict machine. \\
-State space of $lazy-\widehat{CESK}^*_t$ varies just in
-$\mathit{Value}$ (storable values do not change):
-\begin{align*}
-\mval \in \mathit{Value} &::= z \mid b \mid o \mid \saddr{\maddr} \\
-\end{align*}
-
-We define a relation between strict and lazy machines that illustrates
-the quotient that the lazy states embody.
-
-\newcommand{\vapprx}[3]{#1 \cong_{#2} #3}
-\newcommand{\kapprx}[3]{#1 \cong_{#2} #3}
-\newcommand{\lapprx}[3]{#1 \approx_{#2} #3}
-\newcommand{\apprx}[3]{#1 \sim_{#2} #3}
-\newcommand{\capprx}[2]{#1 \sim #2}
-
-\begin{mathpar}
-\inferrule{\mval' \in force(\msto, \mval)}
-          {\vapprx{\mval}{\msto}{\mval'}} \quad
-\inferrule{\ans{\mval} \in cs}{\lapprx{\ans{\mval}}{\msto}{cs}} \quad
-\inferrule{\ev{\mexp, \menv, \mkont, \mcntr} \in cs}
-          {\lapprx{\ev{\mexp, \menv,\mkont, \mcntr}}{\msto}{cs}} \\
-\inferrule{\co{\mkont, \mval'} \in cs \\
-           \vapprx{\mval}{\msto}{\mval'}}
-          {\lapprx{\co{\mkont, \mval}}{\msto}{cs}} \quad
-\inferrule{\ap{\mval_0, \mval_1, \mkont} \in cs \\
-           \vapprx{\mval_1}{\msto}{\mval_1'}}
-          {\lapprx{\ap{\mval_0, \mval_1', \mkont}}{\msto}{cs}} \\
-\inferrule{\ev{\mexp, \menv, \mkont} \in lcs}
-          {\apprx{\ev{\mexp, \menv,\mkont}}{\msto}{lcs}} \quad
-\inferrule{\ans{\mval} \in lcs \\
-           \vapprx{\mval}{\msto}{\mval'}}
-          {\apprx{\ans{\mval'}}{\msto}{lcs}} \\
-\inferrule{\co{\mkont, \mval} \in lcs \\
-           \vapprx{\mval}{\msto}{\mval'}}
-          {\apprx{\co{\mkont, \mval'}}{\msto}{lcs}} \quad
-\inferrule{\ap{\mval_0, \mval_1, \mkont} \in lcs \\
-           \vapprx{\mval_1}{\msto}{\mval_1'}}
-          {\apprx{\ap{\mval_0, \mval_1', \mkont}}{\msto}{lcs}} \\
-\inferrule{\forall lc \in lcs, \lapprx{lc}{\msto}{cs} \\
-           \forall c \in cs, \apprx{c}{\msto}{lcs}}
-          {\capprx{(lcs, \msto)}{(cs, \msto)}}
-\end{mathpar}
-
-We also need metafunctions for adding and removing a store component
-from states. These are straightforwardly defined, so just name them
-$wn$ and $nw$ respectively.
-%% c2vc(<e^l r k d>,s) = <e^l r s k d>
-%% c2vc(<v k d>,s) = <v s k d>
-%% vc2c(<e^l r s k d>) = <e^l r k d>
-%% vc2c(<v s k d>) = <v k d>
-
-The property we show is that all states stay related through
-reduction. The main difficulty is in showing the resulting stores are
-in fact equal. We do this by showing each lazy reduction corresponds
-to a set of strict reductions, for which the union of their output
-stores equal the lazy store. Also for strict equality, every strict
-reduction corresponds to a lazy reduction (immediate from the relation).
-
-\begin{lemma}[Partition step]
-$\forall lc, lcs', cs, \msto, \msto'.$
- if  $\capprx{(\{lc\},\msto)}{(cs,\msto)}$
- and $\forall lc' \in lcs'. wn(lc,\msto) \machstep wn(lc', \msto')$ then
- $\exists cs'. \forall c \in cs. \exists c' \in cs', \msto^c$
- such that $wn(c,\msto) \machstep wn(c', \msto^c)$
- and       $\msto' = \bigcup\limits_{c \in cs}{\msto^c}$
- and       $\capprx{(lcs',\msto')}{(cs',\msto')}$.
-\end{lemma}
-\begin{proof}
-Let $lc' \in lcs'$ be arbitrary.
-By cases on $wn(lc,\msto) \machstep wn(lc', \msto')$:
-\begin{itemize}
-\item{Case $\ev{\svar{\mvar}, \menv, \mkont, \msto} \machstep \co{\mkont,
-    \saddr{\menv(\mvar)}, \msto}$: \\ By definition of $\machstep$,
-  $wn(lc,\msto)$ steps strictly to each of $cs' = \{\co{\mval, \mkont} \mid \mval \in \msto(\menv(\mvar))\}$.  By definition of
-  $\capprx{}{}$, $cs = \{lc\}$. The stores are the same, and by
-  definition of $\lapprx{}{\msto}{}$, $\lapprx{lc'}{\msto}{cs'}$.}
-\item{Case $\ev{\slit{\mlit}, \menv, \msto, \mkont, \msto} \machstep
-            \co{\mkont, \mlit, \msto}$: \\
-      Immeditate.}
-\item{Case $\ev{\slam{\mvar}{\mexp}, \menv, \mkont, \msto} \machstep
-            \co{\mkont, \clos{\mvar, \mexp, \menv}, \msto}$: \\
-      Immediate}
-\item{Case $\ev[^\mcntr]{\sapp[^\mlab]{\mexp_0}{\mexp_1}, \menv, \mkont, \msto} \machstep
-            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kar[^\mcntr_\mlab]{\mexp_1, \menv, \maddr}, \msto'}$: \\
-      Where $\maddr, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$ \\
-      Immediate.}
-\item{Case $\ev[^\mcntr]{\sif[^\mlab]{\mexp_0}{\mexp_1}{\mexp_2}, \menv, \mkont, \msto} \machstep
-            \ev[^\mcntr]{\mexp_0, \menv, \msto', \kif[^\mcntr]{\mexp_1, \mexp_2, \menv, \maddr}, \msto'}$: \\
-      Where $\maddr, \msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)$ \\
-      Immediate.}
-\item{Case $\co{\kmt, \mval, \msto} \machstep \ans{\msto, \mval'}$: \\
-      Where $\mval' \in force(\msto, \mval)$ \\
-      By definition of $\capprx{}{}$, $cs = \{\co{\kmt, \mval'} \mid \mval \in force(\msto, \mval)\}$.
-      By definition of $\machstep$, $wn(lc,\msto)$ steps strictly to
-      each of $cs' = \{\ans{\mval}\}$. The stores are the same.
-      By definition of $\lapprx{}{\msto}{}$, $\forall lc' \in lcs'. \lapprx{lc'}{\msto}{cs'}$.
-      By definition of $\apprx{}{\msto}{}$, $\forall c' \in cs'. \apprx{c'}{\msto}{lcs'}$.
-      Thus $\capprx{(lcs',\msto)}{(cs',\msto)}$.}
-\item{Case $\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval, \msto} \machstep
-            \ev[^\mcntr]{\mexp, \menv, \kfn[^\mcntr_\mlab]{\mval',\maddr}, \msto}$: \\
-      Where $v' \in force(\msto, \mval)$. \\
-      By definition of $\apprx{}{\msto}{}$,
-        $cs = \{\co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval'} \mid \mval \in force(\msto, \mval)\}$.
-      By definition of $\machstep, \lapprx{}{\msto}{}$,
-        $cs' = \{\ev[^\mcntr]{\mexp, \menv, \kfn[^\mcntr_\mlab]{\mval',\maddr}}
-            \mid \co{\kar[^\mcntr_\mlab]{\mexp, \menv, \maddr}, \mval'} \in cs\}$.
-      The stores are the same, and the previous imply $\capprx{(lcs',\msto)}{(cs',\msto)}$.}
-\item{Case $\co{\kfn[^\mcntr_\mlab]{\mvalx{u}, \maddr}, \mval, \msto} \machstep
-            \ap{\mvalx{u}, \mval, \mkont, \msto}$: \\
-      Where $\mkont \in \msto(\maddr)$. \\
-      Immediate.}
-\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
-            \ev{\mexp_0, \menv, \mkont, \msto}$: \\
-      Where $\mkont \in \msto(\maddr)$. \\
-      By definition of $\machstep$, $\strue \in force(\msto, \mval)$, thus by definition of
-      $\capprx{}{}$, the strict machine takes the same step.}
-\item{Case $\co{\kif[^\mcntr]{\mexp_0, \mexp_1, \menv, \maddr}, \mval, \msto} \machstep
-            \ev{\mexp_1, \menv, \mkont, \msto}$: \\
-      Where $\mkont \in \msto(\maddr)$. \\
-      By definition of $\machstep$, $\sfalse \in force(\msto, \mval)$, thus by definition of
-      $\capprx{}{}$, the strict machine takes the same step.}
-\item{Case $\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval, \mkont,\msto} \machstep
-            \ev[^{\mcntr'}]{\mexp, \menv', \mkont, \msto'}$: \\
-      Where $\menv', \msto', \mcntr' = \mathit{bind}^\mcntr_\mlab(\menv, \msto, \mvar, \mval)$
-      By definition of $\apprx{}{\msto}{}$,
-        $cs = \{\ap[^\mcntr_\mlab]{\clos{\mvar, \mexp, \menv}, \mval', \mkont} \mid
-                \mval' \in force(\msto, \mval)\}$.
-      Thus $cs' = \{\ev[^{\mcntr'}]{\mexp, \menv', \mkont}\}$ (clearly $\lapprx{lc'}{\msto'}{cs'}$)  and
-      for all $c \in cs$, $wn(c,\msto) \machstep wn(c',\msto\sqcup[\mvar\mcntr' \mapsto \{c.\mval\}])$.
-      Thus the union of these stores is equal to $\msto'$.}
-\item{Case $\ap[^\mcntr_\mlab]{\mop, \mval, \mkont,\msto} \machstep
-            \co{\mval'', \mkont, \msto}$: \\
-      Where $\mval' \in force(\msto, \mval)$
-        and $\mval'' \in\interpdelta(\mop,\mval')$. \\
-      By definition of $\capprx{}{}$ and $\interpdelta$,
-        $cs = \{\ap[^\mcntr_\mlab]{\mop, \mvalx{u}, \mkont} \mid
-                \mvalx{u} \in force(\msto, \mval),
-                \mval'' \in\interpdelta(\mop,\mvalx{u})\}$.
-      Thus the strict machine takes the same step.}
-\end{itemize}
-\end{proof}
-
-\begin{theorem}[Laziness preserves precision]
-$\forall lcs,cs,lcs',\msto,\msto'$ if $\capprx{(lcs, \msto)}{(cs,\msto)}$
-  and $(lcs,\msto) \machstep (lcs',\msto')$ then there exists $cs'$
-  such that $(cs,\msto) \machstep (cs',\msto')$
-\end{theorem}
-\begin{proof}
-By definition of $\machstep$,
-$lcs' = \{lc' \mid \exists \msto^{lc}. wn(lc,\msto) \machstep wn(lc', \msto^{lc}), lc \in lcs\}$
-For each $lc \in lcs$ let $\hat{c}^{lc} \subseteq cs$ be the
-smallest set such that $\lapprx{lc}{\msto}{\hat{c}^{lc}}$. \\ By the
-previous lemma, there exists a $\hat{c}^{lc}{}'$ such that $\capprx{(\{lc' \mid wn(lc,\msto) \machstep
-  wn(lc', \msto^{lc})\}, \msto^{lc})}{(\hat{c}',\msto^{lc})}$. Let
-$cs' = \bigcup\limits_{lc \in lcs}{\hat{c}^{lc}{}'}$ and $\msto' = \bigcup\limits_{lc \in lcs}{\msto^{lc}}$.
-Since each step individually is related, the union is related: $\capprx{lcs'}{\msto'}{cs'}$.
-By definition of $\machstep$, $(cs, \msto) \machstep (cs', \msto')$ and $(lcs, \msto) \machstep (lcs', \msto')$.
-\end{proof}
-
-%% --Eval
-%% <x r s k d> --> <v s k d> where v in s(r(x))  [VAR]
-%% <(e0^l0 e1^l1)^{l,lf,la} r s k d> --> <e0^l0 r s\sqcup[ld |-> {k}] ar^{l,lf,la}(e1^l1 r ld) d> [APP]
-%% <(lambda x e) r s k d> --> <clos(x e r) s k d> [CLOS]
-%% --Continue
-%% <v s ar^{l,lf,la}(e^le r a) d> --> <e^le r s\sqcup[lfd |-> {v}] fn^{l,la}(lfd, a)> [CO]
-%% --Apply
-%% <v s fn^{l,la}(fa, ka) d> --> <e^fl r[x |-> xd'] s'[xd' |-> s'(lad)] k d'> [AP]
-%%   where clos(x e^fl r) in s(fa)
-%%         k in s(ka)
-%%         s' = s\sqcup[lad |-> {v}]
-%%         d' = truncate(ld, K)
-
-%% This extra binding corresponds directly to the intermediate
-%% bindings introduced by ANF.
-%% Matt's ANF analyses likely have all the flavours of our "laziness" technique, but this is how we get our chocolate with our peanut butter as Olin would say.
-
-%% Compare this to lazyCESK^*t:
-%% v ::= clos(x e r) | addr(a)
-%% Clos ::= clos(x e r)
-%% s : a -> P(Clos)
-%% others the same
-
-%% force(s, addr(a)) = s(a)
-%% force(s, clos(x e r)) = {clos(x e r)}
-
-%% --Eval
-%% <x r s k d> --> <addr(r(x)) s k d>
-%% others the same
-%% --Continue
-%% <v s ar^{l,lf,la}(e^le r a) d> --> <e^le r s\sqcup[lfd |-> force(s,v)] fn^{l,la}(lfd, a)>
-%% --Apply
-%% <v s fn^{l,la}(fa, ka) d> --> <e^fl r[x |-> xd'] s'[xd' |-> s'(lad)] k d'>
-%%   where clos(x e^fl r) in s(fa)
-%%         k in s(ka)
-%%         s' = s\sqcup[lad |-> force(s, v)]
-%%         d' = truncate(ld, K)
-%% The same globalizing stuff applies.
-
-%% Let us relate the state spaces of the two abstract machines for the lambda calculus in the following way
-
-%% Then we can show that if lazyconf ~ conf and lazyconf ==> lazyconf' then there exists a conf' such that conf ==> conf' and lazyconf' ~ conf'
-%% Let lc in lazyconf.cs be arbitrary (call lazyconf.s, s)
-%% By cases on lc:
-%% <x r k d>:
-%% Premises:
-%% <x r k d> in conf.cs
-%% Thus:
-%% Since
-%% <x r s k d> -->lazy   <addr(r(x)) s k d>
-%% <x r s k d> -->strict <v s k d> where v in s(r(x))
-%% By definition of force, v in force(s, addr(r(x)))
-%% Thus the first rule of ~_s applies and the second rule of l~_s applies.
-
-%% <(lambda x e) r k d>:
-%% Trivial
-
-%% <(e0^l0 e1^l1)^{l,lf,la} r k d>:
-%% Premises:
-%% <(e0^l0 e1^l1)^{l,lf,la} r k d> in conf.cs
-%% Thus:
-%% <(e0^l0 e1^l1)^{l,lf,la} r s k d> -->lazy   <e0^l0 r s\sqcup[ld |-> {k}] ar^{l,lf,la}(e1^l1 r ld) d>
-%% <(e0^l0 e1^l1)^{l,lf,la} r s k d> -->strict <e0^l0 r s\sqcup[ld |-> {k}] ar^{l,lf,la}(e1^l1 r ld) d>
-
-%% Stores are same in this case.
-
-%% <v ar^{l,lf,la}(e^le r a) d>:
-%% Premises:
-%% For all v' in force(s, v),
-%%   <v' ar^{l,lf,la}(e^le r a) d> in conf.cs
-%% Thus:
-%% <v ar^{l,lf,la}(e^le r ka) d> -->lazy   <e^le r s\sqcup[lfd |-> force(s, v)] fn^{l,la}(lfd ka) d>
-%% for all v' in force(s,v),
-%% <v' ar^{l,lf,la}(e^le r ka) d> -->strict <e^le r s\sqcup[lfd |-> {v'}] fn^{l,la}(lfd ka) d>
-%% Thus the union of all the strict stores will equal the lazy store.
-
-%% <v fn^{l,la}(fa ka) d>:
-%% Premises:
-%% For all v' in force(s,v),
-%%   <v' fn^{l,la}(fa ka) d> in conf.cs
-%% Thus
-%% <v fn^{l,la}(fa ka) d> -->lazy <e^fl r[x |-> xd'] s'\sqcup[xd' |-> s'(lad)] k d'>
-%% where k in s(ka)
-%%       clos(x e^fl r) in s(fa)
-%%       d' = truncate(ld, K)
-%%       s' = s\sqcup[lad |-> force(s,v)]
-%% For all v' in force(s,v),
-%% <v' fn^{l,la}(fa ka) d> -->strict <e^fl r[x |-> xd'] s'\sqcup[xd' |-> s'(lad)] k d'>
-%% where k in s(ka)
-%%       clos(x e^fl r) in s(fa)
-%%       d' = truncate(ld, K)
-%%       s' = s\sqcup[lad |-> {v'}]
-
-%% Thus the union of all the strict stores will equal the lazy store.
+% \include{appendix}
 

### 2012-11-12 paper/paper.tex
Killed code and most of implementation section. Started elaborating on compound data.

@@
 
-\title{Optimizing Abstract Abstract Machines}
+\title{Optimizing Abstract$^{\text 2}$ Machines}
 
@@
 
-\paragraph{Primitives}
-
-\begin{align*}
-\mnum+1 &\in \interpdelta(\saddone,\mnum) &
-\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
-\strue &\in \interpdelta(\szerohuh,\szero) &
-\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
-\end{align*}
-
-\begin{align*}
-\sNum &\in \hat\interpdelta(\saddone,\mnum) &
-\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
-\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
-\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
-\strue &\in \hat\interpdelta(\szerohuh,\szero) &
-\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
-\end{align*}
-
-%% \begin{figure}
-%% \begin{alltt}
-%%   ;; State \(\to\) \(\Set(\)State\()\)
-%%   (define (step \(\mstate\))
-%%     (match \(\mstate\)
-%%       [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
-%%        (match \(\mexp\)
-%%          [(var\(\superscript\mlab\) \(\mvar\))
-%%           (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
-%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
-%%          [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
-%%          [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
-%%          [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
-%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
-%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
-%%          [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
-%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
-%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
-%%       [(co \(\msto\) \(\mkont\) \(\mval\))
-%%        (match \(\mkont\)
-%%          ['mt (set (ans \(\msto\) \(\mval\)))]
-%%          [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
-%%          [(fn\(\superscript\mlab\) \(\mval'\))
-%%           (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
-%%             (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
-%%          [(fi\(\superscript\mlab\) c a \(\rho\))
-%%           (for/set ((k (get-cont \(\msto\) l)))
-%%             (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
-%%       [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
-%%        (match fun
-%%          [(clos l \(\mvar\) \(\mexp\) \(\rho\))
-%%           (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
-%%           (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
-%%          [(op \(\mop\))
-%%           (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
-%%                      (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
-%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
-%%          [_ (set)]))]))
-%% \end{alltt}
-%% \caption{Implementation of machine transition relation.}
-%% \end{figure}
-
-\newpage
+%% \paragraph{Primitives}
+
+%% \begin{align*}
+%% \mnum+1 &\in \interpdelta(\saddone,\mnum) &
+%% \mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
+%% \strue &\in \interpdelta(\szerohuh,\szero) &
+%% \sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+%% \end{align*}
+
+%% \begin{align*}
+%% \sNum &\in \hat\interpdelta(\saddone,\mnum) &
+%% \sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
+%% \strue &\in \hat\interpdelta(\szerohuh,\sNum) &
+%% \sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
+%% \strue &\in \hat\interpdelta(\szerohuh,\szero) &
+%% \sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+%% \end{align*}
+
+%\newpage
 \section{Baseline analyzer}
@@
 machine will pair the store with each of the states in this set,
-reduce, and combine the output stores to create the next. Let $wn :
-\widehat{State} \times Store \to State$ be the metafunction that adds
-a store to a storeless state, $nw$ removes the store, and $c.\msto$
-projects the store out of a state $c$. Formally,
+reduce, and combine the output stores to create the next. The
+scaffolded reduction relation, $\widehat{\machstep}$, is as follows:
+
+Let $wn : \widehat{State} \times Store \to State$ be the metafunction
+that adds a store to a storeless state and $c.\msto$ projects the
+store out of a state $c$.
 
@@
 \end{align*}
+That is, the next set of states is what we get by stepping each
+given state, without its store. The next store is the combination of all produced
+stores. For simplicity, assume that stuck states step to themselves.
 
@@
 
+The important rules that change are the following, specialized to 0-CFA:
+
 \begin{align*}
@@
 results, and thus the precision aligns with an analysis specialized to
-either ANF or CPS.
+either ANF or CPS. We can also play a representation trick and remove
+$\menv$ since we know it is the identity function in 0-CFA. We will
+refer to this change in the semantics as value indirection.
 
@@
 
-\section{Precision preserving recipe}
+\section{Mostly precision preserving recipe}
 \label{sec:opt}
@@
 \end{itemize}
+
 Each technique requires minimal changes to the reduction
@@
 
-The first problem that we realized with the AAM approach was that much
-of the non-determinism is spurious. For example, in a function
-application,
-\begin{center}{\tt (f x y)}\end{center}
-
+A shortcoming of the AAM approach is that much of the non-determinism
+is spurious. For example, in a function application, {\tt (f x y)},
 where all variables have been joined with multiple values, each
@@
 cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
-only to be joined back together in the application position. We would see transitions of the following shape:
+only to be joined back together in their respective application positions. We would
+see transitions of the following shape:
 \begin{center}
@@
 For this, we introduce a new value for a delayed lookup that acts to
-quotient the state space by the values at the delayed address.
+quotient the state space by the values at the delayed address. Its changes to the semantics are highlighted in the following rules:
 
@@
 
+This semantics introduces a subtle precision difference over the
+baseline that is negligible. Consider a configuration where a
+reference to a variable and a binding of a variable will happen in one
+step. With laziness, the reference will mean the original binding(s)
+of the variable or the new one, because the actual store lookup is
+delayed precisely one step (i.e. laziness is administrative). Without
+laziness, the reference will fan out to all the bindings of the
+variable before the new binding happens and thus might have an
+observable precision difference. In our experience, this situation
+happens too rarely to warrant the significant increase in time and
+memory needed for this eager non-determinism. Indeed, were the
+variable reference a step later and another binding is not made in
+that step, the results of the two approaches are the same.
+
+The administrative nature of laziness means that we could remove the
+loss in precision by essentially duplicating the reduction relation to
+specialize variable lookup. This works since in the semantics of ISWIM
+with value indirection consumes the value component of states in
+one step. This is not the case for semantics that replicate the value
+component across reductions, say for popping off exception handler
+frames. Further convolution is needed to remove the administrative
+nature of laziness in these semantics. Due to the increase of
+conceptual complexity for negligable benefit, we decided against this
+approach.
+
+All following techniques preserve precision.
 % Lazy:  cpu time: 32481 real time: 32881 gc time: 547
@@
 
-\newpage
+%\newpage
 
@@
 many addresses we will need. Thus the store can be represented by a
-vector the size of the largest label assigned, plus 1. This technique
-only works in 0-CFA, since there are many more addresses that can be
-allocated in 1-CFA and higher. To support general $k$-CFA in a vector
-representation, we need a growable vector and a map from addresses to
-assigned indices. When assignable indices run out, we grow the
-vector. The cost of this map is less than a hash representation for
-the whole store since typically there are more references to an address
-than allocations of it.
-
-% Ian's sections: generators, implementation and compound data
-\section{Implementation}
-\label{sec:impl}
-
-All the above techniques can be combined into a single implementation
-of the small step reduction relation and simple variations of a
-general fixpoint combinator. We implemented all of these ideas in an
-analysis of core Scheme (call-by-value lambda calculus extended with
-variable-arity functions, call/cc, vectors, conses, recursive binding,
-various primitives) in just 3000 lines of Racket. Most lines are spent
-defining the primitives and expected library functions used in our
-benchmarks. Here we describe the unifying vocabulary for writing such
-a framework, and some design decisions related to scaling to a
-language with compound data.
-
-\subsection{Vocabulary of Parameterized Optimizations}
-
-We combine the interface of generators~\cite{ianjohnson:cluhistory}
-with a special lambda form ({\tt $\lambda$\%}) that wraps the meanings
-of {\tt ev} states. We also have a special {\tt do} form that performs
-behind-the-scenes state passing and iteration much like {\tt do} in
-Haskell. In the presence of the different optimization
-techniques, {\tt generator}, {\tt yield}, {\tt $\lambda$\%} and {\tt do}
-change meaning slightly, but keep the same interface as their names
-suggest.
-
-\begin{figure}
-\begin{alltt}
-\(e-dispatch =\)
-(match e
-  [(var\(\superscript\mlab\) x)
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-    (do (\(\msto\)) ((v (delay (lookup-env \(\menv\) x))))
-      (yield (co \(\msto\) k v))))]
-  [(lit\(\superscript\mlab\) l)
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-    (do (\(\msto\)) () (yield (co \(\msto\) k n))))]
-  [(lam\(\superscript\mlab\) x e)
-   (define c (compile e))
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-    (do (\(\msto\)) () (yield (co \(\msto\) k (clos x c \(\menv\))))))]
-  [(app\(\superscript\mlab\) f e)
-   (define cf (compile f))
-   (define ce (compile e))
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-    (define-values (\(\msto\)* a) (push state))
-    (yield (ev \(\msto\)* cf \(\menv\) (ar c \(\menv\) a))))]
-  [(ife\(\superscript\mlab\) e0 e1 e2)
-   (define c0 (compile e0))
-   (define c1 (compile e1))
-   (define c2 (compile e2))
-   (\(\lambda\)\% (\(\msto\) \(\menv\) k)
-     (do (\(\msto\)) [(\(\msto\)* a) (push state)]
-      (yield (ev \(\msto\)* c0 \(\menv\) (ifk c1 c2 \(\menv\) a)))))])
-
-(define (step state)
-  (match state
-    [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
-    [(co \(\msto\) k v)
-     (match k
-       ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
-        (yield (ans \(\msto\) v))))]
-       [(ar\(\superscript\mcntr\subscript\mlab\) e \(\menv\) a)
-        (generator
-         (do (\(\msto\)) ([\(\msto*\) (join \(\msto\) a\(\superscript{f}\) (force \(\msto\) v))])
-           (yield (ev \(\msto*\) e \(\menv\) (fn\(\superscript\mcntr\subscript\mlab\) a\(\superscript{f}\) a)))))
-       [(fn\(\superscript\mcntr\subscript\mlab\) fa a)
-        (generator
-         (do (\(\msto\)) ([\(\msto*\) (join \(\msto\) a\(\superscript{a}\) (force \(\msto\) v))]
-                  [k <- (get \(\msto\) a)]
-                  [f <- (get \(\msto\) fa)])
-           (yield (ap\(\superscript\mcntr\subscript\mlab\) \(\msto*\) f a\(\superscript{a}\) k))))]
-       [(fi\(\superscript\mcntr\subscript\mlab\) c a \(\menv\))
-        (generator
-         (do (\(\msto\)) ([k <- (get \(\msto\) l)]
-                  [v <- (force \(\msto\) v)])
-           (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
-    [(ap\(\superscript\mcntr\subscript\mlab\) \(\msto\) fun a k)
-     (match fun
-       [(clos x e \(\menv\))
-        (generator
-         (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
-           (yield (ev \(\msto\)* e \(\menv\)* k))))]
-       [(? op? o)
-        (do (\(\msto\)) ([k <- (get \(\msto\) l)]
-                 [v <- (force \(\msto\) v)]
-                 [v (\(\interpdelta\) o (list v))])
-          (yield (co \(\msto\) k v)))
-       [_ (generator (do (\(\msto\)) () (continue)))]))]))
-\end{alltt}
-\caption{Parameterized Abstract$^2$ machine for ISWIM}
-\label{fig:paam}
-\end{figure}
-
-\begin{figure}
-\begin{alltt}
-(define (compile e) \(e-dispatch\))
-(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\machstep\) (\(\lambda\) (\(\msto\) \(\menv\) k) body \ldots)
-(yield (ev \(\msto\) c . rest)) \(\longmapsto\) (c \(\msto\) . rest)
-(yield e) \(\longmapsto\) (yield-meaning e)
-\end{alltt}
-\caption{Meaning of forms if compiled}
-\label{fig:cfm}
-\end{figure}
-
-\begin{figure}
-\begin{alltt}
-(define (compile e) e)
-(\(\lambda\)\% (\(\msto\) \(\menv\) k) body \ldots) \(\longmapsto\) (let () body ...)
-(yield e) \(\longmapsto\) (yield-meaning e)
-\end{alltt}
-\caption{Meaning of forms if not compiled}
-\label{fig:ncfm}
-\end{figure}
-
-Why generators? Relations in general cannot be written as computable
-functions, but in our case we have a decidable relation with finitely
-many pairs sharing the same left component. There are several
-implementation strategies for such relations, some more effective than
-others. For instance, a silly implementation strategy would be
-defining a binary function that decides membership in the relation. We
-want the definition to drive the interpreter, so we might naturally
-decide to write a function that takes a state and returns a set of all
-the next states. As we described in the imperative workset section,
-this intermediate allocation is unnecessary and wasteful. What we
-really want is a way of pulling on the reduction relation for more
-next states as the fixpoint is ready for them.
-
-For intermediate sets, {\tt yield} is {\tt return} in the set
-monad. For the imperative worklist, {\tt yield} performs mutation
-behind the scenes. Indeed, we could use real generators! This is the
-only case where {\tt generator} constructs a nullary generator; in the
-other cases it evaluates to its body. Racket's generators are not
-tuned for performance, so we actually lose efficiency in this case,
-but the unifying vocabulary is instructive nonetheless.
-
-The {\tt $\lambda$\%} form constructs a real closure in the
-compilation case and evaluates to its body otherwise.
-
-The {\tt do} form has the special job of maintaining the current store
-and accumulated states (if indeed those are being passed around). It
-uses Racket's syntax parameters ~\cite{dvanhorn:eli/stxparam} to
-track the most recent bindings of these accumulators so that {\tt
-  yield-meaning} can refer to them (say {\tt $\sigma$p} and {\tt
-  statep}). In the case of state passing and set-accumulated states, its
-definition is simply
-
-\begin{center}
-\begin{alltt}
-(yield-meaning e) \(\longmapsto\) (values \(\sigma\)p (set-add statep e))
-\end{alltt}
-\end{center}
-
-The linguistic elements described above were all reusable to implement
-the various languages' abstract interpreters. The language of our
-benchmarks, Scheme, has compound data (e.g. vectors and conses), which
-introduces choices for abstraction.
+vector the size of the largest label (call it $\mlab_{max}$) assigned, plus
+1. The parser should be assigning new, fresh names to variables
+anyway, so this change is a one-liner.
+
+This technique by itself only works in 0-CFA, since there are many more
+addresses that can be allocated in 1-CFA and higher. To support
+general $k$-CFA in a vector representation, we use a growable vector
+and a map from addresses to assigned indices. When assignable indices
+run out, we grow the vector. The cost of this map is less than a hash
+representation for the whole store since typically there are more
+references to an address than allocations of it. This is more
+space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
+up-front and addressing by the contour converted to polynomial, since
+many combinations of labels are not likely to occur. The all up-front
+allocation again breaks down when we add compound data, which
+introduce new kinds of labels (e.g. left and right components of a
+tuple).
 
@@
 
-We discovered that the choice of abstract domain for literal data
-(such as the large quoted axiom database in the Boyer benchmark)
-drastically affected the performance of our analysis on the benchmark
-suite. It is worth discussing the ``obvious'' and na\"ive abstraction
-that is natural to a reader of
-~\citep{dvanhorn:VanHorn2011Abstracting}, and other easy-to-implement
-yet effective approaches.
+Practical issues of performance arise when one chooses the
+``obvious,'' or ``natural'' abstraction for compound data. Our running
+example will be the classic Boyer benchmark. The detail of note for
+the Boyer benchmark is that it has large list literals, the exact
+value of which do not particularly matter for control-flow
+analysis. If we interpret a list literal of 30 elements (that again
+have 10 elements each) as 300 applications of {\tt cons}, and then
+interpret a function that traverses the tree, we will end up
+interpreting that function at least 300 times, when once or twice
+would do.
+
+We will first explain the natural abstraction of tuples, and then we
+will explain a less precise allocation strategy from uniform $k-CFA$
+that we use for large compound data literals that we modified from the
+implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.
 
@@
 
-The uniform way they approach a simple abstraction strategy is to cut
+The uniform way AAM approaches a simple abstraction strategy is to cut
 recursion out of the data definition by tying the recursive knot
-through the abstract store. For Scheme, the grammar for values looks like the following:
+through the abstract store. For Scheme, the grammar for values looks
+like the following:
 
@@
 store, and return the flattened {\tt (cons $a$ $d$)} value. Since
-these addresses are all distinguished at different syntactic
-callsites, and quoted lists really are sugar for a sequence of calls
-to {\tt cons}, this abstraction explodes the value space. Analyzing a
-function that counts the number of atoms in a literal s-expression
-would actually interpret that function at least that number of times
-(more because of intermediate conses). Indeed, even in our fastest
-parameterization, this abstraction causes the analysis of Boyer to be
-430 times slower than the approach we will now describe.
+these addresses are all distinguished at different syntactic callsites
+in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
+for a sequence of calls to {\tt cons}, this abstraction explodes the
+value space. Analyzing a function that counts the number of atoms in a
+literal s-expression would actually interpret that function at least
+that number of times (more because of intermediate conses). Indeed,
+even in our fastest parameterization, this abstraction causes the
+analysis of Boyer to be 430 times slower than the approach we will now
+describe.
 
@@
 literal lists is dramatically smaller in typical Scheme programs. We
-use the above abstraction for these syntactic uses, but choose a
-different strategy for quoted lists. Quotation is special because it
-cannot introduce function values or mutable values, which is important
-to enhancing our technique's soundness to conceptual complexity ratio.
+use the above abstraction for these syntactic uses, but choose to
+interpret literal lists as not always sugar for cascading {\tt
+  cons}es. In particular, if a list literal is ``too big'' (in our
+case length > 1), we interpret the list as a circular data structure;
+the right address points back to the {\tt cons} itself, and the left
+address points to all of the elements of the list. We take this
+farther and join the list elements together in with a coarse
+type-based abstraction. Quotation is special because it cannot
+introduce function values, which is important to enhancing our
+technique's soundness to conceptual complexity ratio. If we join two
+values of different type together, we don't get ``anything,'' which
+has complex meaning in higher-order languages (Shivers' escape
+semantics) and is overly approximate. We instead get ``any quotable
+value,'' which has much simpler semantics.
 
@@
        and similar definitions for immutable vectors.}
- \item{Extend the $\interpdelta$ axioms to include conservative meaning for these new values (e.g.
-{\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
+ \item{Extend the $\interpdelta$ axioms to include conservative
+   meaning for these new values (e.g. {\tt (car QData)} = {\tt QData},
+   {\tt (add1 QData)} = {\tt Number} and log ``possible type error'')
+   and allow them to allocate addresses and change the store}
 \end{itemize}
@@
 \end{figure}
-
-The alternative to {\tt QData} is to go higher in the lattice to
-$\top$, which has more complicated meaning (Shivers' escape semantics) and
-is overly approximate.
 
@@
 
-% \include{appendix}
+\appendix
+\section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
+\label{sec:accept}
+
+\cite{dvanhorn:nielson-nielson-popl97} \cite{dvanhorn:Neilson:1999}
+
+This machine's allocation strategy mimics the Uniform k-CFA analysis
+in Principles of Program Analysis, which is defined in terms of
+``$\delta$ contours.''  However, because the machine represetation makes
+context explicit via continuations, we can calculate these contours
+rather than thread them throught the evaluator.  In other words, we
+can use the CESK* machine without modification to obtain Uniform k-CFA
+by way of a simple allocation strategy.  (In this way, it's a
+simplification of the presentation in JFP.)
+
+NNH uses a coinductive acceptability relation to specify Uniform
+k-CFA:
+
+\[
+   C,R \models^{ce}_\delta E
+\]
+
+The cache and global environment form a finite store-like structure
+holding bindings and return values.  The contour environment ce maps
+variables to locations in R which contains their bindings, just as the
+environment of the CESK* machine does.  The current contour delta is a
+string of application labels describing the enclosing context under
+which this term is being analyzed (or evaluated).  If you view the
+acceptability relation as a big-step evalator, the
+$(\widehat C,\widehat\rho)$ component should be seen as a global
+store ce is the environment mapping variables to their locations.
+
+Starting form the initial configuration for a program and iterating
+the machine transition relation until reaching a fixpoint of reachable
+states will \emph{underestimate} the acceptability relation of Uniform
+k-CFA.  You can recover acceptability by feeding this store back into
+the initial configuration and iterating again.  Repeating this process
+until a complete run of the program reaches no new states will be the
+least solution that is acceptable.
+
+HOWEVER.  Why should we care about acceptability?  What this
+machine computes is safe.  In other words, it computes a more
+precise characterization of the run-time behavior of a program.  In
+doing is so, it actually saves work (as can be seen above).
+
+An Example:
+
+\begin{alltt}
+ (let ((id (\(\lambda\) (x) x)))
+   (begin (id 1) (id 2)))
+\end{alltt}
+
+Under Uniform 0-CFA, we would have:
+\[
+   [{\tt x} \mapsto \{{\tt 1}, {\tt 2}\}] \in \widehat\rho
+\]
+
+in the least solution to $\models$.  This says that, when run, 'x' is
+bound to 1 or 2.
+
+Under the machine semantics using a 0CFA allocation policy, the trace
+semantics of the machine show that x is bound to 1, and that at some
+later point, x becomes bound to 1 or 2.  Moreover, the machine would
+show that (id 1) evaluates to 1 and only 1, while Uniform 0CFA must
+give that (id 1) is either 1 or 2 to be acceptable.  We don't see any
+value in these kinds of false flows that are due to the global and
+timeless aspects of C,R which acceptability requires the heap to be
+both finite and unchanging over the course of abstract
+interpretation. (Another view of the difference: the machine abstracts
+a program's execution as a \emph{finite state machine} that mimics the
+machine interpretation of the program; the aceptability relation of
+Uniform \(k\)-CFA abstracts a program's execution as a \emph{finite
+  map} that mimics the big-step evaluator: from terms to (sets of)
+values.)
+
+\subsection{Another problem with acceptability: Temporal ignorance}
+
+The small-step approach to static analysis brings subtle yet important temporal
+richness not found in classical analyses for higher-order programs.
+%
+Classical analyses (ultimately) compute judgments on program terms and
+contexts, e.g., at expression $e$, value $x$ may have value $v$.
+%
+The judgments do not relate the order in which expressions and context may be
+evaluated in a program, e.g., a classical analysis has nothing to say with
+regard to question like, ``Do we always evaluate $e_1$ before $e_2$?'' or ``Is
+it always the case that a file handle is opened, read and then closed in that
+order?''
+
+Small-step analyses, by their nature, encode the temporal relationships between
+abstract states.
+%
+It is sensible to make temporal queries of a small-step analysis.
+%
+Of course, this does not come for free: respecting temporal order imposes an
+order in which states and terms may be evaluated \emph{during} the analysis.
+%
+Classical analyses can (and do) evaluate expressions in any order, or in some
+cases, even in parallel~\cite{might:Prabhu:2010:EigenCFA}.
+%
+Relaxing that restriction on order affords additional optimizations that we
+have \emph{not} performed.
+
+We avoid sacrificing order not simply because we are interested in the
+questions it allows us to ask, but because considering temporal order actually
+improves the precision of the analysis itself.
+
+\section{Pushdown Analysis}
+
+It is straightforward to instantiate a \emph{pushdown} abstraction by
+bounding only the variable binding portion of the heap, but using a
+unique allocation strategy for continuations.  Such a strategy
+abstracts a program's execution as a \emph{pushdown automata}
+that mimics the machine interpretation.  This strategy therefore
+models the abstract stack in a true stack like fashion and always
+properly matches function calls with their return.
+
+Although such analyses can be formulated straightforwardly in the
+abstract machine approach, it is not clear all of the techniques of
+this paper can be applied to similar effect in the pushdown context.
+The main problem is calculation of an analysis can no longer be
+computed as the fixed point of the machine transition relation.
+Although there are several implementations (CFA2,ICFP'12), they
+operate at speeds roughly on par with our starting point: unoptimized
+store widened
+machines. \cite{dvanhorn:Earl2012Introspective,dvanhorn:Vardoulakis2011CFA2}
 

### 2012-11-12 paper/paper.tex

@@
 
-%% \paragraph{Primitives}
-
-%% \begin{align*}
-%% \mnum+1 &\in \interpdelta(\saddone,\mnum) &
-%% \mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
-%% \strue &\in \interpdelta(\szerohuh,\szero) &
-%% \sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
-%% \end{align*}
-
-%% \begin{align*}
-%% \sNum &\in \hat\interpdelta(\saddone,\mnum) &
-%% \sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
-%% \strue &\in \hat\interpdelta(\szerohuh,\sNum) &
-%% \sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
-%% \strue &\in \hat\interpdelta(\szerohuh,\szero) &
-%% \sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
-%% \end{align*}
-
-%\newpage
+\paragraph{Primitives}
+
+\begin{align*}
+\mnum+1 &\in \interpdelta(\saddone,\mnum) &
+\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
+\strue &\in \interpdelta(\szerohuh,\szero) &
+\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+\end{align*}
+
+\begin{align*}
+\sNum &\in \hat\interpdelta(\saddone,\mnum) &
+\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
+\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
+\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
+\strue &\in \hat\interpdelta(\szerohuh,\szero) &
+\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+\end{align*}
+
+%% \begin{figure}
+%% \begin{alltt}
+%%   ;; State \(\to\) \(\Set(\)State\()\)
+%%   (define (step \(\mstate\))
+%%     (match \(\mstate\)
+%%       [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
+%%        (match \(\mexp\)
+%%          [(var\(\superscript\mlab\) \(\mvar\))
+%%           (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
+%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
+%%          [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
+%%          [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
+%%          [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
+%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
+%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
+%%          [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
+%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
+%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
+%%       [(co \(\msto\) \(\mkont\) \(\mval\))
+%%        (match \(\mkont\)
+%%          ['mt (set (ans \(\msto\) \(\mval\)))]
+%%          [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
+%%          [(fn\(\superscript\mlab\) \(\mval'\))
+%%           (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
+%%             (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
+%%          [(fi\(\superscript\mlab\) c a \(\rho\))
+%%           (for/set ((k (get-cont \(\msto\) l)))
+%%             (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
+%%       [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
+%%        (match fun
+%%          [(clos l \(\mvar\) \(\mexp\) \(\rho\))
+%%           (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
+%%           (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
+%%          [(op \(\mop\))
+%%           (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
+%%                      (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
+%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
+%%          [_ (set)]))]))
+%% \end{alltt}
+%% \caption{Implementation of machine transition relation.}
+%% \end{figure}
+
+\newpage
 \section{Baseline analyzer}
@@
 
-\section{Mostly precision preserving recipe}
+\section{Implementation techniques}
 \label{sec:opt}
@@
    the 0-CFA case. (preallocation)}
+ \item{[SS] coarsely abstract literal data (abstracting compound data)}
 \end{itemize}
 
-Each technique requires minimal changes to the reduction
-relation/fixpoint, and each improves performance by integer
-factors. The first three remain in the purely functional realm. The
-final three use mutation in small, localized places, so they do not
-increase conceptual complexity drastically.
+Some techniques preserve the precision of the underlying analysis, and
+others do not. We will discuss the design decisions behind the
+precision-lossy techniques.  Each technique requires minimal changes
+to the reduction relation/fixpoint, and each improves performance by
+integer factors. The first three remain in the purely functional
+realm. The final three use mutation in small, localized places, so
+they do not increase conceptual complexity drastically.
 
@@
 
-All following techniques preserve precision.
 % Lazy:  cpu time: 32481 real time: 32881 gc time: 547
@@
 
-Notice that the only difference between concrete and abstract interpretation
-is which addresses are pushed.
+The compilation step converts expressions into functions that expect
+the other components of the {\tt ev} state. Its definition in figure
+\ref{fig:compile} shows close similarity to the rules for interpreting
+    {\tt ev} states. The next step is to change creation of {\tt ev} states to simply calling these functions. The modified reduction relation is shown in figure \ref{fig:caam}.
 
@@
 \caption{Compilation}
+\label{fig:compile}
 \end{figure}
@@
 
-\appendix
-\section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
-\label{sec:accept}
-
-\cite{dvanhorn:nielson-nielson-popl97} \cite{dvanhorn:Neilson:1999}
-
-This machine's allocation strategy mimics the Uniform k-CFA analysis
-in Principles of Program Analysis, which is defined in terms of
-``$\delta$ contours.''  However, because the machine represetation makes
-context explicit via continuations, we can calculate these contours
-rather than thread them throught the evaluator.  In other words, we
-can use the CESK* machine without modification to obtain Uniform k-CFA
-by way of a simple allocation strategy.  (In this way, it's a
-simplification of the presentation in JFP.)
-
-NNH uses a coinductive acceptability relation to specify Uniform
-k-CFA:
-
-\[
-   C,R \models^{ce}_\delta E
-\]
-
-The cache and global environment form a finite store-like structure
-holding bindings and return values.  The contour environment ce maps
-variables to locations in R which contains their bindings, just as the
-environment of the CESK* machine does.  The current contour delta is a
-string of application labels describing the enclosing context under
-which this term is being analyzed (or evaluated).  If you view the
-acceptability relation as a big-step evalator, the
-$(\widehat C,\widehat\rho)$ component should be seen as a global
-store ce is the environment mapping variables to their locations.
-
-Starting form the initial configuration for a program and iterating
-the machine transition relation until reaching a fixpoint of reachable
-states will \emph{underestimate} the acceptability relation of Uniform
-k-CFA.  You can recover acceptability by feeding this store back into
-the initial configuration and iterating again.  Repeating this process
-until a complete run of the program reaches no new states will be the
-least solution that is acceptable.
-
-HOWEVER.  Why should we care about acceptability?  What this
-machine computes is safe.  In other words, it computes a more
-precise characterization of the run-time behavior of a program.  In
-doing is so, it actually saves work (as can be seen above).
-
-An Example:
-
-\begin{alltt}
- (let ((id (\(\lambda\) (x) x)))
-   (begin (id 1) (id 2)))
-\end{alltt}
-
-Under Uniform 0-CFA, we would have:
-\[
-   [{\tt x} \mapsto \{{\tt 1}, {\tt 2}\}] \in \widehat\rho
-\]
-
-in the least solution to $\models$.  This says that, when run, 'x' is
-bound to 1 or 2.
-
-Under the machine semantics using a 0CFA allocation policy, the trace
-semantics of the machine show that x is bound to 1, and that at some
-later point, x becomes bound to 1 or 2.  Moreover, the machine would
-show that (id 1) evaluates to 1 and only 1, while Uniform 0CFA must
-give that (id 1) is either 1 or 2 to be acceptable.  We don't see any
-value in these kinds of false flows that are due to the global and
-timeless aspects of C,R which acceptability requires the heap to be
-both finite and unchanging over the course of abstract
-interpretation. (Another view of the difference: the machine abstracts
-a program's execution as a \emph{finite state machine} that mimics the
-machine interpretation of the program; the aceptability relation of
-Uniform \(k\)-CFA abstracts a program's execution as a \emph{finite
-  map} that mimics the big-step evaluator: from terms to (sets of)
-values.)
-
-\subsection{Another problem with acceptability: Temporal ignorance}
-
-The small-step approach to static analysis brings subtle yet important temporal
-richness not found in classical analyses for higher-order programs.
-%
-Classical analyses (ultimately) compute judgments on program terms and
-contexts, e.g., at expression $e$, value $x$ may have value $v$.
-%
-The judgments do not relate the order in which expressions and context may be
-evaluated in a program, e.g., a classical analysis has nothing to say with
-regard to question like, ``Do we always evaluate $e_1$ before $e_2$?'' or ``Is
-it always the case that a file handle is opened, read and then closed in that
-order?''
-
-Small-step analyses, by their nature, encode the temporal relationships between
-abstract states.
-%
-It is sensible to make temporal queries of a small-step analysis.
-%
-Of course, this does not come for free: respecting temporal order imposes an
-order in which states and terms may be evaluated \emph{during} the analysis.
-%
-Classical analyses can (and do) evaluate expressions in any order, or in some
-cases, even in parallel~\cite{might:Prabhu:2010:EigenCFA}.
-%
-Relaxing that restriction on order affords additional optimizations that we
-have \emph{not} performed.
-
-We avoid sacrificing order not simply because we are interested in the
-questions it allows us to ask, but because considering temporal order actually
-improves the precision of the analysis itself.
-
-\section{Pushdown Analysis}
-
-It is straightforward to instantiate a \emph{pushdown} abstraction by
-bounding only the variable binding portion of the heap, but using a
-unique allocation strategy for continuations.  Such a strategy
-abstracts a program's execution as a \emph{pushdown automata}
-that mimics the machine interpretation.  This strategy therefore
-models the abstract stack in a true stack like fashion and always
-properly matches function calls with their return.
-
-Although such analyses can be formulated straightforwardly in the
-abstract machine approach, it is not clear all of the techniques of
-this paper can be applied to similar effect in the pushdown context.
-The main problem is calculation of an analysis can no longer be
-computed as the fixed point of the machine transition relation.
-Although there are several implementations (CFA2,ICFP'12), they
-operate at speeds roughly on par with our starting point: unoptimized
-store widened
-machines. \cite{dvanhorn:Earl2012Introspective,dvanhorn:Vardoulakis2011CFA2}
+% \include{appendix}
 

### 2012-11-12 paper/paper.tex
Balance.
Touching up prose.

@@
 \end{align*}
-For the duration of this paper, $\sqcup$ is a point-wise lifting of $\cup$:
+For the duration of this paper, since our example abstract value space
+has no interesting ordering, $\sqcup$ is a point-wise lifting of
+$\cup$:
 \begin{align*}
@@
 \newpage
-\section{Baseline analyzer}
+\section{Reduction semantics to baseline analyzer}
 \label{sec:baseline}
+
+The uniform $k$-CFA allocation strategy would make the fixpoint of the
+reduction relation in figure \ref{fig:aam} a computable analysis, but
+not an effective one. This is not the strategy that AAM, nor we,
+recommend. We explain a succession of approximations to reach the
+baseline analysis that we compare performance against. But first, we
+need a fixpoint combinator to drive the succession of reduction
+relations we define.
 
@@
 The final approximation we make to get to our baseline analysis is
-store-allocating results of application sub-expressions. Notice that
+store-allocating results of application sub-expressions. The semantics
+of the previous section is where AAM leaves off. However, notice that
 the $\kfn{}$ continuation stores a value - this makes the space of
-continuations quadratic rather than linear in 0-CFA. We instead
-allocate an address for the value position when we create the
-continuation. This address and the tail address are both determined by
-the label of the application point, so the space becomes linear and
-the overall complexity drops to cubic. Note that this is a critical
-abstraction in languages with n-ary functions, since otherwise the
-continuation space becomes ${\mathcal O}(n^n)$!
+continuations quadratic rather than linear in 0-CFA. A linear space
+drops the complexity to cubic.
+
+To achieve this, we instead allocate an address for the value position
+when we create the continuation.  We will refer to this change in the
+semantics as store-allocated results. This address and the tail address are
+both determined by the label of the application point, so the space
+becomes linear and the overall complexity drops to cubic. Note that
+this is a critical abstraction in languages with n-ary functions,
+since otherwise the continuation space becomes ${\mathcal O}(n^n)$!
 
@@
 results, and thus the precision aligns with an analysis specialized to
-either ANF or CPS. We can also play a representation trick and remove
-$\menv$ since we know it is the identity function in 0-CFA. We will
-refer to this change in the semantics as value indirection.
+ANF. We can also play a representation trick and remove
+$\menv$ since we know it is the identity function in 0-CFA.
 
@@
 
-We discuss a variety of techniques for increasing performance by
+We discuss a succession of techniques for increasing performance by
 decreasing the interpretive overhead of abstract interpretation. They
 come in two flavors: decreasing the size of the represented state
-space [SS], and decreasing intermediate data structures [DS].
+space [SS], and decreasing the amount of intermediate data structures [DS].
 
@@
  \item{[DS] use a global mutable workset}
- \item{[DS] represent the global store as a global, mutable vector in
-   the 0-CFA case. (preallocation)}
+ \item{[DS] represent the global store as a global, mutable vector (preallocation)}
  \item{[SS] coarsely abstract literal data (abstracting compound data)}
@@
 to the reduction relation/fixpoint, and each improves performance by
-integer factors. The first three remain in the purely functional
-realm. The final three use mutation in small, localized places, so
-they do not increase conceptual complexity drastically.
+integer factors. The first three and last remain in the purely
+functional realm. The final three use mutation in small, localized
+places, so they do not increase conceptual complexity
+drastically. Also, preallocation and mutable workset are standard
+implementation practices \cite{ianjohnson:Shivers:1991:CFA}.
 
@@
 variable reference transitions to several new states that comprise the
-cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
+cartesian product of all the values of {\tt f}, {\tt x} and {\tt y} (or in the our baseline analysis, the sum),
 only to be joined back together in their respective application positions. We would
@@
 For this, we introduce a new value for a delayed lookup that acts to
-quotient the state space by the values at the delayed address. Its changes to the semantics are highlighted in the following rules:
+quotient the state space by the values at the delayed address. Its
+changes to the semantics are highlighted in the following rules:
 
@@
 
-This semantics introduces a subtle precision difference over the
+\paragraph{Precision loss} This semantics introduces a subtle precision difference over the
 baseline that is negligible. Consider a configuration where a
@@
 memory needed for this eager non-determinism. Indeed, were the
-variable reference a step later and another binding is not made in
+variable reference a step later and another binding not made in
 that step, the results of the two approaches are the same.
 
-The administrative nature of laziness means that we could remove the
+\paragraph{Avoiding precision loss} The administrative nature of laziness means that we could remove the
 loss in precision by essentially duplicating the reduction relation to
 specialize variable lookup. This works since in the semantics of ISWIM
-with value indirection consumes the value component of states in
+with store-allocated results consumes the value component of states in
 one step. This is not the case for semantics that replicate the value
@@
 
-We can eliminate the {\tt ev} states from the run-time interpretation
-of a program by specializing the machine transition relation to the
-program being analyzed. This eliminates interpretative overhead by
-first compiling the program into ``byte code'' instructions.
+Laziness removes several represented states that would then have to be
+stepped by the fixpoint combinator.  We can further eliminate the {\tt
+  ev} states from the run-time interpretation of a program by
+specializing the machine transition relation to the program being
+analyzed. This eliminates interpretative overhead by first compiling
+the program into ``byte code'' instructions.
 
@@
 \ref{fig:compile} shows close similarity to the rules for interpreting
-    {\tt ev} states. The next step is to change creation of {\tt ev} states to simply calling these functions. The modified reduction relation is shown in figure \ref{fig:caam}.
+    {\tt ev} states. The next step is to change creation of {\tt ev}
+    states to simply calling these functions. The modified reduction
+    relation is shown in figure \ref{fig:caam}.
 
@@
 
-Each step in the fixpointer for the above techniques requires joining
+Each step in the fixpoint for the above techniques requires joining
 several, commonly large, stores together. Even with immutable hashes
-that utilize sharing, this is memory-intensive. Worse, this
+that utilize sharing, this is memory- and time-intensive. Worse, this
 contributes an obscene fraction of the runtime. If we instead track
@@
 \begin{align*}
-bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
-\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
-               \maddr &= \mvar\mcntr' \\
 (\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
@@
 \end{align*}
-
-We also lift $\machstep$ differently to accommodate for this asymmetry
+\begin{align*}
+bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
+\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
+               \maddr &= \mvar\mcntr'
+\end{align*}
+
+We also lift $\machstep$ to accommodate for this asymmetry
 in the input and output. For each state that is stepped, we feed the
@@
 
-To ensure termination, we guard adding $c$ to {\tt todo} by the following check:
+The new fixpoint calculation is defined in
+figure~\ref{fig:imperative}.  To ensure termination, we guard adding
+$c$ to {\tt todo} by the following check:
 \begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
@@
 
-After all steps complete, we apply $\msdiff$ to $\msto$ and increase
+After all steps complete, we apply $\msdiff$ to $\msto$ imperatively and increase
 {\tt T} as long as there was at least one change in $\msdiff$. This
-logic is only sound if we know that each $\ttuple{\maddr}{\mval{s}}$
+logic leads to termination if we know that each $\ttuple{\maddr}{\mval{s}}$
 in $\msdiff$ would change the value of $\maddr$ in the current
@@
 introduce new kinds of labels (e.g. left and right components of a
-tuple).
+tuple at the contour of creation).
 
@@
 the Boyer benchmark is that it has large list literals, the exact
-value of which do not particularly matter for control-flow
-analysis. If we interpret a list literal of 30 elements (that again
-have 10 elements each) as 300 applications of {\tt cons}, and then
-interpret a function that traverses the tree, we will end up
-interpreting that function at least 300 times, when once or twice
-would do.
+values of which do not particularly matter for many analyses. If we
+interpret a list literal of 30 elements (that again have 10 elements
+each) as 300 applications of {\tt cons}, and then interpret a function
+that traverses the tree, we will end up interpreting that function at
+least 300 times, when once or twice would do.
 
 We will first explain the natural abstraction of tuples, and then we
-will explain a less precise allocation strategy from uniform $k-CFA$
+will explain a less precise allocation strategy from uniform $k$-CFA
 that we use for large compound data literals that we modified from the
@@
 that number of times (more because of intermediate conses). Indeed,
-even in our fastest parameterization, this abstraction causes the
+even in our fastest implementation, this abstraction causes the
 analysis of Boyer to be 430 times slower than the approach we will now

### 2012-11-12 paper/paper.tex
Talked about mutable literal data, gave a breakdown of the implementation (in comments) and gave better explanation for deltas.

@@
 \usepackage{graphicx}
+\usepackage{balance}
 
@@
 
-In section~\ref{sec:impl}, we describe a more complete implementation
-scaled up and capable of analyzing Scheme benchmark programs.
-
-In section~\ref{sec:eval} we evaluate this implementation against a
+In section~\ref{sec:eval} we evaluate our implementation against a
 set of benchmark programs drawn from the literature.  Many benchmarks
@@
 \begin{itemize}
- \item{[SS] quotient by sets of values (lazy non-determinism)}
- \item{[SS] remove {\tt ev} states (abstract compilation)}
- \item{[DS] operate on store changes rather than whole stores (store deltas)}
- \item{[DS] represent the store as a global, mutable hash and
+ \item{[SS] lazy non-determinism: quotient by sets of values}
+ \item{[SS] abstract compilation: eliminate {\tt ev} states}
+ \item{[DS] store deltas: operate on store changes rather than whole stores}
+ \item{[DS] timestamp: represent the store as a global, mutable hash and
    determine a state's store by the number of changes made to the
-   global store at the time of visiting (timestamp)}
- \item{[DS] use a global mutable workset}
- \item{[DS] represent the global store as a global, mutable vector (preallocation)}
- \item{[SS] coarsely abstract literal data (abstracting compound data)}
+   global store at the time of visiting}
+ \item{[DS] preallocation: represent the global store as a global, mutable vector}
+ \item{[SS] uniform literal approximation: coarsely abstract literal data}
 \end{itemize}
@@
 to the reduction relation/fixpoint, and each improves performance by
-integer factors. The first three and last remain in the purely
-functional realm. The final three use mutation in small, localized
-places, so they do not increase conceptual complexity
-drastically. Also, preallocation and mutable workset are standard
-implementation practices \cite{ianjohnson:Shivers:1991:CFA}.
+integer factors.
 
@@
 specializing the machine transition relation to the program being
-analyzed. This eliminates interpretative overhead by first compiling
-the program into ``byte code'' instructions.
+analyzed. This precision-preserving transformation eliminates
+interpretative overhead by first compiling the program into ``byte
+code'' instructions.
 
@@
 \begin{align*}
-\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\mval,\msto} \text{ where }\mval\in\msto(\menv(\mvar))
+\compile{\_} &: Expr \to Env \times Store \times Kont \to State \\
+\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\saddr{\menv(\mvar)},\msto}
 \\
@@
 Each step in the fixpoint for the above techniques requires joining
-several, commonly large, stores together. Even with immutable hashes
-that utilize sharing, this is memory- and time-intensive. Worse, this
-contributes an obscene fraction of the runtime. If we instead track
-the changes to the store made over all steps and afterward update the
-store, we end up with much less allocation and computational overhead.
-
-We represent changes as a list of pairs of address to set of storeable
-values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
-$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins as empty for
-each step. Applying the changes is straightforward:
+several, commonly large, stores together. Each step is unlikely to
+always modify all addresses of the store, so this is needlessly
+memory- and time-intensive. We can instead accumulate store changes
+and apply them after all steps have completed. This uses far fewer
+join operations, leading to much less allocation and computational
+overhead.
+
+We represent changes as $Store\msdiff = List[Addr \times
+  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
+becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
+begins as empty for each step. Applying the changes is
+straightforward:
 
@@
 \end{align*}
+Compilation changes to additionally take a $\msdiff$ component, so the
+above rule's right hand side would instead be $k^{\mcntr'}(\menv',
+\msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.
 
@@
 in the input and output. For each state that is stepped, we feed the
-output changes to the next so that all changes get accumulated.
+output changes to the next so that all changes get accumulated:
 
@@
 but for determining if a state has been visited before (vis. $f(X)
-\setminus S$ in ${\mathcal G}$), we still have to compare entire stores - an
-expensive computation.  We modify a technique created in Shivers'
-dissertation~\cite{ianjohnson:Shivers:1991:CFA}, timestamp
-approximation, that removes the need for joining whole stores by
-manipulating a global store imperatively. The original technique loses
-precision because the starting store for each state in a machine
-configuration can change between stepping all states and introduce
-more flows than originally possible. However, by delaying the store
-changes until after all steps have completed, we can maintain our
-precision. In a variety of experiments, we observed that the overhead
-for storing store-deltas was outweighed by the precision keeping flows
-in check. That is to say, the analysis was faster {\it and} more
-precise.
-
-We pair this technique with a mutable workset. That is, to avoid the
-intermediate sets introduced by the $State \to \Set(State)$ signature
-of $\machstep$, we change the fixpoint algorithm to use a global
+\setminus S$ in ${\mathcal G}$), we still have to compare entire
+stores - an expensive computation.  We modify a technique due to
+Shivers~\cite{ianjohnson:Shivers:1991:CFA}, timestamp approximation,
+that removes the need for joining whole stores by manipulating a
+global store imperatively. The original technique loses precision
+because states are stepped in sequence, modifying the store in place -
+more values or continuations can be added before another state gets to
+step. This interference means more possible flows (i.e. less
+precision) and a nondeterministic analysis - the order the set of states is
+traversed is not necessarily deterministic. However, by delaying the
+store changes until after all steps have completed, we can maintain
+our precision and determinism. In a variety of experiments, we
+observed that the time and space overhead for storing store-deltas was outweighed by
+the precision keeping flows in check. That is to say, the analysis was
+faster {\it and} more precise.
+
+We pair this technique with a mutable workset to avoid the
+intermediate sets introduced by the $State \times Store \to \Set(State) \times Store\msdiff$ signature
+of $\machstep$. This necessitates a change to the fixpoint algorithm to use a global
 workset, {\tt todo}, that $\machstep$ imperatively updates with the
-states that it reaches.
+states that it would have otherwise accumulated in a set that would later be added to the workset.
 
@@
  \item{{\tt todo}, the workset}
- \item{{\tt seen}, a map from state to the timestamp at which it was last seen}
+ \item{{\tt seen}, a map from states to the timestamps at which they were last seen}
  \item{{\tt $\msdiff$}, the store changes for the current step}
@@
 
-\subsection{Abstracting Compound Data}
+\subsection{Abstracting literal compound data}
 
@@
   cons}es. In particular, if a list literal is ``too big'' (in our
-case length > 1), we interpret the list as a circular data structure;
+case length $>$ 1), we interpret the list as a circular data structure;
 the right address points back to the {\tt cons} itself, and the left
@@
 
+\paragraph{Mutable literal data} The Scheme we model is nonstandard in that it has immutable literal
+data (the Scheme benchmarks did not mutate their literal data). To
+support mutable literal data, abstract values like {\tt QPair} would
+be like regular conses, only point to a preallocated pair of addresses
+that hold {\tt QData} as well. This is because conses can be mutated
+to hold any value, not just quoted data.
+
 \section{Evaluation}
 \label{sec:eval}
+
+Our implementation of the analysis used in the evaluation is roughly
+3500 lines of Racket code.
+%% 400 lines for the core, 700 lines for
+%% abstraction over optimizations, 1500 lines for primitives and standard
+%% list operations, 700 lines for instantiations to different
+%% optimizations and 300 for parsing and macro transformers.
 
@@
 
+\balance
 \bibliographystyle{plain}

### 2012-11-12 paper/paper.tex
Removed repetition in baseline.
Start of evaluation section.  Some related work.

--- 2012-11-12 paper/paper.tex
@@
 originally applied to features such as higher-order functions,
-stack-inspection, exceptions, laziness, first-class continuations, and
+stack inspection, exceptions, laziness, first-class continuations, and
 garbage collection.  It has since been used to verify actor-
@@
 
+We have implemented, optimized, and evaluated \emph{all} of the
+machines presented in their JFP paper
+\cite{dvanhorn:VanHorn2012Systematic}, with the exception of the
+machine for performing pushdown flow analysis.  This includes machines
+supporting higher-order functions, state, first-class control,
+laziness, and stack inspection.
+
+For each benchmark, we collect analysis times, peak memory usage, and
+the rate of states-per-second explored by the analysis for each of the
+optimizations discussed in section~\ref{sec:opt}, culmulatively
+applied.  The analysis is stopped after consuming 30 minutes of time
+or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
+use the timeout limits as a lower bound on the actual time required,
+thus giving a conservative estimate of improvements.
+
+All benchmarks are calculated as an average of N runs on a [MACHINE
+  SPECS].
+
+\begin{figure}
+\begin{tiny}
+\begin{alltt}
+Program  |  LOC | B. Time | O. Time | B. Space | O. Space | B. Rate | O. Rate
+----------------------------------------------------------------------------
+nucleic  | 3508 |
+matrix   |  747 |
+nbody    | 1434 |
+earley   |  667 |
+maze     |  681 |
+church   |   41 |
+lattice  |  214 |
+boyer    |  642 |
+mbrotz   |   68 |
+\end{alltt}
+\end{tiny}
+\end{figure}
+
+\begin{itemize}
+
+\item {\bf nucleic}: a floating-point intensive application taken from
+  molecular biology that has been used widely in benchmarking
+  functional language
+  implementations~\cite{dvanhorn:Hartel1996Benchmarking} and analyses
+  (e.g.~\cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+  It is a constraint satisfaction algorithm used to determine the
+  three-dimensional structure of nucleic acids.
+
+\item {\bf matrix} tests whether M is maximal among all matrices of
+  the same dimension obtainable by simple reordering of rows and
+  columns and negation of any subset of rows and columns.  It is
+  written in continuation-passing style (used in
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+
+\item {\bf nbody}: implements the Greengard multipole algorithm for
+  computing gravitational forces on point masses distributed uniformly
+  in a cube (used in
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+
+\item {\bf earley}:  Earley's parser written by Marc Feeley.
+
+\item {\bf maze}: generates a random maze and finds a path (used in
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
+  Makes use of Scheme's {\tt call-with-current-continuation}
+  operation.
+
+\item {\bf church}: tests distributivity of multiplication over
+  addition for a few Church numerals (introduced by
+  \cite{dvanhorn:Vardoulakis2011CFA2}).
+
+\item {\bf lattice}: enumerates the maps between two lattices (used in
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
+
+\item {\bf boyer}: a term-rewriting theorem prover (used in
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
+
+\item {\bf mbrotZ}: generates Mandelbrot set fractal using complex
+  numbers.
+
+\item {\bf graphs}: counts the number of directed graphs with a
+  distinguished root and \(k\) vertices, each having out-degree at
+  most 2. It makes extensive use of higher-order procedures and is
+  written in a continuation-passing style (used by
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
+
+\end{itemize}
+
 \begin{figure*}
@@
 
-\cite{dvanhorn:Earl2012Introspective}
-
-\cite{dvanhorn:wright-jagannathan-toplas98}
-
-Other benchmarks
-
 \section{Related work}
 \label{sec:related}
+
+\paragraph{Abstracting Abstract Machines}
+
+This work clearly closely follows Van Horn and Might's original papers
+on abstracting abstract
+machines~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic},
+which in turn is one piece of the large body of research on flow
+analysis for higher-order languages (see
+Midtgaard~\cite{dvanhorn:Midtgaard2011Controlflow} for a thorough
+survey).  The AAM approach sits at the confluence of two major lines
+of research: (1) the study of abstract
+machines~\cite{dvanhorn:landin-64} and their systematic
+construction~\cite{dvanhorn:reynolds-hosc98}, and (2) the theory of
+abstract interpretation
+\cite{dvanhorn:Cousot:1977:AI,dvanhorn:Cousot:1979:Galois}.
+
+\paragraph{Frameworks for flow analysis of higher-order programs}
+
+The analysis most similar to that presented in section~\ref{sec:aam}
+is the infinitary control-flow analysis of Nielson and
+Nielson~\cite{dvanhorn:nielson-nielson-popl97} and the unified
+treatment of flow analysis by Jagannathan and
+Weeks~\cite{dvanhorn:jagannathan-weeks-popl95}.  Both are
+parameterized in such a way that in the limit, the analysis is
+equivalent to an interreter for the language, just as is the case
+here.  What is different is that both give a constraint-based
+formulation of the abstract semantics rather than a finite machine
+model.
+
+\paragraph{Abstract compilation}
 

### 2012-11-13 paper/paper.tex
Less detail in intro, more in evaluation. Better table format.

@@
 
-\section{At a Glance}
+\section{At a glance}
 
@@
 environment and control structure arising from the use of higher-order
-functions to model arithmetic.
+functions to encode arithmetic.
 \end{enumerate}
@@
 We start, in section~\ref{sec:aam}, by developing an abstract
-interpreter according to the abstracting abstract machines approach
-and observe the resulting implementation cannot analyze the example in
-under 30 minutes.  In section~\ref{sec:baseline}, we perform a further
-abstraction by widening the store.  The resulting analyzer sacrifices
-precision for speed and is able to analyze the example in about 1
-minute.  Without this abstraction step the analysis is exponential due
-to per-state store variance.  This step is described by Van Horn and
-Might~\cite[\S 3.5--6]{dvanhorn:VanHorn2012Systematic} and is
-necessary to make even small examples feasible.  We therefore take the
-widened interpreter as the baseline for our evaluation.
-
-In section~\ref{sec:opt} we give a series of simple abstractions and
-implementation techniques that, in total, drop the analysis time to
-only 138 milliseconds, a speedup of 475 times.
-Figure~\ref{fig:churchtime} shows the step-wise improvement of the
-analysis time for this example.
+interpreter according to the AAM approach.  Without further
+abstraction, the analysis is exponential due to per-state store
+variance and thus cannot analyze the example in a reasonable amount of
+time.  In section~\ref{sec:baseline}, we perform a further abstraction
+by widening the store.  The resulting analyzer sacrifices precision
+for speed and is able to analyze the example in about 1 minute.  This
+step is described by Van Horn and Might~\cite[\S
+  3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make
+even small examples feasible.  We therefore take the widened
+interpreter as the baseline for our evaluation.
+
+Section~\ref{sec:opt} gives a series of simple abstractions and
+implementation techniques that, in total, speed up the analysis by
+nearly a factor of 500, dropping the analys time to a fraction of a
+second.  Figure~\ref{fig:churchtime} shows the step-wise improvement
+of the analysis time for this example.
 
@@
 
-In section~\ref{sec:impl}, we describe a more complete implementation
-scaled up and capable of analyzing Scheme benchmark programs.
-
-In section~\ref{sec:eval} we evaluate this implementation against a
-set of benchmark programs drawn from the literature.  Many benchmarks
-cause the baseline analyzer to take longer than 30 minutes or to
-consume more 1 gigabit of memory, at which point the analysis is
-stopped.  This is the case for the largest benchmark program, which is
-3,500 lines of code and takes under a minute in the most optimized
-analyzer.  For those benchmarks that did complete on the baseline, the
-optimized analyzer outperformed the baseline by a factor of
+Section~\ref{sec:eval} describes the evaluation of each optimiztion
+technique applied to an implementation supporting a more realistic set
+of features, including mutation, first-class control, compound data, a
+full numeric tower and many more forms of primitive data and
+operations.
+%
+We evaluate this implementation against a set of benchmark programs
+drawn from the literature.
+%
+For all benchmarks, the optimized analyzer outperforms the baseline
+by at least a factor of
 % 475
@@
 
+The benchmarks make use of mutation, first-class control (in the form
+of Scheme's {\tt call/cc} operator), compound data such as lists and
+vectors, many more primitive forms of data and operations such as
+Scheme's numeric tower including exact rationals, complex, and
+floating point numbers.
+
+Many benchmarks cause the baseline analyzer to take longer than 30
+minutes or to consume more 1 gigabit of memory, at which point the
+analysis is stopped.  This is the case for the largest benchmark
+program, which is 3,500 lines of code and takes under a minute in the
+most optimized analyzer.  For those benchmarks that did complete on
+the baseline, the optimized analyzer outperformed the baseline by a
+factor of 475 to 4,382.
+
 \begin{figure}
-\begin{tiny}
-\begin{alltt}
-Program  |  LOC | B. Time | O. Time | B. Space | O. Space | B. Rate | O. Rate
-----------------------------------------------------------------------------
-nucleic  | 3508 |
-matrix   |  747 |
-nbody    | 1434 |
-earley   |  667 |
-maze     |  681 |
-church   |   41 |
-lattice  |  214 |
-boyer    |  642 |
-mbrotz   |   68 |
-\end{alltt}
-\end{tiny}
+\begin{tabular}{@{}l|r|r|r|r|r|r|r@{}}
+Program & LOC
+& \multicolumn{2}{c|}{Time (ms)}
+& \multicolumn{2}{c|}{Space (mb)}
+& \multicolumn{2}{c@{}}{Speed (state/sec)}
+\\
+\hline\hline
+nucleic &  3508 & 0 & 0 & 0 & 0 & 0 & 0\\
+matrix  &  747  & ? & ? & ? & ? & ? & ?\\
+nbody   & 1434  & ? & ? & ? & ? & ? & ?\\
+earley  &  667  & ? & ? & ? & ? & ? & ?\\
+maze    &  681  & ? & ? & ? & ? & ? & ?\\
+church  &   41  & ? & ? & ? & ? & ? & ?\\
+lattice &  214  & ? & ? & ? & ? & ? & ?\\
+boyer   &  642  & ? & ? & ? & ? & ? & ?\\
+mbrotz  &   68  & ? & ? & ? & ? & ? & ?
+\end{tabular}
+\caption{Overview performance comparison between baseline and
+  optimized analyzer.}
+\label{fig:bench-overview}
 \end{figure}
 
-\begin{itemize}
+\begin{enumerate}  %% Maybe use ``dictionary'' style enumeration.
 
@@
 
-\end{itemize}
+\end{enumerate}
 
@@
 \end{figure}
+
+\paragraph{Comparison with other flow analysis implementations}
+
+[Computes results more like Earl, et al.'s 0CFA, which times out on
+  the Church numeral benchmark and so even though it offers a fair
+  point of comparison, a more thorough evaluation is likely to
+  uninformative as the other benchmarks are highly likely to timeout
+  as well (and it would require significant effort to extend their
+  implementation with the features needed to analyze our benchmark
+  suite).  That implementation is evaluated against much smaller
+  benchmarks: the largest program consists of 63 \emph{subexpressions}
+  and times out after 30 minutes.]
+
+Vardoulakis and Shivers evaluate their CFA2
+analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0CFA
+defined in their framework.  [NEED TIMING FOR 0CFA ON CHURCH.]  The
+Church example is the largest benchmark Vardoulakis and Shivers
+consider.  More work would be required to scale the analyzer to the
+set of features required by our benchmarks.
+
+[Should probably talk about mCFA too.]
+
+The only analyzers we were able to find that proved capable of
+analyzing the full suite of benchmarks considered here were the Soft
+Typing system of Wright [CITE] and, in many ways its successor, the
+Polymorphic splitting system of Wright and
+Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
+  is not a coincidence; these papers set a high standard for
+  evaluation, which we consciously aimed to approach.  Many of our
+  benchmarks are drawn from these papers.}  Unfortunately, these
+analyses compute an inherently different and uncomparable form of
+analysis.  Consequently, we have omitted a complete comparison with
+these implementations,.  The AAM approach provides more precision in
+terms of temporal-ordering of program states, which comes at a cost
+that can be avoided in constraint-based approaches.  Consequently
+implementation techniques cannot be ``ported'' between these two
+approaches.  However, our optimized implementation is within an order
+of magnitude of the performance of Wright and Jaganathan's analyzer.
+Although we would like to improve this to be more competitive, the
+optimized AAM approach still has many strengths to recommend it in
+terms of precision, ease of implementation and verification, and rapid
+design.
 
@@
 
-The analysis most similar to that presented in section~\ref{sec:aam}
-is the infinitary control-flow analysis of Nielson and
-Nielson~\cite{dvanhorn:nielson-nielson-popl97} and the unified
-treatment of flow analysis by Jagannathan and
+Besides the original AAM work, the analysis most similar to that
+presented in section~\ref{sec:aam} is the infinitary control-flow
+analysis of Nielson and Nielson~\cite{dvanhorn:nielson-nielson-popl97}
+and the unified treatment of flow analysis by Jagannathan and
 Weeks~\cite{dvanhorn:jagannathan-weeks-popl95}.  Both are
@@
 %% \acks Sam Tobin-Hochstadt for encouragement and feedback -- he was
-%% the first to prompt me to look into how make effective
+%% the first to prompt us to look into how make effective
 %% implementations of the AAM approach.
+
+%% Vincent St Amour for feedback on early drafts.
+%% Greg Morrisset and Matthias Felleisen for discussions.
 

### 2012-11-13 paper/paper.tex
Incorporating Matt's comments.

--- 2012-11-12 paper/paper.tex
@@
 \end{align*}
+$\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
 For the duration of this paper, since our example abstract value space
-has no interesting ordering, $\sqcup$ is a point-wise lifting of
-$\cup$:
-\begin{align*}
-\msto\sqcup \varnothing &= \msto \\
-\msto\sqcup \msto'[\maddr \mapsto S] &= \lambda \maddralt. \left\{
-  \begin{array}{ll}
-    \msto(\maddr)\cup S & \mbox{if } \maddralt = \maddr, \maddr\in dom(\msto) \\
-    S & \mbox{if } \maddralt = \maddr, \maddr\notin dom(\msto) \\
-    (\msto\sqcup\msto')(\maddralt) & \mbox{otherwise}
-  \end{array}\right.
-\end{align*}
-
+has no interesting ordering, $\sqcup$ can be read as $\cup$.
 \paragraph{Primitives}
@@
 
-The uniform $k$-CFA allocation strategy would make the fixpoint of the
-reduction relation in figure \ref{fig:aam} a computable analysis, but
-not an effective one. This is not the strategy that AAM, nor we,
+The uniform $k$-CFA allocation strategy would make $eval$ in figure
+\ref{fig:aam} a computable abstraction of reachable states, but not an
+effective one. This is not the strategy that AAM, nor we,
 recommend. We explain a succession of approximations to reach the
@@
 transition relation until a fixed point in the reachable states is
-reached. We begin with a generic fixpoint combinator, ${\mathcal F}$:
+reached. We begin with Kleene's fixpoint combinator, ${\mathcal F}$,
+presented in the frontier/accumulator style:
 
@@
 
-A common technique to faster convergence in flow analyses is to share
+A common technique to accelerate convergence in flow analyses is to share
 a common store that is monotonically updated as interpretation
 continues. In the ISWIM language with only unary operators, this makes
-our 0-CFA quartic in the size of the program. Not great, but
-polynomial.
+our 0-CFA quartic in the size of the program, which in one step drops the
+complexity from intractable exponentiality to merely daunting
+polynomiality.
 
@@
 
-Let $wn : \widehat{State} \times Store \to State$ be the metafunction
-that adds a store to a storeless state and $c.\msto$ projects the
-store out of a state $c$.
-
-\begin{align*}
-(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
+\begin{align*}
+(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
 \mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
-              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
+              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\} \\
+wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
+wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
+wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont}
 \end{align*}
@@
 
-\subsection{Baseline evaluation}
+\subsection{Store-allocated results}
 \label{sec:baselineeval}
@@
 To achieve this, we instead allocate an address for the value position
-when we create the continuation.  We will refer to this change in the
-semantics as store-allocated results. This address and the tail address are
-both determined by the label of the application point, so the space
-becomes linear and the overall complexity drops to cubic. Note that
-this is a critical abstraction in languages with n-ary functions,
+when we create the continuation.  This address and the tail address
+are both determined by the label of the application point, so the
+space becomes linear and the overall complexity drops to cubic. Note
+that this is a critical abstraction in languages with n-ary functions,
 since otherwise the continuation space becomes ${\mathcal O}(n^n)$!
+This amount of distinction between continuations is unnecessary
+in practice.
 
@@
 results, and thus the precision aligns with an analysis specialized to
-ANF. We can also play a representation trick and remove
-$\menv$ since we know it is the identity function in 0-CFA.
+ANF. We can also play a representation trick in 0-CFA and remove
+$\menv$. In monovariant analyses, variables map to themselves, meaning
+$\menv$ is effectively the identity function. That degeneration, in
+turn, allows us to simply discard it from the analysis.
 
@@
 \begin{align*}
-(cs, \msto) &\widehat{\machstep} (cs\cup cs', \msdiff(\msto)) \\
+(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', \msdiff(\msto)) \\
 \mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\

### 2012-11-13 paper/paper.tex
Some of Matt's comments.  Space savings.

--- 2012-11-13 paper/paper.tex
@@
 analyzer derived under the abstracting abstract machine approach, to
-an efficient program analyzer.
+an efficient program analyzer.  The end result of the process is a two
+to three order-of-magnitude improvement over the systematically
+derived analyzer, making it competitive with hand-optimized
+implementations that compute fundamentally less precise results.
 \end{abstract}
@@
 In this article, we develop a systematic approach to deriving the
-feasible implementation of an abstract machine based analyzer.
+feasible implementation of an abstract-machine-based analyzer.
 
@@
 
-In the following sections, we develop an analyzer for a core
-functional language and follow a series of steps of abstraction and
-optimization.  At each step we evaluate the implementation on an
-example that computes with Church numerals, shown in
-figure~\ref{fig:church}.
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=2.5in]{rel-time-church.ps}
+\end{center}
+\caption{Relative run-time of
+optimized analyzers to baseline for Church numeral example.}
+\label{fig:churchtime}
+\end{figure}
+
+This paper is organized in two halfs: in the first, we start with a
+quick review of the AAM approach to develop an analysis framework and
+then apply our step-by-step optimization techniques in the simplified
+setting of a core functional language.  This allows us to explicate
+the optimizations with a minimal amount of inessential technical
+overhead.  In the second half, we scale this approach up to an
+analyzer for a realistic untyped, higher-order imperative language
+with a number of interesting features.
+
+At each step during the initial development, we evaluate the
+implementation on an example, shown in figure~\ref{fig:church}, taken
+from the flow analysis literature~\cite{dvanhorn:Vardoulakis2011CFA2}
+that tests distributivity of multiplication over addition on Church
+numerals.
 %% \footnote{It is
@@
 \begin{enumerate}
-\item it can be written in most modern programming languages, and
+\item it can be written in most modern programming languages,
 %
-\item it stresses an analyzer's ability to deal with complicated
-environment and control structure arising from the use of higher-order
-functions to encode arithmetic.
+\item it was designed to stress an analyzer's ability to deal with
+  complicated environment and control structure arising from the use
+  of higher-order functions to encode arithmetic, and
+%
+\item it proves to be the \emph{least} improved benchmark of the
+  complete suite considered in section~\ref{sec:eval}, and thus it
+  serves as a good sanity check and lower-bound for each of the
+  optimization techniques considered.
 \end{enumerate}
@@
 of the analysis time for this example.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=2.5in]{rel-time-church.ps}
-\end{center}
-\caption{Relative run-time of
-optimized analyzers to baseline for Church numeral example.}
-\label{fig:churchtime}
-\end{figure}
 
@@
 machine is a very slight variation on a standard abstract machine for
-ISWIM in ``eval, continue, apply'' form.
+ISWIM in ``eval, continue, apply''
+form~\cite{dvanhorn:Danvy:DSc}.\marginpar{There's probably a better
+  (more specific) cite for eval/apply/continue.}  It can be
+systematically derived from a definitional interpreter through a
+continuation-passing style transformation and defunctionalization, or
+from a structural operational semantics using the refocusing
+construction of Danvy and
+Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.
 
@@
 
-\begin{figure*}
+\begin{figure}
+\begin{gather*}
 \begin{align*}
 \mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
+\end{align*}
 \\[2mm]
+\begin{array}{@{}r@{\ }c@{\ }l@{}}
 %% EVAL
-\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
+\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
 \co{\mkont,\mval,\msto}
@@
 \\
-\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
+\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
 \co{\mkont,\mlit,\msto}
 \\
-\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
+\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
 \co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
 \\
-\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
+\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
 \ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
+\\
+&&
 \text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
 \\
-\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
+\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
 \ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
+\\
+&&
 \text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
@@
 %% CONTINUE
-\co{\kmt,\mval,\msto} &\machstep
+\co{\kmt,\mval,\msto} &\machstep&
 \ans{\msto,\mval}
 \\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep&
 \ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
 \\
-\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
+\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep&
 \ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
@@
 \\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep&
 \ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
@@
 \\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep&
 \ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
@@
 %% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
 \ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
-\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
-\\
-\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
+\\
+\multicolumn{3}{r@{}}{
+\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)}
+\\
+\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
 \co{\mkont,\mval',\msto}
 \text{ where } \mval'\in\interpdelta(\mop,\mval)
-\end{align*}
-\caption{Abstract$^2$ machine for ISWIM}
+\end{array}
+\end{gather*}
+\caption{Abstract abstract machine for ISWIM}
 \label{fig:aam}
-\end{figure*}
+\end{figure}
 
@@
 It is important to recognize from this definition that \emph{any}
-allocation strategy is an abstract interpretation.  In particular,
-concrete interpretation is a kind of abstract interpretation.  So is an
+allocation strategy is an abstract
+interpretation~\cite{MIGHTANDMANOLIOS}.  In particular, concrete
+interpretation is a kind of abstract interpretation.  So is an
 interpretation that allocates a single cell into which all bindings
@@
 
-\newpage
+%\newpage
 \section{Baseline analyzer}

### 2012-11-13 paper/paper.tex

@@
 
-%\newpage
-\section{Baseline analyzer}
+\section{Reduction semantics to baseline analyzer}
 \label{sec:baseline}
+
+The uniform $k$-CFA allocation strategy would make the fixpoint of the
+reduction relation in figure \ref{fig:aam} a computable analysis, but
+not an effective one. This is not the strategy that AAM, nor we,
+recommend. We explain a succession of approximations to reach the
+baseline analysis that we compare performance against. But first, we
+need a fixpoint combinator to drive the succession of reduction
+relations we define.
 
@@
 The final approximation we make to get to our baseline analysis is
-store-allocating results of application sub-expressions. Notice that
+store-allocating results of application sub-expressions. The semantics
+of the previous section is where AAM leaves off. However, notice that
 the $\kfn{}$ continuation stores a value - this makes the space of
-continuations quadratic rather than linear in 0-CFA. We instead
-allocate an address for the value position when we create the
-continuation. This address and the tail address are both determined by
-the label of the application point, so the space becomes linear and
-the overall complexity drops to cubic. Note that this is a critical
-abstraction in languages with n-ary functions, since otherwise the
-continuation space becomes ${\mathcal O}(n^n)$!
+continuations quadratic rather than linear in 0-CFA. A linear space
+drops the complexity to cubic.
+
+To achieve this, we instead allocate an address for the value position
+when we create the continuation.  This address and the tail address are
+both determined by the label of the application point, so the space
+becomes linear and the overall complexity drops to cubic. Note that
+this is a critical abstraction in languages with n-ary functions,
+since otherwise the continuation space becomes ${\mathcal O}(n^n)$!
 
@@
 
-We discuss a variety of techniques for increasing performance by
+We discuss a succession of techniques for increasing performance by
 decreasing the interpretive overhead of abstract interpretation. They
 come in two flavors: decreasing the size of the represented state
-space [SS], and decreasing intermediate data structures [DS].
+space [SS], and decreasing the amount of intermediate data structures [DS].
 
 \begin{itemize}
- \item{[SS] quotient by sets of values (lazy non-determinism)}
- \item{[SS] remove {\tt ev} states (abstract compilation)}
- \item{[DS] operate on store changes rather than whole stores (store deltas)}
- \item{[DS] represent the store as a global, mutable hash and
+ \item{[SS] lazy non-determinism: quotient by sets of values}
+ \item{[SS] abstract compilation: eliminate {\tt ev} states}
+ \item{[DS] store deltas: operate on store changes rather than whole stores}
+ \item{[DS] timestamp: represent the store as a global, mutable hash and
    determine a state's store by the number of changes made to the
-   global store at the time of visiting (timestamp)}
- \item{[DS] use a global mutable workset}
- \item{[DS] represent the global store as a global, mutable vector in
-   the 0-CFA case. (preallocation)}
- \item{[SS] coarsely abstract literal data (abstracting compound data)}
+   global store at the time of visiting}
+ \item{[DS] preallocation: represent the global store as a global, mutable vector}
+ \item{[SS] uniform literal approximation: coarsely abstract literal data}
 \end{itemize}
@@
 to the reduction relation/fixpoint, and each improves performance by
-integer factors. The first three remain in the purely functional
-realm. The final three use mutation in small, localized places, so
-they do not increase conceptual complexity drastically.
+integer factors.
 
@@
 
-We can eliminate the {\tt ev} states from the run-time interpretation
-of a program by specializing the machine transition relation to the
-program being analyzed. This eliminates interpretative overhead by
-first compiling the program into ``byte code'' instructions.
+Laziness removes several represented states that would then have to be
+stepped by the fixpoint combinator.  We can further eliminate the {\tt
+  ev} states from the run-time interpretation of a program by
+specializing the machine transition relation to the program being
+analyzed. This precision-preserving transformation eliminates
+interpretative overhead by first compiling the program into ``byte
+code'' instructions.
 
@@
 
-Each step in the fixpointer for the above techniques requires joining
-several, commonly large, stores together. Even with immutable hashes
-that utilize sharing, this is memory-intensive. Worse, this
-contributes an obscene fraction of the runtime. If we instead track
-the changes to the store made over all steps and afterward update the
-store, we end up with much less allocation and computational overhead.
-
-We represent changes as a list of pairs of address to set of storeable
-values. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now becomes
-$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins as empty for
-each step. Applying the changes is straightforward:
+Each step in the fixpoint for the above techniques requires joining
+several, commonly large, stores together. Each step is unlikely to
+always modify all addresses of the store, so this is needlessly
+memory- and time-intensive. We can instead accumulate store changes
+and apply them after all steps have completed. This uses far fewer
+join operations, leading to much less allocation and computational
+overhead.
+
+We represent changes as $Store\msdiff = List[Addr \times
+  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
+becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
+begins as empty for each step. Applying the changes is
+straightforward:
 
@@
 \end{align*}
-
-We also lift $\machstep$ differently to accommodate for this asymmetry
+\begin{align*}
+bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
+\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
+               \maddr &= \mvar\mcntr'
+\end{align*}
+Compilation changes to additionally take a $\msdiff$ component, so the
+above rule's right hand side would instead be $k^{\mcntr'}(\menv',
+\msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.
+
+We also lift $\machstep$ to accommodate for this asymmetry
 in the input and output. For each state that is stepped, we feed the
-output changes to the next so that all changes get accumulated.
+output changes to the next so that all changes get accumulated:
 
@@
 introduce new kinds of labels (e.g. left and right components of a
-tuple).
-
-\subsection{Abstracting Compound Data}
+tuple at the contour of creation).
+
+\subsection{Abstracting literal compound data}
 
@@
 \end{figure}
+
+\paragraph{Mutable literal data} The Scheme we model is nonstandard in that it has immutable literal
+data (the Scheme benchmarks did not mutate their literal data). To
+support mutable literal data, abstract values like {\tt QPair} would
+be like regular conses, only point to a preallocated pair of addresses
+that hold {\tt QData} as well. This is because conses can be mutated
+to hold any value, not just quoted data.
 
@@
 \end{enumerate}
+
+Our implementation of the analysis used in the evaluation is roughly
+3500 lines of Racket code.
+%% 400 lines for the core, 700 lines for
+%% abstraction over optimizations, 1500 lines for primitives and standard
+%% list operations, 700 lines for instantiations to different
+%% optimizations and 300 for parsing and macro transformers.
 

### 2012-11-13 paper/paper.tex
Some changes from Matt's comments on section 5

@@
 \begin{align*}
-(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
+(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
 \mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
@@
 
-For this, we introduce a new value for a delayed lookup that acts to
-quotient the state space by the values at the delayed address. Its
-changes to the semantics are highlighted in the following rules:
-
-\begin{align*}
+For this, we introduce a new kind of value, $\saddr{\maddr}$, for delaying looking up $\maddr$.
+This acts to quotient the state space by the values at the delayed address.
+The following rules highlight the changes to the semantics:
+
+\begin{align*}
+force &: Store \times Addr \to \Set(Value) \\
 force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
@@
 
-\paragraph{Precision loss} This semantics introduces a subtle precision difference over the
-baseline that is negligible. Consider a configuration where a
-reference to a variable and a binding of a variable will happen in one
-step. With laziness, the reference will mean the original binding(s)
-of the variable or the new one, because the actual store lookup is
-delayed precisely one step (i.e. laziness is administrative). Without
-laziness, the reference will fan out to all the bindings of the
-variable before the new binding happens and thus might have an
-observable precision difference. In our experience, this situation
-happens too rarely to warrant the significant increase in time and
-memory needed for this eager non-determinism. Indeed, were the
-variable reference a step later and another binding not made in
-that step, the results of the two approaches are the same.
-
-\paragraph{Avoiding precision loss} The administrative nature of laziness means that we could remove the
+\paragraph{Lose precision; simplify implementation}
+This semantics introduces a subtle precision difference over the
+baseline. Consider a configuration where a reference to a variable and
+a binding of a variable will happen in one step. With laziness, the
+reference will mean the original binding(s) of the variable or the new
+one, because the actual store lookup is delayed precisely one step
+(i.e. laziness is administrative). Without laziness, the reference
+will fan out to all the bindings of the variable before the new
+binding happens and thus might have an observable precision
+difference.
+
+\paragraph{Regain precision; complicate implementation}
+The administrative nature of laziness means that we could remove the
 loss in precision by essentially duplicating the reduction relation to
@@
 
+\paragraph{Our choice: option 1}
+The configurations that lead to precision loss happen too rarely to
+warrant the significant increase in time and memory needed for this
+eager non-determinism. Indeed, were the variable reference a step
+later and another binding not made in that step, the results of the
+two approaches are the same.
+
 % Lazy:  cpu time: 32481 real time: 32881 gc time: 547
@@
 
-Laziness removes several represented states that would then have to be
-stepped by the fixpoint combinator.  We can further eliminate the {\tt
-  ev} states from the run-time interpretation of a program by
-specializing the machine transition relation to the program being
-analyzed. This precision-preserving transformation eliminates
-interpretative overhead by first compiling the program into ``byte
-code'' instructions.
-
-The essence of the compilation effect can be seen by considering an example
-such as
+The laziness optimization saved time by doing the same amount of reasoning as before but in fewer transitions.
+We can exploit the same idea - same reasoning, fewer transitions - with abstract compilation.
+Abstract compilation transforms complex expression whose \emph{abstract} evaluation is deterministic into ``abstract bytecodes.''  The abstract interpreter then does in one transition what previously took many.
+Abstract compilation eliminates unnecessary allocation, deallocation and branching.
+
+The essence of the compilation effect can be seen by considering an
+example such as
 \[
@@
     {\tt ev} states. The next step is to change creation of {\tt ev}
-    states to simply calling these functions. The modified reduction
-    relation is shown in figure \ref{fig:caam}.
+    states to simply calling these functions. Figure \ref{fig:caam}
+    shows the modified reduction relation.
 
@@
 \end{align*}
-\caption{Compilation}
+\caption{Abstract compilation}
 \label{fig:compile}
@@
 
-\subsection{Store deltas}
-
-Each step in the fixpoint for the above techniques requires joining
-several, commonly large, stores together. Each step is unlikely to
-always modify all addresses of the store, so this is needlessly
-memory- and time-intensive. We can instead accumulate store changes
-and apply them after all steps have completed. This uses far fewer
-join operations, leading to much less allocation and computational
+\subsection{Locally log-based store deltas}
+
+Every step the analysis makes for the above techniques requires
+joining several, commonly large, stores together. Each step is
+unlikely to always modify all addresses of the store, so this is
+needlessly memory- and time-intensive. We can instead log store
+changes and replay the change log on the full store after all steps
+have completed. This uses far fewer join operations, leading to less
 overhead.
 
-We represent changes as $Store\msdiff = List[Addr \times
+We represent changes as $\msdiff \in Store\msdiff = List[Addr \times
   \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
 becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
-begins as empty for each step. Applying the changes is
-straightforward:
+begins empty ($\mtlst$) for each step. Applying the changes to the
+full store is straightforward - we interpret $\msdiff$ as a $Store \to
+Store$ function:
 
@@
 
-%% Delta Store + Compile + Lazy:
+Here $concat : \Set(Store\msdiff) \to Store\msdiff$ flattens the lists
+of changes to one; order it appends does not matter:
+\begin{align*}
+concat(\varnothing) &= \mtlst \\
+concat(\{\msdiff\}\cup D) &= append(\msdiff, concat(D))
+\end{align*}
+
+%% Delta Store +k Compile + Lazy:
 %%    cpu time: 668 real time: 686 gc time: 41
@@
 
-The store delta technique removes overhead of joining large stores,
-but for determining if a state has been visited before (vis. $f(X)
-\setminus S$ in ${\mathcal G}$), we still have to compare entire
-stores - an expensive computation.  We modify a technique due to
-Shivers~\cite{ianjohnson:Shivers:1991:CFA}, timestamp approximation,
-that removes the need for joining whole stores by manipulating a
-global store imperatively. The original technique loses precision
-because states are stepped in sequence, modifying the store in place -
-more values or continuations can be added before another state gets to
-step. This interference means more possible flows (i.e. less
-precision) and a nondeterministic analysis - the order the set of states is
-traversed is not necessarily deterministic. However, by delaying the
-store changes until after all steps have completed, we can maintain
-our precision and determinism. In a variety of experiments, we
-observed that the time and space overhead for storing store-deltas was outweighed by
-the precision keeping flows in check. That is to say, the analysis was
-faster {\it and} more precise.
-
-We pair this technique with a mutable workset to avoid the
-intermediate sets introduced by the $State \times Store \to \Set(State) \times Store\msdiff$ signature
-of $\machstep$. This necessitates a change to the fixpoint algorithm to use a global
-workset, {\tt todo}, that $\machstep$ imperatively updates with the
-states that it would have otherwise accumulated in a set that would later be added to the workset.
-
-In total, we have six globals:
+Thus far, we have made our optimizations in a purely functional
+manner. For the next series of optimizations, we need to dip into the
+imperative. We can motivate this entire sequence of optimizations by
+focusing on the largest bottleneck in the current state-space
+exploration algorithm: checking to see if a state has already been
+seen. Given two states, checking equality is expensive because the
+stores within each are large, and every entry must be checked against
+every other. Hashes can sometimes rule out inequality relatively
+quickly, but the incidence of collisions and actual equality is
+costly.
+
+And, there is a better way. Shivers' original work on k-CFA was
+susceptible to the same problem, and he suggested three complementary
+optimizations: (1) make the store global; (2) update the store
+imperatively; and (3) associate every change in the store with a
+version number -- its timestamp. Then, put timestamps in states
+where previously there were stores. Given two states, the analysis can
+now compare their stores just by comparing their timestamps -- a
+constant-time operation.
+
+There is a subtle loss of precision and Shivers' original time-stamp
+technique that we can fix. For a given abstract state, all writes to
+the global store need to be delayed until the analysis considers all
+branches from that abstract state. This avoids cross-branch pollution
+which would not otherwise happen, e.g., when one branch writes to
+address $\maddr$ and another branch reads from address
+$\maddr$. Fortunately, given our conversion to log-based stores this
+change is straightforward.
+
+The current state of the state-space search now has six components:
+
 \begin{itemize}

### 2012-11-13 paper/paper.tex

@@
 
-Let $wn : \widehat{State} \times Store \to State$ be the metafunction
-that adds a store to a storeless state and $c.\msto$ projects the
-store out of a state $c$.
-
 \begin{align*}
@@
 \mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
-              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
+              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\} \\
+wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
+wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
+wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont}
 \end{align*}
@@
 To achieve this, we instead allocate an address for the value position
-when we create the continuation.  This address and the tail address are
-both determined by the label of the application point, so the space
-becomes linear and the overall complexity drops to cubic. Note that
-this is a critical abstraction in languages with n-ary functions,
+when we create the continuation.  This address and the tail address
+are both determined by the label of the application point, so the
+space becomes linear and the overall complexity drops to cubic. Note
+that this is a critical abstraction in languages with n-ary functions,
 since otherwise the continuation space becomes ${\mathcal O}(n^n)$!
+This amount of distinction between continuations is unnecessary
+in practice.
 

### 2012-11-13 paper/paper.tex
Rewording in literal data abstraction.

@@
 The $\machstep$ relation is most naturally written as a function from
-a state to a set of all next states. We want to find a fixpoint of
-$\machstep$ on the state space, however, so we must lift it to work on
-sets of states. The lifting is a natural one:
+a state to a set of all next states. To calculate $eval$, we find a fixed point of
+$\machstep$ lifted to step a set of states. The lifting is a natural one:
 
@@
   cons}es. In particular, if a list literal is ``too big'' (in our
-case length $>$ 1), we interpret the list as a circular data structure;
-the right address points back to the {\tt cons} itself, and the left
-address points to all of the elements of the list. We take this
-farther and join the list elements together in with a coarse
-type-based abstraction. Quotation is special because it cannot
-introduce function values, which is important to enhancing our
-technique's soundness to conceptual complexity ratio. If we join two
-values of different type together, we don't get ``anything,'' which
-has complex meaning in higher-order languages (Shivers' escape
-semantics) and is overly approximate. We instead get ``any quotable
-value,'' which has much simpler semantics.
+case length $>$ 1), we interpret the list as a circular data
+structure; the right address points back to the {\tt cons} itself, and
+the left address points to all of the elements of the list. We take
+this farther and join the list elements together in with a coarse
+type-based abstraction. In effect, large lists or vectors of literal
+numbers become unbounded lists/vectors of ``number.'' Heterogeneous
+data is combined to just ``data,'' rather than a union of ``number'',
+``string'', etc. since primitives are interpreted on every combination
+of values that flow to them, and unions lead to more primitive
+interpretations.
+
+Quotation is special because it cannot introduce function values,
+which is important to enhancing our technique's soundness to
+conceptual complexity ratio. If we join two values of different type
+together, we don't get ``anything,'' which has complex meaning in
+higher-order languages (Shivers' escape semantics) and is overly
+approximate. We instead get ``any quotable value,'' which has much
+simpler semantics.
 
@@
 parameterized in such a way that in the limit, the analysis is
-equivalent to an interreter for the language, just as is the case
+equivalent to an interpreter for the language, just as is the case
 here.  What is different is that both give a constraint-based

### 2012-11-13 paper/paper.tex
Figure showing state graphs under base, lazy, compile.

--- 2012-11-13 paper/paper.tex
@@
 \usepackage{balance}
+\usepackage{calc}
 
@@
 \begin{figure}
+\hrule
+\vspace{2mm}
 \begin{alltt}
@@
 
+\begin{figure}
+\hrule
+\vspace{2mm}
+\begin{alltt}
+(define (id x) x)
+(define (f n)
+  (cond [(<= n 1)  1]
+        [else      (* n (f (- n 1)))]))
+(define (g n)
+  (cond [(<= n 1)  1]
+        [else      (+ (* n n) (g (- n 1)))]))
+(print (+ ((id f) 3) ((id g) 4)))
+\end{alltt}
+\hrule
+\begin{center}
+\begin{tabular}{ccc}
+\raisebox{1ex-\height}{
+\includegraphics[height=4in]{introspective-base.pdf}}
+&
+\raisebox{1ex-\height}{
+\includegraphics[height=3.7in]{introspective-lazy.pdf}}
+&
+\raisebox{1ex-\height}{
+\includegraphics[height=2.2in]{introspective-lazyc.pdf}}
+\\
+(a) Baseline
+&
+(b) Lazy
+&
+(c) Compiled
+\end{tabular}
+\end{center}
+\caption{Example state graphs for the program above.  Part (a) shows
+  the result of the baseline analyzer.  It has long ``corridor''
+  transitions and ``diamond'' subgraphs that fan-out from
+  nondeterminism and fan-in from joins.  Part (b) shows the result of
+  performing nondeterminism lazily and thus avoids many of the diamond
+  subgraphs.  Part (c) shows the result of abstract compilation that
+  removes intepretive overhead in the form of intermediate states,
+  thus minimizing the corridor transitions.  The end result is a more
+  compact abstraction of the program that can be generated faster.}
+\end{figure}
+
+\newpage
+\
+\newpage
 \section{An abstract$^{\text 2}$ machine for ISWIM}

### 2012-11-13 paper/paper.tex
Compact state graph figure.

@@
 (define (f n)
-  (cond [(<= n 1)  1]
-        [else      (* n (f (- n 1)))]))
+  (if (<= n 1) 1 (* n (f (- n 1)))))
 (define (g n)
-  (cond [(<= n 1)  1]
-        [else      (+ (* n n) (g (- n 1)))]))
+  (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
 (print (+ ((id f) 3) ((id g) 4)))
@@
 &
-(c) Compiled
+(c) Compiled (\& lazy)
 \end{tabular}
@@
 
-\newpage
-\
-\newpage
 \section{An abstract$^{\text 2}$ machine for ISWIM}
@@
 ISWIM in ``eval, continue, apply''
-form~\cite{dvanhorn:Danvy:DSc}.\marginpar{There's probably a better
-  (more specific) cite for eval/apply/continue.}  It can be
+form~\cite{dvanhorn:Danvy:DSc}.  It can be
 systematically derived from a definitional interpreter through a

### 2012-11-13 paper/paper.tex
More work on Glance section.

@@
 
-Section~\ref{sec:eval} describes the evaluation of each optimiztion
-technique applied to an implementation supporting a more realistic set
-of features, including mutation, first-class control, compound data, a
-full numeric tower and many more forms of primitive data and
-operations.
-%
-We evaluate this implementation against a set of benchmark programs
-drawn from the literature.
-%
-For all benchmarks, the optimized analyzer outperforms the baseline
-by at least a factor of
-% 475
-two to
-% 4,382
-three orders of magnitude.
-
 \newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
-
 \begin{figure}
@@
 
-Section~\ref{sec:related} relates this work to the literature and
-section~\ref{sec:conclusion} concludes.
-
-\begin{figure}
+\begin{figure}[t]
 \hrule
@@
   compact abstraction of the program that can be generated faster.}
+\label{fig:state-graphs}
 \end{figure}
 
+The AAM approach, in essence, does the following: it takes a
+machine-based view of computation and turns it into a \emph{finitary
+  approximation} by bounding the size of the store.  With a limited
+address space, the store must map addresses to \emph{sets} of values.
+Store updates are interpreted as joins, and store dereferences are
+interpreted by non-deterministic choice of an element from a set.  The
+result of analyzing a program is a finite directed graph where nodes
+in the graph are (abstract) machine states and edges denote machine
+transitions between states.
+
+The techniques we propose for optimizing analysis fall into the
+following categories:
+\begin{enumerate}
+\item generate fewer states by avoiding the eager exploration of
+  non-deterministic choices that will later collapse into a single
+  join point.  We accompish this by applying lazy evaluation
+  techniques so that non-determinism is evaluated \emph{by need}.
+
+\item generate fewer states by avoiding unnecessary, intermediate
+  states of a computation.  We accomplish this by applying compilation
+  techniques from functional languages to avoid interpretive overhead
+  in the machine transition system.
+
+\item generate states faster.  We accomplish this by better algorithm
+  design in the fixpoint computation we use to generate state graphs.
+\end{enumerate}
+Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
+small example due to Earl, et
+al.~\cite{dvanhorn:Earl2012Introspective}.
+By generating significantly fewer states at a significantly faster
+rate, we are able to achieve large performance improvements in terms
+of both time and space.
+
+Section~\ref{sec:eval} describes the evaluation of each optimiztion
+technique applied to an implementation supporting a more realistic set
+of features, including mutation, first-class control, compound data, a
+full numeric tower and many more forms of primitive data and
+operations.
+%
+We evaluate this implementation against a set of benchmark programs
+drawn from the literature.
+%
+For all benchmarks, the optimized analyzer outperforms the baseline
+by at least a factor of
+% 475
+two to
+% 4,382
+three orders of magnitude.
+
+Section~\ref{sec:related} relates this work to the literature and
+section~\ref{sec:conclusion} concludes.
+
+\newpage
 \section{An abstract$^{\text 2}$ machine for ISWIM}

### 2012-11-13 paper/paper.tex
Made explicit that laziness does not change the concrete semantics, took a stab at recategorizing the techniques.

--- 2012-11-13 paper/paper.tex
@@
 \usepackage{balance}
-\usepackage{calc}
 
@@
 \begin{figure}
-\hrule
-\vspace{2mm}
 \begin{alltt}
@@
 
-\begin{figure}
-\hrule
-\vspace{2mm}
-\begin{alltt}
-(define (id x) x)
-(define (f n)
-  (cond [(<= n 1)  1]
-        [else      (* n (f (- n 1)))]))
-(define (g n)
-  (cond [(<= n 1)  1]
-        [else      (+ (* n n) (g (- n 1)))]))
-(print (+ ((id f) 3) ((id g) 4)))
-\end{alltt}
-\hrule
-\begin{center}
-\begin{tabular}{ccc}
-\raisebox{1ex-\height}{
-\includegraphics[height=4in]{introspective-base.pdf}}
-&
-\raisebox{1ex-\height}{
-\includegraphics[height=3.7in]{introspective-lazy.pdf}}
-&
-\raisebox{1ex-\height}{
-\includegraphics[height=2.2in]{introspective-lazyc.pdf}}
-\\
-(a) Baseline
-&
-(b) Lazy
-&
-(c) Compiled
-\end{tabular}
-\end{center}
-\caption{Example state graphs for the program above.  Part (a) shows
-  the result of the baseline analyzer.  It has long ``corridor''
-  transitions and ``diamond'' subgraphs that fan-out from
-  nondeterminism and fan-in from joins.  Part (b) shows the result of
-  performing nondeterminism lazily and thus avoids many of the diamond
-  subgraphs.  Part (c) shows the result of abstract compilation that
-  removes intepretive overhead in the form of intermediate states,
-  thus minimizing the corridor transitions.  The end result is a more
-  compact abstraction of the program that can be generated faster.}
-\end{figure}
-
-\newpage
-\
-\newpage
 \section{An abstract$^{\text 2}$ machine for ISWIM}
@@
 baseline analysis that we compare performance against. But first, we
-need a fixpoint combinator to drive the succession of reduction
+need a fixed-point combinator to drive the succession of reduction
 relations we define.
 
-\subsection{Generic fixpoint calculator}
+\subsection{Generic fixed-point calculator}
 \label{sec:fixpoint}
@@
 We compute the semantics of a program by iterating the state
-transition relation until a fixed point in the reachable states is
-reached. We begin with Kleene's fixpoint combinator, ${\mathcal F}$,
-presented in the frontier/accumulator style:
+transition relation until a fixed-point in the reachable states is
+reached. We begin with Kleene's fixed-point combinator, ${\mathcal F}$,
+operationalized in the frontier/accumulator style:
 
@@
 The $\machstep$ relation is most naturally written as a function from
-a state to a set of all next states. To calculate $eval$, we find a fixed point of
+a state to a set of all next states. To calculate $eval$, we find a fixed-point of
 $\machstep$ lifted to step a set of states. The lifting is a natural one:
@@
 (cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
-\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
-              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\} \\
+\mbox{where } cs' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in cs\} \\
+              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in cs\} \\
 wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
 wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
-wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont}
+wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
+wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
 \end{align*}
@@
 $\menv$ is effectively the identity function. That degeneration, in
-turn, allows us to simply discard it from the analysis.
+turn, allows us to discard it from the analysis.
 
@@
 decreasing the interpretive overhead of abstract interpretation. They
-come in two flavors: decreasing the size of the represented state
-space [SS], and decreasing the amount of intermediate data structures [DS].
-
-\begin{itemize}
- \item{[SS] lazy non-determinism: quotient by sets of values}
- \item{[SS] abstract compilation: eliminate {\tt ev} states}
- \item{[DS] store deltas: operate on store changes rather than whole stores}
- \item{[DS] timestamp: represent the store as a global, mutable hash and
-   determine a state's store by the number of changes made to the
-   global store at the time of visiting}
- \item{[DS] preallocation: represent the global store as a global, mutable vector}
- \item{[SS] uniform literal approximation: coarsely abstract literal data}
-\end{itemize}
+come in two flavors: decreasing the size of the state space with domain-specific knowledge [DSK], and
+decreasing the amount of intermediate data structures [Eng](ineering).
+
+\begin{enumerate}
+ \item{[DSK] lazy non-determinism: quotient by sets of values}
+ \item{[DSK] abstract compilation: eliminate {\tt ev} states}
+ \item{[Eng] store deltas: operate on store changes rather than whole stores}
+ \item{[DSK/Eng] timestamp: represent the store as a global, mutable
+   hash and determine a state's store by the number of changes made to
+   the global store at the time of visiting}
+ \item{[Eng] preallocation: represent the global store as a global, mutable vector}
+ \item{[DSK] uniform literal approximation: coarsely abstract literal data}
+\end{enumerate}
 
@@
 precision-lossy techniques.  Each technique requires minimal changes
-to the reduction relation/fixpoint, and each improves performance by
+to the reduction relation, and each improves performance by
 integer factors.
@@
 where all variables have been joined with multiple values, each
-variable reference transitions to several new states that comprise the
-cartesian product of all the values of {\tt f}, {\tt x} and {\tt y} (or in the our baseline analysis, the sum),
-only to be joined back together in their respective application positions. We would
-see transitions of the following shape:
+variable reference transitions to as many new states as there are
+values, only to be joined back together in their respective
+application positions. We would see transitions of the following
+shape:
 \begin{center}
@@
 
-For this, we introduce a new kind of value, $\saddr{\maddr}$, for delaying looking up $\maddr$.
-This acts to quotient the state space by the values at the delayed address.
-The following rules highlight the changes to the semantics:
+This does not change the concrete semantics of the language. It is an
+abstraction of the otherwise several states the original
+non-deterministic semantics steps to. We say the abstraction is
+\emph{lazy} because it delays dereferencing an address until its
+contents are \emph{needed} as values in the semantics. It does not
+change the execution order that leads to the values that are stored in
+the address.
+
+We introduce a new kind of value, $\saddr{\maddr}$, that represents a
+delayed lookup of $\maddr$. This acts to quotient the state space by
+the values at the delayed address. The following rules highlight the
+changes to the semantics:
 
@@
 reference will mean the original binding(s) of the variable or the new
-one, because the actual store lookup is delayed precisely one step
+one, because the actual store lookup is delayed one step
 (i.e. laziness is administrative). Without laziness, the reference
@@
 The administrative nature of laziness means that we could remove the
-loss in precision by essentially duplicating the reduction relation to
-specialize variable lookup. This works since in the semantics of ISWIM
-with store-allocated results consumes the value component of states in
-one step. This is not the case for semantics that replicate the value
+loss in precision by duplicating the reduction relation to specialize
+variable lookup. This works since in the semantics of ISWIM with
+store-allocated results consumes the value component of states in one
+step. This is not the case for semantics that replicate the value
 component across reductions, say for popping off exception handler
@@
 
-The laziness optimization saved time by doing the same amount of reasoning as before but in fewer transitions.
-We can exploit the same idea - same reasoning, fewer transitions - with abstract compilation.
-Abstract compilation transforms complex expression whose \emph{abstract} evaluation is deterministic into ``abstract bytecodes.''  The abstract interpreter then does in one transition what previously took many.
-Abstract compilation eliminates unnecessary allocation, deallocation and branching.
+The laziness optimization saved time by doing the same amount of
+reasoning as before but in fewer transitions. We can exploit the same
+idea - same reasoning, fewer transitions - with abstract
+compilation. Abstract compilation is \emph{precision-preserving} and
+transforms complex expressions whose \emph{abstract} evaluation is
+deterministic into ``abstract bytecodes.''  The abstract interpreter
+then does in one transition what previously took many. Abstract
+compilation eliminates unnecessary allocation, deallocation and
+branching.
 
@@
 \ref{fig:compile} shows close similarity to the rules for interpreting
-    {\tt ev} states. The next step is to change creation of {\tt ev}
-    states to simply calling these functions. Figure \ref{fig:caam}
-    shows the modified reduction relation.
+    {\tt ev} states. The next step is to change reduction rules that
+    create {\tt ev} states to instead call these functions. Figure
+    \ref{fig:caam} shows the modified reduction relation.
 
@@
 Every step the analysis makes for the above techniques requires
-joining several, commonly large, stores together. Each step is
-unlikely to always modify all addresses of the store, so this is
-needlessly memory- and time-intensive. We can instead log store
-changes and replay the change log on the full store after all steps
-have completed. This uses far fewer join operations, leading to less
-overhead.
+joining large stores together. Not every step will modify all
+addresses of the store, so joining entire stores is wasteful in terms
+of memory and time. We can instead log store changes and replay the
+change log on the full store after all steps have completed. This uses
+far fewer join operations, leading to less overhead, and is \emph{precision preserving}.
 
 We represent changes as $\msdiff \in Store\msdiff = List[Addr \times
-  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ now
-becomes $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$
-begins empty ($\mtlst$) for each step. Applying the changes to the
-full store is straightforward - we interpret $\msdiff$ as a $Store \to
-Store$ function:
-
-\begin{align*}
-\mtlst(\msto) &= \msto \\
-\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}(\msto) &= \msdiff(\msto)\sqcup[\maddr \mapsto \mval{s}]
+  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
+instead becomes a log addition
+$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
+empty ($\mtlst$) for each step. Applying the changes to the full store
+is straightforward:
+
+\begin{align*}
+replay &: Store\msdiff \times Store \to Store \\
+replay(\mtlst, \msto) &= \msto \\
+replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
 \end{align*}
@@
 \begin{align*}
-(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', \msdiff(\msto)) \\
+(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', replay(\msdiff,\msto)) \\
 \mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
@@
 
-The new fixpoint calculation is defined in
+The new $eval$ calculation is defined in
 figure~\ref{fig:imperative}.  To ensure termination, we guard adding
@@
 
-After all steps complete, we apply $\msdiff$ to $\msto$ imperatively and increase
-{\tt T} as long as there was at least one change in $\msdiff$. This
-logic leads to termination if we know that each $\ttuple{\maddr}{\mval{s}}$
-in $\msdiff$ would change the value of $\maddr$ in the current
-store. Thus, we also guard additions to $\msdiff$ so that only updates
-that would change the store are permitted. Each time $\msdiff$ is
-successfully extended, we set {\tt $\msdiff$?} to true. Before each
-individual step, we set it to false.
+After all steps complete, we apply $\msdiff$ to $\msto$ imperatively
+(with {\tt replay!}) and increase {\tt T} as long as there was at
+least one change in $\msdiff$. This logic leads to termination if we
+know that each $\ttuple{\maddr}{\mval{s}}$ in $\msdiff$ would change
+the value of $\maddr$ in the current store. Thus, we also guard
+additions to $\msdiff$ so that only updates that would change the
+store are permitted. Each time $\msdiff$ is successfully extended, we
+set {\tt $\msdiff$?} to true. Before each individual step, we set it
+to false.
 
@@
       foreach c \(\in\) old: \(\msdiff\)? := false; c()
-      unless \(\msdiff\) = \(\mtlst\): T += 1; \(\msdiff(\msto)\)
+      unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
 \end{alltt}
-\caption{Imperative fixpoint algorithm}
+\caption{Imperative algorithm}
 \label{fig:imperative}
@@
 
-This technique by itself only works in 0-CFA, since there are many more
-addresses that can be allocated in 1-CFA and higher. To support
+This technique by itself only works in 0-CFA, since there are many
+more addresses that can be allocated in 1-CFA and higher. To support
 general $k$-CFA in a vector representation, we use a growable vector
@@
 up-front and addressing by the contour converted to polynomial, since
-many combinations of labels are not likely to occur. The all up-front
-allocation again breaks down when we add compound data, which
-introduce new kinds of labels (e.g. left and right components of a
-tuple at the contour of creation).
+a majority of combinations of labels will not occur, except in
+pathological programs. The all up-front allocation again breaks down
+when we add compound data, which introduce new kinds of labels
+(e.g. left and right components of a tuple at the contour of
+creation).
 
@@
 the Boyer benchmark is that it has large list literals, the exact
-values of which do not particularly matter for many analyses. If we
+values of which do not matter for many analyses. If we
 interpret a list literal of 30 elements (that again have 10 elements
@@
 The number of syntactic uses of {\tt cons} versus implicit uses via
-literal lists is dramatically smaller in typical Scheme programs. We
-use the above abstraction for these syntactic uses, but choose to
-interpret literal lists as not always sugar for cascading {\tt
-  cons}es. In particular, if a list literal is ``too big'' (in our
-case length $>$ 1), we interpret the list as a circular data
-structure; the right address points back to the {\tt cons} itself, and
-the left address points to all of the elements of the list. We take
-this farther and join the list elements together in with a coarse
-type-based abstraction. In effect, large lists or vectors of literal
-numbers become unbounded lists/vectors of ``number.'' Heterogeneous
-data is combined to just ``data,'' rather than a union of ``number'',
-``string'', etc. since primitives are interpreted on every combination
-of values that flow to them, and unions lead to more primitive
-interpretations.
+literal lists is smaller in typical Scheme programs. We use the above
+abstraction for these syntactic uses, but choose to interpret literal
+lists as not always sugar for cascading {\tt cons}es. In particular,
+if a list literal is ``too big'' (in our case length $>$ 1), we
+interpret the list as a circular data structure; the right address
+points back to the {\tt cons} itself, and the left address points to
+all of the elements of the list. We take this farther and join the
+list elements together in with a coarse type-based abstraction. In
+effect, large lists or vectors of literal numbers become unbounded
+lists/vectors of ``number.'' Heterogeneous data is combined to just
+``data,'' rather than a union of ``number'', ``string'', etc. since
+primitives are interpreted on every combination of values that flow to
+them, and unions lead to more primitive interpretations.
 
@@
 
-\paragraph{Mutable literal data} The Scheme we model is nonstandard in that it has immutable literal
+\paragraph{Mutable literal data}
+The Scheme we model is nonstandard in that it has immutable literal
 data (the Scheme benchmarks did not mutate their literal data). To
@@
 Many benchmarks cause the baseline analyzer to take longer than 30
-minutes or to consume more 1 gigabit of memory, at which point the
+minutes or to consume more 1 gigabyte of memory, at which point the
 analysis is stopped.  This is the case for the largest benchmark

### 2012-11-13 paper/paper.tex

@@
 
+\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
+\begin{figure}
+\hrule
+\vspace{2mm}
+\begin{alltt}
+(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
+(define (((mult m1) m2) mf) (m2 (m1 mf)))
+(define ((((plus p1) p2) pf) x)
+  ((p1 pf) ((p2 pf) x)))
+
+(define (((pred n) rf) rx)
+  (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
+    (\(\lambda\) (ignored) rx))
+   (\(\lambda\) (id) id)))
+
+(define ((church=? e1) e2)
+  (if (zero? e1) (zero? e2)
+      (if (zero? e2) #f
+          ((church=? (pred e1)) (pred e2)))))
+
+;; multiplication distributes over addition
+((church=? ((mult \church2) ((plus \church1) \church3)))
+ ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
+\end{alltt}
+\caption{Church-numeral calculation}
+\label{fig:church}
+\end{figure}
+
+\begin{figure}[t]
+\hrule
+\vspace{2mm}
+\begin{alltt}
+(define (id x) x)
+(define (f n)
+  (if (<= n 1) 1 (* n (f (- n 1)))))
+(define (g n)
+  (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
+(print (+ ((id f) 3) ((id g) 4)))
+\end{alltt}
+\hrule
+\begin{center}
+\begin{tabular}{ccc}
+\raisebox{1ex-\height}{
+\includegraphics[height=4in]{introspective-base.pdf}}
+&
+\raisebox{1ex-\height}{
+\includegraphics[height=3.7in]{introspective-lazy.pdf}}
+&
+\raisebox{1ex-\height}{
+\includegraphics[height=2.2in]{introspective-lazyc.pdf}}
+\\
+(a) Baseline
+&
+(b) Lazy
+&
+(c) Compiled (\& lazy)
+\end{tabular}
+\end{center}
+\caption{Example state graphs for the program above.  Part (a) shows
+  the result of the baseline analyzer.  It has long ``corridor''
+  transitions and ``diamond'' subgraphs that fan-out from
+  nondeterminism and fan-in from joins.  Part (b) shows the result of
+  performing nondeterminism lazily and thus avoids many of the diamond
+  subgraphs.  Part (c) shows the result of abstract compilation that
+  removes intepretive overhead in the form of intermediate states,
+  thus minimizing the corridor transitions.  The end result is a more
+  compact abstraction of the program that can be generated faster.}
+\label{fig:state-graphs}
+\end{figure}
+
+The AAM approach, in essence, does the following: it takes a
+machine-based view of computation and turns it into a \emph{finitary
+  approximation} by bounding the size of the store.  With a limited
+address space, the store must map addresses to \emph{sets} of values.
+Store updates are interpreted as joins, and store dereferences are
+interpreted by non-deterministic choice of an element from a set.  The
+result of analyzing a program is a finite directed graph where nodes
+in the graph are (abstract) machine states and edges denote machine
+transitions between states.
+
+The techniques we propose for optimizing analysis fall into the
+following categories:
+\begin{enumerate}
+\item generate fewer states by avoiding the eager exploration of
+  non-deterministic choices that will later collapse into a single
+  join point.  We accompish this by applying lazy evaluation
+  techniques so that non-determinism is evaluated \emph{by need}.
+
+\item generate fewer states by avoiding unnecessary, intermediate
+  states of a computation.  We accomplish this by applying compilation
+  techniques from functional languages to avoid interpretive overhead
+  in the machine transition system.
+
+\item generate states faster.  We accomplish this by better algorithm
+  design in the fixpoint computation we use to generate state graphs.
+\end{enumerate}
+Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
+small example due to Earl, et
+al.~\cite{dvanhorn:Earl2012Introspective}.
+By generating significantly fewer states at a significantly faster
+rate, we are able to achieve large performance improvements in terms
+of both time and space.
+
 Section~\ref{sec:eval} describes the evaluation of each optimiztion
@@
 
-\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
-
-\begin{figure}
-\begin{alltt}
-(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
-(define (((mult m1) m2) mf) (m2 (m1 mf)))
-(define ((((plus p1) p2) pf) x)
-  ((p1 pf) ((p2 pf) x)))
-
-(define (((pred n) rf) rx)
-  (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
-    (\(\lambda\) (ignored) rx))
-   (\(\lambda\) (id) id)))
-
-(define ((church=? e1) e2)
-  (if (zero? e1) (zero? e2)
-      (if (zero? e2) #f
-          ((church=? (pred e1)) (pred e2)))))
-
-;; multiplication distributes over addition
-((church=? ((mult \church2) ((plus \church1) \church3)))
- ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
-\end{alltt}
-\caption{Church-numeral calculation}
-\label{fig:church}
-\end{figure}
-
 Section~\ref{sec:related} relates this work to the literature and
@@
 
+\newpage
 \section{An abstract$^{\text 2}$ machine for ISWIM}

### 2012-11-14 paper/paper.tex
Fix up figure for labelless nodes in state graphs.
Trying to get state graphs drawn closer to scale.
Cut code, present VS benchmark.

@@
 \begin{center}
-\includegraphics[width=2.5in]{rel-time-church.ps}
+\includegraphics[width=3.2in]{church-relative-space.ps}
+\includegraphics[width=3.2in]{church-relative-speed.ps}
+\includegraphics[width=3.2in]{church-relative-time.ps}
+\vspace{-1.5em}
 \end{center}
-\caption{Relative run-time of
-optimized analyzers to baseline for Church numeral example.}
+\caption{Relative improvements over the baseline analyzer for the
+  Vardoulakis and Shivers benchmark in terms of peak memory usage, the
+  rate of state transitions, and total analysis time.}
 \label{fig:churchtime}
@@
 
-\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
-\begin{figure}
-\hrule
-\vspace{2mm}
-\begin{alltt}
-(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
-(define (((mult m1) m2) mf) (m2 (m1 mf)))
-(define ((((plus p1) p2) pf) x)
-  ((p1 pf) ((p2 pf) x)))
-
-(define (((pred n) rf) rx)
-  (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
-    (\(\lambda\) (ignored) rx))
-   (\(\lambda\) (id) id)))
-
-(define ((church=? e1) e2)
-  (if (zero? e1) (zero? e2)
-      (if (zero? e2) #f
-          ((church=? (pred e1)) (pred e2)))))
-
-;; multiplication distributes over addition
-((church=? ((mult \church2) ((plus \church1) \church3)))
- ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
-\end{alltt}
-\caption{Church-numeral calculation}
-\label{fig:church}
-\end{figure}
+%% \newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
+%% \begin{figure}
+%% \hrule
+%% \vspace{2mm}
+%% \begin{alltt}
+%% (define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
+%% (define (((mult m1) m2) mf) (m2 (m1 mf)))
+%% (define ((((plus p1) p2) pf) x)
+%%   ((p1 pf) ((p2 pf) x)))
+
+%% (define (((pred n) rf) rx)
+%%   (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
+%%     (\(\lambda\) (ignored) rx))
+%%    (\(\lambda\) (id) id)))
+
+%% (define ((church=? e1) e2)
+%%   (if (zero? e1) (zero? e2)
+%%       (if (zero? e2) #f
+%%           ((church=? (pred e1)) (pred e2)))))
+
+%% ;; multiplication distributes over addition
+%% ((church=? ((mult \church2) ((plus \church1) \church3)))
+%%  ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
+%% \end{alltt}
+%% \caption{Church-numeral calculation}
+%% \label{fig:church}
+%% \end{figure}
 
 \begin{figure}[t]
-\hrule
-\vspace{2mm}
-\begin{alltt}
-(define (id x) x)
-(define (f n)
-  (if (<= n 1) 1 (* n (f (- n 1)))))
-(define (g n)
-  (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
-(print (+ ((id f) 3) ((id g) 4)))
-\end{alltt}
-\hrule
+%% \hrule
+%% \vspace{2mm}
+%% \begin{alltt}
+%% (define (id x) x)
+%% (define (f n)
+%%   (if (<= n 1) 1 (* n (f (- n 1)))))
+%% (define (g n)
+%%   (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
+%% (print (+ ((id f) 3) ((id g) 4)))
+%% \end{alltt}
+%% \hrule
 \begin{center}
@@
 \raisebox{1ex-\height}{
-\includegraphics[height=4in]{introspective-base.pdf}}
+\includegraphics[height=3.5in]{introspective-base.pdf}}
 &
 \raisebox{1ex-\height}{
-\includegraphics[height=3.7in]{introspective-lazy.pdf}}
+\includegraphics[height=3.5in]{introspective-lazy.pdf}}
 &
 \raisebox{1ex-\height}{
-\includegraphics[height=2.2in]{introspective-lazyc.pdf}}
+\includegraphics[height=3in]{introspective-lazyc.pdf}}
 \\
@@
 
-\newpage
+%\newpage
 \section{An abstract$^{\text 2}$ machine for ISWIM}

### 2012-11-14 paper/paper.tex
Added macros for names of types, tried to get some overfull hboxes less full.

--- 2012-11-13 paper/paper.tex
@@
 implementation techniques that, in total, speed up the analysis by
-nearly a factor of 500, dropping the analys time to a fraction of a
+nearly a factor of 500, dropping the analysis time to a fraction of a
 second.  Figure~\ref{fig:churchtime} shows the step-wise improvement
@@
   non-deterministic choices that will later collapse into a single
-  join point.  We accompish this by applying lazy evaluation
+  join point.  We accomplish this by applying lazy evaluation
   techniques so that non-determinism is evaluated \emph{by need}.
@@
 \item generate states faster.  We accomplish this by better algorithm
-  design in the fixpoint computation we use to generate state graphs.
+  design in the fixed-point computation we use to generate state graphs.
 \end{enumerate}
@@
 
-\begin{align*}
+\newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
+%\newcommand{\ext}[3]{ext(#1,#2,#3)}
+
+\begin{align*}
+%ext(\msto, \maddr, s) &= \msto \sqcup [\maddr \mapsto s] \\
 \ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
-\ev{\mexpi{0},\menv,\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
+\ev{\mexpi{0},\menv,\ext{\msto}{\mlab}{\{\mkont\}},\kar{\mexpi{1},\mlab}}
 \\
 \co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
-\ev{\mexp,\menv,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
+\ev{\mexp,\menv,\ext{\msto}{\mlab^f}{\{\mval\}},\kfn{\mlab^f,\mlab}}
 \\
 \co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
-\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
+\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\ext{\msto}{\mlab^a}{\{\mval\}}}
 \\
@@
 \ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
-\ev{\mexp,\menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
+\ev{\mexp,\menv[\mvar\mapsto\mvar],\ext{\msto}{\mvar}{\msto(\mlab^a)},\mkont}
 \\
@@
 \begin{align*}
-force &: Store \times Addr \to \Set(Value) \\
+force &: \Store \times \Addr \to \Set(\Value) \\
 force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
-force(\msto,\mval) &= \{\mval\}\\
+force(\msto,\mval) &= \{\mval\} \\
+ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto force(\msto, \mval)] \\
 \ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
@@
 &\machstep\;
-\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
+\ev[^\mcntr]{\mexp,\menv,ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
 bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
-  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
+  \maddr], ext(\msto,\maddr, \mval), \mcntr' \\
 \mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
@@
 \begin{align*}
-\compile{\_} &: Expr \to Env \times Store \times Kont \to State \\
+\compile{\_} &: \Expr \to \Env \times \Store \times \Kont \to \State \\
 \compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\saddr{\menv(\mvar)},\msto}
@@
 
-We represent changes as $\msdiff \in Store\msdiff = List[Addr \times
-  \Set(Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
+We represent changes as $\msdiff \in \StoreDelta = List[\Addr \times
+  \Set(\Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
 instead becomes a log addition
@@
 \begin{align*}
-replay &: Store\msdiff \times Store \to Store \\
+replay &: \StoreDelta \times \Store \to \Store \\
 replay(\mtlst, \msto) &= \msto \\
@@
 
-Here $concat : \Set(Store\msdiff) \to Store\msdiff$ flattens the lists
+Here $concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
 of changes to one; order it appends does not matter:
@@
 
-\begin{alltt}
-Value ::= #t | #f | (cons Value Value) | '() | ...
-\end{alltt}
+\newcommand{\alt}{\;\mid\;}
+\begin{align*}
+\Value &::= \mbox{{\tt \#t}} \alt \mbox{{\tt \#f}} \alt \mbox{{\tt(cons $\Value$ $\Value$)}} \alt \mbox{{\tt '()}} \alt ...
+\end{align*}
 

### 2012-11-14 paper/paper.tex

--- 2012-11-14 paper/paper.tex
@@
 \subsection{Pre-allocating the store}
-Hashes are not the most efficent for raw speed. If we commit to using
-sequential natural numbers for variables and expression labels, by the
-time we finish parsing the input program, we will know exactly how
-many addresses we will need. Thus the store can be represented by a
-vector the size of the largest label (call it $\mlab_{max}$) assigned, plus
-1. The parser should be assigning new, fresh names to variables
-anyway, so this change is a one-liner.
-
-This technique by itself only works in 0-CFA, since there are many
-more addresses that can be allocated in 1-CFA and higher. To support
-general $k$-CFA in a vector representation, we use a growable vector
-and a map from addresses to assigned indices. When assignable indices
-run out, we grow the vector. The cost of this map is less than a hash
-representation for the whole store since typically there are more
-references to an address than allocations of it. This is more
-space-efficient than preallocating $(\mlab_{max}+1)^k$ addresses
-up-front and addressing by the contour converted to polynomial, since
-a majority of combinations of labels will not occur, except in
-pathological programs. The all up-front allocation again breaks down
-when we add compound data, which introduce new kinds of labels
-(e.g. left and right components of a tuple at the contour of
-creation).
+
+Internally, the algorithm at this stage uses hash tables to model the store.
+%
+This is because stores used to be distributed to all states, which
+required a compact, dynamic representation.
+%
+But, such a dynamic structure isn't necessary when we know the
+structure of the store in advance: we know all possible entries, and
+we know its maximum size.
+
+In a monovariant analysis, the domain of the store is
+exactly the set of expressions in the program.
+%
+If we label each expression with a unique natural, the analysis can
+index directly into the store without a hash or a collision.
+%
+Even for polyvariant analyses, it is possible to compute the maximum
+number of addresses and similarly pre-allocate either the spine of the
+store or (if memory is no concern) the entire store.
 

### 2012-11-14 paper/paper.tex
Overhauled fixed point algorithm section

@@
 \ref{fig:aam} a computable abstraction of reachable states, but not an
-effective one. This is not the strategy that AAM, nor we,
-recommend. We explain a succession of approximations to reach the
-baseline analysis that we compare performance against. But first, we
-need a fixed-point combinator to drive the succession of reduction
-relations we define.
-
-\subsection{Generic fixed-point calculator}
+efficient one. This is not the strategy that AAM, nor we, recommend. Through
+this and the following section, we will explain a succession of approximations
+to reach the baseline analysis.  We'll compare performance at each stage to
+identify the criticality of each optimization. 
+%
+We ground this journey by first formulating the analysis in terms of a classic
+fixed-point computation.
+
+\subsection{Static analysis as fixed-point computation}
 \label{sec:fixpoint}
 
-We compute the semantics of a program by iterating the state
-transition relation until a fixed-point in the reachable states is
-reached. We begin with Kleene's fixed-point combinator, ${\mathcal F}$,
-operationalized in the frontier/accumulator style:
-
-\begin{align*}
-{\mathcal F}(f)(X) &= {\mathcal G}(f, X, \varnothing) \\
-{\mathcal G}(f, \varnothing, S) &= S \\
-{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup X) \\
-\end{align*}
-
-The $\machstep$ relation is most naturally written as a function from
-a state to a set of all next states. To calculate $eval$, we find a fixed-point of
-$\machstep$ lifted to step a set of states. The lifting is a natural one:
-
-\begin{align*}
-\lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
-\end{align*}
-
-Then $\eval(e) = {\mathcal F}(\lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.
-
-If we run this on our Church numeral example, we must be
-patient. Indeed, the state space is finite, but it is exponential in
-the size of the program. With $k = 0$, the store can be in one of ${\mathcal
-  O}(2^n)$ many configurations! We take a step toward a polynomial
-complexity baseline by lifting the store out so that there is one
-shared amongst all states.
+Conceptually, the AAM approach calls for computing an analysis as a graph
+exploration: (1) start with an initial state, and (2) compute the transitive
+closure of the transition relation starting from that state.
+
+We can cast this exploration process in terms of a fixed-point calculation.
+%
+Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
+we define the global transfer function:
+\begin{equation*}
+ F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
+ \text.
+\end{equation*}
+Internally, this global transfer function computes the successors of all supplied states, and then includes the inital state:
+\begin{equation*}
+  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
+\end{equation*}
+Then, the evaluator for the analysis computes the least fixed point of the global transfer function:
+\begin{equation*}
+ \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
+\end{equation*}
+where:
+\begin{equation*}
+\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\text.
+\end{equation*}
+
+To conduct this naive exploration on the \Church{} example would require
+considerable time.  Even though the state space is finite, it is exponential in
+the size of the program.  Even with $k = 0$, there are exponentially many
+stores in the AAM framework.
+
+In the next subsection, we'll fix this with a widening and reach polynomial
+(albeit of a high degree) complexity.
+%
+This widening effictively lifts the store out of individual states to create
+a single, global shared store for all.
 

### 2012-11-14 paper/paper.tex
Add section labels in VS graphs.

--- 2012-11-14 paper/paper.tex
@@
 \caption{Relative improvements over the baseline analyzer for the
-  Vardoulakis and Shivers benchmark in terms of peak memory usage, the
-  rate of state transitions, and total analysis time.}
+  \Church{} benchmark in terms of peak memory usage, the rate of state
+  transitions, and total analysis time. Each point is marked with the
+  section that introduces the optimization.}
 \label{fig:churchtime}
@@
 At each step during the initial development, we evaluate the
-implementation on an example, shown in figure~\ref{fig:church}, taken
-from the flow analysis literature~\cite{dvanhorn:Vardoulakis2011CFA2}
-that tests distributivity of multiplication over addition on Church
-numerals.
+implementation on a benchmark from Vardoulakis and
+Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
+of multiplication over addition on Church numerals.
 %% \footnote{It is

### 2012-11-14 paper/paper.tex
Rewrote store-widening section

@@
 \end{equation*}
-where:
-\begin{equation*}
-\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\text.
-\end{equation*}
+where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.
 
@@
 
-A common technique to accelerate convergence in flow analyses is to share
-a common store that is monotonically updated as interpretation
-continues. In the ISWIM language with only unary operators, this makes
-our 0-CFA quartic in the size of the program, which in one step drops the
-complexity from intractable exponentiality to merely daunting
-polynomiality.
-
-We can reuse the reduction relation we have defined, but with some
-scaffolding around it. A state in the store-widened machine is a pair
-of a set of ISWIM states without a store, and a store. A step of this
-machine will pair the store with each of the states in this set,
-reduce, and combine the output stores to create the next. The
-scaffolded reduction relation, $\widehat{\machstep}$, is as follows:
-
-\begin{align*}
-(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
-\mbox{where } cs' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in cs\} \\
-              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in cs\} \\
+A common technique to accelerate convergence in flow analyses is to share a
+common, global store.
+%
+To retain soundness, this store grows monotonically.
+%
+Formally, we can cast this optimization as a second abstraction or as the
+application of a widening operator during the fixed-point iteration.
+%
+In the ISWIM language, such a widening makes 0CFA quartic in the size of the
+program.
+%
+Thus, in one step, complexity drops from intractable exponentiality to merely
+daunting polynomiality.
+
+Since we can cast this optimization as a widening, there is no need to change
+the transition relation itself.
+%
+Rather, what changes is the structure of the fixed-point iteration.
+%
+In each pass, the algorithm will collect all newly produced stores and join
+them together.
+%
+Then, before each transition, it will install this joined store into current
+state.
+
+To describe this process, we'll refactor the transition relation so that it operates on
+a pair of a set of contexts ($C$) and a store ($\sigma$).
+%
+A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
+%
+The refactored relation, $\widehat{\machstep}$, becomes:
+%
+\begin{align*}
+(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
+\mbox{where } C' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in C\} \\
+              \msto' &= \bigsqcup\; \{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in C\} \\
 wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
@@
 \end{align*}
-That is, the next set of states is what we get by stepping each
-given state, without its store. The next store is the combination of all produced
-stores. For simplicity, assume that stuck states step to themselves.
+%
+In effect, the new store is computed as the least upper bound of all subsequent stores.
 

### 2012-11-14 paper/paper.tex
Fixed the table of numbers generator.

--- 2012-11-14 paper/paper.tex
@@
 \caption{Relative improvements over the baseline analyzer for the
-  \Church{} benchmark in terms of peak memory usage, the rate of state
-  transitions, and total analysis time. Each point is marked with the
-  section that introduces the optimization.}
+  Vardoulakis and Shivers benchmark in terms of peak memory usage, the
+  rate of state transitions, and total analysis time.}
 \label{fig:churchtime}
@@
 At each step during the initial development, we evaluate the
-implementation on a benchmark from Vardoulakis and
-Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
-of multiplication over addition on Church numerals.
+implementation on an example, shown in figure~\ref{fig:church}, taken
+from the flow analysis literature~\cite{dvanhorn:Vardoulakis2011CFA2}
+that tests distributivity of multiplication over addition on Church
+numerals.
 %% \footnote{It is
@@
 \begin{figure}
-\begin{tabular}{@{}l|r|r|r|r|r|r|r@{}}
-Program & LOC
-& \multicolumn{2}{c|}{Time (ms)}
-& \multicolumn{2}{c|}{Space (mb)}
-& \multicolumn{2}{c@{}}{Speed (state/sec)}
-\\
-\hline\hline
-nucleic &  3508 & 0 & 0 & 0 & 0 & 0 & 0\\
-matrix  &  747  & ? & ? & ? & ? & ? & ?\\
-nbody   & 1434  & ? & ? & ? & ? & ? & ?\\
-earley  &  667  & ? & ? & ? & ? & ? & ?\\
-maze    &  681  & ? & ? & ? & ? & ? & ?\\
-church  &   41  & ? & ? & ? & ? & ? & ?\\
-lattice &  214  & ? & ? & ? & ? & ? & ?\\
-boyer   &  642  & ? & ? & ? & ? & ? & ?\\
-mbrotz  &   68  & ? & ? & ? & ? & ? & ?
-\end{tabular}
+\include{bench-overview}
 \caption{Overview performance comparison between baseline and

### 2012-11-14 paper/paper.tex
Consistency in naive spelling and fixed-point hyphenation.

--- 2012-11-14 paper/paper.tex
@@
 \newtheorem{lemma}{Lemma}
+
+\newcommand{\naive}{na\"ive}
+\newcommand{\naively}{na\"ively}
+\newcommand{\Naive}{Na\"ive}
+\newcommand{\Naively}{Na\"ively}
 % \usepackage{xltxtra}
@@
 
-This article contributes a step by step process for going from a naive
+This article contributes a step by step process for going from a \naive
 analyzer derived under the abstracting abstract machine approach, to
@@
 \end{equation*}
-Then, the evaluator for the analysis computes the least fixed point of the global transfer function:
+Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
 \begin{equation*}
@@
 
-To conduct this naive exploration on the \Church{} example would require
+To conduct this \naive exploration on the \Church{} example would require
 considerable time.  Even though the state space is finite, it is exponential in
@@
 
-\subsubsection{Na\"ively}
+\subsubsection{\Naively}
 
@@
 
-\subsubsection{Less na\"ively and less precisely}
+\subsubsection{Less \naively and less precisely}
 

### 2012-11-14 paper/paper.tex
Consistent hyphenation for 0-CFA

@@
 %
-In the ISWIM language, such a widening makes 0CFA quartic in the size of the
+In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
 program.
@@
 
-[Computes results more like Earl, et al.'s 0CFA, which times out on
+[Computes results more like Earl, et al.'s 0-CFA, which times out on
   the Church numeral benchmark and so even though it offers a fair
@@
 Vardoulakis and Shivers evaluate their CFA2
-analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0CFA
-defined in their framework.  [NEED TIMING FOR 0CFA ON CHURCH.]  The
+analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0-CFA
+defined in their framework.  [NEED TIMING FOR 0-CFA ON CHURCH.]  The
 Church example is the largest benchmark Vardoulakis and Shivers

### 2012-11-14 paper/paper.tex
Spacing issue with \naive command.

@@
 \newcommand{\naive}{na\"ive}
-\newcommand{\naively}{na\"ively}
+\newcommand{\naively}{ na\"ively}
 \newcommand{\Naive}{Na\"ive}
@@
 
-This article contributes a step by step process for going from a \naive
-analyzer derived under the abstracting abstract machine approach, to
-an efficient program analyzer.  The end result of the process is a two
-to three order-of-magnitude improvement over the systematically
-derived analyzer, making it competitive with hand-optimized
-implementations that compute fundamentally less precise results.
+This article contributes a step by step process for going from a
+\naive analyzer derived under the abstracting abstract machine
+approach, to an efficient program analyzer.  The end result of the
+process is a two to three order-of-magnitude improvement over the
+systematically derived analyzer, making it competitive with
+hand-optimized implementations that compute fundamentally less precise
+results.
 \end{abstract}
@@
 
-To conduct this \naive exploration on the \Church{} example would require
+To conduct this \naive{} exploration on the \Church{} example would require
 considerable time.  Even though the state space is finite, it is exponential in
@@
 
-\subsubsection{Less \naively and less precisely}
+\subsubsection{Less \naively{} and less precisely}
 

### 2012-11-14 paper/paper.tex
Better labelling of timeout and oom.
Macrofied the change? boolean.

@@
 \begin{itemize}
- \item{{\tt $\msto$}, the store}
+ \item{$\msto$, the store}
  \item{{\tt todo}, the workset}
  \item{{\tt seen}, a map from states to the timestamps at which they were last seen}
- \item{{\tt $\msdiff$}, the store changes for the current step}
- \item{{\tt $\msdiff$?}, a boolean tracking whether the stepped state contributed a store change}
+ \item{$\msdiff$, the store changes for the current step}
+ \item{$\changep$, a boolean tracking whether the stepped state contributed a store change}
  \item{{\tt T}, the timestamp of the store}
@@
 $c$ to {\tt todo} by the following check:
-\begin{align*}\msdiff\mbox{{\tt ?}} \vee {\tt seen}(c) \neq {\tt T}\end{align*}
+\begin{align*}\changep \vee {\tt seen}(c) \neq {\tt T}\end{align*}
 If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.
@@
 store are permitted. Each time $\msdiff$ is successfully extended, we
-set {\tt $\msdiff$?} to true. Before each individual step, we set it
+set $\changep$ to true. Before each individual step, we set it
 to false.
@@
       todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
-      foreach c \(\in\) old: \(\msdiff\)? := false; c()
+      foreach c \(\in\) old: \(\changep\) := false; c()
       unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
@@
 \caption{Overview performance comparison between baseline and
-  optimized analyzer.}
+  optimized analyzer (entries of \text{{\small $t$}} mean timeout, and \text{{\small $m$}} mean out of memory).}
 \label{fig:bench-overview}

### 2012-11-14 paper/paper.tex
Made numbers fit with unit trickery.
Multiple character variables wrapped in mathit.

@@
 \begin{align*}
-\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
+\eval(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
 \end{align*}
@@
 
-The uniform $k$-CFA allocation strategy would make $eval$ in figure
+The uniform $k$-CFA allocation strategy would make $\eval$ in figure
 \ref{fig:aam} a computable abstraction of reachable states, but not an
@@
 (C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
-\mbox{where } C' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in C\} \\
-              \msto' &= \bigsqcup\; \{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in C\} \\
-wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
-wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
-wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
-wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
+\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c, \msto^c), c \in C\} \\
+              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
+\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
+\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
+\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
+\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
 \end{align*}
@@
 \ev[^\mcntr]{\mexp,\menv,ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
-bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
+\mathit{bind\,}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
   \maddr], ext(\msto,\maddr, \mval), \mcntr' \\
@@
 \begin{align*}
-replay &: \StoreDelta \times \Store \to \Store \\
-replay(\mtlst, \msto) &= \msto \\
-replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
+\replay &: \StoreDelta \times \Store \to \Store \\
+\replay(\mtlst, \msto) &= \msto \\
+\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
 \end{align*}
@@
 \begin{align*}
-(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', replay(\msdiff,\msto)) \\
-\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
-step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
-step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
-cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
-\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
-\end{align*}
-
-Here $concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
+(\states, \msto) &\mathrel{\widehat{\machstep}} (\states\cup \states', \replay(\msdiff,\msto)) \\
+\mbox{ where } (\states', \msdiff) &= \step^*(\varnothing, \states, \mtlst) \\
+\step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
+\step^*(S, \{c\}\cup \states, \msdiff &= \step^*(S\cup cs^*, \states, \msdiff^*) \\
+\states^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
+\msdiff^* &= \concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
+\end{align*}
+
+Here $\concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
 of changes to one; order it appends does not matter:
 \begin{align*}
-concat(\varnothing) &= \mtlst \\
-concat(\{\msdiff\}\cup D) &= append(\msdiff, concat(D))
+\concat(\varnothing) &= \mtlst \\
+\concat(\{\msdiff\}\cup D) &= \append(\msdiff, \concat(D))
 \end{align*}
@@
 \begin{align*}
-\interpdelta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
-\interpdelta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
+\interpdelta(\msto,\ {\tt qlist}) &= (\mbox{{\tt '()}}, \msto) \\
+\interpdelta(\msto,\ {\tt qlist}, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
 \qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
@@
 \bigsqcup(v) &= v \\
-\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
-merge(v, v) &= v \\
-merge(n, m) &= {\tt Number} \\
-merge(n, v) &= {\tt QData} \\
-merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
-merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
-merge({\tt QPair}, v) &= {\tt QData} \\
+\bigsqcup(v, vs ...) &= \merge(v, \bigsqcup(vs ...)) \\
+\merge(v, v) &= v \\
+\merge(n, m) &= {\tt Number} \\
+\merge(n, v) &= {\tt QData} \\
+\merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
+\merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
+\merge({\tt QPair}, v) &= {\tt QData} \\
 \vdots
@@
 \begin{figure}
+\centering
 \include{bench-overview}

### 2012-11-14 paper/paper.tex
Whoopsie in redefining merge.
Bigger is better.

--- 2012-11-14 paper/paper.tex
@@
 \newtheorem{lemma}{Lemma}
-
-\newcommand{\naive}{na\"ive}
-\newcommand{\naively}{ na\"ively}
-\newcommand{\Naive}{Na\"ive}
-\newcommand{\Naively}{Na\"ively}
 % \usepackage{xltxtra}
@@
 
-This article contributes a step by step process for going from a
-\naive analyzer derived under the abstracting abstract machine
-approach, to an efficient program analyzer.  The end result of the
-process is a two to three order-of-magnitude improvement over the
-systematically derived analyzer, making it competitive with
-hand-optimized implementations that compute fundamentally less precise
-results.
+This article contributes a step by step process for going from a naive
+analyzer derived under the abstracting abstract machine approach, to
+an efficient program analyzer.  The end result of the process is a two
+to three order-of-magnitude improvement over the systematically
+derived analyzer, making it competitive with hand-optimized
+implementations that compute fundamentally less precise results.
 \end{abstract}
@@
   \Church{} benchmark in terms of peak memory usage, the rate of state
-  transitions, and total analysis time. Each point is marked with the
-  section that introduces the optimization.}
+  transitions, and total analysis time. (Bigger is better.) Each point
+  is marked with the section that introduces the optimization.}
 \label{fig:churchtime}
@@
 \end{equation*}
-Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
+Then, the evaluator for the analysis computes the least fixed point of the global transfer function:
 \begin{equation*}
@@
 \end{equation*}
-where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.
-
-To conduct this \naive{} exploration on the \Church{} example would require
+where:
+\begin{equation*}
+\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\text.
+\end{equation*}
+
+To conduct this naive exploration on the \Church{} example would require
 considerable time.  Even though the state space is finite, it is exponential in
@@
 
-A common technique to accelerate convergence in flow analyses is to share a
-common, global store.
-%
-To retain soundness, this store grows monotonically.
-%
-Formally, we can cast this optimization as a second abstraction or as the
-application of a widening operator during the fixed-point iteration.
-%
-In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
-program.
-%
-Thus, in one step, complexity drops from intractable exponentiality to merely
-daunting polynomiality.
-
-Since we can cast this optimization as a widening, there is no need to change
-the transition relation itself.
-%
-Rather, what changes is the structure of the fixed-point iteration.
-%
-In each pass, the algorithm will collect all newly produced stores and join
-them together.
-%
-Then, before each transition, it will install this joined store into current
-state.
-
-To describe this process, we'll refactor the transition relation so that it operates on
-a pair of a set of contexts ($C$) and a store ($\sigma$).
-%
-A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
-%
-The refactored relation, $\widehat{\machstep}$, becomes:
-%
-\begin{align*}
-(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
-\mbox{where } C' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in C\} \\
-              \msto' &= \bigsqcup\; \{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in C\} \\
+A common technique to accelerate convergence in flow analyses is to share
+a common store that is monotonically updated as interpretation
+continues. In the ISWIM language with only unary operators, this makes
+our 0-CFA quartic in the size of the program, which in one step drops the
+complexity from intractable exponentiality to merely daunting
+polynomiality.
+
+We can reuse the reduction relation we have defined, but with some
+scaffolding around it. A state in the store-widened machine is a pair
+of a set of ISWIM states without a store, and a store. A step of this
+machine will pair the store with each of the states in this set,
+reduce, and combine the output stores to create the next. The
+scaffolded reduction relation, $\widehat{\machstep}$, is as follows:
+
+\begin{align*}
+(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
+\mbox{where } cs' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in cs\} \\
+              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in cs\} \\
 wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
@@
 \end{align*}
-%
-In effect, the new store is computed as the least upper bound of all subsequent stores.
+That is, the next set of states is what we get by stepping each
+given state, without its store. The next store is the combination of all produced
+stores. For simplicity, assume that stuck states step to themselves.
 
@@
 
-\subsubsection{\Naively}
+\subsubsection{Na\"ively}
 
@@
 
-\subsubsection{Less \naively{} and less precisely}
+\subsubsection{Less na\"ively and less precisely}
 
@@
 \begin{figure}
-\include{bench-overview}
+\begin{tabular}{@{}l|r|r|r|r|r|r|r@{}}
+Program & LOC
+& \multicolumn{2}{c|}{Time (ms)}
+& \multicolumn{2}{c|}{Space (mb)}
+& \multicolumn{2}{c@{}}{Speed (state/sec)}
+\\
+\hline\hline
+nucleic &  3508 & 0 & 0 & 0 & 0 & 0 & 0\\
+matrix  &  747  & ? & ? & ? & ? & ? & ?\\
+nbody   & 1434  & ? & ? & ? & ? & ? & ?\\
+earley  &  667  & ? & ? & ? & ? & ? & ?\\
+maze    &  681  & ? & ? & ? & ? & ? & ?\\
+church  &   41  & ? & ? & ? & ? & ? & ?\\
+lattice &  214  & ? & ? & ? & ? & ? & ?\\
+boyer   &  642  & ? & ? & ? & ? & ? & ?\\
+mbrotz  &   68  & ? & ? & ? & ? & ? & ?
+\end{tabular}
 \caption{Overview performance comparison between baseline and
@@
 
-[Computes results more like Earl, et al.'s 0-CFA, which times out on
+[Computes results more like Earl, et al.'s 0CFA, which times out on
   the Church numeral benchmark and so even though it offers a fair
@@
 Vardoulakis and Shivers evaluate their CFA2
-analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0-CFA
-defined in their framework.  [NEED TIMING FOR 0-CFA ON CHURCH.]  The
+analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0CFA
+defined in their framework.  [NEED TIMING FOR 0CFA ON CHURCH.]  The
 Church example is the largest benchmark Vardoulakis and Shivers

### 2012-11-14 paper/paper.tex
More mathit wrapping.

--- 2012-11-14 paper/paper.tex
@@
 
-\begin{align*}
-force &: \Store \times \Addr \to \Set(\Value) \\
-force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
-force(\msto,\mval) &= \{\mval\} \\
-ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto force(\msto, \mval)] \\
+\renewcommand{\ext}{\mathit{ext}}
+
+\begin{align*}
+\force &: \Store \times \Addr \to \Set(\Value) \\
+\force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
+\force(\msto,\mval) &= \{\mval\} \\
+\ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto \force(\msto, \mval)] \\
 \ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
@@
 &\machstep\;
-\ev[^\mcntr]{\mexp,\menv,ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
+\ev[^\mcntr]{\mexp,\menv,\ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
 \mathit{bind\,}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
-  \maddr], ext(\msto,\maddr, \mval), \mcntr' \\
+  \maddr], \ext(\msto,\maddr, \mval), \mcntr' \\
 \mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
@@
 \begin{align*}
-bind^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{force(\msto,\mval)}}{\msdiff}, \mcntr' \\
+\mathit{bind\,}^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\msdiff}, \mcntr' \\
 \text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
@@
 \bigsqcup(v) &= v \\
-\bigsqcup(v, vs ...) &= \merge(v, \bigsqcup(vs ...)) \\
-\merge(v, v) &= v \\
-\merge(n, m) &= {\tt Number} \\
-\merge(n, v) &= {\tt QData} \\
-\merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
-\merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
-\merge({\tt QPair}, v) &= {\tt QData} \\
+\bigsqcup(v, vs ...) &= \fmerge(v, \bigsqcup(vs ...)) \\
+\fmerge(v, v) &= v \\
+\fmerge(n, m) &= {\tt Number} \\
+\fmerge(n, v) &= {\tt QData} \\
+\fmerge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
+\fmerge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
+\fmerge({\tt QPair}, v) &= {\tt QData} \\
 \vdots

### 2012-11-14 paper/paper.tex
Spacing fix for (cons a_i d_i)

@@
   \Church{} benchmark in terms of peak memory usage, the rate of state
-  transitions, and total analysis time. (Bigger is better.) Each point
-  is marked with the section that introduces the optimization.}
+  transitions, and total analysis time. Each point is marked with the
+  section that introduces the optimization.}
 \label{fig:churchtime}
@@
  \item{Define special value lattice elements for compound data domains that can be quoted
-       (e.g. {\tt QPair} for $\left\{ ({\tt cons} a_i d_i)\right\}_i$, {\tt QVector} for immutable vectors, etc.)}
+       (e.g. {\tt QPair} for $\left\{ ({\tt cons}\ a_i\ d_i)\right\}_i$, {\tt QVector} for immutable vectors, etc.)}
  \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}

### 2012-11-14 paper/paper.tex
Revamped description of store-allocating results

@@
 
-The final approximation we make to get to our baseline analysis is
-store-allocating results of application sub-expressions. The semantics
-of the previous section is where AAM leaves off. However, notice that
-the $\kfn{}$ continuation stores a value - this makes the space of
-continuations quadratic rather than linear in 0-CFA. A linear space
-drops the complexity to cubic.
-
-To achieve this, we instead allocate an address for the value position
+The final approximation we make to get to our point of departure is
+store-allocating results of application sub-expressions. 
+The AAM approach stops at the previous optimization. 
+However, the $\kfn{}$ continuation stores a value, and this makes the space of
+continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0CFA.
+Having the space of continuations grow linearly with the size of the program 
+will drop the overall complexity to cubic (as expected).
+
+To achieve this linearity for continuation, we allocate an address for the value position
 when we create the continuation.  This address and the tail address
 are both determined by the label of the application point, so the
-space becomes linear and the overall complexity drops to cubic. Note
-that this is a critical abstraction in languages with n-ary functions,
-since otherwise the continuation space becomes ${\mathcal O}(n^n)$!
-This amount of distinction between continuations is unnecessary
-in practice.
+space becomes linear and the overall complexity drops to cubic. 
+This is a critical abstraction in languages with $n$-ary functions,
+since otherwise the continuation space grows super-exponentially.
+% In theory, this aggressive distinction between continuations might buy
+% additional precision, but in practice, it does not.
 

### 2012-11-14 paper/paper.tex
Type error in imperative algo. Forgot to pass in empty store changes

--- 2012-11-14 paper/paper.tex
@@
   \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
-  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \kmt\))
+  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \mtlst, \kmt\))
   while(true):

### 2012-11-14 paper/paper.tex
Plurality of continuations.
Tightend more prose on store-allocated resutls

@@
 
-To achieve this linearity for continuation, we allocate an address for the value position
-when we create the continuation.  This address and the tail address
-are both determined by the label of the application point, so the
-space becomes linear and the overall complexity drops to cubic. 
-This is a critical abstraction in languages with $n$-ary functions,
-since otherwise the continuation space grows super-exponentially.
+To achieve this linearity for continuations, we allocate an address
+for the value position when we create the continuation.  This address
+and the tail address are both determined by the label of the
+application point, so the space becomes linear and the overall
+complexity drops to cubic.  This is a critical abstraction in
+languages with $n$-ary functions, since otherwise the continuation
+space grows super-exponentially.
 % In theory, this aggressive distinction between continuations might buy
@@
 
-The important rules that change are the following, specialized to 0-CFA:
+If we specialize to 0CFA, the evaluation rules become:
 
@@
 
-This extra store-allocation is effectively naming all intermediate
-results, and thus the precision aligns with an analysis specialized to
-ANF. We can also play a representation trick in 0-CFA and remove
-$\menv$. In monovariant analyses, variables map to themselves, meaning
-$\menv$ is effectively the identity function. That degeneration, in
-turn, allows us to discard it from the analysis.
+% This extra store-allocation is effectively naming all intermediate
+% results, and thus the precision aligns with an analysis specialized to
+% ANF. We can also play a representation trick in 0-CFA and remove
+% $\menv$. In monovariant analyses, variables map to themselves, meaning
+% $\menv$ is effectively the identity function. That degeneration, in
+% turn, allows us to discard it from the analysis.
 

### 2012-11-14 paper/paper.tex
Finished intro to implementation techniques

@@
 
-We discuss a succession of techniques for increasing performance by
-decreasing the interpretive overhead of abstract interpretation. They
-come in two flavors: decreasing the size of the state space with domain-specific knowledge [DSK], and
-decreasing the amount of intermediate data structures [Eng](ineering).
-
+In this section, we discuss the optimizations for abstract interpreters that
+yield our ultimate performance gains.
+%
+We have two broad categories of these optimizations: (1) transition elimination
+and (2) pragmatic improvement.
+%
+The transition-elimination optimizations reduce the overall number of transitions
+made by the analyzer by performing:
 \begin{enumerate}
- \item{[DSK] lazy non-determinism: quotient by sets of values}
- \item{[DSK] abstract compilation: eliminate {\tt ev} states}
- \item{[Eng] store deltas: operate on store changes rather than whole stores}
- \item{[DSK/Eng] timestamp: represent the store as a global, mutable
-   hash and determine a state's store by the number of changes made to
-   the global store at the time of visiting}
- \item{[Eng] preallocation: represent the global store as a global, mutable vector}
- \item{[DSK] uniform literal approximation: coarsely abstract literal data}
+ \item lazy non-determinism;
+ \item abstract compilation; and
+ \item uniform literal approximation.
 \end{enumerate}
-
-Some techniques preserve the precision of the underlying analysis, and
-others do not. We will discuss the design decisions behind the
-precision-lossy techniques.  Each technique requires minimal changes
-to the reduction relation, and each improves performance by
-integer factors.
+The pragmatic improvements reduce overhead and trade space for time
+by utilizing:
+\begin{enumerate}
+ \item store deltas;
+ \item timestamped stores; and
+ \item preallocated data structures.
+\end{enumerate}
+
+Some techniques preserve the precision of the underlying analysis, and others
+do not.
+%
+For any technique that loses precision, we will discuss the design rationale
+for the move.
 

### 2012-11-14 paper/paper.tex
Cleaned up lazy non-determinism

@@
 
-A shortcoming of the AAM approach is that much of the non-determinism
-is spurious. For example, in a function application, {\tt (f x y)},
-where all variables have been joined with multiple values, each
-variable reference transitions to as many new states as there are
-values, only to be joined back together in their respective
-application positions. We would see transitions of the following
-shape:
+Tracing the execution of the analysis reveals an immediate shortcoming:
+there is a high degree of branching and merging in the exploration.
+%
+Surveying this branching has no benefit for precision.
+%
+For example, in a function application, {\tt (f x y)},
+where {\tt f}, {\tt x} and {\tt y} each have several values
+each argument evaluation induces two-way branching, only to be ultimately joined back together in their respective
+application positions. 
+%
+Transition patterns of this shape litter the state-graph:
+%
 \begin{center}
@@
 \end{center}
-For this reason, we {\it delay} the non-determinism
-until it is needed in {\it strict contexts} (such as the guard for an
-{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
+To avoid the spurious forking and joining, we {\it delay} the non-determinism
+until and unless it is needed in {\it strict contexts} (such as the guard for an
+{\tt if} a called procedure, or a numerical primitive application). 
+%
+Doing so collapses these forks and joins into a linear sequence of states:
 \begin{center}
@@
 
-This does not change the concrete semantics of the language. It is an
-abstraction of the otherwise several states the original
-non-deterministic semantics steps to. We say the abstraction is
-\emph{lazy} because it delays dereferencing an address until its
-contents are \emph{needed} as values in the semantics. It does not
-change the execution order that leads to the values that are stored in
+In case it's unclear, this shift does not change the concrete semantics of the
+language to be lazy.  Rather, it abstracts over transitions that the original
+non-deterministic semantics steps through. 
+%
+We say the abstraction is \emph{lazy} because it delays dereferencing an
+address until its contents are \emph{needed} as values in the semantics. It
+does not change the execution order that leads to the values that are stored in
 the address.
@@
 We introduce a new kind of value, $\saddr{\maddr}$, that represents a
-delayed lookup of $\maddr$. This acts to quotient the state space by
-the values at the delayed address. The following rules highlight the
+delayed lookup of $\maddr$.  The following rules highlight the
 changes to the semantics:
@@
 
-\paragraph{Lose precision; simplify implementation}
+We have two choices for how to implement lazy non-determinism.
+
+\paragraph{Option 1: Lose precision; simplify implementation}
 This semantics introduces a subtle precision difference over the
@@
 
-\paragraph{Regain precision; complicate implementation}
+\paragraph{Option 2: Regain precision; complicate implementation}
 The administrative nature of laziness means that we could remove the

### 2012-11-14 paper/paper.tex
Fixing up some citations.

--- 2012-11-14 paper/paper.tex
@@
 allocation strategy is an abstract
-interpretation~\cite{MIGHTANDMANOLIOS}.  In particular, concrete
-interpretation is a kind of abstract interpretation.  So is an
-interpretation that allocates a single cell into which all bindings
+interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
+concrete interpretation is a kind of abstract interpretation.  So is
+an interpretation that allocates a single cell into which all bindings
 and continuations are stored.  On the one hand is an abstract
@@
 Vardoulakis and Shivers evaluate their CFA2
-analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0-CFA
-defined in their framework.  [NEED TIMING FOR 0-CFA ON CHURCH.]  The
-Church example is the largest benchmark Vardoulakis and Shivers
-consider.  More work would be required to scale the analyzer to the
-set of features required by our benchmarks.
+analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of
+0-CFA defined in their framework and the example we draw on is the
+largest benchmark Vardoulakis and Shivers consider.  More work would
+be required to scale the analyzer to the set of features required by
+our benchmarks.
 
@@
 analyzing the full suite of benchmarks considered here were the Soft
-Typing system of Wright [CITE] and, in many ways its successor, the
-Polymorphic splitting system of Wright and
+Typing system of Wright and
+Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
+successor, the Polymorphic splitting system of Wright and
 Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This

### 2012-11-14 paper/paper.tex
Cleaned up time-stamping the store

@@
 \begin{align*}
-\force &: \Store \times \Addr \to \Set(\Value) \\
+\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
@@
 
-The laziness optimization saved time by doing the same amount of
+The prior optimization saved time by doing the same amount of
 reasoning as before but in fewer transitions. We can exploit the same
-idea - same reasoning, fewer transitions - with abstract
+idea---same reasoning, fewer transitions---with abstract
 compilation. Abstract compilation is \emph{precision-preserving} and
@@
 deterministic into ``abstract bytecodes.''  The abstract interpreter
-then does in one transition what previously took many. Abstract
+then does in one transition what previously took many. In short, abstract
 compilation eliminates unnecessary allocation, deallocation and
@@
 
-The essence of the compilation effect can be seen by considering an
-example such as
+The following example illustrates
+the essence of abstract compilation effect:
 \[
@@
 \]
-which makes the following transitions:
+makes the following transitions:
 \begin{align}
@@
 change log on the full store after all steps have completed. This uses
-far fewer join operations, leading to less overhead, and is \emph{precision preserving}.
-
-We represent changes as $\msdiff \in \StoreDelta = List[\Addr \times
-  \Set(\Storeable)]$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
-instead becomes a log addition
+far fewer join operations, leading to less overhead, and is \emph{precision-preserving}.
+
+We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
+  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
+ becomes a log addition
 $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
@@
 is straightforward:
-
+% TODO: Turn the recursion here into ...:
 \begin{align*}
@@
 
-The reduction relation does have to change to accommodate this
-technique, but only in this regard. The whole store is still given so
-that lookups are possible. Example rule:
+The transition relation is identical except for the addition of this change log.
+%
+Lookups will never rely on the change log, and can use the originally supplied store unmodified:
 
@@
 Here $\concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
-of changes to one; order it appends does not matter:
+of changes to one; the order in which it appends does not matter:
 \begin{align*}
@@
 
-And, there is a better way. Shivers' original work on k-CFA was
+And, there is a better way. Shivers' original work on $k$-CFA was
 susceptible to the same problem, and he suggested three complementary
@@
 
-There is a subtle loss of precision and Shivers' original time-stamp
+There is a subtle loss of precision in Shivers' original time-stamp
 technique that we can fix. For a given abstract state, all writes to
@@
 
-The current state of the state-space search now has six components:
-
+At this point, we can begin to think of the analysis as an imperative
+algorithm with six major components:
+%
 \begin{itemize}
- \item{$\msto$, the store}
- \item{{\tt todo}, the workset}
- \item{{\tt seen}, a map from states to the timestamps at which they were last seen}
- \item{$\msdiff$, the store changes for the current step}
- \item{$\changep$, a boolean tracking whether the stepped state contributed a store change}
- \item{{\tt T}, the timestamp of the store}
+ \item{$\msto$, the store;}
+ \item{{\tt todo}, the workset;}
+ \item{{\tt seen}, a map from states to the timestamps at which they were last seen;}
+ \item{$\msdiff$, the store changes for the current step;}
+ \item{$\changep$, a boolean tracking whether the stepped state contributed a store change; and}
+ \item{{\tt T}, the timestamp of the store.}
 \end{itemize}
 
-The new $eval$ calculation is defined in
-figure~\ref{fig:imperative}.  To ensure termination, we guard adding
+The new $\eval$ calculation is defined in
+figure~\ref{fig:imperative}.  To ensure termination, we guard against adding
 $c$ to {\tt todo} by the following check:
@@
 
-Practical issues of performance arise when one chooses the
-``obvious,'' or ``natural'' abstraction for compound data. Our running
+Performance issues arise when one chooses the
+``obvious,'' or ``natural'' abstraction for compound data structures like list. 
+%
+Our running
 example will be the classic Boyer benchmark. The detail of note for

### 2012-11-14 paper/paper.tex
Pushing on related work.

--- 2012-11-14 paper/paper.tex
@@
 
-\item {\bf maze}: generates a random maze and finds a path (used in
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
-  Makes use of Scheme's {\tt call-with-current-continuation}
-  operation.
+\item {\bf maze}: generates a random maze using Scheme's {\tt
+  call/cc} operation and finds a path solving
+  the maze (used in
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
 
 \item {\bf church}: tests distributivity of multiplication over
-  addition for a few Church numerals (introduced by
+  addition for Church numerals (introduced by
   \cite{dvanhorn:Vardoulakis2011CFA2}).
@@
 \item {\bf lattice}: enumerates the maps between two lattices (used in
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
 
 \item {\bf boyer}: a term-rewriting theorem prover (used in
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
 
@@
   written in a continuation-passing style (used by
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98})
-
+  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
 \end{enumerate}
@@
 
-[Computes results more like Earl, et al.'s 0-CFA, which times out on
-  the Church numeral benchmark and so even though it offers a fair
-  point of comparison, a more thorough evaluation is likely to
-  uninformative as the other benchmarks are highly likely to timeout
-  as well (and it would require significant effort to extend their
-  implementation with the features needed to analyze our benchmark
-  suite).  That implementation is evaluated against much smaller
-  benchmarks: the largest program consists of 63 \emph{subexpressions}
-  and times out after 30 minutes.]
+The analysis considered here computes results more like Earl, et al.'s
+0-CFA, which times out on the \Church{} benchmark because it does not
+widen the store as described for our baseline evaluator.  So even
+though it offers a fair point of comparison, a more thorough
+evaluation is probably uninformative as the other benchmarks are
+likely to timeout as well (and it would require significant effort to
+extend their implementation with the features needed to analyze our
+benchmark suite).  That implementation is evaluated against much
+smaller benchmarks: the largest program is 30 lines.
 
@@
 our benchmarks.
-
-[Should probably talk about mCFA too.]
 
@@
 abstract interpretation
-\cite{dvanhorn:Cousot:1977:AI,dvanhorn:Cousot:1979:Galois}.
+\cite{dvanhorn:Cousot:1977:AI,dvanhorn:Cousot1979Systematic}.
 
@@
 implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
-the church numeral example of figure~\ref{fig:church} in 7
-milliseconds, about 25 times faster than the fastest implementation
-considered here.  These analyses compute very different things (as
-detailed in section~\ref{sec:accept}), so the performance comparison
-is not apples-to-apples.  We view the primary contribution of this
-work as a systematic path that eases the design, verification, and
-implementation of analyses using the abstracting abstract machine
-approach to within a factor of performant constraint-based analyses.
+the \Church{} benchmark about 25 times faster than the fastest
+implementation considered here.  These analyses compute very different
+things, so the performance comparison is not apples-to-apples.
+
+The AAM approach, and the state transition graphs it generates, encode
+temporal properties not found in classical constraint-based analyses
+for higher-order programs.
+%
+These analyses (ultimately) compute judgments on program terms and
+contexts, e.g., at expression $e$, variable $x$ may have value $v$.
+%
+The judgments do not relate the order in which expressions and context
+may be evaluated in a program, e.g., it has nothing
+to say with regard to question like, ``Do we always evaluate $e_1$
+before $e_2$?'' or ``Is it always the case that a file handle is
+opened, read and then closed in that order?''
+%
+The state transition graphs can answer these kinds of queries, but
+this does not come for free: respecting temporal order imposes an
+order in which states and terms may be evaluated \emph{during} the
+analysis.
+
+We view the primary contribution of this work as a systematic path
+that eases the design, verification, and implementation of analyses
+using the abstracting abstract machine approach to within a factor of
+performant constraint-based analyses.
 

### 2012-11-14 paper/paper.tex
Took a pass on abstracting compound literal data structures

@@
 
-Performance issues arise when one chooses the
-``obvious,'' or ``natural'' abstraction for compound data structures like list. 
-%
-Our running
-example will be the classic Boyer benchmark. The detail of note for
-the Boyer benchmark is that it has large list literals, the exact
-values of which do not matter for many analyses. If we
-interpret a list literal of 30 elements (that again have 10 elements
-each) as 300 applications of {\tt cons}, and then interpret a function
-that traverses the tree, we will end up interpreting that function at
-least 300 times, when once or twice would do.
-
+The abstraction of compound data structures like lists has deep implications 
+for precision and performance.
+%
+The classic list-heavy benchmark {\tt boyer}---a simplified implementation
+of the Boyer-Moore theorem-prover---drove our own considerations for
+abstractions for compound data structures.
+%
+Boyer's fluent use of literal list dooms the ``natural'' abstraction
+of lists to fail from overprecision.
+%
+In short, if we interpret a literal list {\tt '(a b c)} as {\tt (cons 'a (cons 'b
+(cons 'c '())))}, even a monovariant allocation stategy for abstract
+cons cells will precisely and exactly create a 3-cell list.
+
+If code contains a length-300 literal list, then its analysis yields a length-300 abstract
+list as well.
+%
+In many cases, the specific contents of the list add little precision
+to the resulting analysis, and yet the analyzer will dutifully execute
+recursive functions over the entirety of these structures at every
+encounter.
+%
+An alternative in this case is to explicitly flatten literal lists
+into single abstract cells.
+
+We will explore both options here.
+%
 We will first explain the natural abstraction of tuples, and then we
@@
 
-\subsubsection{\Naively}
+\subsubsection{Option 1: The natural abstraction}
 
@@
 
-\subsubsection{Less \naively{} and less precisely}
+\subsubsection{Option 2: A precise yet compact abstraction}
 
@@
 the rate of states-per-second explored by the analysis for each of the
-optimizations discussed in section~\ref{sec:opt}, culmulatively
+optimizations discussed in section~\ref{sec:opt}, cumulatively
 applied.  The analysis is stopped after consuming 30 minutes of time

### 2012-11-14 paper/paper.tex
Getting evaluation in working order.

--- 2012-11-14 paper/paper.tex
@@
 \end{center}
-\caption{Relative improvements over the baseline analyzer for the
+\caption{Factor improvements over the baseline analyzer for the
   \Church{} benchmark in terms of peak memory usage, the rate of state
@@
 
-We have implemented, optimized, and evaluated \emph{all} of the
-machines presented in their JFP paper
-\cite{dvanhorn:VanHorn2012Systematic}, with the exception of the
-machine for performing pushdown flow analysis.  This includes machines
+We have implemented, optimized, and evaluated an analysis framework
 supporting higher-order functions, state, first-class control,
-laziness, and stack inspection.
-
+compound data, and a large number of primitive kinds of data and
+operations such as floating point, complex, and exact rational
+arithmetic.  The analysis is evaluated against a suite of benchmarks
+drawn from the literature.
+%
 For each benchmark, we collect analysis times, peak memory usage, and
@@
 
-All benchmarks are calculated as an average of N runs on a [MACHINE
-  SPECS].
-
-The benchmarks make use of mutation, first-class control (in the form
-of Scheme's {\tt call/cc} operator), compound data such as lists and
-vectors, many more primitive forms of data and operations such as
-Scheme's numeric tower including exact rationals, complex, and
-floating point numbers.
+All benchmarks are calculated as an average of 5 runs, done in
+parallel, on an 11-core, 64-bit Intel Xeon machine running at 2.40GHz
+with 11Gb of memory.
 
@@
 the baseline, the optimized analyzer outperformed the baseline by a
-factor of 475 to 4,382.
+factor of two to three orders of magnitude.
+% safer to not give exact factors
+% 475 to 4,382.
 
@@
 
-\item {\bf matrix} tests whether M is maximal among all matrices of
-  the same dimension obtainable by simple reordering of rows and
-  columns and negation of any subset of rows and columns.  It is
-  written in continuation-passing style (used in
+\item {\bf matrix} tests whether a matrix is maximal among all
+  matrices of the same dimension obtainable by simple reordering of
+  rows and columns and negation of any subset of rows and columns.  It
+  is written in continuation-passing style (used in
   \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
@@
 
-\item {\bf earley}:  Earley's parser written by Marc Feeley.
+\item {\bf earley}: Earley's parsing algorithm, applied to a 15-symbol
+  input according to a simple ambiguous grammar.  A real program,
+  applied to small data whose exponential behavior leads to a peak
+  heap size of half a gigabyte or more during concrete execution.
 
@@
 
-\item {\bf lattice}: enumerates the maps between two lattices (used in
+\item {\bf lattice}: enumerates the order-preserving maps between two
+  finite lattices (used in
   \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
@@
   distinguished root and \(k\) vertices, each having out-degree at
-  most 2. It makes extensive use of higher-order procedures and is
-  written in a continuation-passing style (used by
+  most 2. It is written in a continuation-passing style and makes
+  extensive use of higher-order procedures---it creates closures
+  almost as often as it performs non-tail procedure calls--- (used by
   \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
@@
 
-The analysis considered here computes results more like Earl, et al.'s
-0-CFA, which times out on the \Church{} benchmark because it does not
-widen the store as described for our baseline evaluator.  So even
-though it offers a fair point of comparison, a more thorough
-evaluation is probably uninformative as the other benchmarks are
-likely to timeout as well (and it would require significant effort to
-extend their implementation with the features needed to analyze our
-benchmark suite).  That implementation is evaluated against much
-smaller benchmarks: the largest program is 30 lines.
+The analysis considered here computes results similar Earl, et al.'s
+0-CFA implementation, which times out on the \Church{} benchmark
+because it does not widen the store as described for our baseline
+evaluator.  So even though it offers a fair point of comparison, a
+more thorough evaluation is probably uninformative as the other
+benchmarks are likely to timeout as well (and it would require
+significant effort to extend their implementation with the features
+needed to analyze our benchmark suite).  That implementation is
+evaluated against much smaller benchmarks: the largest program is 30
+lines.
 

### 2012-11-14 paper/paper.tex
Small fixes.

@@
 The AAM approach stops at the previous optimization. 
-However, the $\kfn{}$ continuation stores a value, and this makes the space of
+However, the {\tt fn} continuation stores a value, and this makes the space of
 continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0CFA.
@@
 \end{align}
-
-where $\msto_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
-[\maddr_2 \mapsto \kar{\mexp_3,\menv,\maddr_1}]
-[\maddr_3 \mapsto \kar{\mexp_2,\menv,\maddr_2}]$.
+where \(
+\msto_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
+[\maddr_2 \mapsto \kar{\mexp_3,\menv,\maddr_1}],
+[\maddr_3 \mapsto \kar{\mexp_2,\menv,\maddr_2}]\).
 
@@
 \begin{figure}
+\begin{gather*}
 \begin{align*}
 \mathit{eval}(\mexp) &= \{ \mstate\ |\ \compile{\mexp}(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
+\end{align*}
 \\[2mm]
+\begin{align*}
 %% CONTINUE
@@
 \end{align*}
-\caption{Abstract$^2$ machine for compiled ISWIM}
+\end{gather*}
+\caption{Abstract abstract machine for compiled ISWIM}
 \label{fig:caam}
@@
 \end{figure}
-
-\paragraph{Mutable literal data}
-The Scheme we model is nonstandard in that it has immutable literal
-data (the Scheme benchmarks did not mutate their literal data). To
-support mutable literal data, abstract values like {\tt QPair} would
-be like regular conses, only point to a preallocated pair of addresses
-that hold {\tt QData} as well. This is because conses can be mutated
-to hold any value, not just quoted data.
 

### 2012-11-14 paper/paper.tex
Abstract tweaks and wordsmithing

@@
 \begin{abstract}
-Abstracting abstract machines has recently been proposed as a
+Abstracting abstract machines has been proposed as a
 lightweight approach to designing sound and computable program
-analyses.  The approach involves deriving abstract interpreters from
+analyses.  The approach derives abstract interpreters from
 existing machine semantics and has been applied to a variety of
 languages with features widely considered difficult to analyze.
-Although analyzers are straightforward to build under this approach,
-the result is prohibitively inefficient.
-
-This article contributes a step by step process for going from a
-\naive analyzer derived under the abstracting abstract machine
-approach, to an efficient program analyzer.  The end result of the
+Although sound analyzers are straightforward to build under this approach,
+they are also prohibitively inefficient.
+
+This article contributes a step-by-step process for going from a
+\naive{} analyzer derived under the abstracting abstract machine
+approach to an efficient program analyzer.  The end result of the
 process is a two to three order-of-magnitude improvement over the

### 2012-11-14 paper/paper.tex
Played around with the intro

@@
 
-\newcommand{\naive}{na\"ive}
-\newcommand{\naively}{ na\"ively}
-\newcommand{\Naive}{Na\"ive}
-\newcommand{\Naively}{Na\"ively}
+\newcommand{\naive}{naive}
+\newcommand{\naively}{naively}
+\newcommand{\Naive}{Naive}
+\newcommand{\Naively}{Naively}
 % \usepackage{xltxtra}
@@
 approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
-to deriving program analyses provides a straightforward way of
-transforming a programming language semantics in the form of an
-abstract machine, into a family of abstract interpreters parameterized
-over policies for regulating the analytic precision of the control,
-environment, store, and base value domains.  By taking a
-machine-oriented view of computation, it becomes possible to design,
+to deriving program analyses provides a systematic way of transforming a
+programming language semantics in the form of an abstract machine into a family
+of abstract interpreters.
+%
+The approach parameterizes these families with policies for regulating analytic
+precision.
+%
+While flexible and robust, the AAM approach unfortunately yields analyzers with
+poor performance relative to hand-optimized analyzers.
+%
+Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an
+equally systematic way.
+
+%
+% Specifically, these policies tune precision for control,
+% the environment, the store, and base value domains.
+
+By taking a
+machine-oriented view of computation, AAM makes it possible to design,
 verify, and implement program analyzers for realistic language

### 2012-11-14 paper/paper.tex
A second glance at a glance

@@
 analyzer for a realistic untyped, higher-order imperative language
-with a number of interesting features.
-
-At each step during the initial development, we evaluate the
+with a number of interesting features and then measure
+improvements across a suite of benchmarks.
+
+At each step during the initial presentation and development, we evaluate the
 implementation on a benchmark from Vardoulakis and
@@
 %% \end{alltt}}
-The example, while small, is informative:
+For the step-by-step development, 
+this benchmark is particularly informative:
 \begin{enumerate}

### 2012-11-14 paper/paper.tex
Fixed the machine specs

@@
 All benchmarks are calculated as an average of 5 runs, done in
-parallel, on an 11-core, 64-bit Intel Xeon machine running at 2.40GHz
-with 11Gb of memory.
+parallel, on an 12-core, 64-bit Intel Xeon machine running at 2.40GHz
+with 12Gb of memory.
 

### 2012-11-14 paper/paper.tex
Machine components, work on initial analyzer.

--- 2012-11-14 paper/paper.tex
@@
 approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
-to deriving program analyses provides a systematic way of transforming a
-programming language semantics in the form of an abstract machine into a family
-of abstract interpreters.
-%
-The approach parameterizes these families with policies for regulating analytic
-precision.
-%
-While flexible and robust, the AAM approach unfortunately yields analyzers with
-poor performance relative to hand-optimized analyzers.
-%
-Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an
-equally systematic way.
-
-%
-% Specifically, these policies tune precision for control,
-% the environment, the store, and base value domains.
-
-By taking a
-machine-oriented view of computation, AAM makes it possible to design,
+to deriving program analyses provides a straightforward way of
+transforming a programming language semantics in the form of an
+abstract machine, into a family of abstract interpreters parameterized
+over policies for regulating the analytic precision of the control,
+environment, store, and base value domains.  By taking a
+machine-oriented view of computation, it becomes possible to design,
 verify, and implement program analyzers for realistic language
@@
 
-%% Using this approach has also lead to new insights in pushdown
-%% abstractions of programs capable of precisely modeling the program
-%% stack, even in the presence of control effects.
-
-%% \subsection{Notation and prerequisites}
-
-%% Prereqs: Semantics Engineering \cite{dvanhorn:Felleisen2009Semantics},
-%% AAM \cite{dvanhorn:VanHorn2011Abstracting}
-%% \cite{dvanhorn:VanHorn2012Systematic}.
-
-%% Notation: concrete examples of programs under analysis is given in
-%% (monochrome) Scheme notation.  Code is given in (syntax colored)
-%% Racket.
-
 \section{At a glance}
@@
 of multiplication over addition on Church numerals.
-%% \footnote{It is
-%%   written here using Scheme notation and makes use of the shorthands:
-%% %
-%% \begin{alltt}
-%%    (define (((f x) y) z) ...) \(\equiv\)
-%%
-%%    (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
-%% \end{alltt}
-%%
-%% \begin{alltt}
-%%    \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
-%% \end{alltt}}
 The example, while small, is informative:
@@
 
-%% \newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
-%% \begin{figure}
-%% \hrule
-%% \vspace{2mm}
-%% \begin{alltt}
-%% (define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
-%% (define (((mult m1) m2) mf) (m2 (m1 mf)))
-%% (define ((((plus p1) p2) pf) x)
-%%   ((p1 pf) ((p2 pf) x)))
-
-%% (define (((pred n) rf) rx)
-%%   (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
-%%     (\(\lambda\) (ignored) rx))
-%%    (\(\lambda\) (id) id)))
-
-%% (define ((church=? e1) e2)
-%%   (if (zero? e1) (zero? e2)
-%%       (if (zero? e2) #f
-%%           ((church=? (pred e1)) (pred e2)))))
-
-%% ;; multiplication distributes over addition
-%% ((church=? ((mult \church2) ((plus \church1) \church3)))
-%%  ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
-%% \end{alltt}
-%% \caption{Church-numeral calculation}
-%% \label{fig:church}
-%% \end{figure}
-
 \begin{figure}[t]
-%% \hrule
-%% \vspace{2mm}
-%% \begin{alltt}
-%% (define (id x) x)
-%% (define (f n)
-%%   (if (<= n 1) 1 (* n (f (- n 1)))))
-%% (define (g n)
-%%   (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
-%% (print (+ ((id f) 3) ((id g) 4)))
-%% \end{alltt}
-%% \hrule
 \begin{center}
@@
 %\newpage
-\section{An abstract$^{\text 2}$ machine for ISWIM}
+\section{An analyzer for ISWIM}
 \label{sec:aam}
+
+In this section, we give a brief review of the AAM approach by
+defining a sound analytic framework for a core higher-order functional
+language: Landin's ISWIM~\cite{dvanhorn:Landin1966Next}.  In the
+subsequent sections, we will explore optimizations for the analyzer in
+this simplified setting, but scaling these techniques to realistic
+languages is straightforward and has been done for the analyzer
+evaluated in section~\ref{sec:eval}.
 
@@
 operations.
+%
+Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.  It
+includes variables, literals (either integers, booleans, or
+operations), $\lambda$-expressions for defining procedures, procedure
+applications, and conditionals.  Expressions carry a label, $\mlab$,
+which is drawn from an unspecified set and denotes the source location
+of the expression; labels are used to disambiguate distinct, but
+syntactically identical pieces of syntax.  We omit the label
+annotation in contexts where it is irrelevant.
 
@@
 \begin{array}{l@{\qquad}rcl}
-\text{Expressions} & \mathit{e} &=& \svar[^\mlab]\mvar\\
+\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
 &&|& \slit[^\mlab]\mlit\\
@@
 
-Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.
+\begin{figure}
+\[
+\begin{array}{l@{\qquad}rcl}
+\text{Values} & \mval   &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\\
+\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
+                       &&|& \co{\mkont,\mval,\msto}\\
+                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
+\text{Continuations} & \mkont &=& \kmt\\
+&&|& \kfn{\mval,\maddr}\\
+&&|& \kar{\mexp,\menv,\maddr}\\
+&&|& \kif{\mexp,\mexp,\menv,\maddr}
+\end{array}
+\]
+\caption{Abstract machine components}
+\end{figure}
+
 Figure~\ref{fig:aam} defines the semantics of ISWIM as a machine

### 2012-11-14 paper/paper.tex

@@
 of multiplication over addition on Church numerals.
-The example, while small, is informative:
+For the step-by-step development, 
+this benchmark is particularly informative:
 \begin{enumerate}

### 2012-11-14 paper/paper.tex
Overhaul of AAM review.

@@
 \begin{array}{l@{\qquad}rcl}
-\text{Values} & \mval   &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\\
+\text{Values} & \mval   &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
 \text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
@@
 &&|& \kar{\mexp,\menv,\maddr}\\
-&&|& \kif{\mexp,\mexp,\menv,\maddr}
+&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
+\text{Environments} &\menv&\in& \mathit{Var} \rightharpoonup \mathit{Addr}\\
+\text{Stores} &\msto&\in& \mathit{Addr} \rightharpoonup \mathcal{P}(\mathit{Val})
 \end{array}
@@
 
-Figure~\ref{fig:aam} defines the semantics of ISWIM as a machine
-model.  Evaluation is defined as the set of states reachable by the
+The semantics are defined in terms of a machine model.  The machine
+compoents are defined in figure~\ref{fig:domains} and
+figure~\ref{fig:aam} defines the transition relation.  The evaluation
+of a program is defined as the set of states reachable by the
 reflexive, transitive closure of the machine transition relation.  The
 machine is a very slight variation on a standard abstract machine for
-ISWIM in ``eval, continue, apply''
-form~\cite{dvanhorn:Danvy:DSc}.  It can be
-systematically derived from a definitional interpreter through a
-continuation-passing style transformation and defunctionalization, or
-from a structural operational semantics using the refocusing
+ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
+can be systematically derived from a definitional interpreter through
+a continuation-passing style transformation and defunctionalization,
+or from a structural operational semantics using the refocusing
 construction of Danvy and
@@
 Compared with the standard machine semantics, this definition is
-different in the following ways:
+different in the following ways, which make it abstractable as a
+program analyzer:
 \begin{itemize}
-\item the store maps addresses to \emph{sets} of values,
-\item continuations are heap-allocated,
+\item the store maps addresses to \emph{sets} of values, not
+  single values,
+\item continuations are heap-allocated, not stack-allocated,
 \item there are ``contour values'' (written $\mcntr$) and syntax
@@
 \item the machine is implicitly parameterized by the functions
-  $\mathit{push}$, $\mathit{bind}$, $\interpdelta$ and $\sqcup$.
+  $\mathit{push}$, $\mathit{bind}$, and $\interpdelta$.
 \end{itemize}
@@
 
-\paragraph{Concrete interpretation} can be characterized by setting the implicit
-parameters of the relation given in Figure~\ref{fig:aam} as follows:
+\paragraph{Concrete interpretation} To characterize concrete interpretration, set the implicit
+parameters of the relation given in figure~\ref{fig:aam} as follows:
 \begin{align*}
@@
 \\
-\mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
+\mathit{bind}_\mlab^\mcntr(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
 \mbox{ where }\maddr \notin\msto
@@
 
-\paragraph{Abstract interpretation} can be characterized by setting the implicit
-parameters just as above, but dropping the $\maddr \not\in \msto$
+The interpretation of primitive operations is defined by setting
+$\interpdelta$ as follows:
+\begin{align*}
+\mnum+1 &\in \interpdelta(\saddone,\mnum) &
+\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
+\strue &\in \interpdelta(\szerohuh,\szero) &
+\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
+\end{align*}
+
+\paragraph{Abstract interpretation} To characterize abstract interpretation, set the implicit
+parameters just as above, but drop the $\maddr \not\in \msto$
 condition.  This family of interpreters is also non-deterministic in
@@
 analyses lay somewhere in between and can be characterized by their
-choice of address representation and allocation strategy.
-
-%% We now have a framework for describing program analysis for the ISWIM
-%% family of languages, whereby approximation of both control and
-%% environment structure is regulated by the heap and allocation
-%% policies.
-
-Uniform \(k\)-CFA is one such analysis.
-
-\paragraph{Uniform \(k\)-CFA} can be characterized by the following allocation
-strategy:
+choice of address representation and allocation strategy.  Uniform
+\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97} is one such analysis.
+
+\paragraph{Uniform \(k\)-CFA} To characterize uniform \(k\)-CFA, set the allocation
+strategy as follows, for a fixed constant \(k\):
 
@@
 \end{align*}
-$\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
-For the duration of this paper, since our example abstract value space
-has no interesting ordering, $\sqcup$ can be read as $\cup$.
-\paragraph{Primitives}
-
-\begin{align*}
-\mnum+1 &\in \interpdelta(\saddone,\mnum) &
-\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
-\strue &\in \interpdelta(\szerohuh,\szero) &
-\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
-\end{align*}
-
-\begin{align*}
-\sNum &\in \hat\interpdelta(\saddone,\mnum) &
-\sNum &\in \hat\interpdelta(\ssubone,\mnum)\\
-\strue &\in \hat\interpdelta(\szerohuh,\sNum) &
-\sfalse &\in \hat\interpdelta(\szerohuh,\sNum)\\
-\strue &\in \hat\interpdelta(\szerohuh,\szero) &
-\sfalse &\in \hat\interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
-\end{align*}
-
-%% \begin{figure}
-%% \begin{alltt}
-%%   ;; State \(\to\) \(\Set(\)State\()\)
-%%   (define (step \(\mstate\))
-%%     (match \(\mstate\)
-%%       [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
-%%        (match \(\mexp\)
-%%          [(var\(\superscript\mlab\) \(\mvar\))
-%%           (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
-%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
-%%          [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
-%%          [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
-%%          [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
-%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
-%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
-%%          [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
-%%           (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
-%%           (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
-%%       [(co \(\msto\) \(\mkont\) \(\mval\))
-%%        (match \(\mkont\)
-%%          ['mt (set (ans \(\msto\) \(\mval\)))]
-%%          [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
-%%          [(fn\(\superscript\mlab\) \(\mval'\))
-%%           (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
-%%             (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
-%%          [(fi\(\superscript\mlab\) c a \(\rho\))
-%%           (for/set ((k (get-cont \(\msto\) l)))
-%%             (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
-%%       [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
-%%        (match fun
-%%          [(clos l \(\mvar\) \(\mexp\) \(\rho\))
-%%           (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
-%%           (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
-%%          [(op \(\mop\))
-%%           (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
-%%                      (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
-%%             (co \(\msto\) \(\mkont\) \(\mval\)))]
-%%          [_ (set)]))]))
-%% \end{alltt}
-%% \caption{Implementation of machine transition relation.}
-%% \end{figure}
+where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
+
+All that remains is the interpretation of primitives.  For abstract
+intepretation, we set $\interpdelta$ to the function that returns
+$\sNum$ on all inputs---a symbolic value we interpret as denote the
+set of all integers.
+
+At this point, we have abstracted the original machine to one which
+has a finite state space for any given program, and thus forms the
+basis of a sound, computable program analyzer for ISWIM.
 

### 2012-11-14 paper/paper.tex

@@
 \text{Environments} &\menv&\in& \mathit{Var} \rightharpoonup \mathit{Addr}\\
-\text{Stores} &\msto&\in& \mathit{Addr} \rightharpoonup \mathcal{P}(\mathit{Val})
+\text{Stores} &\msto&\in& \mathit{Addr} \rightharpoonup \mathcal{P}(\mathit{Value})
 \end{array}
@@
 \caption{Abstract machine components}
+\label{fig:domains}
 \end{figure}
@@
 \end{align*}
-where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
+where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
+\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
+The \(\rfloor\cdot\lfloor_k\) notation denotes the truncation of a list
+of symbols to the leftmost \(k\) symbols.
 

### 2012-11-14 paper/paper.tex
Typo.

@@
 intepretation, we set $\interpdelta$ to the function that returns
-$\sNum$ on all inputs---a symbolic value we interpret as denote the
+$\sNum$ on all inputs---a symbolic value we interpret as denoting the
 set of all integers.

### 2012-11-14 paper/paper.tex
Getting figures in order.

@@
 
+We use the following set of benchmarks:
 \begin{figure}
@@
 
-Our implementation of the analysis used in the evaluation is roughly
-3500 lines of Racket code.
 %% 400 lines for the core, 700 lines for
@@
 
+Figure~\ref{fig:bench-overview} gives an overview of the benchmark
+results in terms of absolute time, space, and speed between the
+baseline and most optimized analyzer.  Figure~\ref{fig:bench-all}
+plots the factors of improvement over the baseline for each
+optimization step.
+
 \begin{figure*}
 \begin{center}
-\includegraphics[width=6in]{rel-time.ps}
+  \includegraphics[width=6.5in]{all-relative-memory.ps}
+
+  (a) Peak memory usage
+
+  \vspace{1em}
+  \includegraphics[width=6.5in]{all-relative-speed.ps}
+
+  (b) Rate of state transitions
+
+  \vspace{1em}
+  \includegraphics[width=6.5in]{all-relative-time.ps}
+
+  (c) Total analysis time
 \end{center}
+\caption{Factors of improvement over baseline for each step of
+  optimization.}
+\label{fig:bench-all}
 \end{figure*}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3.2in]{abs-time.ps}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3.2in]{state-per-sec.ps}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=3.2in]{peak-mem.ps}
-\end{center}
-\end{figure}
 

### 2012-11-14 paper/paper.tex
Yes it is David: on this planet, bigger is better.
Added comments on what might be unsettling at first glance about the graphs.

@@
 plots the factors of improvement over the baseline for each
-optimization step.
+optimization step. The dip we see in transition rate even though time
+taken decreases is to be expected - fewer ``easy'' states are added by
+abstract compilation. It increases again with the introduced
+algorithmic improvements. Accumulating store changes in addition to
+maintaining the store accounts for the higher memory usage when using
+the store delta technique without further improvements.
 
@@
 \caption{Factors of improvement over baseline for each step of
-  optimization.}
+  optimization (bigger is better).}
 \label{fig:bench-all}

### 2012-11-14 paper/paper.tex
10 pages.

--- 2012-11-14 paper/paper.tex
@@
 
-\title{Optimizing Abstract Abstract Machines}
+\title{Optimizing Abstract Abstract Machines
+\vskip -1.5cm}
 
@@
 exploration: (1) start with an initial state, and (2) compute the transitive
-closure of the transition relation starting from that state.
+closure of the transition relation from that state.
 
@@
 
+\noindent
 We have two choices for how to implement lazy non-determinism.
@@
 
-We will explore both options here.
+We explore both options here.
 %
@@
   extensive use of higher-order procedures---it creates closures
-  almost as often as it performs non-tail procedure calls--- (used by
+  almost as often as it performs non-tail procedure calls (used by
   \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
@@
   is not a coincidence; these papers set a high standard for
-  evaluation, which we consciously aimed to approach.  Many of our
-  benchmarks are drawn from these papers.}  Unfortunately, these
+  evaluation, which we consciously aimed to approach.}  Unfortunately, these
 analyses compute an inherently different and uncomparable form of
@@
 typically compute sets of abstract values for each program point.
-These values approximate the values arising at run-time for each
+These values approximate values arising at run-time for each
 program point.  Value sets are computed as the least solution to a set
-of constraints (either inclusion or equality constraints).  The
+of (inclusion or equality) constraints.  The
 constraints must be designed and proved as a sound approximation of
@@
 
-The AAM approach, and the state transition graphs it generates, encode
+The AAM approach, and the state transition graphs it generates, encodes
 temporal properties not found in classical constraint-based analyses
@@
 Abstract machines are not only a good model for rapid analysis
-development, they also can be systematically developed into efficient
+development, they can be systematically developed into efficient
 algorithms.
@@
 \bibliographystyle{plain}
-\bibliography{local,bibliography}
+\input{paper.bbl}
+%\bibliography{local,bibliography}
 

### 2012-11-14 paper/paper.tex
Unanon.

@@
 
-\title{Optimizing Abstract Abstract Machines
-\vskip -1.5cm}
-
-%% \authorinfo{J. Ian Johnson}
-%%            {Northeastern University}
-%%            {ianj@ccs.neu.edu}
-%% \authorinfo{Matthew Might}
-%%            {University of Utah}
-%%            {might@cs.utah.edu}
-%% \authorinfo{David Van Horn}
-%%            {Northeastern University}
-%%            {dvanhorn@ccs.neu.edu}
-\authorinfo{}{}{}
+\title{Optimizing Abstract Abstract Machines}
+
+\authorinfo{J. Ian Johnson}
+           {Northeastern University}
+           {ianj@ccs.neu.edu}
+\authorinfo{Matthew Might}
+           {University of Utah}
+           {might@cs.utah.edu}
+\authorinfo{David Van Horn}
+           {Northeastern University}
+           {dvanhorn@ccs.neu.edu}
 
@@
 The analysis considered here computes results similar Earl, et al.'s
-0-CFA implementation, which times out on the \Church{} benchmark
-because it does not widen the store as described for our baseline
-evaluator.  So even though it offers a fair point of comparison, a
-more thorough evaluation is probably uninformative as the other
-benchmarks are likely to timeout as well (and it would require
-significant effort to extend their implementation with the features
-needed to analyze our benchmark suite).  That implementation is
-evaluated against much smaller benchmarks: the largest program is 30
-lines.
+0-CFA implementation~\cite{dvanhorn:Earl2012Introspective}, which
+times out on the \Church{} benchmark because it does not widen the
+store as described for our baseline evaluator.  So even though it
+offers a fair point of comparison, a more thorough evaluation is
+probably uninformative as the other benchmarks are likely to timeout
+as well (and it would require significant effort to extend their
+implementation with the features needed to analyze our benchmark
+suite).  That implementation is evaluated against much smaller
+benchmarks: the largest program is 30 lines.
 

### 2012-11-15 paper/paper.tex
Use PDF for figures.  Better make.

@@
 \begin{center}
-\includegraphics[width=3.2in]{church-relative-space.ps}
-\includegraphics[width=3.2in]{church-relative-speed.ps}
-\includegraphics[width=3.2in]{church-relative-time.ps}
+\includegraphics[width=3.2in]{church-relative-space}
+\includegraphics[width=3.2in]{church-relative-speed}
+\includegraphics[width=3.2in]{church-relative-time}
 \vspace{-1.5em}
@@
 \begin{center}
-  \includegraphics[width=6.5in]{all-relative-memory.ps}
+  \includegraphics[width=6.5in]{all-relative-space}
 
@@
   \vspace{1em}
-  \includegraphics[width=6.5in]{all-relative-speed.ps}
+  \includegraphics[width=6.5in]{all-relative-speed}
 
@@
   \vspace{1em}
-  \includegraphics[width=6.5in]{all-relative-time.ps}
+  \includegraphics[width=6.5in]{all-relative-time}
 

### 2012-11-15 paper/paper.tex
Link to implementation.
ispell.

@@
   subgraphs.  Part (c) shows the result of abstract compilation that
-  removes intepretive overhead in the form of intermediate states,
+  removes interpretive overhead in the form of intermediate states,
   thus minimizing the corridor transitions.  The end result is a more
@@
 
-Section~\ref{sec:eval} describes the evaluation of each optimiztion
+Section~\ref{sec:eval} describes the evaluation of each optimization
 technique applied to an implementation supporting a more realistic set
@@
 The semantics are defined in terms of a machine model.  The machine
-compoents are defined in figure~\ref{fig:domains} and
+components are defined in figure~\ref{fig:domains} and
 figure~\ref{fig:aam} defines the transition relation.  The evaluation
@@
 
-\paragraph{Concrete interpretation} To characterize concrete interpretration, set the implicit
+\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
 parameters of the relation given in figure~\ref{fig:aam} as follows:
@@
 All that remains is the interpretation of primitives.  For abstract
-intepretation, we set $\interpdelta$ to the function that returns
+interpretation, we set $\interpdelta$ to the function that returns
 $\sNum$ on all inputs---a symbolic value we interpret as denoting the
@@
 \end{equation*}
-Internally, this global transfer function computes the successors of all supplied states, and then includes the inital state:
+Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
 \begin{equation*}
@@
 %
-This widening effictively lifts the store out of individual states to create
+This widening effectively lifts the store out of individual states to create
 a single, global shared store for all.
@@
 %
-Thus, in one step, complexity drops from intractable exponentiality to merely
-daunting polynomiality.
+Thus, in one step, complexity drops from intractable exponentially to a merely
+daunting polynomial.
 
@@
 nature of laziness in these semantics. Due to the increase of
-conceptual complexity for negligable benefit, we decided against this
+conceptual complexity for negligible benefit, we decided against this
 approach.
@@
 Boyer's fluent use of literal list dooms the ``natural'' abstraction
-of lists to fail from overprecision.
+of lists to fail from over precision.
 %
 In short, if we interpret a literal list {\tt '(a b c)} as {\tt (cons 'a (cons 'b
-(cons 'c '())))}, even a monovariant allocation stategy for abstract
+(cons 'c '())))}, even a monovariant allocation strategy for abstract
 cons cells will precisely and exactly create a 3-cell list.
@@
 store, and return the flattened {\tt (cons $a$ $d$)} value. Since
-these addresses are all distinguished at different syntactic callsites
+these addresses are all distinguished at different syntactic call sites
 in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
@@
 
+Source code of the implementation and benchmark suite is at:
+
+\begin{center}
+\url{https://github.com/dvanhorn/oaam}
+\end{center}
+
 \paragraph{Comparison with other flow analysis implementations}
@@
   evaluation, which we consciously aimed to approach.}  Unfortunately, these
-analyses compute an inherently different and uncomparable form of
+analyses compute an inherently different and incomparable form of
 analysis.  Consequently, we have omitted a complete comparison with

### 2012-11-20 code/bench/out.sh
Adding the script for distilling benchmark results.

@@ -1,0 +1,17 @@
+#!/bin/bash
+benchmark="../../paper/benchmark"
+echo " " > $benchmark
+# Make sure this is ((i = base-num; i < (base-num + run-num); i+=1))
+# from ../drive-benchmarks.rkt
+for ((i = 0; i < 5; i+=1))
+do
+	grep cpu `find . -name "*.time.$i" -print` >> $benchmark
+	grep "State count" `find . -name "*.time.$i" -print` >> $benchmark
+	grep "Point count" `find . -name "*.time.$i" -print` >> $benchmark
+	grep "States/second" `find . -name "*.time.$i" -print` >> $benchmark
+	grep Timeout `find . -name "*.time.$i" -print` >> $benchmark
+	grep Exhaust `find . -name "*.time.$i" -print` >> $benchmark
+	grep Peak `find . -name "*.mem.$i" -print` >> $benchmark
+	grep Current `find . -name "*.mem.$i" -print` >> $benchmark
+done
+

### 2012-11-26 paper/paper.tex
Drop .pdf from includegraphics.
Fixed makefile to build final visualizations, renamed memory to speed for consistency.
Implemented some of Mitch's comments.

@@
 allocating a continuation or variable binding.  If we consider machine
-states equivalent up to consistent renaming, this relation defines
-a deterministic machine.  (The relation is really a function.)
+states equivalent up to consistent renaming and fix an allocation
+scheme, this relation defines a deterministic machine (the relation is really a function).
 
@@
 \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
-The \(\rfloor\cdot\lfloor_k\) notation denotes the truncation of a list
+The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
 of symbols to the leftmost \(k\) symbols.
@@
 (C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
-\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c, \msto^c), c \in C\} \\
+\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
               \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
@@
 %
-In effect, the new store is computed as the least upper bound of all subsequent stores.
+In effect, the new store is computed as the least upper bound of all stores after stepping.
 
@@
 languages with $n$-ary functions, since otherwise the continuation
-space grows super-exponentially.
+space grows super-exponentially. In our example language, we have only
+unary functions, so application forms always have two positions: the
+function and the argument. Thus for every application labeled $\mlab$,
+we introduce two new labels, $\mlab^f$ and $\mlab^a$.
 % In theory, this aggressive distinction between continuations might buy
@@
 
-If we specialize to 0CFA, the evaluation rules become:
+In $k$-CFA, the evaluation rules become:
 
@@
 %ext(\msto, \maddr, s) &= \msto \sqcup [\maddr \mapsto s] \\
-\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
-\ev{\mexpi{0},\menv,\ext{\msto}{\mlab}{\{\mkont\}},\kar{\mexpi{1},\mlab}}
-\\
-\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
-\ev{\mexp,\menv,\ext{\msto}{\mlab^f}{\{\mval\}},\kfn{\mlab^f,\mlab}}
-\\
-\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
-\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\ext{\msto}{\mlab^a}{\{\mval\}}}
-\\
-\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
-\\
-\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
-\ev{\mexp,\menv[\mvar\mapsto\mvar],\ext{\msto}{\mvar}{\msto(\mlab^a)},\mkont}
-\\
-\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
+\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
+\ev[^\mcntr]{\mexpi{0},\menv,\ext{\msto}{\mlab}{\{\mkont\}},\kar{\mexpi{1},\menv,\mlab,\mcntr}}
+\\
+\co{\kar{\mexp,\menv,\mlab,\mcntr},\mval,\msto} & \machstep
+\ev[^\mcntr]{\mexp,\menv,\ext{\msto}{\mlab^f\mcntr}{\{\mval\}},\kfn{\mlab^f,\mlab,\mcntr}}
+\\
+\co{\kfn{\mlab^f,\mlab,\mcntr},\mval,\msto} & \machstep
+\ap[^\mcntr_\mlab]{\mvalx{u},\mlab^a,\mkont,\ext{\msto}{\mlab^a\mcntr}{\{\mval\}}}
+\\
+\text{ where } &\mkont \in \msto(\mlab\mcntr), \mvalx{u} \in \msto(\mlab^f\mcntr)
+\\
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
+\ev[^{\mcntr'}]{\mexp,\menv[\mvar\mapsto\mvar\mcntr'],\ext{\msto}{\mvar\mcntr'}{\msto(\mlab^a\mcntr)},\mkont}
+\\
+\text{ where }& \mcntr' = \lfloor\mlab\mcntr\rfloor_k
+\\
+\ap[^\mcntr_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
 \co{\mkont,\mval',\msto}
 \\
-\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
+\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a\mcntr)\}
 \end{align*}
@@
 \begin{center}
-\includegraphics[scale=0.3]{fanout.pdf}
+\includegraphics[scale=0.3]{fanout}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.3]{lazy.pdf}
+\includegraphics[scale=0.3]{lazy}
 \end{center}
 
-In case it's unclear, this shift does not change the concrete semantics of the
-language to be lazy.  Rather, it abstracts over transitions that the original
+This shift does not change the concrete semantics of the language to
+be lazy.  Rather, it abstracts over transitions that the original
 non-deterministic semantics steps through.
@@
 Quotation is special because it cannot introduce function values,
-which is important to enhancing our technique's soundness to
-conceptual complexity ratio. If we join two values of different type
-together, we don't get ``anything,'' which has complex meaning in
-higher-order languages (Shivers' escape semantics) and is overly
-approximate. We instead get ``any quotable value,'' which has much
-simpler semantics.
+which is important for a simple semantics. If we join two values of
+different type together, we don't get ``anything,'' which has complex
+meaning in higher-order languages (Shivers' escape semantics) and is
+overly approximate. We instead get ``any quotable value,'' which has
+much simpler semantics.
 
@@
 
+This abstraction was so critical to timely convergence that all of our evaluation uses it.
+
 \section{Evaluation}
@@
 operations such as floating point, complex, and exact rational
-arithmetic.  The analysis is evaluated against a suite of benchmarks
+arithmetic.  The analysis is evaluated against a suite of Scheme benchmarks
 drawn from the literature.
@@
 analysis is stopped.  This is the case for the largest benchmark
-program, which is 3,500 lines of code and takes under a minute in the
+program, {\bf nucleic}, which is 3,500 lines of code and takes under a minute in the
 most optimized analyzer.  For those benchmarks that did complete on
@@
 
-\item {\bf nbody}: implements the Greengard multipole algorithm for
-  computing gravitational forces on point masses distributed uniformly
-  in a cube (used in
+\item {\bf nbody}: implementation~\cite{ianjohnson:nbody87} of the
+  Greengard multipole algorithm for computing gravitational forces on
+  point masses distributed uniformly in a cube (used in
   \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
@@
 
-The analysis considered here computes results similar Earl, et al.'s
+The analysis considered here computes results similar to Earl, et al.'s
 0-CFA implementation~\cite{dvanhorn:Earl2012Introspective}, which
@@
 Constraint-based program analyses
-(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
+(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular,dvanhorn:steckler-wand-toplas97})
 typically compute sets of abstract values for each program point.

### 2013-03-04 paper/precision.tex
Some notes on precision and temporal properties.

\documentclass{llncs}
\usepackage{amsmath,amssymb,pfsteps,multicol,stmaryrd,mathpartir,centernot}
\input{preamble}

\begin{document}

\section{Wide semantics decisions and precision differences}

There are a few straightforward ways to widen the state space to use a
single, monotonically increasing store. By far the most common way is
with a worklist algorithm for computing a least fixed point of a
``frontier-based'' collecting semantics. By dropping all the
intermediate stores, we lose temporal information such as $x$ is never
a number before $y$ is a string. This kind of information is
interesting, e.g., for diagnosing possible errors that the analysis
detects when $x$ is used in a call not defined for numbers. More
usefully, this can help pinpoint what influenced execution that lead
to a tainted value in an information flow analysis. Consider the
following program:

\begin{lstlisting}
(define (f x y)
  (if (bytes? y)
      (f (string->number x) (string-append (bytes->string/utf-8 y) x))
      (if (zero? x)
          y
          (f (sub1 x) (substring x)))))
(f (read-numeric) (read-encoded/unencoded))
\end{lstlisting}

There's a faux recursion to get the input into the right
format. Obviously this is contrived, but in a program that uses
polymorphic functions at different stages of its execution, the
temporal information helps pinpoint when and under what circumstances
an abstract address was poisoned with a value that causes an error.

\subsection{Precision on the frontier}
There is also a precision difference between the frontier-based method
and the state-space stepping semantics that is the initial model for a
widened store. That is, instead of separating states that need
stepping from seen states like the frontier-based method does, the
entire known state space steps with the current store:

\begin{align*}
\System &= \wp(\State) \times \Store \\
(S,\msto) &\machstep (S',\msto') \\
 \text{ where } I &= \setof{\mstate \mid \mastate \in S, \exists \msto'. \wn(\mastate,\msto) \machstep \mstate} \\
                S' &= \setof{\mastate \mid \exists \msto.\wn(\mastate,\msto) \in I} \\
                \msto' &= \bigsqcup\limits_{\msto \mid \exists\mastate.\wn(\mastate,\msto) \in I}{\msto}
\end{align*}
Although simple, this formulation has serious performance implications
and precision deficiencies compared to the frontier-based
semantics. This means the frontier-based algorithm is not just a more
efficient implementation strategy for a widened store semantics, its
results are not equivalent (they are better).

The reason for the difference is that store changes that would affect
the execution of some function might not be exercised before that
function may never be called after that change. The frontier semantics
would not end up stepping the state with the store dependency, whereas
the above semantics would.

A contrived example:
\begin{lstlisting}
(define b (box #f))
(define (foo x)
  (if (unbox b)
      (bar x)
      (bar (- x))))
(define (bar y) (/ (log y) (log 2))
(foo 20)
(set-box! b #t)
<code that doesn't invoke foo>
\end{lstlisting}

The frontier semantics would never find that {\tt bar} could cause an
exception for {\tt log} called on a non-positive number, whereas the
above semantics would, unnecessarily. Notice that not only are
bindings not poisoned, fewer paths are explored.

\end{document}

### 2013-03-15 paper/paper.tex
Starting rewrite.

@@
 
-\title{Optimizing Abstract Abstract Machines}
+\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
 
@@
 %
-In effect, the new store is computed as the least upper bound of all stores after stepping.
+In effect, the new store is computed as the least upper bound of all
+stores after stepping. This definition is wasteful, however. It causes
+all states found so far to step each iteration, even if they are not
+revisited. This has negative performance and precision
+consequences. We instead use a frontier-based semantics that
+corresponds to the classic ``worklist'' algorithms for analysis.
+
+\begin{align*}
+(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
+\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
+              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
+              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
+\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
+\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
+\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
+\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval} \\
+\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
+\text{ where } c_0 &= \ev{e,\bot,\kmt}
+\end{align*}
 
@@
 \begin{enumerate}
- \item lazy non-determinism;
- \item abstract compilation; and
- \item uniform literal approximation.
+ \item lazy non-determinism and
+ \item abstract compilation;
 \end{enumerate}
@@
 
-Some techniques preserve the precision of the underlying analysis, and others
-do not.
-%
-For any technique that loses precision, we will discuss the design rationale
+All but the first technique are precision preserving. We will discuss the design rationale
 for the move.
@@
 \ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto \force(\msto, \mval)] \\
-\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
+\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
 \co{\mkont,\saddr{\menv(\mvar)},\msto} \\
 \co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
-&\machstep\;
+&\lmachstep\;
 \ev[^\mcntr]{\mexp,\menv,\ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
@@
 baseline. Consider a configuration where a reference to a variable and
-a binding of a variable will happen in one step. With laziness, the
-reference will mean the original binding(s) of the variable or the new
-one, because the actual store lookup is delayed one step
+a binding of a variable will happen in one step, since store widening
+leads to stepping several states in one big ``step.'' With laziness,
+the reference will mean the original binding(s) of the variable or the
+new binding, because the actual store lookup is delayed one step
 (i.e. laziness is administrative). Without laziness, the reference
 will fan out to all the bindings of the variable before the new
-binding happens and thus might have an observable precision
-difference.
+binding happens and thus theoretically has an observable precision
+difference. % All our benchmarks maintain their precision with lazy non-determinism, however.
 
@@
 nature of laziness in these semantics. Due to the increase of
-conceptual complexity for negligible benefit, we decided against this
-approach.
+conceptual and implementation complexity for negligible benefit, we
+decided against this approach.
 
@@
 two approaches are the same.
+
+\begin{theorem}[Soundness]
+  If $\mstate \machstep \mstate'$ and $\mstate \sqsubseteq
+  \mastate$ then there exists a $\mastate'$ such that $\mastate
+  \lmachstep \mastate'$ and $\mstate' \sqsubseteq \mastate'$
+\end{theorem}
+Here $\sqsubseteq$ is straightforward - the LHS store must be
+contained in the RHS store, and if values occur in the states, forcing
+the LHS value must be a subset of forcing the corresponding RHS value.
+The proof is by cases on $\mstate \machstep \mstate'$.
 
@@
 %% CONTINUE
-\co{\kmt,\mval,\msto} &\machstep
+\co{\kmt,\mval,\msto} &\cmachstep
 \ans{\msto,\mval}
 \\
-\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \machstep
+\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \cmachstep
 \mcomp^\mcntr(\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr})
 \\
-\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
+\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \cmachstep
 \ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
@@
 \\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \machstep
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \cmachstep
 \mcompi0^\mcntr(\menv,\msto,\mkont)
@@
 \\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \machstep
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \cmachstep
 \mcompi1^\mcntr(\menv,\msto,\mkont)
@@
 %% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \machstep
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \cmachstep
 \mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
@@
 \\
-\ap{\mop,\mval,\msto,\mkont} & \machstep
+\ap{\mop,\mval,\msto,\mkont} & \cmachstep
 \co{\mkont,\mval',\msto} \\
@@
 
+\paragraph{Correctness}
+The correctness of abstract compilation seems obvious, but it has
+never before been rigorously proved. What constitutes correctness in
+the case of dropped states, anyway? Applying an abstract bytecode's
+function does many ``steps'' in one go, at the end of which, the two
+semantics line up again (modulo representation of expressions). This
+constitutes the use of a notion of stuttering. We therefore prove a
+stuttering bisimulation~\cite{clarke-stuttering} between this
+semantics and the previous section's to show precision preservation.
+
+The number of transitions that can occur in succession from an
+abstract bytecode is roughly bounded by the amount of expression
+nesting in the program. The partial order that defines subexpressions
+smaller than the expression in which they occur is well-founded, so we
+use this observation to drive our use of a well-founded equivalence
+bisimulation (WEB). WEBs are equivalent to the notion of a stuttering
+bisimulation, but are more amenable to mechanisation (and thus
+rigorous proof). They also only require reasoning over one step of the
+reduction relation, which makes the proof concise.
+
+We define a refinement from non-compiled to compiled states by
+``committing'' all the actions of an $\ev{}$ state (defined similarly to
+$\compile{\_}$, but immediately applies the functions), and
+subsequently changing all expressions with their compiled
+variants. Since WEBs are for single transition systems, a WEB
+refinement is over the disjoint union of our two semantics, and the
+equivalence relation we use is just that a state is related to its
+refined state (and itself). Call this relation $B$.
+
+Before we prove this setup is indeed a WEB, we need one lemma that
+applying an abstract bytecode's function is equal to refining the
+corresponding $\ev{}$ state:
+%
+\begin{lemma}[Compile/commit]
+For all $\mcntr,\mexp,\menv,\msto,\mkont$,
+$\compile{\mexp}^\mcntr(\menv,r(\msto),r(\mkont)) = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
+\end{lemma}
+The proof is by induction on $\mexp$.
+
+\begin{theorem}
+$B$ is a WEB on $\lmachstep \cup \cmachstep$
+\end{theorem}
+
+The full proof is in the appendix, but it follows by cases on
+$\lmachstep \cup \cmachstep$ with the WEB \emph{witness} being the
+well-order on expressions (with a $\bot$ element), and the following
+$\erankt$, $\erankl$ functions:
+
+\begin{align*}
+\erankt(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mexp \\
+\erankt(\mstate) &= \bot \quad \text{otherwise} \\
+\erankl(s,s') &= 0
+\end{align*}
+All cases are either simple steps or appeals to the well-order on $\erankt$'s range. The other rank function,
+$\erankl$ is unnecessary, so we just make it the constant 0
+function. The $\cmachstep$ cases are trivial.
+
 % Compile: cpu time: 255397 real time: 261532 gc time: 2947
@@
 Every step the analysis makes for the above techniques requires
-joining large stores together. Not every step will modify all
+joining entire (large) stores together. Not every step will modify all
 addresses of the store, so joining entire stores is wasteful in terms
@@
 \replay &: \StoreDelta \times \Store \to \Store \\
-\replay(\mtlst, \msto) &= \msto \\
-\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
-\end{align*}
+\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &= \msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots
+\end{align*}
+%\replay(\mtlst, \msto) &= \msto \\
+%\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
 
@@
 \begin{align*}
-(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \machstep
+(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \dmachstep
 (\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
@@
 
-We also lift $\machstep$ to accommodate for this asymmetry
+We also lift $\dmachstep$ to accommodate for this asymmetry
 in the input and output. For each state that is stepped, we feed the
@@
 \begin{align*}
-(\states, \msto) &\mathrel{\widehat{\machstep}} (\states\cup \states', \replay(\msdiff,\msto)) \\
-\mbox{ where } (\states', \msdiff) &= \step^*(\varnothing, \states, \mtlst) \\
-\step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
-\step^*(S, \{c\}\cup \states, \msdiff &= \step^*(S\cup cs^*, \states, \msdiff^*) \\
-\states^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
-\msdiff^* &= \concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
-\end{align*}
-
-Here $\concat : \Set(\StoreDelta) \to \StoreDelta$ flattens the lists
-of changes to one; the order in which it appends does not matter:
-\begin{align*}
-\concat(\varnothing) &= \mtlst \\
-\concat(\{\msdiff\}\cup D) &= \append(\msdiff, \concat(D))
-\end{align*}
+(S, F, \msto) &\mathrel{\damachstep} (S \cup S', F', \msto') \\
+\mbox{ where } (F', \msto') &= \step^*(\varnothing, F, \msto) \\
+               S' &= \setof{(c,\msto') \mid c \in F'} \\
+\step^*(F', \varnothing, \msto) &= (S, \msdiff) \\
+\step^*(F', \{c\}\cup F, \msto) &= \step^*(F'\cup \states^*, F, \msto^*) \\
+\states^* &= \{c' \mid (c,\msto,\msdiff) \dmachstep (c',\msdiff^c) \} \\
+\msto^* &= \replayall(\{\msdiff^c \mid (c, \msto,\msdiff) \dmachstep (c', \msdiff^c)\}, \msto)
+\end{align*}
+
+Here $\replayall : \Set(\StoreDelta) \times \Store \to \Store$ replays all logs in a set on top of the given store. Since join is associative and commutative, the order does not matter:
+\begin{align*}
+  \replayall(\varnothing,\msto) &= \msto \\
+  \replayall(\setof{\msdiff} \cup \Xi, \msto) &= \replayall(\Xi, \replay(\msdiff, \msto))
+\end{align*}
+
+Correctness here requires a lemma that relates the differences in the compilation functions.
+\begin{lemma}
+For all $\mexp, \mcntr,\menv,\msto,\msdiff,\mkont,\msto^*,\mkont^*$,
+if $\msto \equiv \msto^*$ and $\mkont \equiv \mkont^*$, then
+$\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$ iff
+$\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \mstate$ such that
+$\mastate,\replay(\msdiff',\msto) \equiv \nw(\mstate)$
+\end{lemma}
+Proof is by induction on $\mexp$.
+
+\begin{theorem}
+If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
+$(S,F,\msto) \machstep (S',F',\msto')$ iff
+$\exists S^*_1, F^*_1, \msto^*_1.
+  S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
+  (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
+\end{theorem}
+Here $\equiv$ means structurally the same, only compiled expressions
+are compiled with the updated compilation function that builds store
+deltas. The proof follows by equational reasoning and the previous lemma.
 
@@
 store or (if memory is no concern) the entire store.
-
-\subsection{Abstracting literal compound data}
-
-The abstraction of compound data structures like lists has deep implications 
-for precision and performance.
-%
-The classic list-heavy benchmark {\tt boyer}---a simplified implementation
-of the Boyer-Moore theorem-prover---drove our own considerations for
-abstractions for compound data structures.
-%
-Boyer's fluent use of literal list dooms the ``natural'' abstraction
-of lists to fail from over precision.
-%
-In short, if we interpret a literal list {\tt '(a b c)} as {\tt (cons 'a (cons 'b
-(cons 'c '())))}, even a monovariant allocation strategy for abstract
-cons cells will precisely and exactly create a 3-cell list.
-
-If code contains a length-300 literal list, then its analysis yields a length-300 abstract
-list as well.
-%
-In many cases, the specific contents of the list add little precision
-to the resulting analysis, and yet the analyzer will dutifully execute
-recursive functions over the entirety of these structures at every
-encounter.
-%
-An alternative in this case is to explicitly flatten literal lists
-into single abstract cells.
-
-We explore both options here.
-%
-We will first explain the natural abstraction of tuples, and then we
-will explain a less precise allocation strategy from uniform $k$-CFA
-that we use for large compound data literals that we modified from the
-implementation in \citep{dvanhorn:wright-jagannathan-toplas98}.
-
-\subsubsection{Option 1: The natural abstraction}
-
-The uniform way AAM approaches a simple abstraction strategy is to cut
-recursion out of the data definition by tying the recursive knot
-through the abstract store. For Scheme, the grammar for values looks
-like the following:
-
-\newcommand{\alt}{\;\mid\;}
-\begin{align*}
-\Value &::= \mbox{{\tt \#t}} \alt \mbox{{\tt \#f}} \alt \mbox{{\tt(cons $\Value$ $\Value$)}} \alt \mbox{{\tt '()}} \alt ...
-\end{align*}
-
-Upon evaluating a {\tt cons} application, we instead allocate two
-addresses $a$ and $d$, join them to the respective values in the
-store, and return the flattened {\tt (cons $a$ $d$)} value. Since
-these addresses are all distinguished at different syntactic call sites
-in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
-for a sequence of calls to {\tt cons}, this abstraction explodes the
-value space. Analyzing a function that counts the number of atoms in a
-literal s-expression would actually interpret that function at least
-that number of times (more because of intermediate conses). Indeed,
-even in our fastest implementation, this abstraction causes the
-analysis of Boyer to be 430 times slower than the approach we will now
-describe.
-
-\subsubsection{Option 2: A precise yet compact abstraction}
-
-The number of syntactic uses of {\tt cons} versus implicit uses via
-literal lists is smaller in typical Scheme programs. We use the above
-abstraction for these syntactic uses, but choose to interpret literal
-lists as not always sugar for cascading {\tt cons}es. In particular,
-if a list literal is ``too big'' (in our case length $>$ 1), we
-interpret the list as a circular data structure; the right address
-points back to the {\tt cons} itself, and the left address points to
-all of the elements of the list. We take this farther and join the
-list elements together in with a coarse type-based abstraction. In
-effect, large lists or vectors of literal numbers become unbounded
-lists/vectors of ``number.'' Heterogeneous data is combined to just
-``data,'' rather than a union of ``number'', ``string'', etc. since
-primitives are interpreted on every combination of values that flow to
-them, and unions lead to more primitive interpretations.
-
-Quotation is special because it cannot introduce function values,
-which is important for a simple semantics. If we join two values of
-different type together, we don't get ``anything,'' which has complex
-meaning in higher-order languages (Shivers' escape semantics) and is
-overly approximate. We instead get ``any quotable value,'' which has
-much simpler semantics.
-
-There are a few steps to consider:
-\begin{itemize}
- \item{Define special value lattice elements for compound data domains that can be quoted
-       (e.g. {\tt QPair} for $\left\{ ({\tt cons}\ a_i\ d_i)\right\}_i$, {\tt QVector} for immutable vectors, etc.)}
- \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
- \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
-       and similar definitions for immutable vectors.}
- \item{Extend the $\interpdelta$ axioms to include conservative
-   meaning for these new values (e.g. {\tt (car QData)} = {\tt QData},
-   {\tt (add1 QData)} = {\tt Number} and log ``possible type error'')
-   and allow them to allocate addresses and change the store}
-\end{itemize}
-
-\begin{figure}
-\begin{align*}
-\interpdelta(\msto,\ {\tt qlist}) &= (\mbox{{\tt '()}}, \msto) \\
-\interpdelta(\msto,\ {\tt qlist}, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
-\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
-                            &\qquad \sqcup[d \mapsto \{ \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\}] \\
-\bigsqcup(v) &= v \\
-\bigsqcup(v, vs ...) &= \fmerge(v, \bigsqcup(vs ...)) \\
-\fmerge(v, v) &= v \\
-\fmerge(n, m) &= {\tt Number} \\
-\fmerge(n, v) &= {\tt QData} \\
-\fmerge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
-\fmerge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
-\fmerge({\tt QPair}, v) &= {\tt QData} \\
-\vdots
-\end{align*}
-\caption{Quoted list primitive}
-\label{fig:qlist}
-\end{figure}
-
-This abstraction was so critical to timely convergence that all of our evaluation uses it.
 
@@
 
-%% Vincent St Amour for feedback on early drafts.
+%% Vincent St Amour and Mitchell Wand for feedback on early drafts.
 %% Greg Morrisset and Matthias Felleisen for discussions.

### 2013-03-18 paper/paper.tex
Finished draft of imperative section.

@@
 all states found so far to step each iteration, even if they are not
-revisited. This has negative performance and precision
+revisited. This has negative performance \emph{and} precision
 consequences. We instead use a frontier-based semantics that
-corresponds to the classic ``worklist'' algorithms for analysis.
+corresponds to the classic ``worklist'' algorithms for analysis. The
+difference is that the store is not modified in-place, but updated
+after all frontier states have been processed. This has implications
+for the analysis' precision and determinism. Specifically, higher
+precision, and is deterministic even if set iteration is not.
 
@@
 constitutes the use of a notion of stuttering. We therefore prove a
-stuttering bisimulation~\cite{clarke-stuttering} between this
+stuttering bisimulation~\cite{ianjohnson:BCG88} between this
 semantics and the previous section's to show precision preservation.
@@
 use this observation to drive our use of a well-founded equivalence
-bisimulation (WEB). WEBs are equivalent to the notion of a stuttering
+bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are equivalent to the notion of a stuttering
 bisimulation, but are more amenable to mechanisation (and thus
@@
 %
-\begin{lemma}[Compile/commit]
-For all $\mcntr,\mexp,\menv,\msto,\mkont$,
+\begin{lemma}[Compile/commit]\hfill\\
 $\compile{\mexp}^\mcntr(\menv,r(\msto),r(\mkont)) = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
@@
 
-\begin{theorem}
+\begin{theorem}[Precision preservation]
 $B$ is a WEB on $\lmachstep \cup \cmachstep$
@@
 Correctness here requires a lemma that relates the differences in the compilation functions.
-\begin{lemma}
+\begin{lemma}[Compile coherence]
 For all $\mexp, \mcntr,\menv,\msto,\msdiff,\mkont,\msto^*,\mkont^*$,
@@
 
-\begin{theorem}
+\begin{theorem}[Equivalence]
 If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
@@
 
-There is a subtle loss of precision in Shivers' original time-stamp
-technique that we can fix. For a given abstract state, all writes to
-the global store need to be delayed until the analysis considers all
-branches from that abstract state. This avoids cross-branch pollution
-which would not otherwise happen, e.g., when one branch writes to
-address $\maddr$ and another branch reads from address
-$\maddr$. Fortunately, given our conversion to log-based stores this
-change is straightforward.
-
-At this point, we can begin to think of the analysis as an imperative
-algorithm with six major components:
-%
-\begin{itemize}
- \item{$\msto$, the store;}
- \item{{\tt todo}, the workset;}
- \item{{\tt seen}, a map from states to the timestamps at which they were last seen;}
- \item{$\msdiff$, the store changes for the current step;}
- \item{$\changep$, a boolean tracking whether the stepped state contributed a store change; and}
- \item{{\tt T}, the timestamp of the store.}
-\end{itemize}
-
-The new $\eval$ calculation is defined in
-figure~\ref{fig:imperative}.  To ensure termination, we guard against adding
-$c$ to {\tt todo} by the following check:
-\begin{align*}\changep \vee {\tt seen}(c) \neq {\tt T}\end{align*}
-If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.
-
-After all steps complete, we apply $\msdiff$ to $\msto$ imperatively
-(with {\tt replay!}) and increase {\tt T} as long as there was at
-least one change in $\msdiff$. This logic leads to termination if we
-know that each $\ttuple{\maddr}{\mval{s}}$ in $\msdiff$ would change
-the value of $\maddr$ in the current store. Thus, we also guard
-additions to $\msdiff$ so that only updates that would change the
-store are permitted. Each time $\msdiff$ is successfully extended, we
-set $\changep$ to true. Before each individual step, we set it
-to false.
-
-\begin{figure}
-\begin{alltt}
-eval(\(e\)) :=
-  \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
-  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \mtlst, \kmt\))
-  while(true):
-    if todo = \(\varnothing\): return (keys(seen), \(\msto\))
-    else:
-      let old := todo
-      todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
-      foreach c \(\in\) old: \(\changep\) := false; c()
-      unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
-\end{alltt}
-\caption{Imperative algorithm}
-\label{fig:imperative}
-\end{figure}
+There is two subtle losses of precision in Shivers' original
+time-stamp technique that we can fix. Firstly, for a given abstract
+state, all writes to the global store need to be delayed until the
+analysis considers all branches from that abstract state. This avoids
+cross-branch pollution which would not otherwise happen, e.g., when
+one branch writes to address $\maddr$ and another branch reads from
+address $\maddr$. Fortunately, given our conversion to log-based
+stores this change is straightforward. Secondly, the common
+implementation strategy for timestamps destructively updates each
+state's timestamp. This loses \emph{temporal} information about the
+contexts a state is visited in, and in what order. Thus that semantics
+is not equivalent to the previous section's semantics.
+
+A fix is to keep a stack of the timestamps seen at each state, and
+keep a list of stores corresponding with each timestamp. A more
+space-efficient representation that also allows an imperative store
+would be timestamping each update to the store in the same style as
+states. This changes the notion of store equality, but the new
+definition is straightforward. For analysis consumers that disregard
+temporal information, the space cost in unwarranted, but the
+realization that there is precision to be gained opens possibilities.
+
+The following transfer function has several components that can be
+destructively updated, and intermediate sets can be elided by adding
+to global sets. In fact, the log of store deltas can be removed as
+well, by updating the store in-place, and on lookup, using the first
+value timestamped $\le$ the current timestamp. We start with the
+purely functional view.
+
+\subsubsection{Pure setup for imperative implementation}
+
+The store maps to a stack of timestamped sets of abstract values.
+
+\begin{align*}
+\msto \in \Store &= \Addr \to \Valstack \\
+\mvalstack \in \Valstack &= (\Timestamp \times \wp(\Storeable))^*
+\end{align*}
+
+Here we formally define what lookup and update mean at a given
+timestamp.
+
+\begin{align*}
+\lookup(\mvalstack,t) &=
+  \left\{
+    \begin{array}{ll}
+      \mval{s} & \text{ if } \mvalstack = \cons{\ttuple{t'}{\mval{s}}}{\mvalstack'}, t' \le t \\
+      \mval{s'} & \text{ if } \mvalstack = \cons{\ttuple{t'}{\mval{s}}}{\cons{\ttuple{t''}{\mval{s'}}}{\mvalstack'}}, t' > t
+    \end{array}\right. \\
+\msto \sqcup_t [\maddr \mapsto \mval{s}] &= \msto[\maddr \mapsto \mvalstack],\changep \\
+ (\mvalstack,\changep) &= \msto(\maddr)\sqcup_t \mval{s} \\
+\epsilon \sqcup_t \mval{s} &= \ttuple{t}{\mval{s}} \\
+\cons{\ttuple{t'}{\mval{s}}}{\mvalstack} \sqcup_t \mval{s'} &= \cons{\ttuple{t'}{\mval{s}\sqcup\mval{s'}}}{\mvalstack},\strue \text{ if } t' > t \\
+\mvalstack \sqcup_t \mval{s} &= \cons{\ttuple{t+1}{\mval{s^*}}}{\mvalstack},\strue
+           \text{ if } \mval{s'} \neq \mval{s^*} \\
+ \mval{s'} &= \lookup(\msto(\maddr),t) \\
+ \mval{s^*} &= \mval{s} \sqcup \mval{s'} \\
+\mvalstack \sqcup_t \mval{s} &= \mvalstack,\sfalse \text{ otherwise}
+\end{align*}
+
+For the purposes of space, we reuse the $\dmachstep$ semantics and change the definition $\replay$ to track if the store has changed, in order to increment the global timestamp.
+\begin{align*}
+\replay_t(\epsilon,\msto,\changep) &= \msto,\changep \\
+\replay_t(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff},\msto,\changep) &= \replay_t(\msdiff,\msto',\changep \vee \changep') \\
+\text{ where } (\msto',\changep') &= \msto\sqcup_t[\maddr \mapsto, \mval{s}] \\
+\replayall_t(\varnothing,\msto,\changep) &= \msto,\changep \\
+\replayall_t(\setof{\msdiff}\cup\Xi,\msto,\changep) &= \replayall_t(\Xi, \msto', \changep') \\
+\text{ where } (\msto',\changep') &= \replay_t(\msdiff,\msto,\changep)
+\end{align*}
+
+Because the store has all the temporal information baked into it, we
+rephrase the core semantics in terms of a transfer function. The least
+fixed-point of this function gives a more compact representation of
+the reduction relation of the previous section.
+
+\begin{align*}
+\System &= (\State \to {\Timestamp}^*) \times \wp(\State) \times \Store \times \Timestamp \\
+{\mathcal F} &: \System \to \System \\
+{\mathcal F}(S,F, \msto,t) &= (S',F',\msto', t') \\
+\text{ where }
+I &= \setof{(\mstate',\msdiff') \mid
+       \mstate \in F,
+       (\mstate,\msto^*,\epsilon) \dmachstep
+       (\mstate',\msdiff')} \\
+\msto^* &= \lambda \maddr.\lookup(\msto(\maddr),t) \\
+(\msto',\changep) &= \replayall_t(\setof{\msdiff \mid (\_,\msdiff) \in I},\msto,\sfalse) \\
+t' &= \left\{\begin{array}{ll} t+1 & \text{ if } \changep \\
+              t   & \text{ otherwise}
+             \end{array}\right. \\
+F' &= \setof{\mstate \mid (\mstate,\_) \in I, \changep \vee S(\mstate) \neq \cons{t}{\_}} \\
+S' &= \lambda \mstate. \left\{\begin{array}{ll}
+                               \cons{t'}{S(\mstate)} & \text{ if } \mstate \in F' \\
+                               S(\mstate) & \text{ otherwise}
+                             \end{array}\right.
+\end{align*}
+
+\subsubsection{Pure to imperative}
+
+The intermediate data structures of the above transfer function can all be streamlined into globals that are destructively updated. In particular, there are 5 globals:
+
+\begin{enumerate}
+\item{$S$: the \emph{seen} set, though made a map for faster membership tests and updates.}
+\item{$F$: the \emph{frontier} set, which must be persisent or copied for the iteration through the set to be correct.}
+\item{$\msto$: the store, which represents all stores that occur in the reduction semantics.}
+\item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}
+\item{$\changep$: whether there has been a store change stepping all states in $F$.}
+\end{enumerate}
+
+The reduction relation would then instead of building store deltas,
+update the global store. We would also not view it as a general
+relation, but a function that adds all next states to $F$ if they have
+not already been seen. At the end of iterating through $F$, $S$ is
+updated with the new states at the next timestamp.
+
+% \begin{figure}\label{fig:pure-timestamp}
+% \begin{align*}
+% t \in \Timestamp &= {\mathbb N} \\
+% \msto \in \Store &= \Addr \to (\Timestamp \times \wp(\Storeable))^* \\
+% (S,F,\msto,t) &\tmachstep (S \cup S', F', \msto', t') \\
+% \text{ where } I &= \{(\mstate,\msdiff) \mid \mstate \in F, \mstate \
+% \end{align*}
+% \end{figure}
+
+% \begin{figure}
+% \begin{alltt}
+% eval(\(e\)) :=
+%   \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
+%   \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \mtlst, \kmt\))
+%   while(true):
+%     if todo = \(\varnothing\): return (keys(seen), \(\msto\))
+%     else:
+%       let old := todo
+%       todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
+%       foreach c \(\in\) old: \(\changep\) := false; c()
+%       unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
+% \end{alltt}
+% \caption{Imperative algorithm}
+% \label{fig:imperative}
+% \end{figure}
 
@@
 \bibliographystyle{plain}
-\input{paper.bbl}
-%\bibliography{local,bibliography}
+%\input{paper.bbl}
+\bibliography{local,bibliography}
 

### 2013-03-20 paper/paper.tex
Small change to conclusion.

@@
 \begin{lemma}[Compile coherence]
-For all $\mexp, \mcntr,\menv,\msto,\msdiff,\mkont,\msto^*,\mkont^*$,
-if $\msto \equiv \msto^*$ and $\mkont \equiv \mkont^*$, then
+If $\msto \equiv \msto^*$ and $\mkont \equiv \mkont^*$, then
 $\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$ iff
@@
 development, they can be systematically developed into efficient
-algorithms.
+algorithms that can be proved correct.
 

### 2013-03-24 paper/paper.tex
Claim that church was worst performing was bogus. Cleaned up some grammar.

@@
 
-In this article, we develop a systematic approach to deriving the
+In this article, we develop a systematic approach to deriving a
 feasible implementation of an abstract-machine-based analyzer.
@@
 
-This paper is organized in two halfs: in the first, we start with a
-quick review of the AAM approach to develop an analysis framework and
-then apply our step-by-step optimization techniques in the simplified
-setting of a core functional language.  This allows us to explicate
-the optimizations with a minimal amount of inessential technical
-overhead.  In the second half, we scale this approach up to an
-analyzer for a realistic untyped, higher-order imperative language
-with a number of interesting features and then measure
-improvements across a suite of benchmarks.
-
-At each step during the initial presentation and development, we evaluate the
-implementation on a benchmark from Vardoulakis and
+We start with a quick review of the AAM approach to develop an
+analysis framework and then apply our step-by-step optimization
+techniques in the simplified setting of a core functional language.
+This allows us to explicate the optimizations with a minimal amount of
+inessential technical overhead.  Following that, we scale this
+approach up to an analyzer for a realistic untyped, higher-order
+imperative language with a number of interesting features and then
+measure improvements across a suite of benchmarks.
+
+At each step during the initial presentation and development, we
+evaluated the implementation on a set of benchmarks. The highlighted
+benchmark in figure \ref{fig:churchtime} is from Vardoulakis and
 Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
-of multiplication over addition on Church numerals.
-For the step-by-step development, 
-this benchmark is particularly informative:
+of multiplication over addition on Church numerals.  For the
+step-by-step development, this benchmark is particularly informative:
 \begin{enumerate}
@@
 %
-\item it proves to be the \emph{least} improved benchmark of the
-  complete suite considered in section~\ref{sec:eval}, and thus it
-  serves as a good sanity check and lower-bound for each of the
-  optimization techniques considered.
+\item its improvement is about median in the benchmark suite
+  considered in section~\ref{sec:eval}, and thus it serves as a good
+  sanity check for each of the optimization techniques considered.
 \end{enumerate}
@@
 We start, in section~\ref{sec:aam}, by developing an abstract
-interpreter according to the AAM approach.  Without further
-abstraction, the analysis is exponential due to per-state store
-variance and thus cannot analyze the example in a reasonable amount of
-time.  In section~\ref{sec:baseline}, we perform a further abstraction
-by widening the store.  The resulting analyzer sacrifices precision
-for speed and is able to analyze the example in about 1 minute.  This
-step is described by Van Horn and Might~\cite[\S
-  3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make
-even small examples feasible.  We therefore take the widened
-interpreter as the baseline for our evaluation.
+interpreter according to the AAM approach.  In the initial
+abstraction, each state carries a store (what is called per-state
+store variance). The space of stores is exponential in size; without
+further abstraction, the analysis is exponential and thus cannot
+analyze the example in a reasonable amount of time.  In
+section~\ref{sec:baseline}, we perform a further abstraction by
+widening the store.  The resulting analyzer sacrifices precision for
+speed and is able to analyze the example in about 1 minute.  This step
+is described by Van Horn and Might~\cite[\S
+3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make even
+small examples feasible.  We therefore take a widened interpreter as
+the baseline for our evaluation.
 
@@
 result of analyzing a program is a finite directed graph where nodes
-in the graph are (abstract) machine states and edges denote machine
+in the graph are machine states and edges denote machine
 transitions between states.
@@
 %
-Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.  It
+Figure~\ref{fig:syntax} defines the syntax of ISWIM.  It
 includes variables, literals (either integers, booleans, or
@@
 \begin{array}{l@{\qquad}rcl}
-\text{Values} & \mval   &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
+\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
 \text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
@@
 &&|& \kif{\mexp,\mexp,\menv,\maddr}\\
-\text{Environments} &\menv&\in& \mathit{Var} \rightharpoonup \mathit{Addr}\\
-\text{Stores} &\msto&\in& \mathit{Addr} \rightharpoonup \mathcal{P}(\mathit{Value})
+\text{Addresses} &\maddr&\in&\Addr \\
+\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
+\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
 \end{array}
@@
 figure~\ref{fig:aam} defines the transition relation.  The evaluation
-of a program is defined as the set of states reachable by the
-reflexive, transitive closure of the machine transition relation.  The
+of a program is defined as its traces given by the machine transition relation.  The
 machine is a very slight variation on a standard abstract machine for
@@
 \item continuations are heap-allocated, not stack-allocated,
-\item there are ``contour values'' (written $\mcntr$) and syntax
+\item there are ``timestamps'' (written $\mcntr$) and syntax
   labels ($\mlab$) threaded through the computation, and
 \item the machine is implicitly parameterized by the functions
-  $\mathit{push}$, $\mathit{bind}$, and $\interpdelta$.
+  $\mathit{alloc}$, $\mathit{tick}$,
+  $\mathit{push}$, $\interpdelta$, and spaces
+  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
 \end{itemize}
+
+More generally, we can make stores map to a domain that forms a galois
+connection with sets of values, so that $\interpdelta$ can give more
+elaborate abstractions of base values (e.g., with the interval
+abstraction, $\{0,[0,5]\} \equiv \{[0,5]\}$). We stick with sets for
+simpler exposition.
 
@@
 \begin{align*}
-\eval(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
+\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
 \end{align*}
@@
 \begin{array}{@{}r@{\ }c@{\ }l@{}}
+&&\mcntr' =\tick(\mcntr) \\
+&&\maddr  =\alloc(\mstate) \\
 %% EVAL
@@
 \co{\mkont,\mval,\msto}
-\text{ where }\mval \in \msto(\menv(\mvar))
+\text{ if }\mval \in \msto(\menv(\mvar))
 \\
@@
 \\
-\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
-\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
+\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
+\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
 \\
 \ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
-\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
+\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
 \\
@@
 \ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
-\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
+\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
 \\
@@
 \\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep&
-\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
-\\
-\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep&
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
+\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
+\\
+\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
 \ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
-\text{ where }\mkont \in \msto(\maddr)
-\\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep&
-\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
-\text{ where }\mkont\in\msto(\maddr)
-\\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep&
-\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
-\text{ where }\mkont\in\msto(\maddr)
+\text{ if }\mkont \in \msto(\maddr_\mkont)
+\\
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
+\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
+\text{ if }\mkont\in\msto(\maddr_\mkont)
+\\
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
+\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
+\text{ if }\mkont\in\msto(\maddr_\mkont)
 \\[2mm]
@@
 \\
-\multicolumn{3}{r@{}}{
-\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)}
+&&\text{ where } \menv' = \menv[\mvar\mapsto\maddr] \\
+&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
 \\
@@
 \co{\mkont,\mval',\msto}
-\text{ where } \mval'\in\interpdelta(\mop,\mval)
+\text{ if } \mval'\in\interpdelta(\mop,\mval)
 \end{array}
@@
 \begin{align*}
-\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
+\push_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
 \mbox{ where }\maddr \notin\msto
 \\
-\mathit{bind}_\mlab^\mcntr(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
-\mbox{ where }\maddr \notin\msto
-\end{align*}
-The resulting relation is non-deterministic in its choice of
-addresses, however it must always choose a fresh address when
+\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto
+\end{align*}
+where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
+\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
+\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
+$\mcntr$, but they can be used to determinize the choice of fresh
+addresses.  The resulting relation is non-deterministic in its choice
+of addresses, however it must always choose a fresh address when
 allocating a continuation or variable binding.  If we consider machine
 states equivalent up to consistent renaming and fix an allocation
-scheme, this relation defines a deterministic machine (the relation is really a function).
+scheme, this relation defines a deterministic machine (the relation is
+really a function).
 
@@
 
-\paragraph{Abstract interpretation} To characterize abstract interpretation, set the implicit
-parameters just as above, but drop the $\maddr \not\in \msto$
-condition.  This family of interpreters is also non-deterministic in
-choices of addresses, but it is free to choose addresses that are
-already in use.  Consequently, the machines may be non-deterministic
-when multiple values reside in a store location.
+\paragraph{Abstract interpretation} To characterize abstract
+interpretation, set the implicit parameters just as above, but drop
+the $\maddr \not\in \msto$ condition. $\interpdelta$ takes some care
+to not make the analysis run forever; a simple instantiation is a flat
+abstraction that returns an abstract top element $\mathit{Integer}$ as
+the result of arithmetic operations.  This family of interpreters is
+also non-deterministic in choices of addresses, but it is free to
+choose addresses that are already in use.  Consequently, the machines
+may be non-deterministic when multiple values reside in a store
+location.
 
@@
 \begin{align*}
-\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
+\Counter &= \Label^* \\
+\mtcntr &= \epsilon \\
+\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
+\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
+\push_\mlab^\mcntr(\msto,\mkont) &=
   \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
-\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
-                                           \msto\sqcup[\maddr \mapsto
-                                             \{ \mval\}],
-                                           \mcntr' \\
-\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
-              \maddr &= x\mcntr' \\
-%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
-              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
-              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
-\end{align*}
-where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
-\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
+\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
+\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\msto,\mval}) &= \mcntr \\
+\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
+  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
+  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
+\end{align*}
 The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
@@
 
-The uniform $k$-CFA allocation strategy would make $\eval$ in figure
-\ref{fig:aam} a computable abstraction of reachable states, but not an
+The uniform $k$-CFA allocation strategy would make $\traces$ in figure
+\ref{fig:aam} a computable abstraction of possible executions, but not an
 efficient one. This is not the strategy that AAM, nor we, recommend. Through
-this and the following section, we will explain a succession of approximations
+this section, we will explain a succession of approximations
 to reach the baseline analysis.  We'll compare performance at each stage to
-identify the criticality of each optimization. 
+identify the criticality of each choice. 
 %
@@
 
-Conceptually, the AAM approach calls for computing an analysis as a graph
-exploration: (1) start with an initial state, and (2) compute the transitive
-closure of the transition relation from that state.
+Conceptually, the AAM approach calls for computing an analysis as a
+graph exploration: (1) start with an initial state, and (2) compute
+the transitive closure of the transition relation from that state. All
+visited states are potentially reachable in the concrete, and all
+paths through the graph are possible traces of execution.
 
@@
 %
-Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
+Given the initial state $\mstate_0$ and the transition relation $\machstep$,
 we define the global transfer function:
 \begin{equation*}
- F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
+ F_{\mstate_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
  \text.
@@
 Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
-\begin{equation*}
-  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
-\end{equation*}
+\begin{align*}
+  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
+    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
+    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
+\end{align*}
 Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
 \begin{equation*}
- \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
+ \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
 \end{equation*}
-where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.
+where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.
+
+The possible traces of execution tell us the most about a program, so
+we take $\traces(\mexp)$ to be the (regular) set of paths through the
+computed graph.
 
@@
 %
-Thus, in one step, complexity drops from intractable exponentially to a merely
+Thus, in one step, complexity drops from intractable exponentiality to a merely
 daunting polynomial.
@@
 for the analysis' precision and determinism. Specifically, higher
-precision, and is deterministic even if set iteration is not.
+precision, and it is deterministic even if set iteration is not.
 
@@
 
+For clarity, we will exhibit non-widened semantics unless a technique
+specifically requires it.
+
 \subsection{Store-allocated results}
@@
 However, the {\tt fn} continuation stores a value, and this makes the space of
-continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0CFA.
+continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0-CFA.
 Having the space of continuations grow linearly with the size of the program 
@@
 languages with $n$-ary functions, since otherwise the continuation
-space grows super-exponentially. In our example language, we have only
-unary functions, so application forms always have two positions: the
-function and the argument. Thus for every application labeled $\mlab$,
-we introduce two new labels, $\mlab^f$ and $\mlab^a$.
+space grows super-exponentially. We extend the semantics to
+additionally allocate addresses to store the values of each
+subexpression of an application. The continuations have to store these
+addresses to remember where to get argument values from, though
+sometimes your allocation strategy may allow you to elide that
+representation.
+
+The new evaluation rules follow (where $\mcntr' = \tick(\mlab,\mcntr)$).
 % In theory, this aggressive distinction between continuations might buy
@@
 
-In $k$-CFA, the evaluation rules become:
-
 \newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
@@
 \begin{align*}
-%ext(\msto, \maddr, s) &= \msto \sqcup [\maddr \mapsto s] \\
 \ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
-\ev[^\mcntr]{\mexpi{0},\menv,\ext{\msto}{\mlab}{\{\mkont\}},\kar{\mexpi{1},\menv,\mlab,\mcntr}}
-\\
-\co{\kar{\mexp,\menv,\mlab,\mcntr},\mval,\msto} & \machstep
-\ev[^\mcntr]{\mexp,\menv,\ext{\msto}{\mlab^f\mcntr}{\{\mval\}},\kfn{\mlab^f,\mlab,\mcntr}}
-\\
-\co{\kfn{\mlab^f,\mlab,\mcntr},\mval,\msto} & \machstep
-\ap[^\mcntr_\mlab]{\mvalx{u},\mlab^a,\mkont,\ext{\msto}{\mlab^a\mcntr}{\{\mval\}}}
-\\
-\text{ where } &\mkont \in \msto(\mlab\mcntr), \mvalx{u} \in \msto(\mlab^f\mcntr)
-\\
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
-\ev[^{\mcntr'}]{\mexp,\menv[\mvar\mapsto\mvar\mcntr'],\ext{\msto}{\mvar\mcntr'}{\msto(\mlab^a\mcntr)},\mkont}
-\\
-\text{ where }& \mcntr' = \lfloor\mlab\mcntr\rfloor_k
-\\
-\ap[^\mcntr_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
-\co{\mkont,\mval',\msto}
-\\
-\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a\mcntr)\}
-\end{align*}
+\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar{\mexpi{1},\menv,\maddr_f,\maddr_a,\maddr_\mkont}} \\
+\text{ where }
+  \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) \\
+  \maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
+\end{align*}
+This rule describes how we allocate addresses for subexpressions of an
+application up front. This adds another parameter to the machine, but
+it is very much like $\alloc$. We use $f$ and $a$ subscripts to denote
+function and argument positions, respectively. Whereas ANF names all
+intermediate values in the syntax, we allocate addresses for
+intermediate values. This representational leap is not a big one, but
+is notably important for analyzing direct style terms.
+
+\begin{align*}
+\co[^\mcntr]{\kar{\mexp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \machstep
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_a,\maddr_\mkont}} \\
+\text{ where }
+  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
+\\
+\co[^\mcntr]{\kfn{\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \machstep
+\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr_a,\mkont,\msto'}
+\\
+\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f) \\
+\text{ where }
+ \msto' &= \ext{\msto}{\maddr_a}{\{\mval\}}
+\end{align*}
+We associate the value in hand with the expression that produced
+it. Also, we fan out to all functions that are associated with the
+original function position, because applied functions are in
+\emph{strict position} (more about that later).
+
+\begin{align*}
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\maddr_a, \msto,\mkont} & \machstep
+\ev[^{\mcntr'}]{\mexp,\menv',\msto',\mkont} \\
+\text{ where }
+\maddr &= \alloc(\mstate) \\
+\menv' &= \menv[\mvar\mapsto\maddr] \\
+\msto' &= \msto\sqcup[\maddr \mapsto \msto(\maddr_a)]
+\end{align*}
+Arguments are not passed as values, but as addresses to their values.
+
+\begin{align*}
+\ap[^\mcntr_\mlab]{\mop,\maddr_a,\msto,\mkont} & \machstep
+\co[^{\mcntr'}]{\mkont,\mval',\msto}
+\\
+\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\maddr_a)\}
+\end{align*}
+Primitive application still nondeterministically produces all possible values.
 
@@
 \begin{enumerate}
+  \setcounter{enumi}{2}
  \item store deltas;
@@
 
-All but the first technique are precision preserving. We will discuss the design rationale
-for the move.
+All techniques are precision preserving under reasonable
+assumptions. We will discuss these assumptions as the caveats arise.
 
@@
 where {\tt f}, {\tt x} and {\tt y} each have several values
-each argument evaluation induces two-way branching, only to be ultimately joined back together in their respective
+each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective
 application positions. 
@@
 %
-We say the abstraction is \emph{lazy} because it delays dereferencing an
-address until its contents are \emph{needed} as values in the semantics. It
-does not change the execution order that leads to the values that are stored in
-the address.
-
-We introduce a new kind of value, $\saddr{\maddr}$, that represents a
-delayed lookup of $\maddr$.  The following rules highlight the
-changes to the semantics:
+We say the abstraction is \emph{lazy} because it delays splitting on
+the values in an address until they are \emph{needed} in the
+semantics. It does not change the execution order that leads to the
+values that are stored in the address.
+
+We introduce a new kind of value,
+\spchoice
+{$\saddr{\maddr}$}
+{$\superposition{\mval{s}}$ (for ``superposition'')},
+%
+that represents a delayed non-deterministic choice of a value from
+\spchoice
+{$\msto(\maddr)$}
+{$\mval{s}$}.
+%
+The following rules highlight the changes to the semantics:
 
@@
 \begin{align*}
-\force &: \Store \times \Value \to \Set(\Value) \\
-\force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
-\force(\msto,\mval) &= \{\mval\} \\
-\ext(\msto, \maddr, \mval) &= \msto \sqcup[\maddr \mapsto \force(\msto, \mval)] \\
+\spchoice
+{\force &: \Store \times \Value \to \Set(\Value) \\
+ \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
+ \force(\mval) &= \{\mval\}}
+{\force &: \Value \to \Set(\Value) \\
+ \force(\superposition{\mval{s}}) &= \mval{s} \\
+ \force(\mval) &= \{\mval\}}
+\\
+%
 \ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
-\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
+\spchoice
+{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
+{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
+%
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto}
 &\lmachstep\;
-\ev[^\mcntr]{\mexp,\menv,\ext(\msto, \maddr^f,\mval),\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
-\mathit{bind\,}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
-  \maddr], \ext(\msto,\maddr, \mval), \mcntr' \\
-\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
-              \maddr &= \mvar\mcntr'
-\end{align*}
-
+\ev[^\mcntr]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont}} \\
+\text{ where }
+\msto' &=
+\spchoice
+{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
+{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
+%
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
+&\lmachstep\;
+\ev[^\mcntr]{\mexpi,\menv,\msto,\mkont} \\
+\text{ if } \mkont &\in \msto(\maddr_\mkont),
+            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
+\end{align*}
+Since {\tt if} guards are in strict position, we must force the value
+to determine which branch to to take. The middle rule uses $\force$
+only to combine with values in the store - it does not introduce
+needless non-determinism.
+
+\spchoice{
 \noindent
@@
 two approaches are the same.
-
+}{}
+
+\spchoice
+{
 \begin{theorem}[Soundness]
@@
 the LHS value must be a subset of forcing the corresponding RHS value.
-The proof is by cases on $\mstate \machstep \mstate'$.
+The proof is by cases on $\mstate \machstep \mstate'$.}
+{
+\begin{theorem}[Completeness]For all $\mexp$,
+ $\traces_{\text{Lazy}}(\mexp)$ is a complete abstraction of $\traces_{\text{ISWIM}}(\mexp)$.
+\end{theorem}
+We have a statement about traces because we need induction to show no
+cruft values are in superposition.  The induction hypothesis tells us
+that there are non-lazy traces that lead to all the values in
+superposition, so when we take a lazy step, we are taking several
+non-lazy steps, and we stay in sync. The other direction we just
+collapse the superposition in each possibility to construct the
+non-lazy traces.}
 
@@
 idea---same reasoning, fewer transitions---with abstract
-compilation. Abstract compilation is \emph{precision-preserving} and
-transforms complex expressions whose \emph{abstract} evaluation is
-deterministic into ``abstract bytecodes.''  The abstract interpreter
-then does in one transition what previously took many. In short, abstract
-compilation eliminates unnecessary allocation, deallocation and
-branching.
+compilation. Abstract compilation transforms complex expressions whose
+\emph{abstract} evaluation is deterministic into ``abstract
+bytecodes.''  The abstract interpreter then does in one transition
+what previously took many. In short, abstract compilation eliminates
+unnecessary allocation, deallocation and branching. The technique is
+precision preserving without store widening. We discuss the precision
+differences with store widening at the end of the section.
 
@@
 \[
-\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
+\mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
 \]
@@
 \begin{align}
-& \ev{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\mkont,\msto_0}\\
+& \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
 \machstep\; &
-\ev{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\kar{\mexp_3,\menv,\maddr_1},\msto_1}
+\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_{f1},\maddr_{a1},\maddr_1}}
 \\
 \machstep\; &
-\ev{\sapp\mvar{\mexp_1},\menv,\kar{\mexp_2,\menv,\maddr_2},\msto_2}
+\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_{f2},\maddr_{a2},\maddr_2}}
 \\
 \machstep\; &
-\ev{\mvar, \menv,\kar{\mexp_1,\menv,\maddr_3},\msto_3} % {\mexp_2}
+\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3}} % {\mexp_2}
 \\
 \machstep\; &
-\co{\kar{\mexp_1,\menv},\mval,\msto_4} % {\mexp_1}{\mexp_2}
-\mbox{ where } \mval \in \msto(\menv(\maddr))
+\spchoice
+{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto_4,\saddr{\menv(\mvar)}}}
+{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto_4,\superposition{\msto_4(\menv(\mvar))}}}
 \end{align}
 where \(
-\msto_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
-[\maddr_2 \mapsto \kar{\mexp_3,\menv,\maddr_1}],
-[\maddr_3 \mapsto \kar{\mexp_2,\menv,\maddr_2}]\).
+\msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
+\maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
+\maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).
+
+Whereas abstract compilation gives us in one step:
+\begin{align*}
+\compile{\mexp}^\mcntr(\menv,\msto_0,\mkont) = 
+\spchoice
+{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto'_4,\saddr{\menv(\mvar)}}}
+{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto'_4,\superposition{\msto'_4(\menv(\mvar))}}}
+\end{align*}
+where \(
+\msto'_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
+[\maddr_2 \mapsto \kar{\compile{\mexp_3},\menv,\maddr_1}],
+[\maddr_3 \mapsto \kar{\compile{\mexp_2},\menv,\maddr_2}]\).
 
@@
 create {\tt ev} states to instead call these functions. Figure
-    \ref{fig:caam} shows the modified reduction relation.
+\ref{fig:caam} shows the modified reduction relation. The only change
+from the previous semantics is that $\ev{}$ state construction is
+replaced by calling the compiled expression.
 
@@
 \compile{\_} &: \Expr \to \Env \times \Store \times \Kont \to \State \\
-\compile{\svar\mvar} &= \lambda(\menv,\msto,\mkont) .\co{\mkont,\saddr{\menv(\mvar)},\msto}
+\compile{\svar\mvar} &=
+ \lambda(\menv,\msto,\mkont) .
+\spchoice
+{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
+{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}}
 \\
@@
 \compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr})
-\\
-&
-\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
+\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_f,\maddr_a,\maddr_\mkont})
+\\
+&\setlength\arraycolsep{5pt}\begin{array}{lrl}\text{ where } & \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) 
+\\&
+\maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
+\end{array}
 \\
@@
 \\
-&\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
+&\text{ where }\maddr,\msto' = \push_\mlab^\mcntr(\msto,\mkont)
 \end{align*}
@@
 \begin{align*}
-\mathit{eval}(\mexp) &= \{ \mstate\ |\ \compile{\mexp}(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
+\traces(\mexp) &= \{ \compile{\mexp}^\mtcntr(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
 \end{align*}
@@
 \co{\kmt,\mval,\msto} &\cmachstep
-\ans{\msto,\mval}
-\\
-\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \cmachstep
-\mcomp^\mcntr(\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr})
-\\
-\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \cmachstep
-\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
-\text{ where }\mkont \in \msto(\maddr)
+\ans{\msto,\mval'} \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
+\\
+\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \cmachstep
+\mcomp^\mcntr(\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont}) \\
+\text{ where } \msto' &= \msto\sqcup[\maddr_f \mapsto \spchoice{\force(\msto,\mval)}
+                                                              {\force(\mval)}]
+\\
+\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \cmachstep
+\ap[^\mcntr_\mlab]{\mval,\maddr_a,\mkont,\msto'} \\
+\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\maddr_\mkont) \\
+\text{ where } \msto' &= \msto\sqcup[\maddr_f \mapsto \spchoice{\force(\msto,\mval)}
+                                                               {\force(\mval)}]
 \\
@@
 \mcompi0^\mcntr(\menv,\msto,\mkont)
-\text{ where }\mkont\in\msto(\maddr)
+\text{ if }\mkont\in\msto(\maddr)
 \\
@@
 \mcompi1^\mcntr(\menv,\msto,\mkont)
-\text{ where }\mkont\in\msto(\maddr)
+\text{ if }\mkont\in\msto(\maddr)
 \\[2mm]
 %% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \cmachstep
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\maddr_a,\msto,\mkont} & \cmachstep
 \mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
-\text{ where }\menv',\msto',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
+\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
+              \msto' &= \msto\sqcup[\maddr\mapsto\msto(\maddr_a)]
 \\
@@
 \co{\mkont,\mval',\msto} \\
-\text{ where }\mkont &\in\msto(\maddr)
-\text{ and } \mval'\in\interpdelta(\mop,\mval)
+\text{ where } \mvalx{u} &\in \spchoice{\force(\msto,\mval)}{\force(\mval)}, \mval'\in\interpdelta(\mop,\mvalx{u})
 \end{align*}
@@
 
+\paragraph{Wide store and abstract compilation}
+A small problem arises if we try to widen the store as in section
+\ref{sec:storewiden}. In one step, we can both bind a variable and
+reference it. The problem is that the store does not change until all
+steps are done, so the reference unsoundly returns the previous
+binding (or nothing at all)! There are a couple ways to address
+this. One way is to make all compiled forms that reference the store
+into a new kind of state that has to be stepped in order to reference
+the store and continue. Another way is to make function application
+take one extra step after binding, so that running the body of the
+function sees the changed store. The latter is the more enticing of
+the two, since programs have more variable references than function
+calls - we don't want to slow them down if we don't have to.
+
+The fact that many changes happen in one step also makes the
+comparison with the widened lazy semantics more subtle. It is possible
+for different stores to occur between the different semantics because
+abstract compilation can change the order in which the store is
+changed. Consider the case where before we might call the same
+function from two different places in 3 and 2 steps respectively,
+compilation can make that 2 and 1, reversing the order that we bind
+values in the store, leading to a mismatch from the previous
+semantics. This kind of cross-step store poisoning is inevitable when a
+store is shared between several traces. The binding could have just as
+easily been observed the other way around in the non-compiled
+semantics.
+
 % Compile: cpu time: 255397 real time: 261532 gc time: 2947
@@
 change log on the full store after all steps have completed. This uses
-far fewer join operations, leading to less overhead, and is \emph{precision-preserving}.
+far fewer join operations, leading to less overhead, and is precision-preserving.
 
@@
 (\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
-\text{ where }\menv',\msdiff',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\msdiff,\mvar,\mval)
-\end{align*}
-\begin{align*}
-\mathit{bind\,}^{\mcntr}_\mlab(\menv, \msto, \msdiff, \mvar, \mval) &= \menv[\mvar \mapsto \maddr], \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\msdiff}, \mcntr' \\
-\text{ where } \mcntr' &= \lfloor \mlab \mcntr\rfloor_k \\
-               \maddr &= \mvar\mcntr'
+\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
+              \msdiff' &= 
+\spchoice
+{\cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\msdiff}}
+{\cons{\ttuple{\maddr}{\force(\mval)}}{\msdiff}}
 \end{align*}
@@
 
-There is two subtle losses of precision in Shivers' original
-time-stamp technique that we can fix. Firstly, for a given abstract
-state, all writes to the global store need to be delayed until the
-analysis considers all branches from that abstract state. This avoids
+There are two subtle losses of precision in Shivers' original
+time-stamp technique that we can fix.
+
+\begin{enumerate}
+ \item{For a given abstract
+state, all writes to the global store should be delayed until the
+analysis considers all current work. This avoids
 cross-branch pollution which would not otherwise happen, e.g., when
@@
 address $\maddr$. Fortunately, given our conversion to log-based
-stores this change is straightforward. Secondly, the common
-implementation strategy for timestamps destructively updates each
-state's timestamp. This loses \emph{temporal} information about the
-contexts a state is visited in, and in what order. Thus that semantics
-is not equivalent to the previous section's semantics.
-
-A fix is to keep a stack of the timestamps seen at each state, and
-keep a list of stores corresponding with each timestamp. A more
-space-efficient representation that also allows an imperative store
-would be timestamping each update to the store in the same style as
-states. This changes the notion of store equality, but the new
-definition is straightforward. For analysis consumers that disregard
-temporal information, the space cost in unwarranted, but the
-realization that there is precision to be gained opens possibilities.
+stores this change is straightforward.}
+\item{The common implementation strategy for timestamps destructively
+    updates each state's timestamp. This loses \emph{temporal}
+    information about the contexts a state is visited in, and in what
+    order. The semantics is different - we can't precisely answer
+    questions about the traces of the reduction relation we defined. A
+    simple fix is to keep a stack of the timestamps seen at each
+    state, and keep a list of stores corresponding with each
+    timestamp. A more space-efficient representation that also allows
+    an imperative store would be timestamping each update to the store
+    in the same style as states. This changes the notion of store
+    equality, but the new definition is straightforward. For analysis
+    consumers that disregard temporal information, the space cost in
+    unwarranted, but the realization that there is precision to be
+    gained opens possibilities.}
+\end{enumerate}
 
@@
 \begin{center}
-  \includegraphics[width=6.5in]{all-relative-space}
-
-  (a) Peak memory usage
+  \includegraphics[width=6.5in]{all-relative-time}
+
+  (a) Total analysis time
 
@@
   \vspace{1em}
-  \includegraphics[width=6.5in]{all-relative-time}
-
-  (c) Total analysis time
+  \includegraphics[width=6.5in]{all-relative-space}
+
+  (c) Peak memory usage
 \end{center}
@@
   is not a coincidence; these papers set a high standard for
-  evaluation, which we consciously aimed to approach.}  Unfortunately, these
-analyses compute an inherently different and incomparable form of
-analysis.  Consequently, we have omitted a complete comparison with
-these implementations,.  The AAM approach provides more precision in
+  evaluation, which we consciously aimed to approach.}  Unfortunately,
+these analyses compute an inherently different and incomparable form
+of analysis.  Consequently, we have omitted a complete comparison with
+these implementations.  The AAM approach provides more precision in
 terms of temporal-ordering of program states, which comes at a cost
@@
 terms of precision, ease of implementation and verification, and rapid
-design.
+design. We can get closer to their performance by relying on the
+representation of addresses and the behavior of $\alloc$ to
+pre-allocate most data structures and split the abstract store out
+into parts that are more quickly accessed and updated. The first two
+of our optimizations can still be applied to an analysis that does
+abstract garbage collection~\cite{dvanhorn:Might:2006:GammaCFA},
+whereas the polymorphic splitting implementation is tied strongly to a
+single-threaded store.
 
@@
 Boucher and Feeley \cite{dvanhorn:Boucher1996Abstract} introduced the
-idea of \emph{abstract compilation}, which used closure generation
+idea of abstract compilation, which used closure generation
 \cite{dvanhorn:Feeley1987Using} to improve the performance of control
 flow analysis.  We have adapted the closure generation technique from
-composition evaluators to abstract machines and applied it to similar
+compositional evaluators to abstract machines and applied it to similar
 effect.
@@
 %
-These analyses (ultimately) compute judgments on program terms and
+Such analyses (ultimately) compute judgments on program terms and
 contexts, e.g., at expression $e$, variable $x$ may have value $v$.
@@
 
-We view the primary contribution of this work as a systematic path
-that eases the design, verification, and implementation of analyses
-using the abstracting abstract machine approach to within a factor of
-performant constraint-based analyses.
-
 \section{Conclusion}
@@
 development, they can be systematically developed into efficient
-algorithms that can be proved correct.
+algorithms that can be proved correct. We view the primary
+contribution of this work as a systematic path that eases the design,
+verification, and implementation of analyses using the abstracting
+abstract machine approach to within a factor of performant
+constraint-based analyses.
 

### 2013-03-25 paper/paper.tex
Added more of a teaser to the end of the intro.

@@
 
-However, there is a considerable weakness with the approach: an
+However, there is a considerable weakness to the approach: an
 analyzer designed and implemented by following the AAM recipe is
@@
 In this article, we develop a systematic approach to deriving a
-feasible implementation of an abstract-machine-based analyzer.
+practical implementation of an abstract-machine-based analyzer using
+mostly semantic means - not tricky engineering. Our goal is to empower
+programming language implementers and researchers to explore and
+convincingly exhibit their ideas with a low barrier to entry. The
+optimizations we describe are widely applicable and apparently
+effective to scale far beyond toy examples.
 
@@
 %
-\item its improvement is about median in the benchmark suite
-  considered in section~\ref{sec:eval}, and thus it serves as a good
-  sanity check for each of the optimization techniques considered.
+\item it proves to be the \emph{least} improved benchmark of the
+  complete suite considered in section~\ref{sec:eval}, and thus it
+  serves as a good sanity check and lower-bound for each of the
+  optimization techniques considered.
 \end{enumerate}
@@
 
-Conceptually, the AAM approach calls for computing an analysis as a
-graph exploration: (1) start with an initial state, and (2) compute
-the transitive closure of the transition relation from that state. All
-visited states are potentially reachable in the concrete, and all
-paths through the graph are possible traces of execution.
+Conceptually, the AAM approach calls for computing an analysis as a graph
+exploration: (1) start with an initial state, and (2) compute the transitive
+closure of the transition relation from that state.
 
@@
 %
-Given the initial state $\mstate_0$ and the transition relation $\machstep$,
+Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
 we define the global transfer function:
 \begin{equation*}
- F_{\mstate_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
+ F_{\varsigma_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
  \text.
@@
 Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
-\begin{align*}
-  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
-    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
-    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
-\end{align*}
+\begin{equation*}
+  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
+\end{equation*}
 Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
 \begin{equation*}
- \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
+ \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
 \end{equation*}
-where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.
-
-The possible traces of execution tell us the most about a program, so
-we take $\traces(\mexp)$ to be the (regular) set of paths through the
-computed graph.
+where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.
+
+{\bf FIXME} use $\traces$
 
@@
 %
-Thus, in one step, complexity drops from intractable exponentiality to a merely
+Thus, in one step, complexity drops from intractable exponentially to a merely
 daunting polynomial.
@@
 of analysis.  Consequently, we have omitted a complete comparison with
-these implementations.  The AAM approach provides more precision in
+these implementations,.  The AAM approach provides more precision in
 terms of temporal-ordering of program states, which comes at a cost
@@
 Boucher and Feeley \cite{dvanhorn:Boucher1996Abstract} introduced the
-idea of abstract compilation, which used closure generation
+idea of \emph{abstract compilation}, which used closure generation
 \cite{dvanhorn:Feeley1987Using} to improve the performance of control
 flow analysis.  We have adapted the closure generation technique from
-compositional evaluators to abstract machines and applied it to similar
+composition evaluators to abstract machines and applied it to similar
 effect.

### 2013-03-28 paper/paper.tex
Added Nick to author list.
Trying to reorder for better exposition.

--- 2013-03-24 paper/paper.tex
@@
            {ianj@ccs.neu.edu}
+\authorinfo{Nicholas Labich}
+           {Northeastern University}
+           {labichn@ccs.neu.edu}
 \authorinfo{Matthew Might}
@@
            {dvanhorn@ccs.neu.edu}
-
 \maketitle
@@
 \item continuations are heap-allocated, not stack-allocated,
-\item there are ``timestamps'' (written $\mcntr$) and syntax
+\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
   labels ($\mlab$) threaded through the computation, and
 \item the machine is implicitly parameterized by the functions
-  $\mathit{alloc}$, $\mathit{tick}$,
-  $\mathit{push}$, $\interpdelta$, and spaces
+  $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces
   $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
@@
 \begin{array}{@{}r@{\ }c@{\ }l@{}}
-&&\mcntr' =\tick(\mcntr) \\
-&&\maddr  =\alloc(\mstate) \\
+\mstate &\machstep&\mstate' \text{ defined to be the following} \\
+&&\text{let } \mcntr' =\tick(\mstate) \\
 %% EVAL
@@
 &&
-\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
+\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
 \\
@@
 &&
-\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
+\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
 \\[2mm]
@@
 \\
-&&\text{ where } \menv' = \menv[\mvar\mapsto\maddr] \\
+&&\text{ where } \maddr  =\alloc(\mstate) \\
+&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
 &&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
@@
 \begin{align*}
-\push_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
-\mbox{ where }\maddr \notin\msto
-\\
-\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto
+\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
+\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
 \end{align*}
@@
 to not make the analysis run forever; a simple instantiation is a flat
-abstraction that returns an abstract top element $\mathit{Integer}$ as
-the result of arithmetic operations.  This family of interpreters is
-also non-deterministic in choices of addresses, but it is free to
-choose addresses that are already in use.  Consequently, the machines
-may be non-deterministic when multiple values reside in a store
-location.
+abstraction where arithmetic operations return an abstract top element
+$\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on
+$\sNum$.  This family of interpreters is also non-deterministic in
+choices of addresses, but it is free to choose addresses that are
+already in use.  Consequently, the machines may be non-deterministic
+when multiple values reside in a store location.
 
@@
 \alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
-\push_\mlab^\mcntr(\msto,\mkont) &=
-  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
+\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
 \tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
-\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\msto,\mval}) &= \mcntr \\
+\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\mval,\msto}) &= \mcntr \\
 \tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
@@
 \begin{equation*}
- F_{\mstate_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
- \text.
+ F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
 \end{equation*}
@@
 we take $\traces(\mexp)$ to be the (regular) set of paths through the
-computed graph.
+computed graph. We elide the construction of the set of edges in this paper.
 
@@
 
-To describe this process, we'll refactor the transition relation so that it operates on
+To describe this process, AAM defined a transformation of the reduction relation so that it operates on
 a pair of a set of contexts ($C$) and a store ($\sigma$).
@@
 %
-The refactored relation, $\widehat{\machstep}$, becomes:
+The transformed relation, $\widehat{\machstep}$, is
 %
@@
 \wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
-\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
+\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
 \wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
@@
               \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
-\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
-\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
-\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
-\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval} \\
 \inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
@@
 
+% Notice that now $S$ has several copies of the abstract store in it. As
+% it is, this semantics is much less efficient (but still more precise)
+% than the previously proposed semantics because membership checks have
+% to compare entire stores. We return to this in section
+% \ref{sec:timestamping}.
+
 For clarity, we will exhibit non-widened semantics unless a technique
-specifically requires it.
-
-\subsection{Store-allocated results}
+specifically requires it. The final step to the baseline takes the
+complexity from quartic to cubic in the monovariant case.
+
+\subsection{Store-allocate all values}
 \label{sec:baselineeval}
 
-The final approximation we make to get to our point of departure is
-store-allocating results of application sub-expressions. 
-The AAM approach stops at the previous optimization. 
-However, the {\tt fn} continuation stores a value, and this makes the space of
-continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0-CFA.
-Having the space of continuations grow linearly with the size of the program 
-will drop the overall complexity to cubic (as expected).
+The final approximation we make to get to our point of departure is to
+store-allocate all values that appear, so that any non-machine state
+that contains a value instead contains an address to a value.  The AAM
+approach stops at the previous optimization.  However, the {\tt fn}
+continuation stores a value, and this makes the space of continuations
+quadratic rather than linear in the size of the program, for a
+monovariant analysis like 0-CFA.  Having the space of continuations
+grow linearly with the size of the program will drop the overall
+complexity to cubic (as expected).
 
@@
 space grows super-exponentially. We extend the semantics to
-additionally allocate addresses to store the values of each
-subexpression of an application. The continuations have to store these
-addresses to remember where to get argument values from, though
-sometimes your allocation strategy may allow you to elide that
-representation.
-
-The new evaluation rules follow (where $\mcntr' = \tick(\mlab,\mcntr)$).
+additionally allocate an address for the function value when creating
+the $\kfn{}$ continuation. The continuation has to store this address
+to remember where to get values from, though sometimes your allocation
+strategy may allow you to elide that representation.
+
+The new evaluation rules follow (where $\mcntr' = \tick(\mstate)$).
 % In theory, this aggressive distinction between continuations might buy
@@
 \begin{align*}
-\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
-\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar{\mexpi{1},\menv,\maddr_f,\maddr_a,\maddr_\mkont}} \\
+\co[^\mcntr]{\kar{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_\mkont}} \\
 \text{ where }
-  \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) \\
-  \maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
-\end{align*}
-This rule describes how we allocate addresses for subexpressions of an
-application up front. This adds another parameter to the machine, but
-it is very much like $\alloc$. We use $f$ and $a$ subscripts to denote
-function and argument positions, respectively. Whereas ANF names all
-intermediate values in the syntax, we allocate addresses for
-intermediate values. This representational leap is not a big one, but
-is notably important for analyzing direct style terms.
-
-\begin{align*}
-\co[^\mcntr]{\kar{\mexp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \machstep
-\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_a,\maddr_\mkont}} \\
-\text{ where }
+  \maddr_f &= \alloc(\mstate) \\
   \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
 \\
-\co[^\mcntr]{\kfn{\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \machstep
-\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr_a,\mkont,\msto'}
-\\
-\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f) \\
-\text{ where }
- \msto' &= \ext{\msto}{\maddr_a}{\{\mval\}}
+\co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
+\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
+\\
+\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f)
 \end{align*}
@@
 \emph{strict position} (more about that later).
-
-\begin{align*}
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\maddr_a, \msto,\mkont} & \machstep
-\ev[^{\mcntr'}]{\mexp,\menv',\msto',\mkont} \\
-\text{ where }
-\maddr &= \alloc(\mstate) \\
-\menv' &= \menv[\mvar\mapsto\maddr] \\
-\msto' &= \msto\sqcup[\maddr \mapsto \msto(\maddr_a)]
-\end{align*}
-Arguments are not passed as values, but as addresses to their values.
-
-\begin{align*}
-\ap[^\mcntr_\mlab]{\mop,\maddr_a,\msto,\mkont} & \machstep
-\co[^{\mcntr'}]{\mkont,\mval',\msto}
-\\
-\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\maddr_a)\}
-\end{align*}
-Primitive application still nondeterministically produces all possible values.
 
@@
 %
-We have two broad categories of these optimizations: (1) transition elimination
-and (2) pragmatic improvement.
-%
+We have two broad categories of these optimizations: (1) pragmatic
+improvement, (2) transition elimination.
+%
+The pragmatic improvements reduce overhead and trade space for time
+by utilizing:
+\begin{enumerate}
+ \item timestamped stores;
+ \item store deltas; and
+ \item imperative, preallocated data structures.
+\end{enumerate}
 The transition-elimination optimizations reduce the overall number of transitions
@@
 \begin{enumerate}
+  \setcounter{enumi}{3}
  \item lazy non-determinism and
@@
 \end{enumerate}
-The pragmatic improvements reduce overhead and trade space for time
-by utilizing:
+
+All pragmatic improvements are precision preserving (form complete
+abstractions), but the semantic changes are not in some cases, for reasons we
+will describe. We did not observe the precision differences in our evaluation.
+
+The move to the imperative will be made last in order to show the
+effectiveness of these techniques in the purely functional realm.
+
+\subsection{Timestamping monotonicity}
+
+The frontier-based semantics of the baseline has a severe bottleneck:
+checking if a state has been explored or not. Given two states,
+checking equality is expensive because the stores within each are
+large, and every entry must be checked against every other. Hashes can
+sometimes rule out inequality relatively quickly, but the incidence of
+collisions and actual equality is costly.
+
+And, there is a better way. Shivers' original work on $k$-CFA was
+susceptible to the same problem, and he suggested three complementary
+optimizations: (1) make the store global; (2) update the store
+imperatively; and (3) associate every change in the store with a
+version number -- its timestamp. Then, put timestamps in states
+where previously there were stores. Given two states, the analysis can
+now compare their stores just by comparing their timestamps -- a
+constant-time operation.
+
+There are two subtle losses of precision in Shivers' original
+time-stamp technique that we can fix.
+
 \begin{enumerate}
-  \setcounter{enumi}{2}
- \item store deltas;
- \item timestamped stores; and
- \item preallocated data structures.
+\item{In our semantics, the store does not change until the entire
+    frontier has been explored. This avoids cross-branch pollution
+    which would not otherwise happen, e.g., when one branch writes to
+    address $\maddr$ and another branch reads from address
+    $\maddr$.}
+\item{The common implementation strategy for timestamps destructively
+    updates each state's timestamp. This loses \emph{temporal}
+    information about the contexts a state is visited in, and in what
+    order. The semantics is different - we can't precisely answer
+    questions about the traces of the reduction relation we defined.
+    Our semantics has a drop-in replacement of timestamps for stores
+    in the seen set ($S$), so we do not experience precision loss.}
 \end{enumerate}
-
-All techniques are precision preserving under reasonable
-assumptions. We will discuss these assumptions as the caveats arise.
 
@@
 %
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto}
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto}
 &\lmachstep\;
-\ev[^\mcntr]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont}} \\
+\ev[^\mcntr]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
 \text{ where }
+\maddr_f &= \alloc(\mstate) \\
 \msto' &=
@@
 later and another binding not made in that step, the results of the
-two approaches are the same.
-}{}
+two approaches are the same. Without a widened store, lazy
+non-determinism is not a complete abstraction with either approach,
+but the precision gains are not immediately apparent due to fake
+rebinding (a second reference to the same variable chooses a different
+value from the first) and other problems. The performance and memory
+gains are apparent, so we favor the simplicity over invisible
+gains.}{}
 
@@
 \spchoice
-{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto_4,\saddr{\menv(\mvar)}}}
-{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto_4,\superposition{\msto_4(\menv(\mvar))}}}
+{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
+{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
 \end{align}
@@
 \spchoice
-{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto'_4,\saddr{\menv(\mvar)}}}
-{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto'_4,\superposition{\msto'_4(\menv(\mvar))}}}
+{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\saddr{\menv(\mvar)},\msto'_4}}
+{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
 \end{align*}
@@
 from the previous semantics is that $\ev{}$ state construction is
-replaced by calling the compiled expression.
+replaced by calling the compiled expression. For notational coherence,
+we write $\lambda^\mcntr(\mathit{args} \ldots)$ for
+$\lambda(\mathit{args} \ldots, \mcntr)$ and
+$\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
+\mcntr)$.
 
@@
 \begin{align*}
-\compile{\_} &: \Expr \to \Env \times \Store \times \Kont \to \State \\
+\compile{\_} &: (\Expr \to \Env \times \Store \times \Kont \times \Counter) \to \State \\
+\mcntr' &= \tick(\mlab,\menv,\msto,\mcntr) \\
 \compile{\svar\mvar} &=
- \lambda(\menv,\msto,\mkont) .
+ \lambda^\mcntr(\menv,\msto,\mkont) .
 \spchoice
@@
 \\
-\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
+\compile{\slit\mlit} &= \lambda^\mcntr(\menv,\msto,\mkont) .
 \co{\mkont,\mlit,\msto}
 \\
-\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
-\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
+\compile{\slam\mvar\mexp} &= \lambda^\mcntr(\menv,\msto,\mkont) .
+\co{\mkont,\clos{\mvar,\compile{\mexp},\menv},\msto}
 \\
 \compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_f,\maddr_a,\maddr_\mkont})
-\\
-&\setlength\arraycolsep{5pt}\begin{array}{lrl}\text{ where } & \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) 
-\\&
-\maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
+\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
+\\
+&\setlength\arraycolsep{5pt}
+\begin{array}{lrl}
+\text{ where } & \maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+               & \msto' = \msto\sqcup[\maddr_\mkont \mapsto \setof{\mkont}]
 \end{array}
@@
 \compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
-\\
-&\text{ where }\maddr,\msto' = \push_\mlab^\mcntr(\msto,\mkont)
+\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
+\\
+&\text{ where }\maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr_\mkont \mapsto \setof{\mkont}]
 \end{align*}
@@
 \\
-\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \cmachstep
-\mcomp^\mcntr(\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont}) \\
-\text{ where } \msto' &= \msto\sqcup[\maddr_f \mapsto \spchoice{\force(\msto,\mval)}
+\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_\mkont},\mval,\msto} & \cmachstep
+\mcomp^\mcntr(\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}) \\
+\text{ where } \maddr_f &= \alloc(\mstate) \\
+               \msto' &= \msto\sqcup[\maddr_f \mapsto \spchoice{\force(\msto,\mval)}
                                                                {\force(\mval)}]
 \\
-\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \cmachstep
-\ap[^\mcntr_\mlab]{\mval,\maddr_a,\mkont,\msto'} \\
+\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont},\mval,\msto} & \cmachstep
+\ap[^\mcntr_\mlab]{\mval,\mval\mkont,\msto} \\
 \text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\maddr_\mkont) \\
-\text{ where } \msto' &= \msto\sqcup[\maddr_f \mapsto \spchoice{\force(\msto,\mval)}
-                                                               {\force(\mval)}]
 \\
@@
 %% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\maddr_a,\msto,\mkont} & \cmachstep
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \cmachstep
 \mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
 \text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
-              \msto' &= \msto\sqcup[\maddr\mapsto\msto(\maddr_a)]
+              \msto' &= \msto\sqcup[\maddr\mapsto \spchoice{\force(\msto,\mval)}{\force(\mval)}]
 \\
@@
 \paragraph{Wide store and abstract compilation}
-A small problem arises if we try to widen the store as in section
-\ref{sec:storewiden}. In one step, we can both bind a variable and
-reference it. The problem is that the store does not change until all
-steps are done, so the reference unsoundly returns the previous
-binding (or nothing at all)! There are a couple ways to address
-this. One way is to make all compiled forms that reference the store
-into a new kind of state that has to be stepped in order to reference
-the store and continue. Another way is to make function application
-take one extra step after binding, so that running the body of the
-function sees the changed store. The latter is the more enticing of
-the two, since programs have more variable references than function
-calls - we don't want to slow them down if we don't have to.
-
 The fact that many changes happen in one step also makes the
@@
 is straightforward:
-% TODO: Turn the recursion here into ...:
 \begin{align*}
@@
 %
-Lookups will never rely on the change log, and can use the originally supplied store unmodified:
+We maintain the invariant that lookups will never rely on the change log, so we can use the originally supplied store unmodified:
 
@@
 \end{align*}
+
 Compilation changes to additionally take a $\msdiff$ component, so the
@@
 
-We also lift $\dmachstep$ to accommodate for this asymmetry
+\spchoice{ }{
+\paragraph{Caveat:} With abstract compilation, in one step, we can
+both bind a variable and reference it. This breaks the invariant of
+not relying on the change log. The problem is that the store does not
+change until all steps are done, so the reference unsoundly returns
+the previous binding (or nothing at all)! There are a couple ways to
+address this. One way is to make all compiled forms that reference the
+store into a new kind of state that has to be stepped in order to
+reference the store and continue. Another way is to make function
+application take one extra step after binding, so that running the
+body of the function sees the changed store. The latter is the more
+enticing of the two, since programs have more variable references than
+function calls - we don't want to slow them down if we don't have
+to. This change is also minutely different in precision, so we don't
+have complete abstraction. If the lazy semantics delayed the lookup of
+a variable's address to put in superposition (i.e., produced
+$\saddr{\menv(\mvar)}$ instead of
+$\superposition{\msto(\menv(\mvar))}$, and $\force$ additionally takes
+a $\msto$), then the lazy semantics loses its complete abstraction,
+but this current caveat disappears.
+}
+
+We lift $\dmachstep$ to accommodate for the asymmetry
 in the input and output. For each state that is stepped, we feed the
@@
 (S, F, \msto) &\mathrel{\damachstep} (S \cup S', F', \msto') \\
-\mbox{ where } (F', \msto') &= \step^*(\varnothing, F, \msto) \\
+\mbox{ where }
+ (F', \msdiff') &= \step^*(\varnothing, F, \msto, \epsilon) \\
+ \msto' &= \replay(\msdiff',\msto) \\
                S' &= \setof{(c,\msto') \mid c \in F'} \\
-\step^*(F', \varnothing, \msto) &= (S, \msdiff) \\
-\step^*(F', \{c\}\cup F, \msto) &= \step^*(F'\cup \states^*, F, \msto^*) \\
+\step^*(F', \varnothing, \msdiff) &= (F', \msdiff) \\
+\step^*(F', \{c\}\cup F, \msdiff) &= \step^*(F'\cup \states^*, F, \msdiff^*) \\
 \states^* &= \{c' \mid (c,\msto,\msdiff) \dmachstep (c',\msdiff^c) \} \\
-\msto^* &= \replayall(\{\msdiff^c \mid (c, \msto,\msdiff) \dmachstep (c', \msdiff^c)\}, \msto)
-\end{align*}
-
-Here $\replayall : \Set(\StoreDelta) \times \Store \to \Store$ replays all logs in a set on top of the given store. Since join is associative and commutative, the order does not matter:
-\begin{align*}
-  \replayall(\varnothing,\msto) &= \msto \\
-  \replayall(\setof{\msdiff} \cup \Xi, \msto) &= \replayall(\Xi, \replay(\msdiff, \msto))
-\end{align*}
+\msdiff^* &= \appendall(\setof{\msdiff^c \mid (c, \msto,\msdiff) \dmachstep (c', \msdiff^c)}) \\
+\appendall(\varnothing) &= \epsilon \\
+\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
+\end{align*}
+
+Here $\appendall$ combines change logs across all non-deterministic
+steps for a state to later be replayed. The order the combination
+happens in doesn't matter, because join is associative and
+commutative.
+
+% Here $\replayall : \Set(\StoreDelta) \times \Store \to \Store$ replays all logs in a set on top of the given store. Since join is associative and commutative, the order does not matter:
+% \begin{align*}
+%   \replayall(\varnothing,\msto) &= \msto \\
+%   \replayall(\setof{\msdiff} \cup \Xi, \msto) &= \replayall(\Xi, \replay(\msdiff, \msto))
+% \end{align*}
 
@@
 \begin{lemma}[Compile coherence]
-If $\msto \equiv \msto^*$ and $\mkont \equiv \mkont^*$, then
-$\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$ iff
-$\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \mstate$ such that
-$\mastate,\replay(\msdiff',\msto) \equiv \nw(\mstate)$
+Let $\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$
+and $\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \wn(\mastate',\msto^{*'})$.
+If $\msto \equiv \msto^*$, $\msdiff \equiv \msdiff^*$, and $\mkont \equiv \mkont^*$, then
+$\mastate' \equiv \mastate$
+and there exists an $\msdiff''$ such that \\
+   $\replay(\msdiff'',\replay(\msdiff^*,\msto^*)) \equiv \replay(\msdiff',\msto)$.
 \end{lemma}
-Proof is by induction on $\mexp$.
-
-\begin{theorem}[Equivalence]
+Here $\equiv$ means structurally the same, only compiled expressions
+are compiled with the updated compilation function that builds store
+deltas. Proof is by induction on $\mexp$.
+
+\begin{theorem}[Equivalence with wide abstract compilation]
 If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
@@
 \end{theorem}
-Here $\equiv$ means structurally the same, only compiled expressions
-are compiled with the updated compilation function that builds store
-deltas. The proof follows by equational reasoning and the previous lemma.
+Proof by the above lemma and that replaying in any order is equivalent
+(associativity and commutativity of $\sqcup$).
 
@@
 
-\subsection{Timestamping an imperative store}
+\subsection{Imperative, preallocated data structures}
 
 Thus far, we have made our optimizations in a purely functional
-manner. For the next series of optimizations, we need to dip into the
-imperative. We can motivate this entire sequence of optimizations by
-focusing on the largest bottleneck in the current state-space
-exploration algorithm: checking to see if a state has already been
-seen. Given two states, checking equality is expensive because the
-stores within each are large, and every entry must be checked against
-every other. Hashes can sometimes rule out inequality relatively
-quickly, but the incidence of collisions and actual equality is
-costly.
-
-And, there is a better way. Shivers' original work on $k$-CFA was
-susceptible to the same problem, and he suggested three complementary
-optimizations: (1) make the store global; (2) update the store
-imperatively; and (3) associate every change in the store with a
-version number -- its timestamp. Then, put timestamps in states
-where previously there were stores. Given two states, the analysis can
-now compare their stores just by comparing their timestamps -- a
-constant-time operation.
-
-There are two subtle losses of precision in Shivers' original
-time-stamp technique that we can fix.
-
-\begin{enumerate}
- \item{For a given abstract
-state, all writes to the global store should be delayed until the
-analysis considers all current work. This avoids
-cross-branch pollution which would not otherwise happen, e.g., when
-one branch writes to address $\maddr$ and another branch reads from
-address $\maddr$. Fortunately, given our conversion to log-based
-stores this change is straightforward.}
-\item{The common implementation strategy for timestamps destructively
-    updates each state's timestamp. This loses \emph{temporal}
-    information about the contexts a state is visited in, and in what
-    order. The semantics is different - we can't precisely answer
-    questions about the traces of the reduction relation we defined. A
-    simple fix is to keep a stack of the timestamps seen at each
-    state, and keep a list of stores corresponding with each
-    timestamp. A more space-efficient representation that also allows
-    an imperative store would be timestamping each update to the store
-    in the same style as states. This changes the notion of store
-    equality, but the new definition is straightforward. For analysis
-    consumers that disregard temporal information, the space cost in
-    unwarranted, but the realization that there is precision to be
-    gained opens possibilities.}
-\end{enumerate}
+manner. For the final push for performance, we need to dip into the
+imperative. In this section, we show an alternative representation of
+the store and seen set that are more space-efficient and are amenable
+to destructive updates.
 
@@
 
+We increase the timestamp only if we find that the store has changed
+from any join operation. The frontier set only gets states that we
+haven't seen at the next timestamp. The states we've seen gets updated
+with the frontier states at the next timestamp. We prove semantic
+equivalence with the previous semantics with a lock-step bisimulation with the stack of stores abstraction,
+which follow from equational reasoning from the following lemmas:
+
+\begin{lemma}[Store of value stacks completely abstracts stack of stores]
+TODO
+\end{lemma}
+
+\begin{lemma}[$\changep$ means change]
+  If $\replayall_t(\Xi,\msto,\changep) = \msto',\changep'$ then
+  $\msto' = \replay(\appendall(\Xi),\msto)$ and $\changep \vee
+  (\changep' \iff \msto \neq msto')$.
+\end{lemma}
+By induction on the size of $\Xi$.
+
+\begin{theorem}[Timestamped system with stack of stores completely abstracts wide store deltas]
+TODO
+\end{theorem}
+
 \subsubsection{Pure to imperative}
@@
 
-% \begin{figure}\label{fig:pure-timestamp}
-% \begin{align*}
-% t \in \Timestamp &= {\mathbb N} \\
-% \msto \in \Store &= \Addr \to (\Timestamp \times \wp(\Storeable))^* \\
-% (S,F,\msto,t) &\tmachstep (S \cup S', F', \msto', t') \\
-% \text{ where } I &= \{(\mstate,\msdiff) \mid \mstate \in F, \mstate \
-% \end{align*}
-% \end{figure}
-
-% \begin{figure}
-% \begin{alltt}
-% eval(\(e\)) :=
-%   \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
-%   \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \mtlst, \kmt\))
-%   while(true):
-%     if todo = \(\varnothing\): return (keys(seen), \(\msto\))
-%     else:
-%       let old := todo
-%       todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
-%       foreach c \(\in\) old: \(\changep\) := false; c()
-%       unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
-% \end{alltt}
-% \caption{Imperative algorithm}
-% \label{fig:imperative}
-% \end{figure}
-
-\subsection{Pre-allocating the store}
+\subsubsection{Pre-allocating the store}
 
@@
 \end{figure*}
+
+We evaluated the precision of these techniques with a singleton
+variable analysis to find opportunities to inline constants and closed
+functions. We found no change in the results across all
+implementations, including Shivers' timestamp approximation.
 

### 2013-03-28 paper/paper.tex
Reordering and semantics changes complete. Paper way too long.

@@
               \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
+\wn &: \widehat{\State} \times \Store \to \State \\
 \wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
@@
 In effect, the new store is computed as the least upper bound of all
-stores after stepping. This definition is wasteful, however. It causes
-all states found so far to step each iteration, even if they are not
-revisited. This has negative performance \emph{and} precision
-consequences. We instead use a frontier-based semantics that
-corresponds to the classic ``worklist'' algorithms for analysis. The
-difference is that the store is not modified in-place, but updated
-after all frontier states have been processed. This has implications
-for the analysis' precision and determinism. Specifically, higher
-precision, and it is deterministic even if set iteration is not.
-
-\begin{align*}
-(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
-\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
-              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
-              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
-\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
-\text{ where } c_0 &= \ev{e,\bot,\kmt}
-\end{align*}
-
-% Notice that now $S$ has several copies of the abstract store in it. As
-% it is, this semantics is much less efficient (but still more precise)
-% than the previously proposed semantics because membership checks have
-% to compare entire stores. We return to this in section
-% \ref{sec:timestamping}.
-
-For clarity, we will exhibit non-widened semantics unless a technique
-specifically requires it. The final step to the baseline takes the
-complexity from quartic to cubic in the monovariant case.
+stores after stepping. For clarity, we will exhibit non-widened
+semantics unless a technique specifically requires it. The final step
+to the baseline takes the complexity from quartic to cubic in the
+monovariant case.
 
@@
   \setcounter{enumi}{3}
+ \item frontier-based semantics
  \item lazy non-determinism and
@@
 
-The move to the imperative will be made last in order to show the
-effectiveness of these techniques in the purely functional realm.
-
-\subsection{Timestamping monotonicity}
-
-The frontier-based semantics of the baseline has a severe bottleneck:
-checking if a state has been explored or not. Given two states,
-checking equality is expensive because the stores within each are
+We apply the frontier-based semantics combined with timestamped stores
+as our first step.  The move to the imperative will be made last in
+order to show the effectiveness of these techniques in the purely
+functional realm.
+
+\subsection{Timestamped frontier}
+
+The semantics given for store widening in section \ref{sec:storewiden},
+while simple, is wasteful. It also does not model what typical
+implementations do. It causes all states found so far to step each
+iteration, even if they are not revisited. This has negative
+performance \emph{and} precision consequences (changes to the store
+can travel back in time in straight-line code). We instead use a
+frontier-based semantics that corresponds to the classic ``worklist''
+algorithms for analysis. The difference is that the store is not
+modified in-place, but updated after all frontier states have been
+processed. This has implications for the analysis' precision and
+determinism. Specifically, higher precision, and it is deterministic
+even if set iteration is not.
+
+\begin{align*}
+(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
+\mbox{where } I &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
+              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
+              F' &= \{c \mid (c,\_) \in I, (c,\msto') \notin S\} \\
+              S' &= \setof{(c,\msto') \mid c \in F'} \\
+\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
+\text{ where } c_0 &= \ev{e,\bot,\kmt}
+\end{align*}
+
+Notice that now $S$ has several copies of the abstract store in it. As
+it is, this semantics is much less efficient (but still more precise)
+than the previously proposed semantics because membership checks have
+to compare entire stores. Checking equality is expensive because the stores within each are
 large, and every entry must be checked against every other. Hashes can
@@
 There are two subtle losses of precision in Shivers' original
-time-stamp technique that we can fix.
+timestamp technique that we can fix.
 
@@
 
+\begin{align*}
+\Sigma &\in \Store^* \\
+S &\subseteq {\mathbb N} \times \widehat{\State} \\
+F &\subseteq \widehat{\State} \\
+(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
+\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
+              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
+              (t',\Sigma') &=\left\{\begin{array}{ll}
+                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
+                           (t,\Sigma)   & \text{ otherwise}
+                          \end{array}\right. \\
+              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
+              S' &= \setof{(c,t') \mid c \in F'}\\
+\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
+\text{ where } c_0 &= \ev{e,\bot,\kmt}
+\end{align*}
+
+The observation Shivers made was that the store is increasing
+monotonically, so all stores throughout execution will be totally
+ordered (form a chain). This observation allows you to replace stores
+with pointers into this chain. We keep the stores around in $\Sigma$
+to acheive a complete abstraction. This corresponds to the temporal
+information about the execution's effect on the store.
+
+Note also that $F$ is only populated with states that have not been
+seen at the resulting store. This is what produces the more precise
+abstraction than the baseline widening.
+
+The general fixed-point combinator we showed above can be specialized
+to this semantics, as well. In fact, $\widehat{\machstep}^T$ is a functional
+relation, so we can get the least fixed-point of it directly.
+
+\begin{lemma}
+  $\widehat{\machstep}$ maintains the invariant that all stores in $S$ are
+  totally ordered and $\msto$ is an upper bound of the stores in $S$.
+\end{lemma}
+
+\begin{lemma}
+  $\widehat{\machstep}^T$ maintains the invariant that $\Sigma$ is in
+  order with respect to $\sqsupset$ and $\msto = \hd(\Sigma)$.
+\end{lemma}
+
+\begin{theorem}
+Timestamps are a complete abstraction
+\end{theorem}
+The proof follows from the galois connection that, in one direction,
+sorts all the stores in $S$ to form $\Sigma$, and translates stores in
+$S$ to their distance from the end of $\Sigma$. In the other
+direction, timestamps are replaced by the stores they point to.
+
+\subsection{Locally log-based store deltas}
+
+The above technique requires joining entire (large) stores
+together. Additionally, there is still a comparison of stores, which
+we established is expensive. Not every step will modify all addresses
+of the store, so joining entire stores is wasteful in terms of memory
+and time. We can instead log store changes and replay the change log
+on the full store after all steps have completed, noting when there is
+an actual change. This uses far fewer join and comparison operations,
+leading to less overhead, and is precision-preserving.
+
+We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
+  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
+ becomes a log addition
+$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
+empty ($\mtlst$) for each step. Applying the changes to the full store
+is straightforward:
+\begin{equation*}
+\replay : (\StoreDelta \times \Store) \to (\Store \times \Boolean)
+\end{equation*}
+\begin{align*}
+\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &=
+\ttuple{\msto'}
+       {\mval{s_i} \overset{?}{=} \msto'(\maddr_i) \vee \ldots} \\
+\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots}
+\end{align*}
+
+We change the semantics slightly to add to the change log rather than
+produce an entire modified store.  The transition relation is
+identical except for the addition of this change log.  We maintain the
+invariant that lookups will never rely on the change log, so we can
+use the originally supplied store unmodified.
+
+A taste of the changes to the reduction relation is as follows:
+
+\begin{align*}
+\dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \dmachstep_\msto
+\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
+\text{ where }\maddr &= \alloc(\mstate) \\
+              \menv' &= \menv[\mvar\mapsto\maddr]
+\end{align*}
+
+% Compilation changes to additionally take a $\msdiff$ component, so the
+% above rule's right hand side would instead be $k^{\mcntr'}(\menv',
+% \msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.
+
+We lift $\dmachstep$ to accommodate for the asymmetry
+in the input and output, and change the frontier-based semantics in the following way:
+
+\begin{align*}
+(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
+\mbox{ where }
+ I &= \setof{(\mastate',\msdiff) \mid \mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff} } \\
+ \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
+ \ttuple{t'}{\Sigma'} &=
+     \left\{
+       \begin{array}{ll}
+         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
+         \ttuple{t}{\Sigma} & \text{ otherwise}
+       \end{array}\right. \\
+ F' &= \setof{\mastate \mid (\mastate,\_) \in I, (\mastate,t') \notin S} \\
+ S' &= \setof{(\mastate,t') \mid c \in F'} \\
+\appendall(\varnothing) &= \epsilon \\
+\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
+\end{align*}
+
+Here $\appendall$ combines change logs across all non-deterministic
+steps for a state to later be replayed. The order the combination
+happens in doesn't matter, because join is associative and
+commutative.
+
+\begin{lemma}
+$\mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff}$ iff
+$\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
+\end{lemma}
+By cases on $\dmachstep$ and $\machstep$.
+
+\begin{lemma}[$\changep$ means change]
+Let $\replay(\msdiff,\msto) = \ttuple{\msto'}{\changep}$. $\msto' \neq \msto$ iff $\changep$.
+\end{lemma}
+By induction on $\msdiff$.
+
+\begin{theorem}
+$\damachstep$ is a complete abstraction of $\widehat{\machstep}^T$.
+\end{theorem}
+Follows from previous lemma and that join is associative and commutative.
+
+% Correctness here requires a lemma that relates the differences in the compilation functions.
+
+% \begin{lemma}[Compile coherence]
+% Let $\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$
+% and $\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \wn(\mastate',\msto^{*'})$.
+% If $\msto \equiv \msto^*$, $\msdiff \equiv \msdiff^*$, and $\mkont \equiv \mkont^*$, then
+% $\mastate' \equiv \mastate$
+% and there exists an $\msdiff''$ such that \\
+%    $\replay(\msdiff'',\replay(\msdiff^*,\msto^*)) \equiv \replay(\msdiff',\msto)$.
+% \end{lemma}
+% Here $\equiv$ means structurally the same, only compiled expressions
+% are compiled with the updated compilation function that builds store
+% deltas. Proof is by induction on $\mexp$.
+
+% \begin{theorem}[Equivalence with wide abstract compilation]
+% If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
+% $(S,F,\msto) \machstep (S',F',\msto')$ iff
+% $\exists S^*_1, F^*_1, \msto^*_1.
+%   S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
+%   (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
+% \end{theorem}
+% Proof by the above lemma and that replaying in any order is equivalent
+% (associativity and commutativity of $\sqcup$).
+
 \subsection{Lazy non-determinism}
@@
 until and unless it is needed in {\it strict contexts} (such as the guard for an
-{\tt if} a called procedure, or a numerical primitive application). 
+{\tt if}, a called procedure, or a numerical primitive application).
 %
@@
  \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
- \force(\mval) &= \{\mval\}}
+ \force(\msto,\mval) &= \setof{\mval}}
 {\force &: \Value \to \Set(\Value) \\
@@
 &\lmachstep\;
-\ev[^\mcntr]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
 \text{ where }
@@
 &\lmachstep\;
-\ev[^\mcntr]{\mexpi,\menv,\msto,\mkont} \\
+\ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
 \text{ if } \mkont &\in \msto(\maddr_\mkont),
@@
 \machstep\; &
-\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_{f1},\maddr_{a1},\maddr_1}}
+\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_1}}
 \\
 \machstep\; &
-\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_{f2},\maddr_{a2},\maddr_2}}
+\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_2}}
 \\
 \machstep\; &
-\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3}} % {\mexp_2}
+\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_3}} % {\mexp_2}
 \\
@@
 \spchoice
-{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
+{\co{\kar{\mexp_1,\menv,\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
 {\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
@@
 \begin{align*}
-\compile{\mexp}^\mcntr(\menv,\msto_0,\mkont) = 
+\compile{\mexp}^\mcntr(\msto_0)(\menv,\epsilon,\mkont) =
 \spchoice
-{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\saddr{\menv(\mvar)},\msto'_4}}
+{\co{\kar{\compile{\mexp_1},\menv,\maddr_3},\saddr{\menv(\mvar)}},\msdiff_4}
 {\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
 \end{align*}
-where \(
-\msto'_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
-[\maddr_2 \mapsto \kar{\compile{\mexp_3},\menv,\maddr_1}],
-[\maddr_3 \mapsto \kar{\compile{\mexp_2},\menv,\maddr_2}]\).
+where
+\begin{equation*}
+\msdiff_4 = [\ttuple{\maddr_1}{\setof{\mkont}},
+             \ttuple{\maddr_2}{\setof{\kar{\compile{\mexp_3},\menv,\maddr_1}}},
+             \ttuple{\maddr_3}{\setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}}]\text.
+\end{equation*}
 
@@
 \begin{align*}
-\compile{\_} &: (\Expr \to \Env \times \Store \times \Kont \times \Counter) \to \State \\
+\compile{\_} &: \Expr \to \Store \\
+             &\phantom{: \Expr } \to \Env  \times \StoreDelta \times \Kont \times \Counter \\
+             &\phantom{: \Expr } \to \State \\
 \mcntr' &= \tick(\mlab,\menv,\msto,\mcntr) \\
-\compile{\svar\mvar} &=
- \lambda^\mcntr(\menv,\msto,\mkont) .
+\compile{\svar\mvar}_\msto &=
+ \lambda^\mcntr(\menv,\msdiff,\mkont) .
 \spchoice
-{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
-{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}}
-\\
-\compile{\slit\mlit} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\co{\mkont,\mlit,\msto}
-\\
-\compile{\slam\mvar\mexp} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\co{\mkont,\clos{\mvar,\compile{\mexp},\menv},\msto}
-\\
-\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
+{\co{\mkont,\saddr{\menv(\mvar)}},\msdiff}
+{\co{\mkont,\superposition{\msto(\menv(\mvar))}},\msdiff}
+\\
+\compile{\slit\mlit}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
+\co{\mkont,\mlit},\msdiff
+\\
+\compile{\slam\mvar\mexp}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
+\co{\mkont,\clos{\mvar,\compile{\mexp},\menv}},\msdiff
+\\
+\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
+\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
 \\
@@
 \text{ where } & \maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-               & \msto' = \msto\sqcup[\maddr_\mkont \mapsto \setof{\mkont}]
+               & \msdiff' = \cons{\ttuple{\maddr_\mkont}{\setof{\mkont}}}{\msdiff}
 \end{array}
 \\
-\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
+\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
+\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
 \\
 &\text{ where }\maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr_\mkont \mapsto \setof{\mkont}]
+&\phantom{\text{ where }} \msdiff' = \cons{\ttuple{\maddr_\mkont}{\setof{\mkont}}}{\msdiff}
 \end{align*}
@@
 \begin{align*}
-\traces(\mexp) &= \{ \compile{\mexp}^\mtcntr(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
+\traces(\mexp) &= \setof{ \inject(\compile{\mexp}^\mtcntr_\bot(\bot,\epsilon,\kmt)) \multimachstep \mstate}
+                    \text{ where } \\
+\inject(\mastate,\msdiff) &= \wn(\mastate,\replay(\msdiff,\bot)) \\
+\wn(\mastate,\msto) \machstep \wn(\mastate',\msto') &\iff \mastate \cmachstep_\msto \mastate',\msdiff \\
+\msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
 \end{align*}
@@
 %% CONTINUE
-\co{\kmt,\mval,\msto} &\cmachstep
-\ans{\msto,\mval'} \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
-\\
-\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_\mkont},\mval,\msto} & \cmachstep
-\mcomp^\mcntr(\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}) \\
+\co{\kmt,\mval} &\cmachstep_\msto
+\ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
+\\
+\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_\mkont},\mval} & \cmachstep_\msto
+\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}) \\
 \text{ where } \maddr_f &= \alloc(\mstate) \\
-               \msto' &= \msto\sqcup[\maddr_f \mapsto \spchoice{\force(\msto,\mval)}
-                                                               {\force(\mval)}]
-\\
-\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont},\mval,\msto} & \cmachstep
-\ap[^\mcntr_\mlab]{\mval,\mval\mkont,\msto} \\
-\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\maddr_\mkont) \\
-\\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \cmachstep
-\mcompi0^\mcntr(\menv,\msto,\mkont)
+               \msdiff &= \cons{\ttuple{\maddr_f}{\force(\msto,\mval)}}{\epsilon}
+\\
+\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont},\mval} & \cmachstep_\msto
+\ap[^\mcntr_\mlab]{\mval,\mval\mkont},\epsilon \\
+\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\maddr_\mkont)
+\\
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue} & \cmachstep_\msto
+\mcompi{0}^\mcntr(\msto)(\menv,\epsilon,\mkont)
 \text{ if }\mkont\in\msto(\maddr)
 \\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \cmachstep
-\mcompi1^\mcntr(\menv,\msto,\mkont)
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse} & \cmachstep_\msto
+\mcompi{1}^\mcntr(\msto)(\menv,\epsilon,\mkont)
 \text{ if }\mkont\in\msto(\maddr)
@@
 %% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \cmachstep
-\mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\mkont} & \cmachstep_\msto
+\mcomp^{\mcntr'}(\msto)(\menv',\msdiff,\mkont) \\
 \text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
-              \msto' &= \msto\sqcup[\maddr\mapsto \spchoice{\force(\msto,\mval)}{\force(\mval)}]
-\\
-\ap{\mop,\mval,\msto,\mkont} & \cmachstep
-\co{\mkont,\mval',\msto} \\
+              \msdiff &= \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon}
+\\
+\ap{\mop,\mval,\mkont} & \cmachstep_\msto
+\co{\mkont,\mval'},\epsilon \\
 \text{ where } \mvalx{u} &\in \spchoice{\force(\msto,\mval)}{\force(\mval)}, \mval'\in\interpdelta(\mop,\mvalx{u})
@@
 semantics line up again (modulo representation of expressions). This
-constitutes the use of a notion of stuttering. We therefore prove a
-stuttering bisimulation~\cite{ianjohnson:BCG88} between this
-semantics and the previous section's to show precision preservation.
+constitutes the use of a notion of stuttering. We provide a formal
+analysis of abstract compilation \emph{without} store widening with a proof
+of a stuttering bisimulation~\cite{ianjohnson:BCG88} between this
+semantics and lazy non-determinism without widening to show precision
+preservation. We revisit store widening at the end of the section.
 
@@
 %
-\begin{lemma}[Compile/commit]\hfill\\
-$\compile{\mexp}^\mcntr(\menv,r(\msto),r(\mkont)) = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
+\begin{lemma}[Compile/commit]
+Let $\mastate,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
+Let $\wn(\mastate',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
+$\wn(\mastate,\replay(\msdiff',\msto)) = \wn(\mastate',\replay(\msdiff,\msto'))$.
 \end{lemma}
@@
 \begin{theorem}[Precision preservation]
-$B$ is a WEB on $\lmachstep \cup \cmachstep$
+$B$ is a WEB on $\lmachstep \cup \machstep$
 \end{theorem}
 
-The full proof is in the appendix, but it follows by cases on
-$\lmachstep \cup \cmachstep$ with the WEB \emph{witness} being the
-well-order on expressions (with a $\bot$ element), and the following
-$\erankt$, $\erankl$ functions:
+The proof follows by cases on $\lmachstep \uplus \machstep$ with the
+WEB \emph{witness} being the well-order on expressions (with a $\bot$
+element), and the following $\erankt$, $\erankl$ functions:
 
@@
 values in the store, leading to a mismatch from the previous
-semantics. This kind of cross-step store poisoning is inevitable when a
-store is shared between several traces. The binding could have just as
-easily been observed the other way around in the non-compiled
-semantics.
+semantics. This kind of cross-step store poisoning is inevitable when
+a store is shared between several traces. The binding could have just
+as easily been observed the other way around in the non-compiled
+semantics. Call $\camachstep$ the result of the widening operator from
+the previous section on $\cmachstep$.
 
@@
 %\newpage
-
-\subsection{Locally log-based store deltas}
-
-Every step the analysis makes for the above techniques requires
-joining entire (large) stores together. Not every step will modify all
-addresses of the store, so joining entire stores is wasteful in terms
-of memory and time. We can instead log store changes and replay the
-change log on the full store after all steps have completed. This uses
-far fewer join operations, leading to less overhead, and is precision-preserving.
-
-We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
-  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
- becomes a log addition
-$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
-empty ($\mtlst$) for each step. Applying the changes to the full store
-is straightforward:
-\begin{align*}
-\replay &: \StoreDelta \times \Store \to \Store \\
-\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &= \msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots
-\end{align*}
-%\replay(\mtlst, \msto) &= \msto \\
-%\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
-
-The transition relation is identical except for the addition of this change log.
-%
-We maintain the invariant that lookups will never rely on the change log, so we can use the originally supplied store unmodified:
-
-\begin{align*}
-(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \dmachstep
-(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
-\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
-              \msdiff' &= 
-\spchoice
-{\cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\msdiff}}
-{\cons{\ttuple{\maddr}{\force(\mval)}}{\msdiff}}
-\end{align*}
-
-Compilation changes to additionally take a $\msdiff$ component, so the
-above rule's right hand side would instead be $k^{\mcntr'}(\menv',
-\msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.
-
-\spchoice{ }{
-\paragraph{Caveat:} With abstract compilation, in one step, we can
-both bind a variable and reference it. This breaks the invariant of
-not relying on the change log. The problem is that the store does not
-change until all steps are done, so the reference unsoundly returns
-the previous binding (or nothing at all)! There are a couple ways to
-address this. One way is to make all compiled forms that reference the
-store into a new kind of state that has to be stepped in order to
-reference the store and continue. Another way is to make function
-application take one extra step after binding, so that running the
-body of the function sees the changed store. The latter is the more
-enticing of the two, since programs have more variable references than
-function calls - we don't want to slow them down if we don't have
-to. This change is also minutely different in precision, so we don't
-have complete abstraction. If the lazy semantics delayed the lookup of
-a variable's address to put in superposition (i.e., produced
-$\saddr{\menv(\mvar)}$ instead of
-$\superposition{\msto(\menv(\mvar))}$, and $\force$ additionally takes
-a $\msto$), then the lazy semantics loses its complete abstraction,
-but this current caveat disappears.
-}
-
-We lift $\dmachstep$ to accommodate for the asymmetry
-in the input and output. For each state that is stepped, we feed the
-output changes to the next so that all changes get accumulated:
-
-\begin{align*}
-(S, F, \msto) &\mathrel{\damachstep} (S \cup S', F', \msto') \\
-\mbox{ where }
- (F', \msdiff') &= \step^*(\varnothing, F, \msto, \epsilon) \\
- \msto' &= \replay(\msdiff',\msto) \\
-               S' &= \setof{(c,\msto') \mid c \in F'} \\
-\step^*(F', \varnothing, \msdiff) &= (F', \msdiff) \\
-\step^*(F', \{c\}\cup F, \msdiff) &= \step^*(F'\cup \states^*, F, \msdiff^*) \\
-\states^* &= \{c' \mid (c,\msto,\msdiff) \dmachstep (c',\msdiff^c) \} \\
-\msdiff^* &= \appendall(\setof{\msdiff^c \mid (c, \msto,\msdiff) \dmachstep (c', \msdiff^c)}) \\
-\appendall(\varnothing) &= \epsilon \\
-\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
-\end{align*}
-
-Here $\appendall$ combines change logs across all non-deterministic
-steps for a state to later be replayed. The order the combination
-happens in doesn't matter, because join is associative and
-commutative.
-
-% Here $\replayall : \Set(\StoreDelta) \times \Store \to \Store$ replays all logs in a set on top of the given store. Since join is associative and commutative, the order does not matter:
-% \begin{align*}
-%   \replayall(\varnothing,\msto) &= \msto \\
-%   \replayall(\setof{\msdiff} \cup \Xi, \msto) &= \replayall(\Xi, \replay(\msdiff, \msto))
-% \end{align*}
-
-Correctness here requires a lemma that relates the differences in the compilation functions.
-\begin{lemma}[Compile coherence]
-Let $\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$
-and $\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \wn(\mastate',\msto^{*'})$.
-If $\msto \equiv \msto^*$, $\msdiff \equiv \msdiff^*$, and $\mkont \equiv \mkont^*$, then
-$\mastate' \equiv \mastate$
-and there exists an $\msdiff''$ such that \\
-   $\replay(\msdiff'',\replay(\msdiff^*,\msto^*)) \equiv \replay(\msdiff',\msto)$.
-\end{lemma}
-Here $\equiv$ means structurally the same, only compiled expressions
-are compiled with the updated compilation function that builds store
-deltas. Proof is by induction on $\mexp$.
-
-\begin{theorem}[Equivalence with wide abstract compilation]
-If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
-$(S,F,\msto) \machstep (S',F',\msto')$ iff
-$\exists S^*_1, F^*_1, \msto^*_1.
-  S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
-  (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
-\end{theorem}
-Proof by the above lemma and that replaying in any order is equivalent
-(associativity and commutativity of $\sqcup$).
 
@@
 
-For the purposes of space, we reuse the $\dmachstep$ semantics and change the definition $\replay$ to track if the store has changed, in order to increment the global timestamp.
-\begin{align*}
-\replay_t(\epsilon,\msto,\changep) &= \msto,\changep \\
-\replay_t(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff},\msto,\changep) &= \replay_t(\msdiff,\msto',\changep \vee \changep') \\
-\text{ where } (\msto',\changep') &= \msto\sqcup_t[\maddr \mapsto, \mval{s}] \\
-\replayall_t(\varnothing,\msto,\changep) &= \msto,\changep \\
-\replayall_t(\setof{\msdiff}\cup\Xi,\msto,\changep) &= \replayall_t(\Xi, \msto', \changep') \\
-\text{ where } (\msto',\changep') &= \replay_t(\msdiff,\msto,\changep)
-\end{align*}
-
-Because the store has all the temporal information baked into it, we
-rephrase the core semantics in terms of a transfer function. The least
-fixed-point of this function gives a more compact representation of
-the reduction relation of the previous section.
-
-\begin{align*}
-\System &= (\State \to {\Timestamp}^*) \times \wp(\State) \times \Store \times \Timestamp \\
+For the purposes of space, we reuse the $\cmachstep$ semantics,
+although the $\replay$ of the produced $\msdiff$ objects should be
+in-place, and the $\lookup$ function should be using this
+single-threaded store.  Because the store has all the temporal
+information baked into it, we rephrase the core semantics in terms of
+a transfer function. The least fixed-point of this function gives a
+more compact representation of the reduction relation of the previous
+section.
+
+\begin{align*}
+\System &= (\widehat{\State} \to {\Timestamp}^*) \times \wp(\widehat{\State}) \times \Store \times \Timestamp \\
 {\mathcal F} &: \System \to \System \\
@@
 \text{ where }
-I &= \setof{(\mstate',\msdiff') \mid
-       \mstate \in F,
-       (\mstate,\msto^*,\epsilon) \dmachstep
-       (\mstate',\msdiff')} \\
+I &= \setof{(\mastate',\msdiff) \mid
+       \mastate \in F,
+       \mastate \cmachstep_{\msto^*} \mastate',\msdiff} \\
 \msto^* &= \lambda \maddr.\lookup(\msto(\maddr),t) \\
-(\msto',\changep) &= \replayall_t(\setof{\msdiff \mid (\_,\msdiff) \in I},\msto,\sfalse) \\
+\ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 t' &= \left\{\begin{array}{ll} t+1 & \text{ if } \changep \\
@@
              \end{array}\right. \\
-F' &= \setof{\mstate \mid (\mstate,\_) \in I, \changep \vee S(\mstate) \neq \cons{t}{\_}} \\
-S' &= \lambda \mstate. \left\{\begin{array}{ll}
-                               \cons{t'}{S(\mstate)} & \text{ if } \mstate \in F' \\
-                               S(\mstate) & \text{ otherwise}
+F' &= \setof{\mastate \mid (\mastate,\_) \in I, \changep \vee S(\mastate) \neq \cons{t}{\_}} \\
+S' &= \lambda \mastate. \left\{\begin{array}{ll}
+                               \cons{t'}{S(\mastate)} & \text{ if } \mastate \in F' \\
+                               S(\mastate) & \text{ otherwise}
                              \end{array}\right.
@@
 
-We increase the timestamp only if we find that the store has changed
-from any join operation. The frontier set only gets states that we
-haven't seen at the next timestamp. The states we've seen gets updated
-with the frontier states at the next timestamp. We prove semantic
-equivalence with the previous semantics with a lock-step bisimulation with the stack of stores abstraction,
-which follow from equational reasoning from the following lemmas:
-
-\begin{lemma}[Store of value stacks completely abstracts stack of stores]
-TODO
+We prove semantic equivalence with the previous semantics with a
+lock-step bisimulation with the stack of stores abstraction, which
+follow from equational reasoning from the following lemmas:
+
+\begin{lemma}
+Stores of value stacks completely abstract stacks of stores.
 \end{lemma}
-
-\begin{lemma}[$\changep$ means change]
-  If $\replayall_t(\Xi,\msto,\changep) = \msto',\changep'$ then
-  $\msto' = \replay(\appendall(\Xi),\msto)$ and $\changep \vee
-  (\changep' \iff \msto \neq msto')$.
-\end{lemma}
-By induction on the size of $\Xi$.
-
-\begin{theorem}[Timestamped system with stack of stores completely abstracts wide store deltas]
-TODO
+This depends on some wellformedness conditions about the order of the
+stacks. The store of value stacks can be translated to a stack of
+stores by taking successive ``snapshots'' of the store at different
+timestamps from the max timestamp it holds down to 0. Vice versa, we
+replay the changes across adjacent stores in the stack.
+
+We apply a similar construction to the different representation of seen states in order to get the final result:
+
+\begin{theorem}
+${\mathcal F}$ is a complete abstraction of $\camachstep$.
 \end{theorem}

### 2013-03-28 paper/paper.tex
Change title, no author affiliations.

@@
 
-\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
-
-\authorinfo{J. Ian Johnson}
-           {Northeastern University}
-           {ianj@ccs.neu.edu}
-\authorinfo{Nicholas Labich}
-           {Northeastern University}
-           {labichn@ccs.neu.edu}
-\authorinfo{Matthew Might}
-           {University of Utah}
-           {might@cs.utah.edu}
-\authorinfo{David Van Horn}
-           {Northeastern University}
-           {dvanhorn@ccs.neu.edu}
+%\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
+\title{Systematically Optimizing Abstract Abstract Machines}
+
+\authorinfo{J. Ian Johnson \and Nicholas Labich \and Matthew Might \and David Van Horn}
+           {}
+           {}
+%% \authorinfo{J. Ian Johnson}
+%%            {Northeastern University}
+%%            {ianj@ccs.neu.edu}
+%% \authorinfo{Nicholas Labich}
+%%            {Northeastern University}
+%%            {labichn@ccs.neu.edu}
+%% \authorinfo{Matthew Might}
+%%            {University of Utah}
+%%            {might@cs.utah.edu}
+%% \authorinfo{David Van Horn}
+%%            {Northeastern University}
+%%            {dvanhorn@ccs.neu.edu}
 \maketitle

### 2013-03-29 paper/paper.tex
Some changes to intro.

@@
 \begin{abstract}
-Abstracting abstract machines has been proposed as a
-lightweight approach to designing sound and computable program
-analyses.  The approach derives abstract interpreters from
-existing machine semantics and has been applied to a variety of
-languages with features widely considered difficult to analyze.
-Although sound analyzers are straightforward to build under this approach,
-they are also prohibitively inefficient.
-
-This article contributes a step-by-step process for going from a
-\naive{} analyzer derived under the abstracting abstract machine
-approach to an efficient program analyzer.  The end result of the
-process is a two to three order-of-magnitude improvement over the
-systematically derived analyzer, making it competitive with
+The technique of \emph{abstracting abstract machines} (AAM) provides a
+systematic approach for deriving computable approximations of
+evaluators that are easily proved sound.
+%
+This article contributes a complementary step-by-step process for
+subsequently going from a \naive{} analyzer derived under the AAM
+approach, to an efficient and correct implementation.  The end result
+of the process is a two to three order-of-magnitude improvement over
+the systematically derived analyzer, making it competitive with
 hand-optimized implementations that compute fundamentally less precise
@@
 
+Program analysis provides sound predictive models of program behavior, but
+in order for such models to be effective, they must be efficiently
+computable and correct.  Past approaches to designing program analysis
+have often featured abstractions that are far removed from the
+original language semantics, requiring ingenuity in their construction
+and effort in their verification.
+%
 The \emph{abstracting abstract machines} (AAM)
 approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
-to deriving program analyses provides a systematic way of transforming a
-programming language semantics in the form of an abstract machine into a family
-of abstract interpreters.
-%
-The approach parameterizes these families with policies for regulating analytic
-precision.
-%
-While flexible and robust, the AAM approach unfortunately yields analyzers with
-poor performance relative to hand-optimized analyzers.
-%
-Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an
-equally systematic way.
+to deriving program analyses provides an alternative: a systematic way
+of transforming a programming language semantics in the form of an
+abstract machine into a family of abstract interpreters.  It thus reduces
+the burden of constructing and verifying the soundness of an abstract
+interpreter.
+%
 
@@
 
-However, there is a considerable weakness to the approach: an
-analyzer designed and implemented by following the AAM recipe is
-prohibitively inefficient without both further approximation and
-further implementation effort.
+Unfortunately, the AAM approach yields analyzers with poor performance
+relative to hand-optimized analyzers.
+%
+Our work takes aim squarely at this ``efficiency gap,'' and narrows it
+in an equally systematic way through a number of simple steps, many of
+which are inspired by run-time implementation techniques such as
+laziness and compilation to avoid interpretative overhead.  Each of
+these steps is proven correct, so the end result is an implementation
+that is trustworthy and efficient.
 
@@
 practical implementation of an abstract-machine-based analyzer using
-mostly semantic means - not tricky engineering. Our goal is to empower
-programming language implementers and researchers to explore and
-convincingly exhibit their ideas with a low barrier to entry. The
+mostly semantic means rather than tricky engineering. Our goal is to
+empower programming language implementers and researchers to explore
+and convincingly exhibit their ideas with a low barrier to entry. The
 optimizations we describe are widely applicable and apparently
-effective to scale far beyond toy examples.
+effective to scale far beyond the size of programs typically considered
+in the recent literature on flow analysis for functional languages.
 

### 2013-03-29 paper/paper.tex
Added changes we just discussed.

@@
 \begin{align*}
-\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
-\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
-\end{align*}
-where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
-\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
-\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
-$\mcntr$, but they can be used to determinize the choice of fresh
-addresses.  The resulting relation is non-deterministic in its choice
-of addresses, however it must always choose a fresh address when
-allocating a continuation or variable binding.  If we consider machine
-states equivalent up to consistent renaming and fix an allocation
-scheme, this relation defines a deterministic machine (the relation is
-really a function).
+\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
+\allockont^\mcntr_\mlab(\msto,\mkont) &=\maddr \mbox{ where } \maddr \notin \msto
+\end{align*}
+These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
+used to determinize the choice of fresh addresses. The $\sqcup$ on
+stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
+\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
+relation is non-deterministic in its choice of addresses, however it
+must always choose a fresh address when allocating a continuation or
+variable binding.  If we consider machine states equivalent up to
+consistent renaming and fix an allocation scheme, this relation
+defines a deterministic machine (the relation is really a function).
 
@@
 
-\section{Reduction semantics to baseline analyzer}
+\section{Machine semantics to baseline analyzer}
 \label{sec:baseline}
@@
   \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
-\\
+\end{align*}
+Now instead of storing the evaluated function in the continuation
+frame itself, we indirect it through the store for further control on
+complexity and precision.
+
+\begin{align*}
 \co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
@@
 \end{align*}
-We associate the value in hand with the expression that produced
-it. Also, we fan out to all functions that are associated with the
-original function position, because applied functions are in
+Associated with this indirection, we now apply all functions stored in
+the address. This nondeterminism is demanded in order to continue with
+evaluation. More generally, this is because applied functions are in
 \emph{strict position} (more about that later).
@@
 \dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \dmachstep_\msto
+\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
 \ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
@@
 \mbox{ where }
- I &= \setof{(\mastate',\msdiff) \mid \mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff} } \\
+ I &= \setof{(\mastate',\msdiff) \mid \ttuple{\mastate}{\msto} \dmachstep \ttuple{\mastate'}{\msdiff} } \\
  \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
@@
 \begin{lemma}
-$\mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff}$ iff
+$\ttuple{\mastate}{\msto} \dmachstep \ttuple{\mastate'}{\msdiff}$ iff
 $\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
@@
 semantics and lazy non-determinism without widening to show precision
-preservation. We revisit store widening at the end of the section.
+preservation. %We revisit store widening at the end of the section.
 
@@
 \begin{theorem}[Precision preservation]
-$B$ is a WEB on $\lmachstep \cup \machstep$
+$B$ is a WEB on $\lmachstep \uplus \machstep$
 \end{theorem}
@@
 \item{$F$: the \emph{frontier} set, which must be persisent or copied for the iteration through the set to be correct.}
-\item{$\msto$: the store, which represents all stores that occur in the reduction semantics.}
+\item{$\msto$: the store, which represents all stores that occur in the machine semantics.}
 \item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}

### 2013-03-29 paper/paper.tex
More text fix ups.

--- 2013-03-29 paper/paper.tex
@@
 %\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
-\title{Systematically Optimizing Abstract Abstract Machines}
+\title{Correctly Optimizing Abstract Abstract Machines}
 
@@
 result of analyzing a program is a finite directed graph where nodes
-in the graph are machine states and edges denote machine
+in the graph are (abstract) machine states and edges denote machine
 transitions between states.
@@
 %\newpage
-\section{An analyzer for ISWIM}
+\section{Abstract interpretation of ISWIM}
 \label{sec:aam}
@@
 The semantics are defined in terms of a machine model.  The machine
-components are defined in figure~\ref{fig:domains} and
+components are defined in figure~\ref{fig:domains};
 figure~\ref{fig:aam} defines the transition relation.  The evaluation
-of a program is defined as its traces given by the machine transition relation.  The
+of a program is defined as its set of traces that arise from iterating the
+machine transition relation.  The
 machine is a very slight variation on a standard abstract machine for
@@
 \begin{itemize}
-\item the store maps addresses to \emph{sets} of values, not
+\item the store maps addresses to \emph{sets} of values,\footnote{
+More generally, we can have stores map to any domain that forms
+ a Galois connection with values, enabling $\interpdelta$ to
+produce elaborate abstractions of base values (e.g., interval or octagon
+abstractions). We use sets of values for a simpler exposition.
+} not
   single values,
+
 \item continuations are heap-allocated, not stack-allocated,
@@
 \end{itemize}
-
-More generally, we can make stores map to a domain that forms a galois
-connection with sets of values, so that $\interpdelta$ can give more
-elaborate abstractions of base values (e.g., with the interval
-abstraction, $\{0,[0,5]\} \equiv \{[0,5]\}$). We stick with sets for
-simpler exposition.
 
@@
 \alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
-\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
-\end{align*}
-where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
-\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
-\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
-$\mcntr$, but they can be used to determinize the choice of fresh
+\allockont^\mcntr_\mlab(\msto,\mkont) &= \maddr \mbox{ where } \maddr \notin \msto
+\end{align*}
+where $\sqcup$ on stores is a point-wise lifting of $\sqcup$, \ie,
+$\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
+\msto'(\maddr)$ and $\mstate.\msto$ is notation for projecting out the
+store component of a state. These functions appear to ignore $\mlab$
+and $\mcntr$, but they can be used to determinize the choice of fresh
 addresses.  The resulting relation is non-deterministic in its choice
@@
 interpretation, set the implicit parameters just as above, but drop
-the $\maddr \not\in \msto$ condition. $\interpdelta$ takes some care
+the $\maddr \not\in \msto$ condition. The $\interpdelta$ relation takes some care
 to not make the analysis run forever; a simple instantiation is a flat
@@
 It is important to recognize from this definition that \emph{any}
-allocation strategy is an abstract
+allocation strategy is a sound abstract
 interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
@@
 an interpretation that allocates a single cell into which all bindings
-and continuations are stored.  On the one hand is an abstract
+and continuations are stored.  The former is an abstract
 interpretation that is non-computable and gives only the ground truth
-of a programs behavior; on the other is an abstract interpretation
+of a programs behavior; the latter is an abstract interpretation
 that is easy to compute but gives little information.  Useful program
@@
 choice of address representation and allocation strategy.  Uniform
-\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97} is one such analysis.
+\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97}, presented next, is one such analysis.
 
@@
 
-\section{Reduction semantics to baseline analyzer}
+\section{From machine semantics to baseline analyzer}
 \label{sec:baseline}
@@
 The uniform $k$-CFA allocation strategy would make $\traces$ in figure
-\ref{fig:aam} a computable abstraction of possible executions, but not an
-efficient one. This is not the strategy that AAM, nor we, recommend. Through
-this section, we will explain a succession of approximations
-to reach the baseline analysis.  We'll compare performance at each stage to
-identify the criticality of each choice.
-%
-We ground this journey by first formulating the analysis in terms of a classic
-fixed-point computation.
+\ref{fig:aam} a computable abstraction of possible executions, but one
+that is too inefficient to run, even on small examples.  Through this
+section, we explain a succession of approximations to reach a more
+appropriate baseline analysis.
+%
+We ground this path by first formulating the analysis in terms of a
+classic fixed-point computation.
 
@@
 
-In the next subsection, we'll fix this with a widening and reach polynomial
-(albeit of a high degree) complexity.
+In the next subsection, we fix this with store widening to reach polynomial
+(albeit of high degree) complexity.
 %
@@
 %
-Thus, in one step, complexity drops from intractable exponentiality to a merely
+Thus, complexity drops from intractable exponentiality to a merely
 daunting polynomial.
@@
 \begin{align*}
-(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
+(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
 \mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
@@
 
-The final approximation we make to get to our point of departure is to
+The final approximation we make to get to our baseline is to
 store-allocate all values that appear, so that any non-machine state
@@
 additionally allocate an address for the function value when creating
-the $\kfn{}$ continuation. The continuation has to store this address
-to remember where to get values from, though sometimes your allocation
-strategy may allow you to elide that representation.
-
-The new evaluation rules follow (where $\mcntr' = \tick(\mstate)$).
+the $\kfn{}$ continuation. The continuation has to contain this address
+to remember where to retreive values from in the store.
+
+The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
 % In theory, this aggressive distinction between continuations might buy

### 2013-03-29 paper/paper.tex
Gentler intro to frontier-based semantics.

@@
 \text{Continuations} & \mkont &=& \kmt\\
-&&|& \kfn{\mval,\maddr}\\
-&&|& \kar{\mexp,\menv,\maddr}\\
-&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
+&&|& \kfn{\mval,\mkaddr}\\
+&&|& \kar{\mexp,\menv,\mkaddr}\\
+&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
 \text{Addresses} &\maddr&\in&\Addr \\
@@
 \ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
-\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
+\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
 \\
 &&
-\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
+\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
 \\
 \ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
-\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
+\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
 \\
 &&
-\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
+\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
 \\[2mm]
@@
 \\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
-\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
-\\
-\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
+\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
+\\
+\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
 \ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
-\text{ if }\mkont \in \msto(\maddr_\mkont)
-\\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
+\text{ if }\mkont \in \msto(\mkaddr)
+\\
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
 \ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
-\text{ if }\mkont\in\msto(\maddr_\mkont)
-\\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
+\text{ if }\mkont\in\msto(\mkaddr)
+\\
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
 \ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
-\text{ if }\mkont\in\msto(\maddr_\mkont)
+\text{ if }\mkont\in\msto(\mkaddr)
 \\[2mm]
@@
 \begin{align*}
-\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
-\allockont^\mcntr_\mlab(\msto,\mkont) &= \maddr \mbox{ where } \maddr \notin \msto
-\end{align*}
-where $\sqcup$ on stores is a point-wise lifting of $\sqcup$, \ie,
-$\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
-\msto'(\maddr)$ and $\mstate.\msto$ is notation for projecting out the
-store component of a state. These functions appear to ignore $\mlab$
-and $\mcntr$, but they can be used to determinize the choice of fresh
-addresses.  The resulting relation is non-deterministic in its choice
-of addresses, however it must always choose a fresh address when
-allocating a continuation or variable binding.  If we consider machine
-states equivalent up to consistent renaming and fix an allocation
-scheme, this relation defines a deterministic machine (the relation is
-really a function).
+\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
+\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
+\end{align*}
+These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
+used to determinize the choice of fresh addresses. The $\sqcup$ on
+stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
+\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
+relation is non-deterministic in its choice of addresses, however it
+must always choose a fresh address when allocating a continuation or
+variable binding.  If we consider machine states equivalent up to
+consistent renaming and fix an allocation scheme, this relation
+defines a deterministic machine (the relation is really a function).
 
@@
 \tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
-\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\mval,\msto}) &= \mcntr \\
+\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
 \tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
@@
               \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
-\wn &: \widehat{\State} \times \Store \to \State \\
+\wn &: \Context \times \Store \to \State \\
 \wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
@@
 
-\begin{align*}
-\co[^\mcntr]{\kar{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep
-\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_\mkont}} \\
+HERE
+
+\begin{align*}
+\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\mkaddr}} \\
 \text{ where }
@@
   \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
-\\
-\co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
+\end{align*}
+Now instead of storing the evaluated function in the continuation
+frame itself, we indirect it through the store for further control on
+complexity and precision.
+
+\begin{align*}
+\co[^\mcntr]{\kfn{\maddr_f,\mkaddr},\mval,\msto} & \machstep
 \ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
 \\
-\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f)
-\end{align*}
-We associate the value in hand with the expression that produced
-it. Also, we fan out to all functions that are associated with the
-original function position, because applied functions are in
+\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr_f)
+\end{align*}
+Associated with this indirection, we now apply all functions stored in
+the address. This nondeterminism is demanded in order to continue with
+evaluation. More generally, this is because applied functions are in
 \emph{strict position} (more about that later).
@@
 can travel back in time in straight-line code). We instead use a
-frontier-based semantics that corresponds to the classic ``worklist''
+frontier-based semantics that corresponds to the classic worklist
 algorithms for analysis. The difference is that the store is not
@@
 
-\begin{align*}
-(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
-\mbox{where } I &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
-              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
-              F' &= \{c \mid (c,\_) \in I, (c,\msto') \notin S\} \\
-              S' &= \setof{(c,\msto') \mid c \in F'} \\
+The state space changes from one store plus a set of contexts to a set
+of seen abstract states (context plus store), $S$, a set of contexts to
+step (the frontier), $F$, and the store to step those contexts with, $\msto$.
+
+\begin{equation*}
+(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
+\end{equation*}
+
+We constantly see more states, so $S$ is always growing. The frontier,
+which is what remains to be done, changes. Let's start with the result
+of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
+
+\begin{equation*}
+I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
+\end{equation*}
+
+The next store is the least upper bound of all the stores in $I$ (in
+order to have only one, and keep the semantics sound):
+
+\begin{equation*}
+\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
+\end{equation*}
+
+The next frontier is exactly the states that we found from stepping the last frontier, but have not seen before. They must be \emph{states}, so we pair the contexts with the next store:
+
+\begin{equation*}
+F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
+\end{equation*}
+
+Finally, we add what we know we had not yet seen to the seen set:
+
+\begin{equation*}
+  S' = \setof{(c,\msto') \mid c \in F'}
+\end{equation*}
+
+To inject a program $e$ into this machine, we start off knowing we
+have seen the first state, and that we need to process the first
+state:
+\begin{align*}
 \inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
@@
 \begin{align*}
-\Sigma &\in \Store^* \\
-S &\subseteq {\mathbb N} \times \widehat{\State} \\
-F &\subseteq \widehat{\State} \\
+\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
+\end{align*}
+\begin{align*}
 (S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
@@
 \begin{align*}
-\dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \dmachstep_\msto
+\dmachstep &\subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) \\
+\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
 \ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
@@
 \mbox{ where }
- I &= \setof{(\mastate',\msdiff) \mid \mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff} } \\
+ I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
  \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
@@
        \end{array}\right. \\
- F' &= \setof{\mastate \mid (\mastate,\_) \in I, (\mastate,t') \notin S} \\
- S' &= \setof{(\mastate,t') \mid c \in F'} \\
+ F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
+ S' &= \setof{(c,t') \mid c \in F'} \\
 \appendall(\varnothing) &= \epsilon \\
@@
 \begin{lemma}
-$\mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff}$ iff
-$\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
+$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff
+$\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
 \end{lemma}
@@
 %
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto}
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto}
 &\lmachstep\;
-\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}} \\
 \text{ where }
@@
 %
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\mval,\msto}
 &\lmachstep\;
 \ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
-\text{ if } \mkont &\in \msto(\maddr_\mkont),
+\text{ if } \mkont &\in \msto(\mkaddr),
             \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
@@
 \compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
-\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
+\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\mkaddr})
 \\
@@
 \begin{array}{lrl}
-\text{ where } & \maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-               & \msdiff' = \cons{\ttuple{\maddr_\mkont}{\setof{\mkont}}}{\msdiff}
+\text{ where } & \mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+               & \msdiff' = \cons{\ttuple{\mkaddr}{\setof{\mkont}}}{\msdiff}
 \end{array}
@@
 \\
-&\text{ where }\maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&\phantom{\text{ where }} \msdiff' = \cons{\ttuple{\maddr_\mkont}{\setof{\mkont}}}{\msdiff}
+&\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&\phantom{\text{ where }} \msdiff' = \cons{\ttuple{\mkaddr}{\setof{\mkont}}}{\msdiff}
 \end{align*}
@@
 \\
-\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_\mkont},\mval} & \cmachstep_\msto
-\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}) \\
+\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\mkaddr},\mval} & \cmachstep_\msto
+\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}) \\
 \text{ where } \maddr_f &= \alloc(\mstate) \\
@@
 \\
-\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont},\mval} & \cmachstep_\msto
+\co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
 \ap[^\mcntr_\mlab]{\mval,\mval\mkont},\epsilon \\
-\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\maddr_\mkont)
+\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
 \\

### 2013-03-29 paper/paper.tex
Tightening wording in lazy nondet. section.

@@
 \begin{figure}[t]
+\small
 \begin{center}
@@
 \begin{figure}
+\small
 \[
@@
 &&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
-\text{Addresses} &\maddr&\in&\Addr \\
+\text{Addresses} &\maddr,\mkaddr&\in&\Addr \\
 \text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
@@
 \begin{figure}
+\small
 \begin{gather*}
@@
 
+{\small
 \begin{align*}
@@
   \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
-\end{align*}
+\end{align*}}
 The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
@@
 Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
-\begin{equation*}
- \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
-\end{equation*}
+% \begin{equation*}
+ $\eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}$
+%\end{equation*}
 where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.
@@
 %
-Then, before each transition, it will install this joined store into current
+Then, before each transition, it installs this joined store into current
 state.
@@
 %
+{\small
 \begin{align*}
@@
 \wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
-\end{align*}
+\end{align*}}
 %
@@
 % additional precision, but in practice, it does not.
-
-\newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
-%\newcommand{\ext}[3]{ext(#1,#2,#3)}
-
-HERE
-
 \begin{align*}
 \co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
-\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\mkaddr}} \\
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
 \text{ where }
-  \maddr_f &= \alloc(\mstate) \\
-  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
+  \maddr &= \alloc(\mstate) \\
+  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
 \end{align*}
@@
 frame itself, we indirect it through the store for further control on
-complexity and precision.
-
-\begin{align*}
-\co[^\mcntr]{\kfn{\maddr_f,\mkaddr},\mval,\msto} & \machstep
+complexity and precision:
+\begin{align*}
+\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
 \ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
 \\
-\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr_f)
+\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
 \end{align*}
@@
 
-The state space changes from one store plus a set of contexts to a set
+The state space changes from a store and set of contexts to a set
 of seen abstract states (context plus store), $S$, a set of contexts to
-step (the frontier), $F$, and the store to step those contexts with, $\msto$.
-
+step (the frontier), $F$, and a store to step those contexts with, $\msto$:
 \begin{equation*}
@@
 of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
-
+%
 \begin{equation*}
@@
 \end{equation*}
-
+%
 The next store is the least upper bound of all the stores in $I$ (in
 order to have only one, and keep the semantics sound):
-
+%
 \begin{equation*}
@@
 \end{equation*}
-
-The next frontier is exactly the states that we found from stepping the last frontier, but have not seen before. They must be \emph{states}, so we pair the contexts with the next store:
-
+%
+The next frontier is exactly the states that we found from stepping
+the last frontier, but have not seen before. They must be
+\emph{states}, so we pair the contexts with the next store:
+%
 \begin{equation*}
@@
 \end{equation*}
-
+%
 Finally, we add what we know we had not yet seen to the seen set:
-
 \begin{equation*}
@@
 \end{equation*}
-
+%
 To inject a program $e$ into this machine, we start off knowing we
@@
 \text{ where } c_0 &= \ev{e,\bot,\kmt}
-\end{align*}
+\end{align*}}
 
@@
 We lift $\dmachstep$ to accommodate for the asymmetry
-in the input and output, and change the frontier-based semantics in the following way:
-
+in the input and output, and change the frontier-based semantics in
+the following way:
+{\small
 \begin{align*}
@@
 \appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
-\end{align*}
+\end{align*}}
 
@@
 \renewcommand{\ext}{\mathit{ext}}
-
+{\small
 \begin{align*}
@@
             \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
-\end{align*}
+\end{align*}}
 Since {\tt if} guards are in strict position, we must force the value
@@
 The administrative nature of laziness means that we could remove the
-loss in precision by duplicating the reduction relation to specialize
-variable lookup. This works since in the semantics of ISWIM with
-store-allocated results consumes the value component of states in one
-step. This is not the case for semantics that replicate the value
-component across reductions, say for popping off exception handler
-frames. Further convolution is needed to remove the administrative
-nature of laziness in these semantics. Due to the increase of
-conceptual and implementation complexity for negligible benefit, we
-decided against this approach.
+loss in precision by storing the result of the lookup in a value
+representing a delayed nondeterministic choice. This has problems with
+the next optimization, leading to unsoundness. Regaining soundness
+would involve artificial complications to the semantics.
 
@@
 warrant the significant increase in time and memory needed for this
-eager non-determinism. Indeed, were the variable reference a step
-later and another binding not made in that step, the results of the
-two approaches are the same. Without a widened store, lazy
+eager non-determinism. Without a widened store, lazy
 non-determinism is not a complete abstraction with either approach,
@@
 \begin{figure}
+\small
 \begin{align*}
@@
 \mvalstack \sqcup_t \mval{s} &= \mvalstack,\sfalse \text{ otherwise}
-\end{align*}
+\end{align*}}
 
@@
                              \end{array}\right.
-\end{align*}
+\end{align*}}
 
@@
 \item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}
-\item{$\changep$: whether there has been a store change stepping all states in $F$.}
+\item{$\changep$: whether a store change stepped all states in $F$.}
 \end{enumerate}
@@
 
-\item {\bf mbrotZ}: generates Mandelbrot set fractal using complex
+\item {\bf mbrotZ}: generates Mandelbrot fractal using complex
   numbers.
@@
 %\input{paper.bbl}
+
 \bibliography{local,bibliography}

### 2013-03-29 paper/paper.tex
Changed abstract states to contexts. Tightened section on preallocation.

@@
  \item store deltas; and
- \item imperative, preallocated data structures.
+ \item imperative, pre-allocated data structures.
 \end{enumerate}
@@
                     \text{ where } \\
-\inject(\mastate,\msdiff) &= \wn(\mastate,\replay(\msdiff,\bot)) \\
-\wn(\mastate,\msto) \machstep \wn(\mastate',\msto') &\iff \mastate \cmachstep_\msto \mastate',\msdiff \\
+\inject(c,\msdiff) &= \wn(c,\replay(\msdiff,\bot)) \\
+\wn(c,\msto) \machstep \wn(c',\msto') &\iff c \cmachstep_\msto c',\msdiff \\
 \msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
@@
 \begin{lemma}[Compile/commit]
-Let $\mastate,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
-Let $\wn(\mastate',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
-$\wn(\mastate,\replay(\msdiff',\msto)) = \wn(\mastate',\replay(\msdiff,\msto'))$.
+Let $c,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
+Let $\wn(c',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
+$\wn(c,\replay(\msdiff',\msto)) = \wn(c',\replay(\msdiff,\msto'))$.
 \end{lemma}
@@
 
-\subsection{Imperative, preallocated data structures}
+\subsection{Imperative, pre-allocated data structures}
 
@@
 \text{ where }
-I &= \setof{(\mastate',\msdiff) \mid
-       \mastate \in F,
-       \mastate \cmachstep_{\msto^*} \mastate',\msdiff} \\
+I &= \setof{(c',\msdiff) \mid
+       c \in F,
+       c \cmachstep_{\msto^*} c',\msdiff} \\
 \msto^* &= \lambda \maddr.\lookup(\msto(\maddr),t) \\
@@
              \end{array}\right. \\
-F' &= \setof{\mastate \mid (\mastate,\_) \in I, \changep \vee S(\mastate) \neq \cons{t}{\_}} \\
-S' &= \lambda \mastate. \left\{\begin{array}{ll}
-                               \cons{t'}{S(\mastate)} & \text{ if } \mastate \in F' \\
-                               S(\mastate) & \text{ otherwise}
+F' &= \setof{c \mid (c,\_) \in I, \changep \vee S(c) \neq \cons{t}{\_}} \\
+S' &= \lambda c. \left\{\begin{array}{ll}
+                               \cons{t'}{S(c)} & \text{ if } c \in F' \\
+                               S(c) & \text{ otherwise}
                              \end{array}\right.
@@
 
-Internally, the algorithm at this stage uses hash tables to model the store.
-%
-This is because stores used to be distributed to all states, which
-required a compact, dynamic representation.
+Internally, the algorithm at this stage uses hash tables to model the store to allow arbitrary address representations.
 %
@@
 we know its maximum size.
-
-In a monovariant analysis, the domain of the store is
+%
+In a monovariant allocation strategy, the domain of the store is
 exactly the set of expressions in the program.

### 2013-03-29 paper/paper.tex
Space saving.

--- 2013-03-29 paper/paper.tex
@@
 The administrative nature of laziness means that we could remove the
-loss in precision by storing the result of the lookup in a value
-representing a delayed nondeterministic choice. This has problems with
-the next optimization, leading to unsoundness. Regaining soundness
-would involve artificial complications to the semantics.
+loss in precision by duplicating the reduction relation to specialize
+variable lookup. This works since in the semantics of ISWIM with
+store-allocated results consumes the value component of states in one
+step. This is not the case for semantics that replicate the value
+component across reductions, say for popping off exception handler
+frames. Further convolution is needed to remove the administrative
+nature of laziness in these semantics. Due to the increase of
+conceptual and implementation complexity for negligible benefit, we
+decided against this approach.
 
@@
 warrant the significant increase in time and memory needed for this
-eager non-determinism. Without a widened store, lazy
+eager non-determinism. Indeed, were the variable reference a step
+later and another binding not made in that step, the results of the
+two approaches are the same. Without a widened store, lazy
 non-determinism is not a complete abstraction with either approach,

### 2013-03-29 paper/paper.tex
Merge.

--- 2013-03-29 paper/paper.tex
@@
 %
-In effect, the new store is computed as the least upper bound of all
-stores after stepping. For clarity, we will exhibit non-widened
-semantics unless a technique specifically requires it. The final step
-to the baseline takes the complexity from quartic to cubic in the
-monovariant case.
+The new store is computed as the least upper bound of all
+stores after stepping in order to have only one, and keep the
+semantics sound.
 
@@
 \end{equation*}
-%
-The next store is the least upper bound of all the stores in $I$ (in
-order to have only one, and keep the semantics sound):
-%
+
+The next store is the least upper bound of all the stores in $I$:
+
 \begin{equation*}
@@
 plots the factors of improvement over the baseline for each
-optimization step. The dip we see in transition rate even though time
-taken decreases is to be expected - fewer ``easy'' states are added by
-abstract compilation. It increases again with the introduced
-algorithmic improvements. Accumulating store changes in addition to
-maintaining the store accounts for the higher memory usage when using
-the store delta technique without further improvements.
+optimization step.
+
+% The dip we see in transition rate even though time
+% taken decreases is to be expected - fewer ``easy'' states are added by
+% abstract compilation. It increases again with the introduced
+% algorithmic improvements. Accumulating store changes in addition to
+% maintaining the store accounts for the higher memory usage when using
+% the store delta technique without further improvements.
 

### 2013-03-29 paper/paper.tex
We actually do better than 25 times slower now. Hooray.

@@
 pre-allocate most data structures and split the abstract store out
-into parts that are more quickly accessed and updated. The first two
-of our optimizations can still be applied to an analysis that does
-abstract garbage collection~\cite{dvanhorn:Might:2006:GammaCFA},
-whereas the polymorphic splitting implementation is tied strongly to a
+into parts that are more quickly accessed and updated. Our semantic
+optimizations can still be applied to an analysis that does abstract
+garbage collection~\cite{dvanhorn:Might:2006:GammaCFA}, whereas the
+polymorphic splitting implementation is tied strongly to a
 single-threaded store.
@@
 implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
-the \Church{} benchmark about 25 times faster than the fastest
+the \Church{} benchmark about 5.5 times faster than the fastest
 implementation considered here.  These analyses compute very different

### 2013-03-29 paper/paper.tex
Moved figure.
Some punctuation fixes.
Punctuation.
Some fixes in Section 5

@@
 of the analysis time for this example.
+
+The AAM approach, in essence, does the following: it takes a
+machine-based view of computation and turns it into a \emph{finitary
+  approximation} by bounding the size of the store.  With a limited
+address space, the store must map addresses to \emph{sets} of values.
+Store updates are interpreted as joins, and store dereferences are
+interpreted by non-deterministic choice of an element from a set.  The
+result of analyzing a program is a finite directed graph where nodes
+in the graph are (abstract) machine states and edges denote machine
+transitions between states.
+
+The techniques we propose for optimizing analysis fall into the
+following categories:
+\begin{enumerate}
+\item generate fewer states by avoiding the eager exploration of
+  non-deterministic choices that will later collapse into a single
+  join point.  We accomplish this by applying lazy evaluation
+  techniques so that non-determinism is evaluated \emph{by need}.
+
+\item generate fewer states by avoiding unnecessary, intermediate
+  states of a computation.  We accomplish this by applying compilation
+  techniques from functional languages to avoid interpretive overhead
+  in the machine transition system.
+
+\item generate states faster.  We accomplish this by better algorithm
+  design in the fixed-point computation we use to generate state graphs.
+\end{enumerate}
+Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
+small example due to Earl, et
+al.~\cite{dvanhorn:Earl2012Introspective}.
+By generating significantly fewer states at a significantly faster
+rate, we are able to achieve large performance improvements in terms
+of both time and space.
+
+Section~\ref{sec:eval} describes the evaluation of each optimization
+technique applied to an implementation supporting a more realistic set
+of features, including mutation, first-class control, compound data, a
+full numeric tower and many more forms of primitive data and
+operations.
+%
+We evaluate this implementation against a set of benchmark programs
+drawn from the literature.
+%
+For all benchmarks, the optimized analyzer outperforms the baseline
+by at least a factor of
+% 475
+two to
+% 4,382
+three orders of magnitude.
+
+Section~\ref{sec:related} relates this work to the literature and
+section~\ref{sec:conclusion} concludes.
 
@@
 \end{figure}
-
-The AAM approach, in essence, does the following: it takes a
-machine-based view of computation and turns it into a \emph{finitary
-  approximation} by bounding the size of the store.  With a limited
-address space, the store must map addresses to \emph{sets} of values.
-Store updates are interpreted as joins, and store dereferences are
-interpreted by non-deterministic choice of an element from a set.  The
-result of analyzing a program is a finite directed graph where nodes
-in the graph are (abstract) machine states and edges denote machine
-transitions between states.
-
-The techniques we propose for optimizing analysis fall into the
-following categories:
-\begin{enumerate}
-\item generate fewer states by avoiding the eager exploration of
-  non-deterministic choices that will later collapse into a single
-  join point.  We accomplish this by applying lazy evaluation
-  techniques so that non-determinism is evaluated \emph{by need}.
-
-\item generate fewer states by avoiding unnecessary, intermediate
-  states of a computation.  We accomplish this by applying compilation
-  techniques from functional languages to avoid interpretive overhead
-  in the machine transition system.
-
-\item generate states faster.  We accomplish this by better algorithm
-  design in the fixed-point computation we use to generate state graphs.
-\end{enumerate}
-Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
-small example due to Earl, et
-al.~\cite{dvanhorn:Earl2012Introspective}.
-By generating significantly fewer states at a significantly faster
-rate, we are able to achieve large performance improvements in terms
-of both time and space.
-
-Section~\ref{sec:eval} describes the evaluation of each optimization
-technique applied to an implementation supporting a more realistic set
-of features, including mutation, first-class control, compound data, a
-full numeric tower and many more forms of primitive data and
-operations.
-%
-We evaluate this implementation against a set of benchmark programs
-drawn from the literature.
-%
-For all benchmarks, the optimized analyzer outperforms the baseline
-by at least a factor of
-% 475
-two to
-% 4,382
-three orders of magnitude.
-
-Section~\ref{sec:related} relates this work to the literature and
-section~\ref{sec:conclusion} concludes.
 
@@
   \setcounter{enumi}{3}
- \item frontier-based semantics
- \item lazy non-determinism and
- \item abstract compilation;
+ \item frontier-based semantics;
+ \item lazy non-determinism; and
+ \item abstract compilation.
 \end{enumerate}
@@
 \begin{theorem}
-Timestamps are a complete abstraction
+Timestamps are a complete abstraction.
 \end{theorem}
-The proof follows from the galois connection that, in one direction,
+The proof follows from the Galois connection that, in one direction,
 sorts all the stores in $S$ to form $\Sigma$, and translates stores in
@@
 A taste of the changes to the reduction relation is as follows:
-
-\begin{align*}
-\dmachstep &\subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) \\
+\begin{equation*}
+\dmachstep \subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) 
+\end{equation*}
+\begin{align*}
 \ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
@@
 \end{theorem}
-Here $\sqsubseteq$ is straightforward - the LHS store must be
-contained in the RHS store, and if values occur in the states, forcing
-the LHS value must be a subset of forcing the corresponding RHS value.
+Here $\sqsubseteq$ is straightforward - the left-hand side store must be
+contained in the right-hand-side store, and if values occur in the states, forcing
+the left-hand-side value must be a subset of forcing the corresponding right-hand-side value.
 The proof is by cases on $\mstate \machstep \mstate'$.}

### 2013-03-29 paper/paper.tex
Drop mem from front page.

@@
 \begin{center}
-\includegraphics[width=3.2in]{church-relative-space}
+%\includegraphics[width=3.2in]{church-relative-space}
 \includegraphics[width=3.2in]{church-relative-speed}
@@
 \caption{Factor improvements over the baseline analyzer for the
-  \Church{} benchmark in terms of peak memory usage, the rate of state
-  transitions, and total analysis time. (Bigger is better.) Each point
+  \Church{} benchmark in terms of the rate of state
+  transitions and total analysis time. (Bigger is better.) Each point
   is marked with the section that introduces the optimization.}
@@
 
-We evaluated the precision of these techniques with a singleton
-variable analysis to find opportunities to inline constants and closed
-functions. We found no change in the results across all
-implementations, including Shivers' timestamp approximation.
+To determine the impact of these techniques on precision, we evaluated
+the a singleton variable analysis to find opportunities to inline
+constants and closed functions.  We found no change in the results
+across all implementations, including Shivers' timestamp
+approximation---from an empirical point of view, these techniques are
+precision preserving despite the theoterical loss of precision.
 

### 2013-03-29 paper/paper.tex
Better wording of first theorem.

@@
 \begin{theorem}
-Timestamps are a complete abstraction.
+$\widehat{\machstep}^T$ is a complete abstraction of $\widehat{\machstep}$.
 \end{theorem}
@@
 sorts all the stores in $S$ to form $\Sigma$, and translates stores in
-$S$ to their distance from the end of $\Sigma$. In the other
-direction, timestamps are replaced by the stores they point to.
+$S$ to their distance from the end of $\Sigma$ (their timestamp). In the other
+direction, timestamps in $S$ are replaced by the stores they point to in $\Sigma$.
 

### 2013-03-29 paper/paper.tex
Summary of results.

@@
 precision preserving despite the theoterical loss of precision.
+
+Our step-wise optimizations strictly produce better analysis times
+with no observed loss of precision.  The final result is a
+systematically derived and verified implementation that operates
+within a small factor performance loss compared to a hand-optimized,
+unverified implementation.  Moreover, much of the performance gains
+are achieved with purely functional methods, which suggest potential
+further gains from parellelization.  Peak memory usage is often
+considerably improved by the end of the optimization steps, but the
+effect of mutation has unexpected consequence on the analyzers memory
+consumption.
 

### 2013-03-29 paper/paper.tex
As submitted.

@@
 
-\authorinfo{J. Ian Johnson \and Nicholas Labich \and Matthew Might \and David Van Horn}
-           {}
-           {}
-%% \authorinfo{J. Ian Johnson}
-%%            {Northeastern University}
-%%            {ianj@ccs.neu.edu}
-%% \authorinfo{Nicholas Labich}
-%%            {Northeastern University}
-%%            {labichn@ccs.neu.edu}
-%% \authorinfo{Matthew Might}
-%%            {University of Utah}
-%%            {might@cs.utah.edu}
-%% \authorinfo{David Van Horn}
-%%            {Northeastern University}
-%%            {dvanhorn@ccs.neu.edu}
+\authorinfo{J. Ian Johnson}
+           {Northeastern University}
+           {ianj@ccs.neu.edu}
+\authorinfo{Nicholas Labich}
+           {Northeastern University}
+           {labichn@ccs.neu.edu}
+\authorinfo{Matthew Might}
+           {University of Utah}
+           {might@cs.utah.edu}
+\authorinfo{David Van Horn}
+           {Northeastern University}
+           {dvanhorn@ccs.neu.edu}
 \maketitle

### 2013-04-02 paper/paper.tex
Changed over to alloc/tick, discussed problems with widening with compilation.
Conflicts:

	paper/paper.tex

--- 2013-03-29 paper/paper.tex
@@
 
-%\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
-\title{Correctly Optimizing Abstract Abstract Machines}
-
-\authorinfo{J. Ian Johnson \and Nicholas Labich \and Matthew Might \and David Van Horn}
-           {}
-           {}
-%% \authorinfo{J. Ian Johnson}
-%%            {Northeastern University}
-%%            {ianj@ccs.neu.edu}
-%% \authorinfo{Nicholas Labich}
-%%            {Northeastern University}
-%%            {labichn@ccs.neu.edu}
-%% \authorinfo{Matthew Might}
-%%            {University of Utah}
-%%            {might@cs.utah.edu}
-%% \authorinfo{David Van Horn}
-%%            {Northeastern University}
-%%            {dvanhorn@ccs.neu.edu}
+\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
+
+\authorinfo{J. Ian Johnson}
+           {Northeastern University}
+           {ianj@ccs.neu.edu}
+\authorinfo{Matthew Might}
+           {University of Utah}
+           {might@cs.utah.edu}
+\authorinfo{David Van Horn}
+           {Northeastern University}
+           {dvanhorn@ccs.neu.edu}
+
 \maketitle
@@
 \begin{abstract}
-The technique of \emph{abstracting abstract machines} (AAM) provides a
-systematic approach for deriving computable approximations of
-evaluators that are easily proved sound.
-%
-This article contributes a complementary step-by-step process for
-subsequently going from a \naive{} analyzer derived under the AAM
-approach, to an efficient and correct implementation.  The end result
-of the process is a two to three order-of-magnitude improvement over
-the systematically derived analyzer, making it competitive with
+Abstracting abstract machines has been proposed as a
+lightweight approach to designing sound and computable program
+analyses.  The approach derives abstract interpreters from
+existing machine semantics and has been applied to a variety of
+languages with features widely considered difficult to analyze.
+Although sound analyzers are straightforward to build under this approach,
+they are also prohibitively inefficient.
+
+This article contributes a step-by-step process for going from a
+\naive{} analyzer derived under the abstracting abstract machine
+approach to an efficient program analyzer.  The end result of the
+process is a two to three order-of-magnitude improvement over the
+systematically derived analyzer, making it competitive with
 hand-optimized implementations that compute fundamentally less precise
@@
 
-Program analysis provides sound predictive models of program behavior, but
-in order for such models to be effective, they must be efficiently
-computable and correct.  Past approaches to designing program analysis
-have often featured abstractions that are far removed from the
-original language semantics, requiring ingenuity in their construction
-and effort in their verification.
-%
 The \emph{abstracting abstract machines} (AAM)
 approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
-to deriving program analyses provides an alternative: a systematic way
-of transforming a programming language semantics in the form of an
-abstract machine into a family of abstract interpreters.  It thus reduces
-the burden of constructing and verifying the soundness of an abstract
-interpreter.
-%
+to deriving program analyses provides a systematic way of transforming a
+programming language semantics in the form of an abstract machine into a family
+of abstract interpreters.
+%
+The approach parameterizes these families with policies for regulating analytic
+precision.
+%
+While flexible and robust, the AAM approach unfortunately yields analyzers with
+poor performance relative to hand-optimized analyzers.
+%
+Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an
+equally systematic way.
 
@@
 
-Unfortunately, the AAM approach yields analyzers with poor performance
-relative to hand-optimized analyzers.
-%
-Our work takes aim squarely at this ``efficiency gap,'' and narrows it
-in an equally systematic way through a number of simple steps, many of
-which are inspired by run-time implementation techniques such as
-laziness and compilation to avoid interpretative overhead.  Each of
-these steps is proven correct, so the end result is an implementation
-that is trustworthy and efficient.
+However, there is a considerable weakness with the approach: an
+analyzer designed and implemented by following the AAM recipe is
+prohibitively inefficient without both further approximation and
+further implementation effort.
 
 In this article, we develop a systematic approach to deriving a
-practical implementation of an abstract-machine-based analyzer using
-mostly semantic means rather than tricky engineering. Our goal is to
-empower programming language implementers and researchers to explore
-and convincingly exhibit their ideas with a low barrier to entry. The
-optimizations we describe are widely applicable and apparently
-effective to scale far beyond the size of programs typically considered
-in the recent literature on flow analysis for functional languages.
+feasible implementation of an abstract-machine-based analyzer.
 
@@
 \begin{figure}[t]
-\small
 \begin{center}
-%\includegraphics[width=3.2in]{church-relative-space}
+\includegraphics[width=3.2in]{church-relative-space}
 \includegraphics[width=3.2in]{church-relative-speed}
@@
 \caption{Factor improvements over the baseline analyzer for the
-  \Church{} benchmark in terms of the rate of state
-  transitions and total analysis time. (Bigger is better.) Each point
+  \Church{} benchmark in terms of peak memory usage, the rate of state
+  transitions, and total analysis time. (Bigger is better.) Each point
   is marked with the section that introduces the optimization.}
@@
 
-The AAM approach, in essence, does the following: it takes a
-machine-based view of computation and turns it into a \emph{finitary
-  approximation} by bounding the size of the store.  With a limited
-address space, the store must map addresses to \emph{sets} of values.
-Store updates are interpreted as joins, and store dereferences are
-interpreted by non-deterministic choice of an element from a set.  The
-result of analyzing a program is a finite directed graph where nodes
-in the graph are (abstract) machine states and edges denote machine
-transitions between states.
-
-The techniques we propose for optimizing analysis fall into the
-following categories:
-\begin{enumerate}
-\item generate fewer states by avoiding the eager exploration of
-  non-deterministic choices that will later collapse into a single
-  join point.  We accomplish this by applying lazy evaluation
-  techniques so that non-determinism is evaluated \emph{by need}.
-
-\item generate fewer states by avoiding unnecessary, intermediate
-  states of a computation.  We accomplish this by applying compilation
-  techniques from functional languages to avoid interpretive overhead
-  in the machine transition system.
-
-\item generate states faster.  We accomplish this by better algorithm
-  design in the fixed-point computation we use to generate state graphs.
-\end{enumerate}
-Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
-small example due to Earl, et
-al.~\cite{dvanhorn:Earl2012Introspective}.
-By generating significantly fewer states at a significantly faster
-rate, we are able to achieve large performance improvements in terms
-of both time and space.
-
-Section~\ref{sec:eval} describes the evaluation of each optimization
-technique applied to an implementation supporting a more realistic set
-of features, including mutation, first-class control, compound data, a
-full numeric tower and many more forms of primitive data and
-operations.
-%
-We evaluate this implementation against a set of benchmark programs
-drawn from the literature.
-%
-For all benchmarks, the optimized analyzer outperforms the baseline
-by at least a factor of
-% 475
-two to
-% 4,382
-three orders of magnitude.
-
-Section~\ref{sec:related} relates this work to the literature and
-section~\ref{sec:conclusion} concludes.
-
 \begin{figure}[t]
-\small
 \begin{center}
@@
 
+The AAM approach, in essence, does the following: it takes a
+machine-based view of computation and turns it into a \emph{finitary
+  approximation} by bounding the size of the store.  With a limited
+address space, the store must map addresses to \emph{sets} of values.
+Store updates are interpreted as joins, and store dereferences are
+interpreted by non-deterministic choice of an element from a set.  The
+result of analyzing a program is a finite directed graph where nodes
+in the graph are machine states and edges denote machine
+transitions between states.
+
+The techniques we propose for optimizing analysis fall into the
+following categories:
+\begin{enumerate}
+\item generate fewer states by avoiding the eager exploration of
+  non-deterministic choices that will later collapse into a single
+  join point.  We accomplish this by applying lazy evaluation
+  techniques so that non-determinism is evaluated \emph{by need}.
+
+\item generate fewer states by avoiding unnecessary, intermediate
+  states of a computation.  We accomplish this by applying compilation
+  techniques from functional languages to avoid interpretive overhead
+  in the machine transition system.
+
+\item generate states faster.  We accomplish this by better algorithm
+  design in the fixed-point computation we use to generate state graphs.
+\end{enumerate}
+Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
+small example due to Earl, et
+al.~\cite{dvanhorn:Earl2012Introspective}.
+By generating significantly fewer states at a significantly faster
+rate, we are able to achieve large performance improvements in terms
+of both time and space.
+
+Section~\ref{sec:eval} describes the evaluation of each optimization
+technique applied to an implementation supporting a more realistic set
+of features, including mutation, first-class control, compound data, a
+full numeric tower and many more forms of primitive data and
+operations.
+%
+We evaluate this implementation against a set of benchmark programs
+drawn from the literature.
+%
+For all benchmarks, the optimized analyzer outperforms the baseline
+by at least a factor of
+% 475
+two to
+% 4,382
+three orders of magnitude.
+
+Section~\ref{sec:related} relates this work to the literature and
+section~\ref{sec:conclusion} concludes.
+
 %\newpage
-\section{Abstract interpretation of ISWIM}
+\section{An analyzer for ISWIM}
 \label{sec:aam}
@@
 \begin{figure}
-\small
 \[
@@
 \text{Continuations} & \mkont &=& \kmt\\
-&&|& \kfn{\mval,\mkaddr}\\
-&&|& \kar{\mexp,\menv,\mkaddr}\\
-&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
-\text{Addresses} &\maddr,\mkaddr&\in&\Addr \\
+&&|& \kfn{\mval,\maddr}\\
+&&|& \kar{\mexp,\menv,\maddr}\\
+&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
+\text{Addresses} &\maddr&\in&\Addr \\
 \text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
@@
 The semantics are defined in terms of a machine model.  The machine
-components are defined in figure~\ref{fig:domains};
+components are defined in figure~\ref{fig:domains} and
 figure~\ref{fig:aam} defines the transition relation.  The evaluation
-of a program is defined as its set of traces that arise from iterating the
-machine transition relation.  The
+of a program is defined as its traces given by the machine transition relation.  The
 machine is a very slight variation on a standard abstract machine for
@@
 \begin{itemize}
-\item the store maps addresses to \emph{sets} of values,\footnote{
-More generally, we can have stores map to any domain that forms
- a Galois connection with values, enabling $\interpdelta$ to
-produce elaborate abstractions of base values (e.g., interval or octagon
-abstractions). We use sets of values for a simpler exposition.
-} not
+\item the store maps addresses to \emph{sets} of values, not
   single values,
-
 \item continuations are heap-allocated, not stack-allocated,
-\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
+\item there are ``timestamps'' (written $\mcntr$) and syntax
   labels ($\mlab$) threaded through the computation, and
 \item the machine is implicitly parameterized by the functions
-  $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces
+  $\mathit{alloc}$, $\mathit{tick}$,
+  $\mathit{push}$, $\interpdelta$, and spaces
   $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
@@
 
+More generally, we can make stores map to a domain that forms a galois
+connection with sets of values, so that $\interpdelta$ can give more
+elaborate abstractions of base values (e.g., with the interval
+abstraction, $\{0,[0,5]\} \equiv \{[0,5]\}$). We stick with sets for
+simpler exposition.
+
 \begin{figure}
-\small
 \begin{gather*}
@@
 \begin{array}{@{}r@{\ }c@{\ }l@{}}
-\mstate &\machstep&\mstate' \text{ defined to be the following} \\
-&&\text{let } \mcntr' =\tick(\mstate) \\
+&&\mcntr' =\tick(\mcntr) \\
+&&\maddr  =\alloc(\mstate) \\
 %% EVAL
@@
 \ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
-\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
+\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
 \\
 &&
-\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
+\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
 \\
 \ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
-\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
+\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
 \\
 &&
-\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
+\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
 \\[2mm]
@@
 \\
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
-\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
-\\
-\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
+\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
+\\
+\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
 \ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
-\text{ if }\mkont \in \msto(\mkaddr)
-\\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
+\text{ if }\mkont \in \msto(\maddr_\mkont)
+\\
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
 \ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
-\text{ if }\mkont\in\msto(\mkaddr)
-\\
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
+\text{ if }\mkont\in\msto(\maddr_\mkont)
+\\
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
 \ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
-\text{ if }\mkont\in\msto(\mkaddr)
+\text{ if }\mkont\in\msto(\maddr_\mkont)
 \\[2mm]
@@
 \\
-&&\text{ where } \maddr  =\alloc(\mstate) \\
-&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
+&&\text{ where } \menv' = \menv[\mvar\mapsto\maddr] \\
 &&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
@@
 \begin{align*}
-\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
-\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
-\end{align*}
-These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
-used to determinize the choice of fresh addresses. The $\sqcup$ on
-stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
-\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
-relation is non-deterministic in its choice of addresses, however it
-must always choose a fresh address when allocating a continuation or
-variable binding.  If we consider machine states equivalent up to
-consistent renaming and fix an allocation scheme, this relation
-defines a deterministic machine (the relation is really a function).
+\push_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
+\mbox{ where }\maddr \notin\msto
+\\
+\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto
+\end{align*}
+where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
+\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
+\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
+$\mcntr$, but they can be used to determinize the choice of fresh
+addresses.  The resulting relation is non-deterministic in its choice
+of addresses, however it must always choose a fresh address when
+allocating a continuation or variable binding.  If we consider machine
+states equivalent up to consistent renaming and fix an allocation
+scheme, this relation defines a deterministic machine (the relation is
+really a function).
 
@@
 interpretation, set the implicit parameters just as above, but drop
-the $\maddr \not\in \msto$ condition. The $\interpdelta$ relation takes some care
+the $\maddr \not\in \msto$ condition. $\interpdelta$ takes some care
 to not make the analysis run forever; a simple instantiation is a flat
-abstraction where arithmetic operations return an abstract top element
-$\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on
-$\sNum$.  This family of interpreters is also non-deterministic in
-choices of addresses, but it is free to choose addresses that are
-already in use.  Consequently, the machines may be non-deterministic
-when multiple values reside in a store location.
+abstraction that returns an abstract top element $\mathit{Integer}$ as
+the result of arithmetic operations.  This family of interpreters is
+also non-deterministic in choices of addresses, but it is free to
+choose addresses that are already in use.  Consequently, the machines
+may be non-deterministic when multiple values reside in a store
+location.
 
 It is important to recognize from this definition that \emph{any}
-allocation strategy is a sound abstract
+allocation strategy is an abstract
 interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
@@
 an interpretation that allocates a single cell into which all bindings
-and continuations are stored.  The former is an abstract
+and continuations are stored.  On the one hand is an abstract
 interpretation that is non-computable and gives only the ground truth
-of a programs behavior; the latter is an abstract interpretation
+of a programs behavior; on the other is an abstract interpretation
 that is easy to compute but gives little information.  Useful program
@@
 choice of address representation and allocation strategy.  Uniform
-\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97}, presented next, is one such analysis.
+\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97} is one such analysis.
 
@@
 
-{\small
 \begin{align*}
@@
 \alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
-\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
+\push_\mlab^\mcntr(\msto,\mkont) &=
+  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
 \tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
-\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
+\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\msto,\mval}) &= \mcntr \\
 \tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
@@
   \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
-\end{align*}}
+\end{align*}
 The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
@@
 
-\section{From machine semantics to baseline analyzer}
+\section{Reduction semantics to baseline analyzer}
 \label{sec:baseline}
@@
 The uniform $k$-CFA allocation strategy would make $\traces$ in figure
-\ref{fig:aam} a computable abstraction of possible executions, but one
-that is too inefficient to run, even on small examples.  Through this
-section, we explain a succession of approximations to reach a more
-appropriate baseline analysis.
-%
-We ground this path by first formulating the analysis in terms of a
-classic fixed-point computation.
+\ref{fig:aam} a computable abstraction of possible executions, but not an
+efficient one. This is not the strategy that AAM, nor we, recommend. Through
+this section, we will explain a succession of approximations
+to reach the baseline analysis.  We'll compare performance at each stage to
+identify the criticality of each choice. 
+%
+We ground this journey by first formulating the analysis in terms of a classic
+fixed-point computation.
 
@@
 %
-Given the initial state $\mstate_0$ and the transition relation $\machstep$,
+Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
 we define the global transfer function:
 \begin{equation*}
- F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
+ F_{\varsigma_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
+ \text.
 \end{equation*}
 Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
-\begin{align*}
-  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
-    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
-    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
-\end{align*}
+\begin{equation*}
+  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
+\end{equation*}
 Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
-% \begin{equation*}
- $\eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}$
-%\end{equation*}
-where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.
-
-The possible traces of execution tell us the most about a program, so
-we take $\traces(\mexp)$ to be the (regular) set of paths through the
-computed graph. We elide the construction of the set of edges in this paper.
+\begin{equation*}
+ \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
+\end{equation*}
+where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.
+
+{\bf FIXME} use $\traces$
 
@@
 
-In the next subsection, we fix this with store widening to reach polynomial
-(albeit of high degree) complexity.
+In the next subsection, we'll fix this with a widening and reach polynomial
+(albeit of a high degree) complexity.
 %
@@
 %
-Thus, complexity drops from intractable exponentiality to a merely
+Thus, in one step, complexity drops from intractable exponentially to a merely
 daunting polynomial.
@@
 %
-Then, before each transition, it installs this joined store into current
+Then, before each transition, it will install this joined store into current
 state.
 
-To describe this process, AAM defined a transformation of the reduction relation so that it operates on
+To describe this process, we'll refactor the transition relation so that it operates on
 a pair of a set of contexts ($C$) and a store ($\sigma$).
@@
 %
-The transformed relation, $\widehat{\machstep}$, is
-%
-{\small
-\begin{align*}
-(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
+The refactored relation, $\widehat{\machstep}$, becomes:
+%
+\begin{align*}
+(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
 \mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
               \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
-\wn &: \Context \times \Store \to \State \\
 \wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
-\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
+\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
 \wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
 \wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
-\end{align*}}
-%
-The new store is computed as the least upper bound of all
-stores after stepping in order to have only one, and keep the
-semantics sound.
-
-\subsection{Store-allocate all values}
+\end{align*}
+%
+In effect, the new store is computed as the least upper bound of all
+stores after stepping. This definition is wasteful, however. It causes
+all states found so far to step each iteration, even if they are not
+revisited. This has negative performance \emph{and} precision
+consequences. We instead use a frontier-based semantics that
+corresponds to the classic ``worklist'' algorithms for analysis. The
+difference is that the store is not modified in-place, but updated
+after all frontier states have been processed. This has implications
+for the analysis' precision and determinism. Specifically, higher
+precision, and it is deterministic even if set iteration is not.
+
+\begin{align*}
+(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
+\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
+              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
+              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
+\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
+\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
+\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
+\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval} \\
+\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
+\text{ where } c_0 &= \ev{e,\bot,\kmt}
+\end{align*}
+
+For clarity, we will exhibit non-widened semantics unless a technique
+specifically requires it.
+
+\subsection{Store-allocated results}
 \label{sec:baselineeval}
 
-The final approximation we make to get to our baseline is to
-store-allocate all values that appear, so that any non-machine state
-that contains a value instead contains an address to a value.  The AAM
-approach stops at the previous optimization.  However, the {\tt fn}
-continuation stores a value, and this makes the space of continuations
-quadratic rather than linear in the size of the program, for a
-monovariant analysis like 0-CFA.  Having the space of continuations
-grow linearly with the size of the program will drop the overall
-complexity to cubic (as expected).
+The final approximation we make to get to our point of departure is
+store-allocating results of application sub-expressions. 
+The AAM approach stops at the previous optimization. 
+However, the {\tt fn} continuation stores a value, and this makes the space of
+continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0-CFA.
+Having the space of continuations grow linearly with the size of the program 
+will drop the overall complexity to cubic (as expected).
 
@@
 space grows super-exponentially. We extend the semantics to
-additionally allocate an address for the function value when creating
-the $\kfn{}$ continuation. The continuation has to contain this address
-to remember where to retreive values from in the store.
-
-The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
+additionally allocate addresses to store the values of each
+subexpression of an application. The continuations have to store these
+addresses to remember where to get argument values from, though
+sometimes your allocation strategy may allow you to elide that
+representation.
+
+The new evaluation rules follow (where $\mcntr' = \tick(\mlab,\mcntr)$).
 % In theory, this aggressive distinction between continuations might buy
 % additional precision, but in practice, it does not.
-\begin{align*}
-\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
-\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
+
+\newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
+%\newcommand{\ext}[3]{ext(#1,#2,#3)}
+
+\begin{align*}
+\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
+\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar{\mexpi{1},\menv,\maddr_f,\maddr_a,\maddr_\mkont}} \\
+\text{ where }
+  \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) \\
+  \maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
+\end{align*}
+This rule describes how we allocate addresses for subexpressions of an
+application up front. This adds another parameter to the machine, but
+it is very much like $\alloc$. We use $f$ and $a$ subscripts to denote
+function and argument positions, respectively. Whereas ANF names all
+intermediate values in the syntax, we allocate addresses for
+intermediate values. This representational leap is not a big one, but
+is notably important for analyzing direct style terms.
+
+\begin{align*}
+\co[^\mcntr]{\kar{\mexp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \machstep
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_a,\maddr_\mkont}} \\
+\text{ where }
+  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
+\\
+\co[^\mcntr]{\kfn{\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \machstep
+\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr_a,\mkont,\msto'}
+\\
+\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f) \\
+\text{ where }
+ \msto' &= \ext{\msto}{\maddr_a}{\{\mval\}}
+\end{align*}
+We associate the value in hand with the expression that produced
+it. Also, we fan out to all functions that are associated with the
+original function position, because applied functions are in
+\emph{strict position} (more about that later).
+
+\begin{align*}
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\maddr_a, \msto,\mkont} & \machstep
+\ev[^{\mcntr'}]{\mexp,\menv',\msto',\mkont} \\
 \text{ where }
 \maddr &= \alloc(\mstate) \\
-  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
-\end{align*}
-Now instead of storing the evaluated function in the continuation
-frame itself, we indirect it through the store for further control on
-complexity and precision:
-\begin{align*}
-\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
-\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
-\\
-\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
-\end{align*}
-Associated with this indirection, we now apply all functions stored in
-the address. This nondeterminism is demanded in order to continue with
-evaluation. More generally, this is because applied functions are in
-\emph{strict position} (more about that later).
+\menv' &= \menv[\mvar\mapsto\maddr] \\
+\msto' &= \msto\sqcup[\maddr \mapsto \msto(\maddr_a)]
+\end{align*}
+Arguments are not passed as values, but as addresses to their values.
+
+\begin{align*}
+\ap[^\mcntr_\mlab]{\mop,\maddr_a,\msto,\mkont} & \machstep
+\co[^{\mcntr'}]{\mkont,\mval',\msto}
+\\
+\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\maddr_a)\}
+\end{align*}
+Primitive application still nondeterministically produces all possible values.
 
@@
 %
-We have two broad categories of these optimizations: (1) pragmatic
-improvement, (2) transition elimination.
-%
+We have two broad categories of these optimizations: (1) transition elimination
+and (2) pragmatic improvement.
+%
+The transition-elimination optimizations reduce the overall number of transitions
+made by the analyzer by performing:
+\begin{enumerate}
+ \item lazy non-determinism and
+ \item abstract compilation;
+\end{enumerate}
 The pragmatic improvements reduce overhead and trade space for time
@@
 \begin{enumerate}
- \item timestamped stores;
- \item store deltas; and
- \item imperative, pre-allocated data structures.
+  \setcounter{enumi}{2}
+ \item store deltas;
+ \item timestamped stores; and
+ \item preallocated data structures.
 \end{enumerate}
-The transition-elimination optimizations reduce the overall number of transitions
-made by the analyzer by performing:
-\begin{enumerate}
-  \setcounter{enumi}{3}
- \item frontier-based semantics;
- \item lazy non-determinism; and
- \item abstract compilation.
-\end{enumerate}
-
-All pragmatic improvements are precision preserving (form complete
-abstractions), but the semantic changes are not in some cases, for reasons we
-will describe. We did not observe the precision differences in our evaluation.
-
-We apply the frontier-based semantics combined with timestamped stores
-as our first step.  The move to the imperative will be made last in
-order to show the effectiveness of these techniques in the purely
-functional realm.
-
-\subsection{Timestamped frontier}
-
-The semantics given for store widening in section \ref{sec:storewiden},
-while simple, is wasteful. It also does not model what typical
-implementations do. It causes all states found so far to step each
-iteration, even if they are not revisited. This has negative
-performance \emph{and} precision consequences (changes to the store
-can travel back in time in straight-line code). We instead use a
-frontier-based semantics that corresponds to the classic worklist
-algorithms for analysis. The difference is that the store is not
-modified in-place, but updated after all frontier states have been
-processed. This has implications for the analysis' precision and
-determinism. Specifically, higher precision, and it is deterministic
-even if set iteration is not.
-
-The state space changes from a store and set of contexts to a set
-of seen abstract states (context plus store), $S$, a set of contexts to
-step (the frontier), $F$, and a store to step those contexts with, $\msto$:
-\begin{equation*}
-(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
-\end{equation*}
-
-We constantly see more states, so $S$ is always growing. The frontier,
-which is what remains to be done, changes. Let's start with the result
-of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
-%
-\begin{equation*}
-I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
-\end{equation*}
-
-The next store is the least upper bound of all the stores in $I$:
-
-\begin{equation*}
-\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
-\end{equation*}
-%
-The next frontier is exactly the states that we found from stepping
-the last frontier, but have not seen before. They must be
-\emph{states}, so we pair the contexts with the next store:
-%
-\begin{equation*}
-F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
-\end{equation*}
-%
-Finally, we add what we know we had not yet seen to the seen set:
-\begin{equation*}
-  S' = \setof{(c,\msto') \mid c \in F'}
-\end{equation*}
-%
-To inject a program $e$ into this machine, we start off knowing we
-have seen the first state, and that we need to process the first
-state:
-\begin{align*}
-\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
-\text{ where } c_0 &= \ev{e,\bot,\kmt}
-\end{align*}
-
-Notice that now $S$ has several copies of the abstract store in it. As
-it is, this semantics is much less efficient (but still more precise)
-than the previously proposed semantics because membership checks have
-to compare entire stores. Checking equality is expensive because the stores within each are
-large, and every entry must be checked against every other. Hashes can
-sometimes rule out inequality relatively quickly, but the incidence of
-collisions and actual equality is costly.
-
-And, there is a better way. Shivers' original work on $k$-CFA was
-susceptible to the same problem, and he suggested three complementary
-optimizations: (1) make the store global; (2) update the store
-imperatively; and (3) associate every change in the store with a
-version number -- its timestamp. Then, put timestamps in states
-where previously there were stores. Given two states, the analysis can
-now compare their stores just by comparing their timestamps -- a
-constant-time operation.
-
-There are two subtle losses of precision in Shivers' original
-timestamp technique that we can fix.
-
-\begin{enumerate}
-\item{In our semantics, the store does not change until the entire
-    frontier has been explored. This avoids cross-branch pollution
-    which would not otherwise happen, e.g., when one branch writes to
-    address $\maddr$ and another branch reads from address
-    $\maddr$.}
-\item{The common implementation strategy for timestamps destructively
-    updates each state's timestamp. This loses \emph{temporal}
-    information about the contexts a state is visited in, and in what
-    order. The semantics is different - we can't precisely answer
-    questions about the traces of the reduction relation we defined.
-    Our semantics has a drop-in replacement of timestamps for stores
-    in the seen set ($S$), so we do not experience precision loss.}
-\end{enumerate}
-
-{\small
-\begin{align*}
-\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
-\end{align*}
-\begin{align*}
-(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
-\mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
-              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
-              (t',\Sigma') &=\left\{\begin{array}{ll}
-                           (t+1,\msto'\Sigma') & \text{ if } \msto' \neq \msto \\
-                           (t,\Sigma)   & \text{ otherwise}
-                          \end{array}\right. \\
-              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
-              S' &= \setof{(c,t') \mid c \in F'}\\
-\inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
-\text{ where } c_0 &= \ev{e,\bot,\kmt}
-\end{align*}}
-
-The observation Shivers made was that the store is increasing
-monotonically, so all stores throughout execution will be totally
-ordered (form a chain). This observation allows you to replace stores
-with pointers into this chain. We keep the stores around in $\Sigma$
-to acheive a complete abstraction. This corresponds to the temporal
-information about the execution's effect on the store.
-
-Note also that $F$ is only populated with states that have not been
-seen at the resulting store. This is what produces the more precise
-abstraction than the baseline widening.
-
-The general fixed-point combinator we showed above can be specialized
-to this semantics, as well. In fact, $\widehat{\machstep}^T$ is a functional
-relation, so we can get the least fixed-point of it directly.
-
-\begin{lemma}
-  $\widehat{\machstep}$ maintains the invariant that all stores in $S$ are
-  totally ordered and $\msto$ is an upper bound of the stores in $S$.
-\end{lemma}
-
-\begin{lemma}
-  $\widehat{\machstep}^T$ maintains the invariant that $\Sigma$ is in
-  order with respect to $\sqsupset$ and $\msto = \hd(\Sigma)$.
-\end{lemma}
-
-\begin{theorem}
-$\widehat{\machstep}^T$ is a complete abstraction of $\widehat{\machstep}$.
-\end{theorem}
-The proof follows from the Galois connection that, in one direction,
-sorts all the stores in $S$ to form $\Sigma$, and translates stores in
-$S$ to their distance from the end of $\Sigma$ (their timestamp). In the other
-direction, timestamps in $S$ are replaced by the stores they point to in $\Sigma$.
-
-\subsection{Locally log-based store deltas}
-
-The above technique requires joining entire (large) stores
-together. Additionally, there is still a comparison of stores, which
-we established is expensive. Not every step will modify all addresses
-of the store, so joining entire stores is wasteful in terms of memory
-and time. We can instead log store changes and replay the change log
-on the full store after all steps have completed, noting when there is
-an actual change. This uses far fewer join and comparison operations,
-leading to less overhead, and is precision-preserving.
-
-We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
-  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
- becomes a log addition
-$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
-empty ($\mtlst$) for each step. Applying the changes to the full store
-is straightforward:
-\begin{equation*}
-\replay : (\StoreDelta \times \Store) \to (\Store \times \Boolean)
-\end{equation*}
-\begin{align*}
-\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &=
-\ttuple{\msto'}
-       {\mval{s_i} \overset{?}{=} \msto'(\maddr_i) \vee \ldots} \\
-\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots}
-\end{align*}
-
-We change the semantics slightly to add to the change log rather than
-produce an entire modified store.  The transition relation is
-identical except for the addition of this change log.  We maintain the
-invariant that lookups will never rely on the change log, so we can
-use the originally supplied store unmodified.
-
-A taste of the changes to the reduction relation is as follows:
-\begin{equation*}
-\dmachstep \subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) 
-\end{equation*}
-\begin{align*}
-\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
-\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
-\text{ where }\maddr &= \alloc(\mstate) \\
-              \menv' &= \menv[\mvar\mapsto\maddr]
-\end{align*}
-
-% Compilation changes to additionally take a $\msdiff$ component, so the
-% above rule's right hand side would instead be $k^{\mcntr'}(\menv',
-% \msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.
-
-We lift $\dmachstep$ to accommodate for the asymmetry
-in the input and output, and change the frontier-based semantics in
-the following way:
-{\small
-\begin{align*}
-(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
-\mbox{ where }
- I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
- \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
- \ttuple{t'}{\Sigma'} &=
-     \left\{
-       \begin{array}{ll}
-         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
-         \ttuple{t}{\Sigma} & \text{ otherwise}
-       \end{array}\right. \\
- F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
- S' &= \setof{(c,t') \mid c \in F'} \\
-\appendall(\varnothing) &= \epsilon \\
-\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
-\end{align*}}
-
-Here $\appendall$ combines change logs across all non-deterministic
-steps for a state to later be replayed. The order the combination
-happens in doesn't matter, because join is associative and
-commutative.
-
-\begin{lemma}
-$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff
-$\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
-\end{lemma}
-By cases on $\dmachstep$ and $\machstep$.
-
-\begin{lemma}[$\changep$ means change]
-Let $\replay(\msdiff,\msto) = \ttuple{\msto'}{\changep}$. $\msto' \neq \msto$ iff $\changep$.
-\end{lemma}
-By induction on $\msdiff$.
-
-\begin{theorem}
-$\damachstep$ is a complete abstraction of $\widehat{\machstep}^T$.
-\end{theorem}
-Follows from previous lemma and that join is associative and commutative.
-
-% Correctness here requires a lemma that relates the differences in the compilation functions.
-
-% \begin{lemma}[Compile coherence]
-% Let $\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$
-% and $\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \wn(\mastate',\msto^{*'})$.
-% If $\msto \equiv \msto^*$, $\msdiff \equiv \msdiff^*$, and $\mkont \equiv \mkont^*$, then
-% $\mastate' \equiv \mastate$
-% and there exists an $\msdiff''$ such that \\
-%    $\replay(\msdiff'',\replay(\msdiff^*,\msto^*)) \equiv \replay(\msdiff',\msto)$.
-% \end{lemma}
-% Here $\equiv$ means structurally the same, only compiled expressions
-% are compiled with the updated compilation function that builds store
-% deltas. Proof is by induction on $\mexp$.
-
-% \begin{theorem}[Equivalence with wide abstract compilation]
-% If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
-% $(S,F,\msto) \machstep (S',F',\msto')$ iff
-% $\exists S^*_1, F^*_1, \msto^*_1.
-%   S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
-%   (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
-% \end{theorem}
-% Proof by the above lemma and that replaying in any order is equivalent
-% (associativity and commutativity of $\sqcup$).
+
+All techniques are precision preserving under reasonable
+assumptions. We will discuss these assumptions as the caveats arise.
 
@@
 until and unless it is needed in {\it strict contexts} (such as the guard for an
-{\tt if}, a called procedure, or a numerical primitive application).
+{\tt if} a called procedure, or a numerical primitive application). 
 %
@@
 \renewcommand{\ext}{\mathit{ext}}
-{\small
+
 \begin{align*}
@@
  \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
- \force(\msto,\mval) &= \setof{\mval}}
+ \force(\mval) &= \{\mval\}}
 {\force &: \Value \to \Set(\Value) \\
@@
 %
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto}
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto}
 &\lmachstep\;
-\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}} \\
+\ev[^\mcntr]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont}} \\
 \text{ where }
-\maddr_f &= \alloc(\mstate) \\
 \msto' &=
@@
 %
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\mval,\msto}
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
 &\lmachstep\;
-\ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
-\text{ if } \mkont &\in \msto(\mkaddr),
+\ev[^\mcntr]{\mexpi,\menv,\msto,\mkont} \\
+\text{ if } \mkont &\in \msto(\maddr_\mkont),
             \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
-\end{align*}}
+\end{align*}
 Since {\tt if} guards are in strict position, we must force the value
@@
 warrant the significant increase in time and memory needed for this
-eager non-determinism. Without a widened store, lazy
-non-determinism is not a complete abstraction with either approach,
-but the precision gains are not immediately apparent due to fake
-rebinding (a second reference to the same variable chooses a different
-value from the first) and other problems. The performance and memory
-gains are apparent, so we favor the simplicity over invisible
-gains.}{}
+eager non-determinism. Indeed, were the variable reference a step
+later and another binding not made in that step, the results of the
+two approaches are the same.
+}{}
 
@@
 \end{theorem}
-Here $\sqsubseteq$ is straightforward - the left-hand side store must be
-contained in the right-hand-side store, and if values occur in the states, forcing
-the left-hand-side value must be a subset of forcing the corresponding right-hand-side value.
+Here $\sqsubseteq$ is straightforward - the LHS store must be
+contained in the RHS store, and if values occur in the states, forcing
+the LHS value must be a subset of forcing the corresponding RHS value.
 The proof is by cases on $\mstate \machstep \mstate'$.}
@@
 \machstep\; &
-\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_1}}
+\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_{f1},\maddr_{a1},\maddr_1}}
 \\
 \machstep\; &
-\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_2}}
+\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_{f2},\maddr_{a2},\maddr_2}}
 \\
 \machstep\; &
-\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_3}} % {\mexp_2}
+\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3}} % {\mexp_2}
 \\
@@
 \spchoice
-{\co{\kar{\mexp_1,\menv,\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
-{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
+{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto_4,\saddr{\menv(\mvar)}}}
+{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto_4,\superposition{\msto_4(\menv(\mvar))}}}
 \end{align}
@@
 \begin{align*}
-\compile{\mexp}^\mcntr(\msto_0)(\menv,\epsilon,\mkont) =
+\compile{\mexp}^\mcntr(\menv,\msto_0,\mkont) = 
 \spchoice
-{\co{\kar{\compile{\mexp_1},\menv,\maddr_3},\saddr{\menv(\mvar)}},\msdiff_4}
-{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
-\end{align*}
-where
-\begin{equation*}
-\msdiff_4 = [\ttuple{\maddr_1}{\setof{\mkont}},
-             \ttuple{\maddr_2}{\setof{\kar{\compile{\mexp_3},\menv,\maddr_1}}},
-             \ttuple{\maddr_3}{\setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}}]\text.
-\end{equation*}
+{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto'_4,\saddr{\menv(\mvar)}}}
+{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\msto'_4,\superposition{\msto'_4(\menv(\mvar))}}}
+\end{align*}
+where \(
+\msto'_4 = \msto_0 \sqcup \{ [\maddr_1 \mapsto \{ \mkont \}],
+[\maddr_2 \mapsto \kar{\compile{\mexp_3},\menv,\maddr_1}],
+[\maddr_3 \mapsto \kar{\compile{\mexp_2},\menv,\maddr_2}]\).
 
@@
 from the previous semantics is that $\ev{}$ state construction is
-replaced by calling the compiled expression. For notational coherence,
-we write $\lambda^\mcntr(\mathit{args} \ldots)$ for
-$\lambda(\mathit{args} \ldots, \mcntr)$ and
-$\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
-\mcntr)$.
+replaced by calling the compiled expression.
 
 \begin{figure}
-\small
-\begin{align*}
-\compile{\_} &: \Expr \to \Store \\
-             &\phantom{: \Expr } \to \Env  \times \StoreDelta \times \Kont \times \Counter \\
-             &\phantom{: \Expr } \to \State \\
-\mcntr' &= \tick(\mlab,\menv,\msto,\mcntr) \\
-\compile{\svar\mvar}_\msto &=
- \lambda^\mcntr(\menv,\msdiff,\mkont) .
+\begin{align*}
+\compile{\_} &: \Expr \to \Env \times \Store \times \Kont \to \State \\
+\compile{\svar\mvar} &=
+ \lambda(\menv,\msto,\mkont) .
 \spchoice
-{\co{\mkont,\saddr{\menv(\mvar)}},\msdiff}
-{\co{\mkont,\superposition{\msto(\menv(\mvar))}},\msdiff}
-\\
-\compile{\slit\mlit}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
-\co{\mkont,\mlit},\msdiff
-\\
-\compile{\slam\mvar\mexp}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
-\co{\mkont,\clos{\mvar,\compile{\mexp},\menv}},\msdiff
-\\
-\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
-\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\mkaddr})
-\\
-&\setlength\arraycolsep{5pt}
-\begin{array}{lrl}
-\text{ where } & \mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-               & \msdiff' = \cons{\ttuple{\mkaddr}{\setof{\mkont}}}{\msdiff}
+{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
+{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}}
+\\
+\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
+\co{\mkont,\mlit,\msto}
+\\
+\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
+\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
+\\
+\compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
+\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_f,\maddr_a,\maddr_\mkont})
+\\
+&\setlength\arraycolsep{5pt}\begin{array}{lrl}\text{ where } & \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) 
+\\&
+\maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
 \end{array}
 \\
-\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
-\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
-\\
-&\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&\phantom{\text{ where }} \msdiff' = \cons{\ttuple{\mkaddr}{\setof{\mkont}}}{\msdiff}
+\compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
+\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
+\\
+&\text{ where }\maddr,\msto' = \push_\mlab^\mcntr(\msto,\mkont)
 \end{align*}
@@
 \begin{figure}
-\small
 \begin{gather*}
 \begin{align*}
-\traces(\mexp) &= \setof{ \inject(\compile{\mexp}^\mtcntr_\bot(\bot,\epsilon,\kmt)) \multimachstep \mstate}
-                    \text{ where } \\
-\inject(c,\msdiff) &= \wn(c,\replay(\msdiff,\bot)) \\
-\wn(c,\msto) \machstep \wn(c',\msto') &\iff c \cmachstep_\msto c',\msdiff \\
-\msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
+\traces(\mexp) &= \{ \compile{\mexp}^\mtcntr(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
 \end{align*}
@@
 %% CONTINUE
-\co{\kmt,\mval} &\cmachstep_\msto
-\ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
-\\
-\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\mkaddr},\mval} & \cmachstep_\msto
-\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}) \\
-\text{ where } \maddr_f &= \alloc(\mstate) \\
-               \msdiff &= \cons{\ttuple{\maddr_f}{\force(\msto,\mval)}}{\epsilon}
-\\
-\co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
-\ap[^\mcntr_\mlab]{\mval,\mval\mkont},\epsilon \\
-\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
-\\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue} & \cmachstep_\msto
-\mcompi{0}^\mcntr(\msto)(\menv,\epsilon,\mkont)
+\co{\kmt,\mval,\msto} &\cmachstep
+\ans{\msto,\mval'} \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
+\\
+\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \cmachstep
+\mcomp^\mcntr(\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont}) \\
+\text{ where } \msto' &= \msto\sqcup[\maddr_f \mapsto \spchoice{\force(\msto,\mval)}
+                                                              {\force(\mval)}]
+\\
+\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \cmachstep
+\ap[^\mcntr_\mlab]{\mval,\maddr_a,\mkont,\msto'} \\
+\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\maddr_\mkont) \\
+\text{ where } \msto' &= \msto\sqcup[\maddr_f \mapsto \spchoice{\force(\msto,\mval)}
+                                                               {\force(\mval)}]
+\\
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \cmachstep
+\mcompi0^\mcntr(\menv,\msto,\mkont)
 \text{ if }\mkont\in\msto(\maddr)
 \\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse} & \cmachstep_\msto
-\mcompi{1}^\mcntr(\msto)(\menv,\epsilon,\mkont)
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \cmachstep
+\mcompi1^\mcntr(\menv,\msto,\mkont)
 \text{ if }\mkont\in\msto(\maddr)
@@
 %% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\mkont} & \cmachstep_\msto
-\mcomp^{\mcntr'}(\msto)(\menv',\msdiff,\mkont) \\
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\maddr_a,\msto,\mkont} & \cmachstep
+\mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
 \text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
-              \msdiff &= \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon}
-\\
-\ap{\mop,\mval,\mkont} & \cmachstep_\msto
-\co{\mkont,\mval'},\epsilon \\
+              \msto' &= \msto\sqcup[\maddr\mapsto\msto(\maddr_a)]
+\\
+\ap{\mop,\mval,\msto,\mkont} & \cmachstep
+\co{\mkont,\mval',\msto} \\
 \text{ where } \mvalx{u} &\in \spchoice{\force(\msto,\mval)}{\force(\mval)}, \mval'\in\interpdelta(\mop,\mvalx{u})
@@
 semantics line up again (modulo representation of expressions). This
-constitutes the use of a notion of stuttering. We provide a formal
-analysis of abstract compilation \emph{without} store widening with a proof
-of a stuttering bisimulation~\cite{ianjohnson:BCG88} between this
-semantics and lazy non-determinism without widening to show precision
-preservation. %We revisit store widening at the end of the section.
+constitutes the use of a notion of stuttering. We therefore prove a
+stuttering bisimulation~\cite{ianjohnson:BCG88} between this
+semantics and the previous section's to show precision preservation.
 
@@
 %
-\begin{lemma}[Compile/commit]
-Let $c,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
-Let $\wn(c',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
-$\wn(c,\replay(\msdiff',\msto)) = \wn(c',\replay(\msdiff,\msto'))$.
+\begin{lemma}[Compile/commit]\hfill\\
+$\compile{\mexp}^\mcntr(\menv,r(\msto),r(\mkont)) = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
 \end{lemma}
@@
 \begin{theorem}[Precision preservation]
-$B$ is a WEB on $\lmachstep \uplus \machstep$
+$B$ is a WEB on $\lmachstep \cup \cmachstep$
 \end{theorem}
 
-The proof follows by cases on $\lmachstep \uplus \machstep$ with the
-WEB \emph{witness} being the well-order on expressions (with a $\bot$
-element), and the following $\erankt$, $\erankl$ functions:
+The full proof is in the appendix, but it follows by cases on
+$\lmachstep \cup \cmachstep$ with the WEB \emph{witness} being the
+well-order on expressions (with a $\bot$ element), and the following
+$\erankt$, $\erankl$ functions:
 
@@
 \paragraph{Wide store and abstract compilation}
+A small problem arises if we try to widen the store as in section
+\ref{sec:storewiden}. In one step, we can both bind a variable and
+reference it. The problem is that the store does not change until all
+steps are done, so the reference unsoundly returns the previous
+binding (or nothing at all)! There are a couple ways to address
+this. One way is to make all compiled forms that reference the store
+into a new kind of state that has to be stepped in order to reference
+the store and continue. Another way is to make function application
+take one extra step after binding, so that running the body of the
+function sees the changed store. The latter is the more enticing of
+the two, since programs have more variable references than function
+calls - we don't want to slow them down if we don't have to.
+
 The fact that many changes happen in one step also makes the
@@
 values in the store, leading to a mismatch from the previous
-semantics. This kind of cross-step store poisoning is inevitable when
-a store is shared between several traces. The binding could have just
-as easily been observed the other way around in the non-compiled
-semantics. Call $\camachstep$ the result of the widening operator from
-the previous section on $\cmachstep$.
+semantics. This kind of cross-step store poisoning is inevitable when a
+store is shared between several traces. The binding could have just as
+easily been observed the other way around in the non-compiled
+semantics.
 
@@
 
+\subsection{Locally log-based store deltas}
+
+Every step the analysis makes for the above techniques requires
+joining entire (large) stores together. Not every step will modify all
+addresses of the store, so joining entire stores is wasteful in terms
+of memory and time. We can instead log store changes and replay the
+change log on the full store after all steps have completed. This uses
+far fewer join operations, leading to less overhead, and is precision-preserving.
+
+We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
+  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
+ becomes a log addition
+$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
+empty ($\mtlst$) for each step. Applying the changes to the full store
+is straightforward:
+% TODO: Turn the recursion here into ...:
+\begin{align*}
+\replay &: \StoreDelta \times \Store \to \Store \\
+\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &= \msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots
+\end{align*}
+%\replay(\mtlst, \msto) &= \msto \\
+%\replay(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}, \msto) &= \replay(\msdiff, \msto\sqcup[\maddr \mapsto \mval{s}])
+
+The transition relation is identical except for the addition of this change log.
+%
+Lookups will never rely on the change log, and can use the originally supplied store unmodified:
+
+\begin{align*}
+(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont},\msto,\msdiff) & \dmachstep
+(\ev[^{\mcntr'}]{\mexp,\menv',\mkont},\msdiff') \\
+\text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
+              \msdiff' &= 
+\spchoice
+{\cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\msdiff}}
+{\cons{\ttuple{\maddr}{\force(\mval)}}{\msdiff}}
+\end{align*}
+Compilation changes to additionally take a $\msdiff$ component, so the
+above rule's right hand side would instead be $k^{\mcntr'}(\menv',
+\msto, \msdiff', \mkont)$ where $k = \compile{e}$ would be in the closure.
+
+We also lift $\dmachstep$ to accommodate for this asymmetry
+in the input and output. For each state that is stepped, we feed the
+output changes to the next so that all changes get accumulated:
+
+\begin{align*}
+(S, F, \msto) &\mathrel{\damachstep} (S \cup S', F', \msto') \\
+\mbox{ where } (F', \msto') &= \step^*(\varnothing, F, \msto) \\
+               S' &= \setof{(c,\msto') \mid c \in F'} \\
+\step^*(F', \varnothing, \msto) &= (S, \msdiff) \\
+\step^*(F', \{c\}\cup F, \msto) &= \step^*(F'\cup \states^*, F, \msto^*) \\
+\states^* &= \{c' \mid (c,\msto,\msdiff) \dmachstep (c',\msdiff^c) \} \\
+\msto^* &= \replayall(\{\msdiff^c \mid (c, \msto,\msdiff) \dmachstep (c', \msdiff^c)\}, \msto)
+\end{align*}
+
+Here $\replayall : \Set(\StoreDelta) \times \Store \to \Store$ replays all logs in a set on top of the given store. Since join is associative and commutative, the order does not matter:
+\begin{align*}
+  \replayall(\varnothing,\msto) &= \msto \\
+  \replayall(\setof{\msdiff} \cup \Xi, \msto) &= \replayall(\Xi, \replay(\msdiff, \msto))
+\end{align*}
+
+Correctness here requires a lemma that relates the differences in the compilation functions.
+\begin{lemma}[Compile coherence]
+If $\msto \equiv \msto^*$ and $\mkont \equiv \mkont^*$, then
+$\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$ iff
+$\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \mstate$ such that
+$\mastate,\replay(\msdiff',\msto) \equiv \nw(\mstate)$
+\end{lemma}
+Proof is by induction on $\mexp$.
+
+\begin{theorem}[Equivalence]
+If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
+$(S,F,\msto) \machstep (S',F',\msto')$ iff
+$\exists S^*_1, F^*_1, \msto^*_1.
+  S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
+  (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
+\end{theorem}
+Here $\equiv$ means structurally the same, only compiled expressions
+are compiled with the updated compilation function that builds store
+deltas. The proof follows by equational reasoning and the previous lemma.
+
 %% Delta Store +k Compile + Lazy:
@@
 
-\subsection{Imperative, pre-allocated data structures}
+\subsection{Timestamping an imperative store}
 
 Thus far, we have made our optimizations in a purely functional
-manner. For the final push for performance, we need to dip into the
-imperative. In this section, we show an alternative representation of
-the store and seen set that are more space-efficient and are amenable
-to destructive updates.
+manner. For the next series of optimizations, we need to dip into the
+imperative. We can motivate this entire sequence of optimizations by
+focusing on the largest bottleneck in the current state-space
+exploration algorithm: checking to see if a state has already been
+seen. Given two states, checking equality is expensive because the
+stores within each are large, and every entry must be checked against
+every other. Hashes can sometimes rule out inequality relatively
+quickly, but the incidence of collisions and actual equality is
+costly.
+
+And, there is a better way. Shivers' original work on $k$-CFA was
+susceptible to the same problem, and he suggested three complementary
+optimizations: (1) make the store global; (2) update the store
+imperatively; and (3) associate every change in the store with a
+version number -- its timestamp. Then, put timestamps in states
+where previously there were stores. Given two states, the analysis can
+now compare their stores just by comparing their timestamps -- a
+constant-time operation.
+
+There are two subtle losses of precision in Shivers' original
+time-stamp technique that we can fix.
+
+\begin{enumerate}
+ \item{For a given abstract
+state, all writes to the global store should be delayed until the
+analysis considers all current work. This avoids
+cross-branch pollution which would not otherwise happen, e.g., when
+one branch writes to address $\maddr$ and another branch reads from
+address $\maddr$. Fortunately, given our conversion to log-based
+stores this change is straightforward.}
+\item{The common implementation strategy for timestamps destructively
+    updates each state's timestamp. This loses \emph{temporal}
+    information about the contexts a state is visited in, and in what
+    order. The semantics is different - we can't precisely answer
+    questions about the traces of the reduction relation we defined. A
+    simple fix is to keep a stack of the timestamps seen at each
+    state, and keep a list of stores corresponding with each
+    timestamp. A more space-efficient representation that also allows
+    an imperative store would be timestamping each update to the store
+    in the same style as states. This changes the notion of store
+    equality, but the new definition is straightforward. For analysis
+    consumers that disregard temporal information, the space cost in
+    unwarranted, but the realization that there is precision to be
+    gained opens possibilities.}
+\end{enumerate}
 
@@
 \mvalstack \sqcup_t \mval{s} &= \mvalstack,\sfalse \text{ otherwise}
-\end{align*}}
-
-For the purposes of space, we reuse the $\cmachstep$ semantics,
-although the $\replay$ of the produced $\msdiff$ objects should be
-in-place, and the $\lookup$ function should be using this
-single-threaded store.  Because the store has all the temporal
-information baked into it, we rephrase the core semantics in terms of
-a transfer function. The least fixed-point of this function gives a
-more compact representation of the reduction relation of the previous
-section.
-
-{\small
-\begin{align*}
-\System &= (\widehat{\State} \to {\Timestamp}^*) \times \wp(\widehat{\State}) \times \Store \times \Timestamp \\
+\end{align*}
+
+For the purposes of space, we reuse the $\dmachstep$ semantics and change the definition $\replay$ to track if the store has changed, in order to increment the global timestamp.
+\begin{align*}
+\replay_t(\epsilon,\msto,\changep) &= \msto,\changep \\
+\replay_t(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff},\msto,\changep) &= \replay_t(\msdiff,\msto',\changep \vee \changep') \\
+\text{ where } (\msto',\changep') &= \msto\sqcup_t[\maddr \mapsto, \mval{s}] \\
+\replayall_t(\varnothing,\msto,\changep) &= \msto,\changep \\
+\replayall_t(\setof{\msdiff}\cup\Xi,\msto,\changep) &= \replayall_t(\Xi, \msto', \changep') \\
+\text{ where } (\msto',\changep') &= \replay_t(\msdiff,\msto,\changep)
+\end{align*}
+
+Because the store has all the temporal information baked into it, we
+rephrase the core semantics in terms of a transfer function. The least
+fixed-point of this function gives a more compact representation of
+the reduction relation of the previous section.
+
+\begin{align*}
+\System &= (\State \to {\Timestamp}^*) \times \wp(\State) \times \Store \times \Timestamp \\
 {\mathcal F} &: \System \to \System \\
@@
 \text{ where }
-I &= \setof{(c',\msdiff) \mid
-       c \in F,
-       c \cmachstep_{\msto^*} c',\msdiff} \\
+I &= \setof{(\mstate',\msdiff') \mid
+       \mstate \in F,
+       (\mstate,\msto^*,\epsilon) \dmachstep
+       (\mstate',\msdiff')} \\
 \msto^* &= \lambda \maddr.\lookup(\msto(\maddr),t) \\
-\ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
+(\msto',\changep) &= \replayall_t(\setof{\msdiff \mid (\_,\msdiff) \in I},\msto,\sfalse) \\
 t' &= \left\{\begin{array}{ll} t+1 & \text{ if } \changep \\
@@
              \end{array}\right. \\
-F' &= \setof{c \mid (c,\_) \in I, \changep \vee S(c) \neq \cons{t}{\_}} \\
-S' &= \lambda c. \left\{\begin{array}{ll}
-                               \cons{t'}{S(c)} & \text{ if } c \in F' \\
-                               S(c) & \text{ otherwise}
+F' &= \setof{\mstate \mid (\mstate,\_) \in I, \changep \vee S(\mstate) \neq \cons{t}{\_}} \\
+S' &= \lambda \mstate. \left\{\begin{array}{ll}
+                               \cons{t'}{S(\mstate)} & \text{ if } \mstate \in F' \\
+                               S(\mstate) & \text{ otherwise}
                              \end{array}\right.
-\end{align*}}
-
-We prove semantic equivalence with the previous semantics with a
-lock-step bisimulation with the stack of stores abstraction, which
-follow from equational reasoning from the following lemmas:
-
-\begin{lemma}
-Stores of value stacks completely abstract stacks of stores.
-\end{lemma}
-This depends on some wellformedness conditions about the order of the
-stacks. The store of value stacks can be translated to a stack of
-stores by taking successive ``snapshots'' of the store at different
-timestamps from the max timestamp it holds down to 0. Vice versa, we
-replay the changes across adjacent stores in the stack.
-
-We apply a similar construction to the different representation of seen states in order to get the final result:
-
-\begin{theorem}
-${\mathcal F}$ is a complete abstraction of $\camachstep$.
-\end{theorem}
+\end{align*}
 
@@
 \item{$F$: the \emph{frontier} set, which must be persisent or copied for the iteration through the set to be correct.}
-\item{$\msto$: the store, which represents all stores that occur in the machine semantics.}
+\item{$\msto$: the store, which represents all stores that occur in the reduction semantics.}
 \item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}
-\item{$\changep$: whether a store change stepped all states in $F$.}
+\item{$\changep$: whether there has been a store change stepping all states in $F$.}
 \end{enumerate}
@@
 
-\subsubsection{Pre-allocating the store}
-
-Internally, the algorithm at this stage uses hash tables to model the store to allow arbitrary address representations.
+% \begin{figure}\label{fig:pure-timestamp}
+% \begin{align*}
+% t \in \Timestamp &= {\mathbb N} \\
+% \msto \in \Store &= \Addr \to (\Timestamp \times \wp(\Storeable))^* \\
+% (S,F,\msto,t) &\tmachstep (S \cup S', F', \msto', t') \\
+% \text{ where } I &= \{(\mstate,\msdiff) \mid \mstate \in F, \mstate \
+% \end{align*}
+% \end{figure}
+
+% \begin{figure}
+% \begin{alltt}
+% eval(\(e\)) :=
+%   \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
+%   \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \mtlst, \kmt\))
+%   while(true):
+%     if todo = \(\varnothing\): return (keys(seen), \(\msto\))
+%     else:
+%       let old := todo
+%       todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
+%       foreach c \(\in\) old: \(\changep\) := false; c()
+%       unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
+% \end{alltt}
+% \caption{Imperative algorithm}
+% \label{fig:imperative}
+% \end{figure}
+
+\subsection{Pre-allocating the store}
+
+Internally, the algorithm at this stage uses hash tables to model the store.
+%
+This is because stores used to be distributed to all states, which
+required a compact, dynamic representation.
 %
@@
 we know its maximum size.
-%
-In a monovariant allocation strategy, the domain of the store is
+
+In a monovariant analysis, the domain of the store is
 exactly the set of expressions in the program.
@@
 
-\item {\bf mbrotZ}: generates Mandelbrot fractal using complex
+\item {\bf mbrotZ}: generates Mandelbrot set fractal using complex
   numbers.
@@
 plots the factors of improvement over the baseline for each
-optimization step.
-
-% The dip we see in transition rate even though time
-% taken decreases is to be expected - fewer ``easy'' states are added by
-% abstract compilation. It increases again with the introduced
-% algorithmic improvements. Accumulating store changes in addition to
-% maintaining the store accounts for the higher memory usage when using
-% the store delta technique without further improvements.
+optimization step. The dip we see in transition rate even though time
+taken decreases is to be expected - fewer ``easy'' states are added by
+abstract compilation. It increases again with the introduced
+algorithmic improvements. Accumulating store changes in addition to
+maintaining the store accounts for the higher memory usage when using
+the store delta technique without further improvements.
 
@@
 \end{figure*}
-
-To determine the impact of these techniques on precision, we evaluated
-the a singleton variable analysis to find opportunities to inline
-constants and closed functions.  We found no change in the results
-across all implementations, including Shivers' timestamp
-approximation---from an empirical point of view, these techniques are
-precision preserving despite the theoterical loss of precision.
 
@@
 pre-allocate most data structures and split the abstract store out
-into parts that are more quickly accessed and updated. Our semantic
-optimizations can still be applied to an analysis that does abstract
-garbage collection~\cite{dvanhorn:Might:2006:GammaCFA}, whereas the
-polymorphic splitting implementation is tied strongly to a
+into parts that are more quickly accessed and updated. The first two
+of our optimizations can still be applied to an analysis that does
+abstract garbage collection~\cite{dvanhorn:Might:2006:GammaCFA},
+whereas the polymorphic splitting implementation is tied strongly to a
 single-threaded store.
@@
 implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
-the \Church{} benchmark about 5.5 times faster than the fastest
+the \Church{} benchmark about 25 times faster than the fastest
 implementation considered here.  These analyses compute very different
@@
 %\input{paper.bbl}
-
 \bibliography{local,bibliography}

### 2013-04-02 paper/paper.tex
Backpedaling on shift to superposition semantics. Causes too many problems. Still have results to prove.
Conflicts:

	paper/paper.tex

@@
 \item continuations are heap-allocated, not stack-allocated,
-\item there are ``timestamps'' (written $\mcntr$) and syntax
+\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
   labels ($\mlab$) threaded through the computation, and
@@
 \tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
-\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\msto,\mval}) &= \mcntr \\
+\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\mval,\msto}) &= \mcntr \\
 \tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
@@
 %
-Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
+Given the initial state $\mstate_0$ and the transition relation $\machstep$,
 we define the global transfer function:
 \begin{equation*}
- F_{\varsigma_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
- \text.
+ F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
 \end{equation*}
 Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
-\begin{equation*}
-  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
-\end{equation*}
+\begin{align*}
+  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
+    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
+    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
+\end{align*}
 Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
 \begin{equation*}
- \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
+ \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
 \end{equation*}
-where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.
-
-{\bf FIXME} use $\traces$
+where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.
+
+The possible traces of execution tell us the most about a program, so
+we take $\traces(\mexp)$ to be the (regular) set of paths through the
+computed graph. We elide the construction of the set of edges in this paper.
 
@@
 For clarity, we will exhibit non-widened semantics unless a technique
-specifically requires it.
-
-\subsection{Store-allocated results}
+specifically requires it. The final step to the baseline takes the
+complexity from quartic to cubic in the monovariant case.
+
+\subsection{Store-allocate all values}
 \label{sec:baselineeval}
 
-The final approximation we make to get to our point of departure is
-store-allocating results of application sub-expressions. 
-The AAM approach stops at the previous optimization. 
-However, the {\tt fn} continuation stores a value, and this makes the space of
-continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0-CFA.
-Having the space of continuations grow linearly with the size of the program 
-will drop the overall complexity to cubic (as expected).
+The final approximation we make to get to our point of departure is to
+store-allocate all values that appear, so that any non-machine state
+that contains a value instead contains an address to a value.  The AAM
+approach stops at the previous optimization.  However, the {\tt fn}
+continuation stores a value, and this makes the space of continuations
+quadratic rather than linear in the size of the program, for a
+monovariant analysis like 0-CFA.  Having the space of continuations
+grow linearly with the size of the program will drop the overall
+complexity to cubic (as expected).
 
@@
 
-All techniques are precision preserving under reasonable
-assumptions. We will discuss these assumptions as the caveats arise.
+All pragmatic improvments are precision preserving (form complete
+abstractions), but the semantic changes are not in some cases, for reasons we
+will describe. We did not observe the precision differences in our evaluation.
 
@@
 \\
-\compile{\slit\mlit} &= \lambda(\menv,\msto,\mkont) .
+\compile{\slit\mlit} &= \lambda^\mcntr(\menv,\msto,\mkont) .
 \co{\mkont,\mlit,\msto}
 \\
-\compile{\slam\mvar\mexp} &= \lambda(\menv,\msto,\mkont) .
-\co{\mkont,\clos{\mvar,\compile\mexp,\menv},\msto}
+\compile{\slam\mvar\mexp} &= \lambda^\mcntr(\menv,\msto,\mkont) .
+\co{\mkont,\clos{\mvar,\compile{\mexp},\menv},\msto}
 \\
 \compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^\mcntr(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_f,\maddr_a,\maddr_\mkont})
-\\
-&\setlength\arraycolsep{5pt}\begin{array}{lrl}\text{ where } & \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) 
-\\&
-\maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
-\end{array}
+\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
+\\
+&\setlength\arraycolsep{5pt}
+\begin{array}{lrl}\text{ where } & \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont)\end{array}
 \\
 \compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}} &= \lambda^\mcntr(\menv,\msto,\mkont) .
-\compile{\mexpi0}^\delta(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
+\compile{\mexpi0}^{\mcntr'}(\menv,\msto',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
 \\
@@
 \end{lemma}
+Here $\equiv$ means structurally the same, only compiled expressions
+are compiled with the updated compilation function that builds store
+deltas. The proof follows by equational reasoning and the previous lemma.
 Proof is by induction on $\mexp$.
 
-\begin{theorem}[Equivalence]
+\begin{theorem}[Equivalence with wide abstract compilation]
 If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
@@
 \end{theorem}
-Here $\equiv$ means structurally the same, only compiled expressions
-are compiled with the updated compilation function that builds store
-deltas. The proof follows by equational reasoning and the previous lemma.
+Proof by {\bf TODO}
 
@@
 
+We increase the timestamp only if we find that the store has changed
+from any join operation. The frontier set only gets states that we
+haven't seen at the next timestamp. The states we've seen gets updated
+with the frontier states at the next timestamp. We prove semantic
+equivalence with the previous semantics with a lock-step bisimulation,
+which follow from equational reasoning from the following lemmas:
+
+\begin{lemma}[Store of value stacks completely abstracts stack of stores]
+TODO
+\end{lemma}
+
+\begin{lemma}[$\changep$ means change (a)]
+TODO
+\end{lemma}
+
+\begin{lemma}[$\changep$ means change (b)]
+TODO
+\end{lemma}
+
+\begin{theorem}[Timestamped system with stack of stores completely abstracts wide store deltas]
+TODO
+\end{theorem}
+
 \subsubsection{Pure to imperative}
@@
 updated with the new states at the next timestamp.
-
-% \begin{figure}\label{fig:pure-timestamp}
-% \begin{align*}
-% t \in \Timestamp &= {\mathbb N} \\
-% \msto \in \Store &= \Addr \to (\Timestamp \times \wp(\Storeable))^* \\
-% (S,F,\msto,t) &\tmachstep (S \cup S', F', \msto', t') \\
-% \text{ where } I &= \{(\mstate,\msdiff) \mid \mstate \in F, \mstate \
-% \end{align*}
-% \end{figure}
-
-% \begin{figure}
-% \begin{alltt}
-% eval(\(e\)) :=
-%   \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
-%   \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \mtlst, \kmt\))
-%   while(true):
-%     if todo = \(\varnothing\): return (keys(seen), \(\msto\))
-%     else:
-%       let old := todo
-%       todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
-%       foreach c \(\in\) old: \(\changep\) := false; c()
-%       unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
-% \end{alltt}
-% \caption{Imperative algorithm}
-% \label{fig:imperative}
-% \end{figure}
 
@@
 
+We evaluated the precision of these techniques with a singleton
+variable analysis to find opportunities to inline constants and closed
+functions. We found no change in the results across all
+implementations, including Shivers' timestamp approximation.
+
 Source code of the implementation and benchmark suite is at:
@@
 of analysis.  Consequently, we have omitted a complete comparison with
-these implementations,.  The AAM approach provides more precision in
+these implementations.  The AAM approach provides more precision in
 terms of temporal-ordering of program states, which comes at a cost

### 2013-04-02 paper/paper.tex
More work on initial sections.
Conflicts:

	paper/paper.tex

--- 2013-03-29 paper/paper.tex
@@
 
-\authorinfo{J. Ian Johnson}
-           {Northeastern University}
-           {ianj@ccs.neu.edu}
-\authorinfo{Nicholas Labich}
-           {Northeastern University}
-           {labichn@ccs.neu.edu}
-\authorinfo{Matthew Might}
-           {University of Utah}
-           {might@cs.utah.edu}
-\authorinfo{David Van Horn}
-           {Northeastern University}
-           {dvanhorn@ccs.neu.edu}
+\authorinfo{J. Ian Johnson \and Nicholas Labich \and Matthew Might \and David Van Horn}
+           {}
+           {}
+%% \authorinfo{J. Ian Johnson}
+%%            {Northeastern University}
+%%            {ianj@ccs.neu.edu}
+%% \authorinfo{Nicholas Labich}
+%%            {Northeastern University}
+%%            {labichn@ccs.neu.edu}
+%% \authorinfo{Matthew Might}
+%%            {University of Utah}
+%%            {might@cs.utah.edu}
+%% \authorinfo{David Van Horn}
+%%            {Northeastern University}
+%%            {dvanhorn@ccs.neu.edu}
 \maketitle
@@
 
+The AAM approach, in essence, does the following: it takes a
+machine-based view of computation and turns it into a \emph{finitary
+  approximation} by bounding the size of the store.  With a limited
+address space, the store must map addresses to \emph{sets} of values.
+Store updates are interpreted as joins, and store dereferences are
+interpreted by non-deterministic choice of an element from a set.  The
+result of analyzing a program is a finite directed graph where nodes
+in the graph are (abstract) machine states and edges denote machine
+transitions between states.
+
+The techniques we propose for optimizing analysis fall into the
+following categories:
+\begin{enumerate}
+\item generate fewer states by avoiding the eager exploration of
+  non-deterministic choices that will later collapse into a single
+  join point.  We accomplish this by applying lazy evaluation
+  techniques so that non-determinism is evaluated \emph{by need}.
+
+\item generate fewer states by avoiding unnecessary, intermediate
+  states of a computation.  We accomplish this by applying compilation
+  techniques from functional languages to avoid interpretive overhead
+  in the machine transition system.
+
+\item generate states faster.  We accomplish this by better algorithm
+  design in the fixed-point computation we use to generate state graphs.
+\end{enumerate}
+Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
+small example due to Earl, et
+al.~\cite{dvanhorn:Earl2012Introspective}.
+By generating significantly fewer states at a significantly faster
+rate, we are able to achieve large performance improvements in terms
+of both time and space.
+
+Section~\ref{sec:eval} describes the evaluation of each optimization
+technique applied to an implementation supporting a more realistic set
+of features, including mutation, first-class control, compound data, a
+full numeric tower and many more forms of primitive data and
+operations.
+%
+We evaluate this implementation against a set of benchmark programs
+drawn from the literature.
+%
+For all benchmarks, the optimized analyzer outperforms the baseline
+by at least a factor of
+% 475
+two to
+% 4,382
+three orders of magnitude.
+
+Section~\ref{sec:related} relates this work to the literature and
+section~\ref{sec:conclusion} concludes.
+
 %\newpage
@@
 \begin{figure}
-\small
 \begin{gather*}
@@
 \ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
-\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
+\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
 \\
 &&
-\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
+\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
 \\
 \ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
-\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
+\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
 \\
 &&
-\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
+\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
 \\[2mm]
@@
 \begin{align*}
-\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
-\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
-\end{align*}
-These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
-used to determinize the choice of fresh addresses. The $\sqcup$ on
-stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
-\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
-relation is non-deterministic in its choice of addresses, however it
-must always choose a fresh address when allocating a continuation or
-variable binding.  If we consider machine states equivalent up to
-consistent renaming and fix an allocation scheme, this relation
-defines a deterministic machine (the relation is really a function).
+\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
+\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
+\end{align*}
+where $\sqcup$ on stores is a point-wise lifting of $\sqcup$, \ie, $\msto
+\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
+\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
+$\mcntr$, but they can be used to determinize the choice of fresh
+addresses.  The resulting relation is non-deterministic in its choice
+of addresses, however it must always choose a fresh address when
+allocating a continuation or variable binding.  If we consider machine
+states equivalent up to consistent renaming and fix an allocation
+scheme, this relation defines a deterministic machine (the relation is
+really a function).
 
@@
 \tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
-\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
+\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\mval,\msto}) &= \mcntr \\
 \tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
@@
 The uniform $k$-CFA allocation strategy would make $\traces$ in figure
-\ref{fig:aam} a computable abstraction of possible executions, but one
-that is too inefficient to run, even on small examples.  Through this
-section, we explain a succession of approximations to reach a more
-appropriate baseline analysis.
-%
-We ground this path by first formulating the analysis in terms of a
-classic fixed-point computation.
+\ref{fig:aam} a computable abstraction of possible executions, but not an
+efficient one.  Through
+this section, we explain a succession of approximations
+to reach the baseline analysis.  At each stage, we measure performance to
+identify the importance of each choice.
+%
+We ground this path by first formulating the analysis in terms of a classic
+fixed-point computation.
 
@@
 Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
-% \begin{equation*}
- $\eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}$
-%\end{equation*}
+\begin{equation*}
+ \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
+\end{equation*}
 where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.
@@
 
-In the next subsection, we fix this with store widening to reach polynomial
+In the next subsection, we fix this with store widening and reach a polynomial
 (albeit of high degree) complexity.
@@
 %
-Thus, complexity drops from intractable exponentiality to a merely
+Thus, in one step, complexity drops from intractable exponentiality to a merely
 daunting polynomial.
@@
 %
-Then, before each transition, it installs this joined store into current
+Then, before each transition, it will install this joined store into current
 state.
@@
 %
-{\small
 \begin{align*}
@@
               \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
-\wn &: \Context \times \Store \to \State \\
+\wn &: \widehat{\State} \times \Store \to \State \\
 \wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
@@
 \wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
-\end{align*}}
-%
-The new store is computed as the least upper bound of all
-stores after stepping in order to have only one, and keep the
-semantics sound.
+\end{align*}
+%
+In effect, the new store is computed as the least upper bound of all
+stores after stepping. For clarity, we will exhibit non-widened
+semantics unless a technique specifically requires it. The final step
+to the baseline takes the complexity from quartic to cubic in the
+monovariant case.
 
@@
 additionally allocate an address for the function value when creating
-the $\kfn{}$ continuation. The continuation has to contain this address
-to remember where to retreive values from in the store.
+the $\kfn{}$ continuation. The continuation has to store this address
+to remember where to retreive values from in the store, though some allocation
+strategies allow this representation to be ellided.
 
@@
 % additional precision, but in practice, it does not.
-\begin{align*}
-\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
-\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
+
+\newcommand{\ext}[3]{#1\sqcup[#2\mapsto#3]}
+%\newcommand{\ext}[3]{ext(#1,#2,#3)}
+
+\begin{align*}
+\co[^\mcntr]{\kar{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_\mkont}} \\
 \text{ where }
-  \maddr &= \alloc(\mstate) \\
-  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
-\end{align*}
-Now instead of storing the evaluated function in the continuation
-frame itself, we indirect it through the store for further control on
-complexity and precision:
-\begin{align*}
-\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
+  \maddr_f &= \alloc(\mstate) \\
+  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
+\\
+\co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
 \ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
 \\
-\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
-\end{align*}
-Associated with this indirection, we now apply all functions stored in
-the address. This nondeterminism is demanded in order to continue with
-evaluation. More generally, this is because applied functions are in
+\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f)
+\end{align*}
+We associate the value in hand with the expression that produced
+it. Also, we fan out to all functions that are associated with the
+original function position, because applied functions are in
 \emph{strict position} (more about that later).
@@
  \item store deltas; and
- \item imperative, pre-allocated data structures.
+ \item imperative, preallocated data structures.
 \end{enumerate}
@@
   \setcounter{enumi}{3}
- \item frontier-based semantics;
- \item lazy non-determinism; and
- \item abstract compilation.
+ \item frontier-based semantics
+ \item lazy non-determinism and
+ \item abstract compilation;
 \end{enumerate}
@@
 can travel back in time in straight-line code). We instead use a
-frontier-based semantics that corresponds to the classic worklist
+frontier-based semantics that corresponds to the classic ``worklist''
 algorithms for analysis. The difference is that the store is not
@@
 
-The state space changes from a store and set of contexts to a set
-of seen abstract states (context plus store), $S$, a set of contexts to
-step (the frontier), $F$, and a store to step those contexts with, $\msto$:
-\begin{equation*}
-(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
-\end{equation*}
-
-We constantly see more states, so $S$ is always growing. The frontier,
-which is what remains to be done, changes. Let's start with the result
-of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
-%
-\begin{equation*}
-I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
-\end{equation*}
-
-The next store is the least upper bound of all the stores in $I$:
-
-\begin{equation*}
-\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
-\end{equation*}
-%
-The next frontier is exactly the states that we found from stepping
-the last frontier, but have not seen before. They must be
-\emph{states}, so we pair the contexts with the next store:
-%
-\begin{equation*}
-F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
-\end{equation*}
-%
-Finally, we add what we know we had not yet seen to the seen set:
-\begin{equation*}
-  S' = \setof{(c,\msto') \mid c \in F'}
-\end{equation*}
-%
-To inject a program $e$ into this machine, we start off knowing we
-have seen the first state, and that we need to process the first
-state:
-\begin{align*}
+\begin{align*}
+(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
+\mbox{where } I &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
+              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
+              F' &= \{c \mid (c,\_) \in I, (c,\msto') \notin S\} \\
+              S' &= \setof{(c,\msto') \mid c \in F'} \\
 \inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
@@
 
-{\small
-\begin{align*}
-\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
-\end{align*}
-\begin{align*}
+\begin{align*}
+\Sigma &\in \Store^* \\
+S &\subseteq {\mathbb N} \times \widehat{\State} \\
+F &\subseteq \widehat{\State} \\
 (S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
@@
 \text{ where } c_0 &= \ev{e,\bot,\kmt}
-\end{align*}}
+\end{align*}
 
@@
 \begin{theorem}
-$\widehat{\machstep}^T$ is a complete abstraction of $\widehat{\machstep}$.
+Timestamps are a complete abstraction
 \end{theorem}
-The proof follows from the Galois connection that, in one direction,
+The proof follows from the galois connection that, in one direction,
 sorts all the stores in $S$ to form $\Sigma$, and translates stores in
-$S$ to their distance from the end of $\Sigma$ (their timestamp). In the other
-direction, timestamps in $S$ are replaced by the stores they point to in $\Sigma$.
+$S$ to their distance from the end of $\Sigma$. In the other
+direction, timestamps are replaced by the stores they point to.
 
@@
 A taste of the changes to the reduction relation is as follows:
-\begin{equation*}
-\dmachstep \subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) 
-\end{equation*}
-\begin{align*}
-\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
+
+\begin{align*}
+\dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \dmachstep_\msto
 \ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
@@
 We lift $\dmachstep$ to accommodate for the asymmetry
-in the input and output, and change the frontier-based semantics in
-the following way:
-{\small
+in the input and output, and change the frontier-based semantics in the following way:
+
 \begin{align*}
@@
 \mbox{ where }
- I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
+ I &= \setof{(\mastate',\msdiff) \mid \mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff} } \\
  \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
@@
        \end{array}\right. \\
- F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
- S' &= \setof{(c,t') \mid c \in F'} \\
+ F' &= \setof{\mastate \mid (\mastate,\_) \in I, (\mastate,t') \notin S} \\
+ S' &= \setof{(\mastate,t') \mid c \in F'} \\
 \appendall(\varnothing) &= \epsilon \\
 \appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
-\end{align*}}
+\end{align*}
 
@@
 \begin{lemma}
-$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff
-$\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
+$\mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff}$ iff
+$\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
 \end{lemma}
@@
 %
-\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto}
+\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto}
 &\lmachstep\;
-\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}} \\
+\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
 \text{ where }
@@
 %
-\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\mval,\msto}
+\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
 &\lmachstep\;
 \ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
-\text{ if } \mkont &\in \msto(\mkaddr),
+\text{ if } \mkont &\in \msto(\maddr_\mkont),
             \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
-\end{align*}}
+\end{align*}
 Since {\tt if} guards are in strict position, we must force the value
@@
 \begin{figure}
-\small
 \begin{align*}
@@
 \compile{\sapp[^\mlab]{\mexpi0}{\mexpi1}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
-\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\mkaddr})
+\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kar[_\mlab^\mcntr]{\compile{\mexpi1},\menv,\maddr_\mkont})
 \\
@@
 \begin{array}{lrl}
-\text{ where } & \mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-               & \msdiff' = \cons{\ttuple{\mkaddr}{\setof{\mkont}}}{\msdiff}
+\text{ where } & \maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+               & \msdiff' = \cons{\ttuple{\maddr_\mkont}{\setof{\mkont}}}{\msdiff}
 \end{array}
@@
 \\
-&\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
-&\phantom{\text{ where }} \msdiff' = \cons{\ttuple{\mkaddr}{\setof{\mkont}}}{\msdiff}
+&\text{ where }\maddr_\mkont = \allockont^\mcntr_\mlab(\msto,\mkont) \\
+&\phantom{\text{ where }} \msdiff' = \cons{\ttuple{\maddr_\mkont}{\setof{\mkont}}}{\msdiff}
 \end{align*}
@@
                     \text{ where } \\
-\inject(c,\msdiff) &= \wn(c,\replay(\msdiff,\bot)) \\
-\wn(c,\msto) \machstep \wn(c',\msto') &\iff c \cmachstep_\msto c',\msdiff \\
+\inject(\mastate,\msdiff) &= \wn(\mastate,\replay(\msdiff,\bot)) \\
+\wn(\mastate,\msto) \machstep \wn(\mastate',\msto') &\iff \mastate \cmachstep_\msto \mastate',\msdiff \\
 \msdiff \text{ is such that } &\replay(\msdiff,\msto) = \msto'
@@
 \\
-\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\mkaddr},\mval} & \cmachstep_\msto
-\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}) \\
+\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr_\mkont},\mval} & \cmachstep_\msto
+\mcomp^\mcntr(\msto)(\menv,\msdiff,\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}) \\
 \text{ where } \maddr_f &= \alloc(\mstate) \\
@@
 \\
-\co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
+\co{\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont},\mval} & \cmachstep_\msto
 \ap[^\mcntr_\mlab]{\mval,\mval\mkont},\epsilon \\
-\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
+\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\maddr_\mkont)
 \\
@@
 semantics and lazy non-determinism without widening to show precision
-preservation. %We revisit store widening at the end of the section.
+preservation. We revisit store widening at the end of the section.
 
@@
 \begin{lemma}[Compile/commit]
-Let $c,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
-Let $\wn(c',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
-$\wn(c,\replay(\msdiff',\msto)) = \wn(c',\replay(\msdiff,\msto'))$.
+Let $\mastate,\msdiff' = \compile{\mexp}^\mcntr_{r(\msto)}(\menv,\msdiff,r(\mkont))$.
+Let $\wn(\mastate',\msto') = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
+$\wn(\mastate,\replay(\msdiff',\msto)) = \wn(\mastate',\replay(\msdiff,\msto'))$.
 \end{lemma}
@@
 \begin{theorem}[Precision preservation]
-$B$ is a WEB on $\lmachstep \uplus \machstep$
+$B$ is a WEB on $\lmachstep \cup \machstep$
 \end{theorem}
@@
 
-\subsection{Imperative, pre-allocated data structures}
+\subsection{Imperative, preallocated data structures}
 
@@
 \mvalstack \sqcup_t \mval{s} &= \mvalstack,\sfalse \text{ otherwise}
-\end{align*}}
+\end{align*}
 
@@
 \text{ where }
-I &= \setof{(c',\msdiff) \mid
-       c \in F,
-       c \cmachstep_{\msto^*} c',\msdiff} \\
+I &= \setof{(\mastate',\msdiff) \mid
+       \mastate \in F,
+       \mastate \cmachstep_{\msto^*} \mastate',\msdiff} \\
 \msto^* &= \lambda \maddr.\lookup(\msto(\maddr),t) \\
@@
              \end{array}\right. \\
-F' &= \setof{c \mid (c,\_) \in I, \changep \vee S(c) \neq \cons{t}{\_}} \\
-S' &= \lambda c. \left\{\begin{array}{ll}
-                               \cons{t'}{S(c)} & \text{ if } c \in F' \\
-                               S(c) & \text{ otherwise}
+F' &= \setof{\mastate \mid (\mastate,\_) \in I, \changep \vee S(\mastate) \neq \cons{t}{\_}} \\
+S' &= \lambda \mastate. \left\{\begin{array}{ll}
+                               \cons{t'}{S(\mastate)} & \text{ if } \mastate \in F' \\
+                               S(\mastate) & \text{ otherwise}
                              \end{array}\right.
-\end{align*}}
+\end{align*}
 
@@
 
-Internally, the algorithm at this stage uses hash tables to model the store to allow arbitrary address representations.
+Internally, the algorithm at this stage uses hash tables to model the store.
+%
+This is because stores used to be distributed to all states, which
+required a compact, dynamic representation.
 %
@@
 
-To determine the impact of these techniques on precision, we evaluated
-the a singleton variable analysis to find opportunities to inline
-constants and closed functions.  We found no change in the results
-across all implementations, including Shivers' timestamp
-approximation---from an empirical point of view, these techniques are
-precision preserving despite the theoterical loss of precision.
-
-Our step-wise optimizations strictly produce better analysis times
-with no observed loss of precision.  The final result is a
-systematically derived and verified implementation that operates
-within a small factor performance loss compared to a hand-optimized,
-unverified implementation.  Moreover, much of the performance gains
-are achieved with purely functional methods, which suggest potential
-further gains from parellelization.  Peak memory usage is often
-considerably improved by the end of the optimization steps, but the
-effect of mutation has unexpected consequence on the analyzers memory
-consumption.
+We evaluated the precision of these techniques with a singleton
+variable analysis to find opportunities to inline constants and closed
+functions. We found no change in the results across all
+implementations, including Shivers' timestamp approximation.
 

### 2013-04-02 paper/paper.tex
Use k for continuation addresses.
Conflicts:

	paper/paper.tex

@@
 \begin{align*}
-\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
-\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
-\end{align*}
-where $\sqcup$ on stores is a point-wise lifting of $\sqcup$, \ie, $\msto
-\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
-\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
-$\mcntr$, but they can be used to determinize the choice of fresh
-addresses.  The resulting relation is non-deterministic in its choice
-of addresses, however it must always choose a fresh address when
-allocating a continuation or variable binding.  If we consider machine
-states equivalent up to consistent renaming and fix an allocation
-scheme, this relation defines a deterministic machine (the relation is
-really a function).
+\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
+\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
+\end{align*}
+These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
+used to determinize the choice of fresh addresses. The $\sqcup$ on
+stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
+\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
+relation is non-deterministic in its choice of addresses, however it
+must always choose a fresh address when allocating a continuation or
+variable binding.  If we consider machine states equivalent up to
+consistent renaming and fix an allocation scheme, this relation
+defines a deterministic machine (the relation is really a function).
 
@@
 The uniform $k$-CFA allocation strategy would make $\traces$ in figure
-\ref{fig:aam} a computable abstraction of possible executions, but not an
-efficient one.  Through
-this section, we explain a succession of approximations
-to reach the baseline analysis.  At each stage, we measure performance to
-identify the importance of each choice.
-%
-We ground this path by first formulating the analysis in terms of a classic
-fixed-point computation.
+\ref{fig:aam} a computable abstraction of possible executions, but one
+that is too inefficient to run, even on small examples.  Through this
+section, we explain a succession of approximations to reach a more
+appropriate baseline analysis.
+%
+We ground this path by first formulating the analysis in terms of a
+classic fixed-point computation.
 
@@
 
-In the next subsection, we fix this with store widening and reach a polynomial
+In the next subsection, we fix this with store widening to reach polynomial
 (albeit of high degree) complexity.
@@
 additionally allocate an address for the function value when creating
-the $\kfn{}$ continuation. The continuation has to store this address
-to remember where to retreive values from in the store, though some allocation
-strategies allow this representation to be ellided.
+the $\kfn{}$ continuation. The continuation has to contain this address
+to remember where to retreive values from in the store.
 

### 2013-06-13 paper/paper.tex
Improved visualization for baseline/lazy/lazy&compiled.

@@
 \begin{figure}[t]
+\small
 \begin{center}
-\includegraphics[width=3.2in]{church-relative-space}
+%\includegraphics[width=3.2in]{church-relative-space}
 \includegraphics[width=3.2in]{church-relative-speed}
@@
 
+The AAM approach, in essence, does the following: it takes a
+machine-based view of computation and turns it into a \emph{finitary
+  approximation} by bounding the size of the store.  With a limited
+address space, the store must map addresses to \emph{sets} of values.
+Store updates are interpreted as joins, and store dereferences are
+interpreted by non-deterministic choice of an element from a set.  The
+result of analyzing a program is a finite directed graph where nodes
+in the graph are (abstract) machine states and edges denote machine
+transitions between states.
+
+The techniques we propose for optimizing analysis fall into the
+following categories:
+\begin{enumerate}
+\item generate fewer states by avoiding the eager exploration of
+  non-deterministic choices that will later collapse into a single
+  join point.  We accomplish this by applying lazy evaluation
+  techniques so that non-determinism is evaluated \emph{by need}.
+
+\item generate fewer states by avoiding unnecessary, intermediate
+  states of a computation.  We accomplish this by applying compilation
+  techniques from functional languages to avoid interpretive overhead
+  in the machine transition system.
+
+\item generate states faster.  We accomplish this by better algorithm
+  design in the fixed-point computation we use to generate state graphs.
+\end{enumerate}
+Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for the
+small motivating example in Earl, et
+al.~\cite{dvanhorn:Earl2012Introspective}.  By generating
+significantly fewer states at a significantly faster rate, we are able
+to achieve large performance improvements in terms of both time and
+space.
+
+Section~\ref{sec:eval} describes the evaluation of each optimization
+technique applied to an implementation supporting a more realistic set
+of features, including mutation, first-class control, compound data, a
+full numeric tower and many more forms of primitive data and
+operations.
+%
+We evaluate this implementation against a set of benchmark programs
+drawn from the literature.
+%
+For all benchmarks, the optimized analyzer outperforms the baseline
+by at least a factor of
+% 475
+two to
+% 4,382
+three orders of magnitude.
+
+Section~\ref{sec:related} relates this work to the literature and
+section~\ref{sec:conclusion} concludes.
+
 \begin{figure}[t]
+\small
 \begin{center}
@@
 \end{center}
-\caption{Example state graphs for the program above.  Part (a) shows
-  the result of the baseline analyzer.  It has long ``corridor''
-  transitions and ``diamond'' subgraphs that fan-out from
-  nondeterminism and fan-in from joins.  Part (b) shows the result of
-  performing nondeterminism lazily and thus avoids many of the diamond
-  subgraphs.  Part (c) shows the result of abstract compilation that
-  removes interpretive overhead in the form of intermediate states,
-  thus minimizing the corridor transitions.  The end result is a more
-  compact abstraction of the program that can be generated faster.}
+\caption{Example state graphs for Earl et. al. program.  Gray states
+  follow variable references, {\tt ev} states are black, and all others
+  are white. Part (a) shows the result of the baseline analyzer.  It
+  has long ``corridor'' transitions and ``diamond'' subgraphs that
+  fan-out from nondeterminism and fan-in from joins.  Part (b) shows
+  the result of performing nondeterminism lazily and thus avoids many
+  of the diamond subgraphs.  Part (c) shows the result of abstract
+  compilation that removes interpretive overhead in the form of
+  intermediate states, thus minimizing the corridor transitions.  The
+  end result is a more compact abstraction of the program that can be
+  generated faster.}
 \label{fig:state-graphs}
 \end{figure}
-
-The AAM approach, in essence, does the following: it takes a
-machine-based view of computation and turns it into a \emph{finitary
-  approximation} by bounding the size of the store.  With a limited
-address space, the store must map addresses to \emph{sets} of values.
-Store updates are interpreted as joins, and store dereferences are
-interpreted by non-deterministic choice of an element from a set.  The
-result of analyzing a program is a finite directed graph where nodes
-in the graph are (abstract) machine states and edges denote machine
-transitions between states.
-
-The techniques we propose for optimizing analysis fall into the
-following categories:
-\begin{enumerate}
-\item generate fewer states by avoiding the eager exploration of
-  non-deterministic choices that will later collapse into a single
-  join point.  We accomplish this by applying lazy evaluation
-  techniques so that non-determinism is evaluated \emph{by need}.
-
-\item generate fewer states by avoiding unnecessary, intermediate
-  states of a computation.  We accomplish this by applying compilation
-  techniques from functional languages to avoid interpretive overhead
-  in the machine transition system.
-
-\item generate states faster.  We accomplish this by better algorithm
-  design in the fixed-point computation we use to generate state graphs.
-\end{enumerate}
-Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
-small example due to Earl, et
-al.~\cite{dvanhorn:Earl2012Introspective}.
-By generating significantly fewer states at a significantly faster
-rate, we are able to achieve large performance improvements in terms
-of both time and space.
-
-Section~\ref{sec:eval} describes the evaluation of each optimization
-technique applied to an implementation supporting a more realistic set
-of features, including mutation, first-class control, compound data, a
-full numeric tower and many more forms of primitive data and
-operations.
-%
-We evaluate this implementation against a set of benchmark programs
-drawn from the literature.
-%
-For all benchmarks, the optimized analyzer outperforms the baseline
-by at least a factor of
-% 475
-two to
-% 4,382
-three orders of magnitude.
-
-Section~\ref{sec:related} relates this work to the literature and
-section~\ref{sec:conclusion} concludes.
 
@@
 \text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
-\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
+\text{States} & \mstate &=& \ev[^\mcntr]{\mexp,\menv,\msto,\mkont}\\
                        &&|& \co{\mkont,\mval,\msto}\\
-                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
+                       &&|& \ap[^\mcntr]{\mval,\mval,\msto,\mkont}\\
 \text{Continuations} & \mkont &=& \kmt\\
@@
 \text{Addresses} &\maddr&\in&\Addr \\
+\text{Times} &\mcntr&\in&\Counter \\
 \text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
-\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
+\text{Stores} &\msto&\in& \Addr \rightharpoonup \wp(\Value)
 \end{array}
@@
 components are defined in figure~\ref{fig:domains};
-figure~\ref{fig:aam} defines the transition relation.  The evaluation
-of a program is defined as its set of traces that arise from iterating the
-machine transition relation.  The
-machine is a very slight variation on a standard abstract machine for
-ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
-can be systematically derived from a definitional interpreter through
-a continuation-passing style transformation and defunctionalization,
-or from a structural operational semantics using the refocusing
+figure~\ref{fig:aam} defines the transition relation (unmentioned
+components stay the same).  The evaluation of a program is defined as
+its set of traces that arise from iterating the machine transition
+relation. The $\traces$ function produces the set of all proofs of
+reachability for any state $\mstate$ from the injection of program
+$\mexp$ (from which one could extract a string of states). The machine
+is a very slight variation on a standard abstract machine for ISWIM in
+``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It can be
+systematically derived from a definitional interpreter through a
+continuation-passing style transformation and defunctionalization, or
+from a structural operational semantics using the refocusing
 construction of Danvy and
@@
 More generally, we can have stores map to any domain that forms
- a Galois connection with values, enabling $\interpdelta$ to
+ a Galois connection with sets of values, enabling $\interpdelta$ to
 produce elaborate abstractions of base values (e.g., interval or octagon
@@
 \ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
-\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
+\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
 \\
@@
 %% CONTINUE
-\co{\kmt,\mval,\msto} &\machstep&
-\ans{\msto,\mval}
-\\
+%%\co{\kmt,\mval,\msto} &\machstep& \ans{\msto,\mval}
+%%\\
 \co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
@@
 \mtcntr &= \epsilon \\
-\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
 \alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
@@
 \tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
-  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
+  \kpush[_0]{\mcntr} &= \kpush[_k]{\epsilon} = \mtcntr \\
   \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
-\end{align*}
+\end{align*}}
 The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
@@
 Formally, we can cast this optimization as a second abstraction or as the
-application of a widening operator during the fixed-point iteration.
+application of a widening operator\footnote{Technically, we would have to copy the value of the global store to all states being stepped to fit the formal definition of a widening, but the consequential accelerated convergence fits the M.O.
+of a widening operator.} during the fixed-point iteration.
 %
@@
 \wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
-\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
-\end{align*}
-%
-In effect, the new store is computed as the least upper bound of all
-stores after stepping. For clarity, we will exhibit non-widened
-semantics unless a technique specifically requires it. The final step
-to the baseline takes the complexity from quartic to cubic in the
-monovariant case.
+%%\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
+\end{align*}}
+%
+The new store is computed as the least upper bound of all
+stores after stepping in order to have only one, and keep the
+semantics sound.
 
@@
 languages with $n$-ary functions, since otherwise the continuation
-space grows super-exponentially. We extend the semantics to
+space grows super-exponentially (${\mathcal O}(n^n)$). We extend the semantics to
 additionally allocate an address for the function value when creating
 the $\kfn{}$ continuation. The continuation has to contain this address
-to remember where to retreive values from in the store.
+to remember where to retrieve values from in the store.
 
@@
 frame itself, we indirect it through the store for further control on
-complexity and precision.
-
-\begin{align*}
-\co[^\mcntr]{\kfn{\maddr_f,\mkaddr},\mval,\msto} & \machstep
+complexity and precision:
+\begin{align*}
+\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
 \ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
 \\
-\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr_f)
+\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
 \end{align*}
 Associated with this indirection, we now apply all functions stored in
-the address. This nondeterminism is demanded in order to continue with
-evaluation. More generally, this is because applied functions are in
-\emph{strict position} (more about that later).
+the address. This nondeterminism is necessary in order to continue with
+evaluation.
 
@@
 
-\begin{align*}
-(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
-\mbox{where } I &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
-              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in I\} \\
-              F' &= \{c \mid (c,\_) \in I, (c,\msto') \notin S\} \\
-              S' &= \setof{(c,\msto') \mid c \in F'} \\
+The state space changes from a store and set of contexts to a set
+of seen abstract states (context plus store), $S$, a set of contexts to
+step (the frontier), $F$, and a store to step those contexts with, $\msto$:
+\begin{equation*}
+(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
+\end{equation*}
+
+We constantly see more states, so $S$ is always growing. The frontier,
+which is what remains to be done, changes. Let's start with the result
+of stepping all the contexts in $F$ paired with the current store (call it $I$ for intermediate):
+%
+\begin{equation*}
+I = \setof{(c',\msto') \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto'), c \in F}
+\end{equation*}
+
+The next store is the least upper bound of all the stores in $I$:
+
+\begin{equation*}
+\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
+\end{equation*}
+%
+The next frontier is exactly the states that we found from stepping
+the last frontier, but have not seen before. They must be
+\emph{states}, so we pair the contexts with the next store:
+%
+\begin{equation*}
+F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
+\end{equation*}
+%
+Finally, we add what we know we had not yet seen to the seen set:
+\begin{equation*}
+  S' = \setof{(c,\msto') \mid c \in F'}
+\end{equation*}
+%
+To inject a program $e$ into this machine, we start off knowing we
+have seen the first state, and that we need to process the first
+state:
+\begin{align*}
 \inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
@@
     frontier has been explored. This avoids cross-branch pollution
-    which would not otherwise happen, e.g., when one branch writes to
+    which would otherwise happen in Shivers' semantics, e.g., when one branch writes to
     address $\maddr$ and another branch reads from address
@@
     information about the contexts a state is visited in, and in what
-    order. The semantics is different - we can't precisely answer
-    questions about the traces of the reduction relation we defined.
-    Our semantics has a drop-in replacement of timestamps for stores
-    in the seen set ($S$), so we do not experience precision loss.}
+    order. Our semantics has a drop-in replacement of timestamps for
+    stores in the seen set ($S$), so we do not experience precision
+    loss.}
 \end{enumerate}
 
-\begin{align*}
-\Sigma &\in \Store^* \\
-S &\subseteq {\mathbb N} \times \widehat{\State} \\
-F &\subseteq \widehat{\State} \\
+{\small
+\begin{align*}
+\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
+\end{align*}
+\begin{align*}
 (S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
@@
 with pointers into this chain. We keep the stores around in $\Sigma$
-to acheive a complete abstraction. This corresponds to the temporal
+to achieve a complete abstraction. This corresponds to the temporal
 information about the execution's effect on the store.
@@
 \begin{theorem}
-Timestamps are a complete abstraction
+$\widehat{\machstep}^T$ is a complete abstraction of $\widehat{\machstep}$.
 \end{theorem}
-The proof follows from the galois connection that, in one direction,
+The proof follows from the order isomorphism that, in one direction,
 sorts all the stores in $S$ to form $\Sigma$, and translates stores in
-$S$ to their distance from the end of $\Sigma$. In the other
-direction, timestamps are replaced by the stores they point to.
+$S$ to their distance from the end of $\Sigma$ (their timestamp). In the other
+direction, timestamps in $S$ are replaced by the stores they point to in $\Sigma$.
 
@@
 A taste of the changes to the reduction relation is as follows:
-
-\begin{align*}
-\dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
+\begin{equation*}
+\dmachstep \subseteq (\Context\times\Store) \times (\Context\times\StoreDelta) 
+\end{equation*}
+\begin{align*}
 \ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
@@
 \mbox{ where }
- I &= \setof{(\mastate',\msdiff) \mid \ttuple{\mastate}{\msto} \dmachstep \ttuple{\mastate'}{\msdiff} } \\
+ I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
  \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
@@
        \end{array}\right. \\
- F' &= \setof{\mastate \mid (\mastate,\_) \in I, (\mastate,t') \notin S} \\
- S' &= \setof{(\mastate,t') \mid c \in F'} \\
+ F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
+ S' &= \setof{(c,t') \mid c \in F'} \\
 \appendall(\varnothing) &= \epsilon \\
 \appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
-\end{align*}
+\end{align*}}
 
@@
 \begin{lemma}
-$\ttuple{\mastate}{\msto} \dmachstep \ttuple{\mastate'}{\msdiff}$ iff
-$\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
+$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff
+$\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
 \end{lemma}
@@
 \end{theorem}
-Here $\sqsubseteq$ is straightforward - the LHS store must be
-contained in the RHS store, and if values occur in the states, forcing
-the LHS value must be a subset of forcing the corresponding RHS value.
+Here $\sqsubseteq$ is straightforward --- the left-hand side store must be
+contained in the right-hand-side store, and if values occur in the states,
+the left-hand-side value must be in the forced corresponding right-hand-side value.
 The proof is by cases on $\mstate \machstep \mstate'$.}
@@
 The following example illustrates
-the essence of abstract compilation effect:
+the essence of abstract compilation's effect:
 \[
@@
 \compile{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2}}_\msto &= \lambda^\mcntr(\menv,\msdiff,\mkont) .
-\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\maddr})
+\compile{\mexpi0}^{\mcntr'}(\menv,\msdiff',\kif[^\mcntr]{\compile{\mexpi1},\compile{\mexpi2},\menv,\mkaddr})
 \\
@@
 %% CONTINUE
-\co{\kmt,\mval} &\cmachstep_\msto
-\ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
-\\
+%%\co{\kmt,\mval} &\cmachstep_\msto \ans{\mval'},\epsilon \text{ if } \mval' \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
+%%\\
 \co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\mkaddr},\mval} & \cmachstep_\msto
@@
 \co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
-\ap[^\mcntr_\mlab]{\mval,\mval\mkont},\epsilon \\
-\text{ if }\mval &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
-\\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue} & \cmachstep_\msto
+\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont},\epsilon \\
+\text{ if }\mvalx{u} &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
+\\
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\mkaddr},\strue} & \cmachstep_\msto
 \mcompi{0}^\mcntr(\msto)(\menv,\epsilon,\mkont)
-\text{ if }\mkont\in\msto(\maddr)
-\\
-\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse} & \cmachstep_\msto
+\text{ if }\mkont\in\msto(\mkaddr)
+\\
+\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\mkaddr},\sfalse} & \cmachstep_\msto
 \mcompi{1}^\mcntr(\msto)(\menv,\epsilon,\mkont)
-\text{ if }\mkont\in\msto(\maddr)
+\text{ if }\mkont\in\msto(\mkaddr)
 \\[2mm]
@@
              \end{array}\right. \\
-F' &= \setof{\mastate \mid (\mastate,\_) \in I, \changep \vee S(\mastate) \neq \cons{t}{\_}} \\
-S' &= \lambda \mastate. \left\{\begin{array}{ll}
-                               \cons{t'}{S(\mastate)} & \text{ if } \mastate \in F' \\
-                               S(\mastate) & \text{ otherwise}
+F' &= \setof{c \mid (c,\_) \in I, \changep \vee S(c) \neq \cons{t}{\_}} \\
+S' &= \lambda c. \left\{\begin{array}{ll}
+                               \cons{t'}{S(c)} & \text{ if } c \in F' \\
+                               S(c) & \text{ otherwise}
                              \end{array}\right.
-\end{align*}
+\end{align*}}
 
@@
 \item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}
-\item{$\changep$: whether there has been a store change stepping all states in $F$.}
+\item{$\changep$: whether a store change occurred when stepping states in $F$.}
 \end{enumerate}
@@
 %
-For each benchmark, we collect analysis times, peak memory usage, and
-the rate of states-per-second explored by the analysis for each of the
+For each benchmark, we collect analysis times, peak memory usage (as
+determined by Racket's GC statistics), and the rate of
+states-per-second explored by the analysis for each of the
 optimizations discussed in section~\ref{sec:opt}, cumulatively
@@
 or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
-use the timeout limits as a lower bound on the actual time required,
+use the timeout limits as a lower bound on the actual time required (i.e., one minute versus timeout is at least 30 times faster),
 thus giving a conservative estimate of improvements.
@@
 All benchmarks are calculated as an average of 5 runs, done in
-parallel, on an 12-core, 64-bit Intel Xeon machine running at 2.40GHz
+parallel (each isolated to a core), on an 12-core, 64-bit Intel Xeon machine running at 2.40GHz
 with 12Gb of memory.
@@
 Many benchmarks cause the baseline analyzer to take longer than 30
-minutes or to consume more 1 gigabyte of memory, at which point the
+minutes or to consume more than 1 gigabyte of memory, at which point the
 analysis is stopped.  This is the case for the largest benchmark
@@
 
-  (a) Total analysis time
+  (a) Total analysis time speed-up (baseline / optimized)
 
@@
 
-  (b) Rate of state transitions
+  (b) Rate of state transitions speed-up (optimized / baseline)
 
@@
 
-  (c) Peak memory usage
+  (c) Peak memory usage improvement (baseline / optimized)
 \end{center}
@@
 
-We evaluated the precision of these techniques with a singleton
-variable analysis to find opportunities to inline constants and closed
-functions. We found no change in the results across all
-implementations, including Shivers' timestamp approximation.
+To determine the impact of each section's technique on precision, we evaluated
+a singleton variable analysis to find opportunities to inline
+constants and closed functions.  We found no change in the results
+across all implementations, including Shivers' timestamp
+approximation---from an empirical point of view, these techniques are
+precision preserving despite the theoterical loss of precision.
+
+Our step-wise optimizations strictly produce better analysis times
+with no observed loss of precision.  The final result is a
+systematically derived and verified implementation that operates
+within a small factor performance loss compared to a hand-optimized,
+unverified implementation.  Moreover, much of the performance gains
+are achieved with purely functional methods, which suggest potential
+further gains from parellelization.  Peak memory usage is often
+considerably improved by the end of the optimization steps, but the
+effect of mutation has unexpected consequence on the analyzers memory
+consumption.
 

### 2013-06-17 paper/paper.tex
More reviewer feedback incorporated. Need to start trimming back to 12 pages.

@@
 %\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
-\title{Correctly Optimizing Abstract Abstract Machines}
+\title{Optimizing Abstract Abstract Machines}
 
@@
 %
-Formally, we can cast this optimization as a second abstraction or as the
-application of a widening operator\footnote{Technically, we would have to copy the value of the global store to all states being stepped to fit the formal definition of a widening, but the consequential accelerated convergence fits the M.O.
-of a widening operator.} during the fixed-point iteration.
+Formally, we can cast this optimization as a second abstraction or as
+the application of a widening operator\footnote{Technically, we would
+  have to copy the value of the global store to all states being
+  stepped to fit the formal definition of a widening, but this
+  representation is order-isomorphic to that.} during the fixed-point
+iteration.
 %
@@
 that contains a value instead contains an address to a value.  The AAM
-approach stops at the previous optimization.  However, the {\tt fn}
+approach stops at the previous optimization.  However, the \kfnalone
 continuation stores a value, and this makes the space of continuations
@@
 grow linearly with the size of the program will drop the overall
-complexity to cubic (as expected).
+complexity to cubic (as expected). We also need to allocate an address
+for the argument position in an \apalone state.
 
@@
 additionally allocate an address for the function value when creating
-the $\kfn{}$ continuation. The continuation has to contain this address
+the $\kfnalone$ continuation. The continuation has to contain this address
 to remember where to retrieve values from in the store.
@@
 \co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
-\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
-\\
-\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr)
+\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr,\mkont,\msto'}
+\\
+\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr) \\
+\text{ where } \maddr &= \alloc(\mstate) \\
+               \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
 \end{align*}
+
 Associated with this indirection, we now apply all functions stored in
@@
 The state space changes from a store and set of contexts to a set
-of seen abstract states (context plus store), $S$, a set of contexts to
+of seen abstract states (context plus store), $\mseen$, a set of contexts to
 step (the frontier), $F$, and a store to step those contexts with, $\msto$:
 \begin{equation*}
-(S, F, \msto) \mathrel{\widehat{\machstep}} (S \cup S', F', \msto')
+(\mseen, F, \msto) \mathrel{\widehat{\machstep}} (\mseen \cup \mseen', F', \msto')
 \end{equation*}
 
-We constantly see more states, so $S$ is always growing. The frontier,
+We constantly see more states, so $\mseen$ is always growing. The frontier,
 which is what remains to be done, changes. Let's start with the result
@@
 \begin{equation*}
-F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin S}
+F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin \mseen}
 \end{equation*}
@@
 \begin{equation*}
-  S' = \setof{(c,\msto') \mid c \in F'}
+  \mseen' = \setof{(c,\msto') \mid c \in F'}
 \end{equation*}
@@
 
-Notice that now $S$ has several copies of the abstract store in it. As
+Notice that now $\mseen$ has several copies of the abstract store in it. As
 it is, this semantics is much less efficient (but still more precise)
@@
     order. Our semantics has a drop-in replacement of timestamps for
-    stores in the seen set ($S$), so we do not experience precision
+    stores in the seen set ($\mseentime$), so we do not experience precision
     loss.}
@@
 \begin{align*}
-\Sigma \in \Store^* \qquad S \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
+\Sigma \in \Store^* \qquad \mseentime \subseteq {\mathbb N} \times \Context \qquad F \subseteq \Context
 \end{align*}
 \begin{align*}
-(S, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (S \cup S', F', \msto', \Sigma', t') \\
+(\mseentime, F, \msto, \Sigma, t) &\mathrel{\widehat{\machstep}^T} (\mseentime \cup \mseentime', F', \msto', \Sigma', t') \\
 \mbox{where } I &= \setof{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F} \\
@@
                           \end{array}\right. \\
-              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
-              S' &= \setof{(c,t') \mid c \in F'}\\
+              F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin \mseentime} \\
+              \mseentime' &= \setof{(c,t') \mid c \in F'}\\
 \inject(e) &= (\setof{\ttuple{c_0}{0}},\setof{c_0},\bot,\cons{\bot}{\epsilon},0) \\
@@
 \begin{lemma}
-  $\widehat{\machstep}$ maintains the invariant that all stores in $S$ are
-  totally ordered and $\msto$ is an upper bound of the stores in $S$.
+  $\widehat{\machstep}$ maintains the invariant that all stores in $\mseen$ are
+  totally ordered and $\msto$ is an upper bound of the stores in $\mseen$.
 \end{lemma}
@@
 The proof follows from the order isomorphism that, in one direction,
-sorts all the stores in $S$ to form $\Sigma$, and translates stores in
-$S$ to their distance from the end of $\Sigma$ (their timestamp). In the other
-direction, timestamps in $S$ are replaced by the stores they point to in $\Sigma$.
+sorts all the stores in $\mseen$ to form $\Sigma$, and translates stores in
+$\mseen$ to their distance from the end of $\Sigma$ (their timestamp). In the other
+direction, timestamps in $\mseentime$ are replaced by the stores they point to in $\Sigma$.
 
@@
 \ttuple{\msto'}
-       {\mval{s_i} \overset{?}{=} \msto'(\maddr_i) \vee \ldots} \\
+       {\diffp(\mval{s_i}, \msto(\maddr_i)) \vee \ldots} \\
 \text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots}
+\diffp(\mval{s}, \mval{s'}) &= \mval{s'} \overset{?}{=} \mval{s} \sqcup \mval{s'}
 \end{align*}
@@
 \begin{align*}
-\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
-\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
-\text{ where }\maddr &= \alloc(\mstate) \\
-              \menv' &= \menv[\mvar\mapsto\maddr]
+\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\maddr,\mkont}}{\msto} & \dmachstep
+\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr'}{\msto(\maddr)}}{\epsilon}} \\
+\text{ where }\maddr' &= \alloc(\mstate) \\
+              \menv' &= \menv[\mvar\mapsto\maddr']
 \end{align*}
@@
 \begin{align*}
-(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
+(\mseentime, F, \msto,\Sigma,t) &\mathrel{\damachstep} (\mseentime \cup \mseentime', F', \msto',\Sigma',t') \\
 \mbox{ where }
@@
        \end{array}\right. \\
- F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin S} \\
- S' &= \setof{(c,t') \mid c \in F'} \\
+ F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin \mseentime} \\
+ \mseentime' &= \setof{(c,t') \mid c \in F'} \\
 \appendall(\varnothing) &= \epsilon \\
@@
 \begin{center}
-\includegraphics[scale=0.3]{fanout}
+\includegraphics[scale=0.2]{fanoutdot2}
 \end{center}
@@
 \begin{center}
-\includegraphics[scale=0.3]{lazy}
+\includegraphics[scale=0.2]{lazydot}
 \end{center}
@@
 new binding, because the actual store lookup is delayed one step
-(i.e. laziness is administrative). Without laziness, the reference
-will fan out to all the bindings of the variable before the new
-binding happens and thus theoretically has an observable precision
-difference. % All our benchmarks maintain their precision with lazy non-determinism, however.
+(i.e. laziness is administrative).
 
@@
 loss in precision by storing the result of the lookup in a value
-representing a delayed nondeterministic choice. This has problems with
-the next optimization, leading to unsoundness. Regaining soundness
-would involve artificial complications to the semantics.
+representing a delayed nondeterministic choice. This is a more common
+choice in 0CFA implementations we have seen, but it interferes with
+the next optimization (discussion in next section).
 
 \paragraph{Our choice: option 1}
-The configurations that lead to precision loss happen too rarely to
-warrant the significant increase in time and memory needed for this
-eager non-determinism. Without a widened store, lazy
-non-determinism is not a complete abstraction with either approach,
-but the precision gains are not immediately apparent due to fake
-rebinding (a second reference to the same variable chooses a different
-value from the first) and other problems. The performance and memory
-gains are apparent, so we favor the simplicity over invisible
-gains.}{}
+The conditions for precision loss happen too rarely to warrant the
+added complexity to the next section's semantics to avoid such
+losses. Without abstract compilation, option 2 is the better
+choice.}{}
 
@@
 \ref{fig:caam} shows the modified reduction relation. The only change
-from the previous semantics is that $\ev{}$ state construction is
+from the previous semantics is that $\evalone$ state construction is
 replaced by calling the compiled expression. For notational coherence,
@@
 \co{\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr},\mval} & \cmachstep_\msto
-\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont},\epsilon \\
+\ap[^\mcntr_\mlab]{\mvalx{u},\maddr,\mkont},\cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon} \\
 \text{ if }\mvalx{u} &\in \msto(\maddr_f), \mkont \in \msto(\mkaddr)
@@
 %% APPLY
-\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\mkont} & \cmachstep_\msto
+\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\maddr,\mkont} & \cmachstep_\msto
 \mcomp^{\mcntr'}(\msto)(\menv',\msdiff,\mkont) \\
 \text{ where }\menv' &= \menv[\mvar\mapsto\maddr] \\
-              \msdiff &= \cons{\ttuple{\maddr}{\force(\msto,\mval)}}{\epsilon}
-\\
-\ap{\mop,\mval,\mkont} & \cmachstep_\msto
-\co{\mkont,\mval'},\epsilon \\
-\text{ where } \mvalx{u} &\in \spchoice{\force(\msto,\mval)}{\force(\mval)}, \mval'\in\interpdelta(\mop,\mvalx{u})
+              \msdiff &= \cons{\ttuple{\maddr}{\msto(\maddr)}}{\epsilon}
+\\
+\ap{\mop,\maddr,\mkont} & \cmachstep_\msto
+\co{\mkont,\mvalx{u}},\epsilon \\
+\text{ where } \mval &\in \spchoice{\msto(\maddr)}{\msto(\maddr)}, \mvalx{u}\in\interpdelta(\mop,\mval)
 \end{align*}
@@
 use this observation to drive our use of a well-founded equivalence
-bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are equivalent to the notion of a stuttering
-bisimulation, but are more amenable to mechanisation (and thus
-rigorous proof). They also only require reasoning over one step of the
-reduction relation, which makes the proof concise.
-
-We define a refinement from non-compiled to compiled states by
-``committing'' all the actions of an $\ev{}$ state (defined similarly to
+bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are
+equivalent to the notion of a stuttering bisimulation, but are more
+amenable to mechanisation (and thus rigorous proof). They also only
+require reasoning over one step of the reduction relation, which makes
+the proof concise. The trick is in defining a well-founded ordering
+that determines when the two semantics will match up again, what
+Manolios calls the pair of functions $\erankt$ and $\erankl$ (but we
+don't need the generality that requires $\erankl$ since the uncompiled
+semantics doesn't stutter at all).
+
+We define a refinement, $r$, from non-compiled to compiled states (built structurally) by
+``committing'' all the actions of an $\evalone$ state (defined similarly to
 $\compile{\_}$, but immediately applies the functions), and
@@
 applying an abstract bytecode's function is equal to refining the
-corresponding $\ev{}$ state:
+corresponding $\evalone$ state:
 %
@@
 abstract compilation can change the order in which the store is
-changed. Consider the case where before we might call the same
-function from two different places in 3 and 2 steps respectively,
-compilation can make that 2 and 1, reversing the order that we bind
-values in the store, leading to a mismatch from the previous
-semantics. This kind of cross-step store poisoning is inevitable when
-a store is shared between several traces. The binding could have just
-as easily been observed the other way around in the non-compiled
-semantics. Call $\camachstep$ the result of the widening operator from
-the previous section on $\cmachstep$.
+changed (across steps). Consider the case where before we might call
+the same function from two different places in 3 and 2 steps
+respectively, compilation can make that 2 and 1, reversing the order
+that we bind values in the store, leading to a mismatch from the wide
+lazy semantics.  This kind of cross-step store poisoning is inevitable
+when a store is shared between several traces. The binding could have
+just as easily been observed the other way around in the non-compiled
+semantics. Although there is a difference pre- and post- abstract
+compilation, the result is still deterministic in contrast to Shivers'
+technique. The soundness is in tact since we can add store-widening
+the correct unwidened semantics with an easy correctness proof. Call
+$\camachstep$ the result of the widening operator from the previous
+section on $\cmachstep$.
 
@@
 the store and seen set that are more space-efficient and are amenable
-to destructive updates.
+to destructive updates by adhering to a history for each address.
 
@@
 
-The store maps to a stack of timestamped sets of abstract values.
+The store maps to a stack of timestamped sets of abstract
+values. Throughout this section, we will be taking the parameter $t$
+to be the ``current time,'' or the length of the store chain at the
+beginning of the step.
 
@@
 
-Here we formally define what lookup and update mean at a given
-timestamp.
+To allow imperative store updates, we maintain an invariant that we never look up values tagged at a time in the future:
 
@@
       \mval{s'} & \text{ if } \mvalstack = \cons{\ttuple{t'}{\mval{s}}}{\cons{\ttuple{t''}{\mval{s'}}}{\mvalstack'}}, t' > t
-    \end{array}\right. \\
+    \end{array}\right.
+\end{align*}} To construct this value stack, we have a
+time-parameterized join operation that also tracks changes to the
+store. If joining with a time in the future, we just add to
+it. Otherwise, we're making a change for the future ($t+1$), but only
+if there is an actual change.
+{\small
+\begin{align*}
 \msto \sqcup_t [\maddr \mapsto \mval{s}] &= \msto[\maddr \mapsto \mvalstack],\changep \\
- (\mvalstack,\changep) &= \msto(\maddr)\sqcup_t \mval{s} \\
-\epsilon \sqcup_t \mval{s} &= \ttuple{t}{\mval{s}} \\
-\cons{\ttuple{t'}{\mval{s}}}{\mvalstack} \sqcup_t \mval{s'} &= \cons{\ttuple{t'}{\mval{s}\sqcup\mval{s'}}}{\mvalstack},\strue \text{ if } t' > t \\
+\text{where } (\mvalstack,\changep) &= \msto(\maddr)\sqcup_t \mval{s} \\
+\epsilon \sqcup_t \mval{s} &= \ttuple{t}{\mval{s}},\strue \\
+\cons{\ttuple{t'}{\mval{s}}}{\mvalstack} \sqcup_t \mval{s'} &= \cons{\ttuple{t'}{\mval{s}\sqcup\mval{s'}}}{\mvalstack},\diffp(\mval{s}, \mval{s'}) \text{ if } t' > t \\
 \mvalstack \sqcup_t \mval{s} &= \cons{\ttuple{t+1}{\mval{s^*}}}{\mvalstack},\strue
-           \text{ if } \mval{s'} \neq \mval{s^*} \\
- \mval{s'} &= \lookup(\msto(\maddr),t) \\
- \mval{s^*} &= \mval{s} \sqcup \mval{s'} \\
+           \text{ if } \mval{s_t} \neq \mval{s^*} \\
+ \text{where } &\mval{s_t} = \lookup(\msto(\maddr),t) \\
+ &\mval{s^*} = \mval{s} \sqcup \mval{s_t} \\
 \mvalstack \sqcup_t \mval{s} &= \mvalstack,\sfalse \text{ otherwise}
@@
 {\mathcal F} &: \System \to \System \\
-{\mathcal F}(S,F, \msto,t) &= (S',F',\msto', t') \\
+{\mathcal F}(\mseentime,F, \msto,t) &= (\mseentime',F',\msto', t') \\
 \text{ where }
@@
              \end{array}\right. \\
-F' &= \setof{c \mid (c,\_) \in I, \changep \vee S(c) \neq \cons{t}{\_}} \\
-S' &= \lambda c. \left\{\begin{array}{ll}
-                               \cons{t'}{S(c)} & \text{ if } c \in F' \\
-                               S(c) & \text{ otherwise}
+F' &= \setof{c \mid (c,\_) \in I, \changep \vee \mseentime(c) \neq \cons{t}{\_}} \\
+\mseentime' &= \lambda c. \left\{\begin{array}{ll}
+                               \cons{t'}{\mseentime(c)} & \text{ if } c \in F' \\
+                               \mseentime(c) & \text{ otherwise}
                              \end{array}\right.
@@
 \begin{enumerate}
-\item{$S$: the \emph{seen} set, though made a map for faster membership tests and updates.}
+\item{$\mseentime$: the \emph{seen} set, though made a map for faster membership tests and updates.}
 \item{$F$: the \emph{frontier} set, which must be persisent or copied for the iteration through the set to be correct.}
@@
 relation, but a function that adds all next states to $F$ if they have
-not already been seen. At the end of iterating through $F$, $S$ is
-updated with the new states at the next timestamp.
+not already been seen. At the end of iterating through $F$, $\mseentime$ is
+updated with the new states at the next timestamp. There is no cross-step store poisoning since the lookup is restricted to the current step's time, which points to the same value throughout the step.
 
@@
 effect of mutation has unexpected consequence on the analyzers memory
-consumption.
+consumption, likely due to a larger-than-necessary preallocated store.
 

### 2013-06-27 paper/paper.tex
Started cleaning up sentence breaks for better diffs.

@@
 \begin{abstract}
-The technique of \emph{abstracting abstract machines} (AAM) provides a
-systematic approach for deriving computable approximations of
-evaluators that are easily proved sound.
-%
-This article contributes a complementary step-by-step process for
-subsequently going from a \naive{} analyzer derived under the AAM
-approach, to an efficient and correct implementation.  The end result
-of the process is a two to three order-of-magnitude improvement over
-the systematically derived analyzer, making it competitive with
-hand-optimized implementations that compute fundamentally less precise
-results.
+The technique of \emph{abstracting abstract machines} (AAM) provides a systematic approach for deriving computable approximations of evaluators that are easily proved sound.
+%
+This article contributes a complementary step-by-step process for subsequently going from a \naive{} analyzer derived under the AAM approach, to an efficient and correct implementation.
+%
+The end result of the process is a two to three order-of-magnitude improvement over the systematically derived analyzer, making it competitive with hand-optimized implementations that compute fundamentally less precise results.
 \end{abstract}
@@
 
-Program analysis provides sound predictive models of program behavior, but
-in order for such models to be effective, they must be efficiently
-computable and correct.  Past approaches to designing program analysis
-have often featured abstractions that are far removed from the
-original language semantics, requiring ingenuity in their construction
-and effort in their verification.
-%
-The \emph{abstracting abstract machines} (AAM)
-approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
-to deriving program analyses provides an alternative: a systematic way
-of transforming a programming language semantics in the form of an
-abstract machine into a family of abstract interpreters.  It thus reduces
-the burden of constructing and verifying the soundness of an abstract
-interpreter.
+Program analysis provides sound predictive models of program behavior, but in order for such models to be effective, they must be efficiently computable and correct.
+%
+Past approaches to designing program analysis have often featured abstractions that are far removed from the original language semantics, requiring ingenuity in their construction and effort in their verification.
+%
+The \emph{abstracting abstract machines} (AAM) approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic} to deriving program analyses provides an alternative: a systematic way of transforming a programming language semantics in the form of an abstract machine into a family of abstract interpreters.
+%
+It thus reduces the burden of constructing and verifying the soundness of an abstract interpreter.
 %
@@
 
-By taking a
-machine-oriented view of computation, AAM makes it possible to design,
-verify, and implement program analyzers for realistic language
-features typically considered difficult to model.  The approach was
-originally applied to features such as higher-order functions,
-stack inspection, exceptions, laziness, first-class continuations, and
-garbage collection.  It has since been used to verify actor-
-\cite{local:DOsualdo:12A} and
-thread-based~\cite{dvanhorn:Might2011Family} parallelism and
-behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
-has been used to model Coq~\cite{local:harvard},
-Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
-JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
-Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.
-
-The primary strength of the approach is that abstract interpreters can
-be easily derived through a small number of steps from existing
-machine models.  Since the relationships between abstract machines and
-higher-level semantic models---such as definitional
-interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational
-semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction
-semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well
-understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
-these high-level semantic models to sound program analyzers in a
-systematic way.  Moreover, since these analyses so closely resemble a
-language's interpreter (a) implementing an analysis requires little
-more than implementing an interpreter, (b) a single implementation can
-serve as both an interpreter and analyzer, and (c) verifying the
-correctness of the implementation is straightforward.
-
-Unfortunately, the AAM approach yields analyzers with poor performance
-relative to hand-optimized analyzers.
-%
-Our work takes aim squarely at this ``efficiency gap,'' and narrows it
-in an equally systematic way through a number of simple steps, many of
-which are inspired by run-time implementation techniques such as
-laziness and compilation to avoid interpretative overhead.  Each of
-these steps is proven correct, so the end result is an implementation
-that is trustworthy and efficient.
-
-In this article, we develop a systematic approach to deriving a
-practical implementation of an abstract-machine-based analyzer using
-mostly semantic means rather than tricky engineering. Our goal is to
-empower programming language implementers and researchers to explore
-and convincingly exhibit their ideas with a low barrier to entry. The
-optimizations we describe are widely applicable and apparently
-effective to scale far beyond the size of programs typically considered
-in the recent literature on flow analysis for functional languages.
+By taking a machine-oriented view of computation, AAM makes it possible to design, verify, and implement program analyzers for realistic language features typically considered difficult to model.
+%
+The approach was originally applied to features such as higher-order functions, stack inspection, exceptions, laziness, first-class continuations, and garbage collection.
+%
+It has since been used to verify actor-\cite{local:DOsualdo:12A} and thread-based~\cite{dvanhorn:Might2011Family} parallelism and behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder};
+%
+it has been used to model Coq~\cite{local:harvard}, Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B}, JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.
+%%
+
+%%
+The primary strength of the approach is that abstract interpreters can be easily derived through a small number of steps from existing machine models.
+%
+Since the relationships between abstract machines and higher-level semantic models---such as definitional interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from these high-level semantic models to sound program analyzers in a systematic way.
+%
+Moreover, since these analyses so closely resemble a language's interpreter
+%
+(a) implementing an analysis requires little more than implementing an interpreter,
+%
+(b) a single implementation can serve as both an interpreter and analyzer, and
+%
+(c) verifying the correctness of the implementation is straightforward.
+%%
+
+%%
+Unfortunately, the AAM approach yields analyzers with poor performance relative to hand-optimized analyzers.
+%
+Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an equally systematic way through a number of simple steps, many of which are inspired by run-time implementation techniques such as laziness and compilation to avoid interpretative overhead.
+%
+Each of these steps is proven correct, so the end result is an implementation that is trustworthy and efficient.
+%%
+
+%%
+In this article, we develop a systematic approach to deriving a practical implementation of an abstract-machine-based analyzer using mostly semantic means rather than tricky engineering.
+%
+Our goal is to empower programming language implementers and researchers to explore and convincingly exhibit their ideas with a low barrier to entry.
+%
+The optimizations we describe are widely applicable and apparently effective to scale far beyond the size of programs typically considered in the recent literature on flow analysis for functional languages.
 
@@
 \end{center}
-\caption{Factor improvements over the baseline analyzer for the
-  \Church{} benchmark in terms of the rate of state
-  transitions and total analysis time. (Bigger is better.) Each point
-  is marked with the section that introduces the optimization.}
+\caption{
+Factor improvements over the baseline analyzer for the \Church{} benchmark in terms of the rate of state transitions and total analysis time.
+%
+(Bigger is better.)
+%
+Each point is marked with the section that introduces the optimization.}
 \label{fig:churchtime}
@@
 
-We start with a quick review of the AAM approach to develop an
-analysis framework and then apply our step-by-step optimization
-techniques in the simplified setting of a core functional language.
-This allows us to explicate the optimizations with a minimal amount of
-inessential technical overhead.  Following that, we scale this
-approach up to an analyzer for a realistic untyped, higher-order
-imperative language with a number of interesting features and then
-measure improvements across a suite of benchmarks.
-
-At each step during the initial presentation and development, we
-evaluated the implementation on a set of benchmarks. The highlighted
-benchmark in figure \ref{fig:churchtime} is from Vardoulakis and
-Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
-of multiplication over addition on Church numerals.  For the
-step-by-step development, this benchmark is particularly informative:
+We start with a quick review of the AAM approach to develop an analysis framework and then apply our step-by-step optimization techniques in the simplified setting of a core functional language.
+%
+This allows us to explicate the optimizations with a minimal amount of inessential technical overhead.
+%
+Following that, we scale this approach up to an analyzer for a realistic untyped, higher-order imperative language with a number of interesting features and then measure improvements across a suite of benchmarks.
+%%
+
+%%
+At each step during the initial presentation and development, we evaluated the implementation on a set of benchmarks.
+%
+The highlighted benchmark in figure \ref{fig:churchtime} is from Vardoulakis and Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity of multiplication over addition on Church numerals.
+%
+For the step-by-step development, this benchmark is particularly informative:
 \begin{enumerate}
@@
 %
-\item it was designed to stress an analyzer's ability to deal with
-  complicated environment and control structure arising from the use
-  of higher-order functions to encode arithmetic, and
-%
-\item its improvement is about median in the benchmark suite
-  considered in section~\ref{sec:eval}, and thus it serves as a good
-  sanity check for each of the optimization techniques considered.
+\item it was designed to stress an analyzer's ability to deal with complicated environment and control structure arising from the use of higher-order functions to encode arithmetic, and
+%
+\item its improvement is about median in the benchmark suite considered in section~\ref{sec:eval}, and thus it serves as a good sanity check for each of the optimization techniques considered.
 \end{enumerate}
 
-We start, in section~\ref{sec:aam}, by developing an abstract
-interpreter according to the AAM approach.  In the initial
-abstraction, each state carries a store (what is called per-state
-store variance). The space of stores is exponential in size; without
-further abstraction, the analysis is exponential and thus cannot
-analyze the example in a reasonable amount of time.  In
-section~\ref{sec:baseline}, we perform a further abstraction by
-widening the store.  The resulting analyzer sacrifices precision for
-speed and is able to analyze the example in about 1 minute.  This step
-is described by Van Horn and Might~\cite[\S
-3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make even
-small examples feasible.  We therefore take a widened interpreter as
-the baseline for our evaluation.
-
-Section~\ref{sec:opt} gives a series of simple abstractions and
-implementation techniques that, in total, speed up the analysis by
-nearly a factor of 500, dropping the analysis time to a fraction of a
-second.  Figure~\ref{fig:churchtime} shows the step-wise improvement
-of the analysis time for this example.
-
-The AAM approach, in essence, does the following: it takes a
-machine-based view of computation and turns it into a \emph{finitary
-  approximation} by bounding the size of the store.  With a limited
-address space, the store must map addresses to \emph{sets} of values.
-Store updates are interpreted as joins, and store dereferences are
-interpreted by non-deterministic choice of an element from a set.  The
-result of analyzing a program is a finite directed graph where nodes
-in the graph are (abstract) machine states and edges denote machine
-transitions between states.
-
-The techniques we propose for optimizing analysis fall into the
-following categories:
+We start, in section~\ref{sec:aam}, by developing an abstract interpreter according to the AAM approach.
+%
+In the initial abstraction, each state carries a store (what is called per-state store variance).
+%
+The space of stores is exponential in size; without further abstraction, the analysis is exponential and thus cannot analyze the example in a reasonable amount of time.
+%
+In section~\ref{sec:baseline}, we perform a further abstraction by widening the store.
+%
+The resulting analyzer sacrifices precision for speed and is able to analyze the example in about 1 minute.
+%
+This step is described by Van Horn and Might~\cite[\S 3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make even small examples feasible.
+%
+We therefore take a widened interpreter as the baseline for our evaluation.
+%%
+
+%%
+Section~\ref{sec:opt} gives a series of simple abstractions and implementation techniques that, in total, speed up the analysis by nearly a factor of 500, dropping the analysis time to a fraction of a second.
+%
+Figure~\ref{fig:churchtime} shows the step-wise improvement of the analysis time for this example.
+%%
+
+%%
+The AAM approach, in essence, does the following: it takes a machine-based view of computation and turns it into a \emph{finitary approximation} by bounding the size of the store.
+%
+With a limited address space, the store must map addresses to \emph{sets} of values.
+%
+Store updates are interpreted as joins, and store dereferences are interpreted by non-deterministic choice of an element from a set.
+%
+The result of analyzing a program is a finite directed graph where nodes in the graph are (abstract) machine states and edges denote machine transitions between states.
+%%
+
+%%
+The techniques we propose for optimizing analysis fall into the following categories:
 \begin{enumerate}
-\item generate fewer states by avoiding the eager exploration of
-  non-deterministic choices that will later collapse into a single
-  join point.  We accomplish this by applying lazy evaluation
-  techniques so that non-determinism is evaluated \emph{by need}.
-
-\item generate fewer states by avoiding unnecessary, intermediate
-  states of a computation.  We accomplish this by applying compilation
-  techniques from functional languages to avoid interpretive overhead
-  in the machine transition system.
-
-\item generate states faster.  We accomplish this by better algorithm
-  design in the fixed-point computation we use to generate state graphs.
+\item generate fewer states by avoiding the eager exploration of non-deterministic choices that will later collapse into a single join point.
+%
+  We accomplish this by applying lazy evaluation techniques so that non-determinism is evaluated \emph{by need}.
+
+\item generate fewer states by avoiding unnecessary, intermediate states of a computation.
+%
+  We accomplish this by applying compilation techniques from functional languages to avoid interpretive overhead in the machine transition system.
+
+\item generate states faster.
+%
+  We accomplish this by better algorithm design in the fixed-point computation we use to generate state graphs.
 \end{enumerate}
-Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for the
-small motivating example in Earl, et
-al.~\cite{dvanhorn:Earl2012Introspective}.  By generating
-significantly fewer states at a significantly faster rate, we are able
-to achieve large performance improvements in terms of both time and
-space.
-
-Section~\ref{sec:eval} describes the evaluation of each optimization
-technique applied to an implementation supporting a more realistic set
-of features, including mutation, first-class control, compound data, a
-full numeric tower and many more forms of primitive data and
-operations.
-%
-We evaluate this implementation against a set of benchmark programs
-drawn from the literature.
-%
-For all benchmarks, the optimized analyzer outperforms the baseline
-by at least a factor of
+%
+Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for the small motivating example in Earl, et al.~\cite{dvanhorn:Earl2012Introspective}.
+%
+By generating significantly fewer states at a significantly faster rate, we are able to achieve large performance improvements in terms of both time and space.
+%%
+
+%%
+Section~\ref{sec:eval} describes the evaluation of each optimization technique applied to an implementation supporting a more realistic set of features, including mutation, first-class control, compound data, a full numeric tower and many more forms of primitive data and operations.
+%
+We evaluate this implementation against a set of benchmark programs drawn from the literature.
+%
+For all benchmarks, the optimized analyzer outperforms the baseline by at least a factor of
 % 475
@@
 
-Section~\ref{sec:related} relates this work to the literature and
-section~\ref{sec:conclusion} concludes.
+Section~\ref{sec:related} relates this work to the literature and section~\ref{sec:conclusion} concludes.
 
@@
 \end{center}
-\caption{Example state graphs for Earl et. al. program.  Gray states
-  follow variable references, {\tt ev} states are black, and all others
-  are white. Part (a) shows the result of the baseline analyzer.  It
-  has long ``corridor'' transitions and ``diamond'' subgraphs that
-  fan-out from nondeterminism and fan-in from joins.  Part (b) shows
-  the result of performing nondeterminism lazily and thus avoids many
-  of the diamond subgraphs.  Part (c) shows the result of abstract
-  compilation that removes interpretive overhead in the form of
-  intermediate states, thus minimizing the corridor transitions.  The
-  end result is a more compact abstraction of the program that can be
-  generated faster.}
+\caption{
+Example state graphs for Earl et. al. program.
+%
+  Gray states follow variable references, {\tt ev} states are black, and all others are white.
+%
+ Part (a) shows the result of the baseline analyzer.
+%
+  It has long ``corridor'' transitions and ``diamond'' subgraphs that fan-out from nondeterminism and fan-in from joins.
+%
+  Part (b) shows the result of performing nondeterminism lazily and thus avoids many of the diamond subgraphs.
+%
+  Part (c) shows the result of abstract compilation that removes interpretive overhead in the form of intermediate states, thus minimizing the corridor transitions.
+%
+  The end result is a more compact abstraction of the program that can be generated faster.}
 \label{fig:state-graphs}
@@
 defining a sound analytic framework for a core higher-order functional
-language: Landin's ISWIM~\cite{dvanhorn:Landin1966Next}.  In the
-subsequent sections, we will explore optimizations for the analyzer in
-this simplified setting, but scaling these techniques to realistic
-languages is straightforward and has been done for the analyzer
-evaluated in section~\ref{sec:eval}.
-
-ISWIM is a family of programming languages parameterized by a set of
-base values and operations.  To make things concrete, we consider a
-member of the ISWIM family with integers, booleans, and a few
-operations.
-%
-Figure~\ref{fig:syntax} defines the syntax of ISWIM.  It
-includes variables, literals (either integers, booleans, or
-operations), $\lambda$-expressions for defining procedures, procedure
-applications, and conditionals.  Expressions carry a label, $\mlab$,
-which is drawn from an unspecified set and denotes the source location
-of the expression; labels are used to disambiguate distinct, but
-syntactically identical pieces of syntax.  We omit the label
-annotation in contexts where it is irrelevant.
+language: Landin's ISWIM~\cite{dvanhorn:Landin1966Next}.
+%
+In the subsequent sections, we will explore optimizations for the analyzer in this simplified setting, but scaling these techniques to realistic languages is straightforward and has been done for the analyzer evaluated in section~\ref{sec:eval}.
+%%
+
+%%
+ISWIM is a family of programming languages parameterized by a set of base values and operations.
+%
+To make things concrete, we consider a member of the ISWIM family with integers, booleans, and a few operations.
+%
+Figure~\ref{fig:syntax} defines the syntax of ISWIM.
+%
+It includes variables, literals (either integers, booleans, or operations), $\lambda$-expressions for defining procedures, procedure applications, and conditionals.
+%
+Expressions carry a label, $\mlab$, which is drawn from an unspecified set and denotes the source location of the expression; labels are used to disambiguate distinct, but syntactically identical pieces of syntax.
+%
+We omit the label annotation in contexts where it is irrelevant.
 
@@
 
-The semantics are defined in terms of a machine model.  The machine
-components are defined in figure~\ref{fig:domains};
-figure~\ref{fig:aam} defines the transition relation (unmentioned
-components stay the same).  The evaluation of a program is defined as
-its set of traces that arise from iterating the machine transition
-relation. The $\traces$ function produces the set of all proofs of
-reachability for any state $\mstate$ from the injection of program
-$\mexp$ (from which one could extract a string of states). The machine
-is a very slight variation on a standard abstract machine for ISWIM in
-``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It can be
-systematically derived from a definitional interpreter through a
-continuation-passing style transformation and defunctionalization, or
-from a structural operational semantics using the refocusing
-construction of Danvy and
-Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.
-
+The semantics are defined in terms of a machine model.
+%
+The machine components are defined in figure~\ref{fig:domains};
+%
+figure~\ref{fig:aam} defines the transition relation (unmentioned components stay the same).
+%
+The evaluation of a program is defined as its set of traces that arise from iterating the machine transition relation.
+%
+The $\traces$ function produces the set of all proofs of reachability for any state $\mstate$ from the injection of program $\mexp$ (from which one could extract a string of states).
+%
+ The machine is a very slight variation on a standard abstract machine for ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.
+%
+It can be systematically derived from a definitional interpreter through a continuation-passing style transformation and defunctionalization, or from a structural operational semantics using the refocusing construction of Danvy and Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.
+%%
+
+%%
 Compared with the standard machine semantics, this definition is
@@
 \begin{itemize}
-\item the store maps addresses to \emph{sets} of values,\footnote{
-More generally, we can have stores map to any domain that forms
- a Galois connection with sets of values, enabling $\interpdelta$ to
-produce elaborate abstractions of base values (e.g., interval or octagon
-abstractions). We use sets of values for a simpler exposition.
-} not
-  single values,
+\item the store maps addresses to \emph{sets} of values,
+%
+\footnote{
+More generally, we can have stores map to any domain that forms a Galois connection with sets of values, enabling $\interpdelta$ to produce elaborate abstractions of base values (e.g., interval or octagon abstractions).
+%
+We use sets of values for a simpler exposition.
+}
+%
+not single values,
 
 \item continuations are heap-allocated, not stack-allocated,
-\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
-  labels ($\mlab$) threaded through the computation, and
-\item the machine is implicitly parameterized by the functions
-  $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces
-  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
+%
+\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax labels ($\mlab$) threaded through the computation, and
+%
+\item the machine is implicitly parameterized by the functions $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
 \end{itemize}
@@
 These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
-used to determinize the choice of fresh addresses. The $\sqcup$ on
-stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
-\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
-relation is non-deterministic in its choice of addresses, however it
-must always choose a fresh address when allocating a continuation or
-variable binding.  If we consider machine states equivalent up to
-consistent renaming and fix an allocation scheme, this relation
-defines a deterministic machine (the relation is really a function).
-
-The interpretation of primitive operations is defined by setting
-$\interpdelta$ as follows:
+used to determinize the choice of fresh addresses.
+%
+The $\sqcup$ on stores in the figure is a point-wise lifting of $\cup$:
+%
+ $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$.
+%
+The resulting relation is non-deterministic in its choice of addresses, however it must always choose a fresh address when allocating a continuation or variable binding.
+%
+If we consider machine states equivalent up to consistent renaming and fix an allocation scheme, this relation defines a deterministic machine (the relation is really a function).
+%%
+
+%%
+The interpretation of primitive operations is defined by setting $\interpdelta$ as follows:
 \begin{align*}
@@
 equivalent to the notion of a stuttering bisimulation, but are more
-amenable to mechanisation (and thus rigorous proof). They also only
+amenable to mechanization (and thus rigorous proof). They also only
 require reasoning over one step of the reduction relation, which makes
@@
 \end{lemma}
-This depends on some wellformedness conditions about the order of the
+This depends on some well-formedness conditions about the order of the
 stacks. The store of value stacks can be translated to a stack of
@@
 
-The intermediate data structures of the above transfer function can all be streamlined into globals that are destructively updated. In particular, there are 5 globals:
+The intermediate data structures of the above transfer function can all be streamlined into globals that are destructively updated.
+%
+In particular, there are 5 globals:
 
@@
 
-The reduction relation would then instead of building store deltas,
-update the global store. We would also not view it as a general
-relation, but a function that adds all next states to $F$ if they have
-not already been seen. At the end of iterating through $F$, $\mseentime$ is
-updated with the new states at the next timestamp. There is no cross-step store poisoning since the lookup is restricted to the current step's time, which points to the same value throughout the step.
+The reduction relation would then instead of building store deltas, update the global store.
+%
+We would also not view it as a general relation, but a function that adds all next states to $F$ if they have
+not already been seen.
+%
+At the end of iterating through $F$, $\mseentime$ is updated with the new states at the next timestamp.
+%
+There is no cross-step store poisoning since the lookup is restricted to the current step's time, which points to the same value throughout the step.
 
@@
 %
-But, such a dynamic structure isn't necessary when we know the
-structure of the store in advance: we know all possible entries, and
-we know its maximum size.
-%
-In a monovariant allocation strategy, the domain of the store is
-exactly the set of expressions in the program.
-%
-If we label each expression with a unique natural, the analysis can
-index directly into the store without a hash or a collision.
-%
-Even for polyvariant analyses, it is possible to compute the maximum
-number of addresses and similarly pre-allocate either the spine of the
-store or (if memory is no concern) the entire store.
+But, such a dynamic structure isn't necessary when we know the structure of the store in advance: we know all possible entries, and we know its maximum size.
+%
+In a monovariant allocation strategy, the domain of the store is exactly the set of expressions in the program.
+%
+If we label each expression with a unique natural, the analysis can index directly into the store without a hash or a collision.
+%
+Even for polyvariant analyses, it is possible to compute the maximum number of addresses and similarly pre-allocate either the spine of the store or (if memory is no concern) the entire store.
 
@@
 
-We have implemented, optimized, and evaluated an analysis framework
-supporting higher-order functions, state, first-class control,
-compound data, and a large number of primitive kinds of data and
-operations such as floating point, complex, and exact rational
-arithmetic.  The analysis is evaluated against a suite of Scheme benchmarks
-drawn from the literature.
-%
-For each benchmark, we collect analysis times, peak memory usage (as
-determined by Racket's GC statistics), and the rate of
-states-per-second explored by the analysis for each of the
-optimizations discussed in section~\ref{sec:opt}, cumulatively
-applied.  The analysis is stopped after consuming 30 minutes of time
-or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
-use the timeout limits as a lower bound on the actual time required (i.e., one minute versus timeout is at least 30 times faster),
+We have implemented, optimized, and evaluated an analysis framework supporting higher-order functions, state, first-class control, compound data, and a large number of primitive kinds of data and operations such as floating point, complex, and exact rational arithmetic.
+%
+The analysis is evaluated against a suite of Scheme benchmarks drawn from the literature.
+%
+For each benchmark, we collect analysis times, peak memory usage (as determined by Racket's GC statistics), and the rate of states-per-second explored by the analysis for each of the optimizations discussed in section~\ref{sec:opt}, cumulatively applied.
+%
+The analysis is stopped after consuming 30 minutes of time or 1 gigabyte of space.
+%
+When presenting \emph{relative} numbers, we use the timeout limits as a lower bound on the actual time required (i.e., one minute versus timeout is at least 30 times faster),
 thus giving a conservative estimate of improvements.
-
+%%
+
+%%
 All benchmarks are calculated as an average of 5 runs, done in
@@
 with 12Gb of memory.
-
-Many benchmarks cause the baseline analyzer to take longer than 30
-minutes or to consume more than 1 gigabyte of memory, at which point the
-analysis is stopped.  This is the case for the largest benchmark
-program, {\bf nucleic}, which is 3,500 lines of code and takes under a minute in the
-most optimized analyzer.  For those benchmarks that did complete on
-the baseline, the optimized analyzer outperformed the baseline by a
-factor of two to three orders of magnitude.
+%%
+
+%%
+Many benchmarks cause the baseline analyzer to take longer than 30 minutes or to consume more than 1 gigabyte of memory, at which point the analysis is stopped.
+%
+This is the case for the largest benchmark program, {\bf nucleic}, which is 3,500 lines of code and takes under a minute in the most optimized analyzer.
+%
+For those benchmarks that did complete on the baseline, the optimized analyzer outperformed the baseline by a factor of two to three orders of magnitude.
 % safer to not give exact factors
@@
 
-\item {\bf nucleic}: a floating-point intensive application taken from
-  molecular biology that has been used widely in benchmarking
-  functional language
-  implementations~\cite{dvanhorn:Hartel1996Benchmarking} and analyses
-  (e.g.~\cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
-  It is a constraint satisfaction algorithm used to determine the
-  three-dimensional structure of nucleic acids.
-
-\item {\bf matrix} tests whether a matrix is maximal among all
-  matrices of the same dimension obtainable by simple reordering of
-  rows and columns and negation of any subset of rows and columns.  It
-  is written in continuation-passing style (used in
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
-
-\item {\bf nbody}: implementation~\cite{ianjohnson:nbody87} of the
-  Greengard multipole algorithm for computing gravitational forces on
-  point masses distributed uniformly in a cube (used in
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
-
-\item {\bf earley}: Earley's parsing algorithm, applied to a 15-symbol
-  input according to a simple ambiguous grammar.  A real program,
-  applied to small data whose exponential behavior leads to a peak
-  heap size of half a gigabyte or more during concrete execution.
-
-\item {\bf maze}: generates a random maze using Scheme's {\tt
-  call/cc} operation and finds a path solving
-  the maze (used in
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
-
-\item {\bf church}: tests distributivity of multiplication over
-  addition for Church numerals (introduced by
-  \cite{dvanhorn:Vardoulakis2011CFA2}).
-
-\item {\bf lattice}: enumerates the order-preserving maps between two
-  finite lattices (used in
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
-
-\item {\bf boyer}: a term-rewriting theorem prover (used in
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
-
-\item {\bf mbrotZ}: generates Mandelbrot fractal using complex
-  numbers.
-
-\item {\bf graphs}: counts the number of directed graphs with a
-  distinguished root and \(k\) vertices, each having out-degree at
-  most 2. It is written in a continuation-passing style and makes
-  extensive use of higher-order procedures---it creates closures
-  almost as often as it performs non-tail procedure calls (used by
-  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+\item {\bf nucleic}: a floating-point intensive application taken from molecular biology that has been used widely in benchmarking functional language implementations~\cite{dvanhorn:Hartel1996Benchmarking} and analyses (e.g.~\cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+%
+  It is a constraint satisfaction algorithm used to determine the three-dimensional structure of nucleic acids.
+
+\item {\bf matrix} tests whether a matrix is maximal among all matrices of the same dimension obtainable by simple reordering of rows and columns and negation of any subset of rows and columns.
+%
+  It is written in continuation-passing style (used in \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+
+\item {\bf nbody}: implementation~\cite{ianjohnson:nbody87} of the Greengard multipole algorithm for computing gravitational forces on point masses distributed uniformly in a cube (used in \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+%%
+
+%%
+\item {\bf earley}: Earley's parsing algorithm, applied to a 15-symbol input according to a simple ambiguous grammar.
+%
+  A real program, applied to small data whose exponential behavior leads to a peak heap size of half a gigabyte or more during concrete execution.
+
+\item {\bf maze}: generates a random maze using Scheme's {\tt call/cc} operation and finds a path solving the maze (used in \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+
+\item {\bf church}: tests distributivity of multiplication over addition for Church numerals (introduced by \cite{dvanhorn:Vardoulakis2011CFA2}).
+
+\item {\bf lattice}: enumerates the order-preserving maps between two finite lattices (used in \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+
+\item {\bf boyer}: a term-rewriting theorem prover (used in \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
+
+\item {\bf mbrotZ}: generates Mandelbrot fractal using complex numbers.
+
+\item {\bf graphs}: counts the number of directed graphs with a distinguished root and \(k\) vertices, each having out-degree at most 2.
+%
+ It is written in a continuation-passing style and makes extensive use of higher-order procedures---it creates closures almost as often as it performs non-tail procedure calls (used by \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
 \end{enumerate}
@@
 
-Figure~\ref{fig:bench-overview} gives an overview of the benchmark
-results in terms of absolute time, space, and speed between the
-baseline and most optimized analyzer.  Figure~\ref{fig:bench-all}
-plots the factors of improvement over the baseline for each
-optimization step.
+Figure~\ref{fig:bench-overview} gives an overview of the benchmark results in terms of absolute time, space, and speed between the baseline and most optimized analyzer.
+%
+Figure~\ref{fig:bench-all} plots the factors of improvement over the baseline for each optimization step.
 
@@
 \end{center}
-\caption{Factors of improvement over baseline for each step of
-  optimization (bigger is better).}
+\caption{Factors of improvement over baseline for each step of optimization (bigger is better).}
 \label{fig:bench-all}
@@
 
-To determine the impact of each section's technique on precision, we evaluated
-a singleton variable analysis to find opportunities to inline
-constants and closed functions.  We found no change in the results
-across all implementations, including Shivers' timestamp
-approximation---from an empirical point of view, these techniques are
-precision preserving despite the theoterical loss of precision.
-
-Our step-wise optimizations strictly produce better analysis times
-with no observed loss of precision.  The final result is a
-systematically derived and verified implementation that operates
-within a small factor performance loss compared to a hand-optimized,
-unverified implementation.  Moreover, much of the performance gains
-are achieved with purely functional methods, which suggest potential
-further gains from parellelization.  Peak memory usage is often
-considerably improved by the end of the optimization steps, but the
-effect of mutation has unexpected consequence on the analyzers memory
-consumption, likely due to a larger-than-necessary preallocated store.
-
+To determine the impact of each section's technique on precision, we evaluated a singleton variable analysis to find opportunities to inline constants and closed functions.
+%
+We found no change in the results across all implementations, including Shivers' timestamp approximation---from an empirical point of view, these techniques are precision preserving despite the theoterical loss of precision.
+%%
+
+%%
+Our step-wise optimizations strictly produce better analysis times with no observed loss of precision.
+%
+The final result is a systematically derived and verified implementation that operates within a small factor performance loss compared to a hand-optimized, unverified implementation.
+%
+Moreover, much of the performance gains are achieved with purely functional methods, which suggest potential further gains from parellelization.
+%
+Peak memory usage is often considerably improved by the end of the optimization steps, but the effect of mutation has unexpected consequence on the analyzers memory consumption, likely due to a larger-than-necessary preallocated store.
+%%
+
+%%
 Source code of the implementation and benchmark suite is at:
@@
 
-The analysis considered here computes results similar to Earl, et al.'s
-0-CFA implementation~\cite{dvanhorn:Earl2012Introspective}, which
-times out on the \Church{} benchmark because it does not widen the
-store as described for our baseline evaluator.  So even though it
-offers a fair point of comparison, a more thorough evaluation is
-probably uninformative as the other benchmarks are likely to timeout
-as well (and it would require significant effort to extend their
-implementation with the features needed to analyze our benchmark
-suite).  That implementation is evaluated against much smaller
-benchmarks: the largest program is 30 lines.
-
-Vardoulakis and Shivers evaluate their CFA2
-analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of
-0-CFA defined in their framework and the example we draw on is the
-largest benchmark Vardoulakis and Shivers consider.  More work would
-be required to scale the analyzer to the set of features required by
-our benchmarks.
-
+The analysis considered here computes results similar to Earl, et al.'s 0-CFA implementation~\cite{dvanhorn:Earl2012Introspective}, which times out on the \Church{} benchmark because it does not widen the store as described for our baseline evaluator.
+%
+So even though it offers a fair point of comparison, a more thorough evaluation is probably uninformative as the other benchmarks are likely to timeout as well (and it would require significant effort to extend their implementation with the features needed to analyze our benchmark suite).
+%
+That implementation is evaluated against much smaller benchmarks: the largest program is 30 lines.
+%%
+
+%%
+Vardoulakis and Shivers evaluate their CFA2 analyzer~\cite{dvanhorn:Vardoulakis2011CFA2} against a variant of 0-CFA defined in their framework and the example we draw on is the largest benchmark Vardoulakis and Shivers consider.
+%
+More work would be required to scale the analyzer to the set of features required by our benchmarks.
+%%
+
+%%
 The only analyzers we were able to find that proved capable of
@@
 successor, the Polymorphic splitting system of Wright and
-Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
-  is not a coincidence; these papers set a high standard for
-  evaluation, which we consciously aimed to approach.}  Unfortunately,
-these analyses compute an inherently different and incomparable form
-of analysis.  Consequently, we have omitted a complete comparison with
-these implementations.  The AAM approach provides more precision in
-terms of temporal-ordering of program states, which comes at a cost
-that can be avoided in constraint-based approaches.  Consequently
-implementation techniques cannot be ``ported'' between these two
-approaches.  However, our optimized implementation is within an order
-of magnitude of the performance of Wright and Jaganathan's analyzer.
-Although we would like to improve this to be more competitive, the
-optimized AAM approach still has many strengths to recommend it in
-terms of precision, ease of implementation and verification, and rapid
-design. We can get closer to their performance by relying on the
-representation of addresses and the behavior of $\alloc$ to
-pre-allocate most data structures and split the abstract store out
-into parts that are more quickly accessed and updated. Our semantic
-optimizations can still be applied to an analysis that does abstract
-garbage collection~\cite{dvanhorn:Might:2006:GammaCFA}, whereas the
-polymorphic splitting implementation is tied strongly to a
-single-threaded store.
+Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.
+%
+\footnote{This is not a coincidence; these papers set a high standard for evaluation, which we consciously aimed to approach.}
+%
+Unfortunately, these analyses compute an inherently different and incomparable form of analysis.
+%
+Consequently, we have omitted a complete comparison with these implementations.
+%
+The AAM approach provides more precision in terms of temporal-ordering of program states, which comes at a cost that can be avoided in constraint-based approaches.
+%
+Consequently implementation techniques cannot be ``ported'' between these two approaches.
+%
+  However, our optimized implementation is within an order of magnitude of the performance of Wright and Jaganathan's analyzer.
+%
+Although we would like to improve this to be more competitive, the optimized AAM approach still has many strengths to recommend it in terms of precision, ease of implementation and verification, and rapid design.
+%
+ We can get closer to their performance by relying on the representation of addresses and the behavior of $\alloc$ to pre-allocate most data structures and split the abstract store out into parts that are more quickly accessed and updated.
+%
+Our semantic optimizations can still be applied to an analysis that does abstract garbage collection~\cite{dvanhorn:Might:2006:GammaCFA}, whereas the polymorphic splitting implementation is tied strongly to a single-threaded store.
 
@@
 
-This work clearly closely follows Van Horn and Might's original papers
-on abstracting abstract
-machines~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic},
-which in turn is one piece of the large body of research on flow
-analysis for higher-order languages (see
-Midtgaard~\cite{dvanhorn:Midtgaard2011Controlflow} for a thorough
-survey).  The AAM approach sits at the confluence of two major lines
-of research: (1) the study of abstract
-machines~\cite{dvanhorn:landin-64} and their systematic
-construction~\cite{dvanhorn:reynolds-hosc98}, and (2) the theory of
-abstract interpretation
-\cite{dvanhorn:Cousot:1977:AI,dvanhorn:Cousot1979Systematic}.
+This work clearly closely follows Van Horn and Might's original papers on abstracting abstract machines~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}, which in turn is one piece of the large body of research on flow analysis for higher-order languages (see Midtgaard~\cite{dvanhorn:Midtgaard2011Controlflow} for a thorough survey).
+%
+The AAM approach sits at the confluence of two major lines of research:
+%
+(1) the study of abstract machines~\cite{dvanhorn:landin-64} and their systematic construction~\cite{dvanhorn:reynolds-hosc98},
+%
+and (2) the theory of abstract interpretation \cite{dvanhorn:Cousot:1977:AI,dvanhorn:Cousot1979Systematic}.
 
@@
 
-Besides the original AAM work, the analysis most similar to that
-presented in section~\ref{sec:aam} is the infinitary control-flow
-analysis of Nielson and Nielson~\cite{dvanhorn:nielson-nielson-popl97}
-and the unified treatment of flow analysis by Jagannathan and
-Weeks~\cite{dvanhorn:jagannathan-weeks-popl95}.  Both are
-parameterized in such a way that in the limit, the analysis is
-equivalent to an interpreter for the language, just as is the case
-here.  What is different is that both give a constraint-based
-formulation of the abstract semantics rather than a finite machine
-model.
+Besides the original AAM work, the analysis most similar to that presented in section~\ref{sec:aam} is the infinitary control-flow analysis of Nielson and Nielson~\cite{dvanhorn:nielson-nielson-popl97} and the unified treatment of flow analysis by Jagannathan and Weeks~\cite{dvanhorn:jagannathan-weeks-popl95}.
+%
+Both are parameterized in such a way that in the limit, the analysis is equivalent to an interpreter for the language, just as is the case here.
+%
+What is different is that both give a constraint-based formulation of the abstract semantics rather than a finite machine model.
 
@@
 
-Boucher and Feeley \cite{dvanhorn:Boucher1996Abstract} introduced the
-idea of abstract compilation, which used closure generation
-\cite{dvanhorn:Feeley1987Using} to improve the performance of control
-flow analysis.  We have adapted the closure generation technique from
-compositional evaluators to abstract machines and applied it to similar
-effect.
+Boucher and Feeley \cite{dvanhorn:Boucher1996Abstract} introduced the idea of abstract compilation, which used closure generation \cite{dvanhorn:Feeley1987Using} to improve the performance of control flow analysis.
+%
+We have adapted the closure generation technique from compositional evaluators to abstract machines and applied it to similar effect.
 
@@
 
-Constraint-based program analyses
-(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular,dvanhorn:steckler-wand-toplas97})
-typically compute sets of abstract values for each program point.
-These values approximate values arising at run-time for each
-program point.  Value sets are computed as the least solution to a set
-of (inclusion or equality) constraints.  The
-constraints must be designed and proved as a sound approximation of
-the semantics.  Efficient implementations of these kinds of analyses
-often take the form of worklist-based graph algorithms for constraint
-solving, and are thus quite different from the interpreter
-implementation.  The approach thus requires effort in constraint
-system design and implementation, and the resulting system require
-verification effort to prove the constraint system is sound and that
-the implementation is correct.
-
-This effort increases substantially as the complexity of the analyzed
-language increases.  Both the work of maintaining the concrete
-semantics and constraint system (and the relations between them) must
-be scaled simultaneously.  However, constraint systems, which have
-been extensively studied in their own right, enjoy efficient
-implementation techniques and can be expressed in declarative logic
-languages that are heavily
-optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
-Consequently, constraint-based analyses can be computed quickly.  For
-example, Jagannathan and Wright's polymorphic splitting
-implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
-the \Church{} benchmark about 5.5 times faster than the fastest
-implementation considered here.  These analyses compute very different
-things, so the performance comparison is not apples-to-apples.
-
-The AAM approach, and the state transition graphs it generates, encodes
-temporal properties not found in classical constraint-based analyses
-for higher-order programs.
-%
-Such analyses (ultimately) compute judgments on program terms and
-contexts, e.g., at expression $e$, variable $x$ may have value $v$.
-%
-The judgments do not relate the order in which expressions and context
-may be evaluated in a program, e.g., it has nothing
-to say with regard to question like, ``Do we always evaluate $e_1$
-before $e_2$?'' or ``Is it always the case that a file handle is
-opened, read and then closed in that order?''
-%
-The state transition graphs can answer these kinds of queries, but
-this does not come for free: respecting temporal order imposes an
-order in which states and terms may be evaluated \emph{during} the
-analysis.
+Constraint-based program analyses (e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular,dvanhorn:steckler-wand-toplas97}) typically compute sets of abstract values for each program point.
+%
+These values approximate values arising at run-time for each program point.
+%
+Value sets are computed as the least solution to a set of (inclusion or equality) constraints.
+%
+The constraints must be designed and proved as a sound approximation of the semantics.
+%
+Efficient implementations of these kinds of analyses often take the form of worklist-based graph algorithms for constraint solving, and are thus quite different from the interpreter implementation.
+%
+The approach thus requires effort in constraint system design and implementation, and the resulting system require verification effort to prove the constraint system is sound and that the implementation is correct.
+%%
+
+%%
+This effort increases substantially as the complexity of the analyzed language increases.
+%
+Both the work of maintaining the concrete semantics and constraint system (and the relations between them) must be scaled simultaneously.
+%
+However, constraint systems, which have been extensively studied in their own right, enjoy efficient implementation techniques and can be expressed in declarative logic languages that are heavily optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
+%
+Consequently, constraint-based analyses can be computed quickly.
+%
+For example, Jagannathan and Wright's polymorphic splitting implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses the \Church{} benchmark about 5.5 times faster than the fastest implementation considered here.
+%
+These analyses compute very different things, so the performance comparison is not apples-to-apples.
+%%
+
+%%
+The AAM approach, and the state transition graphs it generates, encodes temporal properties not found in classical constraint-based analyses for higher-order programs.
+%
+Such analyses (ultimately) compute judgments on program terms and contexts, e.g., at expression $e$, variable $x$ may have value $v$.
+%
+The judgments do not relate the order in which expressions and context may be evaluated in a program, e.g., it has nothing to say with regard to question like,
+%
+``Do we always evaluate $e_1$ before $e_2$?'' or
+%
+``Is it always the case that a file handle is opened, read and then closed in that order?''
+%
+The state transition graphs can answer these kinds of queries, but this does not come for free: respecting temporal order imposes an order in which states and terms may be evaluated \emph{during} the analysis.
 
@@
 
-Abstract machines are not only a good model for rapid analysis
-development, they can be systematically developed into efficient
-algorithms that can be proved correct. We view the primary
-contribution of this work as a systematic path that eases the design,
-verification, and implementation of analyses using the abstracting
-abstract machine approach to within a factor of performant
-constraint-based analyses.
+Abstract machines are not only a good model for rapid analysis development, they can be systematically developed into efficient algorithms that can be proved correct.
+%
+We view the primary contribution of this work as a systematic path that eases the design, verification, and implementation of analyses using the abstracting abstract machine approach to within a factor of performant constraint-based analyses.
 
@@
 
-We thank Suresh Jagannathan for providing source code to the
-polymorphic splitting
-analyzer~\cite{dvanhorn:wright-jagannathan-toplas98} and Ilya Sergey
-for the introspective pushdown
-analyzer~\cite{dvanhorn:Earl2012Introspective}.
+We thank Suresh Jagannathan for providing source code to the polymorphic splitting analyzer~\cite{dvanhorn:wright-jagannathan-toplas98} and Ilya Sergey for the introspective pushdown analyzer~\cite{dvanhorn:Earl2012Introspective}.
 

### 2013-07-01 paper/paper.tex
Funding acknowledgements.

@@
 
-We thank Suresh Jagannathan for providing source code to the polymorphic splitting analyzer~\cite{dvanhorn:wright-jagannathan-toplas98} and Ilya Sergey for the introspective pushdown analyzer~\cite{dvanhorn:Earl2012Introspective}.
+We thank Suresh Jagannathan for providing source code to the
+polymorphic splitting
+analyzer~\cite{dvanhorn:wright-jagannathan-toplas98} and Ilya Sergey
+for the introspective pushdown
+analyzer~\cite{dvanhorn:Earl2012Introspective}.
+
+We thank our anonymous reviewers for their detailed comments. This
+material is based on research sponsored by DARPA under the programs
+Auto- mated Program Analysis for Cybersecurity (FA8750-12-2- 0106) and
+Clean-Slate Resilient Adaptive Hosts (CRASH).  The U.S. Government is
+authorized to reproduce and dis- tribute reprints for Governmental
+purposes notwithstanding any copyright notation thereon.
 

### 2013-07-01 paper/paper.tex
Missing newline.
Tweak to acks.

--- 2013-06-27 paper/paper.tex
@@
        {\diffp(\mval{s_i}, \msto(\maddr_i)) \vee \ldots} \\
-\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots}
+\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots} \\
 \diffp(\mval{s}, \mval{s'}) &= \mval{s'} \overset{?}{=} \mval{s} \sqcup \mval{s'}
@@
 
-We thank Suresh Jagannathan for providing source code to the polymorphic splitting analyzer~\cite{dvanhorn:wright-jagannathan-toplas98} and Ilya Sergey for the introspective pushdown analyzer~\cite{dvanhorn:Earl2012Introspective}.
+We thank Suresh Jagannathan for providing source code to the
+polymorphic splitting
+analyzer~\cite{dvanhorn:wright-jagannathan-toplas98} and Ilya Sergey
+for the introspective pushdown
+analyzer~\cite{dvanhorn:Earl2012Introspective}.  We thank our
+anonymous reviewers for their detailed comments.
+
+This material is based on research sponsored by DARPA under the
+programs Automated Program Analysis for Cybersecurity
+(FA8750-12-2-0106) and Clean-Slate Resilient Adaptive Hosts (CRASH).
+The U.S. Government is authorized to reproduce and distribute reprints
+for Governmental purposes notwithstanding any copyright notation
+thereon.
 

### 2013-07-02 paper/paper.tex
Uncomment some acks.
Tightened some wording and moved some less-important comments to footnotes.

@@
-\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
+\documentclass[9pt]{sigplanconf} %{onecol}
 \usepackage{alltt,mathpartir}
@@
 
-\conferenceinfo{WXYZ '05}{date, City.}
-\copyrightyear{2005}
+\conferenceinfo{ICFP '13}{September 25--27, Boston.}
+\copyrightyear{2013}
 \copyrightdata{[to be supplied]}
@@
 %
-Past approaches to designing program analysis have often featured abstractions that are far removed from the original language semantics, requiring ingenuity in their construction and effort in their verification.
+Past approaches to designing program analyses have often featured abstractions that are far removed from the original language semantics, requiring ingenuity in their construction and effort in their verification.
 %
@@
 %
- Part (a) shows the result of the baseline analyzer.
+ Part (a) shows the baseline analyzer result.
 %
@@
 
-The semantics are defined in terms of a machine model.
+The semantics is defined in terms of a machine model.
 %
@@
 
-\paragraph{Abstract interpretation} To characterize abstract
-interpretation, set the implicit parameters just as above, but drop
-the $\maddr \not\in \msto$ condition. The $\interpdelta$ relation takes some care
-to not make the analysis run forever; a simple instantiation is a flat
-abstraction where arithmetic operations return an abstract top element
-$\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on
-$\sNum$.  This family of interpreters is also non-deterministic in
-choices of addresses, but it is free to choose addresses that are
-already in use.  Consequently, the machines may be non-deterministic
-when multiple values reside in a store location.
-
-It is important to recognize from this definition that \emph{any}
-allocation strategy is a sound abstract
-interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
-concrete interpretation is a kind of abstract interpretation.  So is
-an interpretation that allocates a single cell into which all bindings
-and continuations are stored.  The former is an abstract
-interpretation that is non-computable and gives only the ground truth
-of a programs behavior; the latter is an abstract interpretation
-that is easy to compute but gives little information.  Useful program
-analyses lay somewhere in between and can be characterized by their
-choice of address representation and allocation strategy.  Uniform
-\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97}, presented next, is one such analysis.
-
-\paragraph{Uniform \(k\)-CFA} To characterize uniform \(k\)-CFA, set the allocation
-strategy as follows, for a fixed constant \(k\):
-
+\paragraph{Abstract interpretation}
+%
+To characterize abstract interpretation, set the implicit parameters just as above, but drop the $\maddr \not\in \msto$ condition.
+%
+The $\interpdelta$ relation takes some care to not make the analysis run forever; a simple instantiation is a flat abstraction where arithmetic operations return an abstract top element $\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on $\sNum$.
+%
+This family of interpreters is also non-deterministic in choices of addresses, but it is free to choose addresses that are already in use.
+%
+Consequently, the machines may be non-deterministic when multiple values reside in a store location.
+%%
+
+%%
+It is important to recognize from this definition that \emph{any} allocation strategy is a sound abstract interpretation~\cite{dvanhorn:Might2009Posteriori}.
+%
+In particular, concrete interpretation is a kind of abstract interpretation.
+%
+So is an interpretation that allocates a single cell into which all bindings and continuations are stored.
+%
+The former is an abstract interpretation with uncomputable reachability and gives only the ground truth of a program's behavior; the latter is an abstract interpretation that is easy to compute but gives little information.
+%
+Useful program analyses lay somewhere in between and can be characterized by their choice of address representation and allocation strategy.
+%
+Uniform \(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97}, presented next, is one such analysis.
+%%
+
+%%
+\paragraph{Uniform \(k\)-CFA}
+%
+To characterize uniform \(k\)-CFA, set the allocation strategy as follows, for a fixed constant \(k\):
+%%
+
+%%
 {\small
@@
 
-A common technique to accelerate convergence in flow analyses is to share a
-common, global store.
-%
-To retain soundness, this store grows monotonically.
-%
-Formally, we can cast this optimization as a second abstraction or as
-the application of a widening operator\footnote{Technically, we would
-  have to copy the value of the global store to all states being
-  stepped to fit the formal definition of a widening, but this
-  representation is order-isomorphic to that.} during the fixed-point
-iteration.
-%
-In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
-program.
-%
-Thus, complexity drops from intractable exponentiality to a merely
-daunting polynomial.
-
-Since we can cast this optimization as a widening, there is no need to change
-the transition relation itself.
+A common technique to accelerate convergence in flow analyses is to share a common, global store.
+%
+Formally, we can cast this optimization as a second abstraction or as the application of a widening operator
+%
+\footnote{Technically, we would have to copy the value of the global store to all states being stepped to fit the formal definition of a widening, but this representation is order-isomorphic to that.}
+%
+during the fixed-point iteration.
+%
+In the ISWIM language, such a widening makes 0-CFA quartic in the size of the program.
+%
+Thus, complexity drops from intractable exponentiality to a merely daunting polynomial.
+%%
+
+%%
+Since we can cast this optimization as a widening, there is no need to change the transition relation itself.
 %
@@
 %
-In each pass, the algorithm will collect all newly produced stores and join
-them together.
-%
-Then, before each transition, it installs this joined store into current
-state.
-
-To describe this process, AAM defined a transformation of the reduction relation so that it operates on
-a pair of a set of contexts ($C$) and a store ($\sigma$).
+In each pass, the algorithm will collect all newly produced stores and join them together.
+%
+Then, before each transition, it installs this joined store into current state.
+%%
+
+%%
+To describe this process, AAM defined a transformation of the reduction relation so that it operates on a pair of a set of contexts ($C$) and a store ($\sigma$).
 %
@@
 %
-The new store is computed as the least upper bound of all
-stores after stepping in order to have only one, and keep the
-semantics sound.
+To retain soundness, this store grows monotonically as the least upper bound of all occurring stores.
+%
+%The new store is computed as the least upper bound of all stores after stepping in order to have only one, and keep the semantics sound.
 
@@
 
-The final approximation we make to get to our baseline is to
-store-allocate all values that appear, so that any non-machine state
-that contains a value instead contains an address to a value.  The AAM
-approach stops at the previous optimization.  However, the \kfnalone
-continuation stores a value, and this makes the space of continuations
-quadratic rather than linear in the size of the program, for a
-monovariant analysis like 0-CFA.  Having the space of continuations
-grow linearly with the size of the program will drop the overall
-complexity to cubic (as expected). We also need to allocate an address
-for the argument position in an \apalone state.
-
-To achieve this linearity for continuations, we allocate an address
-for the value position when we create the continuation.  This address
-and the tail address are both determined by the label of the
-application point, so the space becomes linear and the overall
-complexity drops to cubic.  This is a critical abstraction in
-languages with $n$-ary functions, since otherwise the continuation
-space grows super-exponentially (${\mathcal O}(n^n)$). We extend the semantics to
-additionally allocate an address for the function value when creating
-the $\kfnalone$ continuation. The continuation has to contain this address
-to remember where to retrieve values from in the store.
-
+The final approximation we make to get to our baseline is to store-allocate all values that appear, so that any non-machine state that contains a value instead contains an address to a value.
+%
+The AAM approach stops at the previous optimization.
+%
+However, the \kfnalone continuation stores a value, and this makes the space of continuations quadratic rather than linear in the size of the program, for a monovariant analysis like 0-CFA.
+%
+Having the space of continuations grow linearly with the size of the program will drop the overall complexity to cubic (as expected).
+%
+We also need to allocate an address for the argument position in an \apalone state.
+%%
+
+%%
+To achieve this linearity for continuations, we allocate an address for the value position when we create the continuation.
+%
+This address and the tail address are both determined by the label of the application point, so the space becomes linear and the overall complexity drops to cubic.
+%
+This is a critical abstraction in languages with $n$-ary functions, since otherwise the continuation space grows super-exponentially (${\mathcal O}(n^n)$).
+%
+We extend the semantics to additionally allocate an address for the function value when creating the $\kfnalone$ continuation.
+%
+The continuation has to contain this address to remember where to retrieve values from in the store.
+%%
+
+%%
 The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
@@
 %
-To inject a program $e$ into this machine, we start off knowing we
-have seen the first state, and that we need to process the first
-state:
+To inject a program $e$ into this machine, we start off knowing we have seen the first state, and that we need to process the first state:
 \begin{align*}
@@
 \end{align*}
-
-Notice that now $\mseen$ has several copies of the abstract store in it. As
-it is, this semantics is much less efficient (but still more precise)
-than the previously proposed semantics because membership checks have
-to compare entire stores. Checking equality is expensive because the stores within each are
-large, and every entry must be checked against every other. Hashes can
-sometimes rule out inequality relatively quickly, but the incidence of
-collisions and actual equality is costly.
-
-And, there is a better way. Shivers' original work on $k$-CFA was
-susceptible to the same problem, and he suggested three complementary
-optimizations: (1) make the store global; (2) update the store
-imperatively; and (3) associate every change in the store with a
-version number -- its timestamp. Then, put timestamps in states
-where previously there were stores. Given two states, the analysis can
-now compare their stores just by comparing their timestamps -- a
+%%
+
+%%
+Notice that now $\mseen$ has several copies of the abstract store in it.
+%
+As it is, this semantics is much less efficient (but still more precise) than the previously proposed semantics because membership checks have to compare entire stores.
+%
+Checking equality is expensive because the stores within each state are large, and nearly every entry must be checked against every other due to high similarities amongst stores.
+%%
+
+%%
+And, there is a better way.
+%
+Shivers' original work on $k$-CFA was susceptible to the same problem, and he suggested three complementary optimizations:
+%
+ (1) make the store global;
+%
+ (2) update the store imperatively; and
+%
+ (3) associate every change in the store with a version number -- its timestamp.
+%
+Then, put timestamps in states where previously there were stores.
+%
+Given two states, the analysis can now compare their stores just by comparing their timestamps -- a
 constant-time operation.
@@
 \end{align*}}
-
-Here $\appendall$ combines change logs across all non-deterministic
-steps for a state to later be replayed. The order the combination
-happens in doesn't matter, because join is associative and
-commutative.
-
+%%
+
+%%
+Here $\appendall$ combines change logs across all non-deterministic steps for a state to later be replayed.
+%
+The order the combination happens in doesn't matter, because join is associative and commutative.
+%%
+
+%%
 \begin{lemma}
-$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff
-$\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
+$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff $\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
 \end{lemma}
 By cases on $\dmachstep$ and $\machstep$.
-
+%%
+
+%%
 \begin{lemma}[$\changep$ means change]
@@
 By induction on $\msdiff$.
-
+%%
+
+%%
 \begin{theorem}
@@
 
-Tracing the execution of the analysis reveals an immediate shortcoming:
-there is a high degree of branching and merging in the exploration.
+Tracing the execution of the analysis reveals an immediate shortcoming: there is a high degree of branching and merging in the exploration.
 %
@@
 %
-For example, in a function application, {\tt (f x y)},
-where {\tt f}, {\tt x} and {\tt y} each have several values
-each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective
-application positions.
+For example, in a function application, {\tt (f x y)}, where {\tt f}, {\tt x} and {\tt y} each have several values each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective application positions.
 %
@@
 \end{center}
-To avoid the spurious forking and joining, we {\it delay} the non-determinism
-until and unless it is needed in {\it strict contexts} (such as the guard for an
-{\tt if}, a called procedure, or a numerical primitive application).
+To avoid the spurious forking and joining, we {\it delay} the non-determinism until and unless it is needed in {\it strict contexts} (such as the guard of an {\tt if}, a called procedure, or a numerical primitive application).
 %
@@
 \end{center}
-
-This shift does not change the concrete semantics of the language to
-be lazy.  Rather, it abstracts over transitions that the original
-non-deterministic semantics steps through.
-%
-We say the abstraction is \emph{lazy} because it delays splitting on
-the values in an address until they are \emph{needed} in the
-semantics. It does not change the execution order that leads to the
-values that are stored in the address.
+%%
+
+%%
+This shift does not change the concrete semantics of the language to be lazy.
+%
+Rather, it abstracts over transitions that the original non-deterministic semantics steps through.
+%
+We say the abstraction is \emph{lazy} because it delays splitting on the values in an address until they are \emph{needed} in the semantics.
+%
+It does not change the execution order that leads to the values that are stored in the address.
 
@@
 &\lmachstep\;
-\ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
+\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont} \\
 \text{ if } \mkont &\in \msto(\mkaddr),
@@
 Since {\tt if} guards are in strict position, we must force the value
-to determine which branch to to take. The middle rule uses $\force$
-only to combine with values in the store - it does not introduce
-needless non-determinism.
-
+to determine which branch to to take.
+%
+The middle rule uses $\force$ only to combine with values in the store - it does not introduce needless non-determinism.
+%%
+
+%%
 \spchoice{
@@
 \paragraph{Option 1: Lose precision; simplify implementation}
-This semantics introduces a subtle precision difference over the
-baseline. Consider a configuration where a reference to a variable and
-a binding of a variable will happen in one step, since store widening
-leads to stepping several states in one big ``step.'' With laziness,
-the reference will mean the original binding(s) of the variable or the
-new binding, because the actual store lookup is delayed one step
-(i.e. laziness is administrative).
+This semantics introduces a subtle precision difference over the baseline.
+%
+Consider a configuration where a reference to a variable and a binding of a variable will happen in one step, since store widening leads to stepping several states in one big ``step.''
+%
+With laziness, the reference will mean the original binding(s) of the variable \emph{or} the new binding, because the actual store lookup is delayed one step (i.e. laziness is administrative).
 
 \paragraph{Option 2: Regain precision; complicate implementation}
-The administrative nature of laziness means that we could remove the
-loss in precision by storing the result of the lookup in a value
-representing a delayed nondeterministic choice. This is a more common
-choice in 0CFA implementations we have seen, but it interferes with
-the next optimization (discussion in next section).
-
-\paragraph{Our choice: option 1}
-The conditions for precision loss happen too rarely to warrant the
-added complexity to the next section's semantics to avoid such
-losses. Without abstract compilation, option 2 is the better
-choice.}{}
+The administrative nature of laziness means that we could remove the loss in precision by storing the result of the lookup in a value representing a delayed nondeterministic choice.
+%
+This is a more common choice in 0CFA implementations we have seen, but it interferes with the next optimization due to the invariant from store deltas we have that lookups must not depend on the change log.
+
+% \paragraph{Our choice: option 1}
+% The conditions for precision loss happen too rarely to warrant the
+% added complexity to the next section's semantics to avoid such
+% losses. Without abstract compilation, option 2 is the better
+% choice.
+}{}
 
@@
 \paragraph{Correctness}
-The correctness of abstract compilation seems obvious, but it has
-never before been rigorously proved. What constitutes correctness in
-the case of dropped states, anyway? Applying an abstract bytecode's
-function does many ``steps'' in one go, at the end of which, the two
-semantics line up again (modulo representation of expressions). This
-constitutes the use of a notion of stuttering. We provide a formal
-analysis of abstract compilation \emph{without} store widening with a proof
-of a stuttering bisimulation~\cite{ianjohnson:BCG88} between this
-semantics and lazy non-determinism without widening to show precision
-preservation. %We revisit store widening at the end of the section.
-
-The number of transitions that can occur in succession from an
-abstract bytecode is roughly bounded by the amount of expression
-nesting in the program. The partial order that defines subexpressions
-smaller than the expression in which they occur is well-founded, so we
-use this observation to drive our use of a well-founded equivalence
-bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are
-equivalent to the notion of a stuttering bisimulation, but are more
-amenable to mechanization (and thus rigorous proof). They also only
-require reasoning over one step of the reduction relation, which makes
-the proof concise. The trick is in defining a well-founded ordering
-that determines when the two semantics will match up again, what
-Manolios calls the pair of functions $\erankt$ and $\erankl$ (but we
-don't need the generality that requires $\erankl$ since the uncompiled
-semantics doesn't stutter at all).
-
-We define a refinement, $r$, from non-compiled to compiled states (built structurally) by
-``committing'' all the actions of an $\evalone$ state (defined similarly to
-$\compile{\_}$, but immediately applies the functions), and
-subsequently changing all expressions with their compiled
-variants. Since WEBs are for single transition systems, a WEB
-refinement is over the disjoint union of our two semantics, and the
-equivalence relation we use is just that a state is related to its
-refined state (and itself). Call this relation $B$.
-
-Before we prove this setup is indeed a WEB, we need one lemma that
-applying an abstract bytecode's function is equal to refining the
-corresponding $\evalone$ state:
+The correctness of abstract compilation seems obvious, but it has never before been rigorously proved.
+%
+What constitutes correctness in the case of dropped states, anyway?
+%
+Applying an abstract bytecode's function does many ``steps'' in one go, at the end of which, the two semantics line up again (modulo representation of expressions).
+%
+ This constitutes the use of a notion of stuttering.
+%
+ We provide a formal analysis of abstract compilation \emph{without} store widening with a proof of a stuttering bisimulation~\cite{ianjohnson:BCG88} between this semantics and lazy non-determinism without widening to show precision preservation.
+%We revisit store widening at the end of the section.
+%%
+
+%%
+The number of transitions that can occur in succession from an abstract bytecode is roughly bounded by the amount of expression nesting in the program.
+%
+We can use the expression containment order to prove stuttering bisimulation with a well-founded equivalence bisimulation (WEB)~\cite{ianjohnson:manolios-diss}.
+%The partial order that defines subexpressions smaller than the expression in which they occur is well-founded, so we use this observation to drive our use of a well-founded equivalence bisimulation (WEB)
+%
+ WEBs are equivalent to the notion of a stuttering bisimulation, but are more amenable to mechanization since they also only require reasoning over one step of the reduction relation.
+%
+The trick is in defining a well-founded ordering that determines when the two semantics will match up again, what Manolios calls the pair of functions $\erankt$ and $\erankl$ (but we don't need $\erankl$ since the uncompiled semantics doesn't stutter).
+%%
+
+%%
+We define a refinement, $r$, from non-compiled to compiled states (built structurally) by ``committing'' all the actions of an $\evalone$ state (defined similarly to $\compile{\_}$, but immediately applies the functions), and subsequently changing all expressions with their compiled variants.
+%
+ Since WEBs are for single transition systems, a WEB refinement is over the disjoint union of our two semantics, and the equivalence relation we use is just that a state is related to its refined state (and itself).
+%
+ Call this relation $B$.
+%%
+
+%%
+Before we prove this setup is indeed a WEB, we need one lemma that applying an abstract bytecode's function is equal to refining the corresponding $\evalone$ state:
 %
@@
 
-The proof follows by cases on $\lmachstep \uplus \machstep$ with the
-WEB \emph{witness} being the well-order on expressions (with a $\bot$
-element), and the following $\erankt$, $\erankl$ functions:
+The proof follows by cases on $\lmachstep \uplus \machstep$ with the WEB \emph{witness} being the well-order on expressions (with a $\bot$ element), and the following $\erankt$, $\erankl$ functions:
 
@@
 \end{align*}
-All cases are either simple steps or appeals to the well-order on $\erankt$'s range. The other rank function,
-$\erankl$ is unnecessary, so we just make it the constant 0
-function. The $\cmachstep$ cases are trivial.
+All cases are either simple steps or appeals to the well-order on $\erankt$'s range.
+%
+The other rank function, $\erankl$ is unnecessary, so we just make it the constant 0 function.
+%
+The $\cmachstep$ cases are trivial.
 
 \paragraph{Wide store and abstract compilation}
-The fact that many changes happen in one step also makes the
-comparison with the widened lazy semantics more subtle. It is possible
-for different stores to occur between the different semantics because
-abstract compilation can change the order in which the store is
-changed (across steps). Consider the case where before we might call
-the same function from two different places in 3 and 2 steps
-respectively, compilation can make that 2 and 1, reversing the order
-that we bind values in the store, leading to a mismatch from the wide
-lazy semantics.  This kind of cross-step store poisoning is inevitable
-when a store is shared between several traces. The binding could have
-just as easily been observed the other way around in the non-compiled
-semantics. Although there is a difference pre- and post- abstract
-compilation, the result is still deterministic in contrast to Shivers'
-technique. The soundness is in tact since we can add store-widening
-the correct unwidened semantics with an easy correctness proof. Call
-$\camachstep$ the result of the widening operator from the previous
-section on $\cmachstep$.
+%The fact that many changes happen in one step also makes the comparison with the widened lazy semantics more subtle.
+%
+ It is possible for different stores to occur between the different semantics because abstract compilation can change the order in which the store is changed (across steps).
+%
+ This is the case because some ``corridor'' expressions may compile down to change the store before some others, meaning there is no stuttering relationship with the wide lazy semantics.
+%
+% This kind of cross-step store poisoning is inevitable when a store is shared between several traces.
+%
+% The binding could have just as easily been observed the other way around in the non-compiled semantics.
+%
+ Although there is a difference pre- and post- abstract compilation, the result is still deterministic in contrast to Shivers' technique.
+%
+ The soundness is in tact since we can add store-widening the correct unwidened semantics with an easy correctness proof.
+%
+ Call $\camachstep$ the result of the widening operator from the previous section on $\cmachstep$.
 
@@
 
-Thus far, we have made our optimizations in a purely functional
-manner. For the final push for performance, we need to dip into the
-imperative. In this section, we show an alternative representation of
-the store and seen set that are more space-efficient and are amenable
-to destructive updates by adhering to a history for each address.
-
-The following transfer function has several components that can be
-destructively updated, and intermediate sets can be elided by adding
-to global sets. In fact, the log of store deltas can be removed as
-well, by updating the store in-place, and on lookup, using the first
-value timestamped $\le$ the current timestamp. We start with the
-purely functional view.
+Thus far, we have made our optimizations in a purely functional manner.
+%
+For the final push for performance, we need to dip into the imperative.
+%
+In this section, we show an alternative representation of the store and seen set that are more space-efficient and are amenable to destructive updates by adhering to a history for each address.
+%%
+
+%%
+The following transfer function has several components that can be destructively updated, and intermediate sets can be elided by adding to global sets.
+%
+In fact, the log of store deltas can be removed as well, by updating the store in-place, and on lookup, using the first value timestamped $\le$ the current timestamp.
+%
+We start with the purely functional view.
 
@@
 
-The store maps to a stack of timestamped sets of abstract
-values. Throughout this section, we will be taking the parameter $t$
-to be the ``current time,'' or the length of the store chain at the
-beginning of the step.
+The store maps to a stack of timestamped sets of abstract values.
+%
+Throughout this section, we will be taking the parameter $t$ to be the ``current time,'' or the length of the store chain at the beginning of the step.
 
@@
     \end{array}\right.
-\end{align*}} To construct this value stack, we have a
-time-parameterized join operation that also tracks changes to the
-store. If joining with a time in the future, we just add to
-it. Otherwise, we're making a change for the future ($t+1$), but only
-if there is an actual change.
+\end{align*}}
+%
+To construct this value stack, we have a time-parameterized join operation that also tracks changes to the store.
+%
+If joining with a time in the future, we just add to it.
+%
+Otherwise, we're making a change for the future ($t+1$), but only if there is an actual change.
 {\small
@@
 
-For the purposes of space, we reuse the $\cmachstep$ semantics,
-although the $\replay$ of the produced $\msdiff$ objects should be
-in-place, and the $\lookup$ function should be using this
-single-threaded store.  Because the store has all the temporal
-information baked into it, we rephrase the core semantics in terms of
-a transfer function. The least fixed-point of this function gives a
-more compact representation of the reduction relation of the previous
-section.
+For the purposes of space, we reuse the $\cmachstep$ semantics, although the $\replay$ of the produced $\msdiff$ objects should be in-place, and the $\lookup$ function should be using this single-threaded store.
+%
+Because the store has all the temporal information baked into it, we rephrase the core semantics in terms of a transfer function.
+%
+The least fixed-point of this function gives a more compact representation of the reduction relation of the previous section.
 
@@
 \item{$\msto$: the store, which represents all stores that occur in the machine semantics.}
-\item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}
-\item{$\changep$: whether a store change occurred when stepping states in $F$.}
+\item{$t$: the timestamp, or length of the store chain.}
+\item{$\changep$: whether the store changed when stepping states in $F$.}
 \end{enumerate}
@@
 %
-But, such a dynamic structure isn't necessary when we know the structure of the store in advance: we know all possible entries, and we know its maximum size.
-%
-In a monovariant allocation strategy, the domain of the store is exactly the set of expressions in the program.
+But, such a dynamic structure isn't necessary when we know the structure of the store in advance.%: we know all possible entries, and we know its maximum size.
+%
+In a monovariant allocation strategy, the domain of the store is bounded by the number of expressions in the program.
 %
@@
 The analysis is evaluated against a suite of Scheme benchmarks drawn from the literature.
+\footnote{Source code of the implementation and benchmark suite available at \url{https://github.com/dvanhorn/oaam}}
 %
@@
 %
-The analysis is stopped after consuming 30 minutes of time or 1 gigabyte of space.
-%
-When presenting \emph{relative} numbers, we use the timeout limits as a lower bound on the actual time required (i.e., one minute versus timeout is at least 30 times faster),
-thus giving a conservative estimate of improvements.
-%%
-
-%%
-All benchmarks are calculated as an average of 5 runs, done in
-parallel (each isolated to a core), on an 12-core, 64-bit Intel Xeon machine running at 2.40GHz
-with 12Gb of memory.
-%%
-
-%%
-Many benchmarks cause the baseline analyzer to take longer than 30 minutes or to consume more than 1 gigabyte of memory, at which point the analysis is stopped.
-%
-This is the case for the largest benchmark program, {\bf nucleic}, which is 3,500 lines of code and takes under a minute in the most optimized analyzer.
+The analysis is stopped after consuming 30 minutes of time or 1 gigabyte of space
+\footnote{All benchmarks are calculated as an average of 5 runs, done in parallel (each isolated to a core), on an 12-core, 64-bit Intel Xeon machine running at 2.40GHz with 12Gb of memory.}.
+%
+When presenting \emph{relative} numbers, we use the timeout limits as a lower bound on the actual time required (i.e., one minute versus timeout is at least 30 times faster), thus giving a conservative estimate of improvements.
+%%
+
+%%
+%%
+%%
+%%
+%% Many benchmarks cause the baseline analyzer to take longer than 30 minutes or to consume more than 1 gigabyte of memory, at which point the analysis is stopped.
+%
+%%This is the case for the largest benchmark program, {\bf nucleic}, which is 3,500 lines of code and takes under a minute in the most optimized analyzer.
 %
@@
 %
-Moreover, much of the performance gains are achieved with purely functional methods, which suggest potential further gains from parellelization.
+Moreover, much of the performance gains are achieved with purely functional methods, which allow the use of these methods in rewriting tools and others with restricted input languages.
 %
@@
 
-%%
-Source code of the implementation and benchmark suite is at:
-
-\begin{center}
-\url{https://github.com/dvanhorn/oaam}
-\end{center}
-
 \paragraph{Comparison with other flow analysis implementations}
@@
 %%
-The only analyzers we were able to find that proved capable of
-analyzing the full suite of benchmarks considered here were the Soft
-Typing system of Wright and
-Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
-successor, the Polymorphic splitting system of Wright and
-Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.
+The only analyzer we were able to find that proved capable of analyzing the full suite of benchmarks considered here
+% were the Soft Typing system of Wright and Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its successor,
+was the Polymorphic splitting system of Wright and Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.
 %
@@
 %
-Unfortunately, these analyses compute an inherently different and incomparable form of analysis.
+Unfortunately, these analyses compute an inherently different and incomparable form of analysis via a global acceptability judgment.
 %
@@
 %
-``Do we always evaluate $e_1$ before $e_2$?'' or
-%
-``Is it always the case that a file handle is opened, read and then closed in that order?''
-%
-The state transition graphs can answer these kinds of queries, but this does not come for free: respecting temporal order imposes an order in which states and terms may be evaluated \emph{during} the analysis.
+``Do we always evaluate $e_1$ before $e_2$?''% or
+%
+%``Is it always the case that a file handle is opened, read and then closed in that order?''
+%
+The state transition graphs can answer these kinds of queries, but evaluation demonstrated this does not come for free.
+%%: respecting temporal order imposes an order in which states and terms may be evaluated \emph{during} the analysis.
 
@@
 
-%% \acks Sam Tobin-Hochstadt for encouragement and feedback -- he was
-%% the first to prompt us to look into how make effective
-%% implementations of the AAM approach.
-
-%% Vincent St Amour and Mitchell Wand for feedback on early drafts.
-%% Greg Morrisset and Matthias Felleisen for discussions.
-
+%% \acks 
 %% NSF, DARPA
@@
 for the introspective pushdown
-analyzer~\cite{dvanhorn:Earl2012Introspective}.  We thank our
-anonymous reviewers for their detailed comments.
-
+analyzer~\cite{dvanhorn:Earl2012Introspective}.  
+%
+We thank Sam Tobin-Hochstadt for encouragement and feedback---he was
+the first to prompt us to look into how make effective
+implementations of the AAM approach.
+%
+We thank Vincent St-Amour and Mitchell Wand for feedback on early drafts
+and Greg Morrisett and Matthias Felleisen for discussions.
+%
+We thank our anonymous reviewers for their detailed comments.
+%
 This material is based on research sponsored by DARPA under the

### 2013-07-02 paper/paper.tex
Removed the convoluted abstract compilation example and referred the reader back to the state graphs instead. One citation over 12 pages

@@
 
-The possible traces of execution tell us the most about a program, so
-we take $\traces(\mexp)$ to be the (regular) set of paths through the
-computed graph. We elide the construction of the set of edges in this paper.
-
-To conduct this \naive{} exploration on the \Church{} example would require
-considerable time.  Even though the state space is finite, it is exponential in
-the size of the program.  Even with $k = 0$, there are exponentially many
-stores in the AAM framework.
-
-In the next subsection, we fix this with store widening to reach polynomial
-(albeit of high degree) complexity.
-%
-This widening effectively lifts the store out of individual states to create
-a single, global shared store for all.
+The possible traces of execution tell us the most about a program, so we take $\traces(\mexp)$ to be the (regular) set of paths through the computed graph.
+%
+ We elide the construction of the set of edges in this paper.
+%%
+
+%%
+To conduct this \naive{} exploration on the \Church{} example would require considerable time.
+%
+  Even though the state space is finite, it is exponential in the size of the program.
+%
+  Even with $k = 0$, there are exponentially many stores in the AAM framework.
+%%
+
+%%
+In the next subsection, we fix this with store widening to reach polynomial (albeit of high degree) complexity.
+%
+This widening effectively lifts the store out of individual states to create a single, global shared store for all.
 
@@
 \end{align*}
-Now instead of storing the evaluated function in the continuation
-frame itself, we indirect it through the store for further control on
-complexity and precision:
+%%
+Now instead of storing the evaluated function in the continuation frame itself, we indirect it through the store for further control on complexity and precision:
+%%
 \begin{align*}
@@
 \end{align*}
-
-Associated with this indirection, we now apply all functions stored in
-the address. This nondeterminism is necessary in order to continue with
-evaluation.
+%%
+
+%%
+Associated with this indirection, we now apply all functions stored in the address.
+%
+This nondeterminism is necessary in order to continue with evaluation.
 
@@
 \end{enumerate}
-
-All pragmatic improvements are precision preserving (form complete
-abstractions), but the semantic changes are not in some cases, for reasons we
-will describe. We did not observe the precision differences in our evaluation.
-
-We apply the frontier-based semantics combined with timestamped stores
-as our first step.  The move to the imperative will be made last in
-order to show the effectiveness of these techniques in the purely
-functional realm.
+%%
+
+%%
+All pragmatic improvements are precision preserving (form complete abstractions), but the semantic changes are not in some cases, for reasons we will describe.
+%
+ We did not observe the precision differences in our evaluation.
+%%
+
+%%
+We apply the frontier-based semantics combined with timestamped stores as our first step.
+%
+  The move to the imperative will be made last in order to show the effectiveness of these techniques in the purely functional realm.
 
@@
 
-The semantics given for store widening in section \ref{sec:storewiden},
-while simple, is wasteful. It also does not model what typical
-implementations do. It causes all states found so far to step each
-iteration, even if they are not revisited. This has negative
-performance \emph{and} precision consequences (changes to the store
-can travel back in time in straight-line code). We instead use a
-frontier-based semantics that corresponds to the classic worklist
-algorithms for analysis. The difference is that the store is not
-modified in-place, but updated after all frontier states have been
-processed. This has implications for the analysis' precision and
-determinism. Specifically, higher precision, and it is deterministic
-even if set iteration is not.
-
-The state space changes from a store and set of contexts to a set
-of seen abstract states (context plus store), $\mseen$, a set of contexts to
-step (the frontier), $F$, and a store to step those contexts with, $\msto$:
+The semantics given for store widening in section \ref{sec:storewiden}, while simple, is wasteful.
+%
+ It also does not model what typical implementations do.
+%
+ It causes all states found so far to step each iteration, even if they are not revisited.
+%
+ This has negative performance \emph{and} precision consequences (changes to the store can travel back in time in straight-line code).
+%
+ We instead use a frontier-based semantics that corresponds to the classic worklist algorithms for analysis.
+%
+ The difference is that the store is not modified in-place, but updated after all frontier states have been processed.
+%
+ This has implications for the analysis' precision and determinism.
+%
+ Specifically, higher precision, and it is deterministic even if set iteration is not.
+%%
+
+%%
+The state space changes from a store and set of contexts to a set of seen abstract states (context plus store), $\mseen$, a set of contexts to step (the frontier), $F$, and a store to step those contexts with, $\msto$:
 \begin{equation*}
@@
 %
-Given two states, the analysis can now compare their stores just by comparing their timestamps -- a
-constant-time operation.
-
-There are two subtle losses of precision in Shivers' original
-timestamp technique that we can fix.
+Given two states, the analysis can now compare their stores just by comparing their timestamps -- a constant-time operation.
+
+There are two subtle losses of precision in Shivers' original timestamp technique that we can fix.
 
 \begin{enumerate}
-\item{In our semantics, the store does not change until the entire
-    frontier has been explored. This avoids cross-branch pollution
-    which would otherwise happen in Shivers' semantics, e.g., when one branch writes to
-    address $\maddr$ and another branch reads from address
-    $\maddr$.}
-\item{The common implementation strategy for timestamps destructively
-    updates each state's timestamp. This loses \emph{temporal}
-    information about the contexts a state is visited in, and in what
-    order. Our semantics has a drop-in replacement of timestamps for
-    stores in the seen set ($\mseentime$), so we do not experience precision
-    loss.}
+\item{In our semantics, the store does not change until the entire frontier has been explored.
+%
+ This avoids cross-branch pollution which would otherwise happen in Shivers' semantics, e.g., when one branch writes to address $\maddr$ and another branch reads from address $\maddr$.}
+\item{The common implementation strategy for timestamps destructively updates each state's timestamp.
+%
+ This loses \emph{temporal} information about the contexts a state is visited in, and in what order.
+%
+ Our semantics has a drop-in replacement of timestamps for stores in the seen set ($\mseentime$), so we do not experience precision loss.}
 \end{enumerate}
@@
 
-The observation Shivers made was that the store is increasing
-monotonically, so all stores throughout execution will be totally
-ordered (form a chain). This observation allows you to replace stores
-with pointers into this chain. We keep the stores around in $\Sigma$
-to achieve a complete abstraction. This corresponds to the temporal
-information about the execution's effect on the store.
-
-Note also that $F$ is only populated with states that have not been
-seen at the resulting store. This is what produces the more precise
-abstraction than the baseline widening.
-
-The general fixed-point combinator we showed above can be specialized
-to this semantics, as well. In fact, $\widehat{\machstep}^T$ is a functional
-relation, so we can get the least fixed-point of it directly.
+The observation Shivers made was that the store is increasing monotonically, so all stores throughout execution will be totally ordered (form a chain).
+%
+ This observation allows you to replace stores with pointers into this chain.
+%
+ We keep the stores around in $\Sigma$ to achieve a complete abstraction.
+%
+ This corresponds to the temporal information about the execution's effect on the store.
+
+Note also that $F$ is only populated with states that have not been seen at the resulting store.
+%
+ This is what produces the more precise abstraction than the baseline widening.
+
+The general fixed-point combinator we showed above can be specialized to this semantics, as well.
+%
+ In fact, $\widehat{\machstep}^T$ is a functional relation, so we can get the least fixed-point of it directly.
 
 \begin{lemma}
-  $\widehat{\machstep}$ maintains the invariant that all stores in $\mseen$ are
-  totally ordered and $\msto$ is an upper bound of the stores in $\mseen$.
+  $\widehat{\machstep}$ maintains the invariant that all stores in $\mseen$ are totally ordered and $\msto$ is an upper bound of the stores in $\mseen$.
 \end{lemma}
@@
 \begin{lemma}
-  $\widehat{\machstep}^T$ maintains the invariant that $\Sigma$ is in
-  order with respect to $\sqsupset$ and $\msto = \hd(\Sigma)$.
+  $\widehat{\machstep}^T$ maintains the invariant that $\Sigma$ is in order with respect to $\sqsupset$ and $\msto = \hd(\Sigma)$.
 \end{lemma}
@@
 \end{theorem}
-The proof follows from the order isomorphism that, in one direction,
-sorts all the stores in $\mseen$ to form $\Sigma$, and translates stores in
-$\mseen$ to their distance from the end of $\Sigma$ (their timestamp). In the other
-direction, timestamps in $\mseentime$ are replaced by the stores they point to in $\Sigma$.
+The proof follows from the order isomorphism that, in one direction, sorts all the stores in $\mseen$ to form $\Sigma$, and translates stores in $\mseen$ to their distance from the end of $\Sigma$ (their timestamp).
+%
+ In the other direction, timestamps in $\mseentime$ are replaced by the stores they point to in $\Sigma$.
 
@@
 
-The above technique requires joining entire (large) stores
-together. Additionally, there is still a comparison of stores, which
-we established is expensive. Not every step will modify all addresses
-of the store, so joining entire stores is wasteful in terms of memory
-and time. We can instead log store changes and replay the change log
-on the full store after all steps have completed, noting when there is
-an actual change. This uses far fewer join and comparison operations,
-leading to less overhead, and is precision-preserving.
-
-We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
-  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
- becomes a log addition
-$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
-empty ($\mtlst$) for each step. Applying the changes to the full store
-is straightforward:
+The above technique requires joining entire (large) stores together.
+%
+ Additionally, there is still a comparison of stores, which we established is expensive.
+%
+ Not every step will modify all addresses of the store, so joining entire stores is wasteful in terms of memory and time.
+%
+ We can instead log store changes and replay the change log on the full store after all steps have completed, noting when there is an actual change.
+%
+ This uses far fewer join and comparison operations, leading to less overhead, and is precision-preserving.
+
+We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times \Set(\Storeable))^*$.
+%
+ Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ becomes a log addition $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins empty ($\mtlst$) for each step.
+%
+ Applying the changes to the full store is straightforward:
 \begin{equation*}
@@
 
-We change the semantics slightly to add to the change log rather than
-produce an entire modified store.  The transition relation is
-identical except for the addition of this change log.  We maintain the
-invariant that lookups will never rely on the change log, so we can
-use the originally supplied store unmodified.
+We change the semantics slightly to add to the change log rather than produce an entire modified store.
+%
+The transition relation is identical except for the addition of this change log.
+%
+  We maintain the invariant that lookups will never rely on the change log, so we can use the originally supplied store unmodified.
 
@@
 
-We lift $\dmachstep$ to accommodate for the asymmetry
-in the input and output, and change the frontier-based semantics in
-the following way:
+We lift $\dmachstep$ to accommodate for the asymmetry in the input and output, and change the frontier-based semantics in the following way:
 {\small
@@
 \end{align*}}
-Since {\tt if} guards are in strict position, we must force the value
-to determine which branch to to take.
+Since {\tt if} guards are in strict position, we must force the value to determine which branch to to take.
 %
@@
 \begin{theorem}[Soundness]
-  If $\mstate \machstep \mstate'$ and $\mstate \sqsubseteq
-  \mastate$ then there exists a $\mastate'$ such that $\mastate
-  \lmachstep \mastate'$ and $\mstate' \sqsubseteq \mastate'$
+  If $\mstate \machstep \mstate'$ and $\mstate \sqsubseteq \mastate$ then there exists a $\mastate'$ such that $\mastate \lmachstep \mastate'$ and $\mstate' \sqsubseteq \mastate'$
 \end{theorem}
-Here $\sqsubseteq$ is straightforward --- the left-hand side store must be
-contained in the right-hand-side store, and if values occur in the states,
-the left-hand-side value must be in the forced corresponding right-hand-side value.
+Here $\sqsubseteq$ is straightforward --- the left-hand side store must be contained in the right-hand-side store, and if values occur in the states, the left-hand-side value must be in the forced corresponding right-hand-side value.
+%
 The proof is by cases on $\mstate \machstep \mstate'$.}
@@
 \end{theorem}
-We have a statement about traces because we need induction to show no
-cruft values are in superposition.  The induction hypothesis tells us
-that there are non-lazy traces that lead to all the values in
-superposition, so when we take a lazy step, we are taking several
-non-lazy steps, and we stay in sync. The other direction we just
-collapse the superposition in each possibility to construct the
-non-lazy traces.}
+We have a statement about traces because we need induction to show no cruft values are in superposition.
+%
+The induction hypothesis tells us that there are non-lazy traces that lead to all the values in superposition, so when we take a lazy step, we are taking several non-lazy steps, and we stay in sync.
+%
+ The other direction we just collapse the superposition in each possibility to construct the non-lazy traces.}
 
@@
 
-The prior optimization saved time by doing the same amount of
-reasoning as before but in fewer transitions. We can exploit the same
-idea---same reasoning, fewer transitions---with abstract
-compilation. Abstract compilation transforms complex expressions whose
-\emph{abstract} evaluation is deterministic into ``abstract
-bytecodes.''  The abstract interpreter then does in one transition
-what previously took many. In short, abstract compilation eliminates
-unnecessary allocation, deallocation and branching. The technique is
-precision preserving without store widening. We discuss the precision
-differences with store widening at the end of the section.
-
-The following example illustrates
-the essence of abstract compilation's effect:
-\[
-\mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
-\]
-makes the following transitions:
-\begin{align}
-& \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
-\machstep\; &
-\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_1}}
-\\
-\machstep\; &
-\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_2}}
-\\
-\machstep\; &
-\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_3}} % {\mexp_2}
-\\
-\machstep\; &
-\spchoice
-{\co{\kar{\mexp_1,\menv,\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
-{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
-\end{align}
-where \(
-\msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
-\maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
-\maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).
-
-Whereas abstract compilation gives us in one step:
-\begin{align*}
-\compile{\mexp}^\mcntr(\msto_0)(\menv,\epsilon,\mkont) =
-\spchoice
-{\co{\kar{\compile{\mexp_1},\menv,\maddr_3},\saddr{\menv(\mvar)}},\msdiff_4}
-{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
-\end{align*}
-where
-\begin{equation*}
-\msdiff_4 = [\ttuple{\maddr_1}{\setof{\mkont}},
-             \ttuple{\maddr_2}{\setof{\kar{\compile{\mexp_3},\menv,\maddr_1}}},
-             \ttuple{\maddr_3}{\setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}}]\text.
-\end{equation*}
-
-The compilation step converts expressions into functions that expect
-the other components of the {\tt ev} state. Its definition in figure
-\ref{fig:compile} shows close similarity to the rules for interpreting
-{\tt ev} states. The next step is to change reduction rules that
-create {\tt ev} states to instead call these functions. Figure
-\ref{fig:caam} shows the modified reduction relation. The only change
-from the previous semantics is that $\evalone$ state construction is
-replaced by calling the compiled expression. For notational coherence,
-we write $\lambda^\mcntr(\mathit{args} \ldots)$ for
-$\lambda(\mathit{args} \ldots, \mcntr)$ and
-$\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
+The prior optimization saved time by doing the same amount of reasoning as before but in fewer transitions.
+%
+ We can exploit the same idea---same reasoning, fewer transitions---with abstract compilation.
+%
+ Abstract compilation transforms complex expressions whose \emph{abstract} evaluation is deterministic into ``abstract bytecodes.''
+%
+  The abstract interpreter then does in one transition what previously took many.
+%
+  Refer back to figure \ref{fig:state-graphs} to see the effect of abstract compilation.
+%
+ In short, abstract compilation eliminates unnecessary allocation, deallocation and branching.
+%
+ The technique is precision preserving without store widening.
+%
+ We discuss the precision differences with store widening at the end of the section.
+%%
+
+% %5
+% The following example illustrates
+% the essence of abstract compilation's effect:
+% \[
+% \mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
+% \]
+% makes the following transitions:
+% \begin{align}
+% & \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
+% \machstep\; &
+% \ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_1}}
+% \\
+% \machstep\; &
+% \ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_2}}
+% \\
+% \machstep\; &
+% \ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_3}} % {\mexp_2}
+% \\
+% \machstep\; &
+% \spchoice
+% {\co{\kar{\mexp_1,\menv,\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
+% {\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
+% \end{align}
+% where \(
+% \msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
+% \maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
+% \maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).
+
+% Whereas abstract compilation gives us in one step:
+% \begin{align*}
+% \compile{\mexp}^\mcntr(\msto_0)(\menv,\epsilon,\mkont) =
+% \spchoice
+% {\co{\kar{\compile{\mexp_1},\menv,\maddr_3},\saddr{\menv(\mvar)}},\msdiff_4}
+% {\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
+% \end{align*}
+% where
+% \begin{equation*}
+% \msdiff_4 = [\ttuple{\maddr_1}{\setof{\mkont}},
+%              \ttuple{\maddr_2}{\setof{\kar{\compile{\mexp_3},\menv,\maddr_1}}},
+%              \ttuple{\maddr_3}{\setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}}]\text.
+% \end{equation*}
+
+The compilation step converts expressions into functions that expect the other components of the {\tt ev} state.
+%
+Its definition in figure \ref{fig:compile} shows close similarity to the rules for interpreting {\tt ev} states.
+%
+ The next step is to change reduction rules that create {\tt ev} states to instead call these functions.
+%
+ Figure \ref{fig:caam} shows the modified reduction relation.
+%
+ The only change from the previous semantics is that $\evalone$ state construction is replaced by calling the compiled expression.
+%
+ For notational coherence, we write $\lambda^\mcntr(\mathit{args} \ldots)$ for $\lambda(\mathit{args} \ldots, \mcntr)$ and $\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
 \mcntr)$.

### 2013-07-02 paper/paper.tex
Tightened some spacing to get into 12 pages.

@@
 %
+\vspace{-1em}
 \begin{center}
@@
 \end{center}
+\vspace{-1em}
 To avoid the spurious forking and joining, we {\it delay} the non-determinism until and unless it is needed in {\it strict contexts} (such as the guard of an {\tt if}, a called procedure, or a numerical primitive application).
@@
 Doing so collapses these forks and joins into a linear sequence of states:
+\vspace{-1em}
 \begin{center}
@@
 \end{center}
+\vspace{-1em}
 %%
@@
 Throughout this section, we will be taking the parameter $t$ to be the ``current time,'' or the length of the store chain at the beginning of the step.
-
+\vspace{-0.9em}
 \begin{align*}

### 2013-07-05 paper/icfp105-johnson.tex
Small typo.

--- 2013-07-02 paper/paper.tex
@@
 \begin{document}
-
-\conferenceinfo{ICFP '13}{September 25--27, Boston.}
+\exclusivelicense
+
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\conferenceinfo{ICFP~'13}{September 25--27, Boston, MA, USA.}
 \copyrightyear{2013}
-\copyrightdata{[to be supplied]}
+\copyrightdata{978-1-4503-2326-0/13/09}
+\doi{2500365.2500604}
 
@@
 
-%% \category{CR-number}{subcategory}{third-level}
-
-%% \terms
-%% term1, term2
-
-%% \keywords
-%% keyword1, keyword2
+\category{F.3.2}{Semantics of Programming Languages}{Program analysis}
+
+\terms
+Performance, Verification
+
+\keywords
+abstract machines, abstract interpretation
 
@@
 %%
-
-%%
-The primary strength of the approach is that abstract interpreters can be easily derived through a small number of steps from existing machine models.
-%
-Since the relationships between abstract machines and higher-level semantic models---such as definitional interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from these high-level semantic models to sound program analyzers in a systematic way.
-%
-Moreover, since these analyses so closely resemble a language's interpreter
-%
-(a) implementing an analysis requires little more than implementing an interpreter,
-%
-(b) a single implementation can serve as both an interpreter and analyzer, and
-%
-(c) verifying the correctness of the implementation is straightforward.
-%%
-
-%%
-Unfortunately, the AAM approach yields analyzers with poor performance relative to hand-optimized analyzers.
-%
-Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an equally systematic way through a number of simple steps, many of which are inspired by run-time implementation techniques such as laziness and compilation to avoid interpretative overhead.
-%
-Each of these steps is proven correct, so the end result is an implementation that is trustworthy and efficient.
-%%
-
-%%
-In this article, we develop a systematic approach to deriving a practical implementation of an abstract-machine-based analyzer using mostly semantic means rather than tricky engineering.
-%
-Our goal is to empower programming language implementers and researchers to explore and convincingly exhibit their ideas with a low barrier to entry.
-%
-The optimizations we describe are widely applicable and apparently effective to scale far beyond the size of programs typically considered in the recent literature on flow analysis for functional languages.
-
-\section{At a glance}
 
@@
 \end{figure}
+
+%%
+The primary strength of the approach is that abstract interpreters can be easily derived through a small number of steps from existing machine models.
+%
+Since the relationships between abstract machines and higher-level semantic models---such as definitional interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from these high-level semantic models to sound program analyzers in a systematic way.
+%
+Moreover, since these analyses so closely resemble a language's interpreter
+%
+(a) implementing an analysis requires little more than implementing an interpreter,
+%
+(b) a single implementation can serve as both an interpreter and analyzer, and
+%
+(c) verifying the correctness of the implementation is straightforward.
+%%
+
+%%
+Unfortunately, the AAM approach yields analyzers with poor performance relative to hand-optimized analyzers.
+%
+Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an equally systematic way through a number of simple steps, many of which are inspired by run-time implementation techniques such as laziness and compilation to avoid interpretative overhead.
+%
+Each of these steps is proven correct, so the end result is an implementation that is trustworthy and efficient.
+%%
+
+%%
+In this article, we develop a systematic approach to deriving a practical implementation of an abstract-machine-based analyzer using mostly semantic means rather than tricky engineering.
+%
+Our goal is to empower programming language implementers and researchers to explore and convincingly exhibit their ideas with a low barrier to entry.
+%
+The optimizations we describe are widely applicable and apparently effective to scale far beyond the size of programs typically considered in the recent literature on flow analysis for functional languages.
+
+\section{At a glance}
 
@@
 We thank Sam Tobin-Hochstadt for encouragement and feedback---he was
-the first to prompt us to look into how make effective
+the first to prompt us to look into how to make effective
 implementations of the AAM approach.
